### 2020-07-01 writing/discussion.tex
paper, the first blood

\section{What We Learned}

\begin{itemize}
  \item Interesting discoveries...?

  \item Comparison of existing GT systems represented in our study.  
    Natural is here, transient is here, erasure is here, and they
        compare as...

      \item Threats to validity:
        \begin{itemize}
          \item Racket stack traces
          \item Mutant sampling
          \item Blame trails do not necessarily capture all aspects of error reporting
          \item Transient blame adaptations
          \item More?
        \end{itemize}
\end{itemize}

### 2020-07-01 writing/experiment.tex
paper, the first blood

\section{The Experimental Setup}

\begin{itemize}

\item Summary of the blame trail experiment. We sample and follow blame trails through configuration
        lattices.

\item Evaluating each point in the design space. We run the blame trail experiment for every point in the
        previously discussed design space.

\item Selection of programs. We are using a set of representative
  programs. Argue the programs are representative of the range of programs written by TR developers; point to the JFP paper.

\item Sampling of mutants. We have too many mutants to run them all, so we sample them with an aim to represent equally all of the mistakes captured by our operators.

\end{itemize}

### 2020-07-01 writing/ideas.tex
paper, the first blood

\section{How to Evaluate Blame}

\begin{itemize}

    \item How to analyze this space?
          \begin{itemize}
            \item We want a quantative measure of error msg utility that allows
                  us to evaluate each point in the above space.
          \end{itemize}

     \item Blame trails
       \begin{itemize}

        \item We want an empirical study to understand how the behavior of blame
        in real systems.

        \item Blame trails provide a quantitative measure of a a system's
        error message utility, and they can be applied to every point
        in the above space.

       \item We generalize the blame trails idea of Lukas et al [POPL '20] to make it
        more systematic, scalable, and general, by:
           \begin{itemize}
             \item Adding a null hypothesis baseline for comparison between multiple systems
             \item Analyzing multiple blame following strategies
             \item Mutating interfaces as well as programs
             \item Incorporating domain knowledge in the selection of blame following
           strategies and mutation
             \item Evaluating systems on the same platform for apples-to-apples comparisons
           \end{itemize}

         \item Null hypothesis. We should have a baseline to tell how all of the points in the
        design space compare to having nothing at all.
\end{itemize}

### 2020-07-01 writing/introduction.tex
paper, the first blood

\section{Does Blame Matter?}

A spectre is haunting gradual typing --- the spectre of blame.  All works
on gradual typing incantate spells either to incarnate blame as they see
fit or to avoid facing it.  But unfortunately, despite all the effort that
has gone to trap blame with models and theorems, its soul remains elusive.

Take for instance the so-called blame theorem~\cite{mf-toplas-2009,
tf-dls-2006, wf-esop-2009}. It reveals blame as an essential indicator of
gradual type soundness; sound gradual type systems never blame typed code.
On one hand, this helps language designers benefit from blame to
(partially) validate their gradual type systems~\cite{gf-icfp-2018}. On
the other hand though, it reduces blame to a proof apparatus.  Worse,
projecting the theorem to a lesson for the programmer leads to misleading
conclusions about where the bug is. The problem  may not be with the
blamed untyped code but with one of the annotations of the typed code.
Similarly, the recent complete monitoring framework~\cite{gfd-oopsla-2019}
sheds limited light to the pragmatic nature of blame for gradual types. It
shows that blame captures some aspects of the provenance of the value that
fails a runtime type check, but again this explanation falls short of
tangible information for the programmer. In effect programmers are left
without any guidance about how to turn information from blame to the
location of bugs.  And this current lack of a principled connection
between blame and bugs ---  the raison d'\^etre of blame --- puts in
question the value of blame for gradual types.

This paper aims to answer a single question: \emph{Does blame add value to
gradual typing?} We posit that blame does add value to gradual typing only
if it conveys to programmers extra information that is useful for
debugging. Specifically, for blame to be valuable there needs to be a
systematic relation between blame and the location of bugs.

To examine the truth of this condition, we follow recent work on the
pragmatics of blame for contracts~\cite{lksfd-popl-2020}. First, we stray
from the usual theoretical treatment of blame in the gradual typing
literature and we investigate the value of blame with an empirical
methodology. Its empirical nature gives us the opportunity to investigate how blame copes
against actual bugs in real implementations of gradual type systems.

Second, the central piece of the methodology is a connection between blame
and bugs inspired straight from the experience of programming with gradual
types.  When facing a run time type error in a mix-typed program, programmers often try to add
type annotations to their programs hoping to turn the  run time type error
into a static one that reveals where the code and its types clash.  In
this setting, blame can suggest what part of the program to migrate from
untyped to typed next.  Therefore, if heeding blame culminates to a static type
error, blame and the location of the bug are directly connected.  Based on
this observation, our methodology explores the migration of mix-typed programs with
run time type errors to validate whether blame guides their migration to an
ill-typed  version that pinpoints the location of the
bug.

Realizing these ideas into a robust experiment requires surpassing three challenges.  First,
gradual typing is a heterogeneous landscape. Herein, we focus on three
points of the landscape: the natural approach of Typed
Racket~\cite{tf-dls-2006,tf-popl-2008,tfffgksst-snapl-2017,tf-icfp-2010},
the transient one of Reticulated
Python~\cite{vsc-dls-2019,vss-popl-2017,vksb-dls-2014} and  the optional
gradual typing of the likes of Typescript. Each of these points comes with
a distinct blaming strategy and together they represent all proposed
strategies so far. In addition, instead of experimenting with the three
approaches in isolation, we have implemented transient and optional
gradual typing as variants of  Typed Racket to seize the opportunity to
compare them.  However, implementations on top of the same platform are
not enough for an apples-to-apples comparison. The three approaches differ
not only on their blaming strategies but also on the type checks they
perform. To untangle the two, we supplement our experiment with two extra
modes. The \emph{vanilla} mode  ignores blame and instead uses information
from dynamic type error messages to select how to proceed with migration. Hence it
separates the effect of blame on the experiment from that of the type
checks. The \emph{control} mode randomly selects which part of the program to migrate
and serves as the null hypothesis control for the overall methodology.

The second challenge stems from the lack of a sufficiently large available
corpus of type errors for a meaningful experiment. Similar to
~\citet{lksfd-popl-2020}, we cover this gap by injecting bugs in
typed programs  with mutation analysis~\cite{lipton1971fault,
demillo1978hints, jia2011analysis}. This results into a large 
number of buggy programs whose mix-typed versions are suitable subjects
for our experiment. However, mutation analysis
traditionally  aims to inject bugs that are sophisticated enough to
challenge extensive test suites. As a consequence, the ill-typed mutants we care about here are
usually deemed \emph{incompetent} and mutation analysis frameworks are
fine-tuned to avoid them. To overcome this problem, we design a
set of mutators specific for type errors inspired by common errors when
programming with in Typed Racket.  These type errors cover
a wide range of Typed Racket's type system and moreover cause reliably runtime
type errors  in mix-typed versions of mutants.

The final challenge deals with the fact that bugs in mix-typed programs
occur not only in the code but also in type
annotations~\cite{sta-nt-base-types, incorrect-ts, wmwz-ecoop-2017}. Thus
our experiment needs to also explore the relation between blame and the
location of bugs in case of wrong type annotations. Similarly to the previous challenge, we
deal with this one by designing a set of mutators that ``break'' type
annotations but leave programs otherwise intact.

To sum up, the maim {\bf contribution} of this paper is an empirical  methodology for
evaluating blame for gradual type systems. The methodology applies across the
landscape of gradual typing and allows to compare its different points. 
  As a result, this paper also contributes the first comparative
empirical study of natural, transient and optional gradual typing
with respect to blame. [A couple of sentences that summarize the results
her.]

The remainder of the paper is organized as follows.
Section~\ref{sec:landscape} revisits the landscape of gradual typing from
the perspective of the different checking and blame strategies.
Section~\ref{sec:ideas} presents in detail the main ideas behind our
methodology and section~\ref{sec:experiment} explains the specifics of our
experiment. Section~\ref{sec:results} gathers our results and
section~\ref{sec:discussion} discusses their implications. The paper
concludes with a survey of related work in section~\ref{sec:related} and 
a few final remarks in section~\ref{sec:conclusion}.

### 2020-07-01 writing/landscape.tex
paper, the first blood

\section{The Landscape of Gradual Typing and Blame}

   By example; adjusted version of the one in OOPSLA sec 2

  \begin{itemize}

  \item Why Blame Assignments Differ.

  \item How Blame Assignments Differ.

  \item A space of GT system design. In terms of error reporting, we identify two axes of GT system
        design: where checking occurs and whether blame is tracked.
    \end{itemize}

### 2020-07-01 writing/paper.tex
paper, the first blood

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[acmsmall,screen]{acmart}\settopmatter{printfolios=true,printccs=true,printacmref=true}
%% For double-blind review submission, w/ CCS and ACM Reference
%%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Note: Authors migrating a paper from PACMPL format to traditional
%% SIGPLAN proceedings format must update the '\documentclass' and
%% topmatter commands above; see 'acmart-sigplanproc-template.tex'.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Title information
\newcommand{\thetitle}{How to Evaluate Blame for Gradual Types}
\title{\thetitle}

%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

%\author[]{}
%\affiliation{
%\institution{}
%% \city{}
%% \city{}
% \country{}
%}
%\email{}          %% \email is recommended

\usepackage{alltt}
\usepackage{wrapfig}

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption

\usepackage{listings}
\usepackage[shortcuts]{extdash}
\usepackage{xcolor}

\usepackage{upgreek}

\input{def.tex}

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
%% \setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations

%%%
\setcopyright{}
\acmPrice{}
\acmDOI{}
\acmYear{}
\copyrightyear{}
\acmJournal{}
\acmVolume{}
\acmNumber{}
\acmArticle{}
\acmMonth{}

\begin{document}

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Gradual types and blame go hand in hand. But what is the value of blame?
\end{abstract}

% CCS concepts: Program specifications < Program reasoning < Semantics and Reasoning < Theory of Computation
% CCS concepts: Empirical software validation < Software verification and validation < Software creation and management < Software and its engineering

% potential choices decided against: pre- and post-conditions, assertions, software testing and debugging

%\begin{CCSXML}
%<ccs2012>
%<concept>
%<concept_id></concept_id>
%<concept_desc></concept_desc>
%<concept_significance></concept_significance>
%</concept>
%<concept>
%<concept_id></concept_id>
%<concept_desc></concept_desc>
%<concept_significance></concept_significance>
%</concept>
%</ccs2012>
%\end{CCSXML}

%\ccsdesc[]{}
%\ccsdesc[]{}

%% Keywords
%% comma separated list
%\keywords{}  %% \keywords are mandatory in final camera-ready submission

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle

\input{introduction.tex} \label{sec:introduction}
\input{landscape.tex} \label{sec:landscape}
\input{ideas.tex} \label{sec:ideas}
\input{experiment.tex} \label{sec:experiment}
\input{results.tex} \label{sec:results}
\input{discussion.tex} \label{sec:discussion}
\input{related.tex} \label{sec:related}
\input{conclusion.tex} \label{sec:conclusion}

%% Acknowledgments
%%\begin{acks}                 
  %% acks environment is optional
  %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
%%\end{acks}

%% Bibliography
\bibliography{cs}

\end{document}

### 2020-07-01 writing/related.tex
paper, the first blood

\section{Related Work}

\begin{itemize}
  \item  GT: systems we don't represent
  \item Mutation testing
  \itme Fault localization
\end{itemize}

### 2020-07-01 writing/challenges.tex
split introduction

--- 2020-07-01 writing/introduction.tex
@@
-\section{Does Blame Matter?}
+\section{Challenges}
 
-A spectre is haunting gradual typing --- the spectre of blame.  All works
-on gradual typing incantate spells either to incarnate blame as they see
-fit or to avoid facing it.  But unfortunately, despite all the effort that
-has gone to trap blame with models and theorems, its soul remains elusive.
-
-Take for instance the so-called blame theorem~\cite{mf-toplas-2009,
-tf-dls-2006, wf-esop-2009}. It reveals blame as an essential indicator of
-gradual type soundness; sound gradual type systems never blame typed code.
-On one hand, this helps language designers benefit from blame to
-(partially) validate their gradual type systems~\cite{gf-icfp-2018}. On
-the other hand though, it reduces blame to a proof apparatus.  Worse,
-projecting the theorem to a lesson for the programmer leads to misleading
-conclusions about where the bug is. The problem  may not be with the
-blamed untyped code but with one of the annotations of the typed code.
-Similarly, the recent complete monitoring framework~\cite{gfd-oopsla-2019}
-sheds limited light to the pragmatic nature of blame for gradual types. It
-shows that blame captures some aspects of the provenance of the value that
-fails a runtime type check, but again this explanation falls short of
-tangible information for the programmer. In effect programmers are left
-without any guidance about how to turn information from blame to the
-location of bugs.  And this current lack of a principled connection
-between blame and bugs ---  the raison d'\^etre of blame --- puts in
-question the value of blame for gradual types.
-
-This paper aims to answer a single question: \emph{Does blame add value to
-gradual typing?} We posit that blame does add value to gradual typing only
-if it conveys to programmers extra information that is useful for
-debugging. Specifically, for blame to be valuable there needs to be a
-systematic relation between blame and the location of bugs.
-
-To examine the truth of this condition, we follow recent work on the
-pragmatics of blame for contracts~\cite{lksfd-popl-2020}. First, we stray
-from the usual theoretical treatment of blame in the gradual typing
-literature and we investigate the value of blame with an empirical
-methodology. Its empirical nature gives us the opportunity to investigate how blame copes
-against actual bugs in real implementations of gradual type systems.
-
-Second, the central piece of the methodology is a connection between blame
-and bugs inspired straight from the experience of programming with gradual
-types.  When facing a run time type error in a mix-typed program, programmers often try to add
-type annotations to their programs hoping to turn the  run time type error
-into a static one that reveals where the code and its types clash.  In
-this setting, blame can suggest what part of the program to migrate from
-untyped to typed next.  Therefore, if heeding blame culminates to a static type
-error, blame and the location of the bug are directly connected.  Based on
-this observation, our methodology explores the migration of mix-typed programs with
-run time type errors to validate whether blame guides their migration to an
-ill-typed  version that pinpoints the location of the
-bug.
+Creating a method for comparing blame assignment strategies poses
+several severe challenges.
 

### 2020-07-01 writing/introduction.tex
split introduction

@@
 
-A spectre is haunting gradual typing --- the spectre of blame.  All works
-on gradual typing incantate spells either to incarnate blame as they see
-fit or to avoid facing it.  But unfortunately, despite all the effort that
-has gone to trap blame with models and theorems, its soul remains elusive.
+Since its very beginning [cite STH], the theory of gradual typing
+comes with a blame theorem. ``Well-typed [components]\footnote{The
+original authors got this word wrong, and more.} can't go wrong'' has
+become the slogan. The idea is that blame assignment can help the
+practicing programmers find impedance mismatches between the types
+they added and the remaining untyped components.
 
-Take for instance the so-called blame theorem~\cite{mf-toplas-2009,
-tf-dls-2006, wf-esop-2009}. It reveals blame as an essential indicator of
-gradual type soundness; sound gradual type systems never blame typed code.
-On one hand, this helps language designers benefit from blame to
-(partially) validate their gradual type systems~\cite{gf-icfp-2018}. On
-the other hand though, it reduces blame to a proof apparatus.  Worse,
-projecting the theorem to a lesson for the programmer leads to misleading
-conclusions about where the bug is. The problem  may not be with the
-blamed untyped code but with one of the annotations of the typed code.
-Similarly, the recent complete monitoring framework~\cite{gfd-oopsla-2019}
-sheds limited light to the pragmatic nature of blame for gradual types. It
-shows that blame captures some aspects of the provenance of the value that
-fails a runtime type check, but again this explanation falls short of
-tangible information for the programmer. In effect programmers are left
-without any guidance about how to turn information from blame to the
-location of bugs.  And this current lack of a principled connection
-between blame and bugs ---  the raison d'\^etre of blame --- puts in
-question the value of blame for gradual types.
+Practical gradual type systems ignore blame completely. Industrial
+systems [lots of citations. footnotes] exploit types for assistance
+with IDE actions and for finding typos in code. Then they remove them
+and rely on the built-in safety checks of the underlying programming
+language to catch problems.
 
-This paper aims to answer a single question: \emph{Does blame add value to
-gradual typing?} We posit that blame does add value to gradual typing only
-if it conveys to programmers extra information that is useful for
-debugging. Specifically, for blame to be valuable there needs to be a
-systematic relation between blame and the location of bugs.
+This contrast raises the question whether
 
-To examine the truth of this condition, we follow recent work on the
-pragmatics of blame for contracts~\cite{lksfd-popl-2020}. First, we stray
-from the usual theoretical treatment of blame in the gradual typing
-literature and we investigate the value of blame with an empirical
-methodology. Its empirical nature gives us the opportunity to investigate how blame copes
-against actual bugs in real implementations of gradual type systems.
+     blame adds value to gradual typing
+     and especially the working programmer.
 
-Second, the central piece of the methodology is a connection between blame
-and bugs inspired straight from the experience of programming with gradual
-types.  When facing a run time type error in a mix-typed program, programmers often try to add
-type annotations to their programs hoping to turn the  run time type error
-into a static one that reveals where the code and its types clash.  In
-this setting, blame can suggest what part of the program to migrate from
-untyped to typed next.  Therefore, if heeding blame culminates to a static type
-error, blame and the location of the bug are directly connected.  Based on
-this observation, our methodology explores the migration of mix-typed programs with
-run time type errors to validate whether blame guides their migration to an
-ill-typed  version that pinpoints the location of the
-bug.
+To obtain an answer, we have developed a method (not methodology!)  to
+evaluate blame assignment systems from the gradual typing
+world. Roughly speaking, the method simulates a programmer who follows
+the above-cited Wadler-Findler slogan. When the run-time system
+signals a violation, the programmer conjectures that the addition of 
+types to untyped components turns the dynamic violation into a static
+type error and thus reveals the impedance mismatch.
 
-Realizing these ideas into a robust experiment requires surpassing three challenges.  First,
-gradual typing is a heterogeneous landscape. Herein, we focus on three
-points of the landscape: the natural approach of Typed
-Racket~\cite{tf-dls-2006,tf-popl-2008,tfffgksst-snapl-2017,tf-icfp-2010},
-the transient one of Reticulated
-Python~\cite{vsc-dls-2019,vss-popl-2017,vksb-dls-2014} and  the optional
-gradual typing of the likes of Typescript. Each of these points comes with
-a distinct blaming strategy and together they represent all proposed
-strategies so far. In addition, instead of experimenting with the three
-approaches in isolation, we have implemented transient and optional
-gradual typing as variants of  Typed Racket to seize the opportunity to
-compare them.  However, implementations on top of the same platform are
-not enough for an apples-to-apples comparison. The three approaches differ
-not only on their blaming strategies but also on the type checks they
-perform. To untangle the two, we supplement our experiment with two extra
-modes. The \emph{vanilla} mode  ignores blame and instead uses information
-from dynamic type error messages to select how to proceed with migration. Hence it
-separates the effect of blame on the experiment from that of the type
-checks. The \emph{control} mode randomly selects which part of the program to migrate
-and serves as the null hypothesis control for the overall methodology.
+The paper presents the evaluation method and the results of applying
+the method to three distinct blame assignment systems.
 
-The second challenge stems from the lack of a sufficiently large available
-corpus of type errors for a meaningful experiment. Similar to
-~\citet{lksfd-popl-2020}, we cover this gap by injecting bugs in
-typed programs  with mutation analysis~\cite{lipton1971fault,
-demillo1978hints, jia2011analysis}. This results into a large 
-number of buggy programs whose mix-typed versions are suitable subjects
-for our experiment. However, mutation analysis
-traditionally  aims to inject bugs that are sophisticated enough to
-challenge extensive test suites. As a consequence, the ill-typed mutants we care about here are
-usually deemed \emph{incompetent} and mutation analysis frameworks are
-fine-tuned to avoid them. To overcome this problem, we design a
-set of mutators specific for type errors inspired by common errors when
-programming with in Typed Racket.  These type errors cover
-a wide range of Typed Racket's type system and moreover cause reliably runtime
-type errors  in mix-typed versions of mutants.
+- explain the systems
+- explain the method
+- preview the results
 
-The final challenge deals with the fact that bugs in mix-typed programs
-occur not only in the code but also in type
-annotations~\cite{sta-nt-base-types, incorrect-ts, wmwz-ecoop-2017}. Thus
-our experiment needs to also explore the relation between blame and the
-location of bugs in case of wrong type annotations. Similarly to the previous challenge, we
-deal with this one by designing a set of mutators that ``break'' type
-annotations but leave programs otherwise intact.
-
-To sum up, the maim {\bf contribution} of this paper is an empirical  methodology for
-evaluating blame for gradual type systems. The methodology applies across the
-landscape of gradual typing and allows to compare its different points. 
-  As a result, this paper also contributes the first comparative
-empirical study of natural, transient and optional gradual typing
-with respect to blame. [A couple of sentences that summarize the results
-her.]
-
-The remainder of the paper is organized as follows.
-Section~\ref{sec:landscape} revisits the landscape of gradual typing from
-the perspective of the different checking and blame strategies.
-Section~\ref{sec:ideas} presents in detail the main ideas behind our
-methodology and section~\ref{sec:experiment} explains the specifics of our
-experiment. Section~\ref{sec:results} gathers our results and
-section~\ref{sec:discussion} discusses their implications. The paper
-concludes with a survey of related work in section~\ref{sec:related} and 
-a few final remarks in section~\ref{sec:conclusion}.
+- weave roadmap into these paragraphs

### 2020-07-01 writing/introduction.tex
first draft of introduction

@@
 
-Since its very beginning [cite STH], the theory of gradual typing
-comes with a blame theorem. ``Well-typed [components]\footnote{The
-original authors got this word wrong, and more.} can't go wrong'' has
-become the slogan. The idea is that blame assignment can help the
-practicing programmers find impedance mismatches between the types
-they added and the remaining untyped components.
+From the very beginning, the theory of gradual typing has come with a blame
+theorem~\cite{mf-toplas-2009, tf-dls-2006}. ``Well-typed [components]\footnote{The
+original authors got this word wrong.} can't go wrong'' turned the theorem into a
+slogan. Indeed, reasonably practical academic systems---for example,, Reticulated
+Python~\cite{vsc-dls-2019, vss-popl-2017, vksb-dls-2014} and Typed
+Racket~\cite{tf-dls-2006,tf-popl-2008,tfffgksst-snapl-2017,tf-icfp-2010}---come
+with (different) blame assignment strategies. The idea is that blame assignment
+can help the practicing programmers find impedance mismatches between the types
+they added and the behavior remaining untyped components.
 
-Practical gradual type systems ignore blame completely. Industrial
-systems [lots of citations. footnotes] exploit types for assistance
-with IDE actions and for finding typos in code. Then they remove them
-and rely on the built-in safety checks of the underlying programming
-language to catch problems.
+Practical gradual type systems ignore blame assignment almost completely.
+Industrial systems such as Flow,\footnote{flow} Hack,\footnote{hack} or
+TypeScript~\cite{ts} exploit types for assistance with IDE actions and for finding
+typos in code. Their compilers remove types and rely on the built-in safety checks
+of the underlying programming language to catch problems.
 
-This contrast raises the question whether
+The contrast between academic and industrial systems raises the question whether
+\begin{quote}
+ \it
+ blame assignment adds any value to a gradual typing system, \\
+ especially for the working programmer.
+\end{quote}
+Given the prominence of the idea and the complete absence of blame in industrial
+systems, it comes as an even bigger surprise that the research literature and the
+industrial blog world do not present or discuss any possible answers.  Instead,
+language designers seem to answer this question one way or another without any
+scientific justification, and even when they have a language, they neglect to look
+into this question. The research community simply does not have a method for
+evaluating such questions, and given the question's practical relevance, this is a
+problem.
 
-     blame adds value to gradual typing
-     and especially the working programmer.
+This paper's first contribution is a method for evaluating the effectiveness of
+blame assignment strategies in the gradual typing world (see
+section~\ref{sec:ideas}). Roughly speaking, the method simulates a programmer who
+follows the Wadler-Findler slogan quoted above. The implied method suggests that
+programmers can find impedance-mismatch bugs by adding types to partially typed
+software system. That is, when the run-time system signals a type-like violation,
+a good blame assignment helps programmers infer where to add an additional type
+specification. Since components that pass the type checker cannot cause run-time
+exceptions; the type checker discovers these mismatches at compile time now, which
+helps the programmer find mistake.  If the type specification implied by the blame
+message does {\em not\/} cause a static type error, the program is run again until
+it signals another violation.  Counting the number of steps it takes to discover
+the mistake is a measure of the effectiveness of the blame assignment.  The method
+is loosely based on the work of Lazarek et al. from last year's
+POPL~\cite{lksfd-popl-2020}.
 
-To obtain an answer, we have developed a method (not methodology!)  to
-evaluate blame assignment systems from the gradual typing
-world. Roughly speaking, the method simulates a programmer who follows
-the above-cited Wadler-Findler slogan. When the run-time system
-signals a violation, the programmer conjectures that the addition of 
-types to untyped components turns the dynamic violation into a static
-type error and thus reveals the impedance mismatch.
+The paper's second contribution is a set of results from applying the evaluation
+method to three distinct blame assignment systems and approximately 20,000
+cases\mf{Replace with real number} (see section~\ref{sec:experiment}): Vitousek et
+al.'s strategy of tracking typed/untyped boundaries in Reticulate; Tobin-Hochstadt
+et al.'s use of Findler et al.'s higher-order contract system~\cite{ff-icfp-2002,
+mf-toplas-2009}; and the error messages from failed safety checks in the
+underlying programming language.
 
-The paper presents the evaluation method and the results of applying
-the method to three distinct blame assignment systems.
+\mf{This is made up and must be fixed for the submission.}
 
-- explain the systems
-- explain the method
-- preview the results
+The results of the experiment (see section~\ref{sec:results}) validate the
+conjecture behind the work of theoreticians. A good blame assignment system
+greatly shortens the search for impedance mismatches between the specified types
+and the behavior of untyped components. Even better, the method clearly identifies
+the wrapper-based blame tracking of Typed Racket as the vastly superior academic
+system. Vitousek et al. lose one more time. Of course, a detailed discussion 
+must acknowledge the threats to validity of these results (see
+section~\ref{sec:discussion}).
 
-- weave roadmap into these paragraphs
+The nest two sections provide an overview of the blame assignment strategies as
+found in existing gradually typed language implementation (see
+section~\ref{sec:landscape}) and the challenges of developing a systematic
+evaluation method (section~\ref{sec:challengs}).

### 2020-07-01 writing/introduction.tex
revisions to the introduction

--- 2020-07-01 writing/introduction.tex
@@
 
-Since its very beginning [cite STH], the theory of gradual typing
-comes with a blame theorem. ``Well-typed [components]\footnote{The
-original authors got this word wrong, and more.} can't go wrong'' has
-become the slogan. The idea is that blame assignment can help the
-practicing programmers find impedance mismatches between the types
-they added and the remaining untyped components.
+From the very beginning, the theory of gradual typing has come with a blame
+theorem~\cite{mf-toplas-2009, tf-dls-2006}. ``Well-typed [components]\footnote{The
+original authors got this word wrong.} can't be blamed'' turned the theorem into a
+slogan~\cite{wf-esop-2009}. Indeed, academic implementationsa---for example,,
+Reticulated Python~\cite{vsc-dls-2019, vss-popl-2017, vksb-dls-2014} and Typed
+Racket~\cite{tf-dls-2006,tf-popl-2008,tfffgksst-snapl-2017,tf-icfp-2010}---come
+with sophisticated blame-assignment subsystems. Theoreticians firmly embrace the
+idea that blame can help practicing programmers find impedance mismatches between
+the types they added to a software system and the behavior remaining untyped
+components.
 
-Practical gradual type systems ignore blame completely. Industrial
-systems [lots of citations. footnotes] exploit types for assistance
-with IDE actions and for finding typos in code. Then they remove them
-and rely on the built-in safety checks of the underlying programming
-language to catch problems.
+Practical gradual type systems ignore blame assignment almost completely.
+Industrial systems such as Flow,\footnote{flow} Hack,\footnote{hack} or
+TypeScript~\cite{ts} exploit types for IDE actions and for finding typos in
+code. Then these compilers remove types and rely on the built-in safety checks of
+the underlying programming language to catch any problems.
 
-This contrast raises the question whether
+The contrast between academic and industrial systems raises the question whether
+\begin{quote}
+ \it
+ blame assignment adds any value to a gradually typed language, \\
+ especially for the benefit of the working programmer.
+\end{quote}
+Given the prominence of blame in academia and its complete absence in
+industrial systems, it comes as an even bigger surprise that the research
+literature and the industrial blog world do not present or discuss any possible
+answers.  Instead, language designers seem to answer this question one way or
+another without any scientific justification, and even when they have a language,
+they neglect to look into this question. The research community simply does not
+have a method for evaluating such questions, and given the question's practical
+relevance, this is a problem.
 
-     blame adds value to gradual typing
-     and especially the working programmer.
+This paper's first contribution is an automated method for evaluating the
+effectiveness of blame assignment strategies in the gradual typing world (see
+section~\ref{sec:ideas}). Roughly speaking, the method simulates a programmer who
+follows the Wadler-Findler slogan quoted above. Its implied method suggests that
+programmers can find impedance-mismatch bugs by adding types to a partially typed
+software system. That is, when the run-time system signals an impedance mismatch,
+a good blame messages helps programmers infer where to add an additional type
+specification. Since components that pass the type checker cannot cause such
+impedance signals; the type checker must discover these mismatches at compile
+time.  If the additional type does {\em not\/} cause a static type error, the
+program is run again until it signals another violation.  Counting the number of
+steps it takes to discover the mistake is a measure of the effectiveness of the
+blame assignment.  The method is loosely based on the work of Lazarek et
+al.~\cite{lksfd-popl-2020}.
 
-To obtain an answer, we have developed a method (not methodology!)  to
-evaluate blame assignment systems from the gradual typing
-world. Roughly speaking, the method simulates a programmer who follows
-the above-cited Wadler-Findler slogan. When the run-time system
-signals a violation, the programmer conjectures that the addition of 
-types to untyped components turns the dynamic violation into a static
-type error and thus reveals the impedance mismatch.
+The paper's second contribution is a set of results from applying the evaluation
+method to three distinct blame assignment systems and approximately 20,000
+cases\mf{Replace with real number} (see section~\ref{sec:experiment}): Vitousek et
+al.'s strategy of tracking typed/untyped boundaries in Reticulated;
+Tobin-Hochstadt et al.'s use of higher-order contract system~\cite{ff-icfp-2002,
+mf-toplas-2009}; and the error messages from failed safety checks in the
+underlying programming language.
 
-The paper presents the evaluation method and the results of applying
-the method to three distinct blame assignment systems.
+\mf{This is made up and must be fixed for the submission.}
 
-- explain the systems
-- explain the method
-- preview the results
+The results of the experiment (see section~\ref{sec:results}) validate the
+conjecture behind the work of theoreticians. A good blame assignment system
+greatly shortens the search for impedance mismatches between the specified types
+and the behavior of untyped components. Even better, the evaluation method clearly
+identifies the wrapper-based blame tracking of Typed Racket as the vastly superior
+academic system. Vitousek et al. lose one more time. Of course, a detailed
+discussion must acknowledge the threats to validity of these results (see
+section~\ref{sec:discussion}).
 
-- weave roadmap into these paragraphs
+Next, the paper provide an overview of the blame assignment strategies as found in
+existing gradually typed language implementation (see section~\ref{sec:landscape})
+and the challenges of developing a systematic evaluation method
+(section~\ref{sec:challengs}).

### 2020-07-01 writing/paper.tex
fix basic latex mistakes in rudimentary way
no ACM footer to make space for real content
abstract

@@
 %% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
-\documentclass[acmsmall,screen]{acmart}\settopmatter{printfolios=true,printccs=true,printacmref=true}
+\documentclass[acmsmall,screen,anonymous,nonacm]{acmart}\settopmatter{printfolios=true,printccs=true,printacmref=true}
 %% For double-blind review submission, w/ CCS and ACM Reference
@@
 
-%\author[]{}
-%\affiliation{
-%\institution{}
-%% \city{}
-%% \city{}
-% \country{}
-%}
-%\email{}          %% \email is recommended
+\author{Nobody In Particular}
+% \affiliation{\institution{PLT} \city{Evanston} \country{Illinois} }
+\affiliation{\relax} % \institution{PLT} \city{Evanston} \country{Illinois} }
+% \email{chrdim@northwestern.edu}          %% \email is recommended
 
@@
 %%%
-\setcopyright{}
-\acmPrice{}
+\setcopyright{none}
+\acmPrice{1}
 \acmDOI{}
@@
 \copyrightyear{}
-\acmJournal{}
-\acmVolume{}
-\acmNumber{}
-\acmArticle{}
-\acmMonth{}
+\acmJournal{pacmpl}
+\acmVolume{1}
+\acmNumber{1}
+\acmArticle{1}
+\acmMonth{1}
 
@@
 \begin{abstract}
-Gradual types and blame go hand in hand. But what is the value of blame?
+Programming language theoreticians develop blame assignment systems
+and prove blame theorems for gradually typed programming languages.
+Practical implementations of gradual typing almost completely ignore
+the idea of blame assignment.  This contrast raises the question
+whether blame assignment provides any value to the working programmer
+and poses the challenge of how to evaluate the effectiveness of blame
+assignment systems. This paper contributes (1) the first evaluation
+method for blame assignment systems and (2) the results from applying
+it to three different systems. \mf{This sentence must be checked
+before final submission.} These results confirm that blame assignment
+is helpful to a rationally acting programmer.
 \end{abstract}
@@
 \input{landscape.tex}    \label{sec:landscape}
+\input{challenges.tex}   \label{sec:challengs}
 \input{ideas.tex}        \label{sec:ideas}

### 2020-07-01 writing/introduction.tex
challenges first

@@
 
-Next, the paper provide an overview of the blame assignment strategies as found in
-existing gradually typed language implementation (see section~\ref{sec:landscape})
-and the challenges of developing a systematic evaluation method
-(section~\ref{sec:challengs}).
+Next, the paper describes the challenges of developing an evaluation method
+(section~\ref{sec:challengs}), which may explain why the question has been neglected
+for such a long time. This description is followed by some background material on the
+existing blame assignment strategies (see section~\ref{sec:landscape}).

### 2020-07-01 writing/challenges.tex
organize the challenge section

@@
 
-Creating a method for comparing blame assignment strategies poses
-several severe challenges.
+Creating a method for comparing blame assignment strategies poses several severe
+challenges.
 
-Realizing these ideas into a robust experiment requires surpassing three challenges.  First,
-gradual typing is a heterogeneous landscape. Herein, we focus on three
-points of the landscape: the natural approach of Typed
-Racket~\cite{tf-dls-2006,tf-popl-2008,tfffgksst-snapl-2017,tf-icfp-2010},
-the transient one of Reticulated
-Python~\cite{vsc-dls-2019,vss-popl-2017,vksb-dls-2014} and  the optional
-gradual typing of the likes of Typescript. Each of these points comes with
-a distinct blaming strategy and together they represent all proposed
-strategies so far. In addition, instead of experimenting with the three
-approaches in isolation, we have implemented transient and optional
-gradual typing as variants of  Typed Racket to seize the opportunity to
-compare them.  However, implementations on top of the same platform are
-not enough for an apples-to-apples comparison. The three approaches differ
-not only on their blaming strategies but also on the type checks they
-perform. To untangle the two, we supplement our experiment with two extra
-modes. The \emph{vanilla} mode  ignores blame and instead uses information
-from dynamic type error messages to select how to proceed with migration. Hence it
-separates the effect of blame on the experiment from that of the type
-checks. The \emph{control} mode randomly selects which part of the program to migrate
-and serves as the null hypothesis control for the overall methodology.
+The most serious challenge concerns the nature of this research. Unlike ordinary
+work on programming languages, the question seems to call for research similar
+to the one in the human-computer interaction area. At the same time, a
+significant evaluation result demands a good number of cases yielding a serious
+amount of data. But, as~\citet{lksfd-popl-2020} already demonstrated, the way
+around this dilemma is to postulate the equivalent of a {\it home economicus\/}
+for programming languages---the {\em rational programmer\/}---and to implement
+this rational programmer as an algorithm. As indicated in the introduction, in
+the context of gradual typing a rationally acting programmer translates the
+Wadler-Findler slogan into a debugging method, searching for the source of the
+impedance mismatch in an incremental fashion.
 
-The second challenge stems from the lack of a sufficiently large available
-corpus of type errors for a meaningful experiment. Similar to
-~\citet{lksfd-popl-2020}, we cover this gap by injecting bugs in
-typed programs  with mutation analysis~\cite{lipton1971fault,
-demillo1978hints, jia2011analysis}. This results into a large 
-number of buggy programs whose mix-typed versions are suitable subjects
-for our experiment. However, mutation analysis
-traditionally  aims to inject bugs that are sophisticated enough to
-challenge extensive test suites. As a consequence, the ill-typed mutants we care about here are
-usually deemed \emph{incompetent} and mutation analysis frameworks are
-fine-tuned to avoid them. To overcome this problem, we design a
+One derivative challenge to the first is the problem of simulating the
+programmer's mistakes. Before a programmer can try to find mistakes, mistakes
+have to be made. Technically the problem is the lack of a sufficiently large
+available corpus of type errors for a meaningful experiment. Similar to
+~\citet{lksfd-popl-2020}, we cover this gap by injecting bugs in typed programs
+with mutation analysis~\cite{lipton1971fault, demillo1978hints,
+jia2011analysis}. This results into a large number of buggy programs whose
+mix-typed versions are suitable subjects for our experiment. However, mutation
+analysis traditionally aims to inject bugs that are sophisticated enough to
+challenge extensive test suites. As a consequence, the ill-typed mutants we care
+about here are usually deemed \emph{incompetent} and mutation analysis
+frameworks are fine-tuned to avoid them. To overcome this problem, we design a
 set of mutators specific for type errors inspired by common errors when
-programming with in Typed Racket.  These type errors cover
-a wide range of Typed Racket's type system and moreover cause reliably runtime
-type errors  in mix-typed versions of mutants.
+programming with in Typed Racket.  These type errors cover a wide range of Typed
+Racket's type system and moreover cause reliably runtime type errors in
+mix-typed versions of mutants.
 
-The final challenge deals with the fact that bugs in mix-typed programs
-occur not only in the code but also in type
-annotations~\cite{sta-nt-base-types, incorrect-ts, wmwz-ecoop-2017}. Thus
-our experiment needs to also explore the relation between blame and the
-location of bugs in case of wrong type annotations. Similarly to the previous challenge, we
-deal with this one by designing a set of mutators that ``break'' type
-annotations but leave programs otherwise intact.
+Another derivative challenge is the problem of mistakes in type specifications.
+Gradual typing enables programmers to move an untyped code base into the typed
+realm on an incremental basis. The retroactive addition of a specific type to a
+component, even a library, may result from a misunderstanding of the code and
+may thus be a mistake itself. Empirical evidence suggests that this scenario is
+reasonably common~\cite{sta-nt-base-types, incorrect-ts, wmwz-ecoop-2017}, and
+hence must be addressed by any evaluation method.  \ldots needs to also explore
+the relation between blame and the location of bugs in case of wrong type
+annotations. Similarly to the previous challenge, we deal with this one by
+designing a set of mutators that ``break'' type annotations but leave programs
+otherwise intact.
 
-To sum up, the maim {\bf contribution} of this paper is an empirical  methodology for
-evaluating blame for gradual type systems. The methodology applies across the
-landscape of gradual typing and allows to compare its different points. 
-  As a result, this paper also contributes the first comparative
-empirical study of natural, transient and optional gradual typing
-with respect to blame. [A couple of sentences that summarize the results
-her.]
-
-The remainder of the paper is organized as follows.
-Section~\ref{sec:landscape} revisits the landscape of gradual typing from
-the perspective of the different checking and blame strategies.
-Section~\ref{sec:ideas} presents in detail the main ideas behind our
-methodology and section~\ref{sec:experiment} explains the specifics of our
-experiment. Section~\ref{sec:results} gathers our results and
-section~\ref{sec:discussion} discusses their implications. The paper
-concludes with a survey of related work in section~\ref{sec:related} and 
-a few final remarks in section~\ref{sec:conclusion}.
+The second big challenge is the set of seemingly incomparable blame-assignment
+strategies. This results in this paper focus on three points of the landscape:
+the {\em natural\/} approach of Typed Racket~\cite{tf-dls-2006, tf-popl-2008,
+tfffgksst-snapl-2017, tf-icfp-2010}, the {\em transient\/} one of Reticulated
+Python~\cite{vsc-dls-2019,vss-popl-2017,vksb-dls-2014}, and the {\em optional\/}
+typing approach of industrial systems. Each of these points comes with a
+distinct blaming strategy and together they represent all proposed strategies so
+far. In addition, instead of experimenting with the three approaches in
+isolation, we have implemented transient and optional gradual typing as variants
+of Typed Racket to seize the opportunity to compare them.  However,
+implementations on top of the same platform are not enough for an
+apples-to-apples comparison. The three approaches differ not only on their
+blaming strategies but also on the type checks they perform. To untangle the
+two, we supplement our experiment with two extra modes. The \emph{vanilla} mode
+ignores blame and instead uses information from dynamic type error messages to
+select how to proceed with migration. Hence it separates the effect of blame on
+the experiment from that of the type checks. The \emph{control} mode randomly
+selects which part of the program to migrate and serves as the null hypothesis
+control for the overall methodology.

### 2020-07-02 writing/challenges.tex
working on challenges

@@
 challenges.
+
+\cd{One point we haven's justified is why empirical. The justification
+seems easy in the sense that we are looking at a real language, programs,
+bugs rather than models.}
+
+\cd{I like this idea of the competent programmer. I couldn't have thought
+about homo economicus. Instead this was partly inspired by Gonen's work.}
+\cd{There they make a hypothesis, try to test it in vivo using designed model organisms
+with a well-specified first-order characteristic (e.g, a deleted gene).
+Then they collect data by placing the organism in a number of scenarios 
+and measuring aspects of its behavior and then they 
+construct an operational model from their data. Due to the synthetic
+nature of CS, for us the process goes in the opposite direction. In that
+sense the competent programmer is a model organism that, as its
+first-order characteristic, implements a direct consequence of the
+semantics. So in a sense for me this has always been about semantics and
+exploring their consequences.}
 
@@
 amount of data. But, as~\citet{lksfd-popl-2020} already demonstrated, the way
-around this dilemma is to postulate the equivalent of a {\it home economicus\/}
+around this dilemma is to postulate the equivalent of a {\it homo economicus\/}
 for programming languages---the {\em rational programmer\/}---and to implement
@@
 jia2011analysis}. This results into a large number of buggy programs whose
-mix-typed versions are suitable subjects for our experiment. However, mutation
+mixed-typed versions are suitable subjects for our experiment. However, mutation
 analysis traditionally aims to inject bugs that are sophisticated enough to
@@
 set of mutators specific for type errors inspired by common errors when
-programming with in Typed Racket.  These type errors cover a wide range of Typed
+programming in Typed Racket.  These type errors cover a wide range of Typed
 Racket's type system and moreover cause reliably runtime type errors in
-mix-typed versions of mutants.
+mixed-typed versions of mutants.
 
@@
 reasonably common~\cite{sta-nt-base-types, incorrect-ts, wmwz-ecoop-2017}, and
-hence must be addressed by any evaluation method.  \ldots needs to also explore
-the relation between blame and the location of bugs in case of wrong type
-annotations. Similarly to the previous challenge, we deal with this one by
-designing a set of mutators that ``break'' type annotations but leave programs
-otherwise intact.
+hence must be addressed by any evaluation method. Similarly to the
+previous challenge, we deal with this one by designing a set of mutators
+that ``break'' type annotations but leave programs otherwise intact.
 
-The second big challenge is the set of seemingly incomparable blame-assignment
-strategies. This results in this paper focus on three points of the landscape:
-the {\em natural\/} approach of Typed Racket~\cite{tf-dls-2006, tf-popl-2008,
-tfffgksst-snapl-2017, tf-icfp-2010}, the {\em transient\/} one of Reticulated
-Python~\cite{vsc-dls-2019,vss-popl-2017,vksb-dls-2014}, and the {\em optional\/}
-typing approach of industrial systems. Each of these points comes with a
-distinct blaming strategy and together they represent all proposed strategies so
-far. In addition, instead of experimenting with the three approaches in
-isolation, we have implemented transient and optional gradual typing as variants
-of Typed Racket to seize the opportunity to compare them.  However,
-implementations on top of the same platform are not enough for an
-apples-to-apples comparison. The three approaches differ not only on their
-blaming strategies but also on the type checks they perform. To untangle the
-two, we supplement our experiment with two extra modes. The \emph{vanilla} mode
+The second big challenge is the set of seemingly incomparable gradual
+typing systems. Despite the fact that the natural, transient and optional
+approaches are all well-understood as variants of the same minimal operational model
+~\cite{gf-icfp-2018, gfd-oopsla-2019}, their actual implementations live on
+different linguistic settings. Hence it is not possible to compare them directly 
+on the same programs; code needs to be adjusted in often subtle ways to
+transfer from one implementation to another. As a first step to overcome the problem, 
+we have implemented a transient and optional variant of Typed Racket.  
+However,
+implementations on top of the same platform are not sufficient for an
+apples-to-apples comparison. As we discuss in detail in
+section~\ref{sec:landscape}, the three strategies differ in the kind of
+blame information they produce. Natural blames one component while
+transient blames a set of components and optional none. To account for these
+differences, our competent programmer comes with \emp{modes} that represent 
+the alternative ways programmers can interpret blame information. For instance, in
+transient, one mode  has the competent programmer pick the oldest element of a blame set 
+as the next component to add types. After all, this corresponds to the
+earliest point in the program's evaluation that can discover the bug. 
+Another mode though, does not distinguish between the elements of the
+blame set and instead it requires that the competent programmer types 
+all blamed components at once. In order to compare different strategies
+across their different modes, we normalize the steps the competent
+programmer takes to find the bug by the effort of each step, that is the
+number of component it has to type. Finally, besides differences in how
+and what they blame, the three gradual typing approaches differ in the type checks they perform. 
+To untangle the two, we supplement our experiment with two extra modes. The \emph{vanilla} mode
 ignores blame and instead uses information from dynamic type error messages to
-select how to proceed with migration. Hence it separates the effect of blame on
-the experiment from that of the type checks. The \emph{control} mode randomly
+select the next component to type. Hence it separates the effect of blame 
+from that of the type checks. The \emph{control} mode randomly
 selects which part of the program to migrate and serves as the null hypothesis
-control for the overall methodology.
+control for the overall method.

### 2020-07-02 writing/introduction.tex
working on challenges

@@
 Tobin-Hochstadt et al.'s use of higher-order contract system~\cite{ff-icfp-2002,
-mf-toplas-2009}; and the error messages from failed safety checks in the
+mf-toplas-2009}; and the approach of industrial systems that forego blame
+and instead rely on error messages from the
 underlying programming language.

### 2020-07-02 writing/challenges.tex
paper, the first blood

@@
 programmer takes to find the bug by the effort of each step, that is the
-number of component it has to type. Finally, besides differences in how
+number of component it has to type. Moreover, besides differences in how
 and what they blame, the three gradual typing approaches differ in the type checks they perform. 
-To untangle the two, we supplement our experiment with two extra modes. The \emph{vanilla} mode
+To untangle the two, we supplement our experiment with an extra mode. The \emph{vanilla} mode
 ignores blame and instead uses information from dynamic type error messages to
 select the next component to type. Hence it separates the effect of blame 
-from that of the type checks. The \emph{control} mode randomly
+from that of the type checks. Finally, to exclude that the experiment
+results are not products of pure chance, the \emph{control} mode randomly
 selects which part of the program to migrate and serves as the null hypothesis
 control for the overall method.
+
+In sum, the peculiarities of the gradual typing landscape demand careful
+design to turn the idea of the competent programmer to a robust method.

### 2020-07-02 writing/challenges.tex
fix latex

@@
 transient blames a set of components and optional none. To account for these
-differences, our competent programmer comes with \emp{modes} that represent 
+differences, our competent programmer comes with \emph{modes} that represent 
 the alternative ways programmers can interpret blame information. For instance, in

### 2020-07-02 writing/challenges.tex
polish challenges

--- 2020-07-02 writing/challenges.tex
@@
-\section{Challenges}
+\section{Why is it hard to evaluate the value of blame?}
 
-Creating a method for comparing blame assignment strategies poses several severe
+Creating a method for comparing blame assignment strategies poses several
 challenges.
 
-\cd{One point we haven's justified is why empirical. The justification
-seems easy in the sense that we are looking at a real language, programs,
-bugs rather than models.}
+% \cd{One point we haven's justified is why empirical. The justification
+% seems easy in the sense that we are looking at a real language, programs,
+% bugs rather than models.}
+%% MF: I edited the text to include this.
 
-\cd{I like this idea of the competent programmer. I couldn't have thought
-about homo economicus. Instead this was partly inspired by Gonen's work.}
-\cd{There they make a hypothesis, try to test it in vivo using designed model organisms
-with a well-specified first-order characteristic (e.g, a deleted gene).
-Then they collect data by placing the organism in a number of scenarios 
-and measuring aspects of its behavior and then they 
-construct an operational model from their data. Due to the synthetic
-nature of CS, for us the process goes in the opposite direction. In that
-sense the competent programmer is a model organism that, as its
-first-order characteristic, implements a direct consequence of the
-semantics. So in a sense for me this has always been about semantics and
-exploring their consequences.}
+% \cd{I like this idea of the competent programmer. I couldn't have thought
+% about homo economicus. Instead this was partly inspired by Gonen's work.}
+% \cd{There they make a hypothesis, try to test it in vivo using designed model organisms
+% with a well-specified first-order characteristic (e.g, a deleted gene).
+% Then they collect data by placing the organism in a number of scenarios 
+% and measuring aspects of its behavior and then they 
+% construct an operational model from their data. Due to the synthetic
+% nature of CS, for us the process goes in the opposite direction. In that
+% sense the competent programmer is a model organism that, as its
+% first-order characteristic, implements a direct consequence of the
+% semantics. So in a sense for me this has always been about semantics and
+% exploring their consequences.}
+
+%% MF says: yes, I figured you were inspired by biology but in a
+% synthetic world, like econ, you need also a rationale for why this
+% thing is worthwhile measuring .. and so they came up with homo
+% econ.
 
 The most serious challenge concerns the nature of this research. Unlike ordinary
-work on programming languages, the question seems to call for research similar
-to the one in the human-computer interaction area. At the same time, a
-significant evaluation result demands a good number of cases yielding a serious
+work on programming languages, the question seems to call for empirical research
+similar to the one in the human-computer interaction area. At the same time, a
+significant evaluation result demands a large number of cases yielding a serious
 amount of data. But, as~\citet{lksfd-popl-2020} already demonstrated, the way
@@
 this rational programmer as an algorithm.  As indicated in the introduction, in
-the context of gradual typing a rationally acting programmer translates the
+the context of gradual typing, the rational programmer translates the
 Wadler-Findler slogan into a debugging method, searching for the source of the
-impedance mismatch in an incremental fashion.
+impedance mismatch in an incremental fashion. Measuring this simulated
+programmer's behavior on a large number of inputs is the equivalent of
+collecting empirical data in a truly human-focused discipline.
 
-One derivative challenge to the first is the problem of simulating the
-programmer's mistakes. Before a programmer can try to find mistakes, mistakes
-have to be made. Technically the problem is the lack of a sufficiently large
-available corpus of type errors for a meaningful experiment. Similar to
-~\citet{lksfd-popl-2020}, we cover this gap by injecting bugs in typed programs
-with mutation analysis~\cite{lipton1971fault, demillo1978hints,
-jia2011analysis}. This results into a large number of buggy programs whose
-mixed-typed versions are suitable subjects for our experiment. However, mutation
-analysis traditionally aims to inject bugs that are sophisticated enough to
-challenge extensive test suites. As a consequence, the ill-typed mutants we care
-about here are usually deemed \emph{incompetent} and mutation analysis
-frameworks are fine-tuned to avoid them. To overcome this problem, we design a
-set of mutators specific for type errors inspired by common errors when
-programming in Typed Racket.  These type errors cover a wide range of Typed
-Racket's type system and moreover cause reliably runtime type errors in
-mixed-typed versions of mutants.
+One derivative challenge of the first one is the problem of simulating the
+programmer's mistakes or, more precisely, the origin of impedance
+mismatches. Before a programmer can try to find mistakes, mistakes have to be
+made. Technically the problem is the lack of a sufficiently large available
+corpus of type errors for a meaningful experiment. Similar to
+~\citet{lksfd-popl-2020}, Following~\citet{lksfd-popl-2020} again, the solution
+is to use mutation analysis~\cite{lipton1971fault, demillo1978hints,
+jia2011analysis} to create mistakes in a synthetic manner. This process yields a
+large number of buggy programs whose mixed-typed versions are suitable subjects
+for our experiment.  The mutation analysis of Lazarek et al, is, however,
+completely useless for creating impedance mismatches in a gradually typed
+context.
 
-Another derivative challenge is the problem of mistakes in type specifications.
-Gradual typing enables programmers to move an untyped code base into the typed
-realm on an incremental basis. The retroactive addition of a specific type to a
-component, even a library, may result from a misunderstanding of the code and
-may thus be a mistake itself. Empirical evidence suggests that this scenario is
-reasonably common~\cite{sta-nt-base-types, incorrect-ts, wmwz-ecoop-2017}, and
-hence must be addressed by any evaluation method. Similarly to the
-previous challenge, we deal with this one by designing a set of mutators
-that ``break'' type annotations but leave programs otherwise intact.
+While mutation analysis traditionally aims to inject bugs that challenge
+extensive test suites, the ones yielding ill-typed mutants are usually deemed
+\emph{incompetent}. As a matter of fact, mutation analysis frameworks are
+fine-tuned to avoid them---yet it is precisely those that are needed for
+evaluating type impedance mismatches. The ones used here are newly designed,
+inspired by the authors extensive experience with programming in Typed Racket.
+These type errors cover a wide range of Typed Racket's type system and moreover
+reliably cause runtime exceptions in mixed-typed versions of mutants.
 
-The second big challenge is the set of seemingly incomparable gradual
-typing systems. Despite the fact that the natural, transient and optional
-approaches are all well-understood as variants of the same minimal operational model
-~\cite{gf-icfp-2018, gfd-oopsla-2019}, their actual implementations live on
-different linguistic settings. Hence it is not possible to compare them directly 
-on the same programs; code needs to be adjusted in often subtle ways to
-transfer from one implementation to another. As a first step to overcome the problem, 
-we have implemented a transient and optional variant of Typed Racket.  
-However,
-implementations on top of the same platform are not sufficient for an
-apples-to-apples comparison. As we discuss in detail in
-section~\ref{sec:landscape}, the three strategies differ in the kind of
-blame information they produce. Natural blames one component while
-transient blames a set of components and optional none. To account for these
-differences, our competent programmer comes with \emp{modes} that represent 
-the alternative ways programmers can interpret blame information. For instance, in
-transient, one mode  has the competent programmer pick the oldest element of a blame set 
-as the next component to add types. After all, this corresponds to the
-earliest point in the program's evaluation that can discover the bug. 
-Another mode though, does not distinguish between the elements of the
-blame set and instead it requires that the competent programmer types 
-all blamed components at once. In order to compare different strategies
-across their different modes, we normalize the steps the competent
-programmer takes to find the bug by the effort of each step, that is the
-number of component it has to type. Finally, besides differences in how
-and what they blame, the three gradual typing approaches differ in the type checks they perform. 
-To untangle the two, we supplement our experiment with two extra modes. The \emph{vanilla} mode
+Another derivative challenge of the first one is the problem of mistakes in type
+annotations themselves.  Gradual typing enables programmers to move an untyped
+code base into the typed realm on an incremental basis. The retroactive addition
+of a specific type to a component, even a library, may result from a
+misunderstanding of the code and may thus be a mistake itself. Empirical
+evidence suggests that this scenario is quite common~\cite{sta-nt-base-types,
+incorrect-ts, wmwz-ecoop-2017}, and hence must be addressed by any evaluation
+method. Conversely, the method not only demands mutators for code that cause
+good impedance mismatches, it also calls for a set of mutators that ``break''
+type annotations without breaking the code itself.
+
+The second big challenge is due to implementation disparity of gradual type
+systems. While the various approaches are well understood in theory, they come
+with vastly different type system, accommodate seemingly incompatible language
+features, and apply to radically different coding styles in the untyped world.
+To eliminate these differences as variables of a comparison, the application of
+an evaluation method must use the same test programs and the same type system.
+The case study presented here builds on Greenman and Felleisen's transient
+variant for functional Typed Racket~\cite{gf-icfp-2018}. the revised transient
+variant accommodates many more linguistic features and calls library code in a
+way that understand the transient semantics.
+
+Beyond the implementation issue, a comparison also faces the problem that the
+three variants---the natural semantics, the transient semantics, and the erasure
+semantics---come with three radically different blame assignment systems.  For
+example, the natural semantics assigns blame to one component while transient
+assigns blame to a set of components. The erasure semantics does not blame a
+component per se, but it comes with an exception location and the usual stack
+traces.
+
+Making these strategies comparable is {\em not\/} an issue of language
+implementation but one about how the rational programmer behaves with regard to
+these messages. Specifically, the rational programmer comes with \emph{modes}
+that represent the alternative ways in which a programmer may interpret blame
+information. For instance, in transient, one mode has the competent programmer
+pick the oldest element of a blame set as the next component to add types. After
+all, this corresponds to the earliest point in the program's evaluation that can
+discover the bug.  Another mode does not distinguish between the elements of the
+blame set and requires that the rational programmer types all blamed components
+at once.  Once information about the search steps in each mode is available, it
+is normalized. In particular, it is not just the steps that are counted but the
+
+Moreover, the three gradual typing systems differ in the type checks they
+perform.  An extra mode aims to untangle the two issues. This \emph{vanilla} mode
 ignores blame and instead uses information from dynamic type error messages to
-select the next component to type. Hence it separates the effect of blame 
-from that of the type checks. The \emph{control} mode randomly
-selects which part of the program to migrate and serves as the null hypothesis
-control for the overall method.
+select the next component to type. Hence it separates the effect of blame from
+that of the type checks.
+
+In sum, the peculiarities of the gradual typing landscape demand a careful
+design of both the evaluation method and the application of the method to
+specific gradual typing systems. Otherwise measuring the rational programmer
+will yield potentially random results.

### 2020-07-06 writing/challenges.tex
working on evaluation description + moderate reorganization

@@
 \section{Why is it hard to evaluate the value of blame?}
+\label{sec:challenges}
 
-Creating a method for comparing blame assignment strategies poses several
-challenges.
-
-% \cd{One point we haven's justified is why empirical. The justification
-% seems easy in the sense that we are looking at a real language, programs,
-% bugs rather than models.}
-%% MF: I edited the text to include this.
-
-% \cd{I like this idea of the competent programmer. I couldn't have thought
-% about homo economicus. Instead this was partly inspired by Gonen's work.}
-% \cd{There they make a hypothesis, try to test it in vivo using designed model organisms
-% with a well-specified first-order characteristic (e.g, a deleted gene).
-% Then they collect data by placing the organism in a number of scenarios 
-% and measuring aspects of its behavior and then they 
-% construct an operational model from their data. Due to the synthetic
-% nature of CS, for us the process goes in the opposite direction. In that
-% sense the competent programmer is a model organism that, as its
-% first-order characteristic, implements a direct consequence of the
-% semantics. So in a sense for me this has always been about semantics and
-% exploring their consequences.}
-
-%% MF says: yes, I figured you were inspired by biology but in a
-% synthetic world, like econ, you need also a rationale for why this
-% thing is worthwhile measuring .. and so they came up with homo
-% econ.
-
-The most serious challenge concerns the nature of this research. Unlike ordinary
+The most serious challenge for comparing blame assignment strategies concerns 
+the nature of this research. Unlike ordinary
 work on programming languages, the question seems to call for empirical research
@@
 programmer's mistakes or, more precisely, the origin of impedance
-mismatches. Before a programmer can try to find mistakes, mistakes have to be
-made. Technically the problem is the lack of a sufficiently large available
-corpus of type errors for a meaningful experiment. Similar to
-~\citet{lksfd-popl-2020}, Following~\citet{lksfd-popl-2020} again, the solution
-is to use mutation analysis~\cite{lipton1971fault, demillo1978hints,
-jia2011analysis} to create mistakes in a synthetic manner. This process yields a
-large number of buggy programs whose mixed-typed versions are suitable subjects
-for our experiment.  The mutation analysis of Lazarek et al, is, however,
-completely useless for creating impedance mismatches in a gradually typed
-context.
+mismatches. Before a programmer can try to find mistakes, mistakes have to
+be made. Technically the problem is the lack of a sufficiently large
+available corpus of type errors for a meaningful experiment.
+Following~\citet{lksfd-popl-2020} again, the solution is to use mutation
+analysis~\cite{lipton1971fault, demillo1978hints, jia2011analysis} to
+create mistakes in a synthetic manner. This process yields a large number
+of buggy programs whose mixed-typed versions are suitable subjects for our
+experiment.  The mutation analysis of Lazarek et al, is, however,
+completely useless for creating impedance mismatches in a gradually
+typed context.
 
@@
 fine-tuned to avoid them---yet it is precisely those that are needed for
-evaluating type impedance mismatches. The ones used here are newly designed,
+evaluating blame assignment strategies. The mutators used here are newly designed,
 inspired by the authors extensive experience with programming in Typed Racket.
-These type errors cover a wide range of Typed Racket's type system and moreover
+They cause type errors that cover a wide range of Typed Racket's type system and moreover
 reliably cause runtime exceptions in mixed-typed versions of mutants.
@@
 
-The second big challenge is due to implementation disparity of gradual type
-systems. While the various approaches are well understood in theory, they come
-with vastly different type system, accommodate seemingly incompatible language
-features, and apply to radically different coding styles in the untyped world.
-To eliminate these differences as variables of a comparison, the application of
-an evaluation method must use the same test programs and the same type system.
-The case study presented here builds on Greenman and Felleisen's transient
-variant for functional Typed Racket~\cite{gf-icfp-2018}. the revised transient
-variant accommodates many more linguistic features and calls library code in a
-way that understand the transient semantics.
+The second big challenge is due to the implementation disparity of gradual
+type systems. While the various approaches are well understood in theory,
+they come with vastly different type systems, accommodate seemingly
+incompatible language features, and apply to radically different coding
+styles in the untyped world.  To eliminate these differences as variables
+of a comparison, the application of an evaluation method to different
+gradual typing systems must use the same test programs and the same type
+system. Otherwise, results for different gradual typing systems are not
+comparable.  For instance, the case study presented here builds on
+Greenman and Felleisen's transient variant for functional Typed
+Racket~\cite{gf-icfp-2018} to obtain a transient semantics that is
+comparable with Typed Racket's default natural one. The revised transient
+variant accommodates many more linguistic features and calls library code
+in a in a way that is faithful to the transient
+semantics~\cite{vss-popl-2017}.
 
-Beyond the implementation issue, a comparison also faces the problem that the
-three variants---the natural semantics, the transient semantics, and the erasure
-semantics---come with three radically different blame assignment systems.  For
-example, the natural semantics assigns blame to one component while transient
-assigns blame to a set of components. The erasure semantics does not blame a
-component per se, but it comes with an exception location and the usual stack
-traces.
+Beyond the implementation issue, a comparison also faces the problem that
+the three variants---the natural semantics, the transient semantics, and
+the erasure semantics---come with three radically different blame
+assignment strategies.  For example, the natural semantics assigns blame
+to one component while transient assigns blame to a set of components. The
+erasure semantics does not blame a component per se, but it comes with an
+exception location and the usual stack traces.
 
@@
 implementation but one about how the rational programmer behaves with regard to
-these messages. Specifically, the rational programmer comes with \emph{modes}
-that represent the alternative ways in which a programmer may interpret blame
-information. For instance, in transient, one mode has the competent programmer
+these messages. Specifically, the rational programmer must come with \emph{modes}
+that represent the alternative ways in which a programmer may interpret blame. 
+For instance, in transient, one mode has the competent programmer
 pick the oldest element of a blame set as the next component to add types. After
 all, this corresponds to the earliest point in the program's evaluation that can
-discover the bug.  Another mode does not distinguish between the elements of the
+discover an impedance mismatch.  Another mode does not distinguish between the elements of the
 blame set and requires that the rational programmer types all blamed components
-at once.  Once information about the search steps in each mode is available, it
-is normalized. In particular, it is not just the steps that are counted but the
+at once. Of course, a step of the rational programmer in the latter mode
+is not as laborious as one in the former. Hence, an evaluation method should take 
+into account the effort of the rational programmer when comparing 
+different modes.
 
-Moreover, the three gradual typing systems differ in the type checks they
-perform.  An extra mode aims to untangle the two issues. This \emph{vanilla} mode
-ignores blame and instead uses information from dynamic type error messages to
-select the next component to type. Hence it separates the effect of blame from
-that of the type checks.
+Besides the blame information they produce, the three gradual typing
+systems also differ in the runtime type checks they perform. Therefore an
+application of the evaluation method to a gradual typing system has to
+factor in the effect of type checks to determine whether blame helps the
+rational programmer find an impedance mismatch. A solution to this issue
+is yet another mode for the rational programmer that uses information only
+from teh underlying languages exceptions to select the next component(s) to type.
+This \emph{vanilla} mode simulates a rational programmer that ignores
+blame and serves as the baseline for blame's value.
 
-In sum, the peculiarities of the gradual typing landscape demand a careful
-design of both the evaluation method and the application of the method to
-specific gradual typing systems. Otherwise measuring the rational programmer
-will yield potentially random results.
+The final big challenge is that the different gradual type
+systems, the number of necessary modes for the rational programmer
+and the large number of mutants lead to an explosion of scenarios the
+method needs to examine. Thus a practical application of the
+method has no option but to carefully sample this space while 
+making sure that it still produces reproducible conclusions.
+
+\smallskip
+
+These challenges expose the subtlety of evaluating blame strategies but
+also suggest a recipe for doing so correctly:
+
+\begin{itemize}
+
+\item Implement all the gradual type systems to be compared on a common
+  basis (section~\ref{sec:landscape}).
+
+\item Construct a large number of suitable benchmarks for all gradual type
+  systems with representative
+  type errors  (section~\ref{sec:mutate}).
+
+\item Simulate the rational programmer with sufficient 
+  modes to (i) match the peculiarities of each gradual type system and 
+    (ii) isolate blame from other confounding factors
+    (section~\ref{sec:rational}).
+
+\item Sample the experimental space  to obtain a computationally feasible
+  but statistically meaningful experiment (section~\ref{sec:sample}).
+
+\end{itemize}

### 2020-07-06 writing/experiment.tex
titles are questions
working on evaluation description + moderate reorganization

@@
-\section{The Experimental Setup}
+\subsection{How to Make Benchmarks for Blame}
 
-\begin{itemize}
-
-\item Summary of the blame trail experiment. We sample and follow blame trails through configuration
-        lattices.
-
-\item Evaluating each point in the design space. We run the blame trail experiment for every point in the
-        previously discussed design space.
-
-\item Selection of programs. We are using a set of representative
-  programs. Argue the programs are representative of the range of programs written by TR developers; point to the JFP paper.
-
-\item Sampling of mutants. We have too many mutants to run them all, so we sample them with an aim to represent equally all of the mistakes captured by our operators.
+\subsection{How to Sample Mutants and Configurations}
 

### 2020-07-06 writing/ideas.tex
fix basic latex mistakes in rudimentary way
working on evaluation description + moderate reorganization

@@
-\section{How to Evaluate Blame}
+\section{How to Simulate the Rational Programmer} \label{sec:ideas}
 
-\begin{itemize}
+Given the comparable implementations of the three gradual typing systems,
+the next step for applying our method is to obtain a set of fully
+typed benchmarks that each comes with a single impedance mismatch between
+its type specifications and its code. As we briefly discuss in
+section~\ref{sec:challenges}, we derive such ill-typed benchmarks with
+mutation analysis from well-typed programs.  We return to mutation
+analysis and the benchmarks in section~\ref{sec:experiment}.  Herein,
+we assume that we have them in hand and instead we focus on the
+main element of our method, the rational programmer.
 
-    \item How to analyze this space?
-          \begin{itemize}
-            \item We want a quantative measure of error msg utility that allows
-                  us to evaluate each point in the above space.
-          \end{itemize}
+The rational programmer exploits blame information to figure out wih. Therefore the way to understand its behavior 
+is to analyze the migratory path it follows through the 
+\emph{configuration space} of the program. The space
+consists of variants of the program, dubbed \emph{configurations}, that 
+differ in their type annotations. 
+Herein we build on top of Typed Racket's migratory typing
+approach,\footnote{The method can be adapted to so-called true gradual typing
+systems~\cite{svcb-snapl-2015} in a straightforward manner.}, hence, each configuration differs from another
+in which of its components are typed and which are not. Formally, we can
+represent a program $\system$ as a set of of components
+$\set{\component}$ and each configuration $\conf$ of $\system$ is a subset of 
+$\set{\component}$
+that contains exactly the components of  $\system$ that are typed in
+$\conf$. Naturally the configurations of  $\system$ are ordered by the subset relation and
+thus form a lattice $\lattice{\system}$ with size $2^{\size{\system}}$.
+The bottom configuration of $\lattice{\system}$ is always $\emptyset$ and
+the top element is $\system$ itself. In between there are all the
+mixed-type configurations. If we assume that  component $\component$ of $\system$
+contains the type error, any configuration $\conf$ that
+includes $\component$ is ill-typed. Any other configuration $\conf$ may blame 
+a $\component_b$, which we denote with $\blame{\conf}$. Let a blametrail
+be a sequence of configurations $\conf_0,...\conf_n$ where 
+for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
+$\conf_{i+1} \setminus \conf_i = \blame{\conf_i}$. That is a blametrail
+corresponds to a succession of a configurations that start with a
+configuration $\conf_0$, dubbed the \emph{root of the blametrail}, and 
+each configuration is derived from its previous one by typing the
+component that the latter blames. A rational programmer is one that
+migrates the variant of a program $\system$ that corresponds to
+configuration $\conf_0$ along the blametrail that starts at $\conf_0$.
+We say a blametrail is final iff its last configuration, $\conf_n$
+doesn't blame a component, $\blame{\conf_i} = \emptyset$. We call a
+blametrail \emph{successful} iff it is final and its last configuration
+$\conf_n$ contains the faulty component of $\system$, $\faulty{\system}
+\in \conf_n$.
 
-     \item Blame trails
-       \begin{itemize}
+Using the notion of successful blametrails we can define precisely what it
+means for blame to add value to a gradual typing system:
 
-        \item We want an empirical study to understand how the behavior of blame
-        in real systems.
+All blametrails are successfull: Given a program P and its configuration
+lattice $\lattice{\system}$, all blametrails in the lattice are successful.
 
-        \item Blame trails provide a quantitative measure of a a system's
-        error message utility, and they can be applied to every point
-        in the above space.
+With these definitions we can describe with precision the experimental
+process of our method. For each benchmark we create its configuration
+lattice. Then we separate the configurations of the lattice that 
+when run result in blame or result in a static type error from  those that do not
+do either. The former are 
+suitable to be roots of a blametrail. Thus we simulate the rational 
+programmer and if the configuration blames we type their blamed component to obtain the next
+configuration in the blametrail and repeat the process until the
+blametrail becomes final. At this point we check whether the last
+configuration fails to type check. If it does for all blametrails in the
+lattice, then the property holds. Otherwise it fails.
 
-       \item We generalize the blame trails idea of Lukas et al [POPL '20] to make it
-        more systematic, scalable, and general, by:
-           \begin{itemize}
-             \item Adding a null hypothesis baseline for comparison between multiple systems
-             \item Analyzing multiple blame following strategies
-             \item Mutating interfaces as well as programs
-             \item Incorporating domain knowledge in the selection of blame following
-           strategies and mutation
-             \item Evaluating systems on the same platform for apples-to-apples comparisons
-           \end{itemize}
+In addition to the property the framework allows us to assign a metric of
+effort to a successful blametrail; $\effort{\conf_0, ...., \conf_n} = n$
 
-         \item Null hypothesis. We should have a baseline to tell how all of the points in the
-        design space compare to having nothing at all.
-\end{itemize}
+\subsection{How to Generalize the Rational Programmer}
+
+As we discuss in section~\ref{sec:challenges}, an important factor that the method need to
+take into account is that different gradual typing systems blame
+differently and perform different checks. For that reason in
+section~\ref{sec:challenges} we introduce the notion of modes for the
+rational programmer. Defining these modes requires a generalization of the 
+discussion above.
+
+In particular it requires revisiting the notion of blametrail and that of
+successful blametrail.
+
+We define a blametrail in a mode  $\modem$ as  
+a sequence of configurations $\conf_0,...\conf_n$ where 
+for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
+$\conf_{i+1} \setminus \conf_i = \mode{\conf_i}$.
+
+   For example we can obtain the simple blame mode
+for the natural system if we use, as above, the selector $\blame{\cdot}$. 
+However, the same mode has a different meaning in transient as this
+semantics results in multiple blamed components.
+
+This requires generalizing the notion of successful blametrail
+to that of successful blametrees: a blametree is the set of all
+blametrails with root a configuration $\conf_0$. A blametree is succesful if at least one
+of its blametrails is succesful.
+
+Of course for systems such as transient we can add modes where the rational
+programmer picks one blametrail as part of the migration process. For
+example the \first{\blame{\cdot}} mode continues with the oldest blame
+entry in the blameset of a configuration, as we discuss in
+section~\ref{sec:challenges}.
+
+Moreover, the \emph{vanilla} mode corresponds to ignoring blame and instead
+using only information from the dynamic type error that a configuration
+produces using the $\error{\cdot}$ selector.
+
+Finally, the \emph{null} mode uses selector $\random{\cdot}$ and it is
+used by our method as the null hypothesis control experiment.
+
+The following table collects all the modes and their selectors for the
+three gradual typing systems we compare in this paper:
+
+\begin{tabular}{c|c|c|c}
+                    & {\bf Natural}  & {\bf Transient}        & {\bf Erasure} \\
+   \hline 
+  {\bf Simple Blame}   & \blame{\conf}  & \blame{\conf}          &               \\
+ {\bf First Blame}  &                 & \first{\blame{\conf}}  &               \\
+  {\bf Vanilla}       &  \error{\conf} & \error{\conf}          & \error{\conf} \\
+  {\bf Null}        & \random{\conf} & \random{\conf}         & \random{\conf} 
+\end{tabular}
+
+As a last step we also need to lift the notion of effort to blametrees.
+The effort of a blametree $\effort{\{\conf_0, ...., \conf_n\},...}$ is the
+aggregate effort of all its blametrails.

### 2020-07-06 writing/introduction.tex
working on evaluation description + moderate reorganization

--- 2020-07-01 writing/introduction.tex
@@
-\section{Does Blame Matter?}
+\section{Does Blame Matter?} \label{sec:introduction}
 
@@
 original authors got this word wrong.} can't be blamed'' turned the theorem into a
-slogan~\cite{wf-esop-2009}. Indeed, academic implementationsa---for example,,
+slogan~\cite{wf-esop-2009}. Indeed, academic implementations---for example,
 Reticulated Python~\cite{vsc-dls-2019, vss-popl-2017, vksb-dls-2014} and Typed
 Racket~\cite{tf-dls-2006,tf-popl-2008,tfffgksst-snapl-2017,tf-icfp-2010}---come
-with sophisticated blame-assignment subsystems. Theoreticians firmly embrace the
+with sophisticated blame assignment strategies. Theoreticians firmly embrace the
 idea that blame can help practicing programmers find impedance mismatches between
-the types they added to a software system and the behavior remaining untyped
+the types they added to a software system and the behavior of the remaining untyped
 components.
@@
 Practical gradual type systems ignore blame assignment almost completely.
-Industrial systems such as Flow,\footnote{flow} Hack,\footnote{hack} or
-TypeScript~\cite{ts} exploit types for IDE actions and for finding typos in
+Industrial systems such as Flow,\footnote{\url{https://flow.org}}
+Hack,\footnote{\url{https://hacklang.org}} or
+TypeScript\footnote{\url{https://www.typescriptlang.org}} exploit types for IDE actions and for finding typos in
 code. Then these compilers remove types and rely on the built-in safety checks of
@@
 industrial systems, it comes as an even bigger surprise that the research
-literature and the industrial blog world do not present or discuss any possible
-answers.  Instead, language designers seem to answer this question one way or
-another without any scientific justification, and even when they have a language,
-they neglect to look into this question. The research community simply does not
-have a method for evaluating such questions, and given the question's practical
-relevance, this is a problem.
+literature and the industrial blog world do not present or discuss any
+possible answers.  Instead, when language designers decide to look into the
+question, they seem to answer it one way or another without any scientific
+justification. Most of the time though they ignore the question altogether. The
+root of the issue is that the research community simply does not have a
+method for evaluating such questions, and given the question's practical
+relevance, this a significant problem.
 
 This paper's first contribution is an automated method for evaluating the
-effectiveness of blame assignment strategies in the gradual typing world (see
-section~\ref{sec:ideas}). Roughly speaking, the method simulates a programmer who
-follows the Wadler-Findler slogan quoted above. Its implied method suggests that
-programmers can find impedance-mismatch bugs by adding types to a partially typed
-software system. That is, when the run-time system signals an impedance mismatch,
-a good blame messages helps programmers infer where to add an additional type
-specification. Since components that pass the type checker cannot cause such
-impedance signals; the type checker must discover these mismatches at compile
-time.  If the additional type does {\em not\/} cause a static type error, the
-program is run again until it signals another violation.  Counting the number of
-steps it takes to discover the mistake is a measure of the effectiveness of the
-blame assignment.  The method is loosely based on the work of Lazarek et
-al.~\cite{lksfd-popl-2020}.
+effectiveness of blame assignment strategies in the gradual typing world
+(sections ~\ref{sec:landscape},~\ref{sec:mutate},~\ref{sec:rational} and \ref{sec:sample}). 
+Roughly speaking, the method, which is
+loosely based on the work of Lazarek et al.~\cite{lksfd-popl-2020},
+simulates a programmer who follows the Wadler-Findler slogan quoted above.
+Such a \emph{rational programmer} can find impedance-mismatch
+bugs by adding types to a partially typed program. That is, when the
+run-time system signals an impedance mismatch, a good blame message helps
+the programmer infer where to add an additional type specification. Since
+components that pass the type checker cannot cause impedance mismatches;
+the type checker discovers these mismatches at compile time and points
+directly to the faulty component. Otherwise, if the
+additional type does {\em not\/} cause a static type error, the program
+ is run again to signal another violation. Then the programmer can heed
+blame one more time to add yet another type specification. Counting the number of
+steps it takes to discover the mistake in this manner is a measure of the effectiveness
+of the blame assignment strategy of the language.
 
 The paper's second contribution is a set of results from applying the evaluation
-method to three distinct blame assignment systems and approximately 20,000
-cases\mf{Replace with real number} (see section~\ref{sec:experiment}): Vitousek et
+method to three distinct blame assignment strategies and approximately 20,000
+cases\mf{Replace with real number} (section~\ref{sec:results}): Vitousek et
 al.'s strategy of tracking typed/untyped boundaries in Reticulated;
 Tobin-Hochstadt et al.'s use of higher-order contract system~\cite{ff-icfp-2002,
-mf-toplas-2009}; and the error messages from failed safety checks in the
+mf-toplas-2009}; and the approach of industrial systems that forego blame
+and instead rely on error messages from the
 underlying programming language.
@@
 
-The results of the experiment (see section~\ref{sec:results}) validate the
-conjecture behind the work of theoreticians. A good blame assignment system
+The results of the experiment (section~\ref{sec:discussion}) validate the
+conjecture behind the work of theoreticians. A good blame assignment
+strategy
 greatly shortens the search for impedance mismatches between the specified types
@@
 academic system. Vitousek et al. lose one more time. Of course, a detailed
-discussion must acknowledge the threats to validity of these results (see
-section~\ref{sec:discussion}).
+discussion must acknowledge the threats to validity of these results.
 
-Next, the paper provide an overview of the blame assignment strategies as found in
-existing gradually typed language implementation (see section~\ref{sec:landscape})
-and the challenges of developing a systematic evaluation method
-(section~\ref{sec:challengs}).
+Next, the paper describes the challenges of developing an evaluation method
+(section~\ref{sec:challenges}), which may explain why the question has been neglected
+for such a long time.

### 2020-07-06 writing/mutate.tex
working on evaluation description + moderate reorganization

--- 2020-07-01 writing/ideas.tex
@@
-\section{How to Evaluate Blame}
+\section{How to Make Benchmarks} \label{sec:mutate}
 
-\begin{itemize}
-
-    \item How to analyze this space?
-          \begin{itemize}
-            \item We want a quantative measure of error msg utility that allows
-                  us to evaluate each point in the above space.
-          \end{itemize}
-
-     \item Blame trails
-       \begin{itemize}
-
-        \item We want an empirical study to understand how the behavior of blame
-        in real systems.
-
-        \item Blame trails provide a quantitative measure of a a system's
-        error message utility, and they can be applied to every point
-        in the above space.
-
-       \item We generalize the blame trails idea of Lukas et al [POPL '20] to make it
-        more systematic, scalable, and general, by:
-           \begin{itemize}
-             \item Adding a null hypothesis baseline for comparison between multiple systems
-             \item Analyzing multiple blame following strategies
-             \item Mutating interfaces as well as programs
-             \item Incorporating domain knowledge in the selection of blame following
-           strategies and mutation
-             \item Evaluating systems on the same platform for apples-to-apples comparisons
-           \end{itemize}
-
-         \item Null hypothesis. We should have a baseline to tell how all of the points in the
-        design space compare to having nothing at all.
-\end{itemize}
+Given the comparable implementations of the three gradual typing systems,
+the next step for applying our method is to obtain a set of fully
+typed benchmarks that each comes with a single impedance mismatch between
+its type specifications and its code. As we briefly discuss in
+section~\ref{sec:challenges}, we derive such ill-typed benchmarks with
+mutation analysis from well-typed programs.

### 2020-07-06 writing/paper.tex
challenges first
working on evaluation description + moderate reorganization

@@
 and poses the challenge of how to evaluate the effectiveness of blame
-assignment systems. This paper contributes (1) the first evaluation
-method for blame assignment systems and (2) the results from applying
-it to three different systems. \mf{This sentence must be checked
+assignment strategies. This paper contributes (1) the first evaluation
+method for blame assignment strategies and (2) the results from applying
+it to three different gradual type systems. \mf{This sentence must be checked
 before final submission.} These results confirm that blame assignment
@@
 
-\input{introduction.tex} \label{sec:introduction}
-\input{landscape.tex}    \label{sec:landscape}
-\input{challenges.tex}   \label{sec:challengs}
-\input{ideas.tex}        \label{sec:ideas}
-\input{experiment.tex}   \label{sec:experiment}
+\input{introduction.tex} 
+\input{challenges.tex}   
+\input{landscape.tex} 
+\input{mutate.tex}
+\input{rational.tex}
+\input{sample.tex}
 \input{results.tex}      \label{sec:results}

### 2020-07-06 writing/rational.tex
working on evaluation description + moderate reorganization

\section{How to Simulate the Rational Programmer} \label{sec:rational}

Consider a rational programmer  that tries to debug a faulty program in
Typed Racket with the default natural semantics. If the Typed Racekt
typechecker detects the issue with the program, the programmer knows where
to look for the bug; in the component that doesn't type check. However, if
the type checker blesses the program, the programmer relies on  runtime
type checks  to  signal an impedance mismatch between the
program's behavior and its types. In these checks raise an error, the programmer ``follows'' blame to
locate where to add a type specification next.  Hopefully, the extra
types are sufficient to turn the run time error into a static one. Otherwise,
the rational programmer starts over with the latest version of the program. 
In essence, at each iteration the
faulty program migrates to a version with more type specifications until
the faulty component becomes typed.

Therefore, a concise way to model rational programmers is with the
migratory paths they follow through the \emph{configuration space} of the
program. This space consists of variants of the program, dubbed
\emph{configurations}, that differ in their type specifications. In the 
setting of Typed Racket's migratory typing,\footnote{The
method can be adapted to so-called true gradual typing
systems~\cite{svcb-snapl-2015} in a straightforward manner.}, each
configuration differs from another in the components that have type
specifications. Hence, formally, we can represent a program
$\system$ as a set of components $\set{\component}$ and each
configuration $\conf$  as a subset of $\set{\component}$ that contains
exactly the components of  $\system$ that are typed in $\conf$. Naturally
the configurations of  $\system$ are ordered by the subset relation and
thus form a lattice $\lattice{\system}$ with $2^{\size{\system}}$ elements.
The bottom configuration of $\lattice{\system}$ is always $\emptyset$ and
the top element is $\system$ itself. In between there are all the
mixed-type configurations. If we assume that  component
$\faulty{\system}$ contains the fault, the type checker detects any
configuration $\conf$ that includes $\faulty{\system}$. Any other
configuration $\conf$ of $\system$ may blame a $\component_b$, which we denote with
$\blame{\system,\conf}$.\footnote{Blame can result either directly from a Typed
Racket type check or from a check in a Racket library function that raises
a blame error or, in the absence of blame, from an
exception that we track back to the component that causes it with the help
of Racket's stack trace.}

With these definitions, the actions of the rational programmer map to an ascending chain in
$\lattice{\system}$, dubbed a \emph{blametrail}. 
The starting configuration $\conf_0$ of a blametrail,  dubbed the \emph{root of the
blametrail}, is the configuration of
$\program$ that the programmer attempts to debug. Any other
configuration $\conf_{i+1}$ in the blametrail differs from its predecessor
$\conf_{i}$ in that 
the untyped component $\blame{\program, \conf_i}$ of $\conf_{i}$
 becomes typed in $\conf_{i+1}$. Putting these pieces together:

\begin{quote}
\it 
A natural blametrail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
$\conf_{i+1} \setminus \conf_i = \{\blame{\system, \conf_i}\}$.
\end{quote}

So far the definition of blametrail does not capture whether blame helps the rational
programmer locate the bug in the root of a blametrail. For that we need 
two properties of blametrails. A blametrail is \emph{final} if
the rational programmer cannot extend it further. Or formally, if 
$\blame{\program, \conf_n}$ is not defined where  $\conf_n$ is the last
configuration of the blametrail. A blametrail is \emph{successful} 
if it is final and  its last configuration
$\conf_n$ contains the faulty component, that is $\faulty{\system}
\in \conf_n$.  We can now give a precise definition of what it
means for blame to help the rational programmer in natural Typed Racket:
\begin{quote}
\it
  Natural blame is useful for a program $\system$ iff 
  all natural blametrails in  $\lattice{\system}$ are successful.
\end{quote}

However, the definition does not give a full picture of how much blame helps the
natural rational programmer. In particular, all blametrails in $\lattice{\system}$
may  be successful but this may well be because of some other aspect of
natural, such as the type checks it performs, or it may even be a result
of pure chance. To exclude these possibilities we define two different
modes for the rational programmer. In the first, the natural programmer
ignores blame and instead decides what component to type next based on
information from the stack trace of the failed runtime type check.  This 
vanilla rational programmer corresponds to \emph{vanilla natural
blametrails}:
\begin{quote}
\it 
A vanlla natural blametrail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
$\conf_{i+1} \setminus \conf_i = \{\error{\system, \conf_i}\}$.
\end{quote}
where $\error{\system, \conf}$ denotes the component that triggers a
failed runtime check according to Racket's stack trace. 
In the second mode, the natural programmer
ignores any kind of information related to the type error and randomly
picks which component to type next.  This 
random rational programmer corresponds to \emph{random natural
blametrails}:
\begin{quote}
\it 
A vanlla natural blametrail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
$\conf_{i+1} \setminus \conf_i = \{\random{\system, \conf_i}\}$.
\end{quote}
where $\random{\system, \conf}$ denotes a randomly selected of $\system$
that is not in $\conf_i$.

With vanilla and random blametrails, we can define the value that blame
adds to natural Typed Rachet: 
\begin{quote}
\it
  Natural blame  adds value to natural for a program $\system$ iff 
 iff 
  all natural blametrails in  $\lattice{\system}$ are successful while 
not all vanilla and random natural blametrails are successful.
\end{quote}

\noindent
{\bf Experimental Process.}~Given a set of faulty benchmarks such as the ones
from~section \ref{sec:mutate}, it is straightforward to turn the above definition
into the falsfiable hypothesis of an experiment that tests whether blame
add value to natural Typed Racket. Specifically, for each benchmark
we first construct its configuration lattice. Then, we determine the
configurations of the lattice that detect the fault in the
program. This can happen either because a
 runtime type check fails or Typed Racket's typechecker
rejects the configuration. In either case,  each of these configurations is the
root of a natural blametrail in the configuration lattice of the benchmark. \emph{We
hypothesize that all such natural blametrails are successful while the
vanilla and random ones are not}. 
To confirm or falsify the hypothesis for a benchmark, we simply extend each blametrail
 according to the mode of the corresponding rational programmer  until the blametrail
becomes final\footnote{All blametrails are final since the lattices are
finite.} and we check if its last configuration contains the faulty
component of the benchmark.

\subsection{How to Generalize the Rational Programmer Beyond Natural Typed
Racket}

Adapting the simulated rational programmer to handle transient
and erasure Typed Racket requires taking into account the distinguishing
features of each semantics. As we discuss in section~\ref{sec:landscape}, 
transient blames a set of components in contrast to natural's
single-component blame. At the other end of the blame spectrum, erasure
blames no component and relies on Racket's exceptions.

Hence, for an erasure rational programmer only the vanilla and random 
rational programmers are meaningful. In contrast, for transient, we 
need two extra transient-specific modes for the rational
programmer. Specifically, in the first mode, the rational programmer types
all blamed components.  In the second mode, the rational programmer types
the component that is added to the blame set first --- and thus can 
detect an impendance mismatch earlier in the evaluation of a program. 
As before, the two modes give
rise to two different notions of blametrail:
\begin{quote}
\it 
A  transient blametrail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
~$\conf_{i+1} \setminus \conf_i \in \{\mblame{\system, \conf_i}\}$;
\end{quote}
\noindent
and
\begin{quote}
\it 
A firstborn transient blametrail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
  $\conf_{i+1} \setminus \conf_i = \oldest{\mblame{\system, \conf_i}}$
\end{quote}
\noindent
where $\mblame{\system, \conf}$ denotes the set of components that $\conf$
blames.

\noindent
{\bf Experimental Process.} The following table summarizes the modes for the rational programmer in
different gradual typing systems:

\center
\begin{tabular}[c]{c|c|c|c}
                       & {\bf Natural}  & {\bf Transient} & {\bf Erasure}  \\
\hline 
{\bf Blame}            &       X        &                 &                \\
  {\bf Multi blame}    &                &        X        &                \\
 {\bf Firstborn blame} &                &        X        &                \\
{\bf Vanilla}          &       X        &        X        &       X        \\
{\bf Random}           &       X        &        X        &       X        \\
\end{tabular}

### 2020-07-06 writing/sample.tex
working on evaluation description + moderate reorganization

\section{How to Sample the Experimental Space} \label{sec:sample}

### 2020-07-06 writing/challenges.tex
draft of rational programmer

@@
 factor in the effect of type checks to determine whether blame helps the
-rational programmer find an impedance mismatch. A solution to this issue
+rational programmer find the cause of an impedance mismatch. A solution to this issue
 is yet another mode for the rational programmer that uses information only
-from teh underlying languages exceptions to select the next component(s) to type.
-This \emph{vanilla} mode simulates a rational programmer that ignores
+from the underlying languages exceptions to select the next component(s) to type.
+This \emph{exception} mode simulates a rational programmer that ignores
 blame and serves as the baseline for blame's value.
@@
 method needs to examine. Thus a practical application of the
-method has no option but to carefully sample this space while 
+method has no option but to carefully sample the space of scenarios while 
 making sure that it still produces reproducible conclusions.
@@
 
-\item Implement all the gradual type systems to be compared on a common
-  basis (section~\ref{sec:landscape}).
+\item Implement all three gradual type systems on a common
+  basis (section~\ref{sec:landscape});
 
@@
   systems with representative
-  type errors  (section~\ref{sec:mutate}).
+  type errors  (section~\ref{sec:mutate});
 
@@
     (ii) isolate blame from other confounding factors
-    (section~\ref{sec:rational}).
+    (section~\ref{sec:rational});
 

### 2020-07-06 writing/rational.tex
draft of rational programmer

@@
 Consider a rational programmer  that tries to debug a faulty program in
-Typed Racket with the default natural semantics. If the Typed Racekt
-typechecker detects the issue with the program, the programmer knows where
-to look for the bug; in the component that doesn't type check. However, if
-the type checker blesses the program, the programmer relies on  runtime
-type checks  to  signal an impedance mismatch between the
-program's behavior and its types. In these checks raise an error, the programmer ``follows'' blame to
+natural Typed Racket. If the Typed Racket
+type checker detects the problem, the programmer knows where
+to look for the bug; in the component that contains the type specification
+the type checker complains about. However, if
+the type checker blesses the program, 
+the programmer relies on  runtime type checks  to  signal the impedance mismatch between the
+program's behavior and its types. If these checks do raise an error that
+blames a component in the program, 
+the rational programmer ``follows'' blame to
 locate where to add a type specification next.  Hopefully, the extra
@@
 the rational programmer starts over with the latest version of the program. 
-In essence, at each iteration the
+In other words, at each iteration the
 faulty program migrates to a version with more type specifications until
-the faulty component becomes typed.
+it becomes ill-typed.
 
 Therefore, a concise way to model rational programmers is with the
-migratory paths they follow through the \emph{configuration space} of the
+migratory paths they follow through the \emph{configuration space} of a
 program. This space consists of variants of the program, dubbed
@@
 method can be adapted to so-called true gradual typing
-systems~\cite{svcb-snapl-2015} in a straightforward manner.}, each
+systems~\cite{svcb-snapl-2015} in a straightforward manner.} each
 configuration differs from another in the components that have type
@@
 the top element is $\system$ itself. In between there are all the
-mixed-type configurations. If we assume that  component
-$\faulty{\system}$ contains the fault, the type checker detects any
-configuration $\conf$ that includes $\faulty{\system}$. Any other
-configuration $\conf$ of $\system$ may blame a $\component_b$, which we denote with
-$\blame{\system,\conf}$.\footnote{Blame can result either directly from a Typed
-Racket type check or from a check in a Racket library function that raises
-a blame error or, in the absence of blame, from an
-exception that we track back to the component that causes it with the help
-of Racket's stack trace.}
+mixed-type configurations.
 
 With these definitions, the actions of the rational programmer map to an ascending chain in
-$\lattice{\system}$, dubbed a \emph{blametrail}. 
-The starting configuration $\conf_0$ of a blametrail,  dubbed the \emph{root of the
-blametrail}, is the configuration of
-$\program$ that the programmer attempts to debug. Any other
-configuration $\conf_{i+1}$ in the blametrail differs from its predecessor
+$\lattice{\system}$, dubbed a \emph{trail}. 
+The starting configuration $\conf_0$ of a trail,  dubbed the \emph{root of the
+trail}, is the configuration that the programmer attempts to debug. Any other
+configuration $\conf_{i+1}$ in the trail differs from its predecessor
 $\conf_{i}$ in that 
-the untyped component $\blame{\program, \conf_i}$ of $\conf_{i}$
- becomes typed in $\conf_{i+1}$. Putting these pieces together:
+the untyped component  that $\conf_i$ blames under the natural semantics, dubbed $\blame{\system, \conf_i}$,
+ becomes typed in $\conf_{i+1}$.\footnote{Blame can result either directly
+ from a run time type check or, in the absence of blame, from an
+ exception that we track back to the component that causes it with the
+ help of the stack trace.} Putting these pieces together:
 
@@
 \it 
-A natural blametrail
+A natural blame trail
 is a sequence of configurations $\conf_0,...\conf_n$ of a program
@@
 
-So far the definition of blametrail does not capture whether blame helps the rational
-programmer locate the bug in the root of a blametrail. For that we need 
-two properties of blametrails. A blametrail is \emph{final} if
-the rational programmer cannot extend it further. Or formally, if 
-$\blame{\program, \conf_n}$ is not defined where  $\conf_n$ is the last
-configuration of the blametrail. A blametrail is \emph{successful} 
-if it is final and  its last configuration
-$\conf_n$ contains the faulty component, that is $\faulty{\system}
-\in \conf_n$.  We can now give a precise definition of what it
-means for blame to help the rational programmer in natural Typed Racket:
+To define capture whether blame helps rational
+programmers in their search for bugs, we define when a trail is successful:
+\begin{quote}
+\it 
+A trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
+\emph{successful}  iff its last configuration $\conf_n$ does not type check.
+  \end{quote}
+  In juxtaposition:
+\begin{quote}
+\it 
+A trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
+\emph{failing}  iff $\conf_n$ type checks and $\blame{\system,\conf_n}$ is
+undefined, meaning $\conf_n$ does not result in blame. 
+ \end{quote}
+That is failing trails are those that we cannot extend
+further following blame and have not reached a configuration that reveals
+the bug at
+compile time.
+
+ We can now give a precise definition of what it
+means for blame to help the rational programmer:
 \begin{quote}
@@
   Natural blame is useful for a program $\system$ iff 
-  all natural blametrails in  $\lattice{\system}$ are successful.
+  $\lattice{\system}$ contains no failing natural blame trails.
 \end{quote}
 
-However, the definition does not give a full picture of how much blame helps the
-natural rational programmer. In particular, all blametrails in $\lattice{\system}$
-may  be successful but this may well be because of some other aspect of
-natural, such as the type checks it performs, or it may even be a result
-of pure chance. To exclude these possibilities we define two different
+However, this definition does not give a full picture of how much blame helps the
+natural rational programmer. In particular,  a $\lattice{\system}$
+may contain no failing natural blame trails because
+of the run time type checks natural performs rather than blame, or even as a result
+of pure chance. To account for these possibilities, we define two extra
 modes for the rational programmer. In the first, the natural programmer
 ignores blame and instead decides what component to type next based on
-information from the stack trace of the failed runtime type check.  This 
-vanilla rational programmer corresponds to \emph{vanilla natural
-blametrails}:
+information from the stack trace of the failed runtime type check of a
+configuration.  This exception mode of the rational programmer corresponds to \emph{natural
+exception trails}:
 \begin{quote}
 \it 
-A vanlla natural blametrail
+A natural exception trail
 is a sequence of configurations $\conf_0,...\conf_n$ of a program
 $\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
-$\conf_{i+1} \setminus \conf_i = \{\error{\system, \conf_i}\}$.
+$\conf_{i+1} \setminus \conf_i = \{\error{\system, \conf_i}\}$
+  where $\error{\system, \conf}$ denotes the component that triggers a
+an exception of the underlying language.
 \end{quote}
-where $\error{\system, \conf}$ denotes the component that triggers a
-failed runtime check according to Racket's stack trace. 
+
 In the second mode, the natural programmer
 ignores any kind of information related to the type error and randomly
-picks which component to type next.  This 
-random rational programmer corresponds to \emph{random natural
-blametrails}:
+picks which component to type next:
 \begin{quote}
 \it 
-A vanlla natural blametrail
+A natural random trail
 is a sequence of configurations $\conf_0,...\conf_n$ of a program
 $\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
-$\conf_{i+1} \setminus \conf_i = \{\random{\system, \conf_i}\}$.
+$\conf_{i+1} \setminus \conf_i = \{\random{\system, \conf_i}\}$
+where $\random{\system, \conf}$ denotes a randomly selected component of $\system$
+that is not in $\conf_i$.
 \end{quote}
-where $\random{\system, \conf}$ denotes a randomly selected of $\system$
-that is not in $\conf_i$.
+\noindent
 
-With vanilla and random blametrails, we can define the value that blame
-adds to natural Typed Rachet: 
+With exception and random trails, we can finaly define the value that blame
+adds to the natural semantics: 
 \begin{quote}
 \it
-  Natural blame  adds value to natural for a program $\system$ iff 
- iff 
-  all natural blametrails in  $\lattice{\system}$ are successful while 
-not all vanilla and random natural blametrails are successful.
+  Blame adds value to natural for a program $\system$ iff 
+  no natural blame trails in  $\lattice{\system}$ are failing while 
+ there exist at least one vanilla and one random natural trail that is
+  failing.
 \end{quote}
@@
 \noindent
-{\bf Experimental Process.}~Given a set of faulty benchmarks such as the ones
-from~section \ref{sec:mutate}, it is straightforward to turn the above definition
-into the falsfiable hypothesis of an experiment that tests whether blame
-add value to natural Typed Racket. Specifically, for each benchmark
-we first construct its configuration lattice. Then, we determine the
-configurations of the lattice that detect the fault in the
-program. This can happen either because a
- runtime type check fails or Typed Racket's typechecker
-rejects the configuration. In either case,  each of these configurations is the
-root of a natural blametrail in the configuration lattice of the benchmark. \emph{We
-hypothesize that all such natural blametrails are successful while the
-vanilla and random ones are not}. 
-To confirm or falsify the hypothesis for a benchmark, we simply extend each blametrail
- according to the mode of the corresponding rational programmer  until the blametrail
-becomes final\footnote{All blametrails are final since the lattices are
-finite.} and we check if its last configuration contains the faulty
-component of the benchmark.
+{\bf Experimental Process.}~Given a set of faulty benchmarks such as the
+ones from section~\ref{sec:mutate}, it is straightforward to turn the
+above definition into the an experimental question for natural Typed Racket. 
+Specifically, for
+each benchmark we first construct its configuration lattice. Then, we
+determine the configurations of the lattice that detect the fault in the
+program. This can happen either because a runtime type check fails or
+because Typed Racket's type checker rejects the configuration. In either
+case,  each of these configurations is the root of a natural blame trail
+in the configuration lattice of the benchmark. We ask whether
+\emph{no such natural blame trail is failing}.  Furthermore  we also
+examine whether \emph{there are failing exception and random trails that start
+from the same roots}.  To answer the questions for a
+benchmark, we simply start from our set of roots and extend each
+trail according to the corresponding mode of the rational programmer
+until no configurations can be added to the trail\footnote{All trails
+reach that point since the lattices are finite.}. Finally we check if the
+last configuration of each trail type checks or not. If it does the trail
+is successful, otherwise it is failing.
 
-\subsection{How to Generalize the Rational Programmer Beyond Natural Typed
-Racket}
+\subsection{How to Generalize the Rational Programmer Beyond Natural}
 
-Adapting the simulated rational programmer to handle transient
-and erasure Typed Racket requires taking into account the distinguishing
+Adapting the simulation of rational programmer to transient
+and erasure requires taking into account the distinguishing
 features of each semantics. As we discuss in section~\ref{sec:landscape}, 
@@
 single-component blame. At the other end of the blame spectrum, erasure
-blames no component and relies on Racket's exceptions.
+blames no component and relies on the exceptions of the underlying
+language.
 
-Hence, for an erasure rational programmer only the vanilla and random 
+Hence, for an erasure rational programmer only the exception and random 
 rational programmers are meaningful. In contrast, for transient, we 
-need two extra transient-specific modes for the rational
-programmer. Specifically, in the first mode, the rational programmer types
+need two transient-specific modes for the rational
+programmer instead of natural's blame mode. Specifically, in the first mode, the rational programmer types
 all blamed components.  In the second mode, the rational programmer types
 the component that is added to the blame set first --- and thus can 
-detect an impendance mismatch earlier in the evaluation of a program. 
+detect an impedance mismatch earlier in the evaluation of a program. 
 As before, the two modes give
-rise to two different notions of blametrail:
+rise to two different notions of trail:
 \begin{quote}
 \it 
-A  transient blametrail
+A  transient all blame trail
 is a sequence of configurations $\conf_0,...\conf_n$ of a program
 $\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
-~$\conf_{i+1} \setminus \conf_i \in \{\mblame{\system, \conf_i}\}$;
+~$\conf_{i+1} \setminus \conf_i \in \mblame{\system, \conf_i}$
+where $\mblame{\system, \conf}$ denotes the set of components that $\conf$
+blames under transient semantics; 
 \end{quote}
@@
 \it 
-A firstborn transient blametrail
+A transient firstborn  trail
 is a sequence of configurations $\conf_0,...\conf_n$ of a program
 $\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
-  $\conf_{i+1} \setminus \conf_i = \oldest{\mblame{\system, \conf_i}}$
+  $\conf_{i+1} \setminus \conf_i = \first{\mblame{\system, \conf_i}}$
+  where $\first{\mblame{\system, \conf}}$ denotes the first component 
+  the transient semantics add to the blame set for $\conf$.
 \end{quote}
-\noindent
-where $\mblame{\system, \conf}$ denotes the set of components that $\conf$
-blames.
 
 \noindent
-{\bf Experimental Process.} The following table summarizes the modes for the rational programmer in
-different gradual typing systems:
+{\bf Experimental Process.} Similar to the experiment for natural, we can
+test whether blame adds value to transient. However, we must ask two
+separate questions about blame; one for each mode of the transient
+rational programmer. Specifically, we have two derivative questions: (i)
+does blame add value to transient if the rational programmer focuses on all
+blamed components of a configuration and (ii) does blame adds value to
+transient if the rational programmer focuses on the earliest entry in a
+configuration's blame set? The two questions boil down to using 
+transient all blame trails and transient
+firstborn trails respectively and performing twice the same experiment as that
+for natural blame.
 
+In the case of erasure, instead of blame, we can examine
+the value of runtime exceptions for the rational programmer. Thus, in the
+same vain as for transient and natural we
+ask whether the configuration lattice of a given program 
+contains only successful erasure exception trail and at least one failing
+erasure random trail. Of course, we can formulate the same questions about 
+ exceptions for natural and transient and query the value of
+exceptions for these semantics too.
+
+Overall our experiment examines six questions for each benchmark: one for the value of natural
+blame, two for the value of transient blame and one for the value of
+exceptions in each of natural, transient and erasure.
+Table~\ref{fig:experiment-outline} summarizes how each question relates to different
+trails/modes of the rational programmer. For example 
+experimental question $Q_1$ asks whether blame adds value to transient and 
+our experiment uses the natural blame, exception and random trails to
+answer it.
+
+\begin{figure}
 \center
-\begin{tabular}[c]{c|c|c|c}
+{\begin{tabular}[c]{c|c|c|c}
                         & {\bf Natural}        & {\bf Transient}          & {\bf Erasure}  \\
 \hline 
-{\bf Blame}            &       X        &                 &                \\
-  {\bf Multi blame}    &                &        X        &                \\
- {\bf Firstborn blame} &                &        X        &                \\
-{\bf Vanilla}          &       X        &        X        &       X        \\
-{\bf Random}           &       X        &        X        &       X        \\
-\end{tabular}
+{\bf Blame}             &       $Q_1$          &                          &               \\
+{\bf All blame}         &                      &     $Q_2$                &               \\
+{\bf Firstborn blame}   &                      &     $Q_3$                &               \\
+{\bf Exception}         &       $Q_1$/$Q_4$    &     $Q_2$/$Q_3$/$Q_5$    &      $Q_6$    \\
+{\bf Random}            &       $Q_1$/$Q_4$    &     $Q_2$/$Q_3$/$Q_5$    &      $Q_6$    \\
+\end{tabular}}
+  \caption{ Experimental questions and rational programmer modes. $Q_1$:
+  Does natural blame add value to natural? $Q_2$: Does transient all blame
+  add value to transient?  $Q_3$: Does transient firstborn blame add value
+  to transient?  $Q_4$: Do exceptions add value to natural?  $Q_5$: Do
+  exceptions add value to transient? $Q_6$: Do exceptions add value to
+  erasure? } 
+  \label{fig:experiment-outline}
+\end{figure}
+
+\noindent
+{\bf Programmer Effort.} In addition to the value of blame, trails permit
+us to examine the effort a rational programmer spends to locate a bug. 
+For instance, the length of a successful natural blame trail tells us how
+many components the programmer has to type before reaching the faulty
+component. This idea also applies to all other modes of the rational
+programmer. The most interesting case is that for transient all blame trails. 
+In detail, given a successful transient all blame trail 
+$\conf_0, ...., \conf_n$ in a configuration lattice $\lattice{\system}$,  
+the effort of the rational programmer is the 
+number of distinct configurations in the all blame trails with root $\conf_n$ and 
+length $n$ in $\lattice{\system}$. Our experiment utilizes this notion of
+effort to analyze burden on the rational programmer for finding a
+bug when the answer to one of the experimental questions from above is
+positive for a benchmark.

### 2020-07-07 writing/challenges.tex
prose touch ups

@@
 impedance mismatch in an incremental fashion. Measuring this simulated
-programmer's behavior on a large number of inputs is the equivalent of
-collecting empirical data in a truly human-focused discipline.
+programmer's behavior on a large number of inputs then yields relevant 
+data, similar to collecting empirical data in a human-focused discipline.
 
@@
 experiment.  The mutation analysis of Lazarek et al, is, however,
-completely useless for creating impedance mismatches in a gradually
-typed context.
+completely useless for creating impedance mismatches.
 
@@
 
-Another derivative challenge of the first one is the problem of mistakes in type
+Another derivative challenge is the problem of mistakes in type
 annotations themselves.  Gradual typing enables programmers to move an untyped
@@
 of a specific type to a component, even a library, may result from a
-misunderstanding of the code and may thus be a mistake itself. Empirical
+misunderstanding of the code and may thus be a mistake itself. Anecdotal 
 evidence suggests that this scenario is quite common~\cite{sta-nt-base-types,
@@
 
-The second big challenge is due to the implementation disparity of gradual
-type systems. While the various approaches are well understood in theory,
+The second big challenge is the disparity of gradually
+typed languages. While the approaches are well understood in theory,
 they come with vastly different type systems, accommodate seemingly
@@
 comparable with Typed Racket's default natural one. The revised transient
-variant accommodates many more linguistic features and calls library code
-in a in a way that is faithful to the transient
+variant accommodates many more linguistic features and library calls
+in a way that is faithful to the transient
 semantics~\cite{vss-popl-2017}.
@@
 Making these strategies comparable is {\em not\/} an issue of language
-implementation but one about how the rational programmer behaves with regard to
-these messages. Specifically, the rational programmer must come with \emph{modes}
-that represent the alternative ways in which a programmer may interpret blame. 
-For instance, in transient, one mode has the competent programmer
-pick the oldest element of a blame set as the next component to add types. After
-all, this corresponds to the earliest point in the program's evaluation that can
+implementation but the behavior of the rational programmer.
+Specifically, the rational programmer must come with \emph{modes}
+that represent the alternative interpretations of blame. 
+For instance, a programmer using transient may
+pick the oldest element of a blame set as the next component to add types because
+it corresponds to the earliest point in the program's evaluation that can
 discover an impedance mismatch.  Another mode does not distinguish between the elements of the
-blame set and requires that the rational programmer types all blamed components
-at once. Of course, a step of the rational programmer in the latter mode
-is not as laborious as one in the former. Hence, an evaluation method should take 
-into account the effort of the rational programmer when comparing 
-different modes.
+blame set and has the rational programmer type all blamed components
+at once. Because this second kind of step 
+is more laborious than the first one, a comparison must take 
+into account the rational programmer's effort.
 
@@
 These challenges expose the subtlety of evaluating blame strategies but
-also suggest a recipe for doing so correctly:
+also suggest a recipe:
+
+%% for doing so correctly
+%% MF: what is correct about this recipe?
 

### 2020-07-07 writing/introduction.tex
prose touch ups

@@
 
-The contrast between academic and industrial systems raises the question whether
+The contrast between academic and industrial systems raises the question 
 \begin{quote}
  \it
- blame assignment adds any value to a gradually typed language, \\
+ whether blame assignment adds any value to a gradually typed language, \\
  especially for the benefit of the working programmer.
@@
 question, they seem to answer it one way or another without any scientific
-justification. Most of the time though they ignore the question altogether. The
-root of the issue is that the research community simply does not have a
-method for evaluating such questions, and given the question's practical
+justification. As a matter of fact, the research community simply does not have a
+method for evaluating blame, and given the question's practical
 relevance, this a significant problem.
@@
 This paper's first contribution is an automated method for evaluating the
-effectiveness of blame assignment strategies in the gradual typing world
-(sections ~\ref{sec:landscape},~\ref{sec:mutate},~\ref{sec:rational} and \ref{sec:sample}). 
+effectiveness of blame assignment strategies in the gradual typing world.
+%% (sections ~\ref{sec:landscape},~\ref{sec:mutate},~\ref{sec:rational} and \ref{sec:sample}). 
+%% MF: says you can't simulate a roadmap by listing all sections.
+%% MF: and for heaven's sake lookup the use of ~ in TeX
 Roughly speaking, the method, which is
@@
 Such a \emph{rational programmer} can find impedance-mismatch
-bugs by adding types to a partially typed program. That is, when the
+bugs by adding types to a partially typed program (section~\ref{sec:landscape}). That is, when the
 run-time system signals an impedance mismatch, a good blame message helps
@@
  is run again to signal another violation. Then the programmer can heed
-blame one more time to add yet another type specification. Counting the number of
+blame one more time to add yet another type specification. Weighing the number of
 steps it takes to discover the mistake in this manner is a measure of the effectiveness
@@
 
-The results of the experiment (section~\ref{sec:discussion}) validate the
-conjecture behind the work of theoreticians. A good blame assignment
+A direct interpretation of the results (section~\ref{sec:discussion}) is that
+they validate the conjecture behind the work of theoreticians. A good blame assignment
 strategy

### 2020-07-07 writing/landscape.tex
titles are questions
working on evaluation description + moderate reorganization
sec 3: draft transient typed racket

@@
-\section{The Landscape of Gradual Typing and Blame}
+\section{How to Bring Natural, Transient and Erasure Under the Same Roof} \label{sec:landscape}
 
@@
         design: where checking occurs and whether blame is tracked.
+
   \end{itemize}
+
+\subsection{Natural Typed Racket}
+
+Typed Racket extends the untyped Racket language with \emph{natural}\/ gradual
+ types~\cite{}.
+Natural types guarantee a strict separation between typed and untyped code;
+ informally, an untyped value can interact with a typed context only if the
+ value behaves in a well-typed way.
+Typed Racket implements this separation by compiling types to higher-order
+ contracts that guard all boundaries to untyped code.
+Any higher-order typed value that exits a boundary receives a protective
+ wrapper, and any untyped value that enters must pass a deep run-time type
+ check.
+These contract checks help predict the behavior of typed code, but
+ may greatly slow down a program.
+
+\subsection{Transient Typed Racket}
+
+Transient gradual types ensure that typed code cannot get stuck on a primitive
+ operation.
+This assurance can be implemented without higher-order
+ contracts~\cite{vss-popl-2017}, and therefore offers a modicum of type
+ soundness for a seemingly-low implementation and performance cost.
+
+For this experiment, we modified Typed Racket to bring transient and natural
+ under the same roof.
+Our version of the library is parameterized by a \emph{type-enforcement strategy}
+ to allow reuse of common components.
+Under the \emph{natural} strategy, our Typed Racket is identical to the latest release.
+By contrast, the \emph{transient} strategy re-uses the static type checker and
+ much of the compiler, but weakens the run-time meaning of types in two ways.
+
+The first major ingredient in transient Typed Racket is a new interpretation
+ of types as first-order shape checks.
+Each shape-check enforces a type constructor.
+For certain types, such as {\tt String} or {\tt (Vectorof Byte)}, transient can
+ re-use a predicate from the underlying Racket language.
+Other types require custom predicates; for example the type {\tt Index} of
+ valid array indices allows a range of positive exact integers.
+Many custom predicates run in constant time, but some are more involved.
+One linear-time check is for {\tt (Listof T)} for any element type {\tt T},
+ because it must distinguish a null-terminated list of pairs from improper
+ list-like structures.
+The key point, however, is that no transient check relies on
+ higher-order contract wrappers.
+
+Second, transient adds a new pass to the compiler that rewrites all typed code.
+Rewriting changes all elimination forms to shape-check the result.
+For example, if the variable {\tt f} has the static type {\tt (-> String String)}
+ then the application {\tt (f x}} must be rewritten to check for a string result
+ because {\tt f} could be an untyped function.
+Rewriting also inserts hooks to update a global blame heap.
+When a value crosses a boundary, the blame heap records a cast assumption.
+Elimination forms and other operations record smaller path assumptions.
+If a transient check ever fails, the language follows path assumptions and
+ reports a set of casts/boundaries to the programmer.

### 2020-07-07 writing/landscape.tex
sec 3: fix cites, add missing paren

@@
 Typed Racket extends the untyped Racket language with \emph{natural}\/ gradual
- types~\cite{}.
+ types~\cite{tf-popl-2008,tfffgksst-snapl-2017}.
 Natural types guarantee a strict separation between typed and untyped code;
@@
 For example, if the variable {\tt f} has the static type {\tt (-> String String)}
- then the application {\tt (f x}} must be rewritten to check for a string result
+ then the application {\tt (f x)} must be rewritten to check for a string result
  because {\tt f} could be an untyped function.

### 2020-07-07 writing/rational.tex
ogranizing section 5

@@
-\section{How to Simulate the Rational Programmer} \label{sec:rational}
+\section{How to Simulate the Rational Programmer}
 
-Consider a rational programmer  that tries to debug a faulty program in
-natural Typed Racket. If the Typed Racket
-type checker detects the problem, the programmer knows where
-to look for the bug; in the component that contains the type specification
-the type checker complains about. However, if
-the type checker blesses the program, 
-the programmer relies on  runtime type checks  to  signal the impedance mismatch between the
-program's behavior and its types. If these checks do raise an error that
-blames a component in the program, 
-the rational programmer ``follows'' blame to
-locate where to add a type specification next.  Hopefully, the extra
-types are sufficient to turn the run time error into a static one. Otherwise,
-the rational programmer starts over with the latest version of the program. 
-In other words, at each iteration the
-faulty program migrates to a version with more type specifications until
-it becomes ill-typed.
+Code migration is the raison d'\^etre of Typed Racket, which is the community
+uses ``migratory typing'' instead of gradual typing. Roughly speaking a
+programmer uses a gradually typed language to incrementally add types to a code
+base. From this perspective, a code base consists of some number of components,
+say modules, and a programmer converts a component at a time from its untyped
+state to a typed one.
 
-Therefore, a concise way to model rational programmers is with the
-migratory paths they follow through the \emph{configuration space} of a
-program. This space consists of variants of the program, dubbed
-\emph{configurations}, that differ in their type specifications. In the 
-setting of Typed Racket's migratory typing,\footnote{The
-method can be adapted to so-called true gradual typing
-systems~\cite{svcb-snapl-2015} in a straightforward manner.} each
-configuration differs from another in the components that have type
-specifications. Hence, formally, we can represent a program
-$\system$ as a set of components $\set{\component}$ and each
-configuration $\conf$  as a subset of $\set{\component}$ that contains
-exactly the components of  $\system$ that are typed in $\conf$. Naturally
-the configurations of  $\system$ are ordered by the subset relation and
-thus form a lattice $\lattice{\system}$ with $2^{\size{\system}}$ elements.
-The bottom configuration of $\lattice{\system}$ is always $\emptyset$ and
-the top element is $\system$ itself. In between there are all the
-mixed-type configurations.
+Based on this idea, a metric for migratory paths is a concise way to measure a
+rational programmer. The components of the program create a \emph{configuration
+space}, like the one introduced by~\citet{tfgnvf-popl-2016}. This space consists
+of variants of the program, dubbed \emph{configurations}, that differ in their
+type specifications. In the setting of migratory typing,\footnote{The method can
+be adapted to so-called true gradual typing systems~\cite{svcb-snapl-2015} in a
+straightforward manner.} each configuration differs from another in the subset
+of typed components. Formally, a program $\system$ is a set of components
+$\set{\component}$, and a configuration $\conf$ is a subset of
+$\set{\component}$ . These configurations of $\system$ are ordered by the subset
+relation and thus form a lattice $\lattice{\system}$ with $2^{\size{\system}}$
+elements.  The bottom configuration of $\lattice{\system}$ is always $\emptyset$
+and the top element is $\system$ itself. In between are all the mixed-type
+configurations.
 
-With these definitions, the actions of the rational programmer map to an ascending chain in
-$\lattice{\system}$, dubbed a \emph{trail}. 
-The starting configuration $\conf_0$ of a trail,  dubbed the \emph{root of the
-trail}, is the configuration that the programmer attempts to debug. Any other
+With these definitions, the actions of the rational programmer map to an
+ascending chain---dubbed a \emph{trail}---in $\lattice{\system}$.  The starting
+configuration $\conf_0$ of a trail---also referred to the \emph{root} (of the
+trail)---is the configuration that the programmer attempts to debug. Any other
 configuration $\conf_{i+1}$ in the trail differs from its predecessor
-$\conf_{i}$ in that 
-the untyped component  that $\conf_i$ blames under the natural semantics, dubbed $\blame{\system, \conf_i}$,
- becomes typed in $\conf_{i+1}$.\footnote{Blame can result either directly
- from a run time type check or, in the absence of blame, from an
- exception that we track back to the component that causes it with the
- help of the stack trace.} Putting these pieces together:
+$\conf_{i}$ in that the untyped component that $\conf_i$ blames under the
+various semantics, dubbed $\blame{\system, \conf_i}$, becomes typed in
+$\conf_{i+1}$.\footnote{Blame can result either directly from a run time type
+check or, in the absence of blame, from an exception that we track back to the
+component that causes it with the help of the stack trace.}
+
+The next three subsections specialize the idea to the various semantics of
+gradual typing. Once this machinery is in place, we use the fourth subsection to
+state the precise experimental research questions that correspond to the
+high-level question on the first page. The final subsection explains how the
+rational programmer's effort is measured.
+
+%% -----------------------------------------------------------------------------
+\subsection{Modes of Natural Type Programming} \label{sub:natural}
+
+Putting these pieces together:
 
@@
 
-\subsection{How to Generalize the Rational Programmer Beyond Natural}
+%% -----------------------------------------------------------------------------
+\subsection{Modes of Optional Type Programming} \label{sub:transient}
+
+The case of erasure semantics is analogous to the third mode of natural.  Since
+gradually typed languages with erasure semantics do not come with blame
+assignment, a rational programmer can only hope that the underlying safety
+checks and their stack traces are helpful.  Thus, in the same veub as for
+natural we ask whether the configuration lattice of a given program contains
+only successful erasure exception trail and at least one failing erasure random
+trail.
+
+%% -----------------------------------------------------------------------------
+\subsection{Modes of Transient Type Programming} \label{sub:transient}
 
 Adapting the simulation of rational programmer to transient
-and erasure requires taking into account the distinguishing
+requires taking into account the distinguishing
 features of each semantics. As we discuss in section~\ref{sec:landscape}, 
 transient blames a set of components in contrast to natural's
-single-component blame. At the other end of the blame spectrum, erasure
-blames no component and relies on the exceptions of the underlying
-language.
+single-component blame.
 
-Hence, for an erasure rational programmer only the exception and random 
-rational programmers are meaningful. In contrast, for transient, we 
+For transient, we 
 need two transient-specific modes for the rational
@@
 
-In the case of erasure, instead of blame, we can examine
-the value of runtime exceptions for the rational programmer. Thus, in the
-same vain as for transient and natural we
-ask whether the configuration lattice of a given program 
-contains only successful erasure exception trail and at least one failing
-erasure random trail. Of course, we can formulate the same questions about 
- exceptions for natural and transient and query the value of
-exceptions for these semantics too.
+%% -----------------------------------------------------------------------------
+\subsection{Questions}
 
-Overall our experiment examines six questions for each benchmark: one for the value of natural
-blame, two for the value of transient blame and one for the value of
-exceptions in each of natural, transient and erasure.
-Table~\ref{fig:experiment-outline} summarizes how each question relates to different
-trails/modes of the rational programmer. For example 
-experimental question $Q_1$ asks whether blame adds value to transient and 
-our experiment uses the natural blame, exception and random trails to
-answer it.
+In sum, our experiment examines four questions for each benchmark:
+\begin{itemize}
+\item[$Q_1$] Does natural blame add value to natural?
+
+\item[$Q_2$] Does transient all blame add value to transient?
+
+\item[$Q_3$] Does transient firstborn blame add value to transient?
+
+\item[$Q_*$] Do exceptions add value? For any of the systems? 
+\end{itemize}
+
+Table~\ref{fig:experiment-outline} summarizes how each question relates to
+different trails/modes of the rational programmer. For example experimental
+question $Q_1$ asks whether blame adds value to transient and our experiment
+uses the natural blame, exception and random trails to answer it.
 
@@
 \center
-{\begin{tabular}[c]{c|c|c|c}
-                        & {\bf Natural}        & {\bf Transient}          & {\bf Erasure}  \\
-\hline 
+{\begin{tabular}[ht]{l|c|c|c}
+%% ---------------------------------------------------------------------------------------------------------------
+                        & {\bf Natural}        & {\bf Transient}          & {\bf Erasure} \\ \hline 
+%% ---------------------------------------------------------------------------------------------------------------
 {\bf Blame}             &       $Q_1$          &                          &               \\
@@
 {\bf Firstborn blame}   &                      &     $Q_3$                &               \\
-{\bf Exception}         &       $Q_1$/$Q_4$    &     $Q_2$/$Q_3$/$Q_5$    &      $Q_6$    \\
-{\bf Random}            &       $Q_1$/$Q_4$    &     $Q_2$/$Q_3$/$Q_5$    &      $Q_6$    \\
+{\bf Exception}         &       $Q_1$/$Q_*$    &     $Q_2$/$Q_3$/$Q_*$    &      $Q_*$    \\
+{\bf Random}            &       $Q_1$/$Q_*$    &     $Q_2$/$Q_3$/$Q_*$    &      $Q_*$    \\
 \end{tabular}}
-  \caption{ Experimental questions and rational programmer modes. $Q_1$:
-  Does natural blame add value to natural? $Q_2$: Does transient all blame
-  add value to transient?  $Q_3$: Does transient firstborn blame add value
-  to transient?  $Q_4$: Do exceptions add value to natural?  $Q_5$: Do
-  exceptions add value to transient? $Q_6$: Do exceptions add value to
-  erasure? } 
+  \caption{ Experimental questions and rational programmer modes.}
   \label{fig:experiment-outline}
@@
 
-\noindent
-{\bf Programmer Effort.} In addition to the value of blame, trails permit
-us to examine the effort a rational programmer spends to locate a bug. 
-For instance, the length of a successful natural blame trail tells us how
-many components the programmer has to type before reaching the faulty
-component. This idea also applies to all other modes of the rational
-programmer. The most interesting case is that for transient all blame trails. 
-In detail, given a successful transient all blame trail 
-$\conf_0, ...., \conf_n$ in a configuration lattice $\lattice{\system}$,  
-the effort of the rational programmer is the 
-number of distinct configurations in the all blame trails with root $\conf_n$ and 
-length $n$ in $\lattice{\system}$. Our experiment utilizes this notion of
-effort to analyze burden on the rational programmer for finding a
-bug when the answer to one of the experimental questions from above is
-positive for a benchmark.
+%% -----------------------------------------------------------------------------
+
+\subsection{Programmer Effort}
+
+In addition to the value of blame, trails permit us to examine the effort a
+rational programmer spends to locate a bug.  For instance, the length of a
+successful natural blame trail tells us how many components the programmer has
+to type before reaching the faulty component. This idea also applies to all
+other modes of the rational programmer. The most interesting case is that for
+transient all blame trails.  In detail, given a successful transient all blame
+trail $\conf_0, ...., \conf_n$ in a configuration lattice $\lattice{\system}$,
+the effort of the rational programmer is the number of distinct configurations
+in the all blame trails with root $\conf_n$ and length $n$ in
+$\lattice{\system}$. Our experiment utilizes this notion of effort to analyze
+burden on the rational programmer for finding a bug when the answer to one of
+the experimental questions from above is positive for a benchmark.

### 2020-07-08 writing/challenges.tex
revising rational section

@@
 
-The most serious challenge for comparing blame assignment strategies concerns 
+The most serious challenge for evaluating blame concerns 
 the nature of this research. Unlike ordinary
@@
 
-Another derivative challenge is the problem of mistakes in type
-annotations themselves.  Gradual typing enables programmers to move an untyped
-code base into the typed realm on an incremental basis. The retroactive addition
-of a specific type to a component, even a library, may result from a
-misunderstanding of the code and may thus be a mistake itself. Anecdotal 
-evidence suggests that this scenario is quite common~\cite{sta-nt-base-types,
-incorrect-ts, wmwz-ecoop-2017}, and hence must be addressed by any evaluation
-method. Conversely, the method not only demands mutators for code that cause
-good impedance mismatches, it also calls for a set of mutators that ``break''
-type annotations without breaking the code itself.
+%Another derivative challenge is the problem of mistakes in type
+%annotations themselves.  Gradual typing enables programmers to move an untyped
+%code base into the typed realm on an incremental basis. The retroactive addition
+%of a specific type to a component, even a library, may result from a
+%misunderstanding of the code and may thus be a mistake itself. Anecdotal 
+%evidence suggests that this scenario is quite common~\cite{sta-nt-base-types,
+%incorrect-ts, wmwz-ecoop-2017}, and hence must be addressed by any evaluation
+%method. Conversely, the method not only demands mutators for code that cause
+%good impedance mismatches, it also calls for a set of mutators that ``break''
+%type annotations without breaking the code itself.
 
@@
 incompatible language features, and apply to radically different coding
-styles in the untyped world.  To eliminate these differences as variables
-of a comparison, the application of an evaluation method to different
+styles in the untyped world.  To eliminate the effect of these differences 
+on its results, the application of an evaluation method to different
 gradual typing systems must use the same test programs and the same type
-system. Otherwise, results for different gradual typing systems are not
-comparable.  For instance, the case study presented here builds on
+system. Otherwise, results for different gradual typing systems 
+do not share a common baseline and, therefore, do not lend support to any
+conclusions about whether the value of blame changes from one system to
+another. For instance, the case study presented here builds on
 Greenman and Felleisen's transient variant for functional Typed
@@
 
-Beyond the implementation issue, a comparison also faces the problem that
+Beyond the implementation issue, the method also faces the problem that
 the three variants---the natural semantics, the transient semantics, and
@@
 
-Making these strategies comparable is {\em not\/} an issue of language
-implementation but the behavior of the rational programmer.
+Supporting these strategies is {\em not\/} just an issue of language
+implementation but affects greatly the behavior of the rational programmer.
 Specifically, the rational programmer must come with \emph{modes}
@@
 at once. Because this second kind of step 
-is more laborious than the first one, a comparison must take 
+is more laborious than the first one, the analysis of the results must take 
 into account the rational programmer's effort.
@@
 application of the evaluation method to a gradual typing system has to
-factor in the effect of type checks to determine whether blame helps the
+factor in the effect of their different type checks to determine whether blame helps the
 rational programmer find the cause of an impedance mismatch. A solution to this issue

### 2020-07-08 writing/mutate.tex
revising rational section

@@
-\section{How to Make Benchmarks} \label{sec:mutate}
+\section{How to Make Benchmarks for the Value of Blame} \label{sec:mutate}
 
@@
 mutation analysis from well-typed programs.
+
+\begin{figure}
+\begin{tabular}{p{1.5cm} | p{10cm} }
+  {\bf  name} & {\bf description (author)}  \\
+
+\hline
+
+acquire & Board game simulation (M. Felleisen)  \\%[1em]
+
+\hline
+  gregor & Utilities for calendar dates (J. Zeppieri) \\%[1em]
+
+\hline
+kcfa & Functional implementation of 2CFA for the lambda calculus (M. Might) \\%[1em]
+
+\hline
+quadT & Converter from S-expresion source code to PDF format (M. Butterick)\\%[1em]
+
+\hline
+quadU & Converter from S-expresion source code to PDF format  (B. Greenman) \\%[1em]
+
+\hline
+snake & Functional implementation of the  Snake video game (D. Van Horn) \\%[1em]
+
+\hline
+suffixtree & Algorithm for common longest subsequences between strings. (D. Yoo) \\%[1em]
+
+\hline
+synth & Converter of description of notes and drum beats to WAV format (V. St-Amour \& N. Toronto) \\%[1em]
+
+\hline
+take5 & Card game simulator (M.Felleisen)  \\%[1em]
+
+\hline
+tetris & Functional implementation of Tetris (D. Van Horn) \\%[1em]
+
+\end{tabular}
+  \caption{Benchmarks summary.}
+  \label{table:benchmark-descriptions}
+\end{figure}
+
+\begin{figure}
+  \begin{tabular}{p{2cm} | p{5cm}  | p{5.5cm} }
+    {\bf operator} & {\bf description} & {\bf example} \\
+
+\hline
+arithmetic 
+& Swap an arithmetic operator with another one of the same (or greater)
+    arity and a possibly different result type. 
+    & \texttt{+} $\rightarrow$ \texttt{-} \\
+\hline
+
+boolean 
+& Swap \texttt{and} and \texttt{or}. Due to Racket's truthiness it
+    possibly changes the type of the operation's result. 
+& \texttt{and} $\rightarrow$ \texttt{or} \\
+
+\hline 
+constant
+& Swap a literal constant with another of the same value but different type.
+& \texttt{5.6} $\rightarrow$ \texttt{5.6+0.0i} \\
+\hline 
+deletion
+    & Deletes the result expression of a \texttt{begin} or \texttt{begin0}
+    sequence.
+    & \texttt{(begin x y z)} $\rightarrow$ \texttt{(begin x y)} \\
+\hline     
+list
+    & Swap \texttt{car} and \texttt{cdr}.
+    & \texttt{car} $\rightarrow$ \texttt{cdr} \\
+ \hline    
+class:public
+    & Make a public method private and vice versa.
+    & 
+    \begin{tabular}[t]{@{}l}
+      \texttt{(class object\%}\\
+        \texttt{\,\,(define/public (m x) x))}\\
+       \rightarrow\\ 
+      \texttt{(class object\%}\\
+        \texttt{\,\,(define/private (m x) x))}\\
+    \end{tabular}\\
+\hline
+
+class:super
+    & Remove \texttt{super-new} calls from class definitions.
+    &  \begin{tabular}[t]{@{}l}
+      \texttt{(class object\% (super-new))}\\
+       \rightarrow\\ 
+      \texttt{(class object\% (void))}\\
+    \end{tabular}\\
+\hline
+
+conditional
+    & Negate conditional test expressions.
+    &  \begin{tabular}[t]{@{}l}
+      \texttt{(cond [(= x 0) 42] ...)}\\ 
+       \rightarrow\\ 
+      \texttt{(cond [(not (= x 0)) 42] ...)}
+    \end{tabular}\\
+\hline
+
+class:parent
+    & Replace the parent of classes with \texttt{object\%}.
+    &  \begin{tabular}[t]{@{}l}
+     \texttt{(class foo\% ...)}\\
+       \rightarrow\\ 
+      \texttt{(class object\% ...)}\\
+    \end{tabular}\\
+\hline
+
+class:init
+  & Swap default values of class fields.
+   &  \begin{tabular}[t]{@{}l}
+     \texttt{(class object\%}\\
+     \texttt{\,\,(field [a 5] [b "hello"]))}\\
+       \rightarrow\\ 
+      \texttt{(class object\%}\\
+     \texttt{\,\,(field [a "hello"] [b 5]))}\\
+    \end{tabular}\\
+\hline
+
+position
+ & Swap the position of subexpressions.
+  &  \begin{tabular}[t]{@{}l}
+     \texttt{(f a 42 "b" 0)}\\
+       \rightarrow\\ 
+      \texttt{(f a 42 0 "b")}
+    \end{tabular}\\
+\hline
+
+class:method
+ & Add an extra trivial public method to a class.
+ &  \begin{tabular}[t]{@{}l}
+   \texttt{(class object\%  ...)}\\
+       \rightarrow\\ 
+     \texttt{(class object\%}\\
+     \texttt{\,\,(define/public (extra x) x)}\\
+     \texttt{\,\,...)}
+    \end{tabular}\\
+\hline
+
+id
+ & Swaps a use of a top-level identifier defined with with another defined
+ in the same module.
+ &\texttt{(f ...)} \rightarrow \texttt{(g ...)}
+
+\end{tabular}
+  \caption{Summary of code mutation operators.}
+  \label{table:mutation-ops}
+\end{figure}

### 2020-07-08 writing/rational.tex
revising rational section

@@
 
-Code migration is the raison d'\^etre of Typed Racket, which is the community
+Code migration is the raison d'\^etre of Typed Racket, which is why the community
 uses ``migratory typing'' instead of gradual typing. Roughly speaking a
@@
 
-Based on this idea, a metric for migratory paths is a concise way to measure a
-rational programmer. The components of the program create a \emph{configuration
+Based on this idea, a metric for migratory paths is a concise way to
+evaluate the effectiveness of a
+rational programmer. The components of each program create a \emph{configuration
 space}, like the one introduced by~\citet{tfgnvf-popl-2016}. This space consists
@@
 elements.  The bottom configuration of $\lattice{\system}$ is always $\emptyset$
-and the top element is $\system$ itself. In between are all the mixed-type
+and the top element is $\system$ itself. In between are all the mix-typed
 configurations.
@@
 state the precise experimental research questions that correspond to the
-high-level question on the first page. The final subsection explains how the
-rational programmer's effort is measured.
+high-level question on the first page and to explain our experimental
+process. The final subsection extends the development with a discussion of
+how we can measure the rational programmer's effort.
 
 %% -----------------------------------------------------------------------------
-\subsection{Modes of Natural Type Programming} \label{sub:natural}
+\subsection{Modes of the Natural Rational Programmer} \label{sub:natural}
 
-Putting these pieces together:
+In natural Typed Racket, rational programmers rely on blame when they
+attempt to debug a mix-typed programmer; they add type specifications to
+the component blame points. In terms of configurations and trails, this
+translates to extending a trail with a configuration that adds to the
+last configuration of the trail so far the component the latter configuration
+blames.  We call trails constructed in this manner \emph{natural blame
+trails} and they capture the essence of the natural rational programmer:
 
@@
 is a sequence of configurations $\conf_0,...\conf_n$ of a program
-$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
-$\conf_{i+1} \setminus \conf_i = \{\blame{\system, \conf_i}\}$.
+$\system$  such that for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
+  $\conf_{i+1} \setminus \conf_i =\{\blame{\system, \conf_i}\}$ where
+  $\blame{\system, \conf}$ denotes the component that $\conf$ blames.
 \end{quote}
 
-To define capture whether blame helps rational
-programmers in their search for bugs, we define when a trail is successful:
+Natural rational programmers migrate faulty programs along natural blame
+trails in order to locate bugs. Specifically, assuming that the types 
+they add during migration are correct, when a buggy component migrates
+from untyped to typed, the typechecker detects the impedance mismatch
+between the broken code of the component and its types and signals a type
+error. Hence, trails that end at an ill-typed configuration successfully 
+pinpoint the location of bugs:
+In formal terms:
 \begin{quote}
 \it 
-A trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
+A natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
 \emph{successful}  iff its last configuration $\conf_n$ does not type check.
@@
 \it 
-A trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
-\emph{failing}  iff $\conf_n$ type checks and $\blame{\system,\conf_n}$ is
-undefined, meaning $\conf_n$ does not result in blame. 
+A natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
+\emph{failing}  iff $\conf_n$ type checks and the trail cannot be extended
+  further. 
  \end{quote}
-That is failing trails are those that we cannot extend
-further following blame and have not reached a configuration that reveals
-the bug at
-compile time.
+\noindent 
+That is failing natural blame trails are those that have not reached a configuration that reveals
+the bug at compile time and their last configuration does not produce
+blame. Thus the rational programmer receives no hint from the gradual type
+system about how to further migrate the faulty program in search for the
+bug.
 
@@
 \it
-  Natural blame is useful for a program $\system$ iff 
+  Natural blame is always useful for a program $\system$ iff 
   $\lattice{\system}$ contains no failing natural blame trails.
+
+   Natural blame is sometimes useful for a program $\system$ iff 
+  $\lattice{\system}$ contains successful natural blame trails.
+
+   Natural blame is never useful for a program $\system$ iff 
+  $\lattice{\system}$ contains no successful natural blame trails.
+
 \end{quote}
 
-However, this definition does not give a full picture of how much blame helps the
-natural rational programmer. In particular,  a $\lattice{\system}$
-may contain no failing natural blame trails because
-of the run time type checks natural performs rather than blame, or even as a result
-of pure chance. To account for these possibilities, we define two extra
-modes for the rational programmer. In the first, the natural programmer
-ignores blame and instead decides what component to type next based on
-information from the stack trace of the failed runtime type check of a
-configuration.  This exception mode of the rational programmer corresponds to \emph{natural
-exception trails}:
+These definitions help a rational programmer gauge whether it pays off to heed blame
+when debugging \system. 
+However, they do not answer whether blame is a critical piece of 
+the rational programmer's process.  For instance, 
+typing a random component   could be equally useful as typing the blamed one. 
+To account for this situation, we devise a new
+mode of the natural rational programmer that 
+ follows a random migration process:
+\begin{quote}
+\it 
+A natural random trail
+is a sequence of configurations $\conf_0,...\conf_n$ of a program
+$\system$  such that for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
+$\conf_{i+1} \setminus \conf_i = \{\random{\system, \conf_i}\}$
+where $\random{\system, \conf}$ denotes a randomly selected component of $\system$
+that is not in $\conf_i$.
+\end{quote}
+\indent
+With blame and random natural trails, we can ask whether 
+blame adds value over random migration: 
+\begin{quote}
+\it
+  Blame adds value to natural for a program $\system$ over pure chance iff 
+  no natural blame trails in  $\lattice{\system}$ are failing while 
+ there exists at least one random natural trail that is
+  failing.
+\end{quote}
+
+Even if the success of the natural rational programmer is not by pure
+chance, it could well be because the natural semantics perform runtime type checks at
+the right time. In other words, one rational programmer that 
+heeds blame could be equally successful as one that only uses information from the exceptions of failed 
+runtime type checks. Similar to before, in response we
+define the exception mode of the rational programmer:
 \begin{quote}
@@
 is a sequence of configurations $\conf_0,...\conf_n$ of a program
-$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
+$\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
 $\conf_{i+1} \setminus \conf_i = \{\error{\system, \conf_i}\}$
@@
 \end{quote}
-
-In the second mode, the natural programmer
-ignores any kind of information related to the type error and randomly
-picks which component to type next:
+\noindent
+We use the new mode to describe what it means for blame to add value over both exceptions and pure
+chance:
 \begin{quote}
 \it
-A natural random trail
-is a sequence of configurations $\conf_0,...\conf_n$ of a program
-$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
-$\conf_{i+1} \setminus \conf_i = \{\random{\system, \conf_i}\}$
-where $\random{\system, \conf}$ denotes a randomly selected component of $\system$
-that is not in $\conf_i$.
+  Blame adds value to natural for a program $\system$ over pure chance and
+  exceptions iff no natural blame trails in  $\lattice{\system}$ are failing while 
+ there exist at least one random natural trail and one natural exception trail
+ that are
+  failing.
+\end{quote}
+\noindent
+If blame does add value to natural for a program $\system$ 
+over pure chance and exceptions then we can finally
+deduce that, for debugging $\system$, blame is a valuable feature of the
+natural semantics.
+
+%% -----------------------------------------------------------------------------
+\subsection{Modes of the Erasure Rational Programmer} \label{sub:erasure}
+
+ Since gradually typed languages with erasure semantics do not come with blame
+assignment, a rational programmer can only hope that the underlying safety
+checks and their exceptions are helpful.  Thus, adapting the corresponding
+definitions from natural to erasure, we ask whether the effect of erasure
+exception in debugging a program $\system$ goes beyond pure chance:
+
+\begin{quote}
+\it
+  Exceptions adds value to erasure for a program $\system$ over pure chance
+  iff no erasure exception trails in  $\lattice{\system}$ are failing while 
+ there exists at least one random erasure trail that is failing.
 \end{quote}
@@
 
-With exception and random trails, we can finaly define the value that blame
-adds to the natural semantics: 
-\begin{quote}
-\it
-  Blame adds value to natural for a program $\system$ iff 
-  no natural blame trails in  $\lattice{\system}$ are failing while 
- there exist at least one vanilla and one random natural trail that is
-  failing.
-\end{quote}
+%% -----------------------------------------------------------------------------
+\subsection{Modes of the Transient Rational Programmer} \label{sub:transient}
 
-\noindent
-{\bf Experimental Process.}~Given a set of faulty benchmarks such as the
-ones from section~\ref{sec:mutate}, it is straightforward to turn the
-above definition into the an experimental question for natural Typed Racket. 
-Specifically, for
-each benchmark we first construct its configuration lattice. Then, we
-determine the configurations of the lattice that detect the fault in the
-program. This can happen either because a runtime type check fails or
-because Typed Racket's type checker rejects the configuration. In either
-case,  each of these configurations is the root of a natural blame trail
-in the configuration lattice of the benchmark. We ask whether
-\emph{no such natural blame trail is failing}.  Furthermore  we also
-examine whether \emph{there are failing exception and random trails that start
-from the same roots}.  To answer the questions for a
-benchmark, we simply start from our set of roots and extend each
-trail according to the corresponding mode of the rational programmer
-until no configurations can be added to the trail\footnote{All trails
-reach that point since the lattices are finite.}. Finally we check if the
-last configuration of each trail type checks or not. If it does the trail
-is successful, otherwise it is failing.
-
-%% -----------------------------------------------------------------------------
-\subsection{Modes of Optional Type Programming} \label{sub:transient}
-
-The case of erasure semantics is analogous to the third mode of natural.  Since
-gradually typed languages with erasure semantics do not come with blame
-assignment, a rational programmer can only hope that the underlying safety
-checks and their stack traces are helpful.  Thus, in the same veub as for
-natural we ask whether the configuration lattice of a given program contains
-only successful erasure exception trail and at least one failing erasure random
-trail.
-
-%% -----------------------------------------------------------------------------
-\subsection{Modes of Transient Type Programming} \label{sub:transient}
-
-Adapting the simulation of rational programmer to transient
-requires taking into account the distinguishing
-features of each semantics. As we discuss in section~\ref{sec:landscape}, 
-transient blames a set of components in contrast to natural's
-single-component blame.
-
-For transient, we 
-need two transient-specific modes for the rational
-programmer instead of natural's blame mode. Specifically, in the first mode, the rational programmer types
-all blamed components.  In the second mode, the rational programmer types
-the component that is added to the blame set first --- and thus can 
-detect an impedance mismatch earlier in the evaluation of a program. 
-As before, the two modes give
-rise to two different notions of trail:
+In contrast to erasure, transient does come with blame. As we discuss in
+section~\ref{sec:landscape} though, distinctly from natural blame,
+transient may blame more than one components. This peculiarity of 
+transient raises the question of how the rational programmer should react
+when the language produces a blame set. Our answer is that the rational
+programmer has least two reasonable options.  The first option is the rational programmer types
+all blamed components at once.  In the second option, the rational
+programmer selects the component that is added to the blame set first and types
+only that --- after all the types of this first component should be able
+to  detect an impedance mismatch earlier in the evaluation of a program. 
+As before, the two modes give rise to two different notions of trail:
 \begin{quote}
@@
 $\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
-~$\conf_{i+1} \setminus \conf_i \in \mblame{\system, \conf_i}$
+~$\conf_{i+1} \setminus \conf_i = \mblame{\system, \conf_i}$
 where $\mblame{\system, \conf}$ denotes the set of components that $\conf$
@@
 \it 
-A transient firstborn  trail
+A transient firstborn blame trail
 is a sequence of configurations $\conf_0,...\conf_n$ of a program
@@
 \end{quote}
-
-\noindent
-{\bf Experimental Process.} Similar to the experiment for natural, we can
-test whether blame adds value to transient. However, we must ask two
-separate questions about blame; one for each mode of the transient
-rational programmer. Specifically, we have two derivative questions: (i)
-does blame add value to transient if the rational programmer focuses on all
-blamed components of a configuration and (ii) does blame adds value to
-transient if the rational programmer focuses on the earliest entry in a
-configuration's blame set? The two questions boil down to using 
-transient all blame trails and transient
-firstborn trails respectively and performing twice the same experiment as that
-for natural blame.
+\noindent As for natural we use the transient all blame trails and 
+transient firstborn  trails to define when  transient all blame and 
+transient firstborn add value to transient over exceptions and pure
+chance.
 
 %% -----------------------------------------------------------------------------
-\subsection{Questions}
+\subsection{Experimental Process.}
 
-In sum, our experiment examines four questions for each benchmark:
+Trails and their properties give us the tools for a rigorous examination
+of blame for natural, transient and erasure Typed Racket and the 
+mutants of section~\ref{sec:mutate}. In particular, our experiment examines 
+four questions for each mutant:
 \begin{itemize}
-\item[$Q_1$] Does natural blame add value to natural?
+\item[$Q_1$] Is blame is valuable for natural?
 
-\item[$Q_2$] Does transient all blame add value to transient?
+\item[$Q_2$] Is all blame valuable for transient?
 
-\item[$Q_3$] Does transient firstborn blame add value to transient?
+\item[$Q_3$] Is firstborn blame valuable for transient?
 
-\item[$Q_*$] Do exceptions add value? For any of the systems? 
+\item[$Q_*$] Are plain exceptions valuable for
+  erasure? For the other two systems? 
 \end{itemize}
@@
 different trails/modes of the rational programmer. For example experimental
-question $Q_1$ asks whether blame adds value to transient and our experiment
+question $Q_1$ asks whether blame is valuable for natural and our experiment
 uses the natural blame, exception and random trails to answer it.
@@
 
+In detail, to answer $Q_1$ for a mutant, we first construct the mutant's configuration 
+lattice. Then, we determine the configurations of the lattice that detect the fault in the
+program. This can happen either because Typed Racket's type checker
+outright rejects the configuration or because natural Typed Racket's runtime 
+performs a type check that fails. In either
+case,  each of these configurations is the root of a trail that we 
+extend according the natural blame . 
+If no configurations can be added to the trail\footnote{All trails
+reach that point since the lattices are finite.}, we check if the
+last configuration of the trail type checks or not. If it does the trail
+is successful, otherwise it is failing. We repeat the same process for the
+roots two more times; once following the natural exception rational programmer and 
+once the natural random rational programmer. Finally, we use the
+success/failure results for the trails to check if they meet the condition
+for natural blame to add value to natural over exceptions and pure chance.
+The process is analogues for the other questions and their corresponding
+modes of the rational programmer.
+
 \subsection{Programmer Effort}
@@
 In addition to the value of blame, trails permit us to examine the effort a
-rational programmer spends to locate a bug.  For instance, the length of a
-successful natural blame trail tells us how many components the programmer has
-to type before reaching the faulty component. This idea also applies to all
-other modes of the rational programmer. The most interesting case is that for
-transient all blame trails.  In detail, given a successful transient all blame
-trail $\conf_0, ...., \conf_n$ in a configuration lattice $\lattice{\system}$,
-the effort of the rational programmer is the number of distinct configurations
-in the all blame trails with root $\conf_n$ and length $n$ in
-$\lattice{\system}$. Our experiment utilizes this notion of effort to analyze
-burden on the rational programmer for finding a bug when the answer to one of
-the experimental questions from above is positive for a benchmark.
+rational programmer spends to locate a bug. As a metric for the effort, 
+we use the number of components the rational programmer has to migrate 
+from untyped to typed to find the bug. We can easily and uniformly calculate the value
+of the metric for any successful trail in any mode of the rational
+programmer (including transient blame all)  as the  size of the difference 
+between the trail's root and its final configuration.
+
+Programmer effort can help shed some light to the relative effectiveness of the
+three gradual typing systems. For example, consider a configuration of the
+lattice of one of our mutants. The configuration corresponds to a
+ scenario that the rational programmer attempts to debug. If the
+configuration results in blame both under, for instance, the natural and transient semantics,
+then the natural blame rational programmer and the two transient blame
+rational programmers (all and firstborn) can compete to see which one
+reaches a configuration that doesn't type check with less effort. In
+general, the effort of trails of different modes that have the same
+root is a metric that allows us to compare the effectiveness of these
+modes even if they correspond to different gradual typing systems.
+Unfortunately, statistically speaking, such comparisons between two modes do not generalize 
+beyond the successful trails that share their root in our experiments.
+Still they allow us to quantify the difference between the modes of the rational
+programmers and thus the different blame assignment strategies  for the specific
+debugging scenarios we analyze in our experiment.

### 2020-07-08 writing/sample.tex
revising rational section

@@
 \section{How to Sample the Experimental Space} \label{sec:sample}
+
+As we mention above we treat each definitions as a component. A
+consequence of this decision is that examining the whole lattice
+for each of the mutants becomes unfeasible.
+As figure~\ref{table:benchmarks} shows,  most of the programs contain dozens of definitions  
+resulting in lattices with as many as $1.41 \times 10^{28}$ configurations.
+Since running a single configuration can take 10 minutes or more 
+(due to contract checking and instrumentation), running every possible configuration is impractical.
+Evidence from similar contexts suggests that randomly sampling the lattice instead of
+an exhaustive exploration is an effective alternative~\cite{greenman2019evaluate}.
+Hence we randomly sample enough significant configurations
+from the lattice to obtain a confidence of 0.95 (with margin of error
+0.05) about whether our hypotheses hold or not.
+See accompanying appendix for the details of the calculations for the
+sufficient size of samples for this estimate.
+Exploring more of the lattice would yield higher confidence in the generalizability
+of our results for each benchmark in cases where no violations of a property
+(e.g. Blame Trail) are found.
+In such cases, we could be more certain that we did not simply miss
+the configuration(s) that reveal the violation by exploring the lattice more fully.
+Our choice of random sample counts reflects the (informal) standard practice of estimating
+results to 95 percent or higher confidence.

### 2020-07-08 writing/conclusion.tex
threat from rationality; point to bad type annotations as future work

%% -----------------------------------------------------------------------------
\section{What to Do Next} \label{sec:conclusion}

The paper provides the first insight into the measurable value of blame
assignment in the gradually typed world. In an implicit manner researchers and
language designers have answered this question one way or another, without any
evidence. Now they may use the evaluation method presented in this paper to
confirm or reject their answers. We already know from~\citet{tgpk-dls-2018}
that, by intuition, programmers prefer the run-time checking of natural over
other soundness methods. But, just because a random set of programmers expresses
this view does not mean that blame assignment adds value. At this point we know
that it does, in some circumstances, and that in others, the answer remains
murky.

The paper does {\em not\/} address a problem in the gradually typed world that
was pointed out early on by practical researchers~\cite{incorrect-ts,
sta-nt-base-types, wmwz-ecoop-2017} and that has recently received theoretical
attention~\cite{gfd-oopsla-2019}: mistakes in type annotations themselves.
Developers use gradual typing to move an untyped code base into the typed realm,
and to this end, they need typed APIs for the vast repositories of
already-existing libraries. Because they can, language implementors merely
create typed facade modules that import untyped functions and exports them with
type annotations---instead of converting the libraries themselves. With those
facades, the compiler can easily type-check typed components. The problem is
that these retroactive additions of a specific type to a component and the
standard run-time functions may result from a misunderstanding of the
code. \emph{They may thus be a mistake themselves.}

The cited evidence suggests that this scenario is quite common, and that a type
soundness check does {\em not\/} at all address the problem.  Hence, a future
evaluation must develop type-level mutators that produce incorrect type
annotations without breaking the code itself. Some preliminary work suggests
that such type mutators are much more difficult to develop than the code
mutators presented in this paper. We conjecture that the natural semantics will
usually discover such mistakes, while the transient and erasure semantics cannot
help with such mistakes at all; indeed, we expect that these latter two raise
misleading exceptions or produce plainly-wrong answers.  Of course, only further
evaluation work can confirm or reject these conjectures.

### 2020-07-08 writing/discussion.tex
titles are questions
threat from rationality; point to bad type annotations as future work

@@
           \item More?
+
+	  \item is the rational programmer rational?
+
         \end{itemize}
 \end{itemize}
+
+Like {\em homo economicus\/}, which decouples the actual behavior of a
+participant in an economy for the sake of mathematical modeling, the
+model of a rational programmer decouples the actual debugging behavior
+of a software developer for the sake of a systematic, large-scale
+analysis. This decoupling comes advantages and disadvantages. In the
+economic realm, mathematical models have provided some predictive
+insights into the market's behavior; but as behavioral economics has
+shown more recently, the mathematical abstraction of a rational actor
+makes predictions also quite unreliable in some situations.
+\footnote{It has also misled economists to focus on just the
+mathematics, though this problem is not relevant here.}  Just like an
+ordinary consumer or producer, an actual software developer is
+unlikely to stick to the exact strategy proposed here. When this
+happens, the predicted benefits of blame assignment may not
+materialize. Indeed, our own personal experience suggests such
+deviations, and it also suggests that deviating is a mistake. To make
+a true judgment of the usefulness of the rational-programmer idea, the
+community will need a lot more experience with this form of evaluation
+and relating the evaluation to the behavior of working programmers.

### 2020-07-08 writing/introduction.tex
threat from rationality; point to bad type annotations as future work

@@
 
-Next, the paper describes the challenges of developing an evaluation method
-(section~\ref{sec:challenges}), which may explain why the question has been neglected
-for such a long time.
+Next, the paper describes the challenges of developing an evaluation
+method, which may explain why the question has been neglected for such
+a long time.

### 2020-07-08 writing/rational.tex
more tex-os

@@
   $\conf_{i+1} \setminus \conf_i = \first{\mblame{\system, \conf_i}}$
-  where $\first{\mblame{\system, \conf}}$ denotes the first component 
+  where $\first{\mblame{\system, \conf}}$ is the first component 
   the transient semantics add to the blame set for $\conf$.

### 2020-07-08 writing/sample.tex
more tex-os

@@
 for each of the mutants becomes unfeasible.
-As figure~\ref{table:benchmarks} shows,  most of the programs contain dozens of definitions  
+As figure~\ref{table:benchmark-descriptions} shows,  most of the programs contain dozens of definitions  
 resulting in lattices with as many as $1.41 \times 10^{28}$ configurations.
@@
 Evidence from similar contexts suggests that randomly sampling the lattice instead of
-an exhaustive exploration is an effective alternative~\cite{greenman2019evaluate}.
+an exhaustive exploration is an effective alternative~\cite{gtnffvf-jfp-2019}.
 Hence we randomly sample enough significant configurations

### 2020-07-08 writing/rational.tex
revision of section 5

@@
 evaluate the effectiveness of a
-rational programmer. The components of each program create a \emph{configuration
+rational programmer. The components of each program create a
+\emph{debugging scenario
 space}, like the one introduced by~\citet{tfgnvf-popl-2016}. This space consists
-of variants of the program, dubbed \emph{configurations}, that differ in their
+of variants of the program, dubbed \emph{scenarios}, that differ in their
 type specifications. In the setting of migratory typing,\footnote{The method can
 be adapted to so-called true gradual typing systems~\cite{svcb-snapl-2015} in a
-straightforward manner.} each configuration differs from another in the subset
-of typed components. Formally, a program $\system$ is a set of components
-$\set{\component}$, and a configuration $\conf$ is a subset of
-$\set{\component}$ . These configurations of $\system$ are ordered by the subset
+straightforward manner.} each scenario differs from another in which
+components are typed. Formally, a program $\system$ is a set of components
+$\set{\component}$, and a scenario $\conf$ is a subset of
+$\set{\component}$ . These scenarios of $\system$ are ordered by the subset
 relation and thus form a lattice $\lattice{\system}$ with $2^{\size{\system}}$
-elements.  The bottom configuration of $\lattice{\system}$ is always $\emptyset$
-and the top element is $\system$ itself. In between are all the mix-typed
-configurations.
+elements.  The bottom scenario of $\lattice{\system}$ is always $\emptyset$
+and the top one is $\system$ itself. In between are all the mix-typed
+scenarios.
 
 With these definitions, the actions of the rational programmer map to an
-ascending chain---dubbed a \emph{trail}---in $\lattice{\system}$.  The starting
-configuration $\conf_0$ of a trail---also referred to the \emph{root} (of the
-trail)---is the configuration that the programmer attempts to debug. Any other
+ascending chain---dubbed a \emph{trail}---in $\lattice{\system}$. 
+Intuitively, with each action the rational programmer  reduces the current
+scenario to an ``easier'' one to debug. The easiest scenarios are those
+that the type checker rejects outright. Assuming the types of the
+program are correct, if the typechecker signals a type error, the error
+pinpoints an impedance mismatch between the broken code of a component 
+and its type. Hence the programmer knows which component to fix. 
+The starting
+scenario $\conf_0$ of a trail---also referred to as the \emph{root} (of the
+trail)---is the scenario that the programmer attempts to debug initially. Any other
 configuration $\conf_{i+1}$ in the trail differs from its predecessor
-$\conf_{i}$ in that the untyped component that $\conf_i$ blames under the
-various semantics, dubbed $\blame{\system, \conf_i}$, becomes typed in
-$\conf_{i+1}$.\footnote{Blame can result either directly from a run time type
+$\conf_{i}$ in that more 
+components compared to  $\conf_{i}$ become typed. The rational programmer 
+decides which are these components based on the feedback it receives from
+the gradual typing system when running $\conf_{i}$.
+
+\footnote{Blame can result either directly from a run time type
 check or, in the absence of blame, from an exception that we track back to the
@@
 attempt to debug a mix-typed programmer; they add type specifications to
-the component blame points. In terms of configurations and trails, this
-translates to extending a trail with a configuration that adds to the
-last configuration of the trail so far the component the latter configuration
+the component blame points. In terms of scenarios and trails, this
+translates to extending a trail with a new scenario that adds to the
+latest scenario from the trail the component that latest scenario
 blames.  We call trails constructed in this manner \emph{natural blame
-trails} and they capture the essence of the natural rational programmer:
+trails} and they capture the essence of the natural rational programmer
+:\footnote{We use trail and rational programmer interchangeably from now
+on.}
 
@@
 A natural blame trail
-is a sequence of configurations $\conf_0,...\conf_n$ of a program
+is a sequence of scenarios $\conf_0,...\conf_n$ of a program
 $\system$  such that for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
@@
 Natural rational programmers migrate faulty programs along natural blame
-trails in order to locate bugs. Specifically, assuming that the types 
-they add during migration are correct, when a buggy component migrates
-from untyped to typed, the typechecker detects the impedance mismatch
-between the broken code of the component and its types and signals a type
-error. Hence, trails that end at an ill-typed configuration successfully 
+trails in order to locate bugs. As we mention above, assuming that the types 
+they add during migration are correct, when the buggy component of a
+program migrates from untyped to typed, the typechecker detects the issue
+at compile time. Hence, trails that end at an ill-typed scenarios successfully 
 pinpoint the location of bugs:
@@
 A natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
-\emph{successful}  iff its last configuration $\conf_n$ does not type check.
+\emph{successful}  iff its last scenario $\conf_n$ does not type check.
   \end{quote}
@@
 \noindent 
-That is failing natural blame trails are those that have not reached a configuration that reveals
-the bug at compile time and their last configuration does not produce
-blame. Thus the rational programmer receives no hint from the gradual type
+That is failing natural blame trails are those that have not reached a scenario that reveals
+the bug at compile time and their last scenario does not produce
+blame. Thus the rational programmer receives no hints from the gradual type
 system about how to further migrate the faulty program in search for the
@@
 
-These definitions help a rational programmer gauge whether it pays off to heed blame
+These definitions allow a rational programmer to gauge whether it pays off to heed blame
 when debugging \system. 
@@
 A natural random trail
-is a sequence of configurations $\conf_0,...\conf_n$ of a program
+is a sequence of scenarios $\conf_0,...\conf_n$ of a program
 $\system$  such that for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
@@
 \indent
-With blame and random natural trails, we can ask whether 
+Similar to natural blame we determine how successful the random
+migration is based on the success of natural random trails: 
+\begin{quote}
+\it
+  Natural random is always useful for a program $\system$ iff 
+  $\lattice{\system}$ contains no failing natural random trails.
+
+   Natural random is sometimes useful for a program $\system$ iff 
+  $\lattice{\system}$ contains successful natural random trails.
+
+   Natural random is never useful for a program $\system$ iff 
+  $\lattice{\system}$ contains no successful natural random trails.
+
+\end{quote}
+
+Putting all these pieces together, we can ask whether 
 blame adds value over random migration: 
@@
 \it
-  Blame adds value to natural for a program $\system$ over pure chance iff 
-  no natural blame trails in  $\lattice{\system}$ are failing while 
- there exists at least one random natural trail that is
-  failing.
+  Natural blame is more successful than natural random for $\system$ if 
+  either (i) natural blame is always successful for $\system$ while natural random 
+  is sometimes or never successful for $\system$, or (ii) 
+  natural blame is sometimes successful for $\system$ while natural random 
+  is never successful for $\system$.
+
+  Natural blame adds value to natural for a program $\system$ over pure chance iff 
+  natural blame is more successful than natural random.
 \end{quote}
@@
 A natural exception trail
-is a sequence of configurations $\conf_0,...\conf_n$ of a program
+is a sequence of scenarios $\conf_0,...\conf_n$ of a program
 $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
@@
 \noindent
-We use the new mode to describe what it means for blame to add value over both exceptions and pure
-chance:
+We use the new mode to ask whether blame adds value over both exceptions and pure
+chance in natural:
 \begin{quote}
 \it
-  Blame adds value to natural for a program $\system$ over pure chance and
-  exceptions iff no natural blame trails in  $\lattice{\system}$ are failing while 
- there exist at least one random natural trail and one natural exception trail
- that are
-  failing.
+  Natural blame adds value to natural for a program $\system$ over pure chance and
+  natural exceptions iff natural blame is more successful than both natural
+  random and natural exception.
 \end{quote}
@@
 
- Since gradually typed languages with erasure semantics do not come with blame
-assignment, a rational programmer can only hope that the underlying safety
-checks and their exceptions are helpful.  Thus, adapting the corresponding
-definitions from natural to erasure, we ask whether the effect of erasure
-exception in debugging a program $\system$ goes beyond pure chance:
+ Since gradually typed languages with erasure semantics do not come with
+ blame assignment, a rational programmer can only hope that the underlying
+ safety checks and their exceptions are helpful.  Thus, adapting the
+ definitions related to exception and random trails from natural to
+ erasure, we ask whether the effect of erasure exceptions in debugging a
+ program $\system$ goes beyond that of pure chance:
 
@@
 \it
-  Exceptions adds value to erasure for a program $\system$ over pure chance
-  iff no erasure exception trails in  $\lattice{\system}$ are failing while 
- there exists at least one random erasure trail that is failing.
+  Erasure exceptions add value to erasure for a program $\system$ over pure chance
+  iff erasure exception is more successful than erasure random.
 \end{quote}
@@
 when the language produces a blame set. Our answer is that the rational
-programmer has least two reasonable options.  The first option is the rational programmer types
+programmer has at least two reasonable options.  The first option is the rational programmer types
 all blamed components at once.  In the second option, the rational
@@
 A  transient all blame trail
-is a sequence of configurations $\conf_0,...\conf_n$ of a program
+is a sequence of scenarios $\conf_0,...\conf_n$ of a program
 $\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
@@
 A transient firstborn blame trail
-is a sequence of configurations $\conf_0,...\conf_n$ of a program
+is a sequence of scenarios $\conf_0,...\conf_n$ of a program
 $\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
@@
 \end{quote}
-\noindent As for natural we use the transient all blame trails and 
-transient firstborn  trails to define when  transient all blame and 
-transient firstborn add value to transient over exceptions and pure
-chance.
+\noindent As for natural we use the transient all blame trail transient
+firstborn, transient erasure and transient random trails to define when
+transient all blame and transient firstborn add value to transient over
+exceptions and pure chance: 
+\begin{quote}
+\it 
+Transient all blame adds value to transient for a program $\system$ over
+  pure chance and transient exceptions iff transient all blame 
+  is more successful than both transient random and transient exception.
+
+Transient firstborn blame adds value to transient for a program $\system$ over
+  pure chance and transient exceptions iff transient all blame 
+  is more successful than both transient random and transient exception.
+
+\end{quote}
 
@@
 Trails and their properties give us the tools for a rigorous examination
-of blame for natural, transient and erasure Typed Racket and the 
-mutants of section~\ref{sec:mutate}. In particular, our experiment examines 
-four questions for each mutant:
+of blame for natural, transient and erasure Typed Racket in the setting of the 
+mutants from section~\ref{sec:mutate}. In particular, our experimental
+process collects data to answer four questions for each mutant:
 \begin{itemize}
-\item[$Q_1$] Is blame is valuable for natural?
+\item[$Q_1$] Is blame valuable for natural?
 
@@
 Table~\ref{fig:experiment-outline} summarizes how each question relates to
-different trails/modes of the rational programmer. For example experimental
+different kinds of trails/modes of the rational programmer. For example experimental
 question $Q_1$ asks whether blame is valuable for natural and our experiment
@@
 
-In detail, to answer $Q_1$ for a mutant, we first construct the mutant's configuration 
-lattice. Then, we determine the configurations of the lattice that detect the fault in the
+In detail, to answer $Q_1$ for a mutant, we have to 
+calculate and log for each natural blame, natural exception and natural
+random trail 
+in the scenario lattice of the mutant whether it is successful or failing. 
+Hence, we first construct the mutant's scenario
+lattice and we determine the scenario of the lattice that can detect the fault in the
 program. This can happen either because Typed Racket's type checker
-outright rejects the configuration or because natural Typed Racket's runtime 
+outright rejects the scenario or because natural Typed Racket's runtime 
 performs a type check that fails. In either
-case,  each of these configurations is the root of a trail that we 
-extend according the natural blame . 
-If no configurations can be added to the trail\footnote{All trails
+case,  each of these scenarios is the root of a trail that we attempt
+extend according to the natural blame rational programmer. 
+If no scenarios can be added to the trail\footnote{All trails
 reach that point since the lattices are finite.}, we check if the
-last configuration of the trail type checks or not. If it does the trail
+last scenario of the trail type checks or not. If it does this
+natural blame trail
 is successful, otherwise it is failing. We repeat the same process for the
 roots two more times; once following the natural exception rational programmer and 
-once the natural random rational programmer. Finally, we use the
-success/failure results for the trails to check if they meet the condition
-for natural blame to add value to natural over exceptions and pure chance.
-The process is analogues for the other questions and their corresponding
+once the natural random rational programmer. Finally, based on the
+collected data, we
+calculate the success/failure results for the trails to determine whether
+ natural blame is more successful than  natural exceptions and pure chance
+ and we obtain the answer for $Q_1$.
+The process is analogues for the other questions employing their corresponding
 modes of the rational programmer.
@@
 
-In addition to the value of blame, trails permit us to examine the effort a
-rational programmer spends to locate a bug. As a metric for the effort, 
-we use the number of components the rational programmer has to migrate 
-from untyped to typed to find the bug. We can easily and uniformly calculate the value
-of the metric for any successful trail in any mode of the rational
-programmer (including transient blame all)  as the  size of the difference 
-between the trail's root and its final configuration.
+In addition to successful and failure information for the various trails,  
+our experiment  records the number of components a rational programmer 
+has to type along each successful trail ($\lvert \conf_n \setminus \conf_0
+\rvert$). We use this number as the metric for the effort 
+of the rational programmer to debug the root scenario of each successful trail.
 
 Programmer effort can help shed some light to the relative effectiveness of the
-three gradual typing systems. For example, consider a configuration of the
-lattice of one of our mutants. The configuration corresponds to a
- scenario that the rational programmer attempts to debug. If the
-configuration results in blame both under, for instance, the natural and transient semantics,
+three gradual typing systems. For example, consider a debugging scenario from  the
+lattice of one of our mutants.  If the
+scenario results in blame both under, for instance, the natural and transient semantics,
 then the natural blame rational programmer and the two transient blame
 rational programmers (all and firstborn) can compete to see which one
-reaches a configuration that doesn't type check with less effort. In
+reaches an ill-typed scenarion with less effort. In
 general, the effort of trails of different modes that have the same
-root is a metric that allows us to compare the effectiveness of these
-modes even if they correspond to different gradual typing systems.
+root is a point of comparison for the effectiveness of these
+modes even if the modes correspond to different gradual typing systems.
 Unfortunately, statistically speaking, such comparisons between two modes do not generalize 
-beyond the successful trails that share their root in our experiments.
+beyond the successful trails with common roots in our experiments.
 Still they allow us to quantify the difference between the modes of the rational

### 2020-07-08 writing/rational.tex
rational programmer touch ups

@@
 Based on this idea, a metric for migratory paths is a concise way to
-evaluate the effectiveness of a
-rational programmer. The components of each program create a
-\emph{debugging scenario
-space}, like the one introduced by~\citet{tfgnvf-popl-2016}. This space consists
-of variants of the program, dubbed \emph{scenarios}, that differ in their
-type specifications. In the setting of migratory typing,\footnote{The method can
-be adapted to so-called true gradual typing systems~\cite{svcb-snapl-2015} in a
+evaluate the effectiveness of a rational programmer. The components of
+each program create a \emph{debugging scenario space}, a lattice like
+the one introduced by~\citet{tfgnvf-popl-2016} for performance
+evaluations. This space consists of variants of the program, dubbed
+\emph{scenarios}, that differ in their type specifications. In the
+setting of migratory typing,\footnote{The method can be adapted to
+so-called true gradual typing systems~\cite{svcb-snapl-2015} in a
 straightforward manner.} each scenario differs from another in which
-components are typed. Formally, a program $\system$ is a set of components
-$\set{\component}$, and a scenario $\conf$ is a subset of
-$\set{\component}$ . These scenarios of $\system$ are ordered by the subset
-relation and thus form a lattice $\lattice{\system}$ with $2^{\size{\system}}$
-elements.  The bottom scenario of $\lattice{\system}$ is always $\emptyset$
-and the top one is $\system$ itself. In between are all the mix-typed
-scenarios.
+components are typed. Formally, a program $\system$ is a set of
+components $\set{\component}$, and a scenario $\conf$ is a subset of
+$\set{\component}$ . These scenarios of $\system$ are ordered by the
+subset relation and thus form a lattice $\lattice{\system}$ with
+$2^{\size{\system}}$ elements.  The bottom scenario of
+$\lattice{\system}$ is always $\emptyset$ and the top one is $\system$
+itself. In between are all the mix-typed scenarios.
 
-With these definitions, the actions of the rational programmer map to an
-ascending chain---dubbed a \emph{trail}---in $\lattice{\system}$. 
-Intuitively, with each action the rational programmer  reduces the current
-scenario to an ``easier'' one to debug. The easiest scenarios are those
-that the type checker rejects outright. Assuming the types of the
-program are correct, if the typechecker signals a type error, the error
-pinpoints an impedance mismatch between the broken code of a component 
-and its type. Hence the programmer knows which component to fix. 
-The starting
-scenario $\conf_0$ of a trail---also referred to as the \emph{root} (of the
-trail)---is the scenario that the programmer attempts to debug initially. Any other
-configuration $\conf_{i+1}$ in the trail differs from its predecessor
-$\conf_{i}$ in that more 
-components compared to  $\conf_{i}$ become typed. The rational programmer 
-decides which are these components based on the feedback it receives from
-the gradual typing system when running $\conf_{i}$.
+With these definitions, the actions of the rational programmer map to
+an ascending chain---dubbed a \emph{trail}---in $\lattice{\system}$.
+The starting scenario $\conf_0$ of a trail---also referred to as the
+\emph{root} (of the trail)---is the scenario that the programmer
+attempts to debug initially. The easiest scenarios are those that the
+type checker rejects outright, because a type error pinpoints an
+impedance mismatch between the broken code of one component and its
+typed interface to others. At this point, the programmer has
+identified the broken component. If the scenario type-checks, the
+programmers runs the program until it raises an exception.  The
+rational programmer uses the blame assignment\footnote{Blame can
+result either directly from a run time type check or, in the absence
+of blame, from an exception that we track back to the component that
+causes it with the help of the stack trace.} to decide which
+component(s) to equip with types. This choice constructs scenario
+$\conf_{i+1}$ from $\conf_{i}$ and implies the creation of an
+ascending chain.  Intuitively, with each step along the chain, the
+rational programmer reduces the current scenario to an ``easier'' one
+to debug.
 
-\footnote{Blame can result either directly from a run time type
-check or, in the absence of blame, from an exception that we track back to the
-component that causes it with the help of the stack trace.}
+% The successor of a scenario on a trail has more typed component;
+% formally, configuration $\conf_{i+1}$ on the trail differs from its
+% predecessor $\conf_{i}$ in that more components are typed.
 
-The next three subsections specialize the idea to the various semantics of
-gradual typing. Once this machinery is in place, we use the fourth subsection to
-state the precise experimental research questions that correspond to the
-high-level question on the first page and to explain our experimental
-process. The final subsection extends the development with a discussion of
-how we can measure the rational programmer's effort.
+The next three subsections specialize the trail-scenario idea to the
+various semantics of gradual typing. Once this machinery is in place,
+we use the fourth subsection to state the precise experimental
+research questions that correspond to the high-level question on the
+first page and to explain our experimental process. The final
+subsection extends the development with a discussion of how we can
+measure the rational programmer's effort along the trail.
 
@@
 
-In natural Typed Racket, rational programmers rely on blame when they
-attempt to debug a mix-typed programmer; they add type specifications to
-the component blame points. In terms of scenarios and trails, this
-translates to extending a trail with a new scenario that adds to the
-latest scenario from the trail the component that latest scenario
-blames.  We call trails constructed in this manner \emph{natural blame
-trails} and they capture the essence of the natural rational programmer
-:\footnote{We use trail and rational programmer interchangeably from now
-on.}
+The natural semantics assigns blame to exactly one boundary.  A blame assignment
+has the following specific meaning: the typed component may make incorrect type
+assumptions about the untyped component in its interface, or the correct
+interface exposes a bug in the untyped component. Our setup rules out the first
+alternative (but see section~\ref{sec:conclusion}), and therefore the rational
+programmer extends the trail to a scenario that swaps out the untyped component
+for its typed counterpart.
 
+We can turn this description into a formal definition of the trail. 
 \begin{quote}
-\it 
-A natural blame trail
-is a sequence of scenarios $\conf_0,...\conf_n$ of a program
-$\system$  such that for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
-  $\conf_{i+1} \setminus \conf_i =\{\blame{\system, \conf_i}\}$ where
+\it A \emph{natural blame trail} is a sequence of scenarios $\conf_0,...\conf_n$ of
+a program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
+\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =\{\blame{\system, \conf_i}\}$ where
 $\blame{\system, \conf}$ denotes the component that $\conf$ blames.
@@
 
-Natural rational programmers migrate faulty programs along natural blame
-trails in order to locate bugs. As we mention above, assuming that the types 
-they add during migration are correct, when the buggy component of a
-program migrates from untyped to typed, the typechecker detects the issue
-at compile time. Hence, trails that end at an ill-typed scenarios successfully 
-pinpoint the location of bugs:
-In formal terms:
+When the buggy untyped component of a program is replaced by the typed counterpart,
+the type checker fails if this component causes the impedance mismatch. Hence, a
+trail that end at an ill-typed scenarios successfully pinpoints the location of
+bugs. 
+\begin{quote}
+\it A natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
+\emph{successful} iff its last scenario $\conf_n$ does not type check.  A natural
+blame trail $\conf_0,..,\conf_n$ in a lattice $\lattice{\system}$ is \emph{failing}
+iff $\conf_n$ type checks and the trail cannot be extended further.
+\end{quote}
+That is failing natural blame trails are those that have not reached a scenario that
+reveals the bug at compile time and their last scenario does not produce blame. Thus
+the rational programmer receives no hints from the gradual type system on how to
+continue the search for the bug.
+
+Equipped with these preliminary definitions, we can precisely define what it means
+for blame to add value to a gradual type system in a specific setting. 
 \begin{quote}
 \it
-A natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
-\emph{successful}  iff its last scenario $\conf_n$ does not type check.
-  \end{quote}
-  In juxtaposition:
-\begin{quote}
-\it 
-A natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
-\emph{failing}  iff $\conf_n$ type checks and the trail cannot be extended
-  further. 
- \end{quote}
-\noindent 
-That is failing natural blame trails are those that have not reached a scenario that reveals
-the bug at compile time and their last scenario does not produce
-blame. Thus the rational programmer receives no hints from the gradual type
-system about how to further migrate the faulty program in search for the
-bug.
-
- We can now give a precise definition of what it
-means for blame to help the rational programmer:
-\begin{quote}
-\it
-  Natural blame is always useful for a program $\system$ iff 
-  $\lattice{\system}$ contains no failing natural blame trails.
-
-   Natural blame is sometimes useful for a program $\system$ iff 
-  $\lattice{\system}$ contains successful natural blame trails.
-
-   Natural blame is never useful for a program $\system$ iff 
-  $\lattice{\system}$ contains no successful natural blame trails.
-
+Given a program $\system$, natural blame is\\
+(1) \emph{always useful} iff $\lattice{\system}$ contains no failing natural blame trails.\\
+(2) \emph{sometimes useful} iff $\lattice{\system}$ contains successful natural blame trails.\\
+(3) \emph{never useful} iff $\lattice{\system}$ contains no successful natural blame
+trails.\\
 \end{quote}
 
-These definitions allow a rational programmer to gauge whether it pays off to heed blame
-when debugging \system. 
-However, they do not answer whether blame is a critical piece of 
-the rational programmer's process.  For instance, 
-typing a random component   could be equally useful as typing the blamed one. 
-To account for this situation, we devise a new
-mode of the natural rational programmer that 
- follows a random migration process:
+While these definitions describe how a rational programmer might gauge whether it
+pays off to heed blame assignments during debugging, they do not answer whether
+blame is a critical piece of the rational programmer's process.  For instance,
+typing a random component might be as useful as typing the blamed one.
+
+To account for this situation, we devise a new mode of the natural rational
+programmer that follows a random migration process:
 \begin{quote}
-\it 
-A natural random trail
-is a sequence of scenarios $\conf_0,...\conf_n$ of a program
-$\system$  such that for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
-$\conf_{i+1} \setminus \conf_i = \{\random{\system, \conf_i}\}$
+\it A natural random trail is a sequence of scenarios $\conf_0,...\conf_n$ of a
+program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
+\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i = \{\random{\system, \conf_i}\}$
 where $\random{\system, \conf}$ denotes a randomly selected component of $\system$
@@
 \end{quote}
-\indent
 Similar to natural blame we determine how successful the random
@@
 \it
-  Natural random is always useful for a program $\system$ iff 
-  $\lattice{\system}$ contains no failing natural random trails.
-
-   Natural random is sometimes useful for a program $\system$ iff 
-  $\lattice{\system}$ contains successful natural random trails.
-
-   Natural random is never useful for a program $\system$ iff 
-  $\lattice{\system}$ contains no successful natural random trails.
-
+Given a program $\system$, natural random blame is\\
+(1) \emph{always useful} iff $\lattice{\system}$ contains no failing natural random blame trails.\\
+(2) \emph{sometimes useful} iff $\lattice{\system}$ contains successful natural random blame trails.\\
+(3) \emph{never useful} iff $\lattice{\system}$ contains no successful natural
+random blame trails.\\
 \end{quote}
 
-Putting all these pieces together, we can ask whether 
-blame adds value over random migration: 
+Putting the pieces together, we can now define that blame assignments adds value in
+the context of a natural semantics if blame-based migration is better than random
+migration.
 \begin{quote}
-\it
-  Natural blame is more successful than natural random for $\system$ if 
-  either (i) natural blame is always successful for $\system$ while natural random 
-  is sometimes or never successful for $\system$, or (ii) 
-  natural blame is sometimes successful for $\system$ while natural random 
-  is never successful for $\system$.
-
-  Natural blame adds value to natural for a program $\system$ over pure chance iff 
-  natural blame is more successful than natural random.
+\it Natural blame adds value to natural for a program $\system$ over pure chance iff
+  natural blame is more successful than natural random. And, natural blame is more
+  successful than natural random for $\system$ if either natural blame is always
+  successful for $\system$ when natural random is sometimes or never successful, or
+  natural blame is sometimes successful for $\system$ when natural random is never
+  successful.
 \end{quote}
 
-Even if the success of the natural rational programmer is not by pure
-chance, it could well be because the natural semantics perform runtime type checks at
-the right time. In other words, one rational programmer that 
-heeds blame could be equally successful as one that only uses information from the exceptions of failed 
-runtime type checks. Similar to before, in response we
-define the exception mode of the rational programmer:
-\begin{quote}
-\it 
-A natural exception trail
-is a sequence of scenarios $\conf_0,...\conf_n$ of a program
-$\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
-$\conf_{i+1} \setminus \conf_i = \{\error{\system, \conf_i}\}$
-  where $\error{\system, \conf}$ denotes the component that triggers a
-an exception of the underlying language.
-\end{quote}
-\noindent
-We use the new mode to ask whether blame adds value over both exceptions and pure
-chance in natural:
-\begin{quote}
-\it
-  Natural blame adds value to natural for a program $\system$ over pure chance and
-  natural exceptions iff natural blame is more successful than both natural
-  random and natural exception.
-\end{quote}
-\noindent
-If blame does add value to natural for a program $\system$ 
-over pure chance and exceptions then we can finally
-deduce that, for debugging $\system$, blame is a valuable feature of the
-natural semantics.
+Even if the success of the natural rational programmer is not by pure chance, it
+could well be because the natural semantics perform runtime type checks at the same
 
@@
  definitions related to exception and random trails from natural to
- erasure, we ask whether the effect of erasure exceptions in debugging a
- program $\system$ goes beyond that of pure chance:
+ erasure, we characterize when the effect of erasure exceptions in debugging a
+ program $\system$ goes beyond that of pure chance.
 
@@
 
-In contrast to erasure, transient does come with blame. As we discuss in
-section~\ref{sec:landscape} though, distinctly from natural blame,
-transient may blame more than one components. This peculiarity of 
-transient raises the question of how the rational programmer should react
-when the language produces a blame set. Our answer is that the rational
-programmer has at least two reasonable options.  The first option is the rational programmer types
-all blamed components at once.  In the second option, the rational
-programmer selects the component that is added to the blame set first and types
-only that --- after all the types of this first component should be able
-to  detect an impedance mismatch earlier in the evaluation of a program. 
-As before, the two modes give rise to two different notions of trail:
+The transient semantics assigns blame to a series of components. In this setting, a
+blame assignment has the following meaning: the witness value to the impedance
+mismatch crossed the boundaries between neighbors in the sequence, and each crossing
+checked a portion of the value's type shape.
+
+This ambiguity in transient blame raises the question of how the rational programmer
+should react when the language produces a blame set. Our answer is that the rational
+programmer has at least two reasonable options.  The first option is to type all
+blamed components at once. The second one is to select the component that is added
+to the blame set first and assign types to only that one---after all the types of
+this first component should be able to detect an impedance mismatch earlier in the
+evaluation of a program than the later ones.
+
+These two modes of rationalizing give rise to two different notions of trail.
 \begin{quote}
-\it 
-A  transient all blame trail
-is a sequence of scenarios $\conf_0,...\conf_n$ of a program
-$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
-~$\conf_{i+1} \setminus \conf_i = \mblame{\system, \conf_i}$
-where $\mblame{\system, \conf}$ denotes the set of components that $\conf$
-blames under transient semantics; 
+\it A \emph{ transient-all blame trail} is a sequence of scenarios
+$\conf_0,...\conf_n$ of a program $\system$ where for all $0 \leq i \leq n - 1$,
+$\conf_i \subset \conf_{i+1}$ and ~$\conf_{i+1} \setminus \conf_i = \mblame{\system,
+\conf_i}$ where $\mblame{\system, \conf}$ denotes the set of components that $\conf$
+blames under transient semantics.
 \end{quote}
@@
 \begin{quote}
-\it 
-A transient firstborn blame trail
-is a sequence of scenarios $\conf_0,...\conf_n$ of a program
-$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
-  $\conf_{i+1} \setminus \conf_i = \first{\mblame{\system, \conf_i}}$
-  where $\first{\mblame{\system, \conf}}$ is the first component 
-  the transient semantics add to the blame set for $\conf$.
+\it A \emph{transient-firstborn blame trail} is a sequence of scenarios
+$\conf_0,...\conf_n$ of a program $\system$ where for all $0 \leq i \leq n - 1$,
+$\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
+\first{\mblame{\system, \conf_i}}$ where $\first{\mblame{\system, \conf}}$ is the
+first component transient adds to the blame set for $\conf$.
 \end{quote}
-\noindent As for natural we use the transient all blame trail transient
-firstborn, transient erasure and transient random trails to define when
-transient all blame and transient firstborn add value to transient over
-exceptions and pure chance: 
+The definitions of \emph{transient erasure trails} and \emph{transient random
+trails} are analogous to those for natural erasure trails and natural random trails,
+respectively.
+
+We also use these four forms of blame trails to define when transient-all and
+transient-firstborn add value to transient over exceptions and pure chance.
 \begin{quote}
@@
 
-\begin{figure}
+\begin{figure}[ht]
 \center
-{\begin{tabular}[ht]{l|c|c|c}
+{\begin{tabular}{l|c|c|c}
 %% ---------------------------------------------------------------------------------------------------------------
@@
 
-In detail, to answer $Q_1$ for a mutant, we have to 
-calculate and log for each natural blame, natural exception and natural
-random trail 
-in the scenario lattice of the mutant whether it is successful or failing. 
-Hence, we first construct the mutant's scenario
-lattice and we determine the scenario of the lattice that can detect the fault in the
-program. This can happen either because Typed Racket's type checker
-outright rejects the scenario or because natural Typed Racket's runtime 
-performs a type check that fails. In either
-case,  each of these scenarios is the root of a trail that we attempt
-extend according to the natural blame rational programmer. 
-If no scenarios can be added to the trail\footnote{All trails
-reach that point since the lattices are finite.}, we check if the
-last scenario of the trail type checks or not. If it does this
-natural blame trail
-is successful, otherwise it is failing. We repeat the same process for the
-roots two more times; once following the natural exception rational programmer and 
-once the natural random rational programmer. Finally, based on the
-collected data, we
-calculate the success/failure results for the trails to determine whether
- natural blame is more successful than  natural exceptions and pure chance
- and we obtain the answer for $Q_1$.
-The process is analogues for the other questions employing their corresponding
+In detail we answer $Q_1$ for a mutant by creating trails for each natural blame,
+natural exception and natural random trail in the scenario lattice of the mutant. A
+log reports on success and failure endings.  The first step is to construct the
+mutant's scenario lattice and determine the scenario that can detect the fault in
+the program. This can happen either because Typed Racket's type checker outright
+rejects the scenario or because natural Typed Racket's runtime performs a type check
+that fails. In either case, each of these scenarios is the root of a trail. Our test
+bed extends these trails according to the natural-blame programmer.  If no scenarios
+can be added to the trail---a trail may the top of the finite lattice---we check if
+the last scenario of the trail type checks or not. If it does, this natural-blame
+trail is successful, otherwise it is failing. We repeat the same process for the
+roots two more times; once following the natural-exception programmer and once the
+natural-random programmer. Finally, based on the collected data, we calculate the
+success/failure results for the trails to determine whether natural blame is more
+successful than natural exceptions and pure chance and we obtain the answer for
+$Q_1$.  The process is analogues for the other questions employing the respective
 modes of the rational programmer.

### 2020-07-08 writing/discussion.tex
many types is a threat to validity

@@
+%% -----------------------------------------------------------------------------
 \section{What We Learned}
@@
 
-	  \item is the rational programmer rational?
-
         \end{itemize}
@@
 Like {\em homo economicus\/}, which decouples the actual behavior of a
-participant in an economy for the sake of mathematical modeling, the
-model of a rational programmer decouples the actual debugging behavior
-of a software developer for the sake of a systematic, large-scale
-analysis. This decoupling comes advantages and disadvantages. In the
-economic realm, mathematical models have provided some predictive
-insights into the market's behavior; but as behavioral economics has
-shown more recently, the mathematical abstraction of a rational actor
-makes predictions also quite unreliable in some situations.
-\footnote{It has also misled economists to focus on just the
-mathematics, though this problem is not relevant here.}  Just like an
-ordinary consumer or producer, an actual software developer is
-unlikely to stick to the exact strategy proposed here. When this
-happens, the predicted benefits of blame assignment may not
-materialize. Indeed, our own personal experience suggests such
-deviations, and it also suggests that deviating is a mistake. To make
-a true judgment of the usefulness of the rational-programmer idea, the
-community will need a lot more experience with this form of evaluation
-and relating the evaluation to the behavior of working programmers.
+participant in an economy for the sake of mathematical modeling, the model of a
+rational programmer decouples the actual debugging behavior of a software
+developer for the sake of a systematic, large-scale analysis. This decoupling
+comes advantages and disadvantages. In the economic realm, mathematical models
+have provided some predictive insights into the market's behavior; but as
+behavioral economics has shown more recently, the mathematical abstraction of a
+rational actor makes predictions also quite unreliable in some situations.
+\footnote{It has also misled economists to focus on just the mathematics, though
+this problem is not relevant here.}  Just like an ordinary consumer or producer,
+an actual software developer is unlikely to stick to the exact strategy proposed
+here. When this happens, the predicted benefits of blame assignment may not
+materialize. Indeed, our own personal experience suggests such deviations, and
+it also suggests that deviating is a mistake. To make a true judgment of the
+usefulness of the rational-programmer idea, the community will need a lot more
+experience with this form of evaluation and relating the evaluation to the
+behavior of working programmers.
+
+Our setup hides the type choice a rational programmer must make. When the
+run-time checks signal an impedance mismatch in the real world, a programmer
+does not have a typed component ready to swap in. Instead, the programmer must
+come up with the next set of types---and this means the programmer must make
+choices. It is normally possible to assign consistent types to variables in a
+component in different ways. The creation and curation of the benchmarks over
+many years has driven home this lesson, but fortunately, it has also shown that
+the types are in most cases reasonably canonical.  We therefore conjecture that
+a rational programmer would in most cases come up with an equivalent type
+assignment for trail extension as our experimental setup describes.

### 2020-07-09 writing/mutate.tex
fix latex in mutate
draft of benchamrks and mutators

--- 2020-07-06 writing/mutate.tex
@@
-\section{How to Make Benchmarks} \label{sec:mutate}
+\section{How to Make Benchmarks for Blame} \label{sec:mutate}
 
-Given the comparable implementations of the three gradual typing systems,
-the next step for applying our method is to obtain a set of fully
-typed benchmarks that each comes with a single impedance mismatch between
-its type specifications and its code. As we briefly discuss in
-section~\ref{sec:challenges}, we derive such ill-typed benchmarks with
-mutation analysis from well-typed programs.
+An important challenge from~\ref{sec:challenges} is the lack of a large and diverse real-world
+ corpus of  mix-typed programs with type errors that we can use to
+ evaluate rational programmers. To overcome this issue, 
+we create a synthetic corpus of such programs. In this section, we argue
+that despite its synthetic nature our corpus of programs is realistic 
+and diverse.
+
+The starting point for our corpus of programs
+is~\citet{gtnffvf-jfp-2019}'s gradual typing benchmark suite. The
+benchmark suite consists of fully typed correct programs that are written by different authors
+and have been used, maintained and evolved by their authors and others over a
+number of years.  They range widely in size, complexity, purpose and
+features of Typed Racket they employ.  Furthermore the benchmarks use advanced aspects 
+of Typed Racket's type system such as occurrence
+typing~\cite{tf-icfp-2010}, types for mutable and immutable data
+structures, polymorphic types, types for first-class classes and objects and types for
+Racket's numeric tower~\cite{stathff-padl-12}. Without loss of the diversity of the benchmarks,
+we select the ten largest in terms of numbers of components (between 6 and 14 components).
+These benchmarks also have the most complex dependency graphs and, hence, 
+can make the debugging process the hardest for the rational programmer. 
+The table in figure~\ref{table:benchmark-descriptions} shows our selection
+together with a short description for each benchmark.
+
+\begin{figure}
+\begin{tabular}{p{2cm} | p{10cm} }
+  {\bf  name} & {\bf description (author)}  \\
+
+\hline
+
+  \textttt{acquire} & Board game simulation (M. Felleisen)  \\%[1em]
+
+\hline
+  \texttt{gregor} & Utilities for calendar dates (J. Zeppieri) \\%[1em]
+
+\hline
+  \texttt{kcfa} & Functional implementation of 2CFA for the lambda calculus (M. Might) \\%[1em]
+
+\hline
+  \texttt{quadT} & Converter from S-expresion source code to PDF format (M. Butterick)\\%[1em]
+
+\hline
+  \texttt{quadU} & Converter from S-expresion source code to PDF format  (B. Greenman) \\%[1em]
+
+\hline
+  \texttt{snake} & Functional implementation of the  Snake video game (D. Van Horn) \\%[1em]
+
+\hline
+  \texttt{suffixtree} & Algorithm for common longest subsequences between strings. (D. Yoo) \\%[1em]
+
+\hline
+  \texttt{synth} & Converter of description of notes and drum beats to WAV format (V. St-Amour \& N. Toronto) \\%[1em]
+
+\hline
+  \texttt{take5} & Card game simulator (M.Felleisen)  \\%[1em]
+
+\hline
+  \texttt{tetris} & Functional implementation of Tetris (D. Van Horn) \\%[1em]
+
+\end{tabular}
+  \caption{Benchmarks summary.}
+  \label{table:benchmark-descriptions}
+\end{figure}
+
+Of course, the gradual typing benchmarks have no mistakes for the rational
+programmer to debug. Therefore, we follow
+~\cite{lksfd-popl-2020} and we inject bugs with mutation analysis. 
+A mutations is a local syntactic change of the code of a program that may
+produce a bug. Usually, the outcome of mutating a program
+once, dubbed a mutant, can be distinguished from the original program with a test case. 
+The better the test suite of a program, the more mutants it can ``kill.''
+Thus a big part of research on mutation analysis goes into designing
+operations that perform mutations, dubbed mutators, that create mutants 
+that are hard to ``kill.'' Unlike the standard mutators, that also ~\cite{lksfd-popl-2020}
+use, the mutators we need to evaluate the rational programmer have to be
+fined-tune to generate shallow errors. Typed Racket's type system detects 
+typos not convoluted logical issues and mutators suitable for our purposes
+should target the former rather than the latter.
+
+The table in figure~\ref{table:mutation-ops}  summarizes our mutators. 
+For each one, the table provides a short description and an example
+mutation it performs. Each mutator is inspired by the more-than-a-decade
+long experience of the authors making type errors in Typed Racket. 
+Our first ten mutators capture usual common in every type language. 
+For instance, position swaps the arguments of a function call.d
+The last three mutator target distinguishing features of Typed Racket. 
+Specifically, \texttt{arithmetic} may replace a \texttt{+} with a \texttt{-} 
+in an attempt to change the type of the result of the arithmetic
+operation. In Typed Racket, \texttt{+}'s result is a
+\texttt{Positive-Integer} when all arguments are
+\texttt{Positive-Integer}. However, the result of \texttt{-} is 
+\texttt{Integer}. In the same spirit, \texttt{boolean} aims to take
+advantage of Typed Racket's ``truthiness.'' Finally,
+\texttt{conditional}'s goal is to confuse occurrence typing.
+
+\begin{figure}
+  \begin{tabular}{p{2.1cm} | p{5cm}  | p{5.5cm} }
+    {\bf name} & {\bf description} & {\bf example} \\
+
+    \texttt{constant}
+& Swap a literal constant with another of the same value but different type.
+& \texttt{5.6} $\rightarrow$ \texttt{5.6+0.0i} \\
+\hline 
+    \texttt{deletion}
+    & Deletes the result expression of a \texttt{begin} or \texttt{begin0}
+    sequence.
+    & \texttt{(begin x y z)} $\rightarrow$ \texttt{(begin x y)} \\
+\hline
+
+ \texttt{id}
+ & Swaps a use of a top-level identifier defined with with another defined
+ in the same module.
+ &\texttt{(f ...)} $\rightarrow$ \texttt{(g ...)}\\
+\hline
+
+    \texttt{list}
+    & Swap \texttt{car} and \texttt{cdr}.
+    & \texttt{car} $\rightarrow$ \texttt{cdr} \\
+ \hline
+
+   \texttt{position}
+ & Swap the position of subexpressions.
+  &  \begin{tabular}[t]{@{}l}
+     \texttt{(f a 42 "b" 0)}\\
+       $\rightarrow$\\ 
+      \texttt{(f a 42 0 "b")}
+    \end{tabular}\\
+\hline
+
+\texttt{class:init}
+  & Swap default values of class fields.
+   &  \begin{tabular}[t]{@{}l}
+     \texttt{(class object\%}\\
+     \texttt{\,\,(field [a 5] [b "hello"]))}\\
+       $\rightarrow$\\ 
+      \texttt{(class object\%}\\
+     \texttt{\,\,(field [a "hello"] [b 5]))}\\
+    \end{tabular}\\
+\hline
+
+\texttt{class:method}
+ & Add an extra trivial public method to a class.
+ &  \begin{tabular}[t]{@{}l}
+   \texttt{(class object\%  ...)}\\
+       $\rightarrow$\\ 
+     \texttt{(class object\%}\\
+     \texttt{\,\,(define/public (extra x) x)}\\
+     \texttt{\,\,...)}
+    \end{tabular}\\
+\hline
+
+\texttt{class:parent}
+    & Replace the parent of classes with \texttt{object\%}.
+    &  \begin{tabular}[t]{@{}l}
+     \texttt{(class foo\% ...)}\\
+       $\rightarrow$\\ 
+      \texttt{(class object\% ...)}\\
+    \end{tabular}\\
+\hline
+
+\texttt{class:public}
+    & Make a public method private and vice versa.
+    & 
+    \begin{tabular}[t]{@{}l}
+      \texttt{(class object\%}\\
+        \texttt{\,\,(define/public (m x) x))}\\
+       $\rightarrow$\\ 
+      \texttt{(class object\%}\\
+        \texttt{\,\,(define/private (m x) x))}\\
+    \end{tabular}\\
+\hline
+
+ \texttt{class:super}
+    & Remove \texttt{super-new} calls from class definitions.
+    &  \begin{tabular}[t]{@{}l}
+      \texttt{(class object\% (super-new))}\\
+       $\rightarrow$\\ 
+      \texttt{(class object\% (void))}\\
+    \end{tabular}\\
+\hline
+
+\texttt{arithmetic} 
+& Swap an arithmetic operator with another one of the same (or greater)
+    arity and a possibly different result type. 
+    & \texttt{+} $\rightarrow$ \texttt{-} \\
+\hline
+
+\texttt{boolean} 
+& Swap \texttt{and} and \texttt{or}. Due to Racket's truthiness it
+    possibly changes the type of the operation's result. 
+& \texttt{and} $\rightarrow$ \texttt{or} \\
+
+\hline
+
+\texttt{conditional}
+    & Negate conditional test expressions.
+    &  \begin{tabular}[t]{@{}l}
+      \texttt{(cond [(= x 0) 42] ...)}\\ 
+       $\rightarrow$\\ 
+      \texttt{(cond [(not (= x 0)) 42] ...)}
+    \end{tabular}\\
+\hline
+
+\end{tabular}
+  \caption{Summary of mutators.}
+  \label{table:mutation-ops}
+\end{figure}
+
+An effective set of mutators for type errors should be able to generate
+a large number of ill-typed programs. Hence, to validate that our experience 
+with Typed Racket translates to effective mutators, we analyze the
+ mutants of our ten benchmarks. Figure~\ref{fig:mutators-static-hit} shows
+ the results. For each benchmark, our mutators generate between 156 and
+ 18165 ill-typed mutants. Most importantly the collective success rate of
+ the mutators  per benchmark, i.e., the percentage of ill-typed mutants 
+ over the number of mutants, ranges between 53\% (\texttt{tetris})
+ and 95\% (\texttt{kCFA}). Finally, the figure depicts the individual success rate 
+ of mutators per benchmark which reveals that there is at
+ least one benchmark where each mutator performs well
+ (>90\%).
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{./plots/code-hit-ratios}
+  \caption{Success rate of mutators (the percentage of ill-typed mutants 
+  over the number of mutants)}
+  \label{fig:mutators-static-hit}
+\end{figure}
+
+So far we have established that our mutators and~\citet{gtnffvf-jfp-2019}'s 
+benchmarks are sufficient to create a large number of realistic ill-typed Type Racket
+programs. However, for ill-typed programs if their types are correct, like
+in our case, it is trivial to figure out which component contains the bug.
+It is the component that doesn't type check. Thus the ill-typed mutants
+do not offer an interesting setting to try out the debugging abilities of
+rational programmers. That said, we can derive interesting debugging
+scenarios by removing type specifications from some of the components of
+the benchmarks. The resulting mix-typed programs may be well-typed and when
+run may raise a type error. Section~\ref{sec:rational} explains how we can
+explore the space of debugging scenarios in a principled manner. As a
+final remark in this section, with the data from
+figure~\ref{fig:mutators-runtime-hit}, we verify  that in each of the three semantics
+we target, our mutators generate mutants that half the time give rise to at least one
+interesting debugging scenario.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{./plots/code-dynamic-error}
+  \caption{}
+  \label{fig:mutators-runtime-hit}
+\end{figure}

### 2020-07-20 writing/mutate.tex
working on mutation

@@
 benchmarks are sufficient to create a large number of realistic ill-typed Type Racket
-programs. However, for ill-typed programs if their types are correct, like
+programs. However, in an ill-typed program with correct types, like
 in our case, it is trivial to figure out which component contains the bug.
@@
 figure~\ref{fig:mutators-runtime-hit}, we verify  that in each of the three semantics
-we target, our mutators generate mutants that half the time give rise to at least one
-interesting debugging scenario.
+we target, between 25\% and
+90\% of the ill-typed mutants each mutator generates give rise to at least one interesting debugging scenario.
 

### 2020-07-20 writing/rational.tex
working on mutation

@@
 evaluate the effectiveness of a rational programmer. The components of
-each program create a \emph{debugging scenario space}, a lattice like
-the one introduced by~\citet{tfgnvf-popl-2016} for performance
-evaluations. This space consists of variants of the program, dubbed
+an ill-typed program, such as one of  our ill-typed mutants, 
+create a \emph{debugging scenario space}, a lattice like
+the one introduced by ~\citet{tfgnvf-popl-2016} for performance
+evaluations. As we discuss briefly at the end of ~\ref{sec:mutate}, 
+this space consists of variants of the program, dubbed
 \emph{scenarios}, that differ in their type specifications. In the
@@
 
-In detail we answer $Q_1$ for a mutant by creating trails for each natural blame,
+In detail we answer $Q_1$ for an ill-typed mutant from
+section~\ref{sec:mutate} by creating trails for each natural blame,
 natural exception and natural random trail in the scenario lattice of the mutant. A
@@
 successful than natural exceptions and pure chance and we obtain the answer for
-$Q_1$.  The process is analogues for the other questions employing the respective
-modes of the rational programmer.
+$Q_1$.  The process is analogous for the other questions employing the respective
+modes of the rational programmer. Figure~\ref{fig:process} summarizes the
+experimental process for one mode of the rational programmer and connects
+it with the mutations from section~\ref{sec:mutation}.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{./Images/process}
+  \caption{The experimental process for one mode of the rational
+  programmer}
+  \label{fig:process}
+\end{figure}
 

### 2020-07-21 writing/mutate.tex
tex typo

@@
 
-  \textttt{acquire} & Board game simulation (M. Felleisen)  \\%[1em]
+  \texttt{acquire} & Board game simulation (M. Felleisen)  \\%[1em]
 
@@
 programmer to debug. Therefore, we follow
-~\cite{lksfd-popl-2020} and we inject bugs with mutation analysis. 
+~\citet{lksfd-popl-2020} and we inject bugs with mutation analysis. 
 A mutations is a local syntactic change of the code of a program that may
@@
 operations that perform mutations, dubbed mutators, that create mutants 
-that are hard to ``kill.'' Unlike the standard mutators, that also ~\cite{lksfd-popl-2020}
-use, the mutators we need to evaluate the rational programmer have to be
-fined-tune to generate shallow errors. Typed Racket's type system detects 
-typos not convoluted logical issues and mutators suitable for our purposes
+that are hard to ``kill.'' Unlike the standard mutators, which \citet{lksfd-popl-2020}
+use, too, the mutators we need to evaluate the rational programmer have to be
+fine-tuned to generate shallow errors. Typed Racket's type system detects 
+typos not logical mistakes, and mutators suitable for our purposes
 should target the former rather than the latter.

### 2020-07-24 writing/challenges.tex
threat from rationality; point to bad type annotations as future work
changing motivation for mutation

@@
 
-One derivative challenge of the first one is the problem of simulating the
-programmer's mistakes or, more precisely, the origin of impedance
-mismatches. Before a programmer can try to find mistakes, mistakes have to
-be made. Technically the problem is the lack of a sufficiently large
-available corpus of type errors for a meaningful experiment.
-Following~\citet{lksfd-popl-2020} again, the solution is to use mutation
-analysis~\cite{lipton1971fault, demillo1978hints, jia2011analysis} to
-create mistakes in a synthetic manner. This process yields a large number
-of buggy programs whose mixed-typed versions are suitable subjects for our
-experiment.  The mutation analysis of Lazarek et al, is, however,
-completely useless for creating impedance mismatches.
+A derivative challenge is the problem of simulating the programmer's
+mistakes or, more precisely, the origin of impedance mismatches. Before a
+programmer can try to find mistakes, mistakes have to be made.
+Specifically, a necessary step to test the rational programmer is to find
+a collection of  type mistakes in  mix-typed programs that is highly
+representantive of both mistakes and programs ``in the wild.''
+Unfortunately no such collection exists.  There is a good reason for that;
+the kind of mistakes the rational programmer targets are easily detected
+by unit tests and, even if they are tough to weed out, they do not make it
+to code repositories.  In other words, the problem is how to generate a
+sufficiently large and representantive corpus of type mistakes for a meaningful
+experiment.  Following~\citet{lksfd-popl-2020} again, the solution is to
+use mutation analysis~\cite{lipton1971fault, demillo1978hints,
+jia2011analysis} to create mistakes in a synthetic manner. This process
+yields a large number of buggy programs whose mixed-typed versions are
+suitable subjects for our experiment.  The mutation analysis of Lazarek et
+al, is, however, completely useless for creating impedance mismatches.
 
@@
 reliably cause runtime exceptions in mixed-typed versions of mutants.
-
-%Another derivative challenge is the problem of mistakes in type
-%annotations themselves.  Gradual typing enables programmers to move an untyped
-%code base into the typed realm on an incremental basis. The retroactive addition
-%of a specific type to a component, even a library, may result from a
-%misunderstanding of the code and may thus be a mistake itself. Anecdotal 
-%evidence suggests that this scenario is quite common~\cite{sta-nt-base-types,
-%incorrect-ts, wmwz-ecoop-2017}, and hence must be addressed by any evaluation
-%method. Conversely, the method not only demands mutators for code that cause
-%good impedance mismatches, it also calls for a set of mutators that ``break''
-%type annotations without breaking the code itself.
 
@@
 
-\item Construct a large number of suitable benchmarks for all gradual type
-  systems with representative
-  type errors  (section~\ref{sec:mutate});
+\item Generate a large set of programs with representative
+  type errors that all three systems can detect (section~\ref{sec:mutate});
 
-\item Simulate the rational programmer with sufficient 
-  modes to (i) match the peculiarities of each gradual type system and 
+\item Simulate the rational programmer on these programs with sufficient 
+  modes to (i) match the peculiarities of each  system and 
     (ii) isolate blame from other confounding factors

### 2020-07-24 writing/mutate.tex
changing motivation for mutation

@@
-\section{How to Make Benchmarks for Blame} \label{sec:mutate}
+\section{How to Make Lots of Mistakes} \label{sec:mutate}
 
-An important challenge from~\ref{sec:challenges} is the lack of a large and diverse real-world
- corpus of  mix-typed programs with type errors that we can use to
- evaluate rational programmers. To overcome this issue, 
-we create a synthetic corpus of such programs. In this section, we argue
-that despite its synthetic nature our corpus of programs is realistic 
-and diverse.
+An important challenge from~\ref{sec:challenges} is the generation of
+mix-typed programs with representative type mistakes. In this section, we
+present our generation process and we argue that despite its synthetic
+nature it results in a large corpus of programs with type mistakes that
+are realistic and diverse.
 

### 2020-07-24 writing/introduction.tex
adding to the intro academic claim about useless blame

@@
 
-From the very beginning, the theory of gradual typing has come with a blame
-theorem~\cite{mf-toplas-2009, tf-dls-2006}. ``Well-typed [components]\footnote{The
-original authors got this word wrong.} can't be blamed'' turned the theorem into a
-slogan~\cite{wf-esop-2009}. Indeed, academic implementations---for example,
-Reticulated Python~\cite{vsc-dls-2019, vss-popl-2017, vksb-dls-2014} and Typed
+From the very beginning, the theory of gradual typing has come with a
+blame theorem~\cite{mf-toplas-2009, tf-dls-2006}. ``Well-typed
+[components]\footnote{The original authors got this word wrong.} can't be
+blamed'' turned the theorem into a slogan~\cite{wf-esop-2009}. Indeed,
+academic implementations---for example, Reticulated
+Python~\cite{vsc-dls-2019, vss-popl-2017, vksb-dls-2014} and Typed
 Racket~\cite{tf-dls-2006,tf-popl-2008,tfffgksst-snapl-2017,tf-icfp-2010}---come
-with sophisticated blame assignment strategies. Theoreticians firmly embrace the
-idea that blame can help practicing programmers find impedance mismatches between
-the types they added to a software system and the behavior of the remaining untyped
-components.
+with sophisticated blame assignment strategies. The academics that
+created them firmly embrace the idea that blame can help practicing
+programmers find impedance mismatches between the types they added to a
+software system and the behavior of the remaining untyped components.
 
-Practical gradual type systems ignore blame assignment almost completely.
+However, many academics view blame as an interesting but laborious
+orthogonal add-on to 
+gradual typing. Thus blame is often omitted from theoretical models and
+academic implementations alike.
+Other academics go one step further and doubt its pragmatic significance altogether~\cite{cc-snapl-19}.
+
+In line with the latter, practical gradual type systems ignore blame assignment almost completely.
 Industrial systems such as Flow,\footnote{\url{https://flow.org}}
@@
 
-The contrast between academic and industrial systems raises the question 
+The divisions in the academic and industrial world of gradual typing raise the question 
 \begin{quote}

### 2020-07-24 writing/introduction.tex
adding to the intro academic claim about useless blame

@@
 \end{quote}
-Given the prominence of blame in academia and its complete absence in
+Given the long standing academic interest in blame  and its complete absence in
 industrial systems, it comes as an even bigger surprise that the research

### 2020-07-27 writing/introduction.tex
Fix the injection of Chen and Campora:
- restore parallelism between the opening of the first two paragraphs
- summarize C & C's minor contribution as a single sentence at bottom of para 2
- restore the "contrast" wording in third paragraph to play on on the parallelism of 1 and 2

@@
 
-From the very beginning, the theory of gradual typing has come with a
-blame theorem~\cite{mf-toplas-2009, tf-dls-2006}. ``Well-typed
+Theoreticians of gradual typing have focused on
+blame theorems from the very beginning~\cite{mf-toplas-2009, tf-dls-2006}. ``Well-typed
 [components]\footnote{The original authors got this word wrong.} can't be
@@
 
-However, many academics view blame as an interesting but laborious
-orthogonal add-on to 
-gradual typing. Thus blame is often omitted from theoretical models and
-academic implementations alike.
-Other academics go one step further and doubt its pragmatic significance altogether~\cite{cc-snapl-19}.
-
-In line with the latter, practical gradual type systems ignore blame assignment almost completely.
-Industrial systems such as Flow,\footnote{\url{https://flow.org}}
+Industrial implementors of  gradual typing systems have almost completely ignored blame assignment.
+Systems such as Flow,\footnote{\url{https://flow.org}}
 Hack,\footnote{\url{https://hacklang.org}} or
 TypeScript\footnote{\url{https://www.typescriptlang.org}} exploit types for IDE actions and for finding typos in
-code. Then these compilers remove types and rely on the built-in safety checks of
+code. Then their compilers remove types and rely on the built-in safety checks of
 the underlying programming language to catch any problems. 
+Recently, \citet{cc-snapl-19} have turned this implicitly expressed
+skepticism into an explicit statement and challenge.
 
-The divisions in the academic and industrial world of gradual typing raise the question 
+This contrast between theoretical research and industrial applications of gradual typing raises the question 
 \begin{quote}

### 2020-07-27 writing/challenges.tex
simple touch ups to prose on challenges and mutate

@@
 the kind of mistakes the rational programmer targets are easily detected
-by unit tests and, even if they are tough to weed out, they do not make it
-to code repositories.  In other words, the problem is how to generate a
+by unit tests and, even if their sources are hard to find, these mistakes do not make it
+into code repositories with appropriate commit messages.  In other words, the problem is how to generate a
 sufficiently large and representantive corpus of type mistakes for a meaningful

### 2020-07-27 writing/mutate.tex
simple touch ups to prose on challenges and mutate

@@
 
-An important challenge from~\ref{sec:challenges} is the generation of
-mix-typed programs with representative type mistakes. In this section, we
-present our generation process and we argue that despite its synthetic
-nature it results in a large corpus of programs with type mistakes that
-are realistic and diverse.
+An important challenge from section~\ref{sec:challenges} is the generation of
+mix-typed programs with representative impedance mismatches between
+the typed components, their types, and the behavior of untyped components. In this section, we
+present a generation process and argue that, despite its synthetic
+nature, it yields a large corpus of programs with  realistic and diverse type mistakes.
 

### 2020-08-10 writing/mutate.tex
Add subsection describing selection of debugging scenarios

@@
 So far we have established that our mutators and~\citet{gtnffvf-jfp-2019}'s 
-benchmarks are sufficient to create a large number of realistic ill-typed Type Racket
+benchmarks are sufficient to create a large number of realistic ill-typed Typed Racket
 programs. However, in an ill-typed program with correct types, like

### 2020-08-10 writing/rational.tex
Add subsection describing selection of debugging scenarios

@@
 
+\subsection{Where do the trails begin?}
+The above descriptions of blame trails under various modes all assume that every trail has a beginning scenario.
+How we select the scenarios from which the trails start has ramifications for the kinds of results we can find, and what we can learn from them.
+We break this decision into three basic factors, each of which specify an element of the selection.
+
+First, not all debugging scenarios are suitable for evaluating blame across all three semantics.
+For instance, there exist scenarios where Natural's checks discover a problem but Transient and Erasure do not.
+Since Transient and Erasure provide no debugging information (and in particular no blame) in such cases, we have nothing to evaluate.
+As alluded to in the previous paragraph, scenarios where the type checker detects the problem statically are also uninteresting.
+Hence, we select only scenarios which are interesting in all three semantics.
+We can do this simply by leveraging~\citet{???
+Ben}'s proof that a dynamic type error in erasure implies a dynamic type error in both Transient and Natural, to only look at scenarios in which Erasure detects the problem.
+In other words, we consider just those debugging scenarios that are interesting under Erasure.
+
+The second factor is that we would like to be able to compare the behavior of blame between each of our three semantics.
+In order to do so, we must evaluate each of the semantics on the same set of scenarios;
+this allows us to compare for any given scenario how Natural compares to Erasure, for instance.
+Hence, we pre-select a sample of scenarios which serve as the starting points to evaluate all three semantics.
+
+Finally, the third factor is that we only want to consider debugging scenarios in which the bug has effects that cross component boundaries.
+In the context of migratory typing, dynamic type enforcement occurs when values cross component boundaries.
+However, a bug may be caught by the runtime before any values it influences flow out of the containing component -- for instance, if the bug is that a direct argument to \texttt{+} becomes a string.
+We consider such scenarios trivial, becase typing the component in which the problem manifests immediately identifies that same component as the buggy one.
+In an effort to avoid these trivial scenarios, we further restrict our sample of debugging scenarios to only those in which the stack trace from Erasure identifies a component other than the buggy one.
+Those scenarios may yet be trivial in the other two semantics -- with more checking and blame tracking -- but this selection ensures that none of our scenarios are trivial ``a priori''.
+
 %% -----------------------------------------------------------------------------

### 2020-08-12 writing/mutate.tex
Move debugging scenarios description into previous section

--- 2020-07-27 writing/mutate.tex
@@
 
+%% ll: Is this still necessary?
+As a
+final remark in this section, with the data from
+figure~\ref{fig:mutators-runtime-hit}, we verify  that in each of the three semantics
+we target, between 25\% and
+90\% of the ill-typed mutants each mutator generates give rise to at least one interesting debugging scenario.
+
+\subsection{What are interesting debugging scenarios?}
 So far we have established that our mutators and~\citet{gtnffvf-jfp-2019}'s 
-benchmarks are sufficient to create a large number of realistic ill-typed Type Racket
+benchmarks are sufficient to create a large number of realistic ill-typed Typed Racket
 programs. However, in an ill-typed program with correct types, like
@@
 run may raise a type error. Section~\ref{sec:rational} explains how we can
-explore the space of debugging scenarios in a principled manner. As a
-final remark in this section, with the data from
-figure~\ref{fig:mutators-runtime-hit}, we verify  that in each of the three semantics
-we target, between 25\% and
-90\% of the ill-typed mutants each mutator generates give rise to at least one interesting debugging scenario.
+explore the space of debugging scenarios in a principled manner starting from
+an interesting scenario.
+In the rest of this section we describe how we select interesting scenarios.
+We break this decision into three basic factors, each of which specify an element defining our selection of interesting scenarios.
+
+First, not all debugging scenarios are suitable for evaluating blame across all three semantics.
+For instance, there exist scenarios where Natural's checks discover a problem but Transient and Erasure do not.
+Since Transient and Erasure provide no debugging information (and in particular no blame) in such cases, we have nothing to evaluate.
+As alluded to in the previous paragraph, scenarios where the type checker detects the problem statically are also uninteresting.
+Hence, we select only scenarios which are interesting in all three semantics.
+Based on~\citet{gfd-oopsla-2019}'s proof that a dynamic type error in erasure implies a dynamic type error in both Transient and Natural, for any scenario that Erasure detects the problem, both Transient and Natural should detect it as well.
+We can therefore define scenarios that raise a dynamic type error in Erasure as interesting, and all such scenarios should raise dynamic type errors in Transient and Natural, too.
+
+The second factor is that we would like to be able to compare the behavior of blame between each of our three semantics.
+In order to do so, we must evaluate each of the semantics on the same set of scenarios;
+this allows us to compare for any given scenario how Natural compares to Erasure, for instance.
+Hence, we pre-select a sample of scenarios which serve as the starting points to evaluate all three semantics.
+
+Finally, the third factor is that we only want to consider debugging scenarios in which the bug has effects that cross component boundaries.
+In the context of migratory typing, dynamic type enforcement occurs when values cross component boundaries.
+However, a bug may be caught by the runtime before any values it influences flow out of the containing component -- for instance, if the bug is that a direct argument to \texttt{+} becomes a string.
+We consider such scenarios trivial, becase typing the component in which the problem manifests immediately identifies that same component as the buggy one.
+In an effort to avoid these trivial scenarios, we further restrict our definition of interesting debugging scenarios to be those in which the stack trace from Erasure identifies a component other than the buggy one.
+
+Thus we have defined the set of interesting debugging scenarios as those for which Erasure raises a dynamic type error in some component other than the buggy one.
 

### 2020-08-15 writing/discussion.tex
discussion: transient blame TR vs RP

@@
 assignment for trail extension as our experimental setup describes.
+
+\subsection{Threat: Is our Transient Correct?}
+%% purpose: talk about transient-blame implementation .... what did we do,
+%%  how did we test, what threats remain?
+%%
+%% the benchmark timeouts look like a threat, but this section shows that
+%%  we've looked carefully at those
+
+An apparent flaw in our transient is the high cost of blame.
+Whereas \citet{vss-popl-2017} report an average slowdown of 2.5x and
+ worst-case of 5.4x on fully-typed benchmarks in Reticulated,
+ our configurations frequently exceed a 10 minute timeout with transient blame.
+Unfortunately, our high cost is closer to the truth;
+ Reticulated has two advantages that give low overhead in the prior work.
+First, Reticulated's type inference often assigns the dynamic type to local
+ variables in "fully typed" code.
+This flexible type gives a programmer fewer static errors and less-precise
+ blame,; but it also makes the benchmarks run faster.
+Second, \citet{vss-popl-2017} use relatively small benchmarks.
+Four have since been retired from the official Python benchmark suite
+ because they are too small, unrealistic, and unstable.
+On larger programs, Reticulated suffers from high overhead.
+For example, a Reticulated version of our \texttt{sieve} benchmark runs in
+ ~40 seconds without blame, and with blame times out after 10 minutes.
+
+%% NOTE: Vitousek's benchmarks are from the Python "pyperformance" suite.
+%%   The version notes in the docs talk about retiring benchmarks, but
+%%   you can also look at the current codebase and see what names from POPL'17
+%%   are missing: callsimple, callmethod, callmethodslots, & pystone
+%% <https://github.com/python/pyperformance/tree/master/pyperformance/benchmarks>

### 2020-08-18 writing/discussion.tex
discussion: note Retic missing casts/checks

@@
 Unfortunately, our high cost is closer to the truth;
- Reticulated has two advantages that give low overhead in the prior work.
-First, Reticulated's type inference often assigns the dynamic type to local
- variables in "fully typed" code.
-This flexible type gives a programmer fewer static errors and less-precise
- blame,; but it also makes the benchmarks run faster.
-Second, \citet{vss-popl-2017} use relatively small benchmarks.
+ there are at least three broad issues that skew the earlier results.
+First, the 2017 implementation of Reticulated fails to insert certain
+ soundness checks\footnote{Missing check: \url{https://github.com/mvitousek/reticulated/issues/36}}
+ and blame-map updates\footnote{Missing cast: \url{https://github.com/mvitousek/reticulated/issues/43}}
+ from the paper.
+Second, Reticulated's type checker assigns the dynamic type to local
+ variables in the "fully typed" benchmarks---often because the type system
+ cannot articulate a more precise type.
+Code that ends up with the dynamic type has fewer constraints to check at run-time.
+Third, \citet{vss-popl-2017} use relatively small benchmarks.
 Four have since been retired from the official Python benchmark suite
- because they are too small, unrealistic, and unstable.
+ because they are too small, unrealistic, and unstable.\footnote{Release notes: \url{https://pyperformance.readthedocs.io/changelog.html}}
 On larger programs, Reticulated suffers from high overhead.

### 2020-11-04 writing/mutate.tex
mutate: Add definition of interesting debugging scenarios

@@
 ~\citet{lksfd-popl-2020} and we inject bugs with mutation analysis.
-A mutations is a local syntactic change of the code of a program that may
-produce a bug. Usually, the outcome of mutating a program
+A mutation is a local syntactic change of the code of a program that may
+produce a bug.
+\ll{Need citations for mutation}
+Usually, the outcome of mutating a program
 once, dubbed a mutant, can be distinguished from the original program with a test case.
@@
 mutation it performs. Each mutator is inspired by the more-than-a-decade
-long experience of the authors making type errors in Typed Racket. 
-Our first ten mutators capture usual common in every type language. 
-For instance, position swaps the arguments of a function call.d
-The last three mutator target distinguishing features of Typed Racket. 
+long experience of the authors making type-level mistakes in Typed Racket.
+We find that these mistakes often take non-trivial effort to debug.
+
+Most of our mutators in figure~\ref{table:mutation-ops} are self-explanatory.
+The first twelve can apply to most gradually typed languages, including those with classes.
+The last four mutators target distinguishing features of Typed Racket.
 Specifically, \texttt{arithmetic} may replace a \texttt{+} with a \texttt{-}
@@
 \texttt{Integer}. In the same spirit, \texttt{boolean} aims to take
-advantage of Typed Racket's ``truthiness.'' Finally,
-\texttt{conditional}'s goal is to confuse occurrence typing.
+advantage of Typed Racket's ``truthiness.'' Finally, \texttt{negate-cond} and \texttt{force-cond}
+aim to confuse occurrence typing.
 
@@
 \hline
+
     \texttt{deletion}
-    & Deletes the result expression of a \texttt{begin} or \texttt{begin0}
-    sequence.
+    & Deletes the result expression of a sequence.
     & \texttt{(begin x y z)} $\rightarrow$ \texttt{(begin x y)} \\
@@
 
- \texttt{id}
- & Swaps a use of a top-level identifier defined with with another defined
- in the same module.
- &\texttt{(f ...)} $\rightarrow$ \texttt{(g ...)}\\
+    \texttt{top-level-id}
+ & Swaps two identifiers defined in the same module.
+ &\texttt{(f x 42)} $\rightarrow$ \texttt{(g x 42)}\\
+\hline
+
+    \texttt{imported-id}
+ & Swaps two identifiers imported from the same module.
+ &\texttt{(f x 42)} $\rightarrow$ \texttt{(g x 42)}\\
+\hline
+
+    \texttt{method-id}
+ & Swaps two method identifiers.
+ &\texttt{(send o f x 42)} $\rightarrow$ \texttt{(send o g x 42)}\\
+\hline
+
+    \texttt{field-id}
+ & Swaps two field identifiers.
+    & \begin{tabular}[t]{@{}l} \texttt{(get-field o f)} $\rightarrow$ \\
+        \texttt{(get-field o g)}
+      \end{tabular}\\
+\hline
+
+   \texttt{position}
+ & Swaps two subexpressions.
+  &  \texttt{(f a 42 "b" 0)} $\rightarrow$\texttt{(f a 42 0 "b")} \\
 \hline
@@
      \texttt{list}
-    & Swap \texttt{car} and \texttt{cdr}.
-    & \texttt{car} $\rightarrow$ \texttt{cdr} \\
- \hline
-
-   \texttt{position}
- & Swap the position of subexpressions.
+    & Replaces \texttt{append} with \texttt{cons}.
     &  \begin{tabular}[t]{@{}l}
-     \texttt{(f a 42 "b" 0)}\\
-       $\rightarrow$\\ 
-      \texttt{(f a 42 0 "b")}
+      \texttt{append} $\rightarrow$ \texttt{cons}
     \end{tabular}\\
@@
      \texttt{class:init}
-  & Swap default values of class fields.
+  & Swaps values of class initializers.
    &  \begin{tabular}[t]{@{}l}
-     \texttt{(class object\%}\\
-     \texttt{\,\,(field [a 5] [b "hello"]))}\\
-       $\rightarrow$\\ 
-      \texttt{(class object\%}\\
-     \texttt{\,\,(field [a "hello"] [b 5]))}\\
-    \end{tabular}\\
-\hline
-
-\texttt{class:method}
- & Add an extra trivial public method to a class.
- &  \begin{tabular}[t]{@{}l}
-   \texttt{(class object\%  ...)}\\
-       $\rightarrow$\\ 
-     \texttt{(class object\%}\\
-     \texttt{\,\,(define/public (extra x) x)}\\
-     \texttt{\,\,...)}
+     \texttt{(new c [a 5] [b "hello"])} $\rightarrow$\\
+     \texttt{(new c [a "hello"] [b 5])}\\
     \end{tabular}\\
@@
      \texttt{class:parent}
-    & Replace the parent of classes with \texttt{object\%}.
+    & Replaces the parent of classes with \texttt{object\%}.
     &  \begin{tabular}[t]{@{}l}
-     \texttt{(class foo\% ...)}\\
-       $\rightarrow$\\ 
-      \texttt{(class object\% ...)}\\
+     \texttt{(class foo\% (super-new))} $\rightarrow$\\
+     \texttt{(class object\% (super-new))}\\
     \end{tabular}\\
@@
      \texttt{class:public}
-    & Make a public method private and vice versa.
-    & 
-    \begin{tabular}[t]{@{}l}
+    & Makes a public method private and vice versa.
+    & \begin{tabular}[t]{@{}l}
       \texttt{(class object\%}\\
-        \texttt{\,\,(define/public (m x) x))}\\
-       $\rightarrow$\\ 
+      \phantom{(d}\texttt{(define/public (m x) x))} $\rightarrow$\\
       \texttt{(class object\%}\\
-        \texttt{\,\,(define/private (m x) x))}\\
+      \phantom{(d}\texttt{(define/private (m x) x))}\\
     \end{tabular}\\
@@
      \texttt{class:super}
-    & Remove \texttt{super-new} calls from class definitions.
+    & Removes \texttt{super-new} calls from class definitions.
     &  \begin{tabular}[t]{@{}l}
-      \texttt{(class object\% (super-new))}\\
-       $\rightarrow$\\ 
-      \texttt{(class object\% (void))}\\
+      \texttt{(class foo\% (super-new))} $\rightarrow$\\
+      \texttt{(class foo\% (void))}\\
     \end{tabular}\\
@@
      \texttt{arithmetic}
-& Swap an arithmetic operator with another one of the same (or greater)
-    arity and a possibly different result type. 
+& Swaps arithmetic operators.
     & \texttt{+} $\rightarrow$ \texttt{-} \\
@@
      \texttt{boolean}
-& Swap \texttt{and} and \texttt{or}. Due to Racket's truthiness it
-    possibly changes the type of the operation's result. 
+& Swaps \texttt{and} and \texttt{or}.
 & \texttt{and} $\rightarrow$ \texttt{or} \\
-
 \hline
 
-\texttt{conditional}
-    & Negate conditional test expressions.
+     \texttt{negate-cond}
+    & Negates conditional test expressions.
     &  \begin{tabular}[t]{@{}l}
-      \texttt{(cond [(= x 0) 42] ...)}\\ 
-       $\rightarrow$\\ 
-      \texttt{(cond [(not (= x 0)) 42] ...)}
+      \texttt{(if (= x 0) t e)} $\rightarrow$\\
+      \texttt{(if (not (= x 0)) t e)}
        \end{tabular}\\
@@
 
+     \texttt{force-cond}
+    & Replaces conditional test expressions with \texttt{\#t}.
+    &  \begin{tabular}[t]{@{}l}
+      \texttt{(if (= x 0) t e)} $\rightarrow$\\
+      \texttt{(if \#t t e)}
+    \end{tabular}\\
+\hline
 \end{tabular}
@@
 
+\subsection{Do these mutators give us interesting debugging scenarios?}
+\begin{itemize}
+\item do these mutators work?
+\begin{itemize}
+\item now we need to see if these mutators actually produce interesting errors
+\begin{itemize}
+\item define interesting errors: type error + dynamic err w/ 3 modules on stack in a mixed configuration
+\item this section will show if they do this on our benchmarks
+\end{itemize}
+\item do they actually produce interesting errors? -- for this need plot that's inverted version dyn-err-heatmap + total of mutants per benchmark
+\item do we need all of the mutators?
+\begin{itemize}
+\item each one produces type errors + interesting scenarios
+\item we conjecture they represent different kinds of mistakes
+\end{itemize}
+\end{itemize}
+\end{itemize}
+
+Having selected a set of mutators based on our experience, we next need to see if the mutators do in fact produce interesting debugging scenarios in our programs.
+We define an interesting debugging scenario as a mutation that
+\begin{enumerate}
+\item can be caught by the type checker if the program is fully typed,
+\item otherwise produces a dynamic error under Erasure, and
+\item the stacktrace of the dynamic error contains at least three distinct modules from the program.
+\end{enumerate}
+The first criterion reflects the fact that a gradual type system can only catch bugs that would have been caught by a type checker.
+The second criterion ensures that the bug is detectable by all three semantics.
+It is based on~\citet{gfd-oopsla-2019}'s proof that a dynamic type error in Erasure implies a dynamic type error in both Transient and Natural.
+The final criterion conservatively approximates scenarios with the opportunity for interactions between typed and untyped modules. % I say opportunity because it depends on the config
+In particular, it takes at least two modules to have typed/untyped module interactions (one of each kind), and having two modules on the stack ensures that the error arose in the interaction of at least those two modules.
+The third module comes in because every one of our programs comes with a driver module which is usually uninteresting but should be present in every stacktrace.
+With this definition in hand, the rest of this section will analyze how well our mutators produce interesting debugging scenarios.
+
+Beginning with the first criterion of the definition, we analyze how well our mutators produce type errors.
+
+%% Cut from above
 An effective set of mutators for type errors should be able to generate

### 2020-11-06 writing/mutate.tex
mutate: Revise description of mutators and interesting scenarios

@@
 The better the test suite of a program, the more mutants it can ``kill.''
-Thus a big part of research on mutation analysis goes into designing
-operations that perform mutations, dubbed mutators, that create mutants
-that are hard to ``kill.'' Unlike the standard mutators, which \citet{lksfd-popl-2020}
-use, too, the mutators we need to evaluate the rational programmer have to be
-fine-tuned to generate shallow errors. Typed Racket's type system detects
-typos not logical mistakes, and mutators suitable for our purposes
-should target the former rather than the latter.
+Mutants that are hard to ``kill'' usually correspond to subtle programming errors, and mutation frameworks typically come with standard mutators carefully designed to create such mutants.
 
+However, Typed Racket's gradual type system detects shallow errors, unlike those introduced by standard mutators.
+Thus, we need different mutators to evaluate the rational programmer, ones which are fine-tuned to generate shallow errors.
 The table in figure~\ref{table:mutation-ops} summarizes our mutators.
@@
 
+   \texttt{position}
+ & Swaps two subexpressions.
+  &  \texttt{(f a 42 "b" 0)} $\rightarrow$\texttt{(f a 42 0 "b")} \\
+\hline
+
+     \texttt{list}
+    & Replaces \texttt{append} with \texttt{cons}.
+    &  \begin{tabular}[t]{@{}l}
+      \texttt{append} $\rightarrow$ \texttt{cons}
+    \end{tabular}\\
+\hline
+
     \texttt{top-level-id}
@@
         \texttt{(get-field o g)}
-      \end{tabular}\\
-\hline
-
-   \texttt{position}
- & Swaps two subexpressions.
-  &  \texttt{(f a 42 "b" 0)} $\rightarrow$\texttt{(f a 42 0 "b")} \\
-\hline
-
-     \texttt{list}
-    & Replaces \texttt{append} with \texttt{cons}.
-    &  \begin{tabular}[t]{@{}l}
-      \texttt{append} $\rightarrow$ \texttt{cons}
       \end{tabular}\\
@@
 
-\subsection{Do these mutators give us interesting debugging scenarios?}
+\subsection{Are these mutators good enough?}
 \begin{itemize}
@@
 
-Having selected a set of mutators based on our experience, we next need to see if the mutators do in fact produce interesting debugging scenarios in our programs.
-We define an interesting debugging scenario as a mutation that
+Of course, mutators that introduce these kinds of shallow errors directly into the benchmarks do not produce bugs that are interesting from the perspective of the rational programmer.
+The benchmarks are fully typed, so Typed Racket's type checker locates them immediately.
+For that reason, we apply the mutators to versions of the benchmarks with some components stripped of type annotations (dubbed \emph{configurations}).
+Without the annotations, the type checker may not detect the errors in these mutant configurations.
+Instead, Typed Racket's gradual type system may detect the errors dynamically.
+Mutant configurations that result in dynamic errors are the ideal ground to exercise the rational programmer's ability to locate errors.
+
+Specifically, we define an \emph{interesting debugging scenario} as a mutant configuration that
 \begin{enumerate}
-\item can be caught by the type checker if the program is fully typed,
-\item otherwise produces a dynamic error under Erasure, and
-\item the stacktrace of the dynamic error contains at least three distinct modules from the program.
+\item would be rejected by the type checker if we restored the missing type annotations,
+\item produces a dynamic error under Erasure as-is, and
+\item the stacktrace of the error mentions at least three distinct components.
 \end{enumerate}
-The first criterion reflects the fact that a gradual type system can only catch bugs that would have been caught by a type checker.
-The second criterion ensures that the bug is detectable by all three semantics.
-It is based on~\citet{gfd-oopsla-2019}'s proof that a dynamic type error in Erasure implies a dynamic type error in both Transient and Natural.
-The final criterion conservatively approximates scenarios with the opportunity for interactions between typed and untyped modules. % I say opportunity because it depends on the config
-In particular, it takes at least two modules to have typed/untyped module interactions (one of each kind), and having two modules on the stack ensures that the error arose in the interaction of at least those two modules.
-The third module comes in because every one of our programs comes with a driver module which is usually uninteresting but should be present in every stacktrace.
-With this definition in hand, the rest of this section will analyze how well our mutators produce interesting debugging scenarios.
 
-Beginning with the first criterion of the definition, we analyze how well our mutators produce type errors.
+The second and third criteria require a little more explanation.
+The second criterion requires that the error is detectable under Erasure to ensure that it is detectable by all three semantics.
+This is a direct consequence of~\citet{gfd-oopsla-2019}'s theorem that a dynamic type error under Erasure implies a dynamic type error under both Transient and Natural.
+It is worth noting that this criterion excludes from consideration all scenarios where Natural or Transient can detect the bug but Erasure cannot, and thus it favors Erasure.
+However, our goal here is to evaluate blame across these three semantics, and thus we focus on systems where all of them exhibit blame.
 
-%% Cut from above
-An effective set of mutators for type errors should be able to generate
-a large number of ill-typed programs. Hence, to validate that our experience
-with Typed Racket translates to effective mutators, we analyze the
- mutants of our ten benchmarks. Figure~\ref{fig:mutators-static-hit} shows
- the results. For each benchmark, our mutators generate between 156 and
- 18165 ill-typed mutants. Most importantly the collective success rate of
- the mutators  per benchmark, i.e., the percentage of ill-typed mutants
- over the number of mutants, ranges between 53\% (\texttt{tetris})
- and 95\% (\texttt{kCFA}). Finally, the figure depicts the individual success rate
- of mutators per benchmark which reveals that there is at
- least one benchmark where each mutator performs well
- (>90\%).
+The third criterion aims to capture debugging scenarios with a specific kind of dynamic error: one produced by the interactions between typed and untyped components.
+In those cases, the rational programmer may need to examine several components to locate the source of the bad interaction.
+In technical terms, we conservatively approximate these scenarios by analyzing the stacktrace of the error.
+If the stack trace mentions at least two components, then this is evidence of interaction between components.
+We require the third component because our benchmarks come with a driver component which is present in every stacktrace.
 
@@
   \centering
-  \includegraphics[width=\textwidth]{./plots/code-hit-ratios}
-  \caption{Success rate of mutators (the percentage of ill-typed mutants
-  over the number of mutants)}
-  \label{fig:mutators-static-hit}
+  \includegraphics[width=\textwidth]{./plots/mutant-breakdown}
+  \caption{Interesting debugging scenarios produced by our mutators. Counts are cut off at 50.}
+  \label{fig:mutant-breakdown}
 \end{figure}
 
-%% ll: Is this still necessary?
-As a
-final remark in this section, with the data from
-figure~\ref{fig:mutators-runtime-hit}, we verify  that in each of the three semantics
-we target, between 25\% and
-90\% of the ill-typed mutants each mutator generates give rise to at least one interesting debugging scenario.
+With this definition in hand, we analyze how effectively our mutators produce interesting debugging scenarios from our benchmarks.
+Figure~\ref{fig:mutant-breakdown} shows the number of distinct interesting debugging scenarios produced by our mutators for each benchmark.
+The columns correspond to our benchmarks, and each rows show the count of interesting debugging scenarios produced by one mutator from figure~\ref{table:mutation-ops}.
+The counts are cut of at 50, so any bar reaching the right side of the plot means 50 or more scenarios.
 
-\subsection{What are interesting debugging scenarios?}
-So far we have established that our mutators and~\citet{gtnffvf-jfp-2019}'s
-benchmarks are sufficient to create a large number of realistic ill-typed Typed Racket
-programs. However, in an ill-typed program with correct types, like
-in our case, it is trivial to figure out which component contains the bug.
-It is the component that doesn't type check. Thus the ill-typed mutants
-do not offer an interesting setting to try out the debugging abilities of
-rational programmers. That said, we can derive interesting debugging
-scenarios by removing type specifications from some of the components of
-the benchmarks. The resulting mix-typed programs may be well-typed and when
-run may raise a type error. Section~\ref{sec:rational} explains how we can
-explore the space of debugging scenarios in a principled manner starting from
-an interesting scenario.
-In the rest of this section we describe how we select interesting scenarios.
-We break this decision into three basic factors, each of which specify an element defining our selection of interesting scenarios.
-
-First, not all debugging scenarios are suitable for evaluating blame across all three semantics.
-For instance, there exist scenarios where Natural's checks discover a problem but Transient and Erasure do not.
-Since Transient and Erasure provide no debugging information (and in particular no blame) in such cases, we have nothing to evaluate.
-As alluded to in the previous paragraph, scenarios where the type checker detects the problem statically are also uninteresting.
-Hence, we select only scenarios which are interesting in all three semantics.
-Based on~\citet{gfd-oopsla-2019}'s proof that a dynamic type error in erasure implies a dynamic type error in both Transient and Natural, for any scenario that Erasure detects the problem, both Transient and Natural should detect it as well.
-We can therefore define scenarios that raise a dynamic type error in Erasure as interesting, and all such scenarios should raise dynamic type errors in Transient and Natural, too.
-
-The second factor is that we would like to be able to compare the behavior of blame between each of our three semantics.
-In order to do so, we must evaluate each of the semantics on the same set of scenarios;
-this allows us to compare for any given scenario how Natural compares to Erasure, for instance.
-Hence, we pre-select a sample of scenarios which serve as the starting points to evaluate all three semantics.
-
-Finally, the third factor is that we only want to consider debugging scenarios in which the bug has effects that cross component boundaries.
-In the context of migratory typing, dynamic type enforcement occurs when values cross component boundaries.
-However, a bug may be caught by the runtime before any values it influences flow out of the containing component -- for instance, if the bug is that a direct argument to \texttt{+} becomes a string.
-We consider such scenarios trivial, becase typing the component in which the problem manifests immediately identifies that same component as the buggy one.
-In an effort to avoid these trivial scenarios, we further restrict our definition of interesting debugging scenarios to be those in which the stack trace from Erasure identifies a component other than the buggy one.
-
-Thus we have defined the set of interesting debugging scenarios as those for which Erasure raises a dynamic type error in some component other than the buggy one.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{./plots/code-dynamic-error}
-  \caption{}
-  \label{fig:mutators-runtime-hit}
-\end{figure}
+Figure~\ref{fig:mutant-breakdown} illustrates that our mutators do indeed produce a large number of interesting debugging scenarios for all of our benchmarks.
+All of the mutators together produce at least 40 interesting scenarios for every benchmark, and those scenarios are produced by at least four different mutators for each benchmark.
+Thus our mutators produce a large and diverse population of scenarios for every benchmark.
+Furthermore, every mutator contributes scenarios to at least one benchmark.
+Some mutators clearly only target one or two of them, though, and in particular the class-focused mutators are primarily effective in \texttt{take5}.
+This reflects that among our benchmarks \texttt{take5} makes the most extensive use of object-oriented features, while the other features targeted by the rest of the mutators are used more broadly.

### 2020-11-11 writing/mutate.tex
format + prose tweaks

@@
 
-\begin{figure}
+\begin{figure*}
 \begin{tabular}{p{2cm} | p{10cm} }
@@
   \label{table:benchmark-descriptions}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure}
+\begin{figure*}
   \begin{tabular}{p{2.1cm} | p{5cm}  | p{5.5cm} }
@@
   \label{table:mutation-ops}
-\end{figure}
+\end{figure*}
 
 \subsection{Are these mutators good enough?}
-\begin{itemize}
-\item do these mutators work?
-\begin{itemize}
-\item now we need to see if these mutators actually produce interesting errors
-\begin{itemize}
-\item define interesting errors: type error + dynamic err w/ 3 modules on stack in a mixed configuration
-\item this section will show if they do this on our benchmarks
-\end{itemize}
-\item do they actually produce interesting errors? -- for this need plot that's inverted version dyn-err-heatmap + total of mutants per benchmark
-\item do we need all of the mutators?
-\begin{itemize}
-\item each one produces type errors + interesting scenarios
-\item we conjecture they represent different kinds of mistakes
-\end{itemize}
-\end{itemize}
-\end{itemize}
 
@@
 
-\begin{figure}
+\begin{figure*}
   \centering
-  \includegraphics[width=\textwidth]{./plots/mutant-breakdown}
+  \includegraphics[scale=0.35]{./plots/mutant-breakdown}
   \caption{Interesting debugging scenarios produced by our mutators. Counts are cut off at 50.}
   \label{fig:mutant-breakdown}
-\end{figure}
+\end{figure*}
 

### 2020-11-14 writing/mutate.tex
revision of mutate

@@
 nature, it yields a large corpus of programs with  realistic and diverse type mistakes.
+
+\begin{figure*}
+\begin{tabular}{p{2cm} | p{10cm} }
+  {\bf  name} & {\bf description (author)}  \\
+
+\hline
+
+  \texttt{acquire} & Board game simulation (M. Felleisen)  \\%[1em]
+
+\hline
+  \texttt{gregor} & Utilities for calendar dates (J. Zeppieri) \\%[1em]
+
+\hline
+  \texttt{kcfa} & Functional implementation of 2CFA for the lambda calculus (M. Might) \\%[1em]
+
+\hline
+  \texttt{quadT} & Converter from S-expression source code to PDF format (M. Butterick)\\%[1em]
+
+\hline
+  \texttt{quadU} & Converter from S-expression source code to PDF format  (B. Greenman) \\%[1em]
+
+\hline
+  \texttt{snake} & Functional implementation of the  Snake video game (D. Van Horn) \\%[1em]
+
+\hline
+  \texttt{suffixtree} & Algorithm for common longest subsequences between strings. (D. Yoo) \\%[1em]
+
+\hline
+  \texttt{synth} & Converter of notes and drum beats to WAV (V. St-Amour \& N. Toronto) \\%[1em]
+
+\hline
+  \texttt{take5} & Card game simulator (M.Felleisen)  \\%[1em]
+
+\hline
+  \texttt{tetris} & Functional implementation of Tetris (D. Van Horn) \\%[1em]
+
+\end{tabular}
+  \caption{Benchmarks summary.}
+  \label{table:benchmark-descriptions}
+\end{figure*}
 
@@
 
-\begin{figure*}
-\begin{tabular}{p{2cm} | p{10cm} }
-  {\bf  name} & {\bf description (author)}  \\
+Of course, the gradual typing benchmarks have no mistakes for the rational
+programmer to debug. Therefore, we follow ~\citet{lksfd-popl-2020} and we
+inject bugs with mutation
+analysis~\cite{jia2011analysis,demillo1978hints,lipton1971fault}.  A
+mutation is a local syntactic change of the code of a program that may
+produce a bug. Usually, the outcome of mutating a program once, dubbed a
+mutant, can be distinguished from the original program with a test case.
+The better the test suite of a program, the more mutants it can ``kill.''
+Mutants that are hard to ``kill'' usually correspond to subtle programming
+errors, and mutation frameworks typically come with standard mutators
+carefully designed to create such mutants.
 
-\hline
+However,  unlike the deep errors introduced by standard mutators, 
+gradual type systems detect shallow errors.  Thus, in order to
+evaluate the rational programmer in the context of gradual typing, 
+we need to develop alternative mutators
+that are fine-tuned to generate shallow errors.  Our inspiration is our
+more-than-a-decade long experience of making type-level mistakes in Typed
+Racket. Some of these mistakes often take non-trivial effort to debug and
+have become the seed for our mutators, which we summarize in
+figure~\ref{table:mutation-ops}.  For each one, the figure provides a
+short description and an example mutation.
 
-  \texttt{acquire} & Board game simulation (M. Felleisen)  \\%[1em]
-
-\hline
-  \texttt{gregor} & Utilities for calendar dates (J. Zeppieri) \\%[1em]
-
-\hline
-  \texttt{kcfa} & Functional implementation of 2CFA for the lambda calculus (M. Might) \\%[1em]
-
-\hline
-  \texttt{quadT} & Converter from S-expresion source code to PDF format (M. Butterick)\\%[1em]
-
-\hline
-  \texttt{quadU} & Converter from S-expresion source code to PDF format  (B. Greenman) \\%[1em]
-
-\hline
-  \texttt{snake} & Functional implementation of the  Snake video game (D. Van Horn) \\%[1em]
-
-\hline
-  \texttt{suffixtree} & Algorithm for common longest subsequences between strings. (D. Yoo) \\%[1em]
-
-\hline
-  \texttt{synth} & Converter of description of notes and drum beats to WAV format (V. St-Amour \& N. Toronto) \\%[1em]
-
-\hline
-  \texttt{take5} & Card game simulator (M.Felleisen)  \\%[1em]
-
-\hline
-  \texttt{tetris} & Functional implementation of Tetris (D. Van Horn) \\%[1em]
-
-\end{tabular}
-  \caption{Benchmarks summary.}
-  \label{table:benchmark-descriptions}
-\end{figure*}
-
-Of course, the gradual typing benchmarks have no mistakes for the rational
-programmer to debug. Therefore, we follow
-~\citet{lksfd-popl-2020} and we inject bugs with mutation analysis.
-A mutation is a local syntactic change of the code of a program that may
-produce a bug.
-\ll{Need citations for mutation}
-Usually, the outcome of mutating a program
-once, dubbed a mutant, can be distinguished from the original program with a test case.
-The better the test suite of a program, the more mutants it can ``kill.''
-Mutants that are hard to ``kill'' usually correspond to subtle programming errors, and mutation frameworks typically come with standard mutators carefully designed to create such mutants.
-
-However, Typed Racket's gradual type system detects shallow errors, unlike those introduced by standard mutators.
-Thus, we need different mutators to evaluate the rational programmer, ones which are fine-tuned to generate shallow errors.
-The table in figure~\ref{table:mutation-ops} summarizes our mutators.
-For each one, the table provides a short description and an example
-mutation it performs. Each mutator is inspired by the more-than-a-decade
-long experience of the authors making type-level mistakes in Typed Racket.
-We find that these mistakes often take non-trivial effort to debug.
-
-Most of our mutators in figure~\ref{table:mutation-ops} are self-explanatory.
-The first twelve can apply to most gradually typed languages, including those with classes.
+Most of our mutators are self-explanatory.
+The first twelve apply to most gradually typed languages, including those with classes.
 The last four mutators target distinguishing features of Typed Racket.
@@
 
-\subsection{Are these mutators good enough?}
+\subsection{Are these mutators any good?}
 
-Of course, mutators that introduce these kinds of shallow errors directly into the benchmarks do not produce bugs that are interesting from the perspective of the rational programmer.
-The benchmarks are fully typed, so Typed Racket's type checker locates them immediately.
-For that reason, we apply the mutators to versions of the benchmarks with some components stripped of type annotations (dubbed \emph{configurations}).
-Without the annotations, the type checker may not detect the errors in these mutant configurations.
-Instead, Typed Racket's gradual type system may detect the errors dynamically.
-Mutant configurations that result in dynamic errors are the ideal ground to exercise the rational programmer's ability to locate errors.
+The astute reader must have observed that mutators that cause interesting
+type-level mistakes do not necessarily turn our benchmarks into
+interesting debugging scenarios for the rational programmer.  The
+benchmarks are fully typed, and type checking flags each impedance at
+compile time.  Hence, we apply the mutators to versions, dubbed
+\dubbed{configurations}, of the benchmarks where some components are
+stripped of their type annotation.  Without the annotations, the type
+checker may not detect an impedance and, instead, the type-level checks
+of the gradual type system signal a run time error.  Mutant configurations
+with type-level mistakes that result in run time errors are the ideal
+ground to exercise the rational programmer's ability to locate errors.
 
-Specifically, we define an \emph{interesting debugging scenario} as a mutant configuration that
+Specifically, we define an \emph{interesting debugging scenario} as a
+mutant configuration that
 \begin{enumerate} 
-\item would be rejected by the type checker if we restored the missing type annotations,
-\item produces a dynamic error under Erasure as-is, and
-\item the stacktrace of the error mentions at least three distinct components.
+  \item is rejected by the type checker when we restore any missing type annotations, 
+  \item produces a run time error under Erasure as-is, and 
+  \item the stack trace of the error mentions at least three distinct components.  
 \end{enumerate}
 
-The second and third criteria require a little more explanation.
-The second criterion requires that the error is detectable under Erasure to ensure that it is detectable by all three semantics.
-This is a direct consequence of~\citet{gfd-oopsla-2019}'s theorem that a dynamic type error under Erasure implies a dynamic type error under both Transient and Natural.
-It is worth noting that this criterion excludes from consideration all scenarios where Natural or Transient can detect the bug but Erasure cannot, and thus it favors Erasure.
-However, our goal here is to evaluate blame across these three semantics, and thus we focus on systems where all of them exhibit blame.
+The first criterion simply validates that the debugging scenario concerns
+a type-level mistake. However, The second and third criteria require a
+little more explanation.  The second one stipulates that the mistake is
+detectable under Erasure to ensure that it is detectable by all three
+semantics.  This is a direct consequence of~\citet{gfd-oopsla-2019}'s
+theorem that a run time type error under Erasure implies a run time type
+error under both Transient and Natural.  It is worth noting that this
+criterion excludes from consideration all scenarios where Natural or
+Transient can detect an impedance but Erasure cannot, and thus it favors
+Erasure.  However, our goal here is to test the rational programmer across
+these three semantics, and thus we opt for a setting where all of them
+notice the impedance.
 
-The third criterion aims to capture debugging scenarios with a specific kind of dynamic error: one produced by the interactions between typed and untyped components.
-In those cases, the rational programmer may need to examine several components to locate the source of the bad interaction.
-In technical terms, we conservatively approximate these scenarios by analyzing the stacktrace of the error.
-If the stack trace mentions at least two components, then this is evidence of interaction between components.
-We require the third component because our benchmarks come with a driver component which is present in every stacktrace.
+The third criterion aims to capture debugging scenarios with a specific
+kind of run time error: one produced by the interaction of a buggy
+component with the rest of the program.  In those cases, the rational programmer
+may need to examine several components to locate the source of the faulty interaction.
+In technical terms, we conservatively mark these scenarios by
+analyzing the stack trace of a run time error.  If the stack trace mentions at
+least two distinct components, then this is evidence of interaction between
+components.  We require a third component because our benchmarks come
+with a driver component which is present in the stack trace of all run
+time errors.
 
@@
 
-With this definition in hand, we analyze how effectively our mutators produce interesting debugging scenarios from our benchmarks.
-Figure~\ref{fig:mutant-breakdown} shows the number of distinct interesting debugging scenarios produced by our mutators for each benchmark.
-The columns correspond to our benchmarks, and each rows show the count of interesting debugging scenarios produced by one mutator from figure~\ref{table:mutation-ops}.
-The counts are cut of at 50, so any bar reaching the right side of the plot means 50 or more scenarios.
-
-Figure~\ref{fig:mutant-breakdown} illustrates that our mutators do indeed produce a large number of interesting debugging scenarios for all of our benchmarks.
-All of the mutators together produce at least 40 interesting scenarios for every benchmark, and those scenarios are produced by at least four different mutators for each benchmark.
-Thus our mutators produce a large and diverse population of scenarios for every benchmark.
-Furthermore, every mutator contributes scenarios to at least one benchmark.
-Some mutators clearly only target one or two of them, though, and in particular the class-focused mutators are primarily effective in \texttt{take5}.
-This reflects that among our benchmarks \texttt{take5} makes the most extensive use of object-oriented features, while the other features targeted by the rest of the mutators are used more broadly.
+With this definition in hand, we analyze how effective our mutators
+are to turn the benchmarks into interesting debugging scenarios.
+Indeed the mutators produce at least 72192 interesting debugging scenarios
+across all mutants --- there are potentially many more but this is how
+many we examined. 
+Figure~\ref{fig:mutant-breakdown} breaks down the interesting
+debugging scenarios produced  for each benchmark bu mutator 
+but counts are cut off at 50. The mutators produce at least 40 interesting
+scenarios for every benchmark, and these scenarios originate from at
+least four different mutators per benchmark.  Thus the mutators result in 
+a large and diverse population of scenarios for every benchmark.
+Furthermore, every mutator contributes scenarios to at least one
+benchmark and some mutators clearly only target one or two of them, which
+justifies their inclusion. For
+instance the class-focused mutators are primarily effective in
+which jives with the fact that \texttt{take5} is the benchmark with 
+the most extensive use of object-oriented features, while the other
+features targeted by the rest of the mutators are used more broadly.

### 2020-11-14 writing/related.tex
fix basic latex mistakes in rudimentary way
draft of benchamrks and mutators
adding related work

@@
 
-\begin{itemize}
-  \item  GT: systems we don't represent
-  \item Mutation testing
-  \itme Fault localization
-\end{itemize}
+This paper builds on~\citet{lksfd-popl-2020}'s one-of-a-kind analysis of
+blame for higher-order contracts. They do not spell out the notion of the
+rational programmer but the basic ingredients of our technique derive
+directly from theirs. Specifically, we adapt their empirical technique to
+gradual typing which requires the development of custom
+mutators, careful selection of debugging scenarios and generalization of
+the technique to allow the comparison of different semantics.
+
+Besides the three semantics we consider, the literature on gradual typing
+comes with a plethora of different approaches. For example, Pyret
+(\shorturl{https://www.}{pyret.org}) treats function types same as Natural
+and data types same as Erasure. The Forgetful~\cite{cl-icfp-2017} and the
+Amnesic~\cite{gfd-oopsla-2019} semantics are similar  to Transient but
+use wrappers instead of inlined checks.  Nom~\cite{mt-oopsla-2017} and
+other \emph{concrete\/} semantics~\cite{wnlov-popl-2010, rsfbv-popl-2015,
+mt-oopsla-2017, rzv-ecoop-2015, rat-oopsla-2017} assume that every type
+maps to a unique known tag and they perform type tag checks  for every
+value that crosses from typed to untyped parts of a program or vice versa.
+With the exception of Amnesic~\cite{gfd-oopsla-2019}, all the above
+semantics forego blame. However, Amnesic is a mere
+theoretical construction and is not meant as a blueprint for an
+implementation.

### 2020-11-14 writing/challenges.tex
edits in intor, challenges and mutate

@@
-\section{Why is it hard to evaluate the value of blame?}
+\section{Why is it hard to evaluate blame?}
 \label{sec:challenges}
@@
 impedance mismatch in an incremental fashion. Measuring this simulated
-programmer's behavior on a large number of inputs then yields relevant 
+programmer's behavior on a large number of inputs yields relevant 
 data, similar to collecting empirical data in a human-focused discipline.
@@
 suitable subjects for our experiment.  The mutation analysis of Lazarek et
-al, is, however, completely useless for creating impedance mismatches.
+al, is, however, useless for creating impedance mismatches.
 
@@
 Racket~\cite{gf-icfp-2018} to obtain a transient semantics that is
-comparable with Typed Racket's default natural one. The revised transient
+comparable with Typed Racket's default natural one. This transient
 variant accommodates many more linguistic features and library calls

### 2020-11-14 writing/introduction.tex
edits in intor, challenges and mutate

@@
 code. Then their compilers remove types and rely on the built-in safety checks of
-the underlying programming language to catch any problems. 
+the underlying language to catch any problems. 
 Recently, \citet{cc-snapl-19} have turned this implicitly expressed
@@
  \it
- whether blame assignment adds any value to a gradually typed language, \\
+ whether blame assignment adds any value to a gradually typed language,
  especially for the benefit of the working programmer.
@@
 method for evaluating blame, and given the question's practical
-relevance, this a significant problem.
+relevance, this is a significant problem.
 
@@
 The paper's second contribution is a set of results from applying the evaluation
-method to three distinct blame assignment strategies and approximately 20,000
-cases\mf{Replace with real number} (section~\ref{sec:results}): Vitousek et
+method to three distinct blame assignment strategies and approximately
+65000 cases (section~\ref{sec:results}): Vitousek et
 al.'s strategy of tracking typed/untyped boundaries in Reticulated;

### 2020-11-14 writing/mutate.tex
edits in intor, challenges and mutate

@@
 
-However,  unlike the deep errors introduced by standard mutators, 
-gradual type systems detect shallow errors.  Thus, in order to
-evaluate the rational programmer in the context of gradual typing, 
+However,  unlike the hard-to-kill mistakes introduced by standard mutators, 
+gradual type systems flag mere typos.  Thus, in order to
+exercise the rational programmer in the context of gradual typing, 
 we need to develop alternative mutators
-that are fine-tuned to generate shallow errors.  Our inspiration is our
+that are fine-tuned to produce type-level mistakes gradual type systems 
+can detect. Our inspiration is our
 more-than-a-decade long experience of making type-level mistakes in Typed
@@
 \texttt{Positive-Integer} when all arguments are
-\texttt{Positive-Integer}. However, the result of \texttt{-} is
+positive integers. However, the result of \texttt{-} is
 \texttt{Integer}. In the same spirit, \texttt{boolean} aims to take
 advantage of Typed Racket's ``truthiness.'' Finally, \texttt{negate-cond} and \texttt{force-cond}
-aim to confuse occurrence typing.
+attempt to confuse occurrence typing.
 
@@
 
-The astute reader must have observed that mutators that cause interesting
+The astute reader must have observed that mutators that produce 
 type-level mistakes do not necessarily turn our benchmarks into
 interesting debugging scenarios for the rational programmer.  The
-benchmarks are fully typed, and type checking flags each impedance at
+benchmarks are fully typed, and type checking detects each mistake at
 compile time.  Hence, we apply the mutators to versions, dubbed
-\dubbed{configurations}, of the benchmarks where some components are
+\dubbed{scenarios}, of the benchmarks where some components are
 stripped of their type annotation.  Without the annotations, the type
-checker may not detect an impedance and, instead, the type-level checks
-of the gradual type system signal a run time error.  Mutant configurations
+checker may not detect a mistake at compile time and, instead, the run
+time checks
+of the gradual type system signal an error.  Scenarios
 with type-level mistakes that result in run time errors are the ideal
@@
 Specifically, we define an \emph{interesting debugging scenario} as a
-mutant configuration that
+scenario that
 \begin{enumerate} 
-  \item is rejected by the type checker when we restore any missing type annotations, 
+  \item is rejected by the type checker when we restore all its missing type annotations, 
   \item produces a run time error under Erasure as-is, and 
@@
 The first criterion simply validates that the debugging scenario concerns
-a type-level mistake. However, The second and third criteria require a
-little more explanation.  The second one stipulates that the mistake is
+a type-level mistake. However, The second and third criteria require 
+more explanation.  The second one stipulates that the mistake is
 detectable under Erasure to ensure that it is detectable by all three
 semantics.  This is a direct consequence of~\citet{gfd-oopsla-2019}'s
-theorem that a run time type error under Erasure implies a run time type
+theorem that a run time error under Erasure implies a run time
 error under both Transient and Natural.  It is worth noting that this
-criterion excludes from consideration all scenarios where Natural or
-Transient can detect an impedance but Erasure cannot, and thus it favors
+criterion excludes from consideration the very interesting scenarios where Natural or
+Transient can detect a mistake but Erasure cannot, and thus it favors
 Erasure.  However, our goal here is to test the rational programmer across
 these three semantics, and thus we opt for a setting where all of them
-notice the impedance.
+notice the mistake.
 
@@
 may need to examine several components to locate the source of the faulty interaction.
-In technical terms, we conservatively mark these scenarios by
+In technical terms, we mark these scenarios by
 analyzing the stack trace of a run time error.  If the stack trace mentions at
-least two distinct components, then this is evidence of interaction between
+least two distinct components, then this is accurate evidence of interaction between
 components.  We require a third component because our benchmarks come
@@
 With this definition in hand, we analyze how effective our mutators
-are to turn the benchmarks into interesting debugging scenarios.
+are for turning the benchmarks into interesting debugging scenarios.
 Indeed the mutators produce at least 72192 interesting debugging scenarios
@@
 Figure~\ref{fig:mutant-breakdown} breaks down the interesting
-debugging scenarios produced  for each benchmark bu mutator 
+debugging scenarios produced  for each benchmark by mutator 
 but counts are cut off at 50. The mutators produce at least 40 interesting
@@
 least four different mutators per benchmark.  Thus the mutators result in 
-a large and diverse population of scenarios for every benchmark.
+a sizeable and diverse population of scenarios for every benchmark.
 Furthermore, every mutator contributes scenarios to at least one
-benchmark and some mutators clearly only target one or two of them, which
+benchmark. Especially some mutators  target specific features of
+benchmarks, which
 justifies their inclusion. For
-instance the class-focused mutators are primarily effective in
+instance the class-focused mutators are effective in \texttt{take5}
 which jives with the fact that \texttt{take5} is the benchmark with 
-the most extensive use of object-oriented features, while the other
-features targeted by the rest of the mutators are used more broadly.
+the most extensive use of object-oriented features.

### 2020-11-14 writing/rational.tex
Move debugging scenarios description into previous section
format + prose tweaks
revising related work

--- 2020-07-20 writing/rational.tex
@@
 the one introduced by ~\citet{tfgnvf-popl-2016} for performance
-evaluations. As we discuss briefly at the end of ~\ref{sec:mutate}, 
+evaluations. As we discuss briefly at the end of section ~\ref{sec:mutate}, 
 this space consists of variants of the program, dubbed
-\emph{scenarios}, that differ in their type specifications. In the
+\emph{scenarios}, that differ in their type annotations. In the
 setting of migratory typing,\footnote{The method can be adapted to
@@
 the type checker fails if this component causes the impedance mismatch. Hence, a
-trail that end at an ill-typed scenarios successfully pinpoints the location of
+trail that ensd at an ill-typed scenarios successfully pinpoints the location of
 bugs. 
@@
 \end{quote}
-
-Even if the success of the natural rational programmer is not by pure chance, it
-could well be because the natural semantics perform runtime type checks at the same
 
@@
 log reports on success and failure endings.  The first step is to construct the
-mutant's scenario lattice and determine the scenario that can detect the fault in
-the program. This can happen either because Typed Racket's type checker outright
-rejects the scenario or because natural Typed Racket's runtime performs a type check
-that fails. In either case, each of these scenarios is the root of a trail. Our test
+mutant's scenario lattice and select an interesting debugging scenario
+as we discuss in section~\red{sec:mutate} to be the root of a trail. Our test
 bed extends these trails according to the natural-blame programmer.  If no scenarios
@@
   \centering
-  \includegraphics[width=\textwidth]{./Images/process}
+  \includegraphics[scale=0.36]{./Images/process}
   \caption{The experimental process for one mode of the rational

### 2020-11-14 writing/related.tex
revising related work

@@
 rational programmer but the basic ingredients of our technique derive
-directly from theirs. Specifically, we adapt their empirical technique to
-gradual typing which requires the development of custom
-mutators, careful selection of debugging scenarios and generalization of
-the technique to allow the comparison of different semantics.
+directly from theirs. Specifically, we adapt their empirical technique
+to gradual typing which requires the development of custom mutators,
+careful selection of debugging scenarios and generalization of the
+technique to allow the comparison of different semantics.
 
@@
 (\shorturl{https://www.}{pyret.org}) treats function types same as Natural
-and data types same as Erasure. The Forgetful~\cite{cl-icfp-2017} and the
-Amnesic~\cite{gfd-oopsla-2019} semantics are similar  to Transient but
+and data types same as Erasure. The Forgetful~\cite{cl-icfp-2017} and
+the Amnesic~\cite{gfd-oopsla-2019} semantics are similar  to Transient but
 use wrappers instead of inlined checks.  Nom~\cite{mt-oopsla-2017} and
 other \emph{concrete\/} semantics~\cite{wnlov-popl-2010, rsfbv-popl-2015,
-mt-oopsla-2017, rzv-ecoop-2015, rat-oopsla-2017} assume that every type
-maps to a unique known tag and they perform type tag checks  for every
-value that crosses from typed to untyped parts of a program or vice versa.
-With the exception of Amnesic~\cite{gfd-oopsla-2019}, all the above
-semantics forego blame. However, Amnesic is a mere
+rzv-ecoop-2015, rat-oopsla-2017} assume that every type maps to a unique
+known tag and they perform type tag checks  for every value that crosses
+from typed to untyped parts of a program or vice versa.  Monotonic
+references~\cite{svctg-esop-2015} and the semantics~\cite{tlt-popl-2019,
+etg-icfp-19, tt-scp-20, tgt-popl-18, tt-sas-17} derived with the
+"turn-the-crank-get-a-POPL-paper" Abstracting Gradual Typing
+technique~\cite{gct-popl-2016} are (optimized) variants of Natural for
+different language features.  With the exception of Amnesic and Nom, all
+the above semantics forego blame. However,  Nom's gradual types impose
+significant restrictions on programmers, while Amnesic is a mere
 theoretical construction and is not meant as a blueprint for an

### 2020-11-15 writing/challenges.tex
working on rational

@@
 comparable with Typed Racket's default natural one. This transient
-variant accommodates many more linguistic features and library calls
+variant accommodates varied linguistic features and library calls
 in a way that is faithful to the transient
@@
 Supporting these strategies is {\em not\/} just an issue of language
-implementation but affects greatly the behavior of the rational programmer.
-Specifically, the rational programmer must come with \emph{modes}
-that represent the alternative interpretations of blame. 
-For instance, a programmer using transient may
-pick the oldest element of a blame set as the next component to add types because
+implementation but affects greatly the behavior of the rational
+programmer.  Specifically, the rational programmer must come with
+\emph{modes} that represent the alternative interpretations of blame.  For
+instance, a mode of the rational programmer using transient may pick the
+oldest element of a blame set as the next component to add types because
 it corresponds to the earliest point in the program's evaluation that can
-discover an impedance mismatch.  Another mode does not distinguish between the elements of the
-blame set and has the rational programmer type all blamed components
-at once. Because this second kind of step 
-is more laborious than the first one, the analysis of the results must take 
-into account the rational programmer's effort.
+discover an impedance mismatch.  Another mode may opt to treat the blame
+set as another form of a stack trace and add types to its newest element.
+When both modes are equally successful in locating an impedance mismatch,
+the method has needs to take into account the rational programmer's effort
+to answer which mode is the most effective.
+
+%Another mode does not distinguish between the elements of the
+%blame set and has the rational programmer type all blamed components
+%at once. Because this second kind of step 
+%is more laborious than the first one, the analysis of the results must take 
+%into account the rational programmer's effort.
 

### 2020-11-15 writing/mutate.tex
working on rational

@@
 
-  \texttt{acquire} & Board game simulation (M. Felleisen)  \\%[1em]
+  \texttt{acquire} & Object-oriented board game implementation (M. Felleisen)  \\%[1em]
 
@@
 \hline
-  \texttt{take5} & Card game simulator (M.Felleisen)  \\%[1em]
+  \texttt{take5} & Mixin-based card game simulator (M.Felleisen)  \\%[1em]
 
@@
 we need to develop alternative mutators
-that are fine-tuned to produce type-level mistakes gradual type systems 
+that are fine-tuned to produce the type-level mistakes gradual type systems 
 can detect. Our inspiration is our
@@
 compile time.  Hence, we apply the mutators to versions, dubbed
-\dubbed{scenarios}, of the benchmarks where some components are
-stripped of their type annotation.  Without the annotations, the type
+\dubbed{scenarios}, of the benchmarks without 
+some  of their type annotation.  Without the annotations, the type
 checker may not detect a mistake at compile time and, instead, the run
@@
 The third criterion aims to capture debugging scenarios with a specific
-kind of run time error: one produced by the interaction of a buggy
-component with the rest of the program.  In those cases, the rational programmer
-may need to examine several components to locate the source of the faulty interaction.
+kind of run time error: one produced because of the interaction between the buggy
+portion of the program with the rest of the program.  In those cases, the rational programmer
+may need to examine several pieces of the program to locate the source of the faulty interaction.
 In technical terms, we mark these scenarios by
 analyzing the stack trace of a run time error.  If the stack trace mentions at
-least two distinct components, then this is accurate evidence of interaction between
-components.  We require a third component because our benchmarks come
+least two distinct components, then this is accurate evidence of
+interaction.  We require a third component because our benchmarks come
 with a driver component which is present in the stack trace of all run

### 2020-11-15 writing/rational.tex
working on rational

@@
 
-Based on this idea, a metric for migratory paths is a concise way to
-evaluate the effectiveness of a rational programmer. The components of
-an ill-typed program, such as one of  our ill-typed mutants, 
-create a \emph{debugging scenario space}, a lattice like
-the one introduced by ~\citet{tfgnvf-popl-2016} for performance
-evaluations. As we discuss briefly at the end of section ~\ref{sec:mutate}, 
-this space consists of variants of the program, dubbed
-\emph{scenarios}, that differ in their type annotations. In the
-setting of migratory typing,\footnote{The method can be adapted to
+Based on this idea, migratory paths offer a concise setting for
+evaluating the effectiveness of a rational programmer. If we group the
+scenarios from section ~\ref{sec:mutate} by mutation, then we obtain a set 
+of programs that all contain the same mistake but differ in their type
+annotations.  In the setting of migratory typing,\footnote{The method can be adapted to
 so-called true gradual typing systems~\cite{svcb-snapl-2015} in a
@@
 components are typed. Formally, a program $\system$ is a set of
-components $\set{\component}$, and a scenario $\conf$ is a subset of
-$\set{\component}$ . These scenarios of $\system$ are ordered by the
-subset relation and thus form a lattice $\lattice{\system}$ with
-$2^{\size{\system}}$ elements.  The bottom scenario of
+components $\set{\component}$, and a scenario $\conf$ is the typed subset of
+$\set{\component}$ . Hence, the scenarios of $\system$ are ordered by the
+subset relation and, naturally,  form a lattice $\lattice{\system}$ with
+$2^{\size{\system}}$ elements, like
+the one introduced by ~\citet{tfgnvf-popl-2016} for performance
+evaluations. The bottom scenario of
 $\lattice{\system}$ is always $\emptyset$ and the top one is $\system$
@@
 
-With these definitions, the actions of the rational programmer map to
-an ascending chain---dubbed a \emph{trail}---in $\lattice{\system}$.
-The starting scenario $\conf_0$ of a trail---also referred to as the
-\emph{root} (of the trail)---is the scenario that the programmer
-attempts to debug initially. The easiest scenarios are those that the
-type checker rejects outright, because a type error pinpoints an
-impedance mismatch between the broken code of one component and its
-typed interface to others. At this point, the programmer has
-identified the broken component. If the scenario type-checks, the
-programmers runs the program until it raises an exception.  The
-rational programmer uses the blame assignment\footnote{Blame can
-result either directly from a run time type check or, in the absence
-of blame, from an exception that we track back to the component that
-causes it with the help of the stack trace.} to decide which
-component(s) to equip with types. This choice constructs scenario
-$\conf_{i+1}$ from $\conf_{i}$ and implies the creation of an
-ascending chain.  Intuitively, with each step along the chain, the
-rational programmer reduces the current scenario to an ``easier'' one
-to debug.
+With these definitions, the actions of the rational programmer map to an
+ascending chain---dubbed a \emph{trail}---in $\lattice{\system}$.  The
+starting scenario $\conf_0$ of a trail---also referred to as the
+\emph{root} (of the trail)---is the initial debugging scenario.  Herein we
+consider only roots that are interesting debugging scenarios as described
+in~\ref{sec:mutate} --- we return to this point further on.  Intuitively,
+starting from such a root, the rational programmer  steps along a trail,
+and at each step, uses reduces the current scenario to an ``easier'' next
+scenario to debug.  The easiest scenarios are those that the type checker
+rejects outright, because a type error pinpoints an impedance mismatch
+between the broken code of one component and its typed interface to
+others. At this point, the programmer has identified the broken component.
+If the scenario type-checks, the programmers runs the program until it
+raises an exception.  The rational programmer uses the blame
+assignment\footnote{Blame can result either directly from a run time type
+check or, in the absence of blame, from an exception that we track back to
+the component that causes it with the help of the stack trace.} to decide
+which component(s) to equip with types. This choice constructs scenario
+$\conf_{i+1}$ from $\conf_{i}$ and implies the creation of trail.
 
@@
 subsection extends the development with a discussion of how we can
-measure the rational programmer's effort along the trail.
+compare the effectiveness of the rational programmer across semantics.
 
@@
 the type checker fails if this component causes the impedance mismatch. Hence, a
-trail that ensd at an ill-typed scenarios successfully pinpoints the location of
+trail that ends at an ill-typed scenarios successfully pinpoints the location of
 bugs. 
@@
 \it
-Given a program $\system$, natural blame is\\
+Given a program $\system$, the natural blame mode is\\
 (1) \emph{always useful} iff $\lattice{\system}$ contains no failing natural blame trails.\\
@@
 blame is a critical piece of the rational programmer's process.  For instance,
-typing a random component might be as useful as typing the blamed one.
+typing the top of stack trace of a failed run time type check might be as useful as typing the blamed one.
 
 To account for this situation, we devise a new mode of the natural rational
-programmer that follows a random migration process:
+programmer that follows a migration process based on exceptions:
 \begin{quote}
-\it A natural random trail is a sequence of scenarios $\conf_0,...\conf_n$ of a
+\it A natural exception trail is a sequence of scenarios $\conf_0,...\conf_n$ of a
 program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
 \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i = \{\random{\system, \conf_i}\}$
-where $\random{\system, \conf}$ denotes a randomly selected component of $\system$
-that is not in $\conf_i$.
+where $\exception{\system, \conf}$ denotes the location of the runtime
+  error of $\conf$.
 \end{quote}
-Similar to natural blame we determine how successful the random
-migration is based on the success of natural random trails: 
+Similar to natural blame we determine how successful the exception
+migration is based on the success of natural exception trails: 
 \begin{quote}
 \it
-Given a program $\system$, natural random blame is\\
-(1) \emph{always useful} iff $\lattice{\system}$ contains no failing natural random blame trails.\\
-(2) \emph{sometimes useful} iff $\lattice{\system}$ contains successful natural random blame trails.\\
+Given a program $\system$, the natural exception mode is\\
+(1) \emph{always useful} iff $\lattice{\system}$ contains no failing
+  natural exception trails.\\
+(2) \emph{sometimes useful} iff $\lattice{\system}$ contains successful
+  natural exception trails.\\
 (3) \emph{never useful} iff $\lattice{\system}$ contains no successful natural
-random blame trails.\\
+exception trails.\\
 \end{quote}
@@
 Putting the pieces together, we can now define that blame assignments adds value in
-the context of a natural semantics if blame-based migration is better than random
-migration.
+the context of a natural semantics if blame mode  is better than the
+exception.
 \begin{quote}
-\it Natural blame adds value to natural for a program $\system$ over pure chance iff
-  natural blame is more successful than natural random. And, natural blame is more
-  successful than natural random for $\system$ if either natural blame is always
-  successful for $\system$ when natural random is sometimes or never successful, or
-  natural blame is sometimes successful for $\system$ when natural random is never
-  successful.
+\it Natural blame adds value to natural for a program $\system$ over
+  exceptions iff the natural blame mode is more successful than the
+  natural exception mode. And, the natural blame mode is more successful
+  than the natural exception mode for $\system$ if either natural blame is
+  always successful for $\system$ when natural exception is sometimes or
+  never successful, or natural blame is sometimes successful for $\system$
+  when natural exception is never successful.
 \end{quote}
-
-%% -----------------------------------------------------------------------------
-\subsection{Modes of the Erasure Rational Programmer} \label{sub:erasure}
-
- Since gradually typed languages with erasure semantics do not come with
- blame assignment, a rational programmer can only hope that the underlying
- safety checks and their exceptions are helpful.  Thus, adapting the
- definitions related to exception and random trails from natural to
- erasure, we characterize when the effect of erasure exceptions in debugging a
- program $\system$ goes beyond that of pure chance.
-
-\begin{quote}
-\it
-  Erasure exceptions add value to erasure for a program $\system$ over pure chance
-  iff erasure exception is more successful than erasure random.
-\end{quote}
-\noindent
 
@@
 should react when the language produces a blame set. Our answer is that the rational
-programmer has at least two reasonable options.  The first option is to type all
-blamed components at once. The second one is to select the component that is added
+programmer has at least two reasonable options. The first one is to select the component that is added
 to the blame set first and assign types to only that one---after all the types of
 this first component should be able to detect an impedance mismatch earlier in the
-evaluation of a program than the later ones.
+evaluation of a program than the later ones. The second option is to
+select the component that is added to the blame set last, effectively
+interpreting the blame set as an enhanced stack trace.
 
@@
 \begin{quote}
-\it A \emph{ transient-all blame trail} is a sequence of scenarios
-$\conf_0,...\conf_n$ of a program $\system$ where for all $0 \leq i \leq n - 1$,
-$\conf_i \subset \conf_{i+1}$ and ~$\conf_{i+1} \setminus \conf_i = \mblame{\system,
-\conf_i}$ where $\mblame{\system, \conf}$ denotes the set of components that $\conf$
-blames under transient semantics.
-\end{quote}
-\noindent
-and
-\begin{quote}
-\it A \emph{transient-firstborn blame trail} is a sequence of scenarios
-$\conf_0,...\conf_n$ of a program $\system$ where for all $0 \leq i \leq n - 1$,
+\it A \emph{transient-first blame trail} is a sequence of scenarios
+$\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
 $\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
@@
 \end{quote}
-The definitions of \emph{transient erasure trails} and \emph{transient random
-trails} are analogous to those for natural erasure trails and natural random trails,
-respectively.
 
-We also use these four forms of blame trails to define when transient-all and
-transient-firstborn add value to transient over exceptions and pure chance.
+and
+
 \begin{quote}
-\it 
-Transient all blame adds value to transient for a program $\system$ over
-  pure chance and transient exceptions iff transient all blame 
-  is more successful than both transient random and transient exception.
+\it A \emph{transient-last blame trail} is a sequence of scenarios
+$\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
+$\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
+  \first{\mblame{\system, \conf_i}}$ where $\last{\mblame{\system, \conf}}$ is the
+last component transient adds to the blame set for $\conf$.
+\end{quote}
 
-Transient firstborn blame adds value to transient for a program $\system$ over
-  pure chance and transient exceptions iff transient all blame 
-  is more successful than both transient random and transient exception.
+The definition of \emph{transient exception trails} 
+is analogous to that for natural and, same as for natural, we use it to  
+specify when the transient-last and transient-first blame modes add
+value to transient for a program.
 
-\end{quote}
+%% -----------------------------------------------------------------------------
+\subsection{Modes of the Erasure Rational Programmer} \label{sub:erasure}
+
+ Since gradually typed languages with erasure semantics do not come with
+ blame assignment, a rational programmer can only hope that the underlying
+ safety checks and their exceptions are helpful.  Thus, the erasure 
+ rational programmer has a single mode; the erasure exception mode.
 
@@
 mutants from section~\ref{sec:mutate}. In particular, our experimental
-process collects data to answer four questions for each mutant:
+process collects data to answer three questions for each mutant:
 \begin{itemize}
@@
 
-\item[$Q_2$] Is all blame valuable for transient?
+\item[$Q_2$] Is first blame valuable for transient?
 
-\item[$Q_3$] Is firstborn blame valuable for transient?
+\item[$Q_3$] Is last blame valuable for transient?
 
-\item[$Q_*$] Are plain exceptions valuable for
-  erasure? For the other two systems? 
 \end{itemize}
@@
 question $Q_1$ asks whether blame is valuable for natural and our experiment
-uses the natural blame, exception and random trails to answer it.
+uses the natural blame and exception trails to answer it.
 
@@
 \center
-{\begin{tabular}{l|c|c|c}
+{\begin{tabular}{l|c|c}
 %% ---------------------------------------------------------------------------------------------------------------
-                        & {\bf Natural}        & {\bf Transient}          & {\bf Erasure} \\ \hline 
+                        & {\bf Natural}        & {\bf Transient}         \\ \hline 
 %% ---------------------------------------------------------------------------------------------------------------
-{\bf Blame}             &       $Q_1$          &                          &               \\
-{\bf All blame}         &                      &     $Q_2$                &               \\
-{\bf Firstborn blame}   &                      &     $Q_3$                &               \\
-{\bf Exception}         &       $Q_1$/$Q_*$    &     $Q_2$/$Q_3$/$Q_*$    &      $Q_*$    \\
-{\bf Random}            &       $Q_1$/$Q_*$    &     $Q_2$/$Q_3$/$Q_*$    &      $Q_*$    \\
+{\bf Blame}             &       $Q_1$          &                           \\
+{\bf First blame}         &                      &     $Q_2$                \\
+{\bf Last blame}   &                      &     $      Q_3$                  \\
+{\bf Exceptions}         &       $Q_1$    &     $Q_2$/$Q_3$    \\
 \end{tabular}}
@@
 In detail we answer $Q_1$ for an ill-typed mutant from
-section~\ref{sec:mutate} by creating trails for each natural blame,
-natural exception and natural random trail in the scenario lattice of the mutant. A
+section~\ref{sec:mutate} by analyzing the natural blame and
+natural exception trails in the scenario lattice of the mutant. A
 log reports on success and failure endings.  The first step is to construct the
 mutant's scenario lattice and select an interesting debugging scenario
-as we discuss in section~\red{sec:mutate} to be the root of a trail. Our test
+as  the root of a trail.
+Our test
 bed extends these trails according to the natural-blame programmer.  If no scenarios
@@
 trail is successful, otherwise it is failing. We repeat the same process for the
-roots two more times; once following the natural-exception programmer and once the
-natural-random programmer. Finally, based on the collected data, we calculate the
+same roots once more following the natural-exception programmer
+. Finally, based on the collected data, we calculate the
 success/failure results for the trails to determine whether natural blame is more
-successful than natural exceptions and pure chance and we obtain the answer for
+successful than natural exceptions  and we obtain the answer for
 $Q_1$.  The process is analogous for the other questions employing the respective
@@
 experimental process for one mode of the rational programmer and connects
-it with the mutations from section~\ref{sec:mutation}.
+it with the mutations from section~\ref{sec:mutate}.
 
@@
 
-\subsection{Programmer Effort}
+\subsection{Comparing Modes through Programmer Effort}
 
@@
 rational programmers (all and firstborn) can compete to see which one
-reaches an ill-typed scenarion with less effort. In
+reaches an ill-typed scenario with less effort. In
 general, the effort of trails of different modes that have the same
@@
 modes even if the modes correspond to different gradual typing systems.
-Unfortunately, statistically speaking, such comparisons between two modes do not generalize 
-beyond the successful trails with common roots in our experiments.
-Still they allow us to quantify the difference between the modes of the rational
-programmers and thus the different blame assignment strategies  for the specific
-debugging scenarios we analyze in our experiment.
+Since in our experiment all roots are interesting debugging scenarios and 
+those are independent of mode or semantics, effort of trails is sufficient 
+to compare the relative value of one mode versus another.

### 2020-11-16 writing/related.tex
related: describe pyret correctly
related: para 2, move 'blame' to thesis sentence

--- 2020-07-01 writing/related.tex
@@
 
-\begin{itemize}
-  \item  GT: systems we don't represent
-  \item Mutation testing
-  \itme Fault localization
-\end{itemize}
+This paper builds on~\citet{lksfd-popl-2020}'s one-of-a-kind analysis of
+blame for higher-order contracts. They do not spell out the notion of the
+rational programmer but the basic ingredients of our technique derive
+directly from theirs. Specifically, we adapt their empirical technique
+to gradual typing which requires the development of custom mutators,
+careful selection of debugging scenarios and generalization of the
+technique to allow the comparison of different semantics.
+
+The literature on gradual typing presents many semantics beyond the three
+that we study, but only two additional blame strategies.
+Pyret (\shorturl{https://www.}{pyret.org}) treats fixed-size data types same as Natural
+and functions same as Transient. The Forgetful~\cite{cl-icfp-2017} and
+the Amnesic~\cite{gfd-oopsla-2019} semantics are similar to Transient but
+use wrappers instead of inlined checks.  Nom~\cite{mt-oopsla-2017} and
+other \emph{concrete\/} semantics~\cite{wnlov-popl-2010, rsfbv-popl-2015,
+rzv-ecoop-2015, rat-oopsla-2017} assume that every value comes with a type tag
+and use tag checks to supervise the interactions between typed and untyped code.
+Monotonic
+references~\cite{svctg-esop-2015} and the semantics~\cite{tlt-popl-2019,
+etg-icfp-19, tt-scp-20, tgt-popl-18, tt-sas-17} derived with the
+Abstracting Gradual Typing
+technique~\cite{gct-popl-2016} are (optimized) variants of Natural.
+Among these representative semantics, only Amnesic and Nom present innovative
+blame strategies.
+Our study nevertheless excludes these strategies because Amnesic is a mere
+theoretical construction and Nom imposes restrictions on untyped
+code that would necessitate new data definitions across our benchmarks.

### 2020-11-17 writing/mutate.tex
adding  better explanation of interesting scenarios

@@
 The first criterion simply validates that the debugging scenario concerns
-a type-level mistake. However, The second and third criteria require 
+an actual type-level mistake. However, The second and third criteria require 
 more explanation.  The second one stipulates that the mistake is
 detectable under Erasure to ensure that it is detectable by all three
-semantics.  This is a direct consequence of~\citet{gfd-oopsla-2019}'s
-theorem that a run time error under Erasure implies a run time
-error under both Transient and Natural.  It is worth noting that this
-criterion excludes from consideration the very interesting scenarios where Natural or
-Transient can detect a mistake but Erasure cannot, and thus it favors
-Erasure.  However, our goal here is to test the rational programmer across
-these three semantics, and thus we opt for a setting where all of them
-notice the mistake.
+semantics.   Whether a scenario produces a run time error largely depends on 
+the underlying semantics but \citet{gfd-oopsla-2019}
+show that a run time error under Erasure implies a run time
+error under both Transient and Natural. Thus the criterion aligns
+interesting debugging scenarios with our lowest denominator, Erasure, 
+and excludes  from consideration the very interesting scenarios indeed where Natural or
+Transient can detect a mistake but where Erasure is powerless. Of course this choice favors
+Erasure over Transient and Natural and, for the same reason, Transient over
+Natural. We consider some form of bias towards one or the other semantics
+unavoidable no matter how we define interesting debugging scenarios and we
+opt for tipping the scales in favor of the theoretically weakest
+semantics.
 
@@
 kind of run time error: one produced because of the interaction between the buggy
-portion of the program with the rest of the program.  In those cases, the rational programmer
+portion of the program with the rest of the program.  In these cases, the rational programmer
 may need to examine several pieces of the program to locate the source of the faulty interaction.
-In technical terms, we mark these scenarios by
+In technical terms, we reliably mark these scenarios by
 analyzing the stack trace of a run time error.  If the stack trace mentions at
-least two distinct components, then this is accurate evidence of
-interaction.  We require a third component because our benchmarks come
+least two distinct components, then this indicates interaction.  We require a third component because our benchmarks come
 with a driver component which is present in the stack trace of all run
-time errors.
+time errors. In essence, this last criterion excludes a large number of
+trivial to debug, and thus ineffective, scenarios that error immediately when their mutated component is
+evaluated.
 
@@
 
-With this definition in hand, we analyze how effective our mutators
+With the definition of interesting debugging scenarios in hand, we analyze how effective our mutators
 are for turning the benchmarks into interesting debugging scenarios.

### 2020-11-18 writing/challenges.tex
revision of rational

@@
 \section{Why is it hard to evaluate blame?}
-\label{sec:challenges}
 
@@
 
-\item Generate a large set of programs with representative
-  type errors that all three systems can detect (section~\ref{sec:mutate});
+\item Generate a large set of programs with interesting
+  type-level errors (section~\ref{sec:mutate});
 
 \item Simulate the rational programmer on these programs with sufficient 
-  modes to (i) match the peculiarities of each  system and 
-    (ii) isolate blame from other confounding factors
+  modes to make meaningful comparisons between modes of the same or
+    different systems
     (section~\ref{sec:rational});
@@
 \item Sample the experimental space  to obtain a computationally feasible
-  but statistically meaningful experiment (section~\ref{sec:sample}).
+  but statistically sound experiment (section~\ref{sec:sample}).
 

### 2020-11-18 writing/rational.tex
revision of rational

@@
 Based on this idea, migratory paths offer a concise setting for
-evaluating the effectiveness of a rational programmer. If we group the
-scenarios from section ~\ref{sec:mutate} by mutation, then we obtain a set 
+evaluating the effectiveness of a rational programmer for dealing with the
+interesting debugging scenarios  from section ~\ref{sec:mutate}.
+If we group debugging scenarios by mutation, then we obtain a set 
 of programs that all contain the same mistake but differ in their type
@@
 consider only roots that are interesting debugging scenarios as described
-in~\ref{sec:mutate} --- we return to this point further on.  Intuitively,
+in~\ref{sec:mutate} --- we expand on this point further on.  Intuitively,
 starting from such a root, the rational programmer  steps along a trail,
-and at each step, uses reduces the current scenario to an ``easier'' next
+and at each step, reduces the current scenario to an ``easier'' next
 scenario to debug.  The easiest scenarios are those that the type checker
@@
 If the scenario type-checks, the programmers runs the program until it
-raises an exception.  The rational programmer uses the blame
+raises a run time error.  The rational programmer uses the blame
 assignment\footnote{Blame can result either directly from a run time type
@@
 first page and to explain our experimental process. The final
-subsection extends the development with a discussion of how we can
-compare the effectiveness of the rational programmer across semantics.
+subsection extends the development with a discussion of programmer effort.
 
@@
 blame is a critical piece of the rational programmer's process.  For instance,
-typing the top of stack trace of a failed run time type check might be as useful as typing the blamed one.
+typing the top of stack trace of a failed run time type check, dubbed the
+location of the exception, might be as useful as typing the blamed one.
 
@@
 \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i = \{\random{\system, \conf_i}\}$
-where $\exception{\system, \conf}$ denotes the location of the runtime
-  error of $\conf$.
+where $\exception{\system, \conf}$ denotes the location of the exception of $\conf$.
 \end{quote}
@@
 
-Putting the pieces together, we can now define that blame assignments adds value in
-the context of a natural semantics if blame mode  is better than the
-exception.
+Putting the pieces together, we can now define when blame 
+is more useful than mere exceptions in the context of a natural semantics.
 \begin{quote}
-\it Natural blame adds value to natural for a program $\system$ over
-  exceptions iff the natural blame mode is more successful than the
-  natural exception mode. And, the natural blame mode is more successful
-  than the natural exception mode for $\system$ if either natural blame is
-  always successful for $\system$ when natural exception is sometimes or
-  never successful, or natural blame is sometimes successful for $\system$
-  when natural exception is never successful.
+\it Natural blame  is more useful
+  than natural exceptions for $\system$ if either natural blame is
+  always useful for $\system$ when natural exception is sometimes or
+  never useful, or natural blame is sometimes useful for $\system$
+  when natural exception is never useful.
 \end{quote}
@@
 
-The transient semantics assigns blame to a series of components. In this setting, a
+The transient semantics assigns blame to a set of components. In this setting, a
 blame assignment has the following meaning: the witness value to the impedance
-mismatch crossed the boundaries between neighbors in the sequence, and each crossing
-checked a portion of the value's type shape.
+mismatch may have crossed the boundaries between elements in the set, and each crossing
+checked only a portion of the value's type shape.
 
@@
 programmer has at least two reasonable options. The first one is to select the component that is added
-to the blame set first and assign types to only that one---after all the types of
+to the blame set first and assign types to only that one---after all, if
+fully checked, the types of
 this first component should be able to detect an impedance mismatch earlier in the
@@
 select the component that is added to the blame set last, effectively
-interpreting the blame set as an enhanced stack trace.
+interpreting the blame set as a boundary-aware stack.
 
@@
 is analogous to that for natural and, same as for natural, we use it to  
-specify when the transient-last and transient-first blame modes add
-value to transient for a program.
+specify when the transient-last and transient-first blame modes are more
+useful than exceptions in the context of a natural semantics.
 
@@
  safety checks and their exceptions are helpful.  Thus, the erasure 
- rational programmer has a single mode; the erasure exception mode.
+ rational programmer has a single mode, the erasure exception mode, and
+ its definition and the accompanying notion of usefulness follow that for
+ natural.
 
@@
 of blame for natural, transient and erasure Typed Racket in the setting of the 
-mutants from section~\ref{sec:mutate}. In particular, our experimental
-process collects data to answer three questions for each mutant:
+mutants from section~\ref{sec:mutate}. In line with the discussion so far, 
+our experimental process collects data to answer three initial questions for each mutant:
 \begin{itemize}
@@
 
+Furthermore, our experiment can compare the relative value of blame
+between any two of the three semantics. 
+\begin{itemize}
+\item[$Q_*$] Is blame for X more valuable than blame for Y where X, Y is natural,
+  erasure or natural?
+\end{itemize}
+
 Table~\ref{fig:experiment-outline} summarizes how each question relates to
@@
 \center
-{\begin{tabular}{l|c|c}
+{\begin{tabular}{l|c|c|c}
 %% ---------------------------------------------------------------------------------------------------------------
-                        & {\bf Natural}        & {\bf Transient}         \\ \hline 
+                        & {\bf Natural}  & {\bf Transient} &  {\bf Erasure} \\ \hline 
 %% ---------------------------------------------------------------------------------------------------------------
-{\bf Blame}             &       $Q_1$          &                           \\
-{\bf First blame}         &                      &     $Q_2$                \\
-{\bf Last blame}   &                      &     $      Q_3$                  \\
-{\bf Exceptions}         &       $Q_1$    &     $Q_2$/$Q_3$    \\
+{\bf Blame}             &  $Q_1/Q_*$    &                  &                \\
+{\bf First blame}       &               &     $Q_2/Q_*$    &                 \\
+{\bf Last blame}        &               &     $Q_3/Q_*$    &                 \\
+{\bf Exceptions}        &      $Q_1$    &     $Q_2/Q_3$    &      $Q_*$      \\
 \end{tabular}}
@@
 log reports on success and failure endings.  The first step is to construct the
-mutant's scenario lattice and select an interesting debugging scenario
-as  the root of a trail.
+mutant's scenario lattice and select one of the interesting debugging
+scenarios from section~\ref{sec:mutate}.
 Our test
 bed extends these trails according to the natural-blame programmer.  If no scenarios
-can be added to the trail---a trail may the top of the finite lattice---we check if
+can be added to the trail, we check if
 the last scenario of the trail type checks or not. If it does, this natural-blame
 trail is successful, otherwise it is failing. We repeat the same process for the
-same roots once more following the natural-exception programmer
-. Finally, based on the collected data, we calculate the
-success/failure results for the trails to determine whether natural blame is more
-successful than natural exceptions  and we obtain the answer for
-$Q_1$.  The process is analogous for the other questions employing the respective
-modes of the rational programmer. Figure~\ref{fig:process} summarizes the
+same roots once more following the natural-exception programmer. 
+Next, based on the collected data, we calculate the
+success/failure results for the trails to determine whether blame
+is more useful than exceptions for this mutant. Finally, we
+obtain a positive answer for
+$Q_1$ when the percentage of mutants where blame is more useful
+than exceptions is greater than the percentage of mutants where exceptions
+are more useful
+than blame.  The process is analogous for $Q_2$ and $Q_3$  employing the respective
+modes of the rational programmer. For $Q_*$, we also proceed in a similar
+manner except that we use modes 
+of the rational programmer from different semantics. For instance to
+decide whether blame for natural is more valuable than blame for
+transient, we compare the usefulness of
+the natural blame mode with that of the transient first and transient last
+modes for all our mutants. 
+Figure~\ref{fig:process} summarizes the
 experimental process for one mode of the rational programmer and connects
@@
 
-\subsection{Comparing Modes through Programmer Effort}
+\subsection{Programmer Effort}
 
@@
 our experiment  records the number of components a rational programmer 
-has to type along each successful trail ($\lvert \conf_n \setminus \conf_0
+has to type along each trail ($\lvert \conf_n \setminus \conf_0
 \rvert$). We use this number as the metric for the effort 
-of the rational programmer to debug the root scenario of each successful trail.
+of the rational programmer to debug an interesting scenario.
 
-Programmer effort can help shed some light to the relative effectiveness of the
-three gradual typing systems. For example, consider a debugging scenario from  the
-lattice of one of our mutants.  If the
-scenario results in blame both under, for instance, the natural and transient semantics,
-then the natural blame rational programmer and the two transient blame
-rational programmers (all and firstborn) can compete to see which one
-reaches an ill-typed scenario with less effort. In
-general, the effort of trails of different modes that have the same
-root is a point of comparison for the effectiveness of these
-modes even if the modes correspond to different gradual typing systems.
-Since in our experiment all roots are interesting debugging scenarios and 
-those are independent of mode or semantics, effort of trails is sufficient 
-to compare the relative value of one mode versus another.
+A comparison of the effort of different modes of the rational programmer
+can help shed some further light to the effectiveness of the
+three gradual typing systems. For example, consider one of 
+our mutants.  If, for instance, both natural blame and transient first
+blame are equally useful for the mutant, the two nodes of the 
+rational programmer can compete to see which one
+debugs the interesting scenarios of the mutant with less effort. In
+general, if the effort distribution of a mode of the rational programmer for a
+mutant has a shorter tail and more
+volume around smaller values compared to the effort distribution of another
+mode, then the first mode is the more effective of the two.
+
+Finally, effort allows us to validate that the observed effectiveness 
+of the rational programmer is not an artifact of pure chance.
+To test the null hypothesis for the effectiveness of the rational
+programmer, we can simply compare the effort distribution of a mode of the
+rational programmer with a rational programmer that ignores blame and
+exceptions and instead selects which component to type next randomly. 
+Of course, since each mutant has a finite number of components, 
+such a random mode is always useful for all mutants in the context of all
+three semantics. At the same time though,
+the natural, transient and erasure random mode exhibit a normal effort distribution for each mutant
+that should be distinct from the effort distribution of all other modes of
+the rational programmer if the effectiveness of the latter is not
+coincidental.

### 2020-11-18 writing/discussion.tex
discussion: why does transient blame 0 boundaries

@@
 
-An apparent flaw in our transient is the high cost of blame.
+Surprisingly, Transient blames an empty set of boundaries in XXXX cases.
+Such cases should never occur, at least in theory, because an empty set
+ means the value has never crossed a boundary---if the value is indeed defined
+ in the current module, then we have typed code blaming itself for a typed
+ value; thus, the type checker must be unsound.
+After careful investigation of these empty blame cases, we found no soundness
+ bugs in Typed Racket.
+Instead, we found scenarios in which Transient lost track of the proper
+ boundaries.
+These scenarios suggest ways to improve the blame algorithm from~\citet{vss-popl-2017}:
+\begin{itemize}
+  \item
+    Entries in a blame map must point to several parent entries.
+    For example, if the function \texttt{f} receives bad input in the call
+    \texttt{(filter f xs)}, then blame should point to both \texttt{filter}
+    and the \texttt{xs} list.
+  \item
+    The construction of blame map entries must be guided by type-like specifications
+     instead of relying on syntax.
+    Aliasing the built-in \texttt{filter} function should not change the shape
+     of the blame map.
+  \item
+    The initial blame map must reflect the initial type environment.
+    Typed Racket trusts that untyped core-library functions, such as \texttt{filter},
+     behave correctly, but does not list the assumption in the blame map.
+\end{itemize}
+\noindent{}Despite these known limitations, we conjecture that these improvements
+ to Transient blame-tracking will not affect our overall conclusions.
+
+A second apparent flaw in our Transient semantics is the high cost of blame.
 Whereas \citet{vss-popl-2017} report an average slowdown of 2.5x and

### 2020-11-19 writing/sample.tex
revision of rational
Draft of sampling section

--- 2020-07-06 writing/sample.tex
@@
-\section{How to Sample the Experimental Space} \label{sec:sample}
+\section{How to Sample the Experimental Space}
+
+The experiment as-outlined has all of the parts necessary to find information, but our mutations generate a lot of interesting debugging scenarios.
+Specifically, they generate 96,515,744 scenarios; far too many to try them all.
+Thus, we sample from the total set of scenarios we generate to obtain a computationally-feasible subset for exploration.
+At a high level, we perform a stratified uniform random sample with two levels of stratification to reduce the total number of scenarios we explore down to 728,000 scenarios.
+This section describes this sampling in detail and how it affects our results.
+
+The first natural place to perform sampling is in the sets of scenarios corresponding to each mutant, because those sets form a coherent and finite population.
+% We seek to understand the usefulness of blame for all of the scenarios of a given mutant, but through sampling 
+To understand usefulness, we formulate a binary question capturing whether or not the trail of a given scenario indicates that blame is useful: i.e. if it terminates successfully or not.
+Then, we can sample from the set of all possible scenarios for a given mutant to get an estimate of whether or not blame is useful for the entire set, i.e.
+if all trails for that mutant terminate successfully or not.
+Thus we can obtain a ``yes or no'' answer for the usefulness of blame in debugging a given mutant without having to explore all of its scenarios, up to some level of confidence based on the number of samples.
+Because these answers reflect whether blame is useful for all trails for a given mutant, we dub them ``always'' and ``never''.
+In fact, this formulation offers an additional implicit answer whenever the answer is neither ``always'' nor ``never'';
+in such cases, some of our samples answer ``yes'' and some answer ``no'', hence we dub this third answer ``sometimes''.
+We randomly select enough samples to obtain a confidence of 95\% in the answer obtained for a given mutant (with a margin of error of 5\%).
+We refer the interested reader to the accompanying appendix for the statistical calculations we used to determine the necessary sample size.
+
+By sampling within the population of possible scenarios for every mutant, we can explore a significantly smaller number of scenarios in exchange for only being able to understand blame's usefulness at the granularity of mutants, and only in terms of the three ``always'', ``sometimes'', and ``never'' answers.
+Even so, our mutators generate 16,800 mutants with interesting scenarios; still far too many to explore.
+Thus the set of mutants forms another natural place to perform sampling again.
+Unlike in the context of mutant scenarios, however, our set of mutants is not really a meaningful population because we don't know how they relate to the set of all possible mutants.
+So this time we simply select a sample size according to convenience.
+We sample five mutants per mutator, aiming to ensure that our sample reflects the diversity of the mutants (with respect to mutators) of the original set of all mutators we generate for a given benchmark.
+We have sixteen mutators, so selecting five mutants per mutator gives us a sample of 80 mutants per benchmark.
+The only wrinkle in this stage of sampling is that every benchmark has mutants produced by different mutators, and in particular no benchmark has five mutants from all of them.
+Therefore we cannot select five mutants per mutator for every benchmark; instead, we select the 80 mutants evenly-distributed across all of the mutators available for a benchmark.
+
+%% we use the following algorithm to select the 80 mutants evenly distributed across all of the mutators that do produce mutants of a benchmark:
+%% 1. Treat each mutator as a bucket containing the mutants produced by that mutator
+%% 2. Let the target-sample-size be 80
+%% 3. Let the bucket-sample-size be target-sample-size/number-of-buckets
+%% 4. For each bucket, do
+%%    1. Randomly remove min(bucket-sample-size, mutants-in(bucket)) mutants from the bucket
+%%    2. Set target-sample-size = target-sample-size - (the number of mutants removed)
+%% 5. If target-sample-size is empty, done; else, go back to 3.

### 2020-11-19 writing/results.tex
results: add usefulness table

\section{What We Discovered}

\begin{figure*}
  \centering
  \includegraphics[width=0.7\textwidth]{./plots/usefulness-table}
  \caption{Usefulness breakdown of all mutants}
  \label{fig:usefulness-table}
\end{figure*}

### 2020-11-19 writing/introduction.tex
revision of rational
adding landscape description

@@
-\section{Does Blame Matter?} \label{sec:introduction}
+\section{Does Blame Matter?}
 
@@
 method to three distinct blame assignment strategies and approximately
-65000 cases (section~\ref{sec:results}): Vitousek et
+72000 cases (section~\ref{sec:results}): Vitousek et
 al.'s strategy of tracking typed/untyped boundaries in Reticulated;

### 2020-11-19 writing/landscape.tex
revision of rational
adding landscape description

--- 2020-07-01 writing/landscape.tex
@@
-\section{The Landscape of Gradual Typing and Blame}
+\section{How to Bring Natural, Transient and Erasure Under the Same Roof}
 
-   By example; adjusted version of the one in OOPSLA sec 2
+%% -----------------------------------------------------------------------------
+\def\fname#1#2{$\underline{\hbox to #2in{#1}}$\\[-2ex]}
 
-  \begin{itemize}
+\begin{figure*}[hbt]\footnotesize
+  \vspace{-1mm}
 
-  \item Why Blame Assignments Differ.
+\begin{minipage}[t]{2.2in}
+\begin{alltt}\fname{untyped/pack-lib}{1.8}
+(provide pack _ _ _)
 
-  \item How Blame Assignments Differ.
+_ _ _  dependencies     _ _ _
+_ _ _  and definitions  _ _ _
 
-  \item A space of GT system design. In terms of error reporting, we identify two axes of GT system
-        design: where checking occurs and whether blame is tracked.
-    \end{itemize}
+(define (pack d)
+  ;; process JSON data and 
+  ;; package it in a dictionary
+  ;; (association list)
+  (make-hash _ _ _) ;; BUG!)
+
+\end{alltt}
+\end{minipage}\begin{minipage}[t]{2.2in}
+\begin{alltt}\fname{typed/pack-lib}{1.7}
+(provide typed-pack)
+
+(require/typed pack-lib
+  [pack 
+  (-> JSON 
+      (Listof 
+       (Pairof Symbol String)))])
+
+(define typed-pack pack) 
+\end{alltt}
+\smallskip
+\begin{alltt}\fname{typed/crypto-pack-lib}{1.7}
+(provide crypto-pack)
+
+(require/typed pack-lib
+ [pack 
+  (-> JSON 
+      (Listof 
+       (Pairof Symbol String)))])
+
+(define (crypto-pack [d : JSON])
+  (pack (encrypt d _ _ _)))
+
+ \end{alltt}
+\end{minipage}\begin{minipage}[t]{2.2in}
+\begin{alltt}\fname{untyped/client}{1.65}
+(require json 
+         typed/pack-lib
+           typed/crypto-pack-lib)
+
+_ _ _  other dependencies  _ _ _
+_ _ _  and definitions     _ _ _
+
+;; read data from files, pack 
+;; and share securely
+
+(define public-data 
+  (typed-pack 
+   (read-json 
+    "public-records")))
+
+(define secret-data 
+  (crypto-pack 
+   (read-json 
+    "medical-records")))
+
+ _ _ _ rest of client _ _ _
+\end{alltt}\end{minipage}
+
+   \caption{One mix-typed program, three interpretations} \label{fig:example}
+
+  \vspace{-1mm}
+\end{figure*}
+%% -----------------------------------------------------------------------------
+
+The  program in figure~\ref{fig:example} demonstrates the differences
+between natural, transient and erasure using Typed Racket syntax.  It
+consists of four modules: the left and right ones are untyped while
+the two modules in the middle are typed.
+
+The \texttt{untyped/pack-lib} module (left) provides, among
+others, a function \texttt{pack} that consumes JSON data and packages it
+in a dictionary. However, by mistake \texttt{pack} returns
+a hash table instead of the association list its documentation promises.
+The \texttt{client} module (right) uses \texttt{pack} indirectly to
+package public and sensitive data it reads from the filesystem. It does so
+through two different intermediary typed modules (center). The first,
+\texttt{typed/pack-lib}, simply imports \texttt{pack} and immediately
+exports as \texttt{typed-pack} asserting that it is a function
+that consumes JSON and returns an association list that maps symbols to
+strings. The second typed module, \texttt{typed/crypto-pack-lb},
+also imports \texttt{pack} and assumes for it the same type as
+\texttt{typed/pack-lb} but instead of exporting it directly it wraps in
+another function, \texttt{crypto-pack}, that encrypts its input before
+passing it to \texttt{pack}.
+
+Under natural semantics~\cite{tf-popl-2008,tfffgksst-snapl-2017}, when
+\texttt{pack} is imported in either \texttt{typed/pack-lb} or
+\texttt{typed/secure-pack-lb}, the runtime checks that \texttt{pack} is a
+function and wraps it in a proxy that enforces its  type.  Analogously
+the run time wraps the exports of the two typed modules in proxies that
+check that untyped code applies \texttt{typed-pack} and
+\texttt{crypto-pack} to arguments of its type. In essence, the
+return type checks protect  the typed modules in case they use
+\texttt{pack} and, as is the case here, it does not return an association
+list. The check of the argument type  treats the
+exports of the two typed modules as typed values that need to be protected
+when they are used by untyped code.  As this analysis implies, if the
+return type check fails then the problem is with
+\texttt{untyped/pack-lib}, while if the argument type check fails
+responsibility lies with \texttt{client}. The natural semantics keep track
+of the responsible party for each check in the corresponding proxy. Thus,
+in our example, when the call to \texttt{typed-pack} returns, the return
+type check fails and the natural semantics blames \texttt{pack-lib}.
+
+Under the transient semantics~\cite{vss-popl-2017}, the call to \texttt{typed-pack} does not
+produce an error. In contrast to natural, transient does not rely on
+proxies. Instead, it rewrites typed modules to inline checks at appropriate
+places. Furthermore, these checks are
+partial, that is, they check only the top constructor of a type. In other words
+they check types in a  \emph{shallow} manner. In our
+example, transient's inlined checks make sure that the imported \texttt{pack} 
+is a function and that the result of its call in \texttt{typed/crypto-pack-lib} is a list.   Exactly this
+last check fails when \texttt{client} calls to \texttt{crypto-pack}. In
+order to pinpoint the responsible party for the failed checked, transient
+keeps on the side a map from values to the boundaries between modules they
+cross and the corresponding types. In the example, the map records that
+\texttt{pack} crosses from \texttt{untyped/pack-lib} to
+\texttt{typed/pack-lib} and from \texttt{typed/pack-lib} to
+\texttt{client} with the type that appears in the \texttt{required/typed}
+forms in the example. Since the failed check corresponds to the return
+type of \texttt{pack}, transient blames the sources of the two boundary
+crosses, which in this case is \texttt{untyped/pack-lib} for both. In
+general though, transient may blame more than one parties. In fact, the
+theoretical work of~\citet{gfd-oopsla-2019} shows that there are programs
+where transient constructs a blame set that includes modules that are not
+relevant to the check failure or that it even misses entirely the responsible
+party ---  our experiment aims to examine the pragmatic impact of exactly these
+issues on debugging type-level mistakes.
+
+Under erasure semantics, the example does not raise an error due to the
+gradual type system; erasure performs no type checks at runtime. Thus the
+program may fail with an exception if \texttt{client} tries to use the
+result of \texttt{typed-pack} or \texttt{crypto-pack}, but it may also not
+fail at all and produce an wrong result that goes undetected by the
+gradual type system .  In either case, the erasure semantics provides no
+help to the programmer for detecting the bug in our program.
+
+Since, the goal of our experiment is to compare the effectiveness of blame
+across the three semantics, it is critical to isolate their
+distinguishing characteristics from superficial implementation
+differences. Thus we implement all three systems on top of the common
+basis of Racket. First, we use the Typed Racket syntax and type
+checker for all three semantics. Second, for natural we use Typed Racket as
+is while for erasure we run programs with plain Racket after erasing any
+type annotations. Third, for transient we implement a variant of Typed
+Racket that follows closely the semantics from~\citet{vss-popl-2017}.
+
+The first major ingredient of transient Typed Racket is to determine the
+Racket implementation of the shallow checks that map to each type.  For
+certain types, such as {\tt String} or {\tt (Vectorof Byte)}, transient
+can re-use a predicate from the Racket language.  Other types require custom
+predicates; for example the type {\tt Index} of valid array indices allows
+a range of positive exact integers.  Many custom predicates run in
+constant time, but some are more involved.  One linear-time check is for
+{\tt (Listof T)} for any element type {\tt T}, because it must distinguish
+proper from improper list-like structures.  The
+key point, however, is that no shallow check inspects beyond the top
+constructor of each type.
+
+The second major ingredient of transient Typed Racket  is the addition of
+a pass to the compiler of Typed Racket that rewrites all typed code.
+Rewriting modifies all elimination forms to shallow-check their result.
+For example, if the variable {\tt f} has the static type {\tt (-> String
+String)} then the application {\tt (f x)} must be rewritten to check for a
+string result because {\tt f} could be an untyped function.  Rewriting
+also inserts hooks to update the blame map.  When a value crosses a module
+boundary, the blame map records a type assumption.  The inlined checks
+for elimination forms also extend the blame map with further information. For
+instance, the check of the result of {\tt (f x)} adds to the blame map
+that the expected type of the result is the return type of the type of
+{\tt f}.  If a transient check ever fails, transient Typed Racket walks over 
+the information in the blame map and reports a set of blamed components.
+
+\smallskip
+\noindent
+{\bf Note.} As the common basis for the three semantics, we choose 
+Typed Racket's migratory typing. In contrast to the initial formulation of
+gradual typing~\cite{st-sfp-2006}, migratory typing does not provide
+support for type \texttt{Dyn}, components are either untyped or fully
+typed and the gradual type system does not affect the compilation of
+untyped components. Despite, these differences, Typed Racekt is the most
+mature system for sound mix-typed code
+and recent theoretical work has established that
+its migratory typing offers a setting for comparing different approaches to gradual
+typing~\cite{gf-icfp-2018,gfd-oopsla-2019}.

### 2020-11-19 writing/mutate.tex
revision of rational
mutate: add line in mutator table seperating labels from contents
adding landscape description

@@
-\section{How to Make Lots of Mistakes} \label{sec:mutate}
+\section{How to Make Lots of Mistakes}
 
@@
     {\bf name} & {\bf description} & {\bf example} \\
+\hline
 
@@
 opt for tipping the scales in favor of the theoretically weakest
-semantics.
+semantics. We discuss further the implications of this choice in
+section~\ref{sec:discussion}.
 
@@
 instance the class-focused mutators are effective in \texttt{take5}
-which jives with the fact that \texttt{take5} is the benchmark with 
+which jives with the fact that it is the benchmark with 
 the most extensive use of object-oriented features.

### 2020-11-19 writing/rational.tex
adding landscape description

@@
-\section{How to Simulate the Rational Programmer}
+\section{How to Test the Rational Programmer}
 
@@
 Equipped with these preliminary definitions, we can precisely define what it means
-for blame to add value to a gradual type system in a specific setting. 
+for blame to add value to a gradual type system for a specific
+type-level mistake independently of a program's type annotations. 
 \begin{quote}

### 2020-11-19 writing/mutate.tex
revising sample

@@
 \begin{figure*}
-  \begin{tabular}{p{2.1cm} | p{5cm}  | p{5.5cm} }
+  \begin{tabular}{p{2.1cm} | p{7cm}  | p{7.5cm} }
     {\bf name} & {\bf description} & {\bf example} \\
@@
     \texttt{constant}
-& Swap a literal constant with another of the same value but different type.
+& Swaps a constant with another of the same value but different type.
 & \texttt{5.6} $\rightarrow$ \texttt{5.6+0.0i} \\
@@
     \texttt{top-level-id}
- & Swaps two identifiers defined in the same module.
+ & Swaps identifiers defined in the same module.
  &\texttt{(f x 42)} $\rightarrow$ \texttt{(g x 42)}\\
@@
     \texttt{imported-id}
- & Swaps two identifiers imported from the same module.
+ & Swaps identifiers imported from the same module.
  &\texttt{(f x 42)} $\rightarrow$ \texttt{(g x 42)}\\
@@
  & Swaps two field identifiers.
-    & \begin{tabular}[t]{@{}l} \texttt{(get-field o f)} $\rightarrow$ \\
-        \texttt{(get-field o g)}
-      \end{tabular}\\
+    &  \texttt{(get-field o f)} $\rightarrow$  \texttt{(get-field o g)}\\
 \hline
@@
     & \begin{tabular}[t]{@{}l}
-      \texttt{(class object\%}\\
-      \phantom{(d}\texttt{(define/public (m x) x))} $\rightarrow$\\
-      \texttt{(class object\%}\\
-      \phantom{(d}\texttt{(define/private (m x) x))}\\
+      \texttt{(class object (define/public (m x) x))} $\rightarrow$\\
+      \texttt{(class object (define/private (m x) x))}\\
     \end{tabular}\\
@@
     & \begin{tabular}[t]{@{}l}
-      \texttt{(class foo\% (super-new))} $\rightarrow$\\
+      \texttt{(class foo\% (super-new))} $\rightarrow$ \\
       \texttt{(class foo\% (void))}\\
@@
     & Negates conditional test expressions.
-    &  \begin{tabular}[t]{@{}l}
-      \texttt{(if (= x 0) t e)} $\rightarrow$\\
-      \texttt{(if (not (= x 0)) t e)}
-       \end{tabular}\\
+    &  \texttt{(if (= x 0) t e)} $\rightarrow$ \texttt{(if (not (= x 0)) t e)}\\
 \hline
@@
     & Replaces conditional test expressions with \texttt{\#t}.
-    &  \begin{tabular}[t]{@{}l}
-      \texttt{(if (= x 0) t e)} $\rightarrow$\\
-      \texttt{(if \#t t e)}
-    \end{tabular}\\
-\hline
+    & \texttt{(if (= x 0) t e)} $\rightarrow$ \texttt{(if \#t t e)}
 \end{tabular}

### 2020-11-19 writing/rational.tex
revising sample

@@
 Equipped with these preliminary definitions, we can precisely define what it means
-for blame to add value to a gradual type system for a specific
+for blame to useful in the context of a gradual type system for a specific
 type-level mistake independently of a program's type annotations. 
@@
 \begin{itemize}
-\item[$Q_1$] Is blame valuable for natural?
+\item[$Q_1$] Is blame useful in the context of natural?
 
-\item[$Q_2$] Is first blame valuable for transient?
+\item[$Q_2$] Is first blame useful in the context of transient?
 
-\item[$Q_3$] Is last blame valuable for transient?
+\item[$Q_3$] Is last blame useful in the context of transient?
 
@@
 
-Furthermore, our experiment can compare the relative value of blame
+Furthermore, our experiment can compare the relative usefulness of blame
 between any two of the three semantics. 
 \begin{itemize}
-\item[$Q_*$] Is blame for X more valuable than blame for Y where X, Y is natural,
+\item[$Q_*$] Is blame for X more useful than blame for Y where X, Y is natural,
   erasure or natural?
@@
 of the rational programmer from different semantics. For instance to
-decide whether blame for natural is more valuable than blame for
+decide whether blame for natural more useful than blame for
 transient, we compare the usefulness of

### 2020-11-19 writing/sample.tex
revising sample

--- 2020-07-08 writing/sample.tex
@@
-\section{How to Sample the Experimental Space} \label{sec:sample}
+\section{How to Sample the Experimental Space}
 
-As we mention above we treat each definitions as a component. A
-consequence of this decision is that examining the whole lattice
-for each of the mutants becomes unfeasible.
-As figure~\ref{table:benchmarks} shows,  most of the programs contain dozens of definitions  
-resulting in lattices with as many as $1.41 \times 10^{28}$ configurations.
-Since running a single configuration can take 10 minutes or more 
-(due to contract checking and instrumentation), running every possible configuration is impractical.
-Evidence from similar contexts suggests that randomly sampling the lattice instead of
-an exhaustive exploration is an effective alternative~\cite{greenman2019evaluate}.
-Hence we randomly sample enough significant configurations
-from the lattice to obtain a confidence of 0.95 (with margin of error
-0.05) about whether our hypotheses hold or not.
-See accompanying appendix for the details of the calculations for the
-sufficient size of samples for this estimate.
-Exploring more of the lattice would yield higher confidence in the generalizability
-of our results for each benchmark in cases where no violations of a property
-(e.g. Blame Trail) are found.
-In such cases, we could be more certain that we did not simply miss
-the configuration(s) that reveal the violation by exploring the lattice more fully.
-Our choice of random sample counts reflects the (informal) standard practice of estimating
-results to 95 percent or higher confidence.
+The mutators for section~\ref{sec:mutate} generate 96,515,744 debugging scenarios; 
+far too many to even identify the interesting ones among
+them. Hence, we perform stratified uniform random sampling with two
+levels of stratification to collect sufficiently many interesting
+debugging scenarios for drawing statistically sound conclusions. 
+The outcome of our sampling process is that we test each mode of the
+rational programmer on 72192 randomly selected interesting debugging
+scenarios in total spread across both benchmarks and mutators.
+
+As outlined in section~\ref{sec:rational}, all our experimental questions
+require that we decide whether a mode of the rational programmer is
+always, sometimes or never useful for all the interesting scenarios of a
+mutant $\system$. Assuming a normal distribution for the 
+interesting scenarios in the lattice $\lattice{\system}$, we can obtain an
+estimate answer with high confidence by randomly picking a sufficient
+number of interesting scenarios $\lattice{\system}$. Thus
+$\lattice{\system}$ becomes the first stratum for our sampling. For
+95\% confidence, that is a 5\% margin, the sample size needs
+to be just 95 out of $2^N$ scenarios in $\lattice{\system}$ where $N$ is the
+number of components of $\system$.
+
+The second sampling is the set of mutants. After all, our mutators
+generate 16,800 mutants with
+interesting scenarios; still far too many to explore even after sampling
+the lattice of each mutant.  Unlike the lattice of scenarios of a mutant
+, the set of mutants is not a coherent and finite population;  it is 
+unclear how we can relate a sample of mutants to the
+of all possible mutants. Thus we sacrifice the ability to generalize 
+beyond the mutants we randomly select and opt for the largest random
+sample that keeps the experiment computationally feasible. 
+In particular,  we sample 80 mutants with interesting scenarios
+per benchmark and for any benchmark, we ensure that our sample reflects the diversity of mutants with respect
+to the mutators that generated them. Specifically, we select the 80 mutants evenly-distributed
+across all of the mutators that contribute mutants for a benchmark.
+
+%% we use the following algorithm to select the 80 mutants evenly distributed across all of the mutators that do produce mutants of a benchmark:
+%% 1. Treat each mutator as a bucket containing the mutants produced by that mutator
+%% 2. Let the target-sample-size be 80
+%% 3. Let the bucket-sample-size be target-sample-size/number-of-buckets
+%% 4. For each bucket, do
+%%    1. Randomly remove min(bucket-sample-size, mutants-in(bucket)) mutants from the bucket
+%%    2. Set target-sample-size = target-sample-size - (the number of mutants removed)
+%% 5. If target-sample-size is empty, done; else, go back to 3.

### 2020-11-19 writing/conclusion.tex
more tex-os
revising conclusion

@@
 %% -----------------------------------------------------------------------------
-\section{What to Do Next} \label{sec:conclusion}
+\section{What to Do Next}
 
 The paper provides the first insight into the measurable value of blame
-assignment in the gradually typed world. In an implicit manner researchers and
+ in the gradually typed world. In an implicit manner researchers and
 language designers have answered this question one way or another, without any
-evidence. Now they may use the evaluation method presented in this paper to
+evidence. Now they may use the evaluation method of this paper to
 confirm or reject their answers. We already know from~\citet{tgpk-dls-2018}
@@
 sta-nt-base-types, wmwz-ecoop-2017} and that has recently received theoretical
-attention~\cite{gfd-oopsla-2019}: mistakes in type annotations themselves.
+attention~\cite{gfd-oopsla-2019, cc-oopsla-20}: mistakes in type annotations themselves.
 Developers use gradual typing to move an untyped code base into the typed realm,
@@
 already-existing libraries. Because they can, language implementors merely
-create typed facade modules that import untyped functions and exports them with
+create typed facade modules that import untyped functions and export them with
 type annotations---instead of converting the libraries themselves. With those
-facades, the compiler can easily type-check typed components. The problem is
-that these retroactive additions of a specific type to a component and the
-standard run-time functions may result from a misunderstanding of the
+facades, the compiler can easily type-check typed components. Problem is
+that these retroactive additions of types to a library may result from a misunderstanding of the
 code. \emph{They may thus be a mistake themselves.}
 
-The cited evidence suggests that this scenario is quite common, and that a type
-soundness check does {\em not\/} at all address the problem.  Hence, a future
-evaluation must develop type-level mutators that produce incorrect type
+The cited evidence suggests that this scenario is quite common and largely unadressed.
+%, and that a type soundness check does {\em not\/} at all address the problem.  
+Hence, a future evaluation must develop type-level mutators that produce incorrect type
 annotations without breaking the code itself. Some preliminary work suggests
 that such type mutators are much more difficult to develop than the code
-mutators presented in this paper. We conjecture that the natural semantics will
+mutators of this paper. We conjecture that the natural semantics will
 usually discover such mistakes, while the transient and erasure semantics cannot

### 2020-11-19 writing/introduction.tex
small edits

@@
 and instead rely on error messages from the
-underlying programming language.
-
-\mf{This is made up and must be fixed for the submission.}
-
-A direct interpretation of the results (section~\ref{sec:discussion}) is that
+underlying programming language. A direct interpretation of the results
+(section~\ref{sec:discussion}) is that
 they validate the conjecture behind the work of theoreticians. A good blame assignment
-strategy
-greatly shortens the search for impedance mismatches between the specified types
-and the behavior of untyped components. Even better, the evaluation method clearly
-identifies the wrapper-based blame tracking of Typed Racket as the vastly superior
-academic system. Vitousek et al. lose one more time. Of course, a detailed
-discussion must acknowledge the threats to validity of these results.
+strategy helps the search for impedance mismatches between the specified types
+and the behavior of untyped components. Even better, the evaluation method 
+identifies the wrapper-based blame tracking of Typed Racket as the most
+useful academic system with that of Reticulated Python just a short
+distance behind.
 

### 2020-11-19 writing/paper.tex
ogranizing section 5
format + prose tweaks
revision of rational
small edits

@@
-%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
-\documentclass[acmsmall,screen,anonymous,nonacm]{acmart}\settopmatter{printfolios=true,printccs=true,printacmref=true}
-%% For double-blind review submission, w/ CCS and ACM Reference
-%%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
-%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
-%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-%% For single-blind review submission, w/ CCS and ACM Reference
-%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
-%% For final camera-ready submission, w/ required CCS and ACM Reference
-%\documentclass[acmsmall]{acmart}\settopmatter{}
+\documentclass[sigplan,10pt,review,anonymous]{acmart}
+\settopmatter{printfolios=true,printccs=false,printacmref=false}
 
@@
 
-%% Journal information
-%% Supplied to authors by publisher for camera-ready submission;
-%% use defaults for review submission.
+\acmConference[]{}{}{}
+\acmYear{}
+\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
+\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
 \startPage{1}
@@
 %% use 'none' for review submission.
-%% \setcopyright{none}
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\copyrightyear{2018}           %% If different from \acmYear
+\setcopyright{none}
 
@@
 \bibliographystyle{ACM-Reference-Format}
-%% Citation style
-%% Note: author/year citations are required for papers published as an
-%% issue of PACMPL.
-\citestyle{acmauthoryear}   %% For author/year citations
-
-%%%
-\setcopyright{none}
-\acmPrice{1}
-\acmDOI{}
-\acmYear{}
-\copyrightyear{}
-\acmJournal{pacmpl}
-\acmVolume{1}
-\acmNumber{1}
-\acmArticle{1}
-\acmMonth{1}
 
@@
 method for blame assignment strategies and (2) the results from applying
-it to three different gradual type systems. \mf{This sentence must be checked
-before final submission.} These results confirm that blame assignment
+it to three different gradual type systems. These results confirm that blame assignment
 is helpful to a rationally acting programmer.
@@
 
-\input{introduction.tex} 
-\input{challenges.tex}   
-\input{landscape.tex} 
-\input{mutate.tex}
-\input{rational.tex}
-\input{sample.tex}
+\input{introduction.tex}    \label{sec:introduction}
+\input{challenges.tex}      \label{sec:challenges}
+\input{landscape.tex}       \label{sec:landscape}
+\input{mutate.tex}          \label{sec:mutate}
+\input{rational.tex}	      \label{sec:rational}
+\input{sample.tex}          \label{sec:sample}
 \input{results.tex}         \label{sec:results}

### 2020-11-19 writing/sample.tex
small edits

@@
 As outlined in section~\ref{sec:rational}, all our experimental questions
-require that we decide whether a mode of the rational programmer is
-always, sometimes or never useful for all the interesting scenarios of a
-mutant $\system$. Assuming a normal distribution for the 
-interesting scenarios in the lattice $\lattice{\system}$, we can obtain an
-estimate answer with high confidence by randomly picking a sufficient
-number of interesting scenarios $\lattice{\system}$. Thus
-$\lattice{\system}$ becomes the first stratum for our sampling. For
-95\% confidence, that is a 5\% margin, the sample size needs
-to be just 95 out of $2^N$ scenarios in $\lattice{\system}$ where $N$ is the
-number of components of $\system$.
+require that we answer first whether a mode of the rational programmer is
+useful or not for all the interesting scenarios of a mutant $\system$.
+Given the binary nature of this basic question, we can rely on random
+sampling of interesting scenarios from $\lattice{\system}$ to obtain an
+approximate answer.  Thus $\lattice{\system}$ becomes the first stratum
+for our sampling and, given a sufficiently large sample, we can get with
+high confidence either a positive answer about the usefulness of a mode of
+the rational programmer or a negative answer or neither, in which case we
+deduce that the mode we examine is some times useful and some times it is
+not. For 95\% confidence interval and 5\% error margin, the sample  needs
+to include just 96 out of the $2^N$ scenarios of $\lattice{\system}$
+where $N$ is the number of components of $\system$. For lattices that have
+less than 96 interesting scenarios we select all their interesting
+scenarios.
 
-The second sampling is the set of mutants. After all, our mutators
-generate 16,800 mutants with
-interesting scenarios; still far too many to explore even after sampling
-the lattice of each mutant.  Unlike the lattice of scenarios of a mutant
-, the set of mutants is not a coherent and finite population;  it is 
-unclear how we can relate a sample of mutants to the
-of all possible mutants. Thus we sacrifice the ability to generalize 
-beyond the mutants we randomly select and opt for the largest random
-sample that keeps the experiment computationally feasible. 
-In particular,  we sample 80 mutants with interesting scenarios
-per benchmark and for any benchmark, we ensure that our sample reflects the diversity of mutants with respect
-to the mutators that generated them. Specifically, we select the 80 mutants evenly-distributed
+The second sampling stratum is the set of mutants. After all, our mutators
+generate 16,800 mutants with interesting scenarios; still far too many to
+explore even after sampling the lattice of each mutant. Thus we randomly
+select form the mutants of each benchmarks to obtain  the largest random
+sample that keeps the experiment computationally feasible  while ensuring
+that, per benchmark, the sample reflects the diversity of mutants per with
+respect to the mutators that generated them.  Specifically,  we sample 80
+mutants with interesting scenarios per benchmark  evenly-distributed
 across all of the mutators that contribute mutants for a benchmark.

### 2020-11-19 writing/results.tex
Add rough bt-lengths table figure
Add draft avo matrix figure
Draft section 7

@@
   \includegraphics[width=0.7\textwidth]{./plots/usefulness-table}
-  \caption{Usefulness breakdown of all mutants}
+  \caption{Usefulness: Each cell plots the proportion of a benchmark's mutants for which a mode is always useful (green), sometimes useful (yellow), and never useful (red).}
   \label{fig:usefulness-table}
 \end{figure*}
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=0.95\textwidth]{./plots/bt-lengths-table}
+  \caption{Programmer effort: Each cell depicts the distribution of trail lengths for a given mode and benchmark.
+The leftmost bar of every distribution shows the proportion of trails with length zero, and each bar to the right shows the same for length one, two, three, and so on.}
+  \label{fig:effort-table}
+\end{figure*}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.45\textwidth]{./plots/avo-matrix}
+  \caption{Adds-value comparisons: Each row depicts the percentage of mutants for which that row's mode adds value over each other mode (in the columns).
+The diagonals are ommitted because they compare a mode with itself.}
+  \label{fig:avo-matrix}
+\end{figure}
+
+We present our results according to the three dimensions for assessing the utility of blame described in section~\ref{sec:rational}: how often blame is always, sometimes, and never useful (figure~\ref{fig:usefulness-table}); programmer effort (figure~\ref{fig:effort-table}); and how often one mode adds value over another mode (figure~\ref{fig:avo-matrix}).
+In this section we analyze each of these dimensions in turn, and in the next we discuss the analysis's high level conclusions.
+
+Figure~\ref{fig:usefulness-table} breaks down how often blame is useful (as defined in section~\ref{sec:rational}) for every benchmark and every mode of the rational programmer.
+Each mode corresponds to a row of the table, while the columns correspond to our benchmarks.
+Each cell plots the proportion of a benchmark's mutants for which a mode is always useful (green), sometimes useful (yellow), and never useful (red).
+Comparing the overall pattern of one row to the others provides a broad understanding of how the usefulness of that mode compares with the others.
+For instance, the large proportions of red across Erasure indicates that for many mutants it is never useful.
+
+Figure~\ref{fig:effort-table} shows the distribution of trail lengths for every benchmark and every mode of the rational programmer.
+The modes correspond to rows and columns to benchmarks, like figure~\ref{fig:usefulness-table}.
+In this case, however, the cells plot trail length distributions starting from length zero in the leftmost bar and going up by one for each bar to the right.
+Each bar is broken into two sub-bars, which are color-coded to indicate the proportion of trails within that bar which are successful (green) and not (red).
+Again, comparing the rows of the table provides a high level view of how the lengths of trails for each mode tend to differ.
+For example, the stark difference between the Null mode's distribution and that of every mode illustrates how using blame information (as opposed to randomly selecting components to type) dramatically shortens the rational programmer's search.
+The Null distributions look approximately normal \emdash reflecting that the Null mode always selects a random next component to type \emdash but shifted left, reflecting that our scenarios start with a randomly-selected subset of components already typed.
+On the other hand, all of the other modes have distributions that are significantly more concentrated around trails of length one and two.
+
+Finally, figure~\ref{fig:avo-matrix} shows a head-to-head comparison of each mode against every other mode.
+Each cell of the table compares one mode to another using section~\ref{sec:rational}'s definition of a mode \emph{adding value over} another mode.
+In particular, a cell in row $R$ and column $C$ plots the percentage of mutants for which $R$ adds value over $C$.
+For example, the second cell in the row of Natural shows that Natural adds value over Natural exceptions for 18.9\% of all the mutants we consider.
+Reading across a given row $R$ provides an overview of how $R$ compares to every other mode;
+the higher the bars, the more mutants for which $R$ is strictly more useful than the other modes.
+Reading down a given row $C$ provides an overview of how every other mode compares to $C$;
+the higher the bars, the more mutants for which $C$ is strictly less useful than the other modes.

### 2020-11-19 writing/sample.tex
Reword description of sampling

@@
 The outcome of our sampling process is that we test each mode of the
-rational programmer on 72192 randomly selected interesting debugging
+rational programmer on 72,192 randomly selected interesting debugging
 scenarios in total spread across both benchmarks and mutators.
 
-As outlined in section~\ref{sec:rational}, all our experimental questions
+As outlined in section~\ref{sec:rational}, our experimental questions
 require that we answer first whether a mode of the rational programmer is
@@
 approximate answer.  Thus $\lattice{\system}$ becomes the first stratum
-for our sampling and, given a sufficiently large sample, we can get with
-high confidence either a positive answer about the usefulness of a mode of
-the rational programmer or a negative answer or neither, in which case we
-deduce that the mode we examine is some times useful and some times it is
-not. For 95\% confidence interval and 5\% error margin, the sample  needs
-to include just 96 out of the $2^N$ scenarios of $\lattice{\system}$
-where $N$ is the number of components of $\system$. For lattices that have
-less than 96 interesting scenarios we select all their interesting
-scenarios.
+for our sampling; with a sufficiently large sample, we can obtain from the
+sample an estimate of the usefulness of a mode for the entire lattice.
+The estimate tells us with some confidence whether the mode is useful for all trails in the lattice (if all of the samples yield positive results) or none of them.
+In fact, the estimate can say neither if some of our samples yield positive results and others yield negative ones; in that case we can say no more than that the mode is sometimes useful.
+We select a sample size of 96 to obtain 95\% confidence and a 5\% margin of error for our estimates
+(using standard statistical methods, the choice does not depend on the population size for sufficiently large populations).
+For lattices that have less than 96 interesting scenarios, we select them all.
 
@@
 explore even after sampling the lattice of each mutant. Thus we randomly
-select form the mutants of each benchmarks to obtain  the largest random
-sample that keeps the experiment computationally feasible  while ensuring
-that, per benchmark, the sample reflects the diversity of mutants per with
+select from the mutants of each benchmark to obtain the largest
+sample that keeps the experiment computationally feasible.
+Our selection of mutants ensures that, per benchmark, the sample reflects the diversity of mutants with
 respect to the mutators that generated them.  Specifically,  we sample 80
-mutants with interesting scenarios per benchmark  evenly-distributed
-across all of the mutators that contribute mutants for a benchmark.
+mutants with interesting scenarios per benchmark,  evenly-distributed
+across all of the mutators that contribute mutants for the benchmark.
 

### 2020-11-19 writing/discussion.tex
Add rough outline of discussion points

@@
 \begin{itemize}
-  \item Interesting discoveries...?
+  \item Any kind of dynamic checking has a large positive impact, whether blame or stacktraces are used.
+    \begin{itemize}
+    \item Both in terms of usefulness and programmer effort
+    \item Erasure extremely often fails to provide any helpful information.
+    \item That said, there are no long trails for Erasure.
+    \end{itemize}
 
-  \item Comparison of existing GT systems represented in our study.  
-    Natural is here, transient is here, erasure is here, and they
-        compare as...
+  \item Blame helps significantly over just using stacktraces for both Natural and Transient.
+    \begin{itemize}
+    \item Natural's blame improves over exceptions by nearly 20\%
+    \item Transient's blame improves over exceptions half as much
+    \end{itemize}
+
+  \item Natural clearly offers the most utility, but Transient is not too far behind.
+    \begin{itemize}
+    \item Natural clearly improves over all other modes
+    \item Transient improves significantly over all other modes except Natural.
+    \item Keep in mind that our selection of scenarios is biased toward Erasure & Transient.
+    \end{itemize}
+
+  \item Discussion of how confidence / margin of error should be interpreted wrt these results?
 

### 2020-11-20 writing/rational.tex
checkpoint

@@
 the rational programmer receives no hints from the gradual type system on how to
-continue the search for the bug.
+continue the search for the bug. In contrast, if the rational programmer
+knows that for a program  $\system$
+the lattice $\lattice{\system}$ contains only successful natural blame
+trails, then the rational programmer can rely on natural blame while debugging
+a type-level mistakes in $\system$ independently of $\system$'s type
+annotations.
 
-Equipped with these preliminary definitions, we can precisely define what it means
-for blame to useful in the context of a gradual type system for a specific
-type-level mistake independently of a program's type annotations. 
-\begin{quote}
-\it
-Given a program $\system$, the natural blame mode is\\
-(1) \emph{always useful} iff $\lattice{\system}$ contains no failing natural blame trails.\\
-(2) \emph{sometimes useful} iff $\lattice{\system}$ contains successful natural blame trails.\\
-(3) \emph{never useful} iff $\lattice{\system}$ contains no successful natural blame
-trails.\\
-\end{quote}
-
-While these definitions describe how a rational programmer might gauge whether it
-pays off to heed blame assignments during debugging, they do not answer whether
+While the absence of failing natural blame trails in $\lattice{\system}$
+imply that
+pays off to heed blame assignments during debugging $\system$, it does not answer whether
 blame is a critical piece of the rational programmer's process.  For instance,
@@
 \end{quote}
-Similar to natural blame we determine how successful the exception
-migration is based on the success of natural exception trails: 
+
+Putting the two modes together, we can now compare the usefulness of blame 
+and mere exceptions.
 \begin{quote}
 \it 
-Given a program $\system$, the natural exception mode is\\
-(1) \emph{always useful} iff $\lattice{\system}$ contains no failing
-  natural exception trails.\\
-(2) \emph{sometimes useful} iff $\lattice{\system}$ contains successful
-  natural exception trails.\\
-(3) \emph{never useful} iff $\lattice{\system}$ contains no successful natural
-exception trails.\\
+  Given a program $\system$, natural exceptions are  \emph{at least as useful} as 
+  natural blame for $\system$ iff  for every interesting scenario 
+  $\conf_0$ in $\lattice{\system}$ when the natural blame trail 
+  that starts at $\conf_0$ is successful then the natural exception trail that
+  starts at $\conf_0$ is also successful.
 \end{quote}
-
-Putting the pieces together, we can now define when blame 
-is more useful than mere exceptions in the context of a natural semantics.
-\begin{quote}
-\it Natural blame  is more useful
-  than natural exceptions for $\system$ if either natural blame is
-  always useful for $\system$ when natural exception is sometimes or
-  never useful, or natural blame is sometimes useful for $\system$
-  when natural exception is never useful.
-\end{quote}
+\indent Of course if natural exceptions are not at least as useful as
+natural blame for $\system$ then we have evidence that blame 
+is \emph{valuable} for the rational programmer in the context of natural.
 
@@
 is analogous to that for natural and, same as for natural, we use it to  
-specify when the transient-last and transient-first blame modes are more
-useful than exceptions in the context of a natural semantics.
+specify when transient exceptions are at least as useful as the 
+transient-last and transient-first blame.
 
@@
 \begin{itemize}
-\item[$Q_1$] Is blame useful in the context of natural?
+\item[$Q_1$] Is blame valuable in the context of natural?
 
-\item[$Q_2$] Is first blame useful in the context of transient?
+\item[$Q_2$] Is first blame valuable in the context of transient?
 
-\item[$Q_3$] Is last blame useful in the context of transient?
+\item[$Q_3$] Is last blame valuable in the context of transient?
 
@@
 \begin{itemize}
-\item[$Q_*$] Is blame for X more useful than blame for Y where X, Y is natural,
+\item[$Q_*$] Is blame for X at least as useful as blame for Y where X, Y is natural,
   erasure or natural?
@@
 Next, based on the collected data, we calculate the
-success/failure results for the trails to determine whether blame
-is more useful than exceptions for this mutant. Finally, we
+success/failure results for the trails to determine whether exceptions are
+as useful as blame for this mutant. Finally, we
 obtain a positive answer for
-$Q_1$ when the percentage of mutants where blame is more useful
-than exceptions is greater than the percentage of mutants where exceptions
-are more useful
-than blame.  The process is analogous for $Q_2$ and $Q_3$  employing the respective
+$Q_1$ when there are mutants where natural exceptions are not as useful 
+as natural blame.  The process is analogous for $Q_2$ and $Q_3$  employing the respective
 modes of the rational programmer. For $Q_*$, we also proceed in a similar
@@
 of the rational programmer from different semantics. For instance to
-decide whether blame for natural more useful than blame for
-transient, we compare the usefulness of
-the natural blame mode with that of the transient first and transient last
-modes for all our mutants. 
+decide whether blame for natural is as useful as blame for
+transient, we compare the success of
+natural blame trails with that of the transient first and transient last
+blame trails and we determine if one mode is as useful as another. 
 Figure~\ref{fig:process} summarizes the

### 2020-11-20 writing/results.tex
checkpoint

@@
   \includegraphics[width=0.7\textwidth]{./plots/usefulness-table}
-  \caption{Usefulness: Each cell plots the proportion of a benchmark's mutants for which a mode is always useful (green), sometimes useful (yellow), and never useful (red).}
+  \caption{Usefulness: Each cell depicts the proportion of a benchmark's mutants for which a mode is always useful (green), sometimes useful (yellow), and never useful (red).}
   \label{fig:usefulness-table}
@@
   \includegraphics[width=0.95\textwidth]{./plots/bt-lengths-table}
-  \caption{Programmer effort: Each cell depicts the distribution of trail lengths for a given mode and benchmark.
-The leftmost bar of every distribution shows the proportion of trails with length zero, and each bar to the right shows the same for length one, two, three, and so on.}
+  \caption{Programmer effort: Each cell depicts the distribution of trail
+  lengths for a given mode and a benchmark's mutants starting from trails
+  of length 0.}
   \label{fig:effort-table}
@@
   \includegraphics[width=0.45\textwidth]{./plots/avo-matrix}
-  \caption{Adds-value comparisons: Each row depicts the percentage of mutants for which that row's mode adds value over each other mode (in the columns).
-The diagonals are ommitted because they compare a mode with itself.}
+  \caption{Usefulness comparisons: Each cell depicts the percentage of a
+  benchmark's mutants for which the row mode is more useful than the column mode.}
   \label{fig:avo-matrix}

### 2020-11-20 writing/sample.tex
checkpoint

@@
 As outlined in section~\ref{sec:rational}, our experimental questions
-require that we answer first whether a mode of the rational programmer is
-useful or not for all the interesting scenarios of a mutant $\system$.
+boil down whether a mode of the rational programmer is
+as useful as another mode for all the interesting scenarios of a mutant $\system$.
 Given the binary nature of this basic question, we can rely on random

### 2020-11-20 writing/rational.tex
reivising rational and sample for the new stats

@@
 the rational programmer receives no hints from the gradual type system on how to
-continue the search for the bug. In contrast, if the rational programmer
-knows that for a program  $\system$
-the lattice $\lattice{\system}$ contains only successful natural blame
-trails, then the rational programmer can rely on natural blame while debugging
-a type-level mistakes in $\system$ independently of $\system$'s type
-annotations.
+continue the search for the bug.
 
-While the absence of failing natural blame trails in $\lattice{\system}$
-imply that
-pays off to heed blame assignments during debugging $\system$, it does not answer whether
+While a successful natural blame trail indicates that it 
+pays off to heed blame assignments during debugging the trail's root, it does not answer whether
 blame is a critical piece of the rational programmer's process.  For instance,
@@
 Putting the two modes together, we can now compare the usefulness of blame 
-and mere exceptions.
+and mere exceptions for debugging a scenario in the context of natural semantics.
 \begin{quote}
 \it 
-  Given a program $\system$, natural exceptions are  \emph{at least as useful} as 
-  natural blame for $\system$ iff  for every interesting scenario 
-  $\conf_0$ in $\lattice{\system}$ when the natural blame trail 
-  that starts at $\conf_0$ is successful then the natural exception trail that
-  starts at $\conf_0$ is also successful.
+  Given a program $\system$ and a root $\conf_0$ in $\lattice{\system}$,
+  natural blame is \emph{more useful} than natural exceptions for
+  debugging $\conf_0$ iff 
+  the natural blame trail 
+  that starts at $\conf_0$ is successful while the natural exception trail that
+  starts at $\conf_0$ is failing.
 \end{quote}
-\indent Of course if natural exceptions are not at least as useful as
-natural blame for $\system$ then we have evidence that blame 
-is \emph{valuable} for the rational programmer in the context of natural.
 
@@
 is analogous to that for natural and, same as for natural, we use it to  
-specify when transient exceptions are at least as useful as the 
-transient-last and transient-first blame.
+specify when  
+transient-last and transient-first blame are more useful than 
+transient exceptions for debugging a scenario.
 
@@
 mutants from section~\ref{sec:mutate}. In line with the discussion so far, 
-our experimental process collects data to answer three initial questions for each mutant:
+our experimental process collects data to answer three initial questions
+for our interesting debugging scenarios:
 \begin{itemize}
-\item[$Q_1$] Is blame valuable in the context of natural?
+\item[$Q_1$] Is blame useful in the context of natural?
 
-\item[$Q_2$] Is first blame valuable in the context of transient?
+\item[$Q_2$] Is first blame useful in the context of transient?
 
-\item[$Q_3$] Is last blame valuable in the context of transient?
+\item[$Q_3$] Is last blame useful in the context of transient?
 
@@
 \begin{itemize}
-\item[$Q_*$] Is blame for X at least as useful as blame for Y where X, Y is natural,
+\item[$Q_*$] Is blame for X more useful than blame for Y where X, Y is natural,
   erasure or natural?
@@
 
-In detail we answer $Q_1$ for an ill-typed mutant from
-section~\ref{sec:mutate} by analyzing the natural blame and
-natural exception trails in the scenario lattice of the mutant. A
-log reports on success and failure endings.  The first step is to construct the
-mutant's scenario lattice and select one of the interesting debugging
-scenarios from section~\ref{sec:mutate}.
+In detail we answer $Q_1$  by comparing the success of the 
+natural blame and natural exception trails for all interesting 
+debugging scenarios from section~\ref{sec:mutate}.
+  The first step is to construct each
+mutant's scenario lattice and identify their interesting debugging
+scenarios.
 Our test
-bed extends these trails according to the natural-blame programmer.  If no scenarios
+bed extends the trails that start from such roots according to the natural-blame programmer.  If no scenarios
 can be added to the trail, we check if
-the last scenario of the trail type checks or not. If it does, this natural-blame
-trail is successful, otherwise it is failing. We repeat the same process for the
-same roots once more following the natural-exception programmer. 
-Next, based on the collected data, we calculate the
-success/failure results for the trails to determine whether exceptions are
-as useful as blame for this mutant. Finally, we
-obtain a positive answer for
-$Q_1$ when there are mutants where natural exceptions are not as useful 
-as natural blame.  The process is analogous for $Q_2$ and $Q_3$  employing the respective
-modes of the rational programmer. For $Q_*$, we also proceed in a similar
-manner except that we use modes 
-of the rational programmer from different semantics. For instance to
-decide whether blame for natural is as useful as blame for
-transient, we compare the success of
-natural blame trails with that of the transient first and transient last
-blame trails and we determine if one mode is as useful as another. 
-Figure~\ref{fig:process} summarizes the
+the last scenario of the trail type checks or not. If it does, we log natural-blame
+trail is successful, otherwise that is failing. We repeat the process for the
+same roots again but this time for the other mode, natural exceptions.
+ Figure~\ref{fig:process}
+summarizes this
 experimental process for one mode of the rational programmer and connects
 it with the mutations from section~\ref{sec:mutate}.
+Next, based on the collected data, we calculate the
+success/failure results for the trails to determine for each root whether natural blame
+is more useful than natural exceptions. We
+obtain a positive answer for
+$Q_1$ when a root exists where the above is true since it is evidence that
+there is at least one interesting scenario that the rational programmer
+manages to debug because of blame. 
+The process is analogous for $Q_2$ and $Q_3$  employing the respective
+modes of the rational programmer.
+
+For $Q_*$, the process is a bit more
+involved. At a first level we proceed in a similar
+manner except that we use two modes 
+of the rational programmer from different semantics.
+However to answer this question we compare the percentage of scenarios
+where one mode is more useful than the other and the inverse.
+For instance to
+decide whether blame for natural is more useful than first blame for transient, 
+we compare the percentage of interesting scenarios where 
+natural blame is more successful than transient first blame 
+with the percentage of interesting scenarios where transient first blame 
+is more successful than natural blame. 
+Finally we repeat these steps one more time to compare natural blame and transient last blame
+and get a complete picture of the comparative usefulness of blame
+in the two semantics.
 
@@
 can help shed some further light to the effectiveness of the
-three gradual typing systems. For example, consider one of 
-our mutants.  If, for instance, both natural blame and transient first
-blame are equally useful for the mutant, the two nodes of the 
+three gradual typing systems. For example, consider one debugging scenario.  
+If, for instance, both  natural blame and transient first
+blame trails are successful, the two modes of the 
 rational programmer can compete to see which one
-debugs the interesting scenarios of the mutant with less effort. In
-general, if the effort distribution of a mode of the rational programmer for a
-mutant has a shorter tail and more
+debugs the interesting scenario with less effort. In
+general, if the effort distribution of a mode of the rational programmer
+has a shorter tail and more
 volume around smaller values compared to the effort distribution of another
-mode, then the first mode is the more effective of the two.
+mode, then this is evidence that the first mode is the more effective of the two.
 
-Finally, effort allows us to validate that the observed effectiveness 
-of the rational programmer is not an artifact of pure chance.
-To test the null hypothesis for the effectiveness of the rational
-programmer, we can simply compare the effort distribution of a mode of the
+Finally, effort allows us to test whether the observed effectiveness 
+of the rational programmer is an artifact of pure chance or not.
+For that, we can simply compare the effort distribution of a mode of the
 rational programmer with a rational programmer that ignores blame and
@@
 Of course, since each mutant has a finite number of components, 
-such a random mode is always useful for all mutants in the context of all
-three semantics. At the same time though,
-the natural, transient and erasure random mode exhibit a normal effort distribution for each mutant
-that should be distinct from the effort distribution of all other modes of
+such the random mode trails are always successful. At the same time though,
+the natural, transient and erasure random mode exhibit a normal effort
+distribution across scenarios that should be distinct from the effort distribution of all other modes of
 the rational programmer if the effectiveness of the latter is not

### 2020-11-20 writing/sample.tex
reivising rational and sample for the new stats

@@
 
-The mutators for section~\ref{sec:mutate} generate 96,515,744 debugging scenarios; 
-far too many to even identify the interesting ones among
-them. Hence, we perform stratified uniform random sampling with two
-levels of stratification to collect sufficiently many interesting
-debugging scenarios for drawing statistically sound conclusions. 
-The outcome of our sampling process is that we test each mode of the
-rational programmer on 72,192 randomly selected interesting debugging
-scenarios in total spread across both benchmarks and mutators.
+The mutators from section~\ref{sec:mutate} generate 96,515,744 debugging
+scenarios; far too many to even identify the interesting ones among them.
+Furthermore this population is heterogeneous; different scenarios come
+from different mutants and different mutants are the result of a
+different mutator and benchmark pair. Hence, to make the experiment
+computationally feasible and statistically precise, we perform stratified
+uniform random sampling with two levels of stratification.  The first
+stratum is the set of mutants and we randomly select from the mutants of
+each benchmark.  Our selection of mutants ensures that, per
+benchmark, the sample reflects the diversity of mutants with respect to
+the mutators that generated them.  Specifically,  we sample 80 mutants
+with interesting scenarios per benchmark,  evenly-distributed across all
+of the mutators that contribute mutants for the benchmark.  The second
+stratum of our sampling is the lattice of a mutant and we draw 96 random
+interesting scenarios from each lattice. For lattices that have less than
+96 interesting scenarios, we select them all.
 
-As outlined in section~\ref{sec:rational}, our experimental questions
-boil down whether a mode of the rational programmer is
-as useful as another mode for all the interesting scenarios of a mutant $\system$.
-Given the binary nature of this basic question, we can rely on random
-sampling of interesting scenarios from $\lattice{\system}$ to obtain an
-approximate answer.  Thus $\lattice{\system}$ becomes the first stratum
-for our sampling; with a sufficiently large sample, we can obtain from the
-sample an estimate of the usefulness of a mode for the entire lattice.
-The estimate tells us with some confidence whether the mode is useful for all trails in the lattice (if all of the samples yield positive results) or none of them.
-In fact, the estimate can say neither if some of our samples yield positive results and others yield negative ones; in that case we can say no more than that the mode is sometimes useful.
-We select a sample size of 96 to obtain 95\% confidence and a 5\% margin of error for our estimates
-(using standard statistical methods, the choice does not depend on the population size for sufficiently large populations).
-For lattices that have less than 96 interesting scenarios, we select them all.
-
-The second sampling stratum is the set of mutants. After all, our mutators
-generate 16,800 mutants with interesting scenarios; still far too many to
-explore even after sampling the lattice of each mutant. Thus we randomly
-select from the mutants of each benchmark to obtain the largest
-sample that keeps the experiment computationally feasible.
-Our selection of mutants ensures that, per benchmark, the sample reflects the diversity of mutants with
-respect to the mutators that generated them.  Specifically,  we sample 80
-mutants with interesting scenarios per benchmark,  evenly-distributed
-across all of the mutators that contribute mutants for the benchmark.
+In summary, our sampling process takes advantage of ancillary information
+to deliver a representative sufficiently large set of scenarios. As a result we test each
+mode of the rational programmer on 72,192 randomly selected interesting
+debugging scenarios spread across both benchmarks and mutators.
+Furthermore stratified random sampling allows us to derive conclusions
+about the proportions of the whole population from the proportions of
+our sample, exactly what we need to project comparisons about the
+usefulness of various modes of the
+rational programmer from the sample to set of all interesting debugging
+scenarios.  Finally, we can also provide separate estimates per stratum and provide different
+viewpoints of our results. For instance, because of the size of the sample
+of scenarios from the lattice of a mutant, we can estimate whether natural
+blame is more useful than natural exceptions for all interesting scenarios
+of each mutant separately with 95\% confidence and 5\% error margin.
 

### 2020-11-20 writing/sample.tex
reivising rational and sample for the new stats

@@
 computationally feasible and statistically precise, we perform stratified
-uniform random sampling with two levels of stratification.  The first
-stratum is the set of mutants and we randomly select from the mutants of
-each benchmark.  Our selection of mutants ensures that, per
+uniform random sampling with three levels of stratification.  The first
+two strata group mutants first by benchmark and then by mutator. 
+We randomly select from the mutants of
+each benchmark ensuring that, per
 benchmark, the sample reflects the diversity of mutants with respect to
@@
 with interesting scenarios per benchmark,  evenly-distributed across all
-of the mutators that contribute mutants for the benchmark.  The second
+of the mutators that contribute mutants for the benchmark.  The third
 stratum of our sampling is the lattice of a mutant and we draw 96 random
@@
 about the proportions of the whole population from the proportions of
-our sample, exactly what we need to project comparisons about the
+our sample despite the population's heterogeneity, exactly what we need to project comparisons about the
 usefulness of various modes of the
-rational programmer from the sample to set of all interesting debugging
-scenarios.  Finally, we can also provide separate estimates per stratum and provide different
+rational programmer from the sample to the set of all interesting debugging
+scenarios.  Finally, we can also calculate separate estimates per stratum and provide different
 viewpoints of our results. For instance, because of the size of the sample
@@
 blame is more useful than natural exceptions for all interesting scenarios
-of each mutant separately with 95\% confidence and 5\% error margin.
+of each mutant with 95\% confidence and 5\% error margin.
 

### 2020-11-20 writing/sample.tex
reivising rational and sample for the new stats

@@
 In summary, our sampling process takes advantage of ancillary information
-to deliver a representative sufficiently large set of scenarios. As a result we test each
+to deliver a  sufficiently large representative set of scenarios.
+Concretely, we test each
 mode of the rational programmer on 72,192 randomly selected interesting

### 2020-11-20 writing/sample.tex
reivising rational and sample for the new stats

@@
 debugging scenarios spread across both benchmarks and mutators.
-Furthermore stratified random sampling allows us to derive conclusions
-about the proportions of the whole population from the proportions of
-our sample despite the population's heterogeneity, exactly what we need to project comparisons about the
+Furthermore stratified random sampling allows us to derive precise conclusions
+about the proportion of the whole population that satisfies a given trait 
+from the proportion of the sample 
+that satisfies the trait despite the population's heterogeneity, exactly what we need to project comparisons about the
 usefulness of various modes of the
 rational programmer from the sample to the set of all interesting debugging
-scenarios.  Finally, we can also calculate separate estimates per stratum and provide different
-viewpoints of our results. For instance, because of the size of the sample
+scenarios.  Finally, we can also calculate separate estimates per stratum
+and provide an alternative
+viewpoint of our results. For instance, because of the size of the sample
 of scenarios from the lattice of a mutant, we can estimate whether natural

### 2020-11-21 writing/results.tex
Replace bt-lengths-table with version aggregating over all benchmarks

@@
 
-\begin{figure*}
+\begin{figure}
   \centering
-  \includegraphics[width=0.95\textwidth]{./plots/bt-lengths-table}
-  \caption{Programmer effort: Each cell depicts the distribution of trail
-  lengths for a given mode and a benchmark's mutants starting from trails
+  \includegraphics[width=0.4\textwidth]{./plots/bt-lengths-table}
+  \caption{Programmer effort: Each plot depicts the distribution of trail
+  lengths for a given mode across all benchmarks, starting from trails
   of length 0.}
   \label{fig:effort-table}
-\end{figure*}
+\end{figure}
 

### 2020-11-21 writing/introduction.tex
almost done with results

@@
 identifies the wrapper-based blame tracking of Typed Racket as the most
-useful academic system with that of Reticulated Python just a short
-distance behind.
+useful academic system with that of Reticulated Python only marginally  behind.
 

### 2020-11-21 writing/results.tex
almost done with results

@@
 
-\begin{figure*}
+\begin{figure}
   \centering
-  \includegraphics[width=0.7\textwidth]{./plots/usefulness-table}
-  \caption{Usefulness: Each cell depicts the proportion of a benchmark's mutants for which a mode is always useful (green), sometimes useful (yellow), and never useful (red).}
-  \label{fig:usefulness-table}
-\end{figure*}
+  \includegraphics[width=0.45\textwidth]{./plots/avo-matrix}
+  \caption{Usefulness comparisons: Each cell depicts the estimated percentage of
+  interesting debugging scenarios for which the row mode is more useful
+  than the column mode.
+  The upper bound of error margin is 0.02\%}
+  \label{fig:avo-matrix}
+\end{figure}
+
+We run the experimental process from section~\ref{sec:rational} on the
+sampled debugging scenarios from section~\ref{sec:sample} on a --- specs
+of fix. Each debugging scenario had a timeout limit of 4 minutes and an
+out of memory limit of 6GB. The complete experiment required
+approximately 270 hours of compute hours on our machine.
+
+Figure~\ref{fig:avo-matrix} summarizes our results and provides answers to
+the four questions from section~\ref{subsec:experiment}.  Specifically, it
+depicts a head-to-head comparison of the usefulness of each mode against
+every other mode.  The cell in row $R$ and column $C$ plots the percentage
+of interesting debugging scenarios for which $R$ is more useful $C$ ---
+this percentage has been calculated as an estimated proportion from the
+corresponding proportion of the debugging scenarios in our sample and the
+upper bound.  In effect, reading across a given row $R$ provides an
+overview of how $R$ compares to every other mode; the higher the bars, the
+more interesting debugging scenarios for which $R$ is strictly more useful
+than the other modes.  Reading down a given row $C$ provides an overview
+of how every other mode compares to $C$; the higher the bars, the more
+interesting debugging scenarios for which $C$ is strictly less useful than
+the other modes. For example, the second cell in the row of Natural blame
+shows that Natural is more useful than Natural exceptions for 12\% of all
+the interesting debugging scenarios we generate from our benchmarks and
+mutators.
+
+Figure~\ref{subsec:experiment} shows that the answers for questions $Q_1$
+to $Q_3$ are positive. In all three semantics, blame modes outperform
+their corresponding exception mode (by 9.44\% to 12.7\%). The natural
+exceptions mode is never more useful than natural blame and  transient
+exceptions are more useful than transient first and transient last blame
+in 2.19\% and 2.8\% of the scenarios respectively.
+
+For the answers to $Q_*$, we compare the percentages in antisymmetric
+positions in the matrix. Blame for all three semantics is significantly
+more useful than Erasure exceptions (by 33\% to 35\%). Natural blame is
+more useful than both versions of transient blame by a small percentage;
+in 4.05\% of scenarios natural blame is more useful while in 1.67\% of the
+scenarios transient blame is more useful. The transient first and
+transient last blame are practically indistinguishable. Finally, there is
+no clear winner between natural exceptions and transient exceptions
+despite the theoretically advantageous additional checks of the natural
+semantics.
 
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=0.45\textwidth]{./plots/avo-matrix}
-  \caption{Usefulness comparisons: Each cell depicts the percentage of a
-  benchmark's mutants for which the row mode is more useful than the column mode.}
-  \label{fig:avo-matrix}
-\end{figure}
-
-We present our results according to the three dimensions for assessing the utility of blame described in section~\ref{sec:rational}: how often blame is always, sometimes, and never useful (figure~\ref{fig:usefulness-table}); programmer effort (figure~\ref{fig:effort-table}); and how often one mode adds value over another mode (figure~\ref{fig:avo-matrix}).
-In this section we analyze each of these dimensions in turn, and in the next we discuss the analysis's high level conclusions.
-
-Figure~\ref{fig:usefulness-table} breaks down how often blame is useful (as defined in section~\ref{sec:rational}) for every benchmark and every mode of the rational programmer.
-Each mode corresponds to a row of the table, while the columns correspond to our benchmarks.
-Each cell plots the proportion of a benchmark's mutants for which a mode is always useful (green), sometimes useful (yellow), and never useful (red).
-Comparing the overall pattern of one row to the others provides a broad understanding of how the usefulness of that mode compares with the others.
-For instance, the large proportions of red across Erasure indicates that for many mutants it is never useful.
-
-Figure~\ref{fig:effort-table} shows the distribution of trail lengths for every benchmark and every mode of the rational programmer.
-The modes correspond to rows and columns to benchmarks, like figure~\ref{fig:usefulness-table}.
-In this case, however, the cells plot trail length distributions starting from length zero in the leftmost bar and going up by one for each bar to the right.
-Each bar is broken into two sub-bars, which are color-coded to indicate the proportion of trails within that bar which are successful (green) and not (red).
-Again, comparing the rows of the table provides a high level view of how the lengths of trails for each mode tend to differ.
-For example, the stark difference between the Null mode's distribution and that of every mode illustrates how using blame information (as opposed to randomly selecting components to type) dramatically shortens the rational programmer's search.
-The Null distributions look approximately normal \emdash reflecting that the Null mode always selects a random next component to type \emdash but shifted left, reflecting that our scenarios start with a randomly-selected subset of components already typed.
-On the other hand, all of the other modes have distributions that are significantly more concentrated around trails of length one and two.
-
-Finally, figure~\ref{fig:avo-matrix} shows a head-to-head comparison of each mode against every other mode.
-Each cell of the table compares one mode to another using section~\ref{sec:rational}'s definition of a mode \emph{adding value over} another mode.
-In particular, a cell in row $R$ and column $C$ plots the percentage of mutants for which $R$ adds value over $C$.
-For example, the second cell in the row of Natural shows that Natural adds value over Natural exceptions for 18.9\% of all the mutants we consider.
-Reading across a given row $R$ provides an overview of how $R$ compares to every other mode;
-the higher the bars, the more mutants for which $R$ is strictly more useful than the other modes.
-Reading down a given row $C$ provides an overview of how every other mode compares to $C$;
-the higher the bars, the more mutants for which $C$ is strictly less useful than the other modes.
+Figure~\ref{fig:effort-table} shows the distribution of programmer effort
+for the sampled interesting debugging --- in contrast to the usefulness
+comparisons these results do not generalize to the full population.

### 2020-11-21 writing/sample.tex
almost done with results

@@
 uniform random sampling with three levels of stratification.  The first
-two strata group mutants first by benchmark and then by mutator. 
+two levels of strata group mutants first by benchmark and then by mutator. 
 We randomly select from the mutants of
@@
 of the mutators that contribute mutants for the benchmark.  The third
-stratum of our sampling is the lattice of a mutant and we draw 96 random
-interesting scenarios from each lattice. For lattices that have less than
-96 interesting scenarios, we select them all.
+level of strata for our sampling is the lattice of a single mutant and we draw 96 random
+interesting scenarios from each lattice with replacement.
 
@@
 rational programmer from the sample to the set of all interesting debugging
-scenarios.  Finally, we can also calculate separate estimates per stratum
-and provide an alternative
-viewpoint of our results. For instance, because of the size of the sample
-of scenarios from the lattice of a mutant, we can estimate whether natural
-blame is more useful than natural exceptions for all interesting scenarios
-of each mutant with 95\% confidence and 5\% error margin.
+scenarios.
 

### 2020-11-21 writing/mutate.tex
working on last bit of results

@@
 
-\begin{figure*}
-  \centering
-  \includegraphics[scale=0.35]{./plots/mutant-breakdown}
-  \caption{Interesting debugging scenarios produced by our mutators. Counts are cut off at 50.}
-  \label{fig:mutant-breakdown}
-\end{figure*}
+%\begin{figure*}
+%  \centering
+%  \includegraphics[scale=0.35]{./plots/mutant-breakdown}
+%  \caption{Interesting debugging scenarios produced by our mutators. Counts are cut off at 50.}
+%  \label{fig:mutant-breakdown}
+%\end{figure*}
 
@@
 across all mutants --- there are potentially many more but this is how
-many we examined. 
-Figure~\ref{fig:mutant-breakdown} breaks down the interesting
-debugging scenarios produced  for each benchmark by mutator 
-but counts are cut off at 50. The mutators produce at least 40 interesting
+many we examined. In more detail, the mutators produce at least 40 interesting
 scenarios for every benchmark, and these scenarios originate from at
@@
 benchmark. Especially some mutators  target specific features of
+benchmarks even if they do not contribute interesting scenarios to other
 benchmarks, which

### 2020-11-21 writing/results.tex
working on last bit of results

@@
 Figure~\ref{fig:effort-table} shows the distribution of programmer effort
-for the sampled interesting debugging --- in contrast to the usefulness
-comparisons these results do not generalize to the full population.
+for  the sampled interesting debugging scenarios . Unlike the usefulness
+comparisons above, these results do not generalize to the full population.
+Still, as we discuss in section~\ref{subsec:effort}, they provide an
+alternative view of the workings of the rational programmer. There are two
+immediate take-aways from the figure. First, the effort for succesfully
+debugging interesting scenarios (in green) for the random mode of the
+rational program  follows  a normal distribution as expected.\footnote{The
+random mode distribution is indistinguishable for all three semantics and thus the figure~\ref{fig:effort-table} 
+only shows one plot.} In contrast, in the other modes, succesful effort coalesces at
+the left side of the plot, meaning that in most cases the programmer needs
+to type a single component to debug a scenario. Even in the minority of
+cases where the blame or exception rational programmer has to work harder, the number of
+modules that that need to be typed is small relative to the random
+rational programmer. A second point of interest is that the exceptions of
+the erasure semantics either help the rational programmer immediately or 
+the rational programmer fails to debug a scenario (in red). This shouldn't
+come as a surprise; no matter how many type annotations the programmer
+adds to an erasure program, if the program

### 2020-11-21 writing/discussion.tex
Add outline items to discussion outline
Add draft of discussion

--- 2020-11-18 writing/discussion.tex
@@
 
-\begin{itemize}
-  \item Interesting discoveries...?
+%% \begin{itemize}
+%% \item Why does the data look like this?
+%%   - why tr fails
+%%     - bad runtime errors
+%%     - TR bug
+%%   - why transient fails
+%%     - bad runtime errors
+%%     - timeouts/ooms: give context wrt fully untyped/typed versions
+%%     - empty blame: trim Ben's prose
 
-  \item Comparison of existing GT systems represented in our study.  
-    Natural is here, transient is here, erasure is here, and they
-        compare as...
+%%   \item Any kind of dynamic checking has a large positive impact, whether blame or stacktraces are used.
+%%     \begin{itemize}
+%%     \item Both in terms of usefulness and programmer effort
+%%     \item Erasure extremely often fails to provide any helpful information.
+%%     \item That said, there are no long trails for Erasure.
+%%     \end{itemize}
 
-      \item Threats to validity:
-        \begin{itemize}
-          \item Racket stack traces
-          \item Mutant sampling
-          \item Blame trails do not necessarily capture all aspects of error reporting
-          \item Transient blame adaptations
-          \item More?
+%%   \item Blame helps significantly over just using stacktraces for both Natural and Transient.
+%%     \begin{itemize}
+%%     \item Natural's blame improves over exceptions by nearly 20\%
+%%     \item Transient's blame improves over exceptions half as much
+%%     \end{itemize}
 
-        \end{itemize}
-\end{itemize}
+%%   \item Natural clearly offers the most utility, but Transient is not too far behind.
+%%     \begin{itemize}
+%%     \item Natural clearly improves over all other modes
+%%     \item Transient improves significantly over all other modes except Natural.
+%%     \item Keep in mind that our selection of scenarios is biased toward Erasure & Transient.
+%%     \end{itemize}
+
+%%   \item Discussion of how confidence / margin of error should be interpreted wrt these results?
+
+%%    \item Threats to validity:
+%%         \begin{itemize}
+%%           \item Racket stack traces
+%%           \item Mutant sampling
+%%           \item Blame trails do not necessarily capture all aspects of error reporting
+%%           \item Transient blame adaptations
+%%           \item More?
+
+%%    \end{itemize}
+%% \end{itemize}
+
+\subsection{Why does the data look like this?}
+
+The figures in the last section raise a key question: why does blame fail in our dataset?
+We analyze the reasons for failure separately for each mode of the rational programmer.
+
+Breaking down the reasons for failure for all scenarios in which Natural is unsuccessful reveals two causes.
+First, most of the failing scenarios begin with a runtime exception, which the rational programmer can use to type components until blame becomes available.
+In these failing cases (there are 1708), however, blame never becomes available; the exception information points to the wrong component and the trail ends in failure.
+A second, much smaller set (40 scenarios) do produce a runtime type error blaming some component, but the blame identifies a non-buggy typed component.
+After careful inspection of these scenarios, we are confident that they reveal a bug in the implementation of Typed Racket.
+
+Transient's blame failures break down in a similar way.
+Like Natural, most failures occur with unhelpful exception information without ever getting blame (1851 scenarios).
+However, Transient also has a substantial number of failures that occur because the programs hit the resource limits (timed out or ran out of memory) and had to be killed (some 770 scenarios).
+Additionally, there are nearly a thousand cases where Transient's checks catch a problem, but the blame set reported is empty.
+We discuss each of these last two failure categories in detail in~\ref{sec:threat:transient}.
+
+\subsection{High level conclusions}
+
+The analysis in the last section indicates a number of interesting conclusions about blame in the gradual typing setting.
+First, figure~\ref{fig:avo-matrix} makes it clear that any kind of dynamic checking has a large positive impact over Erasure, and this is true regardless of whether the rational programmer uses blame or stacktraces.
+Indeed, the bottom row of figure~\ref{fig:avo-matrix} illustrates that Erasure almost never offers more useful information than any other mode.
+Nonetheless, figure~\ref{fig:effort-table} does show that even if erasure often fails to provide helpful information, it provides the rational programmer with a quick answer.
+
+Figure~\ref{fig:avo-matrix} also indicates that blame helps significantly over just using exception information.
+Both Natural and Transient's blame modes improve over their corresponding exception modes by about ten percent.
+This illustrates that blame tracking offers value for locating the source of mistakes beyond the value provided by the dynamic checks alone.
+At only ten percent, however, the amount of added value suggests that foregoing blame tracking in order to obtain performance gains may not hurt debugging all too much.
+
+Finally, the results show that while Natural clearly offers the most utility compared to the other modes, Transient is not too far behind.
+Natural is more useful than every other mode in figure~\ref{fig:avo-matrix}, and by a larger proportion than any other mode.
+That said, the same can be said of Transient's two blame modes as compared to every other mode but Natural.
+Furthermore, the difference in usefulness between Natural and Transient's blame modes is quite small across the board.
+As we point out in~\ref{sec:mutate}, however, we should note that our selection of scenarios for analysis introduces bias toward Erasure and Transient.
+
+\subsection{Threats to validity}
 
@@
 \subsection{Threat: Is our Transient Correct?}
+\label{sec:threat:transient}
 %% purpose: talk about transient-blame implementation .... what did we do,
@@
 
-Surprisingly, Transient blames an empty set of boundaries in XXXX cases.
+Surprisingly, Transient blames an empty set of boundaries in 967 cases.
 Such cases should never occur, at least in theory, because an empty set

### 2020-11-21 writing/results.tex
complete results description

--- 2020-11-21 writing/results.tex
@@
 \section{What We Discovered}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.45\textwidth]{./plots/avo-matrix}
-  \caption{Usefulness comparisons: Each cell depicts the estimated percentage of
-  interesting debugging scenarios for which the row mode is more useful
-  than the column mode.
-  The upper bound of error margin is 0.02\%}
-  \label{fig:avo-matrix}
-\end{figure}
 
@@
   \centering
+  \includegraphics[width=0.45\textwidth]{./plots/avo-matrix}
+  \caption{Usefulness comparisons: Each cell depicts the estimated percentage of
+  interesting debugging scenarios for which the row mode is more useful
+  than the column mode.
+  The upper bound of error margin is 0.02\%}
+  \label{fig:avo-matrix}
+\end{figure}
+
+\begin{figure}
+  \centering
   \includegraphics[width=0.4\textwidth]{./plots/bt-lengths-table}
@@
 Figure~\ref{fig:effort-table} shows the distribution of programmer effort
-for the sampled interesting debugging --- in contrast to the usefulness
-comparisons these results do not generalize to the full population.
+for our sample of interesting debugging scenarios . Unlike the usefulness
+comparisons above, these results do not generalize to the full population.
+Still, as we discuss in section~\ref{subsec:effort}, they provide an
+alternative view of the workings of the rational programmer.
+
+There are
+three immediate take-aways from the figure. First, the effort for successfully
+debugging interesting scenarios (in green) for the random mode of the
+rational program  follows  a normal distribution as expected.\footnote{The
+random mode distribution is indistinguishable for all three semantics and thus the figure~\ref{fig:effort-table} 
+only shows one plot.} In contrast, in the other modes, successful effort coalesces at
+the left side of the plot, meaning that in most cases the programmer needs
+to type a single component to debug a scenario. Even in the minority of
+cases where a blame or exception mode has to work harder, the number of
+modules that that need to be typed is small relative to the random mode.
+
+A second point of interest is that the exceptions of
+the erasure semantics either help the rational programmer immediately or 
+the rational programmer fails to debug a scenario altogether (in red). This shouldn't
+come as a surprise; no matter how many type annotations the programmer
+adds to an erasure program, if the type checker doesn't reject the
+program, running the program produces always the same outcome. Thus an
+exception from the erasure's runtime has to point to the buggy component with
+the first try. Otherwise the rational programmer types an irrelevant
+component, runs the program again and the exception points again to the
+already typed component.
+
+A third interesting piece of information from
+the figure is that every mode has failed debugging scenarios. This should
+not come as a surprise to the astute
+reader After all, as we
+discuss above, some modes are more successful than others in certain
+scenarios. Indeed any mode can fail if running a scenario results in an
+exception and the exception carries no useful information about which
+component the programmer should type next, e.g., because the stack trace 
+does not contain frames from the component of the program.
+This explains the majority of
+failures to debug failed scenarios but not all. The next section takes a
+closer look at these unexpected situations as part of the lessons we
+extract from our experimental results.

### 2020-11-21 writing/discussion.tex
Edit threats

@@
 developer for the sake of a systematic, large-scale analysis. This decoupling
-comes advantages and disadvantages. In the economic realm, mathematical models
+comes with advantages and disadvantages. In the economic realm, mathematical models
 have provided some predictive insights into the market's behavior; but as
@@
 
-Our setup hides the type choice a rational programmer must make. When the
+Relatedly, our setup hides the type choice a rational programmer must make. When the
 run-time checks signal an impedance mismatch in the real world, a programmer
@@
     The initial blame map must reflect the initial type environment.
-    Typed Racket trusts that untyped core-library functions, such as \texttt{filter},
-     behave correctly, but does not list the assumption in the blame map.
+    For instance, Typed Racket trusts that untyped core-library functions, such as \texttt{filter},
+     behave correctly; such assumptions should appear in the blame map.
 \end{itemize}
@@
 worst-case of 5.4x on fully-typed benchmarks in Reticulated,
- our configurations frequently exceed a 10 minute timeout with transient blame.
+some of our configurations exceed a 4 minute timeout or 6GB memory limit with transient blame.
+To put those limits into context, the fully typed and fully untyped benchmarks all complete in a few seconds with minimal memory usage, and none of the mixed Natural configurations hit these limits.
 Unfortunately, our high cost is closer to the truth;

### 2020-11-21 writing/results.tex
small edit in results

@@
 program, running the program produces always the same outcome. Thus an
-exception from the erasure's runtime has to point to the buggy component with
+exception from the erasure's runtime has to point to the buggy module with
 the first try. Otherwise the rational programmer types an irrelevant
-component, runs the program again and the exception points again to the
-already typed component.
+module, runs the program again and the exception points again to the
+already typed module.
 
-A third interesting piece of information from
-the figure is that every mode has failed debugging scenarios. This should
+A third notable piece of information from
+the figure is that every mode has failed debugging scenarios, not just
+erasure. This should
 not come as a surprise to the astute
-reader After all, as we
+reader. After all, as we
 discuss above, some modes are more successful than others in certain
@@
 exception and the exception carries no useful information about which
-component the programmer should type next, e.g., because the stack trace 
-does not contain frames from the component of the program.
-This explains the majority of
-failures to debug failed scenarios but not all. The next section takes a
-closer look at these unexpected situations as part of the lessons we
+module the programmer should type next, e.g., because the stack trace 
+does not contain frames from any module of the program. This situation
+can even happen immediately, which explains failures with 0 effort. 
+However  while the majority of
+failures to debug failed scenarios follow this pattern, a few do not. The next section takes a
+closer look at these unexpected failures of the rational programmer as part of the lessons we
 extract from our experimental results.

### 2020-11-21 writing/sample.tex
sample: small rewording

@@
 debugging scenarios spread across both benchmarks and mutators.
-Furthermore stratified random sampling allows us to derive precise conclusions
+Furthermore, stratified random sampling allows us to derive precise conclusions
 about the proportion of the whole population that satisfies a given trait 
-from the proportion of the sample 
-that satisfies the trait despite the population's heterogeneity, exactly what we need to project comparisons about the
+from the sample despite the population's heterogeneity.
+That property is exactly what we need to project comparisons about the
 usefulness of various modes of the

### 2020-11-21 writing/results.tex
Editing results

@@
 We run the experimental process from section~\ref{sec:rational} on the
-sampled debugging scenarios from section~\ref{sec:sample} on a --- specs
-of fix. Each debugging scenario had a timeout limit of 4 minutes and an
-out of memory limit of 6GB. The complete experiment required
+sampled debugging scenarios from section~\ref{sec:sample} using a machine with two Intel Xeon Gold 6258R processors and 500GB of memory.
+Each debugging scenario had a timeout limit of 4 minutes and a
+memory limit of 6GB. The complete experiment required
 approximately 270 hours of compute hours on our machine.
@@
 to $Q_3$ are positive. In all three semantics, blame modes outperform
-their corresponding exception mode (by 9.44\% to 12.7\%). The natural
-exceptions mode is never more useful than natural blame and  transient
-exceptions are more useful than transient first and transient last blame
+their corresponding exception mode (by 9.44\% to 12.7\%). The Natural
+exceptions mode is never more useful than Natural blame, and  Transient
+exceptions are more useful than Transient first and Transient last blame
 in 2.19\% and 2.8\% of the scenarios respectively.
@@
 more useful than Erasure exceptions (by 33\% to 35\%). Natural blame is
-more useful than both versions of transient blame by a small percentage;
-in 4.05\% of scenarios natural blame is more useful while in 1.67\% of the
-scenarios transient blame is more useful. The transient first and
-transient last blame are practically indistinguishable. Finally, there is
-no clear winner between natural exceptions and transient exceptions
-despite the theoretically advantageous additional checks of the natural
+more useful than both versions of Transient blame by a small percentage;
+in 4.05\% of scenarios Natural blame is more useful while in 1.67\% of the
+scenarios Transient blame is more useful. The Transient first and
+Transient last blame are practically indistinguishable. Finally, there is
+no clear winner between Natural exceptions and Transient exceptions
+despite the theoretically advantageous additional checks of the Natural
 semantics.
@@
 Figure~\ref{fig:effort-table} shows the distribution of programmer effort
-for our sample of interesting debugging scenarios . Unlike the usefulness
-comparisons above, these results do not generalize to the full population.
+for our sample of interesting debugging scenarios. Unlike the usefulness
+comparisons above, these proportions do not generalize to a representation of the full population.
 Still, as we discuss in section~\ref{subsec:effort}, they provide an
@@
 debugging interesting scenarios (in green) for the random mode of the
-rational program  follows  a normal distribution as expected.\footnote{The
+rational programmer follows  a normal distribution, as expected.\footnote{The
 random mode distribution is indistinguishable for all three semantics and thus the figure~\ref{fig:effort-table} 
@@
 A second point of interest is that the exceptions of
-the erasure semantics either help the rational programmer immediately or 
+the Erasure semantics either help the rational programmer immediately or 
 the rational programmer fails to debug a scenario altogether (in red). This shouldn't
 come as a surprise; no matter how many type annotations the programmer
-adds to an erasure program, if the type checker doesn't reject the
-program, running the program produces always the same outcome. Thus an
-exception from the erasure's runtime has to point to the buggy module with
+adds to an Erasure program, if the type checker doesn't reject the
+program, running the program always produces the same outcome. Thus an
+exception from the runtime has to point to the buggy module with
 the first try. Otherwise the rational programmer types an irrelevant
@@
 the figure is that every mode has failed debugging scenarios, not just
-erasure. This should
+Erasure. This should
 not come as a surprise to the astute
@@
 module the programmer should type next, e.g., because the stack trace 
-does not contain frames from any module of the program. This situation
+does not contain frames from any module of the program, or only typed modules. This situation
 can even happen immediately, which explains failures with 0 effort. 
-However  while the majority of
+While the majority of
 failures to debug failed scenarios follow this pattern, a few do not. The next section takes a

### 2020-11-21 writing/results.tex
results: update compute hours
small edits in results

@@
 the Erasure semantics either help the rational programmer immediately or 
-the rational programmer fails to debug a scenario altogether (in red). This shouldn't
-come as a surprise; no matter how many type annotations the programmer
+the rational programmer fails to debug a scenario altogether (in red).
+This is expected; no matter how many type annotations the programmer
 adds to an Erasure program, if the type checker doesn't reject the
@@
 can even happen immediately, which explains failures with 0 effort. 
-While the majority of
-failures to debug failed scenarios follow this pattern, a few do not. The next section takes a
+While the most
+failures to debug scenarios follow this pattern, a few do not. The next section takes a
 closer look at these unexpected failures of the rational programmer as part of the lessons we

### 2020-11-21 writing/challenges.tex
slight adjustment of challenges
Lots of minor changes and rewordings

@@
-\section{Why is it hard to evaluate blame?}
+\section{Why Is It Hard to Evaluate Blame?}
+\label{sec:challenges}
 
@@
 another. For instance, the case study presented here builds on
-Greenman and Felleisen's transient variant for functional Typed
-Racket~\cite{gf-icfp-2018} to obtain a transient semantics that is
-comparable with Typed Racket's default natural one. This transient
+Greenman and Felleisen's Transient variant for functional Typed
+Racket~\cite{gf-icfp-2018} to obtain a Transient semantics that is
+comparable with Typed Racket's default Natural one. This Transient
 variant accommodates varied linguistic features and library calls
-in a way that is faithful to the transient
+in a way that is faithful to the Transient
 semantics~\cite{vss-popl-2017}.
@@
 Beyond the implementation issue, the method also faces the problem that
-the three variants---the natural semantics, the transient semantics, and
-the erasure semantics---come with three radically different blame
-assignment strategies.  For example, the natural semantics assigns blame
-to one component while transient assigns blame to a set of components. The
-erasure semantics does not blame a component per se, but it comes with an
+the three variants---the Natural semantics, the Transient semantics, and
+the Erasure semantics---come with three radically different blame
+assignment strategies.  For example, the Natural semantics assigns blame
+to one component while Transient assigns blame to a set of components. The
+Erasure semantics does not blame a component per se, but it comes with an
 exception location and the usual stack traces.
@@
 \emph{modes} that represent the alternative interpretations of blame.  For
-instance, a mode of the rational programmer using transient may pick the
+instance, a mode of the rational programmer using Transient may pick the
 oldest element of a blame set as the next component to add types because
@@
 When both modes are equally successful in locating an impedance mismatch,
-the method has needs to take into account the rational programmer's effort
+the method needs to take into account the rational programmer's effort
 to answer which mode is the most effective.

### 2020-11-21 writing/conclusion.tex
Lots of minor changes and rewordings

@@
 \section{What to Do Next} 
+\label{sec:conclusion}
 
@@
 confirm or reject their answers. We already know from~\citet{tgpk-dls-2018}
-that, by intuition, programmers prefer the run-time checking of natural over
+that, by intuition, programmers prefer the run-time checking of Natural over
 other soundness methods. But, just because a random set of programmers expresses
@@
 that such type mutators are much more difficult to develop than the code
-mutators of this paper. We conjecture that the natural semantics will
-usually discover such mistakes, while the transient and erasure semantics cannot
+mutators of this paper. We conjecture that the Natural semantics will
+usually discover such mistakes, while the Transient and Erasure semantics cannot
 help with such mistakes at all; indeed, we expect that these latter two raise

### 2020-11-21 writing/discussion.tex
Lots of minor changes and rewordings

@@
 \section{What We Learned}
+\label{sec:discussion}
 
@@
 %%     - TR bug
-%%   - why transient fails
+%%   - why Transient fails
 %%     - bad runtime errors
@@
 Indeed, the bottom row of figure~\ref{fig:avo-matrix} illustrates that Erasure almost never offers more useful information than any other mode.
-Nonetheless, figure~\ref{fig:effort-table} does show that even if erasure often fails to provide helpful information, it provides the rational programmer with a quick answer.
+Nonetheless, figure~\ref{fig:effort-table} does show that even if Erasure often fails to provide helpful information, it provides the rational programmer with a quick answer.
 
@@
 worst-case of 5.4x on fully-typed benchmarks in Reticulated,
-some of our configurations exceed a 4 minute timeout or 6GB memory limit with transient blame.
+some of our configurations exceed a 4 minute timeout or 6GB memory limit with Transient blame.
 To put those limits into context, the fully typed and fully untyped benchmarks all complete in a few seconds with minimal memory usage, and none of the mixed Natural configurations hit these limits.

### 2020-11-21 writing/ideas.tex
Lots of minor changes and rewordings

@@
    For example we can obtain the simple blame mode
-for the natural system if we use, as above, the selector $\blame{\cdot}$. 
-However, the same mode has a different meaning in transient as this
+for the Natural system if we use, as above, the selector $\blame{\cdot}$. 
+However, the same mode has a different meaning in Transient as this
 semantics results in multiple blamed components.
@@
 
-Of course for systems such as transient we can add modes where the rational
+Of course for systems such as Transient we can add modes where the rational
 programmer picks one blametrail as part of the migration process. For

### 2020-11-21 writing/introduction.tex
Lots of minor changes and rewordings

@@
 \section{Does Blame Matter?}
+\label{sec:introduction}
 
@@
 run-time system signals an impedance mismatch, a good blame message helps
-the programmer infer where to add an additional type specification. Since
-components that pass the type checker cannot cause impedance mismatches;
-the type checker discovers these mismatches at compile time and points
+the programmer infer where to add an additional type specification.
+Components that pass the type checker cannot cause impedance mismatches,
+so if the type checker discovers these mismatches at compile time then it points
 directly to the faulty component. Otherwise, if the
@@
 is run again to signal another violation. Then the programmer can heed
-blame one more time to add yet another type specification. Weighing the number of
-steps it takes to discover the mistake in this manner is a measure of the effectiveness
+blame one more time to add yet another type specification.
+Weighing whether we discover the mistake in this manner, and the number of
+steps it takes to do so, are measures of the effectiveness
 of the blame assignment strategy of the language.
@@
 method to three distinct blame assignment strategies and approximately
-72000 cases (section~\ref{sec:results}): Vitousek et
+722,000 cases (section~\ref{sec:results}): Vitousek et
 al.'s strategy of tracking typed/untyped boundaries in Reticulated;

### 2020-11-21 writing/landscape.tex
Lots of minor changes and rewordings

@@
 \section{How to Bring Natural, Transient and Erasure Under the Same Roof} 
+\label{sec:landscape}
 
@@
 The  program in figure~\ref{fig:example} demonstrates the differences
-between natural, transient and erasure using Typed Racket syntax.  It
+between Natural, Transient and Erasure using Typed Racket syntax.  It
 consists of four modules: the left and right ones are untyped while
@@
 
-Under natural semantics~\cite{tf-popl-2008,tfffgksst-snapl-2017}, when
+Under Natural semantics~\cite{tf-popl-2008,tfffgksst-snapl-2017}, when
 \texttt{pack} is imported in either \texttt{typed/pack-lb} or
@@
 \texttt{untyped/pack-lib}, while if the argument type check fails
-responsibility lies with \texttt{client}. The natural semantics keep track
+responsibility lies with \texttt{client}. The Natural semantics keep track
 of the responsible party for each check in the corresponding proxy. Thus,
 in our example, when the call to \texttt{typed-pack} returns, the return
-type check fails and the natural semantics blames \texttt{pack-lib}.
+type check fails and the Natural semantics blames \texttt{pack-lib}.
 
-Under the transient semantics~\cite{vss-popl-2017}, the call to \texttt{typed-pack} does not
-produce an error. In contrast to natural, transient does not rely on
+Under the Transient semantics~\cite{vss-popl-2017}, the call to \texttt{typed-pack} does not
+produce an error. In contrast to Natural, Transient does not rely on
 proxies. Instead, it rewrites typed modules to inline checks at appropriate
@@
 they check types in a  \emph{shallow} manner. In our
-example, transient's inlined checks make sure that the imported \texttt{pack} 
+example, Transient's inlined checks make sure that the imported \texttt{pack} 
 is a function and that the result of its call in \texttt{typed/crypto-pack-lib} is a list.   Exactly this
 last check fails when \texttt{client} calls to \texttt{crypto-pack}. In
-order to pinpoint the responsible party for the failed checked, transient
+order to pinpoint the responsible party for the failed checked, Transient
 keeps on the side a map from values to the boundaries between modules they
@@
 forms in the example. Since the failed check corresponds to the return
-type of \texttt{pack}, transient blames the sources of the two boundary
+type of \texttt{pack}, Transient blames the sources of the two boundary
 crosses, which in this case is \texttt{untyped/pack-lib} for both. In
-general though, transient may blame more than one parties. In fact, the
+general though, Transient may blame more than one parties. In fact, the
 theoretical work of~\citet{gfd-oopsla-2019} shows that there are programs
-where transient constructs a blame set that includes modules that are not
+where Transient constructs a blame set that includes modules that are not
 relevant to the check failure or that it even misses entirely the responsible
@@
 
-Under erasure semantics, the example does not raise an error due to the
-gradual type system; erasure performs no type checks at runtime. Thus the
+Under Erasure semantics, the example does not raise an error due to the
+gradual type system; Erasure performs no type checks at runtime. Thus the
 program may fail with an exception if \texttt{client} tries to use the
@@
 fail at all and produce an wrong result that goes undetected by the
-gradual type system .  In either case, the erasure semantics provides no
+gradual type system .  In either case, the Erasure semantics provides no
 help to the programmer for detecting the bug in our program.
@@
 basis of Racket. First, we use the Typed Racket syntax and type
-checker for all three semantics. Second, for natural we use Typed Racket as
-is while for erasure we run programs with plain Racket after erasing any
-type annotations. Third, for transient we implement a variant of Typed
+checker for all three semantics. Second, for Natural we use Typed Racket as
+is while for Erasure we run programs with plain Racket after erasing any
+type annotations. Third, for Transient we implement a variant of Typed
 Racket that follows closely the semantics from~\citet{vss-popl-2017}.
 
-The first major ingredient of transient Typed Racket is to determine the
+The first major ingredient of Transient Typed Racket is to determine the
 Racket implementation of the shallow checks that map to each type.  For
-certain types, such as {\tt String} or {\tt (Vectorof Byte)}, transient
+certain types, such as {\tt String} or {\tt (Vectorof Byte)}, Transient
 can re-use a predicate from the Racket language.  Other types require custom
@@
 
-The second major ingredient of transient Typed Racket  is the addition of
+The second major ingredient of Transient Typed Racket  is the addition of
 a pass to the compiler of Typed Racket that rewrites all typed code.
@@
 that the expected type of the result is the return type of the type of
-{\tt f}.  If a transient check ever fails, transient Typed Racket walks over 
+{\tt f}.  If a Transient check ever fails, Transient Typed Racket walks over 
 the information in the blame map and reports a set of blamed components.

### 2020-11-21 writing/mutate.tex
Put figure 2 in one column
Lots of minor changes and rewordings

@@
 \section{How to Make Lots of Mistakes} 
+\label{sec:mutate}
 
@@
 
-\begin{figure*}
-\begin{tabular}{p{2cm} | p{10cm} }
+\begin{figure}
+\begin{tabular}{p{1.75cm} | p{6cm} }
   {\bf  name} & {\bf description (author)}  \\
@@
   \label{table:benchmark-descriptions}
-\end{figure*}
+\end{figure}
 
@@
 and have been used, maintained and evolved by their authors and others over a
-number of years.  They range widely in size, complexity, purpose and
+number of years.  They range widely in the size, complexity, purpose and
 features of Typed Racket they employ.  Furthermore the benchmarks use advanced aspects
@@
 typing~\cite{tf-icfp-2010}, types for mutable and immutable data
-structures, polymorphic types, types for first-class classes and objects and types for
+structures, polymorphic types, types for first-class classes and objects, and types for
 Racket's numeric tower~\cite{stathff-padl-12}. Without loss of the diversity of the benchmarks,
@@
 \dubbed{scenarios}, of the benchmarks without 
-some  of their type annotation.  Without the annotations, the type
+some  of their type annotations.  Without the annotations, the type
 checker may not detect a mistake at compile time and, instead, the run
@@
 are for turning the benchmarks into interesting debugging scenarios.
-Indeed the mutators produce at least 72192 interesting debugging scenarios
-across all mutants --- there are potentially many more but this is how
-many we examined. In more detail, the mutators produce at least 40 interesting
+At a high level, the mutators produce 16,800 mutants with at least one interesting debugging scenario each across all of our benchmarks.
+Broken down by benchmark, the mutators produce at least 40 interesting
 scenarios for every benchmark, and these scenarios originate from at
@@
 Furthermore, every mutator contributes scenarios to at least one
-benchmark. Especially some mutators  target specific features of
-benchmarks even if they do not contribute interesting scenarios to other
-benchmarks, which
-justifies their inclusion. For
-instance the class-focused mutators are effective in \texttt{take5}
-which jives with the fact that it is the benchmark with 
-the most extensive use of object-oriented features.
+benchmark.
+That being said, some mutators only apply to a small subset of the benchmarks;
+those mutators target specific features which only those benchmarks make us of, so we include them despite their narrow applicability.
+For instance, the class-focused mutators are mainly effective in \texttt{take5}, since it is the benchmark with the most extensive use of object-oriented features.

### 2020-11-21 writing/rational.tex
almost done with results
working on last bit of results
Lots of minor changes and rewordings

@@
 \section{How to Test the Rational Programmer} 
+\label{sec:rational}
 
@@
 If the scenario type-checks, the programmers runs the program until it
-raises a run time error.  The rational programmer uses the blame
+raises a run time error.  The rational programmer uses the error's blame
 assignment\footnote{Blame can result either directly from a run time type
@@
 
-The natural semantics assigns blame to exactly one boundary.  A blame assignment
+The Natural semantics assigns blame to exactly one boundary.  A blame assignment
 has the following specific meaning: the typed component may make incorrect type
@@
 \begin{quote}
-\it A \emph{natural blame trail} is a sequence of scenarios $\conf_0,...\conf_n$ of
+\it A \emph{Natural blame trail} is a sequence of scenarios $\conf_0,...\conf_n$ of
 a program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
@@
 \begin{quote}
-\it A natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
-\emph{successful} iff its last scenario $\conf_n$ does not type check.  A natural
+\it A Natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
+\emph{successful} iff its last scenario $\conf_n$ does not type check.  A Natural
 blame trail $\conf_0,..,\conf_n$ in a lattice $\lattice{\system}$ is \emph{failing}
@@
 \end{quote}
-That is failing natural blame trails are those that have not reached a scenario that
-reveals the bug at compile time and their last scenario does not produce blame. Thus
+That is, failing Natural blame trails are those that have not reached a scenario that
+reveals the bug at compile time and their last scenario does not blame an untyped component. Thus
 the rational programmer receives no hints from the gradual type system on how to
@@
 
-While a successful natural blame trail indicates that it 
+While a successful Natural blame trail indicates that it 
 pays off to heed blame assignments during debugging the trail's root, it does not answer whether
@@
 
-To account for this situation, we devise a new mode of the natural rational
+To account for this situation, we devise a new mode of the Natural rational
 programmer that follows a migration process based on exceptions:
 \begin{quote}
-\it A natural exception trail is a sequence of scenarios $\conf_0,...\conf_n$ of a
+\it A Natural exception trail is a sequence of scenarios $\conf_0,...\conf_n$ of a
 program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
-\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i = \{\random{\system, \conf_i}\}$
+\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i = \{\exception{\system, \conf_i}\}$
 where $\exception{\system, \conf}$ denotes the location of the exception of $\conf$.
@@
 Putting the two modes together, we can now compare the usefulness of blame 
-and mere exceptions for debugging a scenario in the context of natural semantics.
+and mere exceptions for debugging a scenario in the context of Natural semantics.
 \begin{quote}
@@
   Given a program $\system$ and a root $\conf_0$ in $\lattice{\system}$,
-  natural blame is \emph{more useful} than natural exceptions for
+  Natural blame is \emph{more useful} than Natural exceptions for
   debugging $\conf_0$ iff 
-  the natural blame trail 
-  that starts at $\conf_0$ is successful while the natural exception trail that
+  the Natural blame trail 
+  that starts at $\conf_0$ is successful while the Natural exception trail that
   starts at $\conf_0$ is failing.
@@
 
-This ambiguity in transient blame raises the question of how the rational programmer
+This ambiguity in Transient blame raises the question of how the rational programmer
 should react when the language produces a blame set. Our answer is that the rational
@@
 \begin{quote}
-\it A \emph{transient-first blame trail} is a sequence of scenarios
+\it A \emph{Transient-first blame trail} is a sequence of scenarios
 $\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
@@
 \first{\mblame{\system, \conf_i}}$ where $\first{\mblame{\system, \conf}}$ is the
-first component transient adds to the blame set for $\conf$.
+first component Transient adds to the blame set for $\conf$.
 \end{quote}
@@
 \begin{quote}
-\it A \emph{transient-last blame trail} is a sequence of scenarios
+\it A \emph{Transient-last blame trail} is a sequence of scenarios
 $\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
 $\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
-  \first{\mblame{\system, \conf_i}}$ where $\last{\mblame{\system, \conf}}$ is the
-last component transient adds to the blame set for $\conf$.
+  \last{\mblame{\system, \conf_i}}$ where $\last{\mblame{\system, \conf}}$ is the
+last component Transient adds to the blame set for $\conf$.
 \end{quote}
 
-The definition of \emph{transient exception trails} 
-is analogous to that for natural and, same as for natural, we use it to  
+The definition of \emph{Transient exception trails} 
+is analogous to that for Natural and, same as for Natural, we use it to  
 specify when  
-transient-last and transient-first blame are more useful than 
-transient exceptions for debugging a scenario.
+Transient-first and Transient-last blame are more useful than 
+Transient exceptions for debugging a scenario.
 
@@
 
- Since gradually typed languages with erasure semantics do not come with
+ Since gradually typed languages with Erasure semantics do not come with
  blame assignment, a rational programmer can only hope that the underlying
- safety checks and their exceptions are helpful.  Thus, the erasure 
- rational programmer has a single mode, the erasure exception mode, and
+ safety checks and their exceptions are helpful.  Thus, the Erasure 
+ rational programmer has a single mode, the Erasure exception mode, and
  its definition and the accompanying notion of usefulness follow that for
- natural.
+ Natural.
 
@@
 \subsection{Experimental Process.}
+\label{subsec:experiment}
 
 Trails and their properties give us the tools for a rigorous examination
-of blame for natural, transient and erasure Typed Racket in the setting of the 
+of blame for Natural, Transient and Erasure Typed Racket in the setting of the 
 mutants from section~\ref{sec:mutate}. In line with the discussion so far, 
@@
 \begin{itemize}
-\item[$Q_1$] Is blame useful in the context of natural?
+\item[$Q_1$] Is blame useful in the context of Natural?
 
-\item[$Q_2$] Is first blame useful in the context of transient?
+\item[$Q_2$] Is first blame useful in the context of Transient?
 
-\item[$Q_3$] Is last blame useful in the context of transient?
+\item[$Q_3$] Is last blame useful in the context of Transient?
 
@@
 \begin{itemize}
-\item[$Q_*$] Is blame for X more useful than blame for Y where X, Y is natural,
-  erasure or natural?
+\item[$Q_*$] Is blame for X more useful than blame for Y? (Where X and Y are any of Natural,
+  Erasure or Natural?)
 \end{itemize}
@@
 Table~\ref{fig:experiment-outline} summarizes how each question relates to
-different kinds of trails/modes of the rational programmer. For example experimental
-question $Q_1$ asks whether blame is valuable for natural and our experiment
-uses the natural blame and exception trails to answer it.
+different kinds of trails/modes of the rational programmer. For example, experimental
+question $Q_1$ asks whether blame is valuable for Natural and our experiment
+uses the Natural blame and exception trails to answer it.
 
@@
 In detail we answer $Q_1$  by comparing the success of the 
-natural blame and natural exception trails for all interesting 
+Natural blame and Natural exception trails for all interesting 
 debugging scenarios from section~\ref{sec:mutate}.
@@
 Our test
-bed extends the trails that start from such roots according to the natural-blame programmer.  If no scenarios
+bed extends the trails that start from such roots according to the Natural-blame programmer.  If no scenarios
 can be added to the trail, we check if
-the last scenario of the trail type checks or not. If it does, we log natural-blame
-trail is successful, otherwise that is failing. We repeat the process for the
-same roots again but this time for the other mode, natural exceptions.
+the last scenario of the trail type-checks or not. If it does, we record that the Natural-blame
+trail is successful, and otherwise that fails. We repeat the process for the
+same roots again but for the other mode, Natural exceptions.
  Figure~\ref{fig:process}
@@
 it with the mutations from section~\ref{sec:mutate}.
-Next, based on the collected data, we calculate the
-success/failure results for the trails to determine for each root whether natural blame
-is more useful than natural exceptions. We
+After completing the experiment, we calculate the
+success/failure results of the trails to determine for each root whether Natural blame
+is more useful than Natural exceptions. We
 obtain a positive answer for
@@
 manages to debug because of blame. 
-The process is analogous for $Q_2$ and $Q_3$  employing the respective
+The process is analogous for $Q_2$ and $Q_3$, employing the respective
 modes of the rational programmer.
@@
 of the rational programmer from different semantics.
-However to answer this question we compare the percentage of scenarios
+However, answering this question requires us to compare the percentage of scenarios
 where one mode is more useful than the other and the inverse.
-For instance to
-decide whether blame for natural is more useful than first blame for transient, 
+For instance, to
+decide whether blame for Natural is more useful than first blame for Transient, 
 we compare the percentage of interesting scenarios where 
-natural blame is more successful than transient first blame 
-with the percentage of interesting scenarios where transient first blame 
-is more successful than natural blame. 
-Finally we repeat these steps one more time to compare natural blame and transient last blame
+Natural blame is more successful than Transient first blame 
+with the percentage of interesting scenarios where Transient first blame 
+is more successful than Natural blame. 
+Finally we repeat these steps one more time to compare Natural blame and Transient last blame
 and get a complete picture of the comparative usefulness of blame
@@
 \subsection{Programmer Effort}
+\label{subsec:effort}
 
@@
 three gradual typing systems. For example, consider one debugging scenario.  
-If, for instance, both  natural blame and transient first
+If, for instance, both  Natural blame and Transient first
 blame trails are successful, the two modes of the 
@@
 debugs the interesting scenario with less effort. In
-general, if the effort distribution of a mode of the rational programmer
+general, if the effort distribution for a mode of the rational programmer
 has a shorter tail and more
@@
 of the rational programmer is an artifact of pure chance or not.
-For that, we can simply compare the effort distribution of a mode of the
-rational programmer with a rational programmer that ignores blame and
-exceptions and instead selects which component to type next randomly. 
+That is, we can compare the effort distribution for a mode of the
+rational programmer with another mode that ignores error information entirely
+and instead selects which component to type next randomly. 
 Of course, since each mutant has a finite number of components, 
-such the random mode trails are always successful. At the same time though,
-the natural, transient and erasure random mode exhibit a normal effort
-distribution across scenarios that should be distinct from the effort distribution of all other modes of
-the rational programmer if the effectiveness of the latter is not
+such random mode trails are always successful. At the same time though,
+the Natural, Transient, and Erasure random modes exhibit a normal effort
+distribution across scenarios that should be distinct from the effort distribution for all other modes of
+the rational programmer if the effectiveness of the modes are not
 coincidental.

### 2020-11-21 writing/introduction.tex
Introduce transient in intro, reword "mere typos"

@@
 722,000 cases (section~\ref{sec:results}): Vitousek et
-al.'s strategy of tracking typed/untyped boundaries in Reticulated;
+al.'s strategy of tracking typed/untyped boundaries in Reticulated, called Transient;
 Tobin-Hochstadt et al.'s use of higher-order contract system~\cite{ff-icfp-2002,

### 2020-11-21 writing/landscape.tex
Fix typo

@@
 typed and the gradual type system does not affect the compilation of
-untyped components. Despite, these differences, Typed Racekt is the most
+untyped components. Despite, these differences, Typed Racket is the most
 mature system for sound mix-typed code

### 2020-11-21 writing/discussion.tex
revision of results

--- 2020-11-21 writing/discussion.tex
@@
 
-\subsection{Why does the data look like this?}
+The analysis in the previous section suggests a number of interesting
+high-level conclusions about blame in the gradual typing setting.  First,
+run time type checks have a large positive impact. This is the case
+regardless of whether these checks issue blame or throw plain exceptions.
+Second, in many cases type checks that issue blame are more helpful than
+those that do not. However, our results indicate that in other cases blame
+is not critical and thus it is worth further investigation whether
+disabling blame tracking for the sake of performance may be   an
+acceptable trade off. Third, from the three different blame strategies the
+Natural approach of Typed Racket fares better than the Transient approach
+of Reticulated Python but by a small margin. Since Natural offers complete
+and sound path-based blame while Transient incomplete but sound heap-based
+blame~\cite{gfd-oopsla-2019}, our results call for a deeper understanding
+of the relative strengths of the two models for blame.  Fourth, given that
+Transient's sound but shallow run time type checks do not seem to hamper
+debugging, a version of Typed Racket that disables some of its wrappers
+may offer an answer to the performance issues of the Natural semantics. 
+Fifth, just as for Typed Racket, the experiment points to possible
+improvements for Reticulated Python. In particular, the experimental 
+data indicates that the usefulness of transient blame does not depend on
+whether  the rational programmer follows first or the last element from a
+Transient blame set. Hence, given the performance issues of Transient we discuss further on, 
+a possible optimization is to limit the size of blame sets
+using timestamps.
 
-The figures in the last section raise a key question: why does blame fail in our dataset?
-We analyze the reasons for failure separately for each mode of the rational programmer.
+There are a number of threats to the
 
-Breaking down the reasons for failure for all scenarios in which Natural is unsuccessful reveals two causes.
-First, most of the failing scenarios begin with a runtime exception, which the rational programmer can use to type components until blame becomes available.
-In these failing cases (there are 1708), however, blame never becomes available; the exception information points to the wrong component and the trail ends in failure.
-A second, much smaller set (40 scenarios) do produce a runtime type error blaming some component, but the blame identifies a non-buggy typed component.
-After careful inspection of these scenarios, we are confident that they reveal a bug in the implementation of Typed Racket.
-
-Transient's blame failures break down in a similar way.
-Like Natural, most failures occur with unhelpful exception information without ever getting blame (1851 scenarios).
-However, Transient also has a substantial number of failures that occur because the programs hit the resource limits (timed out or ran out of memory) and had to be killed (some 770 scenarios).
-Additionally, there are nearly a thousand cases where Transient's checks catch a problem, but the blame set reported is empty.
-We discuss each of these last two failure categories in detail in~\ref{sec:threat:transient}.
-
-\subsection{High level conclusions}
-
-The analysis in the last section indicates a number of interesting conclusions about blame in the gradual typing setting.
-First, figure~\ref{fig:avo-matrix} makes it clear that any kind of dynamic checking has a large positive impact over Erasure, and this is true regardless of whether the rational programmer uses blame or stacktraces.
-Indeed, the bottom row of figure~\ref{fig:avo-matrix} illustrates that Erasure almost never offers more useful information than any other mode.
-Nonetheless, figure~\ref{fig:effort-table} does show that even if erasure often fails to provide helpful information, it provides the rational programmer with a quick answer.
-
-Figure~\ref{fig:avo-matrix} also indicates that blame helps significantly over just using exception information.
-Both Natural and Transient's blame modes improve over their corresponding exception modes by about ten percent.
-This illustrates that blame tracking offers value for locating the source of mistakes beyond the value provided by the dynamic checks alone.
-At only ten percent, however, the amount of added value suggests that foregoing blame tracking in order to obtain performance gains may not hurt debugging all too much.
-
-Finally, the results show that while Natural clearly offers the most utility compared to the other modes, Transient is not too far behind.
-Natural is more useful than every other mode in figure~\ref{fig:avo-matrix}, and by a larger proportion than any other mode.
-That said, the same can be said of Transient's two blame modes as compared to every other mode but Natural.
-Furthermore, the difference in usefulness between Natural and Transient's blame modes is quite small across the board.
-As we point out in~\ref{sec:mutate}, however, we should note that our selection of scenarios for analysis introduces bias toward Erasure and Transient.
-
-\subsection{Threats to validity}
+\subsection{Threat: Is the Rational Programmer Realistic?}
 
@@
 \label{sec:threat:transient}
-%% purpose: talk about transient-blame implementation .... what did we do,
-%%  how did we test, what threats remain?
-%%
-%% the benchmark timeouts look like a threat, but this section shows that
-%%  we've looked carefully at those
 

### 2020-11-21 writing/results.tex
Lots of minor changes and rewordings
revision of results

@@
 \section{What We Discovered}
+\label{sec:results}
 
@@
 memory limit of 6GB. The complete experiment required
-approximately 270 hours of compute hours on our machine.
+over 30,000 hours of compute hours on our machine.
 
@@
 can even happen immediately, which explains failures with 0 effort.
-While the most
-failures to debug scenarios follow this pattern, a few do not. The next section takes a
-closer look at these unexpected failures of the rational programmer as part of the lessons we
-extract from our experimental results.
+
+While most failures to debug scenarios follow the above pattern, a few do not.
+Breaking down the reasons for failure for Natural blame (1748 in total)
+reveals an additional cause. For a small set of
+debugging scenarios (40), Natural produces a run time type error
+blaming a non-buggy already typed
+component. We tracked down all these cases to known open issues with Typed
+Racket and class contracts.
+
+In Transient, similar to Natural,
+most failures are due to unhelpful exception information (1851 for both
+Transient first and last blame).  
+However, Transient also has a substantial
+number of failures because scenarios hit the time and memory
+limits of our experiment (~ 770 scenarios).  Additionally, there are nearly a 1000 cases where
+Transient reports an empty blame set which leaves the rational programmer
+without hints about how to proceed.
+In~\ref{sec:threat:transient}, we discuss how both of these causes of
+failure for transient affect our experiment.

### 2020-11-21 writing/introduction.tex
Minor rewordings in intro and challenges

--- 2020-11-21 writing/introduction.tex
@@
 722,000 cases (section~\ref{sec:results}): Vitousek et
-al.'s strategy of tracking typed/untyped boundaries in Reticulated;
+al.'s strategy of tracking typed/untyped boundaries in Reticulated, called Transient;
 Tobin-Hochstadt et al.'s use of higher-order contract system~\cite{ff-icfp-2002,
-mf-toplas-2009}; and the approach of industrial systems that forego blame
+mf-toplas-2009}, called Natural; and the approach of industrial systems that forego blame
 and instead rely on error messages from the
-underlying programming language. A direct interpretation of the results
+underlying programming language, called Erasure. A direct interpretation of the results
 (section~\ref{sec:discussion}) is that
@@
 strategy helps the search for impedance mismatches between the specified types
-and the behavior of untyped components. Even better, the evaluation method 
+and the behavior of untyped components.
+Furthermore, the evaluation method
 identifies the wrapper-based blame tracking of Typed Racket as the most
-useful academic system with that of Reticulated Python only marginally  behind.
+useful approach, with that of Reticulated Python close behind.
 
 Next, the paper describes the challenges of developing an evaluation
-method, which may explain why the question has been neglected for such
+method for blame, which may explain why the question has been neglected for such
 a long time.

### 2020-11-21 writing/discussion.tex
revision of discussion

@@
 
-There are a number of threats to the
+There are a number of threats to the validity of our conclusions that we
+describe in the previous section our attempts to mitigate: (i) the
+representativeness of benchmarks; (ii) the relation between mutations and
+real programming mistakes; (iii) the definition of interesting debugging
+scenarios; and (iv) our approach to sampling scenarios. In the remainder
+of this section, we discuss extensively two additional threats.
 
@@
 
-Surprisingly, Transient blames an empty set of boundaries in 967 cases.
+As we discuss in section~\ref{sec:results}, surprisingly Transient blames
+an empty set of boundaries in 967 scenarios.
 Such cases should never occur, at least in theory, because an empty set
@@
  in the current module, then we have typed code blaming itself for a typed
- value; thus, the type checker must be unsound.
+ value as all the checks are inlined in typed code. Thus, the type checker must be unsound.
 After careful investigation of these empty blame cases, we found no soundness
- bugs in Typed Racket.
+ bugs in Transient Typed Racket.
 Instead, we found scenarios in which Transient lost track of the proper
  boundaries.
-These scenarios suggest ways to improve the blame algorithm from~\citet{vss-popl-2017}:
+These scenarios suggest ways to improve~\citet{vss-popl-2017}'s blame
+algorithm:
 \begin{itemize}
@@
 \end{itemize}
-\noindent{}Despite these known limitations, we conjecture that these improvements
- to Transient blame-tracking will not affect our overall conclusions.
+\noindent{}Despite these known limitations, based on the relatively small
+number of debugging failures, we conjecture that these improvements
+ to Transient blame-tracking do not affect our overall conclusions.
 

### 2020-11-21 writing/results.tex
Small changes to sec 7

@@
 the left side of the plot, meaning that in most cases the programmer needs
-to type a single component to debug a scenario. Even in the minority of
-cases where a blame or exception mode has to work harder, the number of
-modules that that need to be typed is small relative to the random mode.
+to type a single component to debug a scenario.
 
@@
 In~\ref{sec:threat:transient}, we discuss how both of these causes of
-failure for transient affect our experiment.
+failure for Transient affect our experiment.

### 2020-11-21 writing/results.tex
results: reword to save space

@@
 memory limit of 6GB. The complete experiment required
-over 30,000 hours of compute hours on our machine.
+over 30,000 hours of compute hours.
 
@@
 of interesting debugging scenarios for which $R$ is more useful $C$ ---
-this percentage has been calculated as an estimated proportion from the
-corresponding proportion of the debugging scenarios in our sample and the
-upper bound.  In effect, reading across a given row $R$ provides an
+we calculate this percentage as an estimated proportion from our sample.
+In effect, reading across a given row $R$ provides an
 overview of how $R$ compares to every other mode; the higher the bars, the

### 2020-11-21 writing/discussion.tex
Rewording and grammar fixes in discussion

@@
 disabling blame tracking for the sake of performance may be   an
-acceptable trade off. Third, from the three different blame strategies the
+acceptable trade off. Third, from the three different blame strategies, the
 Natural approach of Typed Racket fares better than the Transient approach
-of Reticulated Python but by a small margin. Since Natural offers complete
-and sound path-based blame while Transient incomplete but sound heap-based
+of Reticulated Python, but only by a small margin. Since Natural offers complete
+and sound path-based blame while Transient offers incomplete but sound heap-based
 blame~\cite{gfd-oopsla-2019}, our results call for a deeper understanding
@@
 
-There are a number of threats to the validity of our conclusions that we
-describe in the previous section our attempts to mitigate: (i) the
+There are a number of threats to the validity of our conclusions: (i) the
 representativeness of benchmarks; (ii) the relation between mutations and
 real programming mistakes; (iii) the definition of interesting debugging
-scenarios; and (iv) our approach to sampling scenarios. In the remainder
-of this section, we discuss extensively two additional threats.
+scenarios; and (iv) our approach to sampling scenarios. The previous sections
+have described the various ways we attempt to mitigate these threats;
+we discuss two additional threats in the remainder of this section.
 
@@
 
-As we discuss in section~\ref{sec:results}, surprisingly Transient blames
+As we discuss in section~\ref{sec:results}, Transient surprisingly blames
 an empty set of boundaries in 967 scenarios.
@@
 worst-case of 5.4x on fully-typed benchmarks in Reticulated,
-some of our configurations exceed a 4 minute timeout or 6GB memory limit with transient blame.
+some of our configurations exceed our 4 minute timeout or 6GB memory limit with transient blame.
 To put those limits into context, the fully typed and fully untyped benchmarks all complete in a few seconds with minimal memory usage, and none of the mixed Natural configurations hit these limits.
-Unfortunately, our high cost is closer to the truth;
+Unfortunately, our high cost for Transient seems closer to the truth;
  there are at least three broad issues that skew the earlier results.
@@
 On larger programs, Reticulated suffers from high overhead.
-For example, a Reticulated version of our \texttt{sieve} benchmark runs in
+For example, a Reticulated version of our simplest benchmark runs in
  ~40 seconds without blame, and with blame times out after 10 minutes.

### 2020-11-21 writing/related.tex
Lots of minor changes and rewordings
edits in related work

@@
 \section{Related Work}
+\label{sec:related}
 
-This paper builds on~\citet{lksfd-popl-2020}'s one-of-a-kind analysis of
+This paper builds on~\citet{lksfd-popl-2020}'s human-out-of-the-loop
+empirical analysis of
 blame for higher-order contracts. They do not spell out the notion of the
 rational programmer but the basic ingredients of our technique derive
-directly from theirs. Specifically, we adapt their empirical technique
-to gradual typing which requires the development of custom mutators,
+directly from theirs. Specifically, we revisit their  technique
+to bring it to the world of gradual typing which requires the development of custom mutators,
 careful selection of debugging scenarios and generalization of the

### 2020-11-21 writing/introduction.tex
Fix missing section label, number of scenarios

@@
 method to three distinct blame assignment strategies and approximately
-722,000 cases (section~\ref{sec:results}): Vitousek et
+72,200 cases (section~\ref{sec:results}): Vitousek et
 al.'s strategy of tracking typed/untyped boundaries in Reticulated, called Transient;

### 2020-11-21 writing/discussion.tex
Fix missing section label, number of scenarios
typos + small edits

@@
 debugging, a version of Typed Racket that disables some of its wrappers
-may offer an answer to the performance issues of the Natural semantics. 
+may offer an answer to the performance issues of Natural. 
 Fifth, just as for Typed Racket, the experiment points to possible
@@
 whether  the rational programmer follows first or the last element from a
-Transient blame set. Hence, given the performance issues of Transient we discuss further on, 
+blame set. Hence, given the performance issues of Transient we discuss further on, 
 a possible optimization is to limit the size of blame sets
@@
 
-As we discuss in section~\ref{sec:results}, Transient surprisingly blames
-an empty set of boundaries in 967 scenarios.
+As we discuss in section~\ref{sec:results}, Transient surprisingly
+produces
+an empty blame set in 967 scenarios.
 Such cases should never occur, at least in theory, because an empty set
@@
 
-A second apparent flaw in our Transient semantics is the high cost of blame.
+A second apparent flaw in our Transient implementation is the high cost of blame.
 Whereas \citet{vss-popl-2017} report an average slowdown of 2.5x and

### 2020-11-21 writing/landscape.tex
typos + small edits

@@
 also imports \texttt{pack} and assumes for it the same type as
-\texttt{typed/pack-lb} but instead of exporting it directly it wraps in
+\texttt{typed/pack-lb} but instead of exporting it directly \texttt{typed/pack-lb} wraps it in
 another function, \texttt{crypto-pack}, that encrypts its input before
@@
 function and wraps it in a proxy that enforces its  type.  Analogously
-the run time wraps the exports of the two typed modules in proxies that
+the runtime wraps the exports of the two typed modules in proxies that
 check that untyped code applies \texttt{typed-pack} and
-\texttt{crypto-pack} to arguments of its type. In essence, the
+\texttt{crypto-pack} to arguments of the expected type. In essence, the
 return type checks protect  the typed modules in case they use
@@
 result of \texttt{typed-pack} or \texttt{crypto-pack}, but it may also not
-fail at all and produce an wrong result that goes undetected by the
+fail at all and produce a wrong result that goes undetected by the
 gradual type system .  In either case, the Erasure semantics provides no

### 2020-11-21 writing/rational.tex
Fix another typo
typos + small edits

@@
 reveals the bug at compile time and their last scenario does not blame an untyped component. Thus
-the rational programmer receives no hints from the gradual type system on how to
+the rational programmer gets no hints from the gradual type system on how to
 continue the search for the bug.
@@
 mismatch may have crossed the boundaries between elements in the set, and each crossing
-checked only a portion of the value's type shape.
+checked the value's type in a shallow manner.
 
@@
 \begin{itemize}
-\item[$Q_*$] Is blame for X more useful than blame for Y? (Where X and Y are any of Natural,
-  Erasure or Natural?)
+\item[$Q_*$] Is blame for X more useful than blame for Y? (Where X and Y are any of Natural, Transient, or Erasure)
 \end{itemize}

### 2020-11-21 writing/landscape.tex
Fix typos

@@
 that consumes JSON and returns an association list that maps symbols to
-strings. The second typed module, \texttt{typed/crypto-pack-lb},
+strings. The second typed module, \texttt{typed/crypto-pack-lib},
 also imports \texttt{pack} and assumes for it the same type as
-\texttt{typed/pack-lb} but instead of exporting it directly \texttt{typed/pack-lb} wraps it in
+\texttt{typed/pack-lib} but instead of exporting it directly \texttt{typed/pack-lib} wraps it in
 another function, \texttt{crypto-pack}, that encrypts its input before
@@
 
-Under Natural semantics~\cite{tf-popl-2008,tfffgksst-snapl-2017}, when
-\texttt{pack} is imported in either \texttt{typed/pack-lb} or
-\texttt{typed/secure-pack-lb}, the runtime checks that \texttt{pack} is a
+Under the Natural semantics~\cite{tf-popl-2008,tfffgksst-snapl-2017}, when
+\texttt{pack} is imported in either \texttt{typed/pack-lib} or
+\texttt{typed/secure-pack-lib}, the runtime checks that \texttt{pack} is a
 function and wraps it in a proxy that enforces its  type.  Analogously
@@
 last check fails when \texttt{client} calls to \texttt{crypto-pack}. In
-order to pinpoint the responsible party for the failed checked, Transient
+order to pinpoint the responsible party for the failed check, Transient
 keeps on the side a map from values to the boundaries between modules they
@@
 crosses, which in this case is \texttt{untyped/pack-lib} for both. In
-general though, Transient may blame more than one parties. In fact, the
+general though, Transient may blame more than one party. In fact, the
 theoretical work of~\citet{gfd-oopsla-2019} shows that there are programs
@@
 
-Under Erasure semantics, the example does not raise an error due to the
+Under the Erasure semantics, the example does not raise an error due to the
 gradual type system; Erasure performs no type checks at runtime. Thus the

### 2020-11-21 writing/rational.tex
Fix typos

@@
 
-Putting the two modes together, we can now compare the usefulness of blame 
-and mere exceptions for debugging a scenario in the context of Natural semantics.
+Same as for success for a Natural blame trail, we define success for a
+Natural exception trail.
+Putting the definitions for the two modes together, we can now compare the usefulness of blame 
+with that for mere exceptions for debugging a scenario in the context of Natural semantics.
 \begin{quote}
@@
  rational programmer has a single mode, the Erasure exception mode, and
- its definition and the accompanying notion of usefulness follow that for
- Natural.
+ its definition follows that for
+ the Natural exception mode.
 
@@
 the last scenario of the trail type-checks or not. If it does, we record that the Natural-blame
-trail is successful, and otherwise that fails. We repeat the process for the
+trail is successful, and otherwise that it fails. We repeat the process for the
 same roots again but for the other mode, Natural exceptions.

### 2021-02-09 writing/reviews/pldi-rebuttal-notes.md
Move & change review to .md, start on rebuttal notes

Response notes format: indented after ⟶ or TODO.
⟶ is a drafted response, TODO records notes about what to respond to or how.

PLDI 2021 Paper #66 Reviews and Comments
===========================================================================
Paper #66 How to Evaluate Blame for Gradual Types

Review #66A
===========================================================================

Overall merit
-------------
A. I will champion accepting this paper.

Reviewer expertise
------------------
Z. **Some familiarity** = "I have a passing knowledge of the topic(s) but
   do not follow the relevant literature"

Paper summary
-------------
This paper presents a methodology for evaluating blame assignment strategies
based on the notion of "the rational programmer". The paper develops the
methodology in detail and then then uses it to evaluate three different blame
strategies: "Transient", "Natural" and "Erasure".

In gradual typing, a blame assignment is used to show where the program went
wrong. Importantly, at least on the theory side, a well-typed component cannot
be blamed. The authors argue that such blame assignment has received a lot of
academic attention, yet industrial implementations of programming languages with
gradual typing completely forego blame. Why is this? Is it because blame is not
useful in practice? The paper aims to answer this question by applying the
developed methodology to three strategies: The "Natural" blame strategy relies
on proxies to perform additional type checks at run-time. The "Transient" blame
strategy inlines these checks and does not perform "deep" type checks. The
"Erasure" blame "strategy" performs no checks: it only fails if execution of the
program fails.

The contribution of the paper is the methodology for evaluating blame (and
secondary its application to the three strategies.) The idea is to design an
algorithm that behaves as a programmer would debug a gradually typed program
that crashes at run-time with a type error: by incrementally adding types to the
untyped parts of the program until the source of the type error is revealed. In
this way, the question is: Does the blame assignment (or lack thereof in the
case of "Erasure") help guide the programmer? How many "steps" must he take to
find the defect and can he find it at all?

Trying to summarize the methodology:

- Take a collection of fully typed programs.
- Consider mix-typed versions of those programs.
- Develop mutators (like in mutation testing) that inject type errors.
- Emulate the rational programmer who follows the "trail" in the spirit of
  Figure 5.
  - At each step the programmer decides on an untyped component and then types
    it, reruns the program, and either sees a type error or decides to type
    another untyped component.
- Take into account how the "next untyped component" is selected based on the
  specific system at hand. 
- Experimentally evaluate whether the above method is able to find the fault and
  through how many steps.

Comments for author
-------------------
This paper is exceptionally well-written and was a joy to read. Every time I had
a concern the paper seemed to anticipate it and provide useful commentary.

The paper does a very good job at describing the challenges of how to evaluate
blame assignment strategies and then sets out to develop an automated evaluation
methodology based on the notion of "the rational programmer". The paper
carefully lays out the foundation for the generation of buggy mix-type programs,
the actions and variants ("modes") of the rational programmer which can depend
on the specific blame system, and finally how to sample the "experimental
space". The paper also includes a clear section on threats to validity.

I think it is important to have a methodology to evaluate the usefulness of
blame, and in lieu of large-scale and costly user studies, I think this paper
provides a very well thought out automatic approach (with the natural
limitations that such an approach must inevitably have.)

Points for:

- Addresses a problem that has not been addressed before.
- Clear, concise, and meaningful methodology.
- Clear description of the threats to validity.
- Very well-written paper.

Points against:
- none

# Questions

- Could you comment a bit more on what makes a mutator "good"? (other than
  experimentally seeing that it is good.) Perhaps, what would be a really bad
  mutator for evaluating blame assignment?

  ⟶ Abstractly, a good mutator reliably generates bugs that are detectable by the type system and by all three semantics' runtime type checks; ideally the bugs also manifest during the interaction of multiple components.
    We rely on the experimental check of our mutators since this criteria is difficult or impossible to decide otherwise.
    That said, we can reason that some mutators are clearly unlikely or impossible to be good.
	...
	We will extend the discussion at the start of section 4 to clarify this point.

  TODO: Rewrite response along these lines...
    A common mutator in the literature is swapping constants of the same type, but this is not a good mutator for us because type system can't catch it.
	Example of an obviously bad mutator: replacing any string with some pre-defined non-empty string (e.g. "foobar") would be a bad mutator.
	Examples of mutators we cut: `wrap-cond` (tricks occurrence typing by hiding the result of conditionals to produce a type error, but never a dynamic error), `add-extra-method`, swapping `car` and `cdr`, `set!`ing an ID to itself (only causes a type error), swapping mutable and immutable hashes+vectors.
	# This answer should serve to justify our mutators and the challenge of making good mutators more broadly, to answer reviewer D's concerns about the delta of making mutators.

- I don't understand where the 16,800 number comes from. Is it an exhaustive
  enumeration of all possible mutations? (Or the number of mutants constructed
  in X amount of time?)
  ⟶ The former, we will clarify in the text.

- How large are the programs in the benchmark suite?
  ⟶ The program sizes without type annotations, as reported by Racket's `syntax-sloc` tool, are below. We will add this information to the table in Figure 2.
    benchmark | LOC | components
	----------|-----|-----------
	acquire | 1941 | 9
	gregor | 2336 | 13
	kcfa | 328 | 7
	quadT | 7396 | 14
	quadU | 7282 | 14
	snake | 182 | 8
	suffixtree | 1500 | 6
	synth | 871 | 10
	take5 | 465 | 8
	tetris | 280 | 9

# Minor Comments

- L38: Should there have been a colon? Is the text "This contrast... " and
  challenge quote from [3]? Who is saying what here...
  ⟶ Thanks, we will reword the text from "This contrast" to clarify that this our research question, it is not coming from [3].

- L104: I was confused by the word "Next". Maybe instead: We now ..."
  ⟶ Thanks, we will change the wording.

- L268: Grammar issue.
  ⟶ Thanks, we will fix this.

- L314: Floating period.
  ⟶ Thanks, we will fix this.

- L441: Add line numbers (or other metric of size).
  ⟶ See above, we will add this information.

- L956: 30,000 hours of compute time is 1,250 days. I assume that the CPU has
  many cores. Could you put that number?
  ⟶ Each CPU has 28 cores and 56 threads; when we last ran the experiment, it took about two weeks of real time from start to end. 
  We will add this information at the start of section 7.

- L980: Figure 5.4??
  ⟶ Thanks, this is a bad reference; it should be figure 6. We will fix it.

- L991: Figure 6: It took me some time to understand the figure. I wonder if
  there is a better way to represent the same data?
  TODO: come up with some thoughts for this figure
  ⟶ Thanks, we can see how it's difficult to read. We will consider some different presentations, such as...
    - a bunch of two-sided bar charts? Where there's one bar pair per pair of modes. Above the x-axis is % A > B, and below the axis is % B > A?
	- a bunch of stacked histograms? Where there's one stack per pair of modes. A bar has three parts: A > B, A = B, and B > A.

Review #66B
===========================================================================

Overall merit
-------------
B. I support accepting this paper but will not champion it.

Reviewer expertise
------------------
Y. **Knowledgeable** = "I follow the literature on this paper's topic(s)
   but may have missed some relevant developments"

Paper summary
-------------
The paper proposes an evaluation method for blame assignment
strategies and evaluates three different strategies with the
method. While many theoretical gradual typing systems have adopted
blame, blame has been neither shown to be practically useful nor
applied to real-world languages because the systematic evaluation of blame
is difficult. The authors introduce a notion of rational programmers
and simulate their behaviors to evaluate blame. A rational programmer
is assumed to iteratively add type annotations to an untyped portion
of his or her program according to the feedback given by a run-time
system. This behavior can be simulated even without real human
programmers. If a run-time system leads a programmer to the correct
locations of a bug, its blame strategy can be considered more useful
than a strategy that cannot do so. The paper compares the Natural,
Transient, and Erasure strategies. The result shows that both Natural
and Transient outperform Erasure, which does not assign any blame,
and, therefore, implies usefulness of blame assignments.

  TODO: This seems to imply that we are proposing or assuming the way that real "rational programmers" would behave. Should we try to answer or clarify about that?

Comments for author
-------------------
This paper tackles a long-standing difficult problem, which is the
problem of evaluating blame. Many existing gradual typing papers
provide blame systems, but because the papers do not evaluate blame
systems, it is not obvious whether blame itself is indeed helpful for
debugging. Among many theoretical gradual type systems adopting blame
assignments, no paper discussed the evaluation of blame systems. It is an
important problem because, without any evaluation, theoretical aspects
like gradual guarantee and the blame theorem may not be applicable to
practical systems combining static and dynamic typing.

This paper shows that blame assignments do help programmers, and it
may serve as a first step to close the gap between the theory and
the practice of gradual typing.

The experiment results show that there are debugging scenarios where
Transient is more useful than Natural. Does it happen simply because
of luck? Or, does it imply that Transient is more suitable than
Natural for a particular kind of bugs for some reason? Also, can there
be a new blame strategy that outperforms both Transient and Natural
all the time?

  ⟶ 
	Response idea: This is an example where the experiment seems to contradict the theory. 
	The models don't include all of the features in the benchmarks. <enumerate them here> It's possible that the Natural, Transient theories don't scale as expected to rich PL features.
	Make analogy to physics: theory makes predictions, we have found failures of the predictions. We have two courses of action: report the inconsistency to re-eval the theory, and do more experiments.
	It's also possible we have a bug in TR, or in Transient implementation.

    Later: TODO: Scenarios where Transient is more useful than Natural imply that Transient locates the bug while Natural gets stuck.
	      check what the reasons for those scenarios are;
	      - the % is too high to all be because of the classes bug,
		  - how else could Natural get stuck if Transient isn't? It can't be the bad exn cases, because then Transient should be stuck as well because Transient shouldn't be able to get blame in those cases.

Theoretical gradual type systems usually allow programmers to control
the precision of type annotations in a fine-grained way with the Dyn
type rather than distinguishing typed and untyped modules. Would the
same evaluation method be able to directly appliable to blame in such systems?

  TODO: The answer is yes, but this needs more thought to frame the answer the right way.
  ⟶ Our evaluation framework should be directly applicable to blame in gradual typing systems using Dyn.
    The details of how it applies depends on the granularity of the blame information offered by the system -- in other words, what the system considers components.
	If components are modules, as in Typed Racket, then nothing changes.
	Systems with Dyn more likely treat top-level definitions as components, in which case definitions are individually typed (the whole definition is annotated) rather than modules.

Please explain what impedance mismatches are in the paper. Is it
simply type mismatches?

  ⟶ Impedance mismatches are mismatches between a piece of code's true type and its annotation.
    For instance, the code `(λ (x) (+ x 42))` has type `Number -> Number` but might be annotated as `Number -> String`, causing an impedance mismatch.
    We will adjust the text to clarify or remove this terminology.

# Strengths

* The paper is well-organized.
* It proposes a novel solution to a long-standing difficult problem.
* Threats to validity are discussed in detail.

# Minor comments

The paper states that the effort distribution of the random mode
follows a normal distribution, as expected. Please explain why a
normal distribution is a reasonable expectation. When there are $n$
untyped components, the random mode randomly selects one of them, and
only one of them is a buggy component. Therefore, in the first
iteration, the random mode reaches the end of the trail with the
probability of $1/n$. In the second iteration, there are $(n-1)$
untyped components and the probability of a failure at the first
iteration is $(n-1)/n$, so the probability of effort $2$ is $1/n$
again. In this way, it seems to be that the probability of effort $i$
is $1/n$ where $1<=i<=n$ and $n$ is the number of initially untyped
components. Each trail has a different number of untyped components,
so the resulting distribution would be the average of such
distributions. Based on this reasoning, it is unclear why the
distribution is a normal distribution. This point does not affect the
contribution of the paper, but it is not clear from the paper.

  TODO: response sketch: you're right, we misspoke, this is not the normal distribution; our reasoning for why the normal-ish shape of distribution we see makes sense is that given a fixed trail the probability at each step of the trail ending is 1/n, but we randomly select trails so we're multiplying the random distribution of trail lengths by that 1/n one? Not clear on the details of this reasoning.
  ⟶

* page 9 line 980: Figure 5.4 -> Figure 6
  ⟶ Thanks, we will fix this.
* page 10 Figure 7: Null -> Random
  ⟶ Thanks, we will fix this.
* page 11 line 1103: ( 770 -> (770
  ⟶ Thanks, we will fix this.

Review #66C
===========================================================================

Overall merit
-------------
C. I would not accept this paper but will not argue strongly against
   accepting it.

Reviewer expertise
------------------
Z. **Some familiarity** = "I have a passing knowledge of the topic(s) but
   do not follow the relevant literature"

Paper summary
-------------
This paper develops an approach for empirically evaluating the effectiveness of blame assignment in gradually typed languages. Roughly speaking, the idea is to model the steps taken by a "rational programmer" who systematically debugs errors in a gradually typed programming. The question is whether different blame assignment strategies are able to help this rational programmer find the "true" source of errors faster, requiring fewer steps.

To evaluate this approach, the paper also presents empirical results. Starting with a small corpus of programs, the authors use mutations to systematically (but synthetically) introduce type errors -- e.g., one mutation replaces a constants with another of a different type, and another mutation swaps method identifiers. They then conduct experiments to quantify the number of steps required by a rational programmer for several blame assignment schemes: Natural (used in Typed Racket, which tracks at most one location for blame), Transient (used in Reticulated Python, which tracks a set of locations), and Erasure (which does not actually assign blame, and is widely used in industry).

  TODO: do we need to clarify that effort is a secondary aspect of the evaluation, and more important is whether blame leads to the bug at all?

Unsurprisingly, the evaluation shows that blame is effective, when used by a rational programmer. The results are inconclusive, but some trends are clear. Most significant, all of the schemes that assign blame are more useful than Erasure. And Natural blame more often more useful than Transient blame.

Comments for author
-------------------
This is a neat paper that develops an approach for evaluating gradual typing and blame assignment. Thanks for submitting it to PLDI '21! The results are presented in an extremely clear way that draws out the key insights without compromising on technical detail. The notion of a rational programmer -- loosely modeled on rational agents from economic theory -- is a neat conceptual device. And the empirical approach does seem like the right way to address the question of the usefulness of blame.

But while there is a lot to like about the paper, it also has some shortcomings. The suite of programs used for the evaluation is fairly small (c.f., the SIGPLAN checklist).
  TODO: respond to this ^, saying that we use the standard suite of benchmarks for evaluating GT -- point to Ben's POPL paper and others, and we use all of the largest benchmarks in the suite
The faults are introduced using synthetic mutations that may or may not correspond to the kinds of errors that arise in practice. This can be seen from the results in Figure 7 -- the lengths of the "trails" is quite small.
  TODO: unclear that syntheticness relates to trail length at all, and synthetic bugs are the only option here -- there is no corpus of documented GT bugs in the wild anywhere. Indeed, most type errors never even make it into code repositories. Note that we are the first to evaluate blame in this setting.
While the study is mostly well done, it only considers two blame assignment strategies. For example, it would be interesting to consider the "omniscient programmer" to get a bound on the value of blame. But this is not done.
  TODO: We don't understand what is meant by "omniscient programmer", but our best understanding is the ideal oracle strategy that always picks the right module right away and thus always finds the bug in one step. We didn't show this because it's trivial. We do have other baselines to understand blame's performance: the random mode, and the exceptions modes for each of the semantics.
And unfortunately the results are inconclusive -- Section 8 says "our results call for a deeper understanding of the two models for blame".
  TODO: This is what makes our results surprising: they challenge the theoretical conclusions that Natural is superior to Transient, suggesting that the theory is not good enough to understand the pragmatics of blame in reality.
Also, the published Transient scheme seems to have a minor bug, which might affect the outcomes when fixed (Section 8.2).
  TODO: 8.2 does not indicate a bug in Transient, but rather a defect in the design of the system. ... something about a gap between Reticulated Python and 
Most if not all of these shortcomings are discussed in the paper, which is much appreciated. But they do risk undermine the value of the contributions to some degree.

To explain one comment above: I was surprised not to see a comparison against an "omniscient programmer" to establish a baseline (c.f., the SIGPLAN checklist). That is, model a blame assignment scheme in which blame could be placed on any component. And always pick the component that minimizes the length of the trail. Of course, this would be infeasible to implement in practice, but it would place an upper bound on the utility of blame.

Similarly, I wondered why the modes of the transient programmer only consider picking the first/last blame, and not other choices. Section 5.2 just says "our answer" is that there are at least "two reasonable options." Why are other choices not reasonable?
  TODO: Answer suggested by Ben: it's unspecified how one should interpret Transient blame, and there isn't time to evaluate all possible choices, so we pick the most reasonable interpretations.

Review #66D
===========================================================================

Overall merit
-------------
B. I support accepting this paper but will not champion it.

Reviewer expertise
------------------
Y. **Knowledgeable** = "I follow the literature on this paper's topic(s)
   but may have missed some relevant developments"

Paper summary
-------------
Several contract and gradual type systems provide mechanisms for
assigning blame: When encountering a run-time error signifying the
violation of a contract or type signature, which pieces of code should
be scrutinized by the programmer in seeking a fix?

In a recent POPL 2020 paper, Lazarek et al. provide the first attempt
to evaluate if and how blame assignment may help in actual debugging.
In that paper, bugs are systematically injected (via syntactic
mutation) into a suite of Racket programs, and the corresponding blame
trails are analyzed to see whether they, by writing more precise
contracts to "shift" blame from the assigned component, eventually
lead to the actual bug.

The current paper extends the POPL 2020 methodology from contracts in
Racket to gradual (or migratory) typing in Typed Racket, in several
steps:

First, to complement the "Natural" higher-order contract system of
(Typed) Racket, this work implements a "Transient" blame assignment
strategy, which unlike Natural inserts only shallow run-time checks.
(To date, Transient has appeared only in the separate design and
implementation of Reticulated Python, which prevents a head-to-head
comparison as desired in this work.)

Second, compared to POPL 2020, new mutators are defined to inject the
kinds of type-mismatch bugs that may be caught by the gradual type
system and corresponding run-time checks and blame assignment.

Third, the paper runs a large experiment on 10 benchmark programs,
inserting mutations and analyzing whether the blame trails produced by
each debugging mode --- following (a) the component blamed by Natural,
(b/c) the first or last component, respectively, blamed by Transient,
or (d) an exception raised by the underyling language --- conclude
with the actual bug. Results show that Natural, Transient-First, and
Transient-Last all produce short successful blame trails (typically of
length one or two) on most debugging tasks, and that they are all
similarly effective with no clear winner for all tasks.

Comments for author
-------------------
This work advances a compelling approach to evaluating a usability
question through a systematic exploration of synthetic bugs and fixes.
The large-scale experiment is impressive, and I find both of the main
results surprising:

1. That Natural is not much more effective than Transient. I thought
the shallow checks of Transient would certainly be less effective than
the full checks of Natural. It will be interesting to see if and how
this holds up in subsequent usability studies, both with synthetic and
real users and tasks.

2. That Transient --- as implemented in Typed Racket --- is much
slower than Natural. This also goes against intuition, as well as
previous results reported about Reticulated Python (but apparently
there were flaws in those experiments, as noted in this paper). Given
that the experiments are developed and run in the context of the
large, real-world Racket ecosystem, I wonder whether about the chance
of incidental implementation choices or bugs affecting this result.
  TODO: We should clarify that checks alone is faster, but adding blame is slower (bc blame map is linear).

Overall, I think this is a useful reference for further usability work
on contracts and gradual type systems, and am generally in favor of
acceptance.

At the same time, however, I believe the paper overstates the novelty
and challenges to "bring [the POPL 2020 approach] to the world of
gradual typing" (L1252). Comments about each of the three main
components of the methodology in turn below.

*Natural and Transient Under the Same Roof (Section 3)*

This seems to be "only" an engineering challenge.

And the presentation in this section is rather informal. As someone
with passable knowledge of Natural and Transient, I would have liked
to see a concise formalism with clear "knobs" for choosing between
blame assignment strategies. Trying to imagine readers with less
familiarity, I wonder how well this discussion serves as a primer.

  TODO: this is a minor part of the paper (not a result), there is a large gap between the model and what's needed to scale to a full lang -- it's not clear from the simple model how to design check insertion and blame tracking for new features. This required research, will be published elsewhere

*Custom Mutators (Section 4)*

The paper suggests that the mutators required here are significantly
different and more subtle, but this does not seem to be the case when
comparing Figure 3 and Table 3 of the POPL 2020 paper.

Of the 12 (of 16) gradual-typing related mutators listed in Figure 3:

* `constant`: Similar to previous, but now with a type error
* `deletion`: Similar to previous
* `public`: Similar to previous "hide-method"
* Four others involving swapping identifiers: Similar in spirit to
"swap argument" in previous

Of the 4 (of 16) Typed Racket related mutators:

* `arithmetic`: Simpler version of previous
* `boolean`: Same as in previous
* `negate-cond`: Same as in previous

So there is quite a bit of similarity. And on the flip side, why are
not all (8) mutators from the POPL 2020 experiment included here?

These questions are not so important in an absolute sense; _some_
specific mutators need to be chosen to generate some bugs, and these
seem to work fine. But they are salient given how much the paper
emphasizes the new mutators as a contribution.

*Debugging Strategies (Section 5)*

These also seem to be overemphasized. The main design choice seems to
be in how to use Transient blame, which reports multiple components,
and picking the first and last are pretty intuitive.

  TODO: the analytical part is important: we described all of these things in a common framework, allowing for comparison

I'm also not sure why the exception mode is needed in addition to
Erasure (I was confused in Section 2 L204-215 and Section 5.1 L732).
Indeed, Section 5.3 defines the Erasure mode the follow the Natural
exceptional mode.

  TODO: the exception modes are a key part of the analysis framework, enabling understanding of the value of blame separately from checking

*Additional Comments and Typos*

L69: I wish here, and elsewhere, the comparison to [13] had been made
more explicit. As discussed above, there seem to be more similarities
than suggested.

L92: "use of _a_ higher-order contract system"

L155: "authors extensive"

L253: "exports _it_ as"

L280: Missing `untyped-`

L314: " . "

L382: "Dyn,"

Fig 2: Presumably authors were included to indicate the variety of
sources, but I'm not sure that was necessary.

L597: "make us_e_ of"

L664: "described in _Section_ 4"

L980: "Figure 5.4" ==> "Figure 6"

L986: "2.8%" ==> "2.18%" ?

L1017: "33%" ==> "26%" ?

L1304: "Problem is ...."

Review #66E
===========================================================================

Overall merit
-------------
C. I would not accept this paper but will not argue strongly against
   accepting it.

Reviewer expertise
------------------
Z. **Some familiarity** = "I have a passing knowledge of the topic(s) but
   do not follow the relevant literature"

Paper summary
-------------
This paper seeks to establish (or refute) the benefit of blame tracking in gradually typed programming environments. It takes several Typed Racket benchmarks and considers multiple approaches to blame tracking. The central idea is to use a fault injection method to create variant programs along two dimensions: type errors inserted according to some crafted heuristics, and static typing selectively removed on a component-by-component granularity. This creates a large number of variant programs which may produce run-time type errors with blame; this is then iterated to simulate a "rational programmer", i.e. such that blame is used to select where more static checking will be added next; at any given iteration this may or may not catch the injected error, and the number of iterations is counted as a proxy for effort. A further contribution is implementing the different forms of blame-tracking in the same framework so that they can be compared. The finding is that blame is found to offer some positive advantage in effort (over simply being guided by the stack trace of a run-time exception) in 9--13% of the initial sample of variants, with some minor differences between the forms of blame tracking tried.

  TODO: misunderstanding: we propose a methodology

Comments for author
-------------------
I commend the authors for asking the question. Their method is clearly the product of much careful thought. I understand the difficulty in doing experiments about errors that tend not to hit 'real' source code repositories, and I appreciate the effort taken to re-create the different approaches to blame in a comparable setting.

However, I don't find the results to be as conclusive as the authors state. Aside from the Erasure case, whose relevance I found unclear (see below), conclusions rest on some small-valued "more useful than" percentages. These are pretty hard to interpret, because the "usefulness" metrics necessarily build in a lot of simplifying assumptions. E.g. in 5.4, the "percentage of scenarios where ... is more effective" doesn't account for *extent* of difference, and the trail-length "programmer effort" metric in 5.5 is also unlikely to be a great proxy. These limitations are understandable, method-wise, but overall the results are fairly null... they let us continue believing "what we'd expect" but not with any great sense of added confidence. That doesn't entirely diminish the work, of course, but I'd say the current write-up overplays its hand a little. I appreciate the discussion of threats to validity.
  TODO: misunderstanding: the results are not at all "what we'd expect" based on the theory

The presentation of the results is very much around aggregates and summaries. Indeed the whole method is about having run a huge compute job over a large number of variants. Perhaps it's paranoia but I'd be interested to see some specific examples walked through by hand (in the paper) and some smaller sample manually classified (as results). That would add an extra sanity check that the metrics do correspond to some meaningful reality.

About Erasure: if I'm reading this correctly, the key thing here is that in Typed Racket, the set of static types is more expressive than the set of dynamic types. For example, there is a static notion of non-negative integer that is distinct from plain integer. This sort of design isn't universal -- in some languages/systems static types closely mirror the classification of objects in the language's dynamic semantics. Writing these stricter contracts into a program brings its own benefits, separate from blame or indeed from static checking. It feels like the paper doesn't take enough care to distinguish the two effects: the effect of systematically applying more refined contracts over program values, and the effect of gradually enabling static checking of those contracts (iteratively, guided by blame). The authors do mention this around line 204, and the comparisons in Figure 6 between "_ exceptions" and "Erasure" seem to be measuring this -- the gain from checking these extra annotations, with the more precise and/or more timely checks that they imply, relative to the erased case where only the language baseline contracts are checked. Indeed that's the point of 'exception' experiments. Since the biggest effect sizes on display are these ones -- between Erasure and anything else -- this seems at best distracting. I'd be glad to hear from the authors if I'm misunderstanding anything here. In the detailed comments below, I have noted some places in the text where it would be useful to remind the reader that this design property of Typed Racket is at play.

It's interesting to note that there seems to be no convincing analogous experimental or modelling-based justification for plain old static typing, i.e. showing that it somehow presents a net gain to the programmer. Rather, this has simply been posited/assumed by a very long line of work. I am not defending that state of affairs, but it points to the difficulty of showing conclusively that something truly helps programmers. I can see value in this sort of simulation-style approach, and it is no less convincing than user studies. So I'd be interested to hear from the author(s) if they have any more arguments that (1) I've underestimated the results' conclusiveness, or (2) this is a novel/interesting family of methods that might be pursued more widely, or (3) that deeper experiments building on these ideas might yield more compelling insights.

Detailed comments:

In the title, "evaluating blame" reads oddly. It is ambiguous, because "evaluate" sometimes means to compute a result. A fuller phrase, like "evaluating the usefulness of blame tracking...", would probably be worth the words.

The abstract doesn't say much about the work. It would be better written for experts to quickly gather an overview what the paper contributes.

Conversely, the main body of the paper skimps a bit on background. It never explicitly covers what "blame" means and how it works in practice. Similarly, it repeatedly talks about "impedance mismatches" without defining them. Perhaps this phrase is now standard in the gradual typing literature, which I haven't kept up with (hence my Z expertise). But I did read the Wadler/Findler paper carefully at the time. It did not talk about impedance mismatches. In any case, it is a fuzzy metaphor... please say exactly what it means here.

line 34: "then their compilers remove types and rely on the built-in safety checks of the underlying language to catch any problems". This reads oddly in context. Didn't they just do a bunch of static checking? So they are *not* relying just on dynamic checks to catch problems? Maybe there is something more accurate to say here... e.g. no dynamic check is removed, or something like that.

line 38: "explicit statement and challenge" -- what is it?

line 49: I agree they got the word wrong, but you should explain this

line 60: around this paragraph the writing started to grate. There's no need to generalise about what people do or think, and the "As a matter of fact... simply..." style is somewhere between laboured and patronising. It's better to more plainly state the gap in the literature that you're addressing. Throughout the paper, much space could be saved by writing in a more direct style.

line 69: be explicit that Lazarek et al were (as I later gathered) doing something about blame to do with higher-order contracts but not gradual types

line 70: what does it mean to "follow" the slogan? Was unclear to me.

line 90: at first I wondered: what is a case? Maybe say "program variant" to foreshadow the idea of generating mutants etc?

line 91-ish: "Transient, "Natural", "Erasure" -- be explicit that these are names that *you* are introducing

line 94: "forego" => "forgo"

line 119: "homo economicus" needs glossing or a reference

line 125: was wondering whether "impedance mismatch" just means "feasible run-time type error".

line 148: last sentence appears to contradict the preceding paragraph, and is left hanging oddly. Instead, make it the start of a new paragraph.

line 160: "disparity" is a strange word here. "diversity"?

line 170: "Otherwise, results from..." -- the point was already clear

line 174: ... instead of just saying what you built on, first say what you did! It's really not clear at this point.

line 187: what is a "component"? Should be easy to define.

line 199: "blame set as another form of a stack trace" -- yes. I was hoping to see a clearer example of debugging with and without blame, i.e. something making explicit the similarities and differences between having blame info and having only a stack trace at the error site.

line 211: "languages exceptions" typo

line 268: this explanation of the workings of proxies seemed overwrought

line 277: what is "responsibility" of a "party", exactly? Are "party", "module" and "component" all the same things?

line 283: "rewrites typed modules to inline checks" -- so a module gets turned into a check? Clearly not, but that's how it reads

line 293: I was wondering what constitutes a boundary crossing. Clearly, passing by function call or return crosses from the caller's module to the callee's. What about values exchanged through reads/writes to shared state?

line 299: "crosses" => "crossings" (probably)

line 313: it threw me that a program might not fail but "produce a wrong result". If it could be caught by the gradual type system, why can't it be caught at run time? My best get at explaining this is by what I wrote above, i.e. it's a consequence of Typed Racket's more refined static notion of type. In certain other systems this wouldn't be possible, because the static checker would only catch (albeit earlier) errors that would be caught at run time, so there would be no basis to call the result "wrong".

lien 352: "three interpretations" -- what are they? I don't see them in the figure.

line 380: from this I inferred that "migratory typing" means "gradual typing applied at modulewise granularity". Assuming that's correct, it's worth saying directly.

line 421: "type mistake" -- does this mean "feasible run-time type error"?

line 424: "fully typed correct programs" -- presumably your method could also work with not-yet-fully-typed correct programs, just not ranging over the entire lattice in those cases. I was wondering whether that might give different/interesting results.

line 433: "without loss of diversity" -- this is glib. Clearly diversity is lost; just claim that what remains is still diverse enough.

line 497: "truthiness" needs explaining

line 517: the difference between #1 and #2 here again relies on the surprising property (to the unfamiliar) that Typed Racket has a notion of "type-level mistake" that doesn't surface under erasure (as an exception or whatever) but also is not a false positive (i.e. you're not talking about conservativeness of static checking). Line 538's "unavoidable" claim is probably also true only in such a context.

line 519: "at least three" -- better to claim this is only two here, then explain later that the driver doesn't count. Mentioning "three" up-front just raises an unnecessary question in the reader's mind.

line 634: earlier I had inferred that the distinction between "migratory" and "gradual" was more than just preferring one word; this seems to contradict that

line 640: what makes it "concise"?

line 673: "programmers runs" typo

line 727: "dubbed" -- reads oddly. Maybe italicise the "location", but it seems overkill. The phrase is pretty self-suggesting as it is.

line 756: "checked the value's type.." -- and the check passed!?

line 752: "added to the blame set first" -- should it be a blame list, then?

line 936: no need for hyphen after the adverb

line 966: "more useful C" -- missing "than"

line 982: don't think these percentages deserve 3 significant figures

line 1304: "Problem is" -- missing "The". Also, no need to italicise the next sentence... it is really not that deep or surprising.

### 2021-02-09 writing/reviews/pldi-rebuttal-notes.md
Change empty response to a TODO, remove note to check on it later

@@
 
-  ⟶ 
-	Response idea: This is an example where the experiment seems to contradict the theory. 
+  TODO: Response idea: This is an example where the experiment seems to contradict the theory.
 	The models don't include all of the features in the benchmarks. <enumerate them here> It's possible that the Natural, Transient theories don't scale as expected to rich PL features.
@@
 	It's also possible we have a bug in TR, or in Transient implementation.
-
-    Later: TODO: Scenarios where Transient is more useful than Natural imply that Transient locates the bug while Natural gets stuck.
-	      check what the reasons for those scenarios are;
-	      - the % is too high to all be because of the classes bug,
-		  - how else could Natural get stuck if Transient isn't?
-It can't be the bad exn cases, because then Transient should be stuck as well because Transient shouldn't be able to get blame in those cases.
 

### 2021-02-09 writing/reviews/pldi-rebuttal-notes.md
working on the rebuttal

@@
 ⟶ is a drafted response, TODO records notes about what to respond to or how.
+
+chrdimo notes format: unindented after chrdimo.
 
@@
 
+chrdimo: A bad mutator results in a type-level bug that is trivial to
+locate. For instance a bug that leads to a type-level error during the
+evaluation of the component that contains it is trivial to locate. In
+contrast, a bug that is non-trivial to locate is one  that causes the
+component that mutates it to evaluate without an error but produce instead
+a value of the wrong type.  An interesting bug is one where the value with
+the wrong type manages also to cross undetected to at least another
+component than the one that produces it until a type-level checks flags
+it. These latter bugs are those that good mutators should be able to
+produce.
+
 - I don't understand where the 16,800 number comes from. Is it an exhaustive
@@
 
+chrdimo: It is an exhaustive enumeration. We will clarify the prose.
+
 - How large are the programs in the benchmark suite?
   ⟶ The program sizes without type annotations, as reported by Racket's `syntax-sloc` tool, are below. We will add this information to the table in Figure 2.
+    benchmark | LOC | components
+	----------|-----|-----------
+	acquire | 1941 | 9
+	gregor | 2336 | 13
+	kcfa | 328 | 7
+	quadT | 7396 | 14
+	quadU | 7282 | 14
+	snake | 182 | 8
+	suffixtree | 1500 | 6
+	synth | 871 | 10
+	take5 | 465 | 8
+	tetris | 280 | 9
+
+chrdimo: Please find below the loc and number of components for each
+benchmark. We will add the information to figure 2.
+
    benchmark | LOC | components
@@
 
+chrdimo: No, the quote is not from [3]. We will rephrase to clarify that
+it is our statement of the research question.
+
 - L104: I was confused by the word "Next". Maybe instead: We now ..."
   ⟶ Thanks, we will change the wording.
+
+chrdimo: Thank you. We will reword.
 
@@
 
+- L980: Figure 5.4??
+  ⟶ Thanks, this is a bad reference; it should be figure 6. We will fix it.
+
+chrdimo: We will fix these. Thanks.
+
 - L956: 30,000 hours of compute time is 1,250 days. I assume that the CPU has
@@
 
-- L980: Figure 5.4??
-  ⟶ Thanks, this is a bad reference; it should be figure 6. We will fix it.
+chrdimo: Sure. Each CPU has 28 double cores.
 
@@
 	- a bunch of stacked histograms? Where there's one stack per pair of modes. A bar has three parts: A > B, A = B, and B > A.
+
+chrdimo: We have reached the same conclusion and we are considering
+alternatives such as a series of two-sided bar charts that make it easier
+to spot how two modes compare to each other.
 
@@
 
+chrdimo: We were also surprised to discover that transient outperforms
+natural in some scenarios. After all, the theory of blame says otherwise. Even thought
+we haven't identified specific patterns, we believe that the reason behind 
+transient's unexpected success is due to the advanced features of Typed
+Racket's type system that haven't appeared in models that compare Natural
+and Transient: polymorphism, row types, mixins etc. Such features require 
+custom loopholes to be compatible with Racket idioms. In turn these
+loopholes can result in subpar checks in Natural. Our experimentation with 
+transient indicates that the same problem doesn't occur in transient. 
+We do not have any clues about a strategy that outperforms both Natural
+and Erasure. But our current conjecture is that a simple strategy that
+just points somewhere in the blame trail may be good enough and enable
+downstream optimization that are keeping track of blame in Typed Racket
+inhibits. The good thing is is that now we have a method to compare
+different strategies.
+
+chrdimo? We need more details from Ben here.
+
 Theoretical gradual type systems usually allow programmers to control
@@
 
+chrdimo: Yes. Incorporating type Dyn boils down to considering a hierarchy
+of types for each component rather than a single type. At the same time,
+our method is parametric to the ``size'' of a component and adpats in a
+straightforward manner to a setting where each definition can be
+considered a component instead of a whole module. Of course both of these
+adjustments lead to significantly larger scenario lattices than the ones we
+investigate and they have to be sampled aggresively and carefully.
+
 Please explain what impedance mismatches are in the paper. Is it
@@
     We will adjust the text to clarify or remove this terminology.
+
+chrdimo: An impednace mismathc denotes a type annotation that doesn't
+match the actual type of the annotated code. We will adjust the prose to
+clarify.
 
@@
 
+chrdimo: You are right that for a given trail of length n the probability
+that the random rational programmer discovers the buggy component after k 
+attempts is always 1/n and independent of k. However we sample randomly
+from the set of sceharios and as a result the set of trails. So the curve 
+of the random programmer matches our sampling distribution. That said, as
+you observe, what is the distribution of the random programmer is an
+irrelevant point to our results and we will remove the characterization.
+
 * page 9 line 980: Figure 5.4 -> Figure 6
@@
   ⟶ Thanks, we will fix this.
+
+chrdimo: Thank you. We will fix all these.
 
@@
 But while there is a lot to like about the paper, it also has some shortcomings. The suite of programs used for the evaluation is fairly small (c.f., the SIGPLAN checklist).
+
   TODO: respond to this ^, saying that we use the standard suite of benchmarks for evaluating GT -- point to Ben's POPL paper and others, and we use all of the largest benchmarks in the suite
+
+chrdimo: Coupld you please clarify what you mean by fairly small with
+respect to the SIGPLAN checklist? We are the first to conduct an empirical
+study for blame and gradual typing so we repurposed the standard
+performance benchmark suite for Typed Racket and, in particular, its
+largest and most complex benchmarks. Our inspection of the benchmarks
+indicates that they are representative set of Racket programs in terms of
+features and structure.
+
 The faults are introduced using synthetic mutations that may or may not correspond to the kinds of errors that arise in practice.
 This can be seen from the results in Figure 7 -- the lengths of the "trails" is quite small.
+
   TODO: unclear that syntheticness relates to trail length at all, and synthetic bugs are the only option here -- there is no corpus of documented GT bugs in the wild anywhere.
@@
 Note that we are the first to evaluate blame in this setting.
+
+chrdimo: There is no exisitng catalogue of bugs in gradually typed
+programs. Furthermore, most type-level bugs do not survive after decent tsting and 
+thus do not make it to code repos. There are a few anecdotes from the literature but they are mostly
+artificial. Hence, since we need a large number of buggy scenarions for a
+meaningful study, a synthetic approach is the only way forward. In
+addtion, with our mutation we can tune the set of scenarios to get a
+diverse set of bugs.
+
 While the study is mostly well done, it only considers two blame assignment strategies.
-For example, it would be interesting to consider the "omniscient programmer" to get a bound on the value of blame.
-But this is not done.
+For example, it would be interesting to consider the "omniscient programmer" to get a bound on the value of blame. But this is not done.
+
   TODO: We don't understand what is meant by "omniscient programmer", but our best understanding is the ideal oracle strategy that always picks the right module right away and thus always finds the bug in one step.
@@
 We do have other baselines to understand blame's performance: the random mode, and the exceptions modes for each of the semantics.
+
+chrdimo: Could you please explain what you mean by "omniscient
+programmer"? If this is the programmer that always makes the best choice,
+i.e., finds the bug at the first try then this programmer is trivial. It
+has 100% success rate and a constant blame trail of 1. Thus comparing it
+with any strategy amounts to the number of scenarios where a strategy
+fails and those where it succeeds but the blame trail has length greater
+than one. This information is in the paper byt If you think it would be
+useful, we point it out. Moreover, the experiment considers 10 different
+modes of the rational programmer spread across 3 different gradaul typing
+systems.
+
 And unfortunately the results are inconclusive -- Section 8 says "our results call for a deeper understanding of the two models for blame".
+
   TODO: This is what makes our results surprising: they challenge the theoretical conclusions that Natural is superior to Transient, suggesting that the theory is not good enough to understand the pragmatics of blame in reality.
+
+chrdimo: The quote from section 8 refers to the theory of blame rather
+than our results. Our results show in clear terms that, at least for our
+benchmarks, a blame strategy is superior that exceptions and that natural
+and transient have comparable efficacy. The latter is surprisng given that
+theoretical models (and our intiail expectations) point to the opposite.
+Thus a side-effect of our experiment is a call to theoriticians for models
+that are atuned with practice.
+
 Also, the published Transient scheme seems to have a minor bug, which might affect the outcomes when fixed (Section 8.2).
   TODO: 8.2 does not indicate a bug in Transient, but rather a defect in the design of the system. ... something about a gap between Reticulated Python and
+
+chrdimo: Section 8.2 does not discuss a bug in Transient but rather a
+limitation of its current design that we discovered exactly becasue of the
+experiement.
+
 Most if not all of these shortcomings are discussed in the paper, which is much appreciated. But they do risk undermine the value of the contributions to some degree.
+
+chrdimo: We would like to point out that the resutls of the comparison
+between the tree systems are only one part of the contribution of the
+paper.  The other part is the method itself. Benchmarks, source of
+interesting bugs and rational programmer modes are all parameters that can
+be instantiated differently without affecting the recipe we discuss at the
+end of section 2 nor the analytical framework from section 5 that
+underlies our experiment.
 
@@
 
+chrdimo: Our experiemnt does have a baseline. The random mode of the
+rational programmer serves as the ``control'' of our experiment. 
+Furthermore the exception modes serve as further ``controls'' to
+distinguish the effect fo checks from that of blame,
+
 Similarly, I wondered why the modes of the transient programmer only consider picking the first/last blame, and not other choices.
@@
   TODO: Answer suggested by Ben: it's unspecified how one should interpret Transient blame, and there isn't time to evaluate all possible choices, so we pick the most reasonable interpretations.
+
+chrdimo: The transient semantics provide no interpratation for the blame
+set. We picked the first and last because they match the intuitive
+interpretation of the blame set as a list. We didn't mean to imply that
+other options are unreasonable. We just needed to make a reasonable choice 
+and keep the duration of the experiemtn within reasonbale limits.
 
@@
 of incidental implementation choices or bugs affecting this result.
+
   TODO: We should clarify that checks alone is faster, but adding blame is slower (bc blame map is linear).

### 2021-02-09 writing/reviews/pldi-rebuttal-notes.md
Transfer responses from notes to pldi-rebuttal.md for reviews A,B,C

@@
 chrdimo notes format: unindented after chrdimo.
+
+addressed: marks todos that have answers in `pldi-rebuttal.md`
 
@@
 
+  addressed
   TODO: Rewrite response along these lines...
@@
   there is a better way to represent the same data?
+  addressed
   TODO: come up with some thoughts for this figure
@@
 
+  addressed
   TODO: Response idea: This is an example where the experiment seems to contradict the theory.
@@
 
+  addressed
   TODO: The answer is yes, but this needs more thought to frame the answer the right way.
@@
 
+  addressed
   TODO: response sketch: you're right, we misspoke, this is not the normal distribution; our reasoning for why the normal-ish shape of distribution we see makes sense is that given a fixed trail the probability at each step of the trail ending is 1/n, but we randomly select trails so we're multiplying the random distribution of trail lengths by that 1/n one?
@@
 
+  addressed
   TODO: respond to this ^, saying that we use the standard suite of benchmarks for evaluating GT -- point to Ben's POPL paper and others, and we use all of the largest benchmarks in the suite
@@
 
+  addressed
   TODO: unclear that syntheticness relates to trail length at all, and synthetic bugs are the only option here -- there is no corpus of documented GT bugs in the wild anywhere.
@@
 
+  addressed
   TODO: We don't understand what is meant by "omniscient programmer", but our best understanding is the ideal oracle strategy that always picks the right module right away and thus always finds the bug in one step.
@@
 
+  addressed
   TODO: This is what makes our results surprising: they challenge the theoretical conclusions that Natural is superior to Transient, suggesting that the theory is not good enough to understand the pragmatics of blame in reality.
@@
 Also, the published Transient scheme seems to have a minor bug, which might affect the outcomes when fixed (Section 8.2).
+  addressed
   TODO: 8.2 does not indicate a bug in Transient, but rather a defect in the design of the system. ... something about a gap between Reticulated Python and
@@
 Section 5.2 just says "our answer" is that there are at least "two reasonable options." Why are other choices not reasonable?
+  addressed
   TODO: Answer suggested by Ben: it's unspecified how one should interpret Transient blame, and there isn't time to evaluate all possible choices, so we pick the most reasonable interpretations.

### 2021-02-09 writing/reviews/pldi-rebuttal.md
Transfer responses from notes to pldi-rebuttal.md for reviews A,B,C

--- 2021-02-09 writing/reviews/pldi-rebuttal-notes.md
@@
-Response notes format: indented after ⟶ or TODO.
-⟶ is a drafted response, TODO records notes about what to respond to or how.
+PLDI 2021 Paper #66 Author Response
+===========================================================================
 
-PLDI 2021 Paper #66 Reviews and Comments
-===========================================================================
-Paper #66 How to Evaluate Blame for Gradual Types
+First of all, we thank all of the reviewers for reading our submission, asking
+critical questions, and offering constructive feedback. Our responses to all
+comments and questions follow, in which we quote each comment/question and then
+respond below.
 
@@
 ===========================================================================
+> # Questions
+> - Could you comment a bit more on what makes a mutator "good"? (other than
+>   experimentally seeing that it is good.) Perhaps, what would be a really bad
+>   mutator for evaluating blame assignment?
 
-Overall merit
--------------
-A. I will champion accepting this paper.
+A bad mutator results in a type-level bug that is trivial to locate. For
+instance a bug that leads to a type-level error during the evaluation of the
+component that contains it is trivial to locate. In contrast, a bug that is
+non-trivial to locate is one that causes the component that mutates it to
+evaluate without an error but produce instead a value of the wrong type. An
+interesting bug is one where the value with the wrong type manages also to cross
+undetected to at least another component than the one that produces it until a
+type-level checks flags it. These latter bugs are those that good mutators
+should be able to produce.
 
-Reviewer expertise
-------------------
-Z. **Some familiarity** = "I have a passing knowledge of the topic(s) but
-   do not follow the relevant literature"
+> - I don't understand where the 16,800 number comes from. Is it an exhaustive
+>   enumeration of all possible mutations? (Or the number of mutants constructed
+>   in X amount of time?)
 
-Paper summary
--------------
-This paper presents a methodology for evaluating blame assignment strategies
-based on the notion of "the rational programmer". The paper develops the
-methodology in detail and then then uses it to evaluate three different blame
-strategies: "Transient", "Natural" and "Erasure".
+It is an exhaustive enumeration. We will clarify the prose.
 
-In gradual typing, a blame assignment is used to show where the program went
-wrong. Importantly, at least on the theory side, a well-typed component cannot
-be blamed. The authors argue that such blame assignment has received a lot of
-academic attention, yet industrial implementations of programming languages with
-gradual typing completely forego blame. Why is this? Is it because blame is not
-useful in practice? The paper aims to answer this question by applying the
-developed methodology to three strategies: The "Natural" blame strategy relies
-on proxies to perform additional type checks at run-time. The "Transient" blame
-strategy inlines these checks and does not perform "deep" type checks. The
-"Erasure" blame "strategy" performs no checks: it only fails if execution of the
-program fails.
+> - How large are the programs in the benchmark suite?
 
-The contribution of the paper is the methodology for evaluating blame (and
-secondary its application to the three strategies.) The idea is to design an
-algorithm that behaves as a programmer would debug a gradually typed program
-that crashes at run-time with a type error: by incrementally adding types to the
-untyped parts of the program until the source of the type error is revealed. In
-this way, the question is: Does the blame assignment (or lack thereof in the
-case of "Erasure") help guide the programmer? How many "steps" must he take to
-find the defect and can he find it at all?
+Here are the LOC counts and number of components for each benchmark. We will add
+the information to figure 2.
 
-Trying to summarize the methodology:
-
-- Take a collection of fully typed programs.
-- Consider mix-typed versions of those programs.
-- Develop mutators (like in mutation testing) that inject type errors.
-- Emulate the rational programmer who follows the "trail" in the spirit of
-  Figure 5.
-  - At each step the programmer decides on an untyped component and then types
-    it, reruns the program, and either sees a type error or decides to type
-    another untyped component.
-- Take into account how the "next untyped component" is selected based on the
-  specific system at hand. 
-- Experimentally evaluate whether the above method is able to find the fault and
-  through how many steps.
-
-Comments for author
--------------------
-This paper is exceptionally well-written and was a joy to read. Every time I had
-a concern the paper seemed to anticipate it and provide useful commentary.
-
-The paper does a very good job at describing the challenges of how to evaluate
-blame assignment strategies and then sets out to develop an automated evaluation
-methodology based on the notion of "the rational programmer". The paper
-carefully lays out the foundation for the generation of buggy mix-type programs,
-the actions and variants ("modes") of the rational programmer which can depend
-on the specific blame system, and finally how to sample the "experimental
-space". The paper also includes a clear section on threats to validity.
-
-I think it is important to have a methodology to evaluate the usefulness of
-blame, and in lieu of large-scale and costly user studies, I think this paper
-provides a very well thought out automatic approach (with the natural
-limitations that such an approach must inevitably have.)
-
-Points for:
-
-- Addresses a problem that has not been addressed before.
-- Clear, concise, and meaningful methodology.
-- Clear description of the threats to validity.
-- Very well-written paper.
-
-Points against:
-- none
-
-# Questions
-
-- Could you comment a bit more on what makes a mutator "good"? (other than
-  experimentally seeing that it is good.) Perhaps, what would be a really bad
-  mutator for evaluating blame assignment?
-
-  ⟶ Abstractly, a good mutator reliably generates bugs that are detectable by the type system and by all three semantics' runtime type checks; ideally the bugs also manifest during the interaction of multiple components.
-    We rely on the experimental check of our mutators since this criteria is difficult or impossible to decide otherwise.
-    That said, we can reason that some mutators are clearly unlikely or impossible to be good.
-	...
-	We will extend the discussion at the start of section 4 to clarify this point.
-
-  TODO: Rewrite response along these lines...
-    A common mutator in the literature is swapping constants of the same type, but this is not a good mutator for us because type system can't catch it.
-	Example of an obviously bad mutator: replacing any string with some pre-defined non-empty string (e.g. "foobar") would be a bad mutator.
-	Examples of mutators we cut: `wrap-cond` (tricks occurrence typing by hiding the result of conditionals to produce a type error, but never a dynamic error), `add-extra-method`, swapping `car` and `cdr`, `set!`ing an ID to itself (only causes a type error), swapping mutable and immutable hashes+vectors.
-	# This answer should serve to justify our mutators and the challenge of making good mutators more broadly, to answer reviewer D's concerns about the delta of making mutators.
-
-- I don't understand where the 16,800 number comes from. Is it an exhaustive
-  enumeration of all possible mutations? (Or the number of mutants constructed
-  in X amount of time?)
-  ⟶ The former, we will clarify in the text.
-
-- How large are the programs in the benchmark suite?
-  ⟶ The program sizes without type annotations, as reported by Racket's `syntax-sloc` tool, are below. We will add this information to the table in Figure 2.
   benchmark    | LOC  | components
-	----------|-----|-----------
+  -------------|------|-----------
   acquire      | 1941 | 9
@@
 
-# Minor Comments
+> # Minor Comments
+> - L38: Should there have been a colon? Is the text "This contrast... " and
+>   challenge quote from [3]? Who is saying what here...
 
-- L38: Should there have been a colon? Is the text "This contrast... " and
-  challenge quote from [3]? Who is saying what here...
-  ⟶ Thanks, we will reword the text from "This contrast" to clarify that this our research question, it is not coming from [3].
+No, the quote is not from [3]. We will rephrase to clarify that it is our
+statement of the research question.
 
-- L104: I was confused by the word "Next". Maybe instead: We now ..."
-  ⟶ Thanks, we will change the wording.
+> - L104: I was confused by the word "Next". Maybe instead: We now ..."
 
-- L268: Grammar issue.
-  ⟶ Thanks, we will fix this.
+Thank you. We will reword.
 
-- L314: Floating period.
-  ⟶ Thanks, we will fix this.
+> - L268: Grammar issue.
+> - L314: Floating period.
+> - L441: Add line numbers (or other metric of size).
+> - L980: Figure 5.4??
 
-- L441: Add line numbers (or other metric of size).
-  ⟶ See above, we will add this information.
+We will fix these. Thanks.
 
-- L956: 30,000 hours of compute time is 1,250 days. I assume that the CPU has
-  many cores. Could you put that number?
-  ⟶ Each CPU has 28 cores and 56 threads; when we last ran the experiment, it took about two weeks of real time from start to end. 
-  We will add this information at the start of section 7.
+> - L956: 30,000 hours of compute time is 1,250 days. I assume that the CPU has
+>   many cores. Could you put that number?
 
-- L980: Figure 5.4??
-  ⟶ Thanks, this is a bad reference; it should be figure 6. We will fix it.
+Sure. Each CPU has 28 double-threaded cores.
 
-- L991: Figure 6: It took me some time to understand the figure. I wonder if
-  there is a better way to represent the same data?
-  TODO: come up with some thoughts for this figure
-  ⟶ Thanks, we can see how it's difficult to read. We will consider some different presentations, such as...
-    - a bunch of two-sided bar charts? Where there's one bar pair per pair of modes. Above the x-axis is % A > B, and below the axis is % B > A?
-	- a bunch of stacked histograms? Where there's one stack per pair of modes. A bar has three parts: A > B, A = B, and B > A.
+> - L991: Figure 6: It took me some time to understand the figure. I wonder if
+>   there is a better way to represent the same data?
+
+We have reached the same conclusion and we are considering alternatives such as
+a series of two-sided bar charts that make it easier to spot how two modes
+compare to each other.
 
@@
 ===========================================================================
+> The experiment results show that there are debugging scenarios where
+> Transient is more useful than Natural. Does it happen simply because
+> of luck? Or, does it imply that Transient is more suitable than
+> Natural for a particular kind of bugs for some reason? Also, can there
+> be a new blame strategy that outperforms both Transient and Natural
+> all the time?
 
-Overall merit
--------------
-B. I support accepting this paper but will not champion it.
+We were also surprised to discover that transient outperforms natural in some
+scenarios. After all, the theory of blame says otherwise. Even thought we
+haven't identified specific patterns, we believe that the reason behind
+transient's unexpected success is due to the advanced features of Typed Racket's
+type system that haven't appeared in models that compare Natural and Transient:
+polymorphism, row types, mixins, and more. Such features require custom
+loopholes to be compatible with Racket idioms. In turn these loopholes can
+result in subpar checks in Natural. Our experimentation with transient indicates
+that the same problem doesn't occur in transient. We do not have any clues about
+a strategy that outperforms both Natural and Erasure. But our current conjecture
+is that a simple strategy that just points somewhere in the blame trail may be
+good enough and enable downstream optimization that are keeping track of blame
+in Typed Racket inhibits. The good thing is is that now we have a method to
+compare different strategies.
 
-Reviewer expertise
-------------------
-Y. **Knowledgeable** = "I follow the literature on this paper's topic(s)
-   but may have missed some relevant developments"
+chrdimo: TODO We need more details from Ben here.
 
-Paper summary
--------------
-The paper proposes an evaluation method for blame assignment
-strategies and evaluates three different strategies with the
-method. While many theoretical gradual typing systems have adopted
-blame, blame has been neither shown to be practically useful nor
-applied to real-world languages because the systematic evaluation of blame
-is difficult. The authors introduce a notion of rational programmers
-and simulate their behaviors to evaluate blame. A rational programmer
-is assumed to iteratively add type annotations to an untyped portion
-of his or her program according to the feedback given by a run-time
-system. This behavior can be simulated even without real human
-programmers. If a run-time system leads a programmer to the correct
-locations of a bug, its blame strategy can be considered more useful
-than a strategy that cannot do so. The paper compares the Natural,
-Transient, and Erasure strategies. The result shows that both Natural
-and Transient outperform Erasure, which does not assign any blame,
-and, therefore, implies usefulness of blame assignments.
+> Theoretical gradual type systems usually allow programmers to control
+> the precision of type annotations in a fine-grained way with the Dyn
+> type rather than distinguishing typed and untyped modules. Would the
+> same evaluation method be able to directly appliable to blame in such systems?
 
-  TODO: This seems to imply that we are proposing or assuming the way that real "rational programmers" would behave. Should we try to answer or clarify about that?
+Yes. Incorporating type Dyn boils down to considering a hierarchy of types for
+each component rather than a single type. At the same time, our method is
+parametric to the ``size'' of a component and adapts in a straightforward manner
+to a setting where each definition can be considered a component instead of a
+whole module. Of course both of these adjustments lead to significantly larger
+scenario lattices than the ones we investigate and they have to be sampled
+aggressively and carefully.
 
-Comments for author
--------------------
-This paper tackles a long-standing difficult problem, which is the
-problem of evaluating blame. Many existing gradual typing papers
-provide blame systems, but because the papers do not evaluate blame
-systems, it is not obvious whether blame itself is indeed helpful for
-debugging. Among many theoretical gradual type systems adopting blame
-assignments, no paper discussed the evaluation of blame systems. It is an
-important problem because, without any evaluation, theoretical aspects
-like gradual guarantee and the blame theorem may not be applicable to
-practical systems combining static and dynamic typing.
+> Please explain what impedance mismatches are in the paper. Is it
+> simply type mismatches?
 
-This paper shows that blame assignments do help programmers, and it
-may serve as a first step to close the gap between the theory and
-the practice of gradual typing.
+An impedance mismatch denotes a type annotation that doesn't match the actual
+type of the annotated code. We will adjust the prose to clarify.
 
-The experiment results show that there are debugging scenarios where
-Transient is more useful than Natural. Does it happen simply because
-of luck? Or, does it imply that Transient is more suitable than
-Natural for a particular kind of bugs for some reason? Also, can there
-be a new blame strategy that outperforms both Transient and Natural
-all the time?
+> The paper states that the effort distribution of the random mode
+> follows a normal distribution, as expected. Please explain why a
+> normal distribution is a reasonable expectation. ...
+> This point does not affect the contribution of the paper, but it
+> is not clear from the paper.
 
-  TODO: Response idea: This is an example where the experiment seems to contradict the theory.
-	The models don't include all of the features in the benchmarks. <enumerate them here> It's possible that the Natural, Transient theories don't scale as expected to rich PL features.
-	Make analogy to physics: theory makes predictions, we have found failures of the predictions. We have two courses of action: report the inconsistency to re-eval the theory, and do more experiments.
-	It's also possible we have a bug in TR, or in Transient implementation.
+You are right that for a given trail of length n the probability that the random
+rational programmer discovers the buggy component after k attempts is always 1/n
+and independent of k. However we sample randomly from the set of scenarios and
+as a result the set of trails. So the curve of the random programmer matches our
+sampling distribution. That said, as you observe, what is the distribution of
+the random programmer is an irrelevant point to our results and we will remove
+the characterization.
 
-Theoretical gradual type systems usually allow programmers to control
-the precision of type annotations in a fine-grained way with the Dyn
-type rather than distinguishing typed and untyped modules. Would the
-same evaluation method be able to directly appliable to blame in such systems?
+> * page 9 line 980: Figure 5.4 -> Figure 6
+> * page 10 Figure 7: Null -> Random
+> * page 11 line 1103: ( 770 -> (770
 
-  TODO: The answer is yes, but this needs more thought to frame the answer the right way.
-  ⟶ Our evaluation framework should be directly applicable to blame in gradual typing systems using Dyn.
-    The details of how it applies depends on the granularity of the blame information offered by the system -- in other words, what the system considers components.
-	If components are modules, as in Typed Racket, then nothing changes.
-	Systems with Dyn more likely treat top-level definitions as components, in which case definitions are individually typed (the whole definition is annotated) rather than modules.
-
-Please explain what impedance mismatches are in the paper. Is it
-simply type mismatches?
-
-  ⟶ Impedance mismatches are mismatches between a piece of code's true type and its annotation.
-    For instance, the code `(λ (x) (+ x 42))` has type `Number -> Number` but might be annotated as `Number -> String`, causing an impedance mismatch.
-    We will adjust the text to clarify or remove this terminology.
-
-# Strengths
-
-* The paper is well-organized.
-* It proposes a novel solution to a long-standing difficult problem.
-* Threats to validity are discussed in detail.
-
-# Minor comments
-
-The paper states that the effort distribution of the random mode
-follows a normal distribution, as expected. Please explain why a
-normal distribution is a reasonable expectation. When there are $n$
-untyped components, the random mode randomly selects one of them, and
-only one of them is a buggy component. Therefore, in the first
-iteration, the random mode reaches the end of the trail with the
-probability of $1/n$. In the second iteration, there are $(n-1)$
-untyped components and the probability of a failure at the first
-iteration is $(n-1)/n$, so the probability of effort $2$ is $1/n$
-again. In this way, it seems to be that the probability of effort $i$
-is $1/n$ where $1<=i<=n$ and $n$ is the number of initially untyped
-components. Each trail has a different number of untyped components,
-so the resulting distribution would be the average of such
-distributions. Based on this reasoning, it is unclear why the
-distribution is a normal distribution. This point does not affect the
-contribution of the paper, but it is not clear from the paper.
-
-  TODO: response sketch: you're right, we misspoke, this is not the normal distribution; our reasoning for why the normal-ish shape of distribution we see makes sense is that given a fixed trail the probability at each step of the trail ending is 1/n, but we randomly select trails so we're multiplying the random distribution of trail lengths by that 1/n one?
-Not clear on the details of this reasoning.
-  ⟶
-
-* page 9 line 980: Figure 5.4 -> Figure 6
-  ⟶ Thanks, we will fix this.
-* page 10 Figure 7: Null -> Random
-  ⟶ Thanks, we will fix this.
-* page 11 line 1103: ( 770 -> (770
-  ⟶ Thanks, we will fix this.
+Thank you. We will fix all these.
 
@@
 ===========================================================================
+> But while there is a lot to like about the paper, it also has some shortcomings. The suite of programs used for the evaluation is fairly small (c.f., the SIGPLAN checklist).
 
-Overall merit
--------------
-C. I would not accept this paper but will not argue strongly against
-   accepting it.
+Could you clarify what you mean by fairly small with respect to the SIGPLAN
+checklist? We are the first to conduct an empirical study for blame and gradual
+typing, so we repurposed the standard performance benchmark suite for Typed
+Racket -- in particular, its largest and most complex benchmarks. Our inspection
+of the benchmarks indicates that they are representative set of Racket programs
+in terms of features and structure.
 
-Reviewer expertise
-------------------
-Z. **Some familiarity** = "I have a passing knowledge of the topic(s) but
-   do not follow the relevant literature"
+> The faults are introduced using synthetic mutations that may or may not correspond to the kinds of errors that arise in practice.
+This can be seen from the results in Figure 7 -- the lengths of the "trails" is quite small.
 
-Paper summary
--------------
-This paper develops an approach for empirically evaluating the effectiveness of blame assignment in gradually typed languages.
-Roughly speaking, the idea is to model the steps taken by a "rational programmer" who systematically debugs errors in a gradually typed programming.
-The question is whether different blame assignment strategies are able to help this rational programmer find the "true" source of errors faster, requiring fewer steps.
+There is no existing catalog of bugs in gradually typed programs. Furthermore,
+most type-level bugs do not survive after decent testing and thus do not make it
+to code repos. There are a few anecdotes from the literature but they are mostly
+artificial. Hence, since we need a large number of buggy scenarios for a
+meaningful study, a synthetic approach is the only way forward. In addition,
+mutation allows us to tune the set of scenarios to get a diverse set of bugs.
 
-To evaluate this approach, the paper also presents empirical results.
-Starting with a small corpus of programs, the authors use mutations to systematically (but synthetically) introduce type errors -- e.g., one mutation replaces a constants with another of a different type, and another mutation swaps method identifiers.
-They then conduct experiments to quantify the number of steps required by a rational programmer for several blame assignment schemes: Natural (used in Typed Racket, which tracks at most one location for blame), Transient (used in Reticulated Python, which tracks a set of locations), and Erasure (which does not actually assign blame, and is widely used in industry).
+> While the study is mostly well done, it only considers two blame assignment strategies.
+> For example, it would be interesting to consider the "omniscient programmer" to get a bound on the value of blame. But this is not done.
 
-  TODO: do we need to clarify that effort is a secondary aspect of the evaluation, and more important is whether blame leads to the bug at all?
+Could you elaborate what you mean by "omniscient programmer"? If this is the
+programmer that always makes the best choice, i.e. finds the bug at the first
+try, then this programmer is trivial. It has 100% success rate and a constant
+blame trail length of 1. Thus, comparing its success to any other strategy
+amounts to counting the number of scenarios where the other strategy fails;
+comparing effort is similarly trivial. This information is in the paper
+implicitly, but we can certainly add an explicit mention if it would be useful.
 
-Unsurprisingly, the evaluation shows that blame is effective, when used by a rational programmer.
-The results are inconclusive, but some trends are clear.
-Most significant, all of the schemes that assign blame are more useful than Erasure.
-And Natural blame more often more useful than Transient blame.
+Secondly, we would like to point out that the experiment considers then
+different modes of the rational programmer spread across three different gradual
+typing systems.
 
-Comments for author
--------------------
-This is a neat paper that develops an approach for evaluating gradual typing and blame assignment.
-Thanks for submitting it to PLDI '21!
-The results are presented in an extremely clear way that draws out the key insights without compromising on technical detail.
-The notion of a rational programmer -- loosely modeled on rational agents from economic theory -- is a neat conceptual device.
-And the empirical approach does seem like the right way to address the question of the usefulness of blame.
+> And unfortunately the results are inconclusive -- Section 8 says "our results call for a deeper understanding of the two models for blame".
 
-But while there is a lot to like about the paper, it also has some shortcomings. The suite of programs used for the evaluation is fairly small (c.f., the SIGPLAN checklist).
-  TODO: respond to this ^, saying that we use the standard suite of benchmarks for evaluating GT -- point to Ben's POPL paper and others, and we use all of the largest benchmarks in the suite
-The faults are introduced using synthetic mutations that may or may not correspond to the kinds of errors that arise in practice.
-This can be seen from the results in Figure 7 -- the lengths of the "trails" is quite small.
-  TODO: unclear that syntheticness relates to trail length at all, and synthetic bugs are the only option here -- there is no corpus of documented GT bugs in the wild anywhere.
-Indeed, most type errors never even make it into code repositories.
-Note that we are the first to evaluate blame in this setting.
-While the study is mostly well done, it only considers two blame assignment strategies.
-For example, it would be interesting to consider the "omniscient programmer" to get a bound on the value of blame.
-But this is not done.
-  TODO: We don't understand what is meant by "omniscient programmer", but our best understanding is the ideal oracle strategy that always picks the right module right away and thus always finds the bug in one step.
-We didn't show this because it's trivial.
-We do have other baselines to understand blame's performance: the random mode, and the exceptions modes for each of the semantics.
-And unfortunately the results are inconclusive -- Section 8 says "our results call for a deeper understanding of the two models for blame".
-  TODO: This is what makes our results surprising: they challenge the theoretical conclusions that Natural is superior to Transient, suggesting that the theory is not good enough to understand the pragmatics of blame in reality.
-Also, the published Transient scheme seems to have a minor bug, which might affect the outcomes when fixed (Section 8.2).
-  TODO: 8.2 does not indicate a bug in Transient, but rather a defect in the design of the system. ... something about a gap between Reticulated Python and 
-Most if not all of these shortcomings are discussed in the paper, which is much appreciated. But they do risk undermine the value of the contributions to some degree.
+The quote from section 8 refers to the theory of blame rather than our results.
+Our results show in clear terms (at least for our benchmarks) that a blame
+strategy is superior to exceptions, and that Natural and Transient have
+comparable efficacy. The latter is surprising given that theoretical models
+suggest the opposite conclusion. Thus, a side-effect of our experiment is a call
+to theoreticians for models that are aligned with practice.
 
-To explain one comment above: I was surprised not to see a comparison against an "omniscient programmer" to establish a baseline (c.f., the SIGPLAN checklist).
+> Also, the published Transient scheme seems to have a minor bug, which might affect the outcomes when fixed (Section 8.2).
+
+Section 8.2 does not discuss a bug in Transient but rather a limitation of its
+current design that we discovered exactly because of the experiment.
+
+> Most if not all of these shortcomings are discussed in the paper, which is much appreciated. But they do risk undermine the value of the contributions to some degree.
+
+We would like to point out that the results of the comparison between the three
+systems are only one part of the contribution of the paper. The other part is
+the method itself. The benchmarks, source of interesting bugs, and rational
+programmer modes are all parameters that can be instantiated differently without
+affecting either the recipe we discuss at the end of section 2 or the analytical
+framework from section 5 that underlies our experiment.
+
+> To explain one comment above: I was surprised not to see a comparison against an "omniscient programmer" to establish a baseline (c.f., the SIGPLAN checklist).
 That is, model a blame assignment scheme in which blame could be placed on any component.
@@
 
-Similarly, I wondered why the modes of the transient programmer only consider picking the first/last blame, and not other choices.
+Our experiment does have a baseline. The random mode of the rational programmer
+serves as the "control" of our experiment. Furthermore the exception modes each
+serve as further "controls" to distinguish the effect of checks from that of
+blame within each semantics.
+
+> Similarly, I wondered why the modes of the transient programmer only consider picking the first/last blame, and not other choices.
 Section 5.2 just says "our answer" is that there are at least "two reasonable options." Why are other choices not reasonable?
-  TODO: Answer suggested by Ben: it's unspecified how one should interpret Transient blame, and there isn't time to evaluate all possible choices, so we pick the most reasonable interpretations.
+
+The transient semantics provide no interpretation for the blame set. We picked
+the first and last because they match the intuitive interpretation of the blame
+set as a list. We didn't mean to imply that other options are unreasonable. We
+just needed to make a reasonable choice and keep the duration of the experiment
+within reasonable limits. We will clarify this point in the prose.
 

### 2021-02-09 writing/reviews/pldi-rebuttal-notes.md
Add Matthias's note on reviewer E's comment to -notes.md

@@
 
+  mf: As posed, the question expresses a significant mis-characterization of
+	  the submission. It is not the benefits of a gradual type system that
+	  are in doubt. The experimental setup keeps the type system constant
+	  but allows to answer the question which of several run-time checking
+	  regimes (for enforcing type consistency) provides the best explanatory
+	  messages in case of violations.
+	  Hence the analogous question for purely static type systems would ask
+	  which of several reporting schemes provides the best explanatory
+	  message for type-errors. In the case of simply typed languages and
+	  even language with local type inference, this question is basically
+	  meaningless. In the case of languages with HM type inference, the
+	  question has been implicitly raised for four decades with the
+	  development of alternative ways of finding the source of inference
+	  conflicts.
+	  If the reviewer is indeed interested in the question of whether type
+	  systems help programmers---a question that this submission does _not_
+	  ask---the recent OOPSLA literature contains several user studies. A
+	  simple Google query will suggest a short list of these publications.
+
 Detailed comments:

### 2021-02-09 writing/reviews/pldi-rebuttal-notes.md
Add draft responses to reviewer E's comments

@@
   TODO: misunderstanding: we propose a methodology
+  ⟶ The primary intended contribution of this paper is the methodology, and the experimental results serve as secondary contributions that validate the methodology's utility.
+    We will rephrase the introduction to further emphasize that focus.
+	Furthermore, we would like to emphasize that the two key findings in the experimental results are that 1) both Natural and Transient's dynamic type enforcement with blame tracking improve significantly over Erasure, and 2) Natural's blame tracking is not significantly better than Transient's, despite the theoretical indications to the contrary.
 
@@
 I appreciate the discussion of threats to validity.
+
   TODO: misunderstanding: the results are not at all "what we'd expect" based on the theory
+  ⟶ The status quo of what we would expect based on the current theory of Natural and Transient is that 1) either kind of blame tracking should be better than none, and 2) Natural should provide significantly better blame information than Transient and Erasure.
+    Our results validate the first expectation that blame provides benefits over plain exceptions.
+    Surprisingly, our results contradict the second expectation and suggest that in fact Natural and Transient offer comparable blame information.
 
@@
 
+  TODO: misunderstanding about gradual typing as a whole, and the point of the exception modes
+  ⟶ The idea of the Erasure semantics is not related to the expressiveness of the type system or how the types relate to kinds of runtime values.
+    Rather, the key idea is that Erasure does not enforce type annotations with dynamic checks at all.
+	Running a gradual program using the Erasure semantics is the same as stripping all annotations and running the resulting (completely dynamic) program.
+	Thus, Erasure programs can only raise exceptions from either the checks performed by the language's primitive operations (e.g. `+`), or from the program itself raising an exception.
+	In contrast, the other two semantics do enforce type annotations dynamically; both Natural and Transient insert dynamic checks verifying that the shape of values at runtime match their static annotations.
+	Thus, in addition to exceptions, gradual programs using these semantics may raise a second kind of error: a dynamic type check failure.
+	These dynamic errors can include blame information if the dynamic type checks are designed to track blame, or they can forego blame and just offer a stack trace.
+	In the case of Natural, these two options are captured by the "Natural" and "Natural exceptions" modes, respectively.
+	Thus the "Erasure" mode represents a completely different kind of checking (and thus error reporting) compared to both the "Natural" and "Natural exceptions" modes (the two of which share the same kind checking, but different error reporting).
+
 It's interesting to note that there seems to be no convincing analogous experimental or modelling-based justification for plain old static typing, i.e.
@@
 
-  mf: As posed, the question expresses a significant mis-characterization of
+  ⟶ (Matthias) As posed, the question expresses a significant mis-characterization of
     the submission. It is not the benefits of a gradual type system that
@@
 
+  TODO ll: added responses to anything below that seems to warrant more than "Thanks, we'll fix it".
+
 In the title, "evaluating blame" reads oddly.
@@
 please say exactly what it means here.
+  ⟶ We will adjust the prose to more explicitly describe blame and impedance mismatches.
 
@@
 
+  ⟶ The static checking performed by languages using the Erasure semantics (like Flow, Hack, and TypeScript) is done in a best-effort fashion; the fact that a program passes type checking has absolutely no effect on how the program runs.
+    These languages literally erase the types to obtain a program of the underlying language (e.g. TypeScript or PHP) to run.
+	Those underlying languages perform all of the dynamic checks that a typical dynamic language performs for primitive operations (e.g. that `+` is only used with numbers).
+
 line 38: "explicit statement and challenge" -- what is it?
+  ⟶ The challenge questions whether blame is useful at all in gradual typing.
+    We will clarify this in the prose.
 
@@
 Throughout the paper, much space could be saved by writing in a more direct style.
+  ⟶ Noted, we will simplify the language where possible.
 
@@
 line 70: what does it mean to "follow" the slogan? Was unclear to me.
+  ⟶ The following sentences are meant to explain this, so we will adjust the prose to make that clear.
 
@@
 line 91-ish: "Transient, "Natural", "Erasure" -- be explicit that these are names that *you* are introducing
+  ⟶ These are names from the Gradual Typing literature (see for instance [9, 10, 37]).
 
@@
 line 277: what is "responsibility" of a "party", exactly? Are "party", "module" and "component" all the same things?
+  ⟶ Components are parties, and in the case of Typed Racket modules are components because they are the units of migration.
+Other gradual type systems, like those that do not require annotated entire modules, may consider smaller units of code (e.g.
+definitions) as components.
+    We will clarify this in the prose.
 
@@
 What about values exchanged through reads/writes to shared state?
+  ⟶ Yes, communicating a value via state also constitutes a boundary crossing.
+    We will clarify what a boundary crossing is in the prose.
 
@@
 In certain other systems this wouldn't be possible, because the static checker would only catch (albeit earlier) errors that would be caught at run time, so there would be no basis to call the result "wrong".
+  ⟶ The result can be wrong in that it doesn't agree with the annotations (which we assume express programmer intent), even if none of the primitive operations applied to the result raise an error.
+    For example, imagine the the client saves the result to a file.
+    The language primitives for writing values to a file (typically) can write any kind of value to a file, so they have no checks that could fail due to the result being of the wrong type.
+    Instead, the resulting file just ends up with the incorrect contents and the program terminates normally.
 
@@
 line 380: from this I inferred that "migratory typing" means "gradual typing applied at modulewise granularity". Assuming that's correct, it's worth saying directly.
+  ⟶ The three choices of using Dyn, the granularity of components, and how compilation affects completely untyped components represent orthogonal aspects of the design of a Gradual Typing system.
+    In the case of Typed Racket's migratory typing, those three choices amount to more than just applying the ideas of [21] at the granularity of modules.
 
 line 421: "type mistake" -- does this mean "feasible run-time type error"?
+  ⟶ Could you clarify what a makes a runtime type error "feasible"?
 
@@
 I was wondering whether that might give different/interesting results.
+  ⟶ Yes, the approach could be applied, but the results may be mostly inconclusive depending on how many trails reach the point where the blamed component lacks type informtion.
+    All trails that reach that point provide no useful information.
 
@@
 Line 538's "unavoidable" claim is probably also true only in such a context.
+  TODO: Don't respond to this in light of above explanation(s) of why they misunderstand erasure and TR?
 
@@
 line 640: what makes it "concise"?
+  TODO ll: Why does it? I'm not sure about the word here either. Just say we'll change it?
 
@@
 line 756: "checked the value's type.." -- and the check passed!?
+  ⟶ Indeed, Transient checks only check the top-level type constructor of a value, so many checks can pass for a bad value before it is unwrapped to discover the problem.
+    We will add a reminder of this fact to the prose here.
 
@@
 line 982: don't think these percentages deserve 3 significant figures
+  TODO: "deserve"?
+  ⟶ We include the hundreths place to highlight in particular that Transient first and last blame are not exactly the same.
+    If this information isn't useful, we can use just the tenths place.
 

### 2021-02-09 writing/reviews/pldi-rebuttal.md
Edit responses to reviewer A for clarity and flow

@@
 ===========================================================================
-> # Questions
 > - Could you comment a bit more on what makes a mutator "good"? (other than
@@
 A bad mutator results in a type-level bug that is trivial to locate. For
-instance a bug that leads to a type-level error during the evaluation of the
+instance, a bug that leads to a type-level error during the evaluation of the
 component that contains it is trivial to locate. In contrast, a bug that is
-non-trivial to locate is one that causes the component that mutates it to
-evaluate without an error but produce instead a value of the wrong type. An
-interesting bug is one where the value with the wrong type manages also to cross
-undetected to at least another component than the one that produces it until a
-type-level checks flags it. These latter bugs are those that good mutators
-should be able to produce.
+non-trivial to locate is one that causes the buggy component to evaluate without
+an error but produce instead a value of the wrong type. An interesting bug is
+one where the value with the wrong type manages also to cross undetected to at
+least another component before a type-level check flags it. Good mutators should
+be able to produce these interesting bugs.
 
@@
 
-Sure. Each CPU has 28 double-threaded cores.
+Sure. Each CPU has 28 doubly-threaded cores.
 
@@
 
-We have reached the same conclusion and we are considering alternatives such as
-a series of two-sided bar charts that make it easier to spot how two modes
-compare to each other.
+We have reached the same conclusion and we are considering alternatives that
+will make it easier to see how two modes compare to each other, such as a series
+of two-sided bar charts.
 
@@
 
-We were also surprised to discover that transient outperforms natural in some
-scenarios. After all, the theory of blame says otherwise. Even thought we
-haven't identified specific patterns, we believe that the reason behind
-transient's unexpected success is due to the advanced features of Typed Racket's
-type system that haven't appeared in models that compare Natural and Transient:
-polymorphism, row types, mixins, and more. Such features require custom
-loopholes to be compatible with Racket idioms. In turn these loopholes can
-result in subpar checks in Natural. Our experimentation with transient indicates
-that the same problem doesn't occur in transient. We do not have any clues about
-a strategy that outperforms both Natural and Erasure. But our current conjecture
-is that a simple strategy that just points somewhere in the blame trail may be
-good enough and enable downstream optimization that are keeping track of blame
-in Typed Racket inhibits. The good thing is is that now we have a method to
-compare different strategies.
+We were also surprised to discover that Transient outperforms Natural in some
+scenarios. After all, the theory of blame says otherwise. Even though we haven't
+identified specific patterns, we believe that the reason behind Transient's
+unexpected success is due to the advanced features of Typed Racket's type system
+that haven't appeared in models that compare Natural and Transient:
+polymorphism, row types, mixins, and more.
+
+We do not have any clues about a strategy that outperforms both Natural and
+Erasure. That said, our current conjecture is that a simple strategy that just
+points somewhere in the blame trail may be good enough. Such a strategy would
+also enable optimizations that Typed Racket currently foregoes in order to track
+blame. The good thing is that now we have a method to compare different
+strategies.
 
@@
 each component rather than a single type. At the same time, our method is
-parametric to the ``size'' of a component and adapts in a straightforward manner
+parametric to the "size" of a component and adapts in a straightforward manner
 to a setting where each definition can be considered a component instead of a
 whole module. Of course both of these adjustments lead to significantly larger
-scenario lattices than the ones we investigate and they have to be sampled
-aggressively and carefully.
+scenario lattices than the ones we investigate, so keeping the experiment
+computationally feasible would require more aggressive sampling.
 
@@
 
-An impedance mismatch denotes a type annotation that doesn't match the actual
-type of the annotated code. We will adjust the prose to clarify.
+An impedance mismatch denotes a type annotation that is detected to disagree
+with the actual type of the annotated code. We will adjust the prose to clarify.
 
@@
 
-You are right that for a given trail of length n the probability that the random
-rational programmer discovers the buggy component after k attempts is always 1/n
-and independent of k. However we sample randomly from the set of scenarios and
-as a result the set of trails. So the curve of the random programmer matches our
-sampling distribution. That said, as you observe, what is the distribution of
-the random programmer is an irrelevant point to our results and we will remove
-the characterization.
+You are right that for a given trail of length $n$ the probability that the
+random rational programmer discovers the buggy component on every step is $1/n$.
+The missing piece is that we randomly sample from the set of scenarios -- and by
+extension the set of trails. So the shape of the random programmer's effort
+matches our sampling distribution. That said, as you observe, the distribution
+of the random programmer's effort is not relevant to our results, so we will
+remove the characterization.
 

### 2021-02-09 writing/reviews/pldi-rebuttal-notes.md
working on review D
draft of reply to review D

@@
 of types for each component rather than a single type. At the same time,
-our method is parametric to the ``size'' of a component and adpats in a
+our method is parametric to the ``size'' of a component and adapts in a
 straightforward manner to a setting where each definition can be
@@
 adjustments lead to significantly larger scenario lattices than the ones we
-investigate and they have to be sampled aggresively and carefully.
+investigate and they have to be sampled aggressively and carefully.
 
@@
 
-chrdimo: An impednace mismathc denotes a type annotation that doesn't
+chrdimo: An impedance mismatch denotes a type annotation that doesn't
 match the actual type of the annotated code. We will adjust the prose to
@@
 attempts is always 1/n and independent of k. However we sample randomly
-from the set of sceharios and as a result the set of trails. So the curve 
+from the set of scenarios and as a result the set of trails. So the curve 
 of the random programmer matches our sampling distribution. That said, as
@@
 
-chrdimo: Coupld you please clarify what you mean by fairly small with
+chrdimo: Could you please clarify what you mean by fairly small with
 respect to the SIGPLAN checklist? We are the first to conduct an empirical
@@
 
-chrdimo: There is no exisitng catalogue of bugs in gradually typed
-programs. Furthermore, most type-level bugs do not survive after decent tsting and 
+chrdimo: There is no existing catalogue of bugs in gradually typed
+programs. Furthermore, most type-level bugs do not survive after decent testing and 
 thus do not make it to code repos. There are a few anecdotes from the literature but they are mostly
-artificial. Hence, since we need a large number of buggy scenarions for a
+artificial. Hence, since we need a large number of buggy scenarios for a
 meaningful study, a synthetic approach is the only way forward. In
-addtion, with our mutation we can tune the set of scenarios to get a
+addition, with our mutation we can tune the set of scenarios to get a
 diverse set of bugs.
@@
 fails and those where it succeeds but the blame trail has length greater
-than one. This information is in the paper byt If you think it would be
+than one. This information is in the paper but If you think it would be
 useful, we point it out. Moreover, the experiment considers 10 different
-modes of the rational programmer spread across 3 different gradaul typing
+modes of the rational programmer spread across 3 different gradual typing
 systems.
@@
 benchmarks, a blame strategy is superior that exceptions and that natural
-and transient have comparable efficacy. The latter is surprisng given that
-theoretical models (and our intiail expectations) point to the opposite.
-Thus a side-effect of our experiment is a call to theoriticians for models
-that are atuned with practice.
+and transient have comparable efficacy. The latter is surprising given that
+theoretical models (and our initial expectations) point to the opposite.
+Thus a side-effect of our experiment is a call to theoreticians for models
+that are attuned with practice.
 
@@
 chrdimo: Section 8.2 does not discuss a bug in Transient but rather a
-limitation of its current design that we discovered exactly becasue of the
-experiement.
+limitation of its current design that we discovered exactly because of the
+experiment.
 
@@
 
-chrdimo: We would like to point out that the resutls of the comparison
+chrdimo: We would like to point out that the results of the comparison
 between the tree systems are only one part of the contribution of the
@@
 
-chrdimo: Our experiemnt does have a baseline. The random mode of the
+chrdimo: Our experiment does have a baseline. The random mode of the
 rational programmer serves as the ``control'' of our experiment. 
@@
 
-chrdimo: The transient semantics provide no interpratation for the blame
+chrdimo: The transient semantics provide no interpretation for the blame
 set. We picked the first and last because they match the intuitive
@@
 other options are unreasonable. We just needed to make a reasonable choice 
-and keep the duration of the experiemtn within reasonbale limits.
+and keep the duration of the experiment within reasonable limits.
 
@@
 (b/c) the first or last component, respectively, blamed by Transient,
-or (d) an exception raised by the underyling language --- conclude
+or (d) an exception raised by the underling language --- conclude
 with the actual bug. Results show that Natural, Transient-First, and
@@
 
-Overall, I think this is a useful reference for further usability work
-on contracts and gradual type systems, and am generally in favor of
-acceptance.
+chrdimo: A side note here: Transient Typed Racket is faster than Natural
+Typed Racket when we turn off blame. Exactly as a sanity check that this is not 
+an artifact of our implementation we have coded up benchmarks in Python
+and we confirmed that in Reticulated the performance of transient with blame is 
+even worse than what we have observed in Typed Racket.
 
@@
 This required research, will be published elsewhere
+
+chrdimo:
+
+We disagree that implementing Transient Typed Racket is mere engineering.
+Typed Racket comes with a way more sophisticated type system than
+Reticulated and there are a lot of points around type checks and blame
+that neither Reticulated nor the Transient model provide answers for.  We
+are happy to share with the chair a recent dissertation chapter that
+describes all the ideas and work that went into Transient Typed Racket.
+
+We believe that adding a formal model to the paper would distract
+from the main contribution of the paper as there are already multiple
+publications on the semantics of Transient, Natural and Erasure and at
+least one that compares them formally [Greenman et al. OOPSLA 2019]. That said, we
+agree that we should improve the presentation in section 3 to introduce
+more context.
 
@@
 
+chrdimo: While we agree that there is overlap between our mutators
+and those from Lazarek's POPL work (and the common set of mutators where
+the latter come from), we disagree that the construction of custom
+mutators is straightforward. The sixteen we include in the paper are 
+not the original set we tried and others that seemed like good candidates
+have proven to be ineffective. For instance replacing car with cdr should
+lead to type-level bugs however none of them are interesting as we define
+interesting in the paper. Similarly, swapping mutable vectors and hashes
+with immutable ones is an interesting mutation in the traditional sense
+and should lead to a type error in Typed Racket as its type system
+distinguishes the two, however we discovered that again is not interesting
+in our setting. In general, there is no methodology for constructing a new
+set of mutators, so narrowing down the ones in the paper required
+that we first characterize precisely what makes a bug interesting and then 
+rounds of trials to end up with a set of mutators that leads to a 
+sufficiently large set of interesting diverse bugs.
+
 *Debugging Strategies (Section 5)*
@@
   TODO: the exception modes are a key part of the analysis framework, enabling understanding of the value of blame separately from checking
+
+chrdimo:
+
+We disagree that section 5 boils down to picking the first and last modes
+for Transient. The core of this section is the analysis of of the
+  different systems and their unification in a common precise framework as
+  modes. Because of the common substrate of blame trails modes enable
+  the comparison between the different systems even when sampling is
+  involved.  Furthermore we use modes to introduce in a uniform manner in
+  the framework baselines that help us isolate confounding factors from
+  the actual effectiveness of blame. This si exactly the role of the
+  exception modes for Natural and Transient. Those two perform checks at
+  runtime and the exception modes help us determine what part of the
+  success of Natural and Transient is due to the extra checks or due to
+  blame (and also whether blame masks useful information from exceptions).
+  For example, we see that Natural without blame but with the same checks
+  outperforms Erasure in ~25% of the scenarios (figure 6). Blame add
+  another ~11% on top of that. For Transient there is a small percentage
+  of scenarios where exceptions due better than blame but still blame
+  improves over exceptions by ~11% compared to Erasure.
+
+chrdimo: As a final note, even though each of the above points
+individually may not have sufficient technical depth for a separate
+contribution (which we do not claim), together they exhibit the creative
+steps that are necessary to lift the scope and magnitude of the idea
+from Lazarek's POPL paper on evaluating blame from a single behavioral
+contract system to a framework for evaluating and comparing different
+gradual typing systems.
 
@@
 L1304: "Problem is ...."
+
+chrdimo: Thank you. We will fix these.
 

### 2021-02-09 writing/reviews/pldi-rebuttal-notes.md
Incorporate responses to reviewers D and E into rebuttal from notes

@@
 
+  addressed
   TODO: We should clarify that checks alone is faster, but adding blame is slower (bc blame map is linear).
@@
 
+  addressed
   TODO: this is a minor part of the paper (not a result), there is a large gap between the model and what's needed to scale to a full lang -- it's not clear from the simple model how to design check insertion and blame tracking for new features.
@@
 
+  addressed
   TODO: the analytical part is important: we described all of these things in a common framework, allowing for comparison
@@
 
+  addressed
   TODO: the exception modes are a key part of the analysis framework, enabling understanding of the value of blame separately from checking

### 2021-02-09 writing/reviews/pldi-rebuttal.md
Incorporate responses to reviewers D and E into rebuttal from notes

@@
 
-Overall merit
--------------
-B. I support accepting this paper but will not champion it.
+> Given that the experiments are developed and run in the context of the
+> large, real-world Racket ecosystem, I wonder whether about the chance
+> of incidental implementation choices or bugs affecting this result.
 
-Reviewer expertise
-------------------
-Y. **Knowledgeable** = "I follow the literature on this paper's topic(s)
-   but may have missed some relevant developments"
+A side note here: Transient Typed Racket is faster than Natural Typed Racket
+when we turn off blame. Exactly as a sanity check that this is not an artifact
+of our implementation we have coded up benchmarks in Python and we confirmed
+that in Reticulated the performance of transient with blame is even worse than
+what we have observed in Typed Racket.
 
-Paper summary
--------------
-Several contract and gradual type systems provide mechanisms for
-assigning blame: When encountering a run-time error signifying the
-violation of a contract or type signature, which pieces of code should
-be scrutinized by the programmer in seeking a fix?
+> *Natural and Transient Under the Same Roof (Section 3)*
+>
+> This seems to be "only" an engineering challenge.
 
-In a recent POPL 2020 paper, Lazarek et al. provide the first attempt
-to evaluate if and how blame assignment may help in actual debugging.
-In that paper, bugs are systematically injected (via syntactic
-mutation) into a suite of Racket programs, and the corresponding blame
-trails are analyzed to see whether they, by writing more precise
-contracts to "shift" blame from the assigned component, eventually
-lead to the actual bug.
+We disagree that implementing Transient Typed Racket is mere engineering.
+Typed Racket comes with a way more sophisticated type system than
+Reticulated and there are a lot of points around type checks and blame
+that neither Reticulated nor the Transient model provide answers for.  We
+are happy to share with the chair a recent dissertation chapter that
+describes all the ideas and work that went into Transient Typed Racket.
 
-The current paper extends the POPL 2020 methodology from contracts in
-Racket to gradual (or migratory) typing in Typed Racket, in several
-steps:
+We believe that adding a formal model to the paper would distract
+from the main contribution of the paper as there are already multiple
+publications on the semantics of Transient, Natural and Erasure and at
+least one that compares them formally [Greenman et al. OOPSLA 2019]. That said, we
+agree that we should improve the presentation in section 3 to introduce
+more context.
 
-First, to complement the "Natural" higher-order contract system of
-(Typed) Racket, this work implements a "Transient" blame assignment
-strategy, which unlike Natural inserts only shallow run-time checks.
-(To date, Transient has appeared only in the separate design and
-implementation of Reticulated Python, which prevents a head-to-head
-comparison as desired in this work.)
+> *Custom Mutators (Section 4)*
+> The paper suggests that the mutators required here are significantly
+> different and more subtle, but this does not seem to be the case when
+> comparing Figure 3 and Table 3 of the POPL 2020 paper.
+> ...
 
-Second, compared to POPL 2020, new mutators are defined to inject the
-kinds of type-mismatch bugs that may be caught by the gradual type
-system and corresponding run-time checks and blame assignment.
+While we agree that there is overlap between our mutators and those from
+Lazarek's POPL work (and the common set of mutators where the latter come from),
+we disagree that the construction of custom mutators is straightforward. The
+sixteen we include in the paper are not the original set we tried and others
+that seemed like good candidates have proven to be ineffective. For instance
+replacing car with cdr should lead to type-level bugs however none of them are
+interesting as we define interesting in the paper. Similarly, swapping mutable
+vectors and hashes with immutable ones is an interesting mutation in the
+traditional sense and should lead to a type error in Typed Racket as its type
+system distinguishes the two, however we discovered that again is not
+interesting in our setting. In general, there is no methodology for constructing
+a new set of mutators, so narrowing down the ones in the paper required that we
+first characterize precisely what makes a bug interesting and then rounds of
+trials to end up with a set of mutators that leads to a sufficiently large set
+of interesting diverse bugs.
 
-Third, the paper runs a large experiment on 10 benchmark programs,
-inserting mutations and analyzing whether the blame trails produced by
-each debugging mode --- following (a) the component blamed by Natural,
-(b/c) the first or last component, respectively, blamed by Transient,
-or (d) an exception raised by the underyling language --- conclude
-with the actual bug. Results show that Natural, Transient-First, and
-Transient-Last all produce short successful blame trails (typically of
-length one or two) on most debugging tasks, and that they are all
-similarly effective with no clear winner for all tasks.
+> *Debugging Strategies (Section 5)*
+>
+> These also seem to be overemphasized. The main design choice seems to
+> be in how to use Transient blame, which reports multiple components,
+> and picking the first and last are pretty intuitive.
+>
+> I'm also not sure why the exception mode is needed in addition to
+> Erasure (I was confused in Section 2 L204-215 and Section 5.1 L732).
+> Indeed, Section 5.3 defines the Erasure mode the follow the Natural
+> exceptional mode.
 
-Comments for author
--------------------
-This work advances a compelling approach to evaluating a usability
-question through a systematic exploration of synthetic bugs and fixes.
-The large-scale experiment is impressive, and I find both of the main
-results surprising:
+We disagree that section 5 boils down to picking the first and last modes for
+Transient. The core of this section is the analysis of of the different systems
+and their unification in a common precise framework as modes. Because of the
+common substrate of blame trails modes enable the comparison between the
+different systems even when sampling is involved. Furthermore we use modes to
+introduce in a uniform manner in the framework baselines that help us isolate
+confounding factors from the actual effectiveness of blame. This si exactly the
+role of the exception modes for Natural and Transient. Those two perform checks
+at runtime and the exception modes help us determine what part of the success of
+Natural and Transient is due to the extra checks or due to blame (and also
+whether blame masks useful information from exceptions). For example, we see
+that Natural without blame but with the same checks outperforms Erasure in ~25%
+of the scenarios (figure 6). Blame add another ~11% on top of that. For
+Transient there is a small percentage of scenarios where exceptions due better
+than blame but still blame improves over exceptions by ~11% compared to Erasure.
 
-1. That Natural is not much more effective than Transient. I thought
-the shallow checks of Transient would certainly be less effective than
-the full checks of Natural. It will be interesting to see if and how
-this holds up in subsequent usability studies, both with synthetic and
-real users and tasks.
+As a final note, even though each of the above points individually may not have
+sufficient technical depth for a separate contribution (which we do not claim),
+together they exhibit the creative steps that are necessary to lift the scope
+and magnitude of the idea from Lazarek's POPL paper on evaluating blame from a
+single behavioral contract system to a framework for evaluating and comparing
+different gradual typing systems.
 
-2. That Transient --- as implemented in Typed Racket --- is much
-slower than Natural. This also goes against intuition, as well as
-previous results reported about Reticulated Python (but apparently
-there were flaws in those experiments, as noted in this paper). Given
-that the experiments are developed and run in the context of the
-large, real-world Racket ecosystem, I wonder whether about the chance
-of incidental implementation choices or bugs affecting this result.
+> *Additional Comments and Typos*
+>
+> L69: I wish here, and elsewhere, the comparison to [13] had been made
+> more explicit. As discussed above, there seem to be more similarities
+> than suggested.
+> L92: "use of _a_ higher-order contract system"
+> L155: "authors extensive"
+> L253: "exports _it_ as"
+> L280: Missing `untyped-`
+> L314: " . "
+> L382: "Dyn,"
+> Fig 2: Presumably authors were included to indicate the variety of
+> sources, but I'm not sure that was necessary.
+> L597: "make us_e_ of"
+> L664: "described in _Section_ 4"
+> L980: "Figure 5.4" ==> "Figure 6"
+> L986: "2.8%" ==> "2.18%" ?
+> L1017: "33%" ==> "26%" ?
+> L1304: "Problem is ...."
 
-  TODO: We should clarify that checks alone is faster, but adding blame is slower (bc blame map is linear).
-
-Overall, I think this is a useful reference for further usability work
-on contracts and gradual type systems, and am generally in favor of
-acceptance.
-
-At the same time, however, I believe the paper overstates the novelty
-and challenges to "bring [the POPL 2020 approach] to the world of
-gradual typing" (L1252). Comments about each of the three main
-components of the methodology in turn below.
-
-*Natural and Transient Under the Same Roof (Section 3)*
-
-This seems to be "only" an engineering challenge.
-
-And the presentation in this section is rather informal. As someone
-with passable knowledge of Natural and Transient, I would have liked
-to see a concise formalism with clear "knobs" for choosing between
-blame assignment strategies. Trying to imagine readers with less
-familiarity, I wonder how well this discussion serves as a primer.
-
-  TODO: this is a minor part of the paper (not a result), there is a large gap between the model and what's needed to scale to a full lang -- it's not clear from the simple model how to design check insertion and blame tracking for new features.
-This required research, will be published elsewhere
-
-*Custom Mutators (Section 4)*
-
-The paper suggests that the mutators required here are significantly
-different and more subtle, but this does not seem to be the case when
-comparing Figure 3 and Table 3 of the POPL 2020 paper.
-
-Of the 12 (of 16) gradual-typing related mutators listed in Figure 3:
-
-* `constant`: Similar to previous, but now with a type error
-* `deletion`: Similar to previous
-* `public`: Similar to previous "hide-method"
-* Four others involving swapping identifiers: Similar in spirit to
-"swap argument" in previous
-
-Of the 4 (of 16) Typed Racket related mutators:
-
-* `arithmetic`: Simpler version of previous
-* `boolean`: Same as in previous
-* `negate-cond`: Same as in previous
-
-So there is quite a bit of similarity. And on the flip side, why are
-not all (8) mutators from the POPL 2020 experiment included here?
-
-These questions are not so important in an absolute sense; _some_
-specific mutators need to be chosen to generate some bugs, and these
-seem to work fine. But they are salient given how much the paper
-emphasizes the new mutators as a contribution.
-
-*Debugging Strategies (Section 5)*
-
-These also seem to be overemphasized. The main design choice seems to
-be in how to use Transient blame, which reports multiple components,
-and picking the first and last are pretty intuitive.
-
-  TODO: the analytical part is important: we described all of these things in a common framework, allowing for comparison
-
-I'm also not sure why the exception mode is needed in addition to
-Erasure (I was confused in Section 2 L204-215 and Section 5.1 L732).
-Indeed, Section 5.3 defines the Erasure mode the follow the Natural
-exceptional mode.
-
-  TODO: the exception modes are a key part of the analysis framework, enabling understanding of the value of blame separately from checking
-
-*Additional Comments and Typos*
-
-L69: I wish here, and elsewhere, the comparison to [13] had been made
-more explicit. As discussed above, there seem to be more similarities
-than suggested.
-
-L92: "use of _a_ higher-order contract system"
-
-L155: "authors extensive"
-
-L253: "exports _it_ as"
-
-L280: Missing `untyped-`
-
-L314: " . "
-
-L382: "Dyn,"
-
-Fig 2: Presumably authors were included to indicate the variety of
-sources, but I'm not sure that was necessary.
-
-L597: "make us_e_ of"
-
-L664: "described in _Section_ 4"
-
-L980: "Figure 5.4" ==> "Figure 6"
-
-L986: "2.8%" ==> "2.18%" ?
-
-L1017: "33%" ==> "26%" ?
-
-L1304: "Problem is ...."
+Thank you. We will fix these.
 
@@
 
-Overall merit
--------------
-C. I would not accept this paper but will not argue strongly against
-   accepting it.
-
-Reviewer expertise
-------------------
-Z. **Some familiarity** = "I have a passing knowledge of the topic(s) but
-   do not follow the relevant literature"
-
-Paper summary
--------------
-This paper seeks to establish (or refute) the benefit of blame tracking in gradually typed programming environments.
+> This paper seeks to establish (or refute) the benefit of blame tracking in gradually typed programming environments.
 It takes several Typed Racket benchmarks and considers multiple approaches to blame tracking.
@@
 
-Comments for author
--------------------
-I commend the authors for asking the question.
-Their method is clearly the product of much careful thought.
-I understand the difficulty in doing experiments about errors that tend not to hit 'real' source code repositories, and I appreciate the effort taken to re-create the different approaches to blame in a comparable setting.
+The primary intended contribution of this paper is the methodology, and the
+experimental results serve as secondary contributions that validate the
+methodology's utility. We will rephrase the introduction to further emphasize
+that focus. Furthermore, we would like to emphasize that the two key findings in
+the experimental results are that 1) both Natural and Transient's dynamic type
+enforcement with blame tracking improve significantly over Erasure, and 2)
+Natural's blame tracking is not significantly better than Transient's, despite
+the theoretical indications to the contrary.
 
-However, I don't find the results to be as conclusive as the authors state.
+> However, I don't find the results to be as conclusive as the authors state.
 Aside from the Erasure case, whose relevance I found unclear (see below), conclusions rest on some small-valued "more useful than" percentages.
@@
 I appreciate the discussion of threats to validity.
+
   TODO: misunderstanding: the results are not at all "what we'd expect" based on the theory
 
-The presentation of the results is very much around aggregates and summaries.
-Indeed the whole method is about having run a huge compute job over a large number of variants.
-Perhaps it's paranoia but I'd be interested to see some specific examples walked through by hand (in the paper) and some smaller sample manually classified (as results).
-That would add an extra sanity check that the metrics do correspond to some meaningful reality.
+The status quo of what we would expect based on the current theory of Natural
+and Transient is that 1) either kind of blame tracking should be better than
+none, and 2) Natural should provide significantly better blame information than
+Transient and Erasure. Our results validate the first expectation that blame
+provides benefits over plain exceptions. Surprisingly, our results contradict
+the second expectation and suggest that in fact Natural and Transient offer
+comparable blame information.
 
-About Erasure: if I'm reading this correctly, the key thing here is that in Typed Racket, the set of static types is more expressive than the set of dynamic types.
+> About Erasure: if I'm reading this correctly, the key thing here is that in Typed Racket, the set of static types is more expressive than the set of dynamic types.
 For example, there is a static notion of non-negative integer that is distinct from plain integer.
@@
 
-It's interesting to note that there seems to be no convincing analogous experimental or modelling-based justification for plain old static typing, i.e.
+TODO: misunderstanding about gradual typing as a whole, and the point of the exception modes
+
+The idea of the Erasure semantics is not related to the expressiveness of the
+type system or how the types relate to kinds of runtime values. Rather, the key
+idea is that Erasure does not enforce type annotations with dynamic checks at
+all. Running a gradual program using the Erasure semantics is the same as
+stripping all annotations and running the resulting (completely dynamic)
+program. Thus, Erasure programs can only raise exceptions from either the checks
+performed by the language's primitive operations (e.g. `+`), or from the program
+itself raising an exception. In contrast, the other two semantics do enforce
+type annotations dynamically; both Natural and Transient insert dynamic checks
+verifying that the shape of values at runtime match their static annotations.
+Thus, in addition to exceptions, gradual programs using these semantics may
+raise a second kind of error: a dynamic type check failure. These dynamic errors
+can include blame information if the dynamic type checks are designed to track
+blame, or they can forego blame and just offer a stack trace. In the case of
+Natural, these two options are captured by the "Natural" and "Natural
+exceptions" modes, respectively. Thus the "Erasure" mode represents a completely
+different kind of checking (and thus error reporting) compared to both the
+"Natural" and "Natural exceptions" modes (the two of which share the same kind
+checking, but different error reporting).
+
+> It's interesting to note that there seems to be no convincing analogous experimental or modelling-based justification for plain old static typing, i.e.
 showing that it somehow presents a net gain to the programmer.
@@
 So I'd be interested to hear from the author(s) if they have any more arguments that (1) I've underestimated the results' conclusiveness, or (2) this is a novel/interesting family of methods that might be pursued more widely, or (3) that deeper experiments building on these ideas might yield more compelling insights.
+
+(Matthias) As posed, the question expresses a significant mis-characterization
+of the submission. It is not the benefits of a gradual type system that are in
+doubt. The experimental setup keeps the type system constant but allows to
+answer the question which of several run-time checking regimes (for enforcing
+type consistency) provides the best explanatory messages in case of violations.
+Hence the analogous question for purely static type systems would ask which of
+several reporting schemes provides the best explanatory message for type-errors.
+In the case of simply typed languages and even language with local type
+inference, this question is basically meaningless. In the case of languages with
+HM type inference, the question has been implicitly raised for four decades with
+the development of alternative ways of finding the source of inference
+conflicts.
+
+If the reviewer is indeed interested in the question of whether type systems
+help programmers---a question that this submission does _not_ ask---the recent
+OOPSLA literature contains several user studies. A simple Google query will
+suggest a short list of these publications.
+
+  TODO ll: I added responses (after ⟶) to anything below that seems to warrant more than "Thanks, we'll fix it".
 

### 2021-02-09 writing/reviews/pldi-rebuttal.md
make it concise!

@@
 
-First of all, we thank all of the reviewers for reading our submission, asking
-critical questions, and offering constructive feedback. Our responses to all
-comments and questions follow, in which we quote each comment/question and then
-respond below.
+Thank you to al reviewers for the helpful feedback. 
+critical questions, and offering constructive feedback. Our rebuttal combines each comment/question with a response.
 

### 2021-02-09 writing/reviews/pldi-rebuttal-high-level-first.md
Add new version of rebuttal with opening high level response

--- 2021-02-09 writing/reviews/pldi-rebuttal.md
@@
 
-Thank you to al reviewers for the helpful feedback. 
-critical questions, and offering constructive feedback. Our rebuttal combines each comment/question with a response.
+Thank you to all reviewers for the helpful feedback.
+Our rebuttal starts with a high level response, then responds to each remaining comment/question in turn.
 
-Review #66A
-===========================================================================
-> - Could you comment a bit more on what makes a mutator "good"? (other than
->   experimentally seeing that it is good.) Perhaps, what would be a really bad
->   mutator for evaluating blame assignment?
+High level response
+-------------------
 
-A bad mutator results in a type-level bug that is trivial to locate. For
-instance, a bug that leads to a type-level error during the evaluation of the
-component that contains it is trivial to locate. In contrast, a bug that is
-non-trivial to locate is one that causes the buggy component to evaluate without
-an error but produce instead a value of the wrong type. An interesting bug is
-one where the value with the wrong type manages also to cross undetected to at
-least another component before a type-level check flags it. Good mutators should
-be able to produce these interesting bugs.
+As review `D` observes, the results of our experiment are surprising;
+they provide clear evidence that Natural and Transient blame have comparable utility in practice.
+This conclusion flies in the face of theoretical predictions of Natural's superiority.
+The tools of gradual typing theory don't immediately explain why we would see this disparity, perhaps because many of the advanced features of Typed Racket's type system haven't appeared in models that compare Natural and Transient: polymorphism, row types, mixins, and more.
+The disparity between the theory's predictions and our experiment's conclusions demands two responses:
+we must re-evaluate and extend the theory to improve its predictive power,
+and we must perform more extensive experimentation to better understand the extent and nature of the divergence.
+The latter response can already be addressed by applying the methodology we propose, and this applicability is why the methodology is the paper's primary contribution.
 
-> - I don't understand where the 16,800 number comes from. Is it an exhaustive
->   enumeration of all possible mutations? (Or the number of mutants constructed
->   in X amount of time?)
+The methodology is the main contribution because it provides novel insights for evaluating blame in gradual typing along several dimensions.
+- bringing different semantics into the same implementation
+- designing custom mutators
+- designing the framework for understanding and comparing utility of different strategies in terms of modes, and with many baselines against which to interpret the results
 
-It is an exhaustive enumeration. We will clarify the prose.
+Then, clarify and strengthen position of reviewer A, using those comments as hooks to argue further against other reviews' criticisms.
+In particular,
+- Novelty of designing mutators
+- Size and standardness of the benchmarks
+- ... I think that's all?
 
-> - How large are the programs in the benchmark suite?
+Points already drafted to incorporate into the high level response
+==================================================================
 
-Here are the LOC counts and number of components for each benchmark. We will add
-the information to figure 2.
-
-  benchmark    | LOC  | components
-  -------------|------|-----------
-  acquire      | 1941 | 9
-  gregor       | 2336 | 13
-  kcfa         | 328  | 7
-  quadT        | 7396 | 14
-  quadU        | 7282 | 14
-  snake        | 182  | 8
-  suffixtree   | 1500 | 6
-  synth        | 871  | 10
-  take5        | 465  | 8
-  tetris       | 280  | 9
-
-> # Minor Comments
-> - L38: Should there have been a colon? Is the text "This contrast... " and
->   challenge quote from [3]? Who is saying what here...
-
-No, the quote is not from [3]. We will rephrase to clarify that it is our
-statement of the research question.
-
-> - L104: I was confused by the word "Next". Maybe instead: We now ..."
-
-Thank you. We will reword.
-
-> - L268: Grammar issue.
-> - L314: Floating period.
-> - L441: Add line numbers (or other metric of size).
-> - L980: Figure 5.4??
-
-We will fix these. Thanks.
-
-> - L956: 30,000 hours of compute time is 1,250 days. I assume that the CPU has
->   many cores. Could you put that number?
-
-Sure. Each CPU has 28 doubly-threaded cores.
-
-> - L991: Figure 6: It took me some time to understand the figure. I wonder if
->   there is a better way to represent the same data?
-
-We have reached the same conclusion and we are considering alternatives that
-will make it easier to see how two modes compare to each other, such as a series
-of two-sided bar charts.
-
-Review #66B
-===========================================================================
+Review #66B (for surprisingness that Natural is not >> Transient):
+--------------------------------------------------
 > The experiment results show that there are debugging scenarios where
@@
 
-> Theoretical gradual type systems usually allow programmers to control
-> the precision of type annotations in a fine-grained way with the Dyn
-> type rather than distinguishing typed and untyped modules. Would the
-> same evaluation method be able to directly appliable to blame in such systems?
-
-Yes. Incorporating type Dyn boils down to considering a hierarchy of types for
-each component rather than a single type. At the same time, our method is
-parametric to the "size" of a component and adapts in a straightforward manner
-to a setting where each definition can be considered a component instead of a
-whole module. Of course both of these adjustments lead to significantly larger
-scenario lattices than the ones we investigate, so keeping the experiment
-computationally feasible would require more aggressive sampling.
-
-> Please explain what impedance mismatches are in the paper. Is it
-> simply type mismatches?
-
-An impedance mismatch denotes a type annotation that is detected to disagree
-with the actual type of the annotated code. We will adjust the prose to clarify.
-
-> The paper states that the effort distribution of the random mode
-> follows a normal distribution, as expected. Please explain why a
-> normal distribution is a reasonable expectation. ...
-> This point does not affect the contribution of the paper, but it
-> is not clear from the paper.
-
-You are right that for a given trail of length $n$ the probability that the
-random rational programmer discovers the buggy component on every step is $1/n$.
-The missing piece is that we randomly sample from the set of scenarios -- and by
-extension the set of trails. So the shape of the random programmer's effort
-matches our sampling distribution. That said, as you observe, the distribution
-of the random programmer's effort is not relevant to our results, so we will
-remove the characterization.
-
-> * page 9 line 980: Figure 5.4 -> Figure 6
-> * page 10 Figure 7: Null -> Random
-> * page 11 line 1103: ( 770 -> (770
-
-Thank you. We will fix all these.
-
-Review #66C
-===========================================================================
-> But while there is a lot to like about the paper, it also has some shortcomings. The suite of programs used for the evaluation is fairly small (c.f., the SIGPLAN checklist).
-
-Could you clarify what you mean by fairly small with respect to the SIGPLAN
-checklist? We are the first to conduct an empirical study for blame and gradual
-typing, so we repurposed the standard performance benchmark suite for Typed
-Racket -- in particular, its largest and most complex benchmarks. Our inspection
-of the benchmarks indicates that they are representative set of Racket programs
-in terms of features and structure.
-
-> The faults are introduced using synthetic mutations that may or may not correspond to the kinds of errors that arise in practice.
-This can be seen from the results in Figure 7 -- the lengths of the "trails" is quite small.
-
-There is no existing catalog of bugs in gradually typed programs. Furthermore,
-most type-level bugs do not survive after decent testing and thus do not make it
-to code repos. There are a few anecdotes from the literature but they are mostly
-artificial. Hence, since we need a large number of buggy scenarios for a
-meaningful study, a synthetic approach is the only way forward. In addition,
-mutation allows us to tune the set of scenarios to get a diverse set of bugs.
-
-> While the study is mostly well done, it only considers two blame assignment strategies.
-> For example, it would be interesting to consider the "omniscient programmer" to get a bound on the value of blame. But this is not done.
-
-Could you elaborate what you mean by "omniscient programmer"? If this is the
-programmer that always makes the best choice, i.e. finds the bug at the first
-try, then this programmer is trivial. It has 100% success rate and a constant
-blame trail length of 1. Thus, comparing its success to any other strategy
-amounts to counting the number of scenarios where the other strategy fails;
-comparing effort is similarly trivial. This information is in the paper
-implicitly, but we can certainly add an explicit mention if it would be useful.
-
-Secondly, we would like to point out that the experiment considers then
-different modes of the rational programmer spread across three different gradual
-typing systems.
-
+Review #66C (for conclusiveness of results):
+--------------------------------------------------
 > And unfortunately the results are inconclusive -- Section 8 says "our results call for a deeper understanding of the two models for blame".
@@
 
-> Also, the published Transient scheme seems to have a minor bug, which might affect the outcomes when fixed (Section 8.2).
+Review #66E (for conclusiveness and surprisingness of results):
+--------------------------------------------------
+> However, I don't find the results to be as conclusive as the authors state.
+Aside from the Erasure case, whose relevance I found unclear (see below), conclusions rest on some small-valued "more useful than" percentages.
+These are pretty hard to interpret, because the "usefulness" metrics necessarily build in a lot of simplifying assumptions.
+E.g.
+in 5.4, the "percentage of scenarios where ...
+is more effective" doesn't account for *extent* of difference, and the trail-length "programmer effort" metric in 5.5 is also unlikely to be a great proxy.
+These limitations are understandable, method-wise, but overall the results are fairly null...
+they let us continue believing "what we'd expect" but not with any great sense of added confidence.
+That doesn't entirely diminish the work, of course, but I'd say the current write-up overplays its hand a little.
+I appreciate the discussion of threats to validity.
 
-Section 8.2 does not discuss a bug in Transient but rather a limitation of its
-current design that we discovered exactly because of the experiment.
+  TODO: misunderstanding: the results are not at all "what we'd expect" based on the theory
 
-> Most if not all of these shortcomings are discussed in the paper, which is much appreciated. But they do risk undermine the value of the contributions to some degree.
+The status quo of what we would expect based on the current theory of Natural
+and Transient is that 1) either kind of blame tracking should be better than
+none, and 2) Natural should provide significantly better blame information than
+Transient and Erasure. Our results validate the first expectation that blame
+provides benefits over plain exceptions. Surprisingly, our results contradict
+the second expectation and suggest that in fact Natural and Transient offer
+comparable blame information.
 
-We would like to point out that the results of the comparison between the three
-systems are only one part of the contribution of the paper. The other part is
-the method itself. The benchmarks, source of interesting bugs, and rational
-programmer modes are all parameters that can be instantiated differently without
-affecting either the recipe we discuss at the end of section 2 or the analytical
-framework from section 5 that underlies our experiment.
+Review #66E (for main contribution being method):
+--------------------------------------------------
+> This paper seeks to establish (or refute) the benefit of blame tracking in gradually typed programming environments.
+It takes several Typed Racket benchmarks and considers multiple approaches to blame tracking.
+The central idea is to use a fault injection method to create variant programs along two dimensions: type errors inserted according to some crafted heuristics, and static typing selectively removed on a component-by-component granularity.
+This creates a large number of variant programs which may produce run-time type errors with blame; this is then iterated to simulate a "rational programmer", i.e.
+such that blame is used to select where more static checking will be added next; at any given iteration this may or may not catch the injected error, and the number of iterations is counted as a proxy for effort.
+A further contribution is implementing the different forms of blame-tracking in the same framework so that they can be compared.
+The finding is that blame is found to offer some positive advantage in effort (over simply being guided by the stack trace of a run-time exception) in 9--13% of the initial sample of variants, with some minor differences between the forms of blame tracking tried.
 
-> To explain one comment above: I was surprised not to see a comparison against an "omniscient programmer" to establish a baseline (c.f., the SIGPLAN checklist).
-That is, model a blame assignment scheme in which blame could be placed on any component.
-And always pick the component that minimizes the length of the trail.
-Of course, this would be infeasible to implement in practice, but it would place an upper bound on the utility of blame.
+  TODO: misunderstanding: we propose a methodology
 
-Our experiment does have a baseline. The random mode of the rational programmer
-serves as the "control" of our experiment. Furthermore the exception modes each
-serve as further "controls" to distinguish the effect of checks from that of
-blame within each semantics.
+The primary intended contribution of this paper is the methodology, and the
+experimental results serve as secondary contributions that validate the
+methodology's utility. We will rephrase the introduction to further emphasize
+that focus. Furthermore, we would like to emphasize that the two key findings in
+the experimental results are that 1) both Natural and Transient's dynamic type
+enforcement with blame tracking improve significantly over Erasure, and 2)
+Natural's blame tracking is not significantly better than Transient's, despite
+the theoretical indications to the contrary.
 
-> Similarly, I wondered why the modes of the transient programmer only consider picking the first/last blame, and not other choices.
-Section 5.2 just says "our answer" is that there are at least "two reasonable options." Why are other choices not reasonable?
-
-The transient semantics provide no interpretation for the blame set. We picked
-the first and last because they match the intuitive interpretation of the blame
-set as a list. We didn't mean to imply that other options are unreasonable. We
-just needed to make a reasonable choice and keep the duration of the experiment
-within reasonable limits. We will clarify this point in the prose.
-
-Review #66D
-===========================================================================
-
-> Given that the experiments are developed and run in the context of the
-> large, real-world Racket ecosystem, I wonder whether about the chance
-> of incidental implementation choices or bugs affecting this result.
-
-A side note here: Transient Typed Racket is faster than Natural Typed Racket
-when we turn off blame. Exactly as a sanity check that this is not an artifact
-of our implementation we have coded up benchmarks in Python and we confirmed
-that in Reticulated the performance of transient with blame is even worse than
-what we have observed in Typed Racket.
-
+Review #66D (for novelty of the method):
+--------------------------------------------------
 > *Natural and Transient Under the Same Roof (Section 3)*
@@
 
+Review #66C (for baselines):
+--------------------------------------------------
+> While the study is mostly well done, it only considers two blame assignment strategies.
+> For example, it would be interesting to consider the "omniscient programmer" to get a bound on the value of blame. But this is not done.
+
+Could you elaborate what you mean by "omniscient programmer"? If this is the
+programmer that always makes the best choice, i.e. finds the bug at the first
+try, then this programmer is trivial. It has 100% success rate and a constant
+blame trail length of 1. Thus, comparing its success to any other strategy
+amounts to counting the number of scenarios where the other strategy fails;
+comparing effort is similarly trivial. This information is in the paper
+implicitly, but we can certainly add an explicit mention if it would be useful.
+
+Secondly, we would like to point out that the experiment considers then
+different modes of the rational programmer spread across three different gradual
+typing systems.
+
+> To explain one comment above: I was surprised not to see a comparison against an "omniscient programmer" to establish a baseline (c.f., the SIGPLAN checklist).
+That is, model a blame assignment scheme in which blame could be placed on any component.
+And always pick the component that minimizes the length of the trail.
+Of course, this would be infeasible to implement in practice, but it would place an upper bound on the utility of blame.
+
+Our experiment does have a baseline. The random mode of the rational programmer
+serves as the "control" of our experiment. Furthermore the exception modes each
+serve as further "controls" to distinguish the effect of checks from that of
+blame within each semantics.
+
+Review #66A
+===========================================================================
+> - Could you comment a bit more on what makes a mutator "good"? (other than
+>   experimentally seeing that it is good.) Perhaps, what would be a really bad
+>   mutator for evaluating blame assignment?
+
+A bad mutator results in a type-level bug that is trivial to locate. For
+instance, a bug that leads to a type-level error during the evaluation of the
+component that contains it is trivial to locate. In contrast, a bug that is
+non-trivial to locate is one that causes the buggy component to evaluate without
+an error but produce instead a value of the wrong type. An interesting bug is
+one where the value with the wrong type manages also to cross undetected to at
+least another component before a type-level check flags it. Good mutators should
+be able to produce these interesting bugs.
+
+> - I don't understand where the 16,800 number comes from. Is it an exhaustive
+>   enumeration of all possible mutations? (Or the number of mutants constructed
+>   in X amount of time?)
+
+It is an exhaustive enumeration. We will clarify the prose.
+
+> - How large are the programs in the benchmark suite?
+
+Here are the LOC counts and number of components for each benchmark. We will add
+the information to figure 2.
+
+  benchmark    | LOC  | components
+  -------------|------|-----------
+  acquire      | 1941 | 9
+  gregor       | 2336 | 13
+  kcfa         | 328  | 7
+  quadT        | 7396 | 14
+  quadU        | 7282 | 14
+  snake        | 182  | 8
+  suffixtree   | 1500 | 6
+  synth        | 871  | 10
+  take5        | 465  | 8
+  tetris       | 280  | 9
+
+> # Minor Comments
+> - L38: Should there have been a colon? Is the text "This contrast... " and
+>   challenge quote from [3]? Who is saying what here...
+
+No, the quote is not from [3]. We will rephrase to clarify that it is our
+statement of the research question.
+
+> - L104: I was confused by the word "Next". Maybe instead: We now ..."
+
+Thank you. We will reword.
+
+> - L268: Grammar issue.
+> - L314: Floating period.
+> - L441: Add line numbers (or other metric of size).
+> - L980: Figure 5.4??
+
+We will fix these. Thanks.
+
+> - L956: 30,000 hours of compute time is 1,250 days. I assume that the CPU has
+>   many cores. Could you put that number?
+
+Sure. Each CPU has 28 doubly-threaded cores.
+
+> - L991: Figure 6: It took me some time to understand the figure. I wonder if
+>   there is a better way to represent the same data?
+
+We have reached the same conclusion and we are considering alternatives that
+will make it easier to see how two modes compare to each other, such as a series
+of two-sided bar charts.
+
+Review #66B
+===========================================================================
+> Theoretical gradual type systems usually allow programmers to control
+> the precision of type annotations in a fine-grained way with the Dyn
+> type rather than distinguishing typed and untyped modules. Would the
+> same evaluation method be able to directly appliable to blame in such systems?
+
+Yes. Incorporating type Dyn boils down to considering a hierarchy of types for
+each component rather than a single type. At the same time, our method is
+parametric to the "size" of a component and adapts in a straightforward manner
+to a setting where each definition can be considered a component instead of a
+whole module. Of course both of these adjustments lead to significantly larger
+scenario lattices than the ones we investigate, so keeping the experiment
+computationally feasible would require more aggressive sampling.
+
+> Please explain what impedance mismatches are in the paper. Is it
+> simply type mismatches?
+
+An impedance mismatch denotes a type annotation that is detected to disagree
+with the actual type of the annotated code. We will adjust the prose to clarify.
+
+> The paper states that the effort distribution of the random mode
+> follows a normal distribution, as expected. Please explain why a
+> normal distribution is a reasonable expectation. ...
+> This point does not affect the contribution of the paper, but it
+> is not clear from the paper.
+
+You are right that for a given trail of length $n$ the probability that the
+random rational programmer discovers the buggy component on every step is $1/n$.
+The missing piece is that we randomly sample from the set of scenarios -- and by
+extension the set of trails. So the shape of the random programmer's effort
+matches our sampling distribution. That said, as you observe, the distribution
+of the random programmer's effort is not relevant to our results, so we will
+remove the characterization.
+
+> * page 9 line 980: Figure 5.4 -> Figure 6
+> * page 10 Figure 7: Null -> Random
+> * page 11 line 1103: ( 770 -> (770
+
+Thank you. We will fix all these.
+
+Review #66C
+===========================================================================
+> But while there is a lot to like about the paper, it also has some shortcomings. The suite of programs used for the evaluation is fairly small (c.f., the SIGPLAN checklist).
+
+Could you clarify what you mean by fairly small with respect to the SIGPLAN
+checklist? We are the first to conduct an empirical study for blame and gradual
+typing, so we repurposed the standard performance benchmark suite for Typed
+Racket -- in particular, its largest and most complex benchmarks. Our inspection
+of the benchmarks indicates that they are representative set of Racket programs
+in terms of features and structure.
+
+> The faults are introduced using synthetic mutations that may or may not correspond to the kinds of errors that arise in practice.
+This can be seen from the results in Figure 7 -- the lengths of the "trails" is quite small.
+
+There is no existing catalog of bugs in gradually typed programs. Furthermore,
+most type-level bugs do not survive after decent testing and thus do not make it
+to code repos. There are a few anecdotes from the literature but they are mostly
+artificial. Hence, since we need a large number of buggy scenarios for a
+meaningful study, a synthetic approach is the only way forward. In addition,
+mutation allows us to tune the set of scenarios to get a diverse set of bugs.
+
+> Also, the published Transient scheme seems to have a minor bug, which might affect the outcomes when fixed (Section 8.2).
+
+Section 8.2 does not discuss a bug in Transient but rather a limitation of its
+current design that we discovered exactly because of the experiment.
+
+> Most if not all of these shortcomings are discussed in the paper, which is much appreciated. But they do risk undermine the value of the contributions to some degree.
+
+We would like to point out that the results of the comparison between the three
+systems are only one part of the contribution of the paper. The other part is
+the method itself. The benchmarks, source of interesting bugs, and rational
+programmer modes are all parameters that can be instantiated differently without
+affecting either the recipe we discuss at the end of section 2 or the analytical
+framework from section 5 that underlies our experiment.
+
+> Similarly, I wondered why the modes of the transient programmer only consider picking the first/last blame, and not other choices.
+Section 5.2 just says "our answer" is that there are at least "two reasonable options." Why are other choices not reasonable?
+
+The transient semantics provide no interpretation for the blame set. We picked
+the first and last because they match the intuitive interpretation of the blame
+set as a list. We didn't mean to imply that other options are unreasonable. We
+just needed to make a reasonable choice and keep the duration of the experiment
+within reasonable limits. We will clarify this point in the prose.
+
+Review #66D
+===========================================================================
+
+> Given that the experiments are developed and run in the context of the
+> large, real-world Racket ecosystem, I wonder whether about the chance
+> of incidental implementation choices or bugs affecting this result.
+
+A side note here: Transient Typed Racket is faster than Natural Typed Racket
+when we turn off blame. Exactly as a sanity check that this is not an artifact
+of our implementation we have coded up benchmarks in Python and we confirmed
+that in Reticulated the performance of transient with blame is even worse than
+what we have observed in Typed Racket.
+
 > *Additional Comments and Typos*
@@
 ===========================================================================
-
-> This paper seeks to establish (or refute) the benefit of blame tracking in gradually typed programming environments.
-It takes several Typed Racket benchmarks and considers multiple approaches to blame tracking.
-The central idea is to use a fault injection method to create variant programs along two dimensions: type errors inserted according to some crafted heuristics, and static typing selectively removed on a component-by-component granularity.
-This creates a large number of variant programs which may produce run-time type errors with blame; this is then iterated to simulate a "rational programmer", i.e.
-such that blame is used to select where more static checking will be added next; at any given iteration this may or may not catch the injected error, and the number of iterations is counted as a proxy for effort.
-A further contribution is implementing the different forms of blame-tracking in the same framework so that they can be compared.
-The finding is that blame is found to offer some positive advantage in effort (over simply being guided by the stack trace of a run-time exception) in 9--13% of the initial sample of variants, with some minor differences between the forms of blame tracking tried.
-
-  TODO: misunderstanding: we propose a methodology
-
-The primary intended contribution of this paper is the methodology, and the
-experimental results serve as secondary contributions that validate the
-methodology's utility. We will rephrase the introduction to further emphasize
-that focus. Furthermore, we would like to emphasize that the two key findings in
-the experimental results are that 1) both Natural and Transient's dynamic type
-enforcement with blame tracking improve significantly over Erasure, and 2)
-Natural's blame tracking is not significantly better than Transient's, despite
-the theoretical indications to the contrary.
-
-> However, I don't find the results to be as conclusive as the authors state.
-Aside from the Erasure case, whose relevance I found unclear (see below), conclusions rest on some small-valued "more useful than" percentages.
-These are pretty hard to interpret, because the "usefulness" metrics necessarily build in a lot of simplifying assumptions.
-E.g.
-in 5.4, the "percentage of scenarios where ...
-is more effective" doesn't account for *extent* of difference, and the trail-length "programmer effort" metric in 5.5 is also unlikely to be a great proxy.
-These limitations are understandable, method-wise, but overall the results are fairly null...
-they let us continue believing "what we'd expect" but not with any great sense of added confidence.
-That doesn't entirely diminish the work, of course, but I'd say the current write-up overplays its hand a little.
-I appreciate the discussion of threats to validity.
-
-  TODO: misunderstanding: the results are not at all "what we'd expect" based on the theory
-
-The status quo of what we would expect based on the current theory of Natural
-and Transient is that 1) either kind of blame tracking should be better than
-none, and 2) Natural should provide significantly better blame information than
-Transient and Erasure. Our results validate the first expectation that blame
-provides benefits over plain exceptions. Surprisingly, our results contradict
-the second expectation and suggest that in fact Natural and Transient offer
-comparable blame information.
 

### 2021-02-10 writing/reviews/pldi-rebuttal-high-level-first.md
adding rebittal draft

--- 2021-02-09 writing/reviews/pldi-rebuttal.md
@@
 
-Thank you to al reviewers for the helpful feedback. 
-critical questions, and offering constructive feedback. Our rebuttal combines each comment/question with a response.
+Thank you to all the reviewers for the helpful feedback.
+
+We start off with a couple of common themes. Then we reply to all points
+individually.
+
+The conclusions from our experiment
+===========================================================================
+
+There are three clear major conclusions from our experiment:
+
+1) The Transient and Natural rational programmers are more effective than
+the Erasure one;
+
+2) Part of  the additional effectiveness of the former is due to runtime
+checks (that Erasure foregoes) but another part is due to blame;
+
+3) The effectiveness of the Natural and Transient rational programmers is
+comparable.
+
+The first two results confirm what the vast literature on gradual typing
+claims without systematic evidence: blame is useful. However, as review
+`D` astutely points out, the third conclusion is surprising in two
+different ways:
+
+(i) An important motivation for Transient is performance but our
+experiment provides evidence that performance issues, which stem from the
+design of Transient, have a negative effect on the effectiveness of blame.
+
+(ii) Theory claims that natural blame is more accurate than transient
+blame but our experiment provides evidence for the opposite;
+
+Thus our results point to two directions for further research in gradual
+typing:
+
+(i) Implementors can experiment with checks and blame to find a sweet spot
+between type soundness, blame effectiveness and performance. Our method
+provides an automatic tool to guide their exploration.
+
+(ii) The predictive power of gradual typing formal models (the reason
+d'etre of scientific models) fails when it comes to blame. This may be
+because the current models are too small and do not include crucial
+features or because the current models are inherently limited. In any case
+our results are a call to theoreticians to re-examine their assumptions
+about blame.  Our method provides an automatic tool to validate their
+models similar to the way the methods of experimental physicists help
+prove or disprove the  math of theoretical physicists.
+
+Challenges for our work compared to Lazarek's POPL paper
+===========================================================================
+
+Our work builds on the idea of Lazarek et al.'s paper that it is possible
+to evaluate the effectiveness of blame by simulating a debugging process.
+Both their work and our work is inspired by Greenman et al's work that
+uses a lattice of configurations to systematically explore variants of the
+same program with specifications of different precision (for performance
+reasons in Greenman et al).  Both the ideas of ``following`` blame while
+debugging and of gradually adding specifications to tweak the evaluation
+of the program originate from the practice of contracts (see the story in
+the intro of Lazarek's paper and similar anecdotes in Meyer's writings).
+
+That said,
+
+1) In the context of Lazarek et al.'s work interesting bugs  are those
+detectable by behavioral contracts. Hence, they can use off-the-shelf
+mutators. In fact, their method does not detect the overwhelming majority
+of bugs unless if contracts describe the full functional correctness of
+components.
+
+2) Lazarek et al.'s method does not isolate the effect of blame from that
+of checks or mere luck. Thus they cannot really conclude that blame is the
+reason for the success of the debugging process --- only that blame
+shifting works. In other words, their work is lacking proper `controls'.
+
+3) Lazarek et al. deal with a single contact system and examine a yes or
+no question: does blame-shifting find the bug for all debugging scenarios
+of a benchmark. They do not provide any guidance how to generalize to
+multiple systems with different notions of checks and/or blame in a way
+that allows comparisons.
+
+This brings us to the three problems we had to solve in order to design
+our experiment:
+
+1) Natural and Transient under the same roof (section 3).
+
+Implementing Transient Typed Racket is not mere engineering.  Neither
+Reticulated nor the Transient model provide answers about what shallow
+check corresponds to Typed Racket's sophisticated types (e.g.
+polymorphism, row types for mixins, occurrence typing) and what information
+is necessary to record in the blame map. Moreover in the process of the
+implementation we discovered issues with the published Transient model and
+the Reticulated model that we had to fix.  We are happy to share with the
+chair a recent dissertation chapter that describes all the ideas and work
+that went into Transient Typed Racket. We do not think that this paper is
+the appropriate place to describe the implementation in detail.
+
+2) Custom mutators (section 4).
+
+The identification of good mutators for gradual typing is not
+straightforward. A bad mutator results in a type-level bug that is trivial
+to locate. For instance, a bug that leads to a type-level error during the
+evaluation of the component that contains it is trivial to locate. In
+contrast, a bug that is non-trivial to locate is one that causes the buggy
+component to evaluate without an error but produce instead a value of the
+wrong type. Furthermore an interesting bug is one where the value with the
+wrong type manages also to cross undetected to at least another component
+before a type-level check flags it. Good mutators should be able to
+produce these interesting bugs in our benchmarks. Finally, they should
+produce interesting bugs that are detectable by all three systems so that
+we can try all systems on the same debugging scenarios.  In general, there
+is no methodology for constructing a new set of mutators that produce
+interesting bugs for any definition of interesting. So narrowing down the
+ones in the paper required rounds of trials to end up with a set of
+mutators that leads to a sufficiently large set of interesting and diverse
+bugs.
+
+The sixteen mutators we include in the paper are not the original set we
+tried and others that seemed like good candidates proved to be
+ineffective. For instance replacing car with cdr should lead to type-level
+bugs however none of them are interesting --- they blow up immediately.
+Similarly, swapping mutable vectors and hashes with immutable ones is an
+interesting mutation in the traditional sense and should lead to a type
+error in Typed Racket since its type system distinguishes the two, however
+it turns out such mutations  are not interesting in our setting.
+
+3) Modes of the rational programmer (section 5).
+
+A key step for realizing our experiment is the precise and unified
+description of debugging in different gradual typing system  as  the
+different modes of the rational programmer. Because of their common
+definitional substrate of blame trails, modes enable the comparison
+between the different systems even when sampling is involved.  Furthermore
+we use modes to introduce in the framework baselines that help us isolate
+confounding factors from the actual effectiveness of blame. This is
+exactly the role of the exception modes for Natural and Transient; the
+exception modes allow us to determine what part of the success of Natural
+and Transient is due to the checks or due to blame (and also whether blame
+masks useful information from exceptions). For example, we see that
+Natural without blame but with the same checks outperforms Erasure in ~25%
+of the scenarios (figure 6).  Blame add another ~11% on top of that. For
+Transient there is a small percentage of scenarios where exceptions due
+better than blame but still blame improves over exceptions by ~11%
+compared to Erasure. Finally, we introduce yet another mode, the random
+one, to serve as the baseline for the all modes and exclude that our
+observations are due to mere luck.
+
+As a final note, even though each of the above points individually may not
+have sufficient technical depth for a separate contribution (which we do
+not claim), together they exhibit the creative steps that are necessary to
+lift the scope and magnitude of the idea from Lazarek's POPL paper  from
+evaluating blame in a single behavioral contract system to a framework
+for evaluating and comparing blame in different gradual typing systems.
+
+Replies to individual points
+==================================================================
 
@@
 ===========================================================================
-> - Could you comment a bit more on what makes a mutator "good"? (other than
->   experimentally seeing that it is good.) Perhaps, what would be a really bad
->   mutator for evaluating blame assignment?
+> - Could you comment a bit more on what makes a mutator "good"? (other
+>   than experimentally seeing that it is good.) Perhaps, what would be a
+>   really bad mutator for evaluating blame assignment?
 
-A bad mutator results in a type-level bug that is trivial to locate. For
-instance, a bug that leads to a type-level error during the evaluation of the
-component that contains it is trivial to locate. In contrast, a bug that is
-non-trivial to locate is one that causes the buggy component to evaluate without
-an error but produce instead a value of the wrong type. An interesting bug is
-one where the value with the wrong type manages also to cross undetected to at
-least another component before a type-level check flags it. Good mutators should
-be able to produce these interesting bugs.
+Please see our reply to the second part of the discussion at the beginning
+of the rebuttal.
 
-> - I don't understand where the 16,800 number comes from. Is it an exhaustive
->   enumeration of all possible mutations? (Or the number of mutants constructed
->   in X amount of time?)
+> - I don't understand where the 16,800 number comes from. Is it an
+>   exhaustive enumeration of all possible mutations? (Or the number of
+>   mutants constructed in X amount of time?)
 
@@
 
-Here are the LOC counts and number of components for each benchmark. We will add
-the information to figure 2.
+Here are the LOC counts and number of components for each benchmark. We
+will add the information to figure 2.
 
-  benchmark    | LOC  | components
-  -------------|------|-----------
-  acquire      | 1941 | 9
-  gregor       | 2336 | 13
-  kcfa         | 328  | 7
-  quadT        | 7396 | 14
-  quadU        | 7282 | 14
-  snake        | 182  | 8
-  suffixtree   | 1500 | 6
-  synth        | 871  | 10
-  take5        | 465  | 8
-  tetris       | 280  | 9
+  benchmark    | LOC  | components -------------|------|-----------
+  acquire      | 1941 | 9 gregor       | 2336 | 13 kcfa         | 328  | 7
+  quadT        | 7396 | 14 quadU        | 7282 | 14 snake        | 182  |
+  8 suffixtree   | 1500 | 6 synth        | 871  | 10 take5        | 465  |
+  8 tetris       | 280  | 9
 
 > # Minor Comments
-> - L38: Should there have been a colon? Is the text "This contrast... " and
->   challenge quote from [3]? Who is saying what here...
+> - L38: Should there have been a colon? Is the text "This contrast... "
+>   and challenge quote from [3]? Who is saying what here...
 
@@
 
-> - L956: 30,000 hours of compute time is 1,250 days. I assume that the CPU has
->   many cores. Could you put that number?
+> - L956: 30,000 hours of compute time is 1,250 days. I assume that the
+>   CPU has many cores. Could you put that number?
 
@@
 
-> - L991: Figure 6: It took me some time to understand the figure. I wonder if
->   there is a better way to represent the same data?
+> - L991: Figure 6: It took me some time to understand the figure. I
+>   wonder if there is a better way to represent the same data?
 
-We have reached the same conclusion and we are considering alternatives that
-will make it easier to see how two modes compare to each other, such as a series
-of two-sided bar charts.
+We have reached the same conclusion and we are considering alternatives
+that will make it easier to see how two modes compare to each other, such
+as a series of two-sided bar charts.
 
@@
 > The experiment results show that there are debugging scenarios where
-> Transient is more useful than Natural. Does it happen simply because
-> of luck? Or, does it imply that Transient is more suitable than
-> Natural for a particular kind of bugs for some reason? Also, can there
-> be a new blame strategy that outperforms both Transient and Natural
-> all the time?
+> Transient is more useful than Natural. Does it happen simply because of
+> luck? Or, does it imply that Transient is more suitable than Natural for
+> a particular kind of bugs for some reason? Also, can there be a new
+> blame strategy that outperforms both Transient and Natural all the time?
 
-We were also surprised to discover that Transient outperforms Natural in some
-scenarios. After all, the theory of blame says otherwise. Even though we haven't
-identified specific patterns, we believe that the reason behind Transient's
-unexpected success is due to the advanced features of Typed Racket's type system
-that haven't appeared in models that compare Natural and Transient:
-polymorphism, row types, mixins, and more.
-
-We do not have any clues about a strategy that outperforms both Natural and
-Erasure. That said, our current conjecture is that a simple strategy that just
-points somewhere in the blame trail may be good enough. Such a strategy would
-also enable optimizations that Typed Racket currently foregoes in order to track
-blame. The good thing is that now we have a method to compare different
-strategies.
-
-chrdimo: TODO We need more details from Ben here.
+We were also surprised to discover that Transient outperforms Natural in
+some scenarios. After all, theory says otherwise. We haven't identified
+any patterns for these scenarios. We conjecture that they point to
+limitations of the theory but we can;t exclude that some of them aren't
+because of some unknown Typed Racket bug. Please see the first part of the
+discussion at the beginning of our response for more details and for our
+take on your last question.
 
@@
 > type rather than distinguishing typed and untyped modules. Would the
-> same evaluation method be able to directly appliable to blame in such systems?
+> same evaluation method be able to directly appliable to blame in such
+> systems?
 
-Yes. Incorporating type Dyn boils down to considering a hierarchy of types for
-each component rather than a single type. At the same time, our method is
-parametric to the "size" of a component and adapts in a straightforward manner
-to a setting where each definition can be considered a component instead of a
-whole module. Of course both of these adjustments lead to significantly larger
-scenario lattices than the ones we investigate, so keeping the experiment
-computationally feasible would require more aggressive sampling.
+Yes. Incorporating type Dyn boils down to considering a hierarchy of types
+for each component rather than a single type. At the same time, our method
+  is parametric to the "size" of a component and adapts in a
+  straightforward manner to a setting where each definition can be
+  considered a component instead of a whole module. Of course both of
+  these adjustments lead to significantly larger scenario lattices than
+  the ones we investigate, so keeping the experiment computationally
+  feasible would require more aggressive sampling.
 
-> Please explain what impedance mismatches are in the paper. Is it
-> simply type mismatches?
+> Please explain what impedance mismatches are in the paper. Is it simply
+> type mismatches?
 
-An impedance mismatch denotes a type annotation that is detected to disagree
-with the actual type of the annotated code. We will adjust the prose to clarify.
+An impedance mismatch denotes a type annotation that is detected to
+disagree with the actual type of the annotated code. We will adjust the
+prose to clarify.
 
-> The paper states that the effort distribution of the random mode
-> follows a normal distribution, as expected. Please explain why a
-> normal distribution is a reasonable expectation. ...
-> This point does not affect the contribution of the paper, but it
-> is not clear from the paper.
+> The paper states that the effort distribution of the random mode follows
+> a normal distribution, as expected. Please explain why a normal
+> distribution is a reasonable expectation. ...  This point does not
+> affect the contribution of the paper, but it is not clear from the
+> paper.
 
-You are right that for a given trail of length $n$ the probability that the
-random rational programmer discovers the buggy component on every step is $1/n$.
-The missing piece is that we randomly sample from the set of scenarios -- and by
-extension the set of trails. So the shape of the random programmer's effort
-matches our sampling distribution. That said, as you observe, the distribution
-of the random programmer's effort is not relevant to our results, so we will
-remove the characterization.
+You are right that for a given trail of length $n$ the probability that
+the random rational programmer discovers the buggy component on every step
+is $1/n$.  The missing piece is that we randomly sample from the set of
+scenarios -- and by extension the set of trails. So the shape of the
+random programmer's effort matches our sampling distribution. That said,
+as you observe, the distribution of the random programmer's effort is not
+relevant to our results, so we will remove the characterization.
 
@@
 ===========================================================================
-> But while there is a lot to like about the paper, it also has some shortcomings. The suite of programs used for the evaluation is fairly small (c.f., the SIGPLAN checklist).
+> But while there is a lot to like about the paper, it also has some
+> shortcomings. The suite of programs used for the evaluation is fairly
+> small (c.f., the SIGPLAN checklist).
 
-Could you clarify what you mean by fairly small with respect to the SIGPLAN
-checklist? We are the first to conduct an empirical study for blame and gradual
-typing, so we repurposed the standard performance benchmark suite for Typed
-Racket -- in particular, its largest and most complex benchmarks. Our inspection
-of the benchmarks indicates that they are representative set of Racket programs
-in terms of features and structure.
+Could you clarify what you mean by fairly small with respect to the
+SIGPLAN checklist? We are the first to conduct an empirical study for
+blame and gradual typing, so we repurposed the standard performance
+benchmark suite for Typed Racket -- in particular, its largest and most
+complex benchmarks. Our inspection of the benchmarks indicates that they
+are representative set of Racket programs in terms of features and
+structure.
 
-> The faults are introduced using synthetic mutations that may or may not correspond to the kinds of errors that arise in practice.
-This can be seen from the results in Figure 7 -- the lengths of the "trails" is quite small.
+> The faults are introduced using synthetic mutations that may or may not
+> correspond to the kinds of errors that arise in practice. This can be
+> seen from the results in Figure 7 -- the lengths of the "trails" is
+> quite small.
 
-There is no existing catalog of bugs in gradually typed programs. Furthermore,
-most type-level bugs do not survive after decent testing and thus do not make it
-to code repos. There are a few anecdotes from the literature but they are mostly
-artificial. Hence, since we need a large number of buggy scenarios for a
-meaningful study, a synthetic approach is the only way forward. In addition,
-mutation allows us to tune the set of scenarios to get a diverse set of bugs.
+There is no existing catalog of bugs in gradually typed programs.
+Furthermore, most type-level bugs do not survive  decent testing and thus
+do not make it to code repos. There are a few anecdotes from the
+  literature but they are mostly artificial. Hence, since we need a large
+  number of buggy scenarios for a meaningful study, a synthetic approach
+  is the only way forward. In addition, mutation allows us to tune the set
+  of scenarios to get a diverse set of interesting bugs.  We discuss in
+  section 8 how our approach to bug injection is a threat to validity,
 
-> While the study is mostly well done, it only considers two blame assignment strategies.
-> For example, it would be interesting to consider the "omniscient programmer" to get a bound on the value of blame. But this is not done.
+> While the study is mostly well done, it only considers two blame
+> assignment strategies.
 
-Could you elaborate what you mean by "omniscient programmer"? If this is the
-programmer that always makes the best choice, i.e. finds the bug at the first
-try, then this programmer is trivial. It has 100% success rate and a constant
-blame trail length of 1. Thus, comparing its success to any other strategy
-amounts to counting the number of scenarios where the other strategy fails;
-comparing effort is similarly trivial. This information is in the paper
-implicitly, but we can certainly add an explicit mention if it would be useful.
-
-Secondly, we would like to point out that the experiment considers then
-different modes of the rational programmer spread across three different gradual
+We would like to point out that the experiment considers ten different
+modes of the rational programmer spread across three different gradual
 typing systems.
 
-> And unfortunately the results are inconclusive -- Section 8 says "our results call for a deeper understanding of the two models for blame".
+> For example, it would be interesting to consider the "omniscient
+> programmer" to get a bound on the value of blame. But this is not done.
 
-The quote from section 8 refers to the theory of blame rather than our results.
-Our results show in clear terms (at least for our benchmarks) that a blame
-strategy is superior to exceptions, and that Natural and Transient have
-comparable efficacy. The latter is surprising given that theoretical models
-suggest the opposite conclusion. Thus, a side-effect of our experiment is a call
-to theoreticians for models that are aligned with practice.
+If  an  "omniscient programmer"  is the programmer that always makes the
+best choice, i.e. finds the bug at the first try, then this corresponds to
+a trivial mode; It has 100% success rate and a constant blame trail length
+of 1.  Thus, comparing its success to any other strategy amounts to
+counting the number of scenarios where the other strategy fails; comparing
+effort is similarly trivial. This information is in the paper implicitly,
+but we can certainly add an explicit mention if it would be useful.
 
-> Also, the published Transient scheme seems to have a minor bug, which might affect the outcomes when fixed (Section 8.2).
+> And unfortunately the results are inconclusive -- Section 8 says "our
+> results call for a deeper understanding of the two models for blame".
 
-Section 8.2 does not discuss a bug in Transient but rather a limitation of its
-current design that we discovered exactly because of the experiment.
+The quote from section 8 refers to the theory of blame rather than our
+results.  Please see our discussion of the first theme above for more
+details on this.
 
-> Most if not all of these shortcomings are discussed in the paper, which is much appreciated. But they do risk undermine the value of the contributions to some degree.
+> Also, the published Transient scheme seems to have a minor bug, which
+> might affect the outcomes when fixed (Section 8.2).
 
-We would like to point out that the results of the comparison between the three
-systems are only one part of the contribution of the paper. The other part is
-the method itself. The benchmarks, source of interesting bugs, and rational
-programmer modes are all parameters that can be instantiated differently without
-affecting either the recipe we discuss at the end of section 2 or the analytical
-framework from section 5 that underlies our experiment.
+Section 8.2 does not discuss a bug in Transient but rather a limitation of
+its current design that we discovered exactly because of the experiment.
 
-> To explain one comment above: I was surprised not to see a comparison against an "omniscient programmer" to establish a baseline (c.f., the SIGPLAN checklist).
-That is, model a blame assignment scheme in which blame could be placed on any component.
-And always pick the component that minimizes the length of the trail.
-Of course, this would be infeasible to implement in practice, but it would place an upper bound on the utility of blame.
+> Most if not all of these shortcomings are discussed in the paper, which
+> is much appreciated. But they do risk undermine the value of the
+> contributions to some degree.
 
-Our experiment does have a baseline. The random mode of the rational programmer
-serves as the "control" of our experiment. Furthermore the exception modes each
-serve as further "controls" to distinguish the effect of checks from that of
-blame within each semantics.
+We would like to point out that the results of the comparison between the
+three systems are only one part of the contribution of the paper. The
+other part is the method itself.
 
-> Similarly, I wondered why the modes of the transient programmer only consider picking the first/last blame, and not other choices.
-Section 5.2 just says "our answer" is that there are at least "two reasonable options." Why are other choices not reasonable?
+> To explain one comment above: I was surprised not to see a comparison
+> against an "omniscient programmer" to establish a baseline (c.f., the
+> SIGPLAN checklist). That is, model a blame assignment scheme in which
+> blame could be placed on any component. And always pick the component
+> that minimizes the length of the trail. Of course, this would be
+> infeasible to implement in practice, but it would place an upper bound
+> on the utility of blame.
 
-The transient semantics provide no interpretation for the blame set. We picked
-the first and last because they match the intuitive interpretation of the blame
-set as a list. We didn't mean to imply that other options are unreasonable. We
-just needed to make a reasonable choice and keep the duration of the experiment
-within reasonable limits. We will clarify this point in the prose.
+Our experiment does have a baseline. The random mode of the rational
+programmer serves as the "control" of our experiment. Furthermore the
+exception modes each serve as further "controls" to distinguish the effect
+of checks from that of blame within each semantics. Please also see our
+discussion of modes in the second theme at the beginning of our reply,
+
+> Similarly, I wondered why the modes of the transient programmer only
+> consider picking the first/last blame, and not other choices. Section
+> 5.2 just says "our answer" is that there are at least "two reasonable
+> options." Why are other choices not reasonable?
+
+The transient semantics provide no interpretation for the blame set. We
+picked the first and last because they match the intuitive interpretation
+of the blame set as a list. We didn't mean to imply that other options are
+unreasonable. We just needed to make a reasonable choice to keep the
+experiment feasible.
 
@@
 > Given that the experiments are developed and run in the context of the
-> large, real-world Racket ecosystem, I wonder whether about the chance
-> of incidental implementation choices or bugs affecting this result.
+> large, real-world Racket ecosystem, I wonder whether about the chance of
+> incidental implementation choices or bugs affecting this result.
 
-A side note here: Transient Typed Racket is faster than Natural Typed Racket
-when we turn off blame. Exactly as a sanity check that this is not an artifact
-of our implementation we have coded up benchmarks in Python and we confirmed
-that in Reticulated the performance of transient with blame is even worse than
-what we have observed in Typed Racket.
+A side note here: Transient Typed Racket is faster than Natural Typed
+Racket when we turn off blame. Exactly as a sanity check that this is not
+an artifact of our implementation we have coded up benchmarks in Python
+and we confirmed that in Reticulated the performance of transient with
+blame is even worse than what we have observed in Typed Racket.
 
@@
 
-We disagree that implementing Transient Typed Racket is mere engineering.
-Typed Racket comes with a way more sophisticated type system than
-Reticulated and there are a lot of points around type checks and blame
-that neither Reticulated nor the Transient model provide answers for.  We
-are happy to share with the chair a recent dissertation chapter that
-describes all the ideas and work that went into Transient Typed Racket.
-
-We believe that adding a formal model to the paper would distract
-from the main contribution of the paper as there are already multiple
-publications on the semantics of Transient, Natural and Erasure and at
-least one that compares them formally [Greenman et al. OOPSLA 2019]. That said, we
-agree that we should improve the presentation in section 3 to introduce
-more context.
-
-> *Custom Mutators (Section 4)*
-> The paper suggests that the mutators required here are significantly
-> different and more subtle, but this does not seem to be the case when
-> comparing Figure 3 and Table 3 of the POPL 2020 paper.
-> ...
-
-While we agree that there is overlap between our mutators and those from
-Lazarek's POPL work (and the common set of mutators where the latter come from),
-we disagree that the construction of custom mutators is straightforward. The
-sixteen we include in the paper are not the original set we tried and others
-that seemed like good candidates have proven to be ineffective. For instance
-replacing car with cdr should lead to type-level bugs however none of them are
-interesting as we define interesting in the paper. Similarly, swapping mutable
-vectors and hashes with immutable ones is an interesting mutation in the
-traditional sense and should lead to a type error in Typed Racket as its type
-system distinguishes the two, however we discovered that again is not
-interesting in our setting. In general, there is no methodology for constructing
-a new set of mutators, so narrowing down the ones in the paper required that we
-first characterize precisely what makes a bug interesting and then rounds of
-trials to end up with a set of mutators that leads to a sufficiently large set
-of interesting diverse bugs.
+> *Custom Mutators (Section 4)* The paper suggests that the mutators
+> required here are significantly different and more subtle, but this does
+> not seem to be the case when comparing Figure 3 and Table 3 of the POPL
+> 2020 paper.  ...
 
@@
 >
-> These also seem to be overemphasized. The main design choice seems to
-> be in how to use Transient blame, which reports multiple components,
-> and picking the first and last are pretty intuitive.
+> These also seem to be overemphasized. The main design choice seems to be
+> in how to use Transient blame, which reports multiple components, and
+> picking the first and last are pretty intuitive.
 >
@@
 
-We disagree that section 5 boils down to picking the first and last modes for
-Transient. The core of this section is the analysis of of the different systems
-and their unification in a common precise framework as modes. Because of the
-common substrate of blame trails modes enable the comparison between the
-different systems even when sampling is involved. Furthermore we use modes to
-introduce in a uniform manner in the framework baselines that help us isolate
-confounding factors from the actual effectiveness of blame. This si exactly the
-role of the exception modes for Natural and Transient. Those two perform checks
-at runtime and the exception modes help us determine what part of the success of
-Natural and Transient is due to the extra checks or due to blame (and also
-whether blame masks useful information from exceptions). For example, we see
-that Natural without blame but with the same checks outperforms Erasure in ~25%
-of the scenarios (figure 6). Blame add another ~11% on top of that. For
-Transient there is a small percentage of scenarios where exceptions due better
-than blame but still blame improves over exceptions by ~11% compared to Erasure.
-
-As a final note, even though each of the above points individually may not have
-sufficient technical depth for a separate contribution (which we do not claim),
-together they exhibit the creative steps that are necessary to lift the scope
-and magnitude of the idea from Lazarek's POPL paper on evaluating blame from a
-single behavioral contract system to a framework for evaluating and comparing
-different gradual typing systems.
+Please see our discussion of the challenges for our method at the
+beginning of our response.
 
@@
 > more explicit. As discussed above, there seem to be more similarities
-> than suggested.
-> L92: "use of _a_ higher-order contract system"
-> L155: "authors extensive"
-> L253: "exports _it_ as"
-> L280: Missing `untyped-`
-> L314: " . "
-> L382: "Dyn,"
-> Fig 2: Presumably authors were included to indicate the variety of
-> sources, but I'm not sure that was necessary.
-> L597: "make us_e_ of"
-> L664: "described in _Section_ 4"
-> L980: "Figure 5.4" ==> "Figure 6"
-> L986: "2.8%" ==> "2.18%" ?
-> L1017: "33%" ==> "26%" ?
+> than suggested.  L92: "use of _a_ higher-order contract system" L155:
+> "authors extensive" L253: "exports _it_ as" L280: Missing `untyped-`
+> L314: " . " L382: "Dyn," Fig 2: Presumably authors were included to
+> indicate the variety of sources, but I'm not sure that was necessary.
+> L597: "make us_e_ of" L664: "described in _Section_ 4" L980: "Figure
+> 5.4" ==> "Figure 6" L986: "2.8%" ==> "2.18%" ?  L1017: "33%" ==> "26%" ?
 > L1304: "Problem is ...."
@@
 
-> This paper seeks to establish (or refute) the benefit of blame tracking in gradually typed programming environments.
-It takes several Typed Racket benchmarks and considers multiple approaches to blame tracking.
-The central idea is to use a fault injection method to create variant programs along two dimensions: type errors inserted according to some crafted heuristics, and static typing selectively removed on a component-by-component granularity.
-This creates a large number of variant programs which may produce run-time type errors with blame; this is then iterated to simulate a "rational programmer", i.e.
-such that blame is used to select where more static checking will be added next; at any given iteration this may or may not catch the injected error, and the number of iterations is counted as a proxy for effort.
-A further contribution is implementing the different forms of blame-tracking in the same framework so that they can be compared.
-The finding is that blame is found to offer some positive advantage in effort (over simply being guided by the stack trace of a run-time exception) in 9--13% of the initial sample of variants, with some minor differences between the forms of blame tracking tried.
+> However, I don't find the results to be as conclusive as the authors
+> state. Aside from the Erasure case, whose relevance I found unclear (see
+> below), conclusions rest on some small-valued "more useful than"
+> percentages. These are pretty hard to interpret, because the
+> "usefulness" metrics necessarily build in a lot of simplifying
+> assumptions. E.g. in 5.4, the "percentage of scenarios where ... is more
+> effective" doesn't account for *extent* of difference, and the
+> trail-length "programmer effort" metric in 5.5 is also unlikely to be a
+> great proxy. These limitations are understandable, method-wise, but
+> overall the results are fairly null... they let us continue believing
+> "what we'd expect" but not with any great sense of added confidence.
+> That doesn't entirely diminish the work, of course, but I'd say the
+> current write-up overplays its hand a little. I appreciate the
+> discussion of threats to validity.
 
-  TODO: misunderstanding: we propose a methodology
+ We absolutely disagree with you. Please see our discussion of the
+ conclusions form our experiment at the beginning of our reply.
 
-The primary intended contribution of this paper is the methodology, and the
-experimental results serve as secondary contributions that validate the
-methodology's utility. We will rephrase the introduction to further emphasize
-that focus. Furthermore, we would like to emphasize that the two key findings in
-the experimental results are that 1) both Natural and Transient's dynamic type
-enforcement with blame tracking improve significantly over Erasure, and 2)
-Natural's blame tracking is not significantly better than Transient's, despite
-the theoretical indications to the contrary.
+ chrdimo: Should we be more assertive here about the metrics? I think this
+ is a hard to defend position because of the non-constructive manner of the
+ criticism --- it is plain dismissive without explanation.
 
-> However, I don't find the results to be as conclusive as the authors state.
-Aside from the Erasure case, whose relevance I found unclear (see below), conclusions rest on some small-valued "more useful than" percentages.
-These are pretty hard to interpret, because the "usefulness" metrics necessarily build in a lot of simplifying assumptions.
-E.g.
-in 5.4, the "percentage of scenarios where ...
-is more effective" doesn't account for *extent* of difference, and the trail-length "programmer effort" metric in 5.5 is also unlikely to be a great proxy.
-These limitations are understandable, method-wise, but overall the results are fairly null...
-they let us continue believing "what we'd expect" but not with any great sense of added confidence.
-That doesn't entirely diminish the work, of course, but I'd say the current write-up overplays its hand a little.
-I appreciate the discussion of threats to validity.
+> About Erasure: if I'm reading this correctly, the key thing here is that
+> in Typed Racket, the set of static types is more expressive than the set
+> of dynamic types. For example, there is a static notion of non-negative
+> integer that is distinct from plain integer. This sort of design isn't
+> universal -- in some languages/systems static types closely mirror the
+> classification of objects in the language's dynamic semantics. Writing
+> these stricter contracts into a program brings its own benefits,
+> separate from blame or indeed from static checking. It feels like the
+> paper doesn't take enough care to distinguish the two effects: the
+> effect of systematically applying more refined contracts over program
+> values, and the effect of gradually enabling static checking of those
+> contracts (iteratively, guided by blame). The authors do mention this
+> around line 204, and the comparisons in Figure 6 between "_ exceptions"
+> and "Erasure" seem to be measuring this -- the gain from checking these
+> extra annotations, with the more precise and/or more timely checks that
+> they imply, relative to the erased case where only the language baseline
+> contracts are checked. Indeed that's the point of 'exception'
+> experiments. Since the biggest effect sizes on display are these ones --
+> between Erasure and anything else -- this seems at best distracting. I'd
+> be glad to hear from the authors if I'm misunderstanding anything here.
+> In the detailed comments below, I have noted some places in the text
+> where it would be useful to remind the reader that this design property
+> of Typed Racket is at play.
 
-  TODO: misunderstanding: the results are not at all "what we'd expect" based on the theory
+We do think there is a misunderstanding here about what gradual typing is
+in general.
 
-The status quo of what we would expect based on the current theory of Natural
-and Transient is that 1) either kind of blame tracking should be better than
-none, and 2) Natural should provide significantly better blame information than
-Transient and Erasure. Our results validate the first expectation that blame
-provides benefits over plain exceptions. Surprisingly, our results contradict
-the second expectation and suggest that in fact Natural and Transient offer
-comparable blame information.
+1) Erasure is the established name of the semantics implemented by
+TypeScript, Flow etc. These languages come with an equally sophisticated
+type system as Typed Racket. Programmers add annotations to their program
+and the type checker does its best to type check the program in order to
+discover mistakes and help the IDE. Then types are erased and the program
+is run. Since the program is not fully typed, the run time type checks of
+primitive operations may detect an incompatible argument and raise an
+exception.
 
-> About Erasure: if I'm reading this correctly, the key thing here is that in Typed Racket, the set of static types is more expressive than the set of dynamic types.
-For example, there is a static notion of non-negative integer that is distinct from plain integer.
-This sort of design isn't universal -- in some languages/systems static types closely mirror the classification of objects in the language's dynamic semantics.
-Writing these stricter contracts into a program brings its own benefits, separate from blame or indeed from static checking.
-It feels like the paper doesn't take enough care to distinguish the two effects: the effect of systematically applying more refined contracts over program values, and the effect of gradually enabling static checking of those contracts (iteratively, guided by blame).
-The authors do mention this around line 204, and the comparisons in Figure 6 between "_ exceptions" and "Erasure" seem to be measuring this -- the gain from checking these extra annotations, with the more precise and/or more timely checks that they imply, relative to the erased case where only the language baseline contracts are checked.
-Indeed that's the point of 'exception' experiments.
-Since the biggest effect sizes on display are these ones -- between Erasure and anything else -- this seems at best distracting.
-I'd be glad to hear from the authors if I'm misunderstanding anything here.
-In the detailed comments below, I have noted some places in the text where it would be useful to remind the reader that this design property of Typed Racket is at play.
+2) The other two semantics introduce extra run time checks (in the form of
+proxies for Natural and inlined checks for Transient) to enforce that
+values untyped code provides to typed code have the expected type. when
+one of these checks fails they also issue blame (with different
+mechanisms).
 
-TODO: misunderstanding about gradual typing as a whole, and the point of the exception modes
+3) In Typed Racket programmers cannot add contracts. The precision of the
+checks is determined by the types and matches those. In turn the types of
+primitive operations much the checks that their Racket counterparts
+perform. Finally our benchmarks come with the most precise types that
+allow them to type check (where precision is based on subtyping).
 
-The idea of the Erasure semantics is not related to the expressiveness of the
-type system or how the types relate to kinds of runtime values. Rather, the key
-idea is that Erasure does not enforce type annotations with dynamic checks at
-all. Running a gradual program using the Erasure semantics is the same as
-stripping all annotations and running the resulting (completely dynamic)
-program. Thus, Erasure programs can only raise exceptions from either the checks
-performed by the language's primitive operations (e.g. `+`), or from the program
-itself raising an exception. In contrast, the other two semantics do enforce
-type annotations dynamically; both Natural and Transient insert dynamic checks
-verifying that the shape of values at runtime match their static annotations.
-Thus, in addition to exceptions, gradual programs using these semantics may
-raise a second kind of error: a dynamic type check failure. These dynamic errors
-can include blame information if the dynamic type checks are designed to track
-blame, or they can forego blame and just offer a stack trace. In the case of
-Natural, these two options are captured by the "Natural" and "Natural
-exceptions" modes, respectively. Thus the "Erasure" mode represents a completely
-different kind of checking (and thus error reporting) compared to both the
-"Natural" and "Natural exceptions" modes (the two of which share the same kind
-checking, but different error reporting).
+4) The Natural and Transient exception modes perform the same checks as
+Natural and Transient correspondingly. The only difference is that they do
+not produce blame but instead signal an exception with the stack trace of
+the failed check. We use these modes to separate the effectiveness of
+blame from that of the checks.
 
-> It's interesting to note that there seems to be no convincing analogous experimental or modelling-based justification for plain old static typing, i.e.
-showing that it somehow presents a net gain to the programmer.
-Rather, this has simply been posited/assumed by a very long line of work.
-I am not defending that state of affairs, but it points to the difficulty of showing conclusively that something truly helps programmers.
-I can see value in this sort of simulation-style approach, and it is no less convincing than user studies.
-So I'd be interested to hear from the author(s) if they have any more arguments that (1) I've underestimated the results' conclusiveness, or (2) this is a novel/interesting family of methods that might be pursued more widely, or (3) that deeper experiments building on these ideas might yield more compelling insights.
+chrdimo: I am not sure if I understand what this guy is trying to say
+here. It seems a bit like rambling to me sprinkled with bold statements.
 
-(Matthias) As posed, the question expresses a significant mis-characterization
-of the submission. It is not the benefits of a gradual type system that are in
-doubt. The experimental setup keeps the type system constant but allows to
-answer the question which of several run-time checking regimes (for enforcing
-type consistency) provides the best explanatory messages in case of violations.
-Hence the analogous question for purely static type systems would ask which of
-several reporting schemes provides the best explanatory message for type-errors.
-In the case of simply typed languages and even language with local type
-inference, this question is basically meaningless. In the case of languages with
-HM type inference, the question has been implicitly raised for four decades with
-the development of alternative ways of finding the source of inference
-conflicts.
+> It's interesting to note that there seems to be no convincing analogous
+> experimental or modelling-based justification for plain old static
+> typing, i.e. showing that it somehow presents a net gain to the
+> programmer. Rather, this has simply been posited/assumed by a very long
+> line of work. I am not defending that state of affairs, but it points to
+> the difficulty of showing conclusively that something truly helps
+> programmers. I can see value in this sort of simulation-style approach,
+> and it is no less convincing than user studies. So I'd be interested to
+> hear from the author(s) if they have any more arguments that (1) I've
+> underestimated the results' conclusiveness, or (2) this is a
+> novel/interesting family of methods that might be pursued more widely,
+> or (3) that deeper experiments building on these ideas might yield more
+> compelling insights.
 
-If the reviewer is indeed interested in the question of whether type systems
-help programmers---a question that this submission does _not_ ask---the recent
-OOPSLA literature contains several user studies. A simple Google query will
-suggest a short list of these publications.
+For point 1 and 3 please see our discusssion at the beginning of the
+reply. For point 2, the idea of the rational programmer is a new one so
+far applied to blame in contracts (Lazarek et al. POPL 2020) and blame in
+gradual typing (our work). We don't know the future but we believe it can
+have a similar effect in programming languages as homo economicus in
+economics.
 
-  TODO ll: I added responses (after ⟶) to anything below that seems to warrant more than "Thanks, we'll fix it".
+That said, as posed, the comment expresses a significant
+mis-characterization of the submission. It is not the benefits of a
+gradual type system that are in doubt. The experimental setup keeps the
+type system constant but allows to answer the question which of several
+run-time checking regimes (for enforcing types) provides the best
+explanatory messages in case of violations.  Hence the analogous question
+for purely static type systems would ask which of several reporting
+  schemes provides the best explanatory message for type errors.  In the
+  case of simply typed languages and even languages with local type
+  inference, this question is basically meaningless. In the case of
+  languages with HM type inference, the question has been implicitly
+  raised for four decades with the development of alternative ways of
+  finding the source of inference conflicts.
 
-Detailed comments:
+If the reviewer is indeed interested in the question of whether type
+systems help programmers---a question that this submission does _not_
+ask---the recent OOPSLA literature contains several user studies. A simple
+Google query will suggest a short list of these publications.
 
-In the title, "evaluating blame" reads oddly.
-It is ambiguous, because "evaluate" sometimes means to compute a result.
-A fuller phrase, like "evaluating the usefulness of blame tracking...", would probably be worth the words.
+> line 34: "then their compilers remove types and rely on the built-in
+> safety checks of the underlying language to catch any problems". This
+> reads oddly in context. Didn't they just do a bunch of static checking?
+> So they are *not* relying just on dynamic checks to catch problems?
+> Maybe there is something more accurate to say here... e.g. no dynamic
+> check is removed, or something like that.
 
-The abstract doesn't say much about the work. It would be better written for experts to quickly gather an overview what the paper contributes.
+In Erasure, a program can type check and still have type-level errors
+because some type annotations may be missing. In contrast to Natural and
+Transient, types do not result in run time checks so the only way to
+detect type-level errors that the type checker couldn't find is the
+built-in checks of the language's primitives.
 
-Conversely, the main body of the paper skimps a bit on background.
-It never explicitly covers what "blame" means and how it works in practice.
-Similarly, it repeatedly talks about "impedance mismatches" without defining them.
-Perhaps this phrase is now standard in the gradual typing literature, which I haven't kept up with (hence my Z expertise).
-But I did read the Wadler/Findler paper carefully at the time.
-It did not talk about impedance mismatches.
-In any case, it is a fuzzy metaphor...
-please say exactly what it means here.
-  ⟶ We will adjust the prose to more explicitly describe blame and impedance mismatches.
+> line 38: "explicit statement and challenge" -- what is it?
 
-line 34: "then their compilers remove types and rely on the built-in safety checks of the underlying language to catch any problems".
-This reads oddly in context.
-Didn't they just do a bunch of static checking?
-So they are *not* relying just on dynamic checks to catch problems?
-Maybe there is something more accurate to say here...
-e.g.
-no dynamic check is removed, or something like that.
-
-  ⟶ The static checking performed by languages using the Erasure semantics (like Flow, Hack, and TypeScript) is done in a best-effort fashion; the fact that a program passes type checking has absolutely no effect on how the program runs.
-    These languages literally erase the types to obtain a program of the underlying language (e.g. TypeScript or PHP) to run.
-	Those underlying languages perform all of the dynamic checks that a typical dynamic language performs for primitive operations (e.g. that `+` is only used with numbers).
-
-line 38: "explicit statement and challenge" -- what is it?
-  ⟶ The challenge questions whether blame is useful at all in gradual typing.
+The challenge questions whether blame is useful at all in gradual typing.
 We will clarify this in the prose.
 
-line 49: I agree they got the word wrong, but you should explain this
+> line 49: I agree they got the word wrong, but you should explain this
 
-line 60: around this paragraph the writing started to grate.
-There's no need to generalise about what people do or think, and the "As a matter of fact...
-simply..." style is somewhere between laboured and patronising.
-It's better to more plainly state the gap in the literature that you're addressing.
-Throughout the paper, much space could be saved by writing in a more direct style.
-  ⟶ Noted, we will simplify the language where possible.
+> line 60: around this paragraph the writing started to grate. There's no
+> need to generalise about what people do or think, and the "As a matter
+> of fact... simply..." style is somewhere between laboured and
+> patronising. It's better to more plainly state the gap in the literature
+> that you're addressing. Throughout the paper, much space could be saved
+> by writing in a more direct style.
 
-line 69: be explicit that Lazarek et al were (as I later gathered) doing something about blame to do with higher-order contracts but not gradual types
+Noted, we will simplify the language where possible.
 
-line 70: what does it mean to "follow" the slogan? Was unclear to me.
-  ⟶ The following sentences are meant to explain this, so we will adjust the prose to make that clear.
+> line 69: be explicit that Lazarek et al were (as I later gathered) doing
+> something about blame to do with higher-order contracts but not gradual
+> types
 
-line 90: at first I wondered: what is a case? Maybe say "program variant" to foreshadow the idea of generating mutants etc?
+Thanks we will add a more detailed comparison along the lines of the
+discussion at the beginning of our response.
 
-line 91-ish: "Transient, "Natural", "Erasure" -- be explicit that these are names that *you* are introducing
-  ⟶ These are names from the Gradual Typing literature (see for instance [9, 10, 37]).
+> line 70: what does it mean to "follow" the slogan? Was unclear to me.
 
-line 94: "forego" => "forgo"
+The following sentences are meant to explain this, so we will adjust the
+prose to make that clear.
 
-line 119: "homo economicus" needs glossing or a reference
+> line 90: at first I wondered: what is a case? Maybe say "program
+> variant" to foreshadow the idea of generating mutants etc?
 
-line 125: was wondering whether "impedance mismatch" just means "feasible run-time type error".
+It means debugging scenario. We will clarify.
 
-line 148: last sentence appears to contradict the preceding paragraph, and is left hanging oddly. Instead, make it the start of a new paragraph.
+> line 91-ish: "Transient, "Natural", "Erasure" -- be explicit that these
+> are names that *you* are introducing
 
-line 160: "disparity" is a strange word here. "diversity"?
+These are standard names for these three systems [9, 10, 37]. We didn't
+introduce them.
 
-line 170: "Otherwise, results from..." -- the point was already clear
+> line 187: what is a "component"? Should be easy to define.
 
-line 174: ... instead of just saying what you built on, first say what you did! It's really not clear at this point.
+In this context a component is a module.
 
-line 187: what is a "component"? Should be easy to define.
+> line 199: "blame set as another form of a stack trace" -- yes. I was
+> hoping to see a clearer example of debugging with and without blame,
+> i.e. something making explicit the similarities and differences between
+> having blame info and having only a stack trace at the error site.
 
-line 199: "blame set as another form of a stack trace" -- yes.
-I was hoping to see a clearer example of debugging with and without blame, i.e.
-something making explicit the similarities and differences between having blame info and having only a stack trace at the error site.
+> line 277: what is "responsibility" of a "party", exactly? Are "party",
+> "module" and "component" all the same things?
 
-line 211: "languages exceptions" typo
+Yes in this context. Blame refers to the unit of blame which in Typed
+Racket is a module.
 
-line 268: this explanation of the workings of proxies seemed overwrought
+> line 293: I was wondering what constitutes a boundary crossing. Clearly,
+> passing by function call or return crosses from the caller's module to
+> the callee's. What about values exchanged through reads/writes to shared
+> state?
 
-line 277: what is "responsibility" of a "party", exactly? Are "party", "module" and "component" all the same things?
-  ⟶ Components are parties, and in the case of Typed Racket modules are components because they are the units of migration.
-Other gradual type systems, like those that do not require annotated entire modules, may consider smaller units of code (e.g.
-definitions) as components.
-    We will clarify this in the prose.
-
-line 283: "rewrites typed modules to inline checks" -- so a module gets turned into a check? Clearly not, but that's how it reads
-
-line 293: I was wondering what constitutes a boundary crossing.
-Clearly, passing by function call or return crosses from the caller's module to the callee's.
-What about values exchanged through reads/writes to shared state?
-  ⟶ Yes, communicating a value via state also constitutes a boundary crossing.
+Yes, communicating a value via state also constitutes a boundary crossing.
 We will clarify what a boundary crossing is in the prose.
 
-line 299: "crosses" => "crossings" (probably)
+> In the title, "evaluating blame" reads oddly. It is ambiguous, because
+> "evaluate" sometimes means to compute a result. A fuller phrase, like
+> "evaluating the usefulness of blame tracking...", would probably be
+> worth the words.
 
-line 313: it threw me that a program might not fail but "produce a wrong result".
-If it could be caught by the gradual type system, why can't it be caught at run time?
-My best get at explaining this is by what I wrote above, i.e.
-it's a consequence of Typed Racket's more refined static notion of type.
-In certain other systems this wouldn't be possible, because the static checker would only catch (albeit earlier) errors that would be caught at run time, so there would be no basis to call the result "wrong".
-  ⟶ The result can be wrong in that it doesn't agree with the annotations (which we assume express programmer intent), even if none of the primitive operations applied to the result raise an error.
-    For example, imagine the the client saves the result to a file.
-    The language primitives for writing values to a file (typically) can write any kind of value to a file, so they have no checks that could fail due to the result being of the wrong type.
-    Instead, the resulting file just ends up with the incorrect contents and the program terminates normally.
+> The abstract doesn't say much about the work. It would be better written
+> for experts to quickly gather an overview what the paper contributes.
 
-lien 352: "three interpretations" -- what are they? I don't see them in the figure.
+> Conversely, the main body of the paper skimps a bit on background. It
+> never explicitly covers what "blame" means and how it works in practice.
+> Similarly, it repeatedly talks about "impedance mismatches" without
+> defining them. Perhaps this phrase is now standard in the gradual typing
+> literature, which I haven't kept up with (hence my Z expertise). But I
+> did read the Wadler/Findler paper carefully at the time. It did not talk
+> about impedance mismatches. In any case, it is a fuzzy metaphor...
+> please say exactly what it means here.
 
-line 380: from this I inferred that "migratory typing" means "gradual typing applied at modulewise granularity". Assuming that's correct, it's worth saying directly.
-  ⟶ The three choices of using Dyn, the granularity of components, and how compilation affects completely untyped components represent orthogonal aspects of the design of a Gradual Typing system.
-    In the case of Typed Racket's migratory typing, those three choices amount to more than just applying the ideas of [21] at the granularity of modules.
+> line 313: it threw me that a program might not fail but "produce a wrong
+> result". If it could be caught by the gradual type system, why can't it
+> be caught at run time? My best get at explaining this is by what I wrote
+> above, i.e. it's a consequence of Typed Racket's more refined static
+> notion of type. In certain other systems this wouldn't be possible,
+> because the static checker would only catch (albeit earlier) errors that
+> would be caught at run time, so there would be no basis to call the
+> result "wrong".
 
-line 421: "type mistake" -- does this mean "feasible run-time type error"?
-  ⟶ Could you clarify what a makes a runtime type error "feasible"?
+Consider a function with a type annotation that says its a function from
+integers to string that formats its argument to a string and returns it.
+If we import the function in an untyped module under Erasure there are no
+checks to make sure that the function is going to be applied to integers.
+So we can apply it to character or a string and still get back a result
+except that it is not the string representation of a number.
 
-line 424: "fully typed correct programs" -- presumably your method could also work with not-yet-fully-typed correct programs, just not ranging over the entire lattice in those cases.
-I was wondering whether that might give different/interesting results.
-  ⟶ Yes, the approach could be applied, but the results may be mostly inconclusive depending on how many trails reach the point where the blamed component lacks type informtion.
-    All trails that reach that point provide no useful information.
+> line 380: from this I inferred that "migratory typing" means "gradual
+> typing applied at modulewise granularity". Assuming that's correct, it's
+> worth saying directly.
 
-line 433: "without loss of diversity" -- this is glib. Clearly diversity is lost; just claim that what remains is still diverse enough.
+There are two points that distinguish gradual typing from migratory
+typing: the existence of Dyn and the granularity of components.  We will
+clarify the prose.
 
-line 497: "truthiness" needs explaining
+> line 424: "fully typed correct programs" -- presumably your method could
+> also work with not-yet-fully-typed correct programs, just not ranging
+> over the entire lattice in those cases. I was wondering whether that
+> might give different/interesting results.
 
-line 517: the difference between #1 and #2 here again relies on the surprising property (to the unfamiliar) that Typed Racket has a notion of "type-level mistake" that doesn't surface under erasure (as an exception or whatever) but also is not a false positive (i.e.
-you're not talking about conservativeness of static checking).
-Line 538's "unavoidable" claim is probably also true only in such a context.
-  TODO: Don't respond to this in light of above explanation(s) of why they misunderstand erasure and TR?
+Yes, the approach could be applied.
 
-line 519: "at least three" -- better to claim this is only two here, then explain later that the driver doesn't count.
-Mentioning "three" up-front just raises an unnecessary question in the reader's mind.
+> line 517: the difference between #1 and #2 here again relies on the
+> surprising property (to the unfamiliar) that Typed Racket has a notion
+> of "type-level mistake" that doesn't surface under erasure (as an
+> exception or whatever) but also is not a false positive (i.e. you're not
+> talking about conservativeness of static checking). Line 538's
+> "unavoidable" claim is probably also true only in such a context.
 
-line 634: earlier I had inferred that the distinction between "migratory" and "gradual" was more than just preferring one word; this seems to contradict that
+Yes please see our answer to a relevant comment above.
 
-line 640: what makes it "concise"?
-  TODO ll: Why does it? I'm not sure about the word here either. Just say we'll change it?
+> line 756: "checked the value's type.." -- and the check passed!?
 
-line 673: "programmers runs" typo
+Indeed, Transient checks only check the top-level type constructor of a
+value.  We will add a reminder of this fact to the prose here.
 
-line 727: "dubbed" -- reads oddly. Maybe italicise the "location", but it seems overkill. The phrase is pretty self-suggesting as it is.
+> line 119: "homo economicus" needs glossing or a reference
 
-line 756: "checked the value's type.." -- and the check passed!?
-  ⟶ Indeed, Transient checks only check the top-level type constructor of a value, so many checks can pass for a bad value before it is unwrapped to discover the problem.
-    We will add a reminder of this fact to the prose here.
+> line 94: "forego" => "forgo"
 
-line 752: "added to the blame set first" -- should it be a blame list, then?
+> line 125: was wondering whether "impedance mismatch" just means
+> "feasible run-time type error".
 
-line 936: no need for hyphen after the adverb
+> line 148: last sentence appears to contradict the preceding paragraph,
+> and is left hanging oddly. Instead, make it the start of a new
+> paragraph.
 
-line 966: "more useful C" -- missing "than"
+> line 160: "disparity" is a strange word here. "diversity"?
 
-line 982: don't think these percentages deserve 3 significant figures
-  TODO: "deserve"?
-  ⟶ We include the hundreths place to highlight in particular that Transient first and last blame are not exactly the same.
-    If this information isn't useful, we can use just the tenths place.
+> line 170: "Otherwise, results from..." -- the point was already clear
 
-line 1304: "Problem is" -- missing "The". Also, no need to italicise the next sentence... it is really not that deep or surprising.
+> line 174: ... instead of just saying what you built on, first say what
+> you did! It's really not clear at this point.
+
+> line 211: "languages exceptions" typo
+
+> line 268: this explanation of the workings of proxies seemed overwrought
+
+> line 283: "rewrites typed modules to inline checks" -- so a module gets
+> turned into a check? Clearly not, but that's how it reads
+
+> line 299: "crosses" => "crossings" (probably)
+
+> lien 352: "three interpretations" -- what are they? I don't see them in
+> the figure.
+
+> line 421: "type mistake" -- does this mean "feasible run-time type
+> error"?
+
+> line 433: "without loss of diversity" -- this is glib. Clearly diversity
+> is lost; just claim that what remains is still diverse enough.
+
+> line 497: "truthiness" needs explaining
+
+> line 519: "at least three" -- better to claim this is only two here,
+> then explain later that the driver doesn't count. Mentioning "three"
+> up-front just raises an unnecessary question in the reader's mind.
+
+> line 634: earlier I had inferred that the distinction between
+> "migratory" and "gradual" was more than just preferring one word; this
+> seems to contradict that
+
+> line 640: what makes it "concise"?
+
+> line 673: "programmers runs" typo
+
+> line 727: "dubbed" -- reads oddly. Maybe italicise the "location", but
+> it seems overkill. The phrase is pretty self-suggesting as it is.
+
+> line 752: "added to the blame set first" -- should it be a blame list,
+> then?
+
+> line 936: no need for hyphen after the adverb
+
+> line 966: "more useful C" -- missing "than"
+
+> line 982: don't think these percentages deserve 3 significant figures
+
+> line 1304: "Problem is" -- missing "The". Also, no need to italicise the
+> next sentence... it is really not that deep or surprising.
+
+Thanks for all these detailed feedback points. We will use them to improve
+the prose.

### 2021-02-10 writing/reviews/pldi-rebuttal-high-level-first.md
adding rebittal draft

@@
 mutators. In fact, their method does not detect the overwhelming majority
-of bugs unless if contracts describe the full functional correctness of
+of bugs except when contracts describe the full functional correctness of
 components.
@@
 of a benchmark. They do not provide any guidance how to generalize to
-multiple systems with different notions of checks and/or blame in a way
-that allows comparisons.
+multiple systems with different notions of checks and/or blame in a
+uniform way that also allows comparisons.
 
@@
 Reticulated nor the Transient model provide answers about what shallow
-check corresponds to Typed Racket's sophisticated types (e.g.
-polymorphism, row types for mixins, occurrence typing) and what information
-is necessary to record in the blame map. Moreover in the process of the
-implementation we discovered issues with the published Transient model and
-the Reticulated model that we had to fix.  We are happy to share with the
-chair a recent dissertation chapter that describes all the ideas and work
-that went into Transient Typed Racket. We do not think that this paper is
-the appropriate place to describe the implementation in detail.
+checks correspond to Typed Racket's sophisticated types (e.g.
+polymorphism, row types for mixins, occurrence typing) and what
+information is necessary to record in the blame map. Moreover in the
+process of the implementation we discovered issues with the published
+Transient model and Reticulated that we had to fix. We do not think that
+this paper is the appropriate place to describe the implementation in
+detail.
 
@@
 A key step for realizing our experiment is the precise and unified
-description of debugging in different gradual typing system  as  the
+description of debugging in the three gradual typing system  as  the
 different modes of the rational programmer. Because of their common

### 2021-02-10 writing/reviews/pldi-rebuttal-checklist.md
Add checklist file recording if each point in reviews are addressed

--- 2021-02-09 writing/reviews/pldi-rebuttal-notes.md
@@
-Response notes format: indented after ⟶ or TODO.
-⟶ is a drafted response, TODO records notes about what to respond to or how.
-
-chrdimo notes format: unindented after chrdimo.
-
-addressed: marks todos that have answers in `pldi-rebuttal.md`
+Un-addressed points: TODO
+Addressed points: DONE
 
@@
 
+DONE
 - Could you comment a bit more on what makes a mutator "good"? (other than
@@
 
-  ⟶ Abstractly, a good mutator reliably generates bugs that are detectable by the type system and by all three semantics' runtime type checks; ideally the bugs also manifest during the interaction of multiple components.
-    We rely on the experimental check of our mutators since this criteria is difficult or impossible to decide otherwise.
-    That said, we can reason that some mutators are clearly unlikely or impossible to be good.
-	...
-	We will extend the discussion at the start of section 4 to clarify this point.
-
-  addressed
-  TODO: Rewrite response along these lines...
-    A common mutator in the literature is swapping constants of the same type, but this is not a good mutator for us because type system can't catch it.
-	Example of an obviously bad mutator: replacing any string with some pre-defined non-empty string (e.g. "foobar") would be a bad mutator.
-	Examples of mutators we cut: `wrap-cond` (tricks occurrence typing by hiding the result of conditionals to produce a type error, but never a dynamic error), `add-extra-method`, swapping `car` and `cdr`, `set!`ing an ID to itself (only causes a type error), swapping mutable and immutable hashes+vectors.
-	# This answer should serve to justify our mutators and the challenge of making good mutators more broadly, to answer reviewer D's concerns about the delta of making mutators.
-
-chrdimo: A bad mutator results in a type-level bug that is trivial to
-locate. For instance a bug that leads to a type-level error during the
-evaluation of the component that contains it is trivial to locate. In
-contrast, a bug that is non-trivial to locate is one  that causes the
-component that mutates it to evaluate without an error but produce instead
-a value of the wrong type.  An interesting bug is one where the value with
-the wrong type manages also to cross undetected to at least another
-component than the one that produces it until a type-level checks flags
-it. These latter bugs are those that good mutators should be able to
-produce.
-
+DONE
 - I don't understand where the 16,800 number comes from. Is it an exhaustive
@@
   in X amount of time?)
-  ⟶  The former, we will clarify in the text.
 
-chrdimo: It is an exhaustive enumeration. We will clarify the prose.
-
+DONE
 - How large are the programs in the benchmark suite?
-  ⟶ The program sizes without type annotations, as reported by Racket's `syntax-sloc` tool, are below. We will add this information to the table in Figure 2.
-    benchmark | LOC | components
-	----------|-----|-----------
-	acquire | 1941 | 9
-	gregor | 2336 | 13
-	kcfa | 328 | 7
-	quadT | 7396 | 14
-	quadU | 7282 | 14
-	snake | 182 | 8
-	suffixtree | 1500 | 6
-	synth | 871 | 10
-	take5 | 465 | 8
-	tetris | 280 | 9
-
-chrdimo: Please find below the loc and number of components for each
-benchmark. We will add the information to figure 2.
-
-   benchmark | LOC | components
-	----------|-----|-----------
-	acquire | 1941 | 9
-	gregor | 2336 | 13
-	kcfa | 328 | 7
-	quadT | 7396 | 14
-	quadU | 7282 | 14
-	snake | 182 | 8
-	suffixtree | 1500 | 6
-	synth | 871 | 10
-	take5 | 465 | 8
-	tetris | 280 | 9
 
@@
 
+DONE
 - L38: Should there have been a colon? Is the text "This contrast... " and
   challenge quote from [3]? Who is saying what here...
-  ⟶ Thanks, we will reword the text from "This contrast" to clarify that this our research question, it is not coming from [3].
 
-chrdimo: No, the quote is not from [3]. We will rephrase to clarify that
-it is our statement of the research question.
+DONE
+- L104: I was confused by the word "Next". Maybe instead: We now ..."
 
-- L104: I was confused by the word "Next". Maybe instead: We now ..."
-  ⟶ Thanks, we will change the wording.
+DONE
+- L268: Grammar issue.
 
-chrdimo: Thank you. We will reword.
+DONE
+- L314: Floating period.
 
-- L268: Grammar issue.
-  ⟶ Thanks, we will fix this.
+DONE
+- L441: Add line numbers (or other metric of size).
 
-- L314: Floating period.
-  ⟶ Thanks, we will fix this.
-
-- L441: Add line numbers (or other metric of size).
-  ⟶ See above, we will add this information.
-
-- L980: Figure 5.4??
-  ⟶ Thanks, this is a bad reference; it should be figure 6. We will fix it.
-
-chrdimo: We will fix these. Thanks.
-
+DONE
 - L956: 30,000 hours of compute time is 1,250 days. I assume that the CPU has
   many cores. Could you put that number?
-  ⟶ Each CPU has 28 cores and 56 threads; when we last ran the experiment, it took about two weeks of real time from start to end. 
-  We will add this information at the start of section 7.
 
-chrdimo: Sure. Each CPU has 28 double cores.
+DONE
+- L980: Figure 5.4??
 
+DONE
 - L991: Figure 6: It took me some time to understand the figure. I wonder if
   there is a better way to represent the same data?
-  addressed
-  TODO: come up with some thoughts for this figure
-  ⟶ Thanks, we can see how it's difficult to read. We will consider some different presentations, such as...
-    - a bunch of two-sided bar charts? Where there's one bar pair per pair of modes. Above the x-axis is % A > B, and below the axis is % B > A?
-	- a bunch of stacked histograms? Where there's one stack per pair of modes. A bar has three parts: A > B, A = B, and B > A.
-
-chrdimo: We have reached the same conclusion and we are considering
-alternatives such as a series of two-sided bar charts that make it easier
-to spot how two modes compare to each other.
 
@@
 
+DONE: this is directly addressed in minor points, and vaguely by the high lvl response
 The experiment results show that there are debugging scenarios where
@@
 
-  addressed
-  TODO: Response idea: This is an example where the experiment seems to contradict the theory.
-	The models don't include all of the features in the benchmarks. <enumerate them here> It's possible that the Natural, Transient theories don't scale as expected to rich PL features.
-	Make analogy to physics: theory makes predictions, we have found failures of the predictions. We have two courses of action: report the inconsistency to re-eval the theory, and do more experiments.
-	It's also possible we have a bug in TR, or in Transient implementation.
-
-chrdimo: We were also surprised to discover that transient outperforms
-natural in some scenarios. After all, the theory of blame says otherwise. Even thought
-we haven't identified specific patterns, we believe that the reason behind 
-transient's unexpected success is due to the advanced features of Typed
-Racket's type system that haven't appeared in models that compare Natural
-and Transient: polymorphism, row types, mixins etc. Such features require 
-custom loopholes to be compatible with Racket idioms. In turn these
-loopholes can result in subpar checks in Natural. Our experimentation with 
-transient indicates that the same problem doesn't occur in transient. 
-We do not have any clues about a strategy that outperforms both Natural
-and Erasure. But our current conjecture is that a simple strategy that
-just points somewhere in the blame trail may be good enough and enable
-downstream optimization that are keeping track of blame in Typed Racket
-inhibits. The good thing is is that now we have a method to compare
-different strategies.
-
-chrdimo? We need more details from Ben here.
-
+DONE
 Theoretical gradual type systems usually allow programmers to control
@@
 
-  addressed
-  TODO: The answer is yes, but this needs more thought to frame the answer the right way.
-  ⟶ Our evaluation framework should be directly applicable to blame in gradual typing systems using Dyn.
-    The details of how it applies depends on the granularity of the blame information offered by the system -- in other words, what the system considers components.
-	If components are modules, as in Typed Racket, then nothing changes.
-	Systems with Dyn more likely treat top-level definitions as components, in which case definitions are individually typed (the whole definition is annotated) rather than modules.
-
-chrdimo: Yes. Incorporating type Dyn boils down to considering a hierarchy
-of types for each component rather than a single type. At the same time,
-our method is parametric to the ``size'' of a component and adpats in a
-straightforward manner to a setting where each definition can be
-considered a component instead of a whole module. Of course both of these
-adjustments lead to significantly larger scenario lattices than the ones we
-investigate and they have to be sampled aggresively and carefully.
-
+DONE
 Please explain what impedance mismatches are in the paper. Is it
 simply type mismatches?
-
-  ⟶ Impedance mismatches are mismatches between a piece of code's true type and its annotation.
-    For instance, the code `(λ (x) (+ x 42))` has type `Number -> Number` but might be annotated as `Number -> String`, causing an impedance mismatch.
-    We will adjust the text to clarify or remove this terminology.
-
-chrdimo: An impednace mismathc denotes a type annotation that doesn't
-match the actual type of the annotated code. We will adjust the prose to
-clarify.
 
@@
 
+DONE
 The paper states that the effort distribution of the random mode
@@
 
-  addressed
-  TODO: response sketch: you're right, we misspoke, this is not the normal distribution; our reasoning for why the normal-ish shape of distribution we see makes sense is that given a fixed trail the probability at each step of the trail ending is 1/n, but we randomly select trails so we're multiplying the random distribution of trail lengths by that 1/n one?
-Not clear on the details of this reasoning.
-  ⟶
-
-chrdimo: You are right that for a given trail of length n the probability
-that the random rational programmer discovers the buggy component after k 
-attempts is always 1/n and independent of k. However we sample randomly
-from the set of sceharios and as a result the set of trails. So the curve 
-of the random programmer matches our sampling distribution. That said, as
-you observe, what is the distribution of the random programmer is an
-irrelevant point to our results and we will remove the characterization.
-
+DONE
 * page 9 line 980: Figure 5.4 -> Figure 6
-  ⟶ Thanks, we will fix this.
 * page 10 Figure 7: Null -> Random
-  ⟶ Thanks, we will fix this.
 * page 11 line 1103: ( 770 -> (770
-  ⟶ Thanks, we will fix this.
-
-chrdimo: Thank you. We will fix all these.
 
@@
 They then conduct experiments to quantify the number of steps required by a rational programmer for several blame assignment schemes: Natural (used in Typed Racket, which tracks at most one location for blame), Transient (used in Reticulated Python, which tracks a set of locations), and Erasure (which does not actually assign blame, and is widely used in industry).
-
-  TODO: do we need to clarify that effort is a secondary aspect of the evaluation, and more important is whether blame leads to the bug at all?
+TODO: do we need to clarify that effort is a secondary aspect of the evaluation, and more important is whether blame leads to the bug at all? This is not explicitly clarified anywhere in the rebuttal.
 
@@
 But while there is a lot to like about the paper, it also has some shortcomings. The suite of programs used for the evaluation is fairly small (c.f., the SIGPLAN checklist).
-
-  addressed
-  TODO: respond to this ^, saying that we use the standard suite of benchmarks for evaluating GT -- point to Ben's POPL paper and others, and we use all of the largest benchmarks in the suite
-
-chrdimo: Coupld you please clarify what you mean by fairly small with
-respect to the SIGPLAN checklist? We are the first to conduct an empirical
-study for blame and gradual typing so we repurposed the standard
-performance benchmark suite for Typed Racket and, in particular, its
-largest and most complex benchmarks. Our inspection of the benchmarks
-indicates that they are representative set of Racket programs in terms of
-features and structure.
-
+DONE ^
 The faults are introduced using synthetic mutations that may or may not correspond to the kinds of errors that arise in practice.
 This can be seen from the results in Figure 7 -- the lengths of the "trails" is quite small.
-
-  addressed
-  TODO: unclear that syntheticness relates to trail length at all, and synthetic bugs are the only option here -- there is no corpus of documented GT bugs in the wild anywhere.
-Indeed, most type errors never even make it into code repositories.
-Note that we are the first to evaluate blame in this setting.
-
-chrdimo: There is no exisitng catalogue of bugs in gradually typed
-programs. Furthermore, most type-level bugs do not survive after decent tsting and 
-thus do not make it to code repos. There are a few anecdotes from the literature but they are mostly
-artificial. Hence, since we need a large number of buggy scenarions for a
-meaningful study, a synthetic approach is the only way forward. In
-addtion, with our mutation we can tune the set of scenarios to get a
-diverse set of bugs.
-
+DONE ^
 While the study is mostly well done, it only considers two blame assignment strategies.
+DONE ^
 For example, it would be interesting to consider the "omniscient programmer" to get a bound on the value of blame. But this is not done.
-
-  addressed
-  TODO: We don't understand what is meant by "omniscient programmer", but our best understanding is the ideal oracle strategy that always picks the right module right away and thus always finds the bug in one step.
-We didn't show this because it's trivial.
-We do have other baselines to understand blame's performance: the random mode, and the exceptions modes for each of the semantics.
-
-chrdimo: Could you please explain what you mean by "omniscient
-programmer"? If this is the programmer that always makes the best choice,
-i.e., finds the bug at the first try then this programmer is trivial. It
-has 100% success rate and a constant blame trail of 1. Thus comparing it
-with any strategy amounts to the number of scenarios where a strategy
-fails and those where it succeeds but the blame trail has length greater
-than one. This information is in the paper byt If you think it would be
-useful, we point it out. Moreover, the experiment considers 10 different
-modes of the rational programmer spread across 3 different gradaul typing
-systems.
-
+DONE ^
 And unfortunately the results are inconclusive -- Section 8 says "our results call for a deeper understanding of the two models for blame".
-
-  addressed
-  TODO: This is what makes our results surprising: they challenge the theoretical conclusions that Natural is superior to Transient, suggesting that the theory is not good enough to understand the pragmatics of blame in reality.
-
-chrdimo: The quote from section 8 refers to the theory of blame rather
-than our results. Our results show in clear terms that, at least for our
-benchmarks, a blame strategy is superior that exceptions and that natural
-and transient have comparable efficacy. The latter is surprisng given that
-theoretical models (and our intiail expectations) point to the opposite.
-Thus a side-effect of our experiment is a call to theoriticians for models
-that are atuned with practice.
-
+DONE ^
 Also, the published Transient scheme seems to have a minor bug, which might affect the outcomes when fixed (Section 8.2).
-  addressed
-  TODO: 8.2 does not indicate a bug in Transient, but rather a defect in the design of the system. ... something about a gap between Reticulated Python and
-
-chrdimo: Section 8.2 does not discuss a bug in Transient but rather a
-limitation of its current design that we discovered exactly becasue of the
-experiement.
-
+DONE ^
 Most if not all of these shortcomings are discussed in the paper, which is much appreciated. But they do risk undermine the value of the contributions to some degree.
 
-chrdimo: We would like to point out that the resutls of the comparison
-between the tree systems are only one part of the contribution of the
-paper.  The other part is the method itself. Benchmarks, source of
-interesting bugs and rational programmer modes are all parameters that can
-be instantiated differently without affecting the recipe we discuss at the
-end of section 2 nor the analytical framework from section 5 that
-underlies our experiment.
-
+DONE
 To explain one comment above: I was surprised not to see a comparison against an "omniscient programmer" to establish a baseline (c.f., the SIGPLAN checklist).
@@
 
-chrdimo: Our experiemnt does have a baseline. The random mode of the
-rational programmer serves as the ``control'' of our experiment. 
-Furthermore the exception modes serve as further ``controls'' to
-distinguish the effect fo checks from that of blame,
-
+DONE
 Similarly, I wondered why the modes of the transient programmer only consider picking the first/last blame, and not other choices.
 Section 5.2 just says "our answer" is that there are at least "two reasonable options." Why are other choices not reasonable?
-  addressed
-  TODO: Answer suggested by Ben: it's unspecified how one should interpret Transient blame, and there isn't time to evaluate all possible choices, so we pick the most reasonable interpretations.
-
-chrdimo: The transient semantics provide no interpratation for the blame
-set. We picked the first and last because they match the intuitive
-interpretation of the blame set as a list. We didn't mean to imply that
-other options are unreasonable. We just needed to make a reasonable choice 
-and keep the duration of the experiemtn within reasonbale limits.
 
@@
 
+DONE
 2. That Transient --- as implemented in Typed Racket --- is much
@@
 of incidental implementation choices or bugs affecting this result.
-
-  TODO: We should clarify that checks alone is faster, but adding blame is slower (bc blame map is linear).
 
@@
 
+DONE
 *Natural and Transient Under the Same Roof (Section 3)*
@@
 
-  TODO: this is a minor part of the paper (not a result), there is a large gap between the model and what's needed to scale to a full lang -- it's not clear from the simple model how to design check insertion and blame tracking for new features.
-This required research, will be published elsewhere
-
+DONE
 *Custom Mutators (Section 4)*
@@
 
+DONE
 *Debugging Strategies (Section 5)*
@@
 
-  TODO: the analytical part is important: we described all of these things in a common framework, allowing for comparison
-
 I'm also not sure why the exception mode is needed in addition to
@@
 
-  TODO: the exception modes are a key part of the analysis framework, enabling understanding of the value of blame separately from checking
-
+DONE
 *Additional Comments and Typos*
@@
 -------------
+DONE: clarifying that we propose a methodology, and what the key experimental results are
 This paper seeks to establish (or refute) the benefit of blame tracking in gradually typed programming environments.
@@
 
-  TODO: misunderstanding: we propose a methodology
-  ⟶ The primary intended contribution of this paper is the methodology, and the experimental results serve as secondary contributions that validate the methodology's utility.
-    We will rephrase the introduction to further emphasize that focus.
-	Furthermore, we would like to emphasize that the two key findings in the experimental results are that 1) both Natural and Transient's dynamic type enforcement with blame tracking improve significantly over Erasure, and 2) Natural's blame tracking is not significantly better than Transient's, despite the theoretical indications to the contrary.
-
 Comments for author
@@
 
+DONE
 However, I don't find the results to be as conclusive as the authors state.
@@
 
-  TODO: misunderstanding: the results are not at all "what we'd expect" based on the theory
-  ⟶ The status quo of what we would expect based on the current theory of Natural and Transient is that 1) either kind of blame tracking should be better than none, and 2) Natural should provide significantly better blame information than Transient and Erasure.
-    Our results validate the first expectation that blame provides benefits over plain exceptions.
-    Surprisingly, our results contradict the second expectation and suggest that in fact Natural and Transient offer comparable blame information.
-
 The presentation of the results is very much around aggregates and summaries.
@@
 
+DONE
 About Erasure: if I'm reading this correctly, the key thing here is that in Typed Racket, the set of static types is more expressive than the set of dynamic types.
@@
 
-  TODO: misunderstanding about gradual typing as a whole, and the point of the exception modes
-  ⟶ The idea of the Erasure semantics is not related to the expressiveness of the type system or how the types relate to kinds of runtime values.
-    Rather, the key idea is that Erasure does not enforce type annotations with dynamic checks at all.
-	Running a gradual program using the Erasure semantics is the same as stripping all annotations and running the resulting (completely dynamic) program.
-	Thus, Erasure programs can only raise exceptions from either the checks performed by the language's primitive operations (e.g. `+`), or from the program itself raising an exception.
-	In contrast, the other two semantics do enforce type annotations dynamically; both Natural and Transient insert dynamic checks verifying that the shape of values at runtime match their static annotations.
-	Thus, in addition to exceptions, gradual programs using these semantics may raise a second kind of error: a dynamic type check failure.
-	These dynamic errors can include blame information if the dynamic type checks are designed to track blame, or they can forego blame and just offer a stack trace.
-	In the case of Natural, these two options are captured by the "Natural" and "Natural exceptions" modes, respectively.
-	Thus the "Erasure" mode represents a completely different kind of checking (and thus error reporting) compared to both the "Natural" and "Natural exceptions" modes (the two of which share the same kind checking, but different error reporting).
-
+DONE
 It's interesting to note that there seems to be no convincing analogous experimental or modelling-based justification for plain old static typing, i.e.
@@
 
-  ⟶ (Matthias) As posed, the question expresses a significant mis-characterization of
-    the submission. It is not the benefits of a gradual type system that
-	are in doubt. The experimental setup keeps the type system constant
-	but allows to answer the question which of several run-time checking
-	regimes (for enforcing type consistency) provides the best explanatory
-	messages in case of violations.
-	Hence the analogous question for purely static type systems would ask
-	which of several reporting schemes provides the best explanatory
-	message for type-errors. In the case of simply typed languages and
-	even language with local type inference, this question is basically
-	meaningless. In the case of languages with HM type inference, the
-	question has been implicitly raised for four decades with the
-	development of alternative ways of finding the source of inference
-	conflicts.
-	If the reviewer is indeed interested in the question of whether type
-	systems help programmers---a question that this submission does _not_
-	ask---the recent OOPSLA literature contains several user studies. A
-	simple Google query will suggest a short list of these publications.
-
 Detailed comments:
 
-  TODO ll: added responses to anything below that seems to warrant more than "Thanks, we'll fix it".
-
+TODO
 In the title, "evaluating blame" reads oddly.
@@
 
+TODO
 The abstract doesn't say much about the work. It would be better written for experts to quickly gather an overview what the paper contributes.
 
+TODO
 Conversely, the main body of the paper skimps a bit on background.
@@
 please say exactly what it means here.
-  ⟶ We will adjust the prose to more explicitly describe blame and impedance mismatches.
 
+DONE
 line 34: "then their compilers remove types and rely on the built-in safety checks of the underlying language to catch any problems".
@@
 
-  ⟶ The static checking performed by languages using the Erasure semantics (like Flow, Hack, and TypeScript) is done in a best-effort fashion; the fact that a program passes type checking has absolutely no effect on how the program runs.
-    These languages literally erase the types to obtain a program of the underlying language (e.g. TypeScript or PHP) to run.
-	Those underlying languages perform all of the dynamic checks that a typical dynamic language performs for primitive operations (e.g. that `+` is only used with numbers).
+DONE
+line 38: "explicit statement and challenge" -- what is it?
 
-line 38: "explicit statement and challenge" -- what is it?
-  ⟶ The challenge questions whether blame is useful at all in gradual typing.
-    We will clarify this in the prose.
-
+DONE
 line 49: I agree they got the word wrong, but you should explain this
 
+DONE
 line 60: around this paragraph the writing started to grate.
@@
 Throughout the paper, much space could be saved by writing in a more direct style.
-  ⟶ Noted, we will simplify the language where possible.
 
+DONE
 line 69: be explicit that Lazarek et al were (as I later gathered) doing something about blame to do with higher-order contracts but not gradual types
 
+DONE
 line 70: what does it mean to "follow" the slogan? Was unclear to me.
-  ⟶ The following sentences are meant to explain this, so we will adjust the prose to make that clear.
 
+DONE
 line 90: at first I wondered: what is a case? Maybe say "program variant" to foreshadow the idea of generating mutants etc?
 
+DONE
 line 91-ish: "Transient, "Natural", "Erasure" -- be explicit that these are names that *you* are introducing
-  ⟶ These are names from the Gradual Typing literature (see for instance [9, 10, 37]).
 
+DONE
 line 94: "forego" => "forgo"
 
+DONE
 line 119: "homo economicus" needs glossing or a reference
 
+DONE
 line 125: was wondering whether "impedance mismatch" just means "feasible run-time type error".
 
+DONE
 line 148: last sentence appears to contradict the preceding paragraph, and is left hanging oddly. Instead, make it the start of a new paragraph.
 
+DONE
 line 160: "disparity" is a strange word here. "diversity"?
 
+DONE
 line 170: "Otherwise, results from..." -- the point was already clear
 
+DONE
 line 174: ... instead of just saying what you built on, first say what you did! It's really not clear at this point.
 
+DONE
 line 187: what is a "component"? Should be easy to define.
 
+TODO
 line 199: "blame set as another form of a stack trace" -- yes.
@@
 
+DONE
 line 211: "languages exceptions" typo
 
+DONE
 line 268: this explanation of the workings of proxies seemed overwrought
 
+DONE
 line 277: what is "responsibility" of a "party", exactly? Are "party", "module" and "component" all the same things?
-  ⟶ Components are parties, and in the case of Typed Racket modules are components because they are the units of migration.
-Other gradual type systems, like those that do not require annotated entire modules, may consider smaller units of code (e.g.
-definitions) as components.
-    We will clarify this in the prose.
 
+DONE
 line 283: "rewrites typed modules to inline checks" -- so a module gets turned into a check? Clearly not, but that's how it reads
 
+DONE
 line 293: I was wondering what constitutes a boundary crossing.
@@
 What about values exchanged through reads/writes to shared state?
-  ⟶ Yes, communicating a value via state also constitutes a boundary crossing.
-    We will clarify what a boundary crossing is in the prose.
 
+DONE
 line 299: "crosses" => "crossings" (probably)
 
+DONE
 line 313: it threw me that a program might not fail but "produce a wrong result".
@@
 In certain other systems this wouldn't be possible, because the static checker would only catch (albeit earlier) errors that would be caught at run time, so there would be no basis to call the result "wrong".
-  ⟶ The result can be wrong in that it doesn't agree with the annotations (which we assume express programmer intent), even if none of the primitive operations applied to the result raise an error.
-    For example, imagine the the client saves the result to a file.
-    The language primitives for writing values to a file (typically) can write any kind of value to a file, so they have no checks that could fail due to the result being of the wrong type.
-    Instead, the resulting file just ends up with the incorrect contents and the program terminates normally.
 
+DONE
 lien 352: "three interpretations" -- what are they? I don't see them in the figure.
 
+DONE
 line 380: from this I inferred that "migratory typing" means "gradual typing applied at modulewise granularity". Assuming that's correct, it's worth saying directly.
-  ⟶ The three choices of using Dyn, the granularity of components, and how compilation affects completely untyped components represent orthogonal aspects of the design of a Gradual Typing system.
-    In the case of Typed Racket's migratory typing, those three choices amount to more than just applying the ideas of [21] at the granularity of modules.
 
+DONE
 line 421: "type mistake" -- does this mean "feasible run-time type error"?
-  ⟶ Could you clarify what a makes a runtime type error "feasible"?
 
+DONE
 line 424: "fully typed correct programs" -- presumably your method could also work with not-yet-fully-typed correct programs, just not ranging over the entire lattice in those cases.
 I was wondering whether that might give different/interesting results.
-  ⟶ Yes, the approach could be applied, but the results may be mostly inconclusive depending on how many trails reach the point where the blamed component lacks type informtion.
-    All trails that reach that point provide no useful information.
 
+DONE
 line 433: "without loss of diversity" -- this is glib. Clearly diversity is lost; just claim that what remains is still diverse enough.
 
+DONE
 line 497: "truthiness" needs explaining
 
+DONE
 line 517: the difference between #1 and #2 here again relies on the surprising property (to the unfamiliar) that Typed Racket has a notion of "type-level mistake" that doesn't surface under erasure (as an exception or whatever) but also is not a false positive (i.e.
@@
 Line 538's "unavoidable" claim is probably also true only in such a context.
-  TODO: Don't respond to this in light of above explanation(s) of why they misunderstand erasure and TR?
 
+DONE
 line 519: "at least three" -- better to claim this is only two here, then explain later that the driver doesn't count.
@@
 
+DONE
 line 634: earlier I had inferred that the distinction between "migratory" and "gradual" was more than just preferring one word; this seems to contradict that
 
+DONE
 line 640: what makes it "concise"?
-  TODO ll: Why does it? I'm not sure about the word here either. Just say we'll change it?
 
+DONE
 line 673: "programmers runs" typo
 
+DONE
 line 727: "dubbed" -- reads oddly. Maybe italicise the "location", but it seems overkill. The phrase is pretty self-suggesting as it is.
 
+DONE
 line 756: "checked the value's type.." -- and the check passed!?
-  ⟶ Indeed, Transient checks only check the top-level type constructor of a value, so many checks can pass for a bad value before it is unwrapped to discover the problem.
-    We will add a reminder of this fact to the prose here.
 
+DONE
 line 752: "added to the blame set first" -- should it be a blame list, then?
 
+DONE
 line 936: no need for hyphen after the adverb
 
+DONE
 line 966: "more useful C" -- missing "than"
 
+DONE
 line 982: don't think these percentages deserve 3 significant figures
-  TODO: "deserve"?
-  ⟶ We include the hundreths place to highlight in particular that Transient first and last blame are not exactly the same.
-    If this information isn't useful, we can use just the tenths place.
 
+DONE
 line 1304: "Problem is" -- missing "The". Also, no need to italicise the next sentence... it is really not that deep or surprising.

### 2021-02-10 writing/reviews/pldi-rebuttal-high-level-first.md
Minor edits and rewordings

@@
 experiment provides evidence that performance issues, which stem from the
-design of Transient, have a negative effect on the effectiveness of blame.
+design of Transient, negatively impact the effectiveness of blame.
 
-(ii) Theory claims that natural blame is more accurate than transient
+(ii) Theory claims that Natural blame is more accurate than Transient
 blame but our experiment provides evidence for the opposite;
+ll: TODO is "the opposite" the right phrasing? It seems more accurate to say evidence that doesn't support the prediction.
 
@@
 detectable by behavioral contracts. Hence, they can use off-the-shelf
-mutators. In fact, their method does not detect the overwhelming majority
-of bugs except when contracts describe the full functional correctness of
-components.
+mutators. In fact, their contracts do not detect the overwhelming majority
+of bugs except at the highest precision, where the contracts describe the full
+functional correctness of components.
 
@@
 
+ll: TODO phrasing of "original set we tried" seems off
 The sixteen mutators we include in the paper are not the original set we
@@
 Natural without blame but with the same checks outperforms Erasure in ~25%
-of the scenarios (figure 6).  Blame add another ~11% on top of that. For
+of the scenarios (figure 6).  Blame adds another ~11% on top of that. For
 Transient there is a small percentage of scenarios where exceptions due
@@
 
-  benchmark    | LOC  | components -------------|------|-----------
-  acquire      | 1941 | 9 gregor       | 2336 | 13 kcfa         | 328  | 7
-  quadT        | 7396 | 14 quadU        | 7282 | 14 snake        | 182  |
-  8 suffixtree   | 1500 | 6 synth        | 871  | 10 take5        | 465  |
-  8 tetris       | 280  | 9
+  benchmark    | LOC  | components
+  -------------|------|-----------
+  acquire      | 1941 | 9
+  gregor       | 2336 | 13
+  kcfa         | 328  | 7
+  quadT        | 7396 | 14
+  quadU        | 7282 | 14
+  snake        | 182  | 8
+  suffixtree   | 1500 | 6
+  synth        | 871  | 10
+  take5        | 465  | 8
 
@@
 any patterns for these scenarios. We conjecture that they point to
-limitations of the theory but we can;t exclude that some of them aren't
+limitations of the theory but we can't exclude that some of them aren't
 because of some unknown Typed Racket bug. Please see the first part of the
@@
 of scenarios to get a diverse set of interesting bugs.  We discuss in
-  section 8 how our approach to bug injection is a threat to validity,
+section 8 how our approach to bug injection is a threat to validity.
+ll: TODO the review already acknowledges that we discuss this point in the threats; do we need to repeat that?
 
@@
 
-We would like to point out that the experiment considers ten different
+We would like to point out that the experiment considers nine different
 modes of the rational programmer spread across three different gradual
 typing systems.
+
+ll: TODO are we calling the random modes for each of natural, transient, and erasure distinct? I'm not sure that makes sense, in which case it's seven modes.
 
@@
 of checks from that of blame within each semantics. Please also see our
-discussion of modes in the second theme at the beginning of our reply,
+discussion of modes in the second theme at the beginning of our reply.
 
@@
 
-The transient semantics provide no interpretation for the blame set. We
+The Transient semantics provide no interpretation for the blame set. We
 picked the first and last because they match the intuitive interpretation
@@
 
+> 2. That Transient --- as implemented in Typed Racket --- is much
+> slower than Natural. ...
 > Given that the experiments are developed and run in the context of the
@@
 an artifact of our implementation we have coded up benchmarks in Python
-and we confirmed that in Reticulated the performance of transient with
+and we confirmed that in Reticulated the performance of Transient with
 blame is even worse than what we have observed in Typed Racket.
@@
 
+ll: TODO "absolutely disagree with *you*" seems too personal and confrontational. I'm not sure, but just referring to the beginning should speak for itself?
+
 > About Erasure: if I'm reading this correctly, the key thing here is that
@@
 discover mistakes and help the IDE. Then types are erased and the program
-is run. Since the program is not fully typed, the run time type checks of
-primitive operations may detect an incompatible argument and raise an
-exception.
+is run in the underlying dynamic language. That language's primitive
+operations have runtime checks that may detect an incompatible argument
+and raise an exception.
 
@@
 checks is determined by the types and matches those. In turn the types of
-primitive operations much the checks that their Racket counterparts
+primitive operations match the checks that their Racket counterparts
 perform. Finally our benchmarks come with the most precise types that
@@
 Yes, the approach could be applied.
+ll: TODO worth a sentence saying that this would probably not be worthwhile to do?
 

### 2021-02-10 writing/reviews/pldi-rebuttal-checklist.md
bg edits, started small but got excited

@@
 
-TODO: This seems to imply that we are proposing or assuming the way that real "rational programmers" would behave. Should we try to answer or clarify about that?
-
 Comments for author
@@
 
-To evaluate this approach, the paper also presents empirical results.
-Starting with a small corpus of programs, the authors use mutations to systematically (but synthetically) introduce type errors -- e.g., one mutation replaces a constants with another of a different type, and another mutation swaps method identifiers.
-They then conduct experiments to quantify the number of steps required by a rational programmer for several blame assignment schemes: Natural (used in Typed Racket, which tracks at most one location for blame), Transient (used in Reticulated Python, which tracks a set of locations), and Erasure (which does not actually assign blame, and is widely used in industry).
-TODO: do we need to clarify that effort is a secondary aspect of the evaluation, and more important is whether blame leads to the bug at all? This is not explicitly clarified anywhere in the rebuttal.
+To evaluate this approach, the paper also presents empirical results. Starting
+with a small corpus of programs, the authors use mutations to systematically
+(but synthetically) introduce type errors -- e.g., one mutation replaces a
+constants with another of a different type, and another mutation swaps method
+identifiers. They then conduct experiments to quantify the number of steps
+required by a rational programmer for several blame assignment schemes: Natural
+(used in Typed Racket, which tracks at most one location for blame), Transient
+(used in Reticulated Python, which tracks a set of locations), and Erasure
+(which does not actually assign blame, and is widely used in industry).
 
@@
 The presentation of the results is very much around aggregates and summaries.
-Indeed the whole method is about having run a huge compute job over a large number of variants.
-Perhaps it's paranoia but I'd be interested to see some specific examples walked through by hand (in the paper) and some smaller sample manually classified (as results).
-That would add an extra sanity check that the metrics do correspond to some meaningful reality.
+Indeed the whole method is about having run a huge compute job over a large
+number of variants. Perhaps it's paranoia but I'd be interested to see some
+specific examples walked through by hand (in the paper) and some smaller sample
+manually classified (as results). That would add an extra sanity check that the
+metrics do correspond to some meaningful reality.
 
@@
 TODO
-line 199: "blame set as another form of a stack trace" -- yes.
-I was hoping to see a clearer example of debugging with and without blame, i.e.
-something making explicit the similarities and differences between having blame info and having only a stack trace at the error site.
+line 199: "blame set as another form of a stack trace" -- yes. I was hoping to
+see a clearer example of debugging with and without blame, i.e. something
+making explicit the similarities and differences between having blame info and
+having only a stack trace at the error site.
 

### 2021-02-10 writing/reviews/pldi-rebuttal-high-level-first.md
bg edits, started small but got excited

@@
-PLDI 2021 Paper #66 Author Response
-===========================================================================
-
-Thank you to all the reviewers for the helpful feedback.
-
-We start off with a couple of common themes. Then we reply to all points
-individually.
-
-The conclusions from our experiment
-===========================================================================
-
-There are three clear major conclusions from our experiment:
-
-1) The Transient and Natural rational programmers are more effective than
-the Erasure one;
-
-2) Part of  the additional effectiveness of the former is due to runtime
-checks (that Erasure foregoes) but another part is due to blame;
-
-3) The effectiveness of the Natural and Transient rational programmers is
-comparable.
-
-The first two results confirm what the vast literature on gradual typing
-claims without systematic evidence: blame is useful. However, as review
-`D` astutely points out, the third conclusion is surprising in two
-different ways:
-
-(i) An important motivation for Transient is performance but our
-experiment provides evidence that performance issues, which stem from the
-design of Transient, negatively impact the effectiveness of blame.
-
-(ii) Theory claims that Natural blame is more accurate than Transient
-blame but our experiment provides evidence for the opposite;
-ll: TODO is "the opposite" the right phrasing? It seems more accurate to say evidence that doesn't support the prediction.
-
-Thus our results point to two directions for further research in gradual
-typing:
-
-(i) Implementors can experiment with checks and blame to find a sweet spot
-between type soundness, blame effectiveness and performance. Our method
-provides an automatic tool to guide their exploration.
-
-(ii) The predictive power of gradual typing formal models (the reason
-d'etre of scientific models) fails when it comes to blame. This may be
-because the current models are too small and do not include crucial
-features or because the current models are inherently limited. In any case
-our results are a call to theoreticians to re-examine their assumptions
-about blame.  Our method provides an automatic tool to validate their
-models similar to the way the methods of experimental physicists help
-prove or disprove the  math of theoretical physicists.
-
-Challenges for our work compared to Lazarek's POPL paper
-===========================================================================
-
-Our work builds on the idea of Lazarek et al.'s paper that it is possible
-to evaluate the effectiveness of blame by simulating a debugging process.
-Both their work and our work is inspired by Greenman et al's work that
-uses a lattice of configurations to systematically explore variants of the
-same program with specifications of different precision (for performance
-reasons in Greenman et al).  Both the ideas of ``following`` blame while
-debugging and of gradually adding specifications to tweak the evaluation
-of the program originate from the practice of contracts (see the story in
-the intro of Lazarek's paper and similar anecdotes in Meyer's writings).
-
-That said,
-
-1) In the context of Lazarek et al.'s work interesting bugs  are those
-detectable by behavioral contracts. Hence, they can use off-the-shelf
-mutators. In fact, their contracts do not detect the overwhelming majority
-of bugs except at the highest precision, where the contracts describe the full
-functional correctness of components.
-
-2) Lazarek et al.'s method does not isolate the effect of blame from that
-of checks or mere luck. Thus they cannot really conclude that blame is the
-reason for the success of the debugging process --- only that blame
-shifting works. In other words, their work is lacking proper `controls'.
-
-3) Lazarek et al. deal with a single contact system and examine a yes or
-no question: does blame-shifting find the bug for all debugging scenarios
-of a benchmark. They do not provide any guidance how to generalize to
-multiple systems with different notions of checks and/or blame in a
-uniform way that also allows comparisons.
-
-This brings us to the three problems we had to solve in order to design
-our experiment:
-
-1) Natural and Transient under the same roof (section 3).
-
-Implementing Transient Typed Racket is not mere engineering.  Neither
-Reticulated nor the Transient model provide answers about what shallow
-checks correspond to Typed Racket's sophisticated types (e.g.
-polymorphism, row types for mixins, occurrence typing) and what
-information is necessary to record in the blame map. Moreover in the
-process of the implementation we discovered issues with the published
-Transient model and Reticulated that we had to fix. We do not think that
-this paper is the appropriate place to describe the implementation in
-detail.
-
-2) Custom mutators (section 4).
-
-The identification of good mutators for gradual typing is not
-straightforward. A bad mutator results in a type-level bug that is trivial
-to locate. For instance, a bug that leads to a type-level error during the
-evaluation of the component that contains it is trivial to locate. In
-contrast, a bug that is non-trivial to locate is one that causes the buggy
-component to evaluate without an error but produce instead a value of the
-wrong type. Furthermore an interesting bug is one where the value with the
-wrong type manages also to cross undetected to at least another component
-before a type-level check flags it. Good mutators should be able to
-produce these interesting bugs in our benchmarks. Finally, they should
-produce interesting bugs that are detectable by all three systems so that
-we can try all systems on the same debugging scenarios.  In general, there
-is no methodology for constructing a new set of mutators that produce
-interesting bugs for any definition of interesting. So narrowing down the
-ones in the paper required rounds of trials to end up with a set of
-mutators that leads to a sufficiently large set of interesting and diverse
-bugs.
-
-ll: TODO phrasing of "original set we tried" seems off
-The sixteen mutators we include in the paper are not the original set we
-tried and others that seemed like good candidates proved to be
-ineffective. For instance replacing car with cdr should lead to type-level
-bugs however none of them are interesting --- they blow up immediately.
-Similarly, swapping mutable vectors and hashes with immutable ones is an
-interesting mutation in the traditional sense and should lead to a type
-error in Typed Racket since its type system distinguishes the two, however
-it turns out such mutations  are not interesting in our setting.
-
-3) Modes of the rational programmer (section 5).
-
-A key step for realizing our experiment is the precise and unified
-description of debugging in the three gradual typing system  as  the
-different modes of the rational programmer. Because of their common
-definitional substrate of blame trails, modes enable the comparison
-between the different systems even when sampling is involved.  Furthermore
-we use modes to introduce in the framework baselines that help us isolate
-confounding factors from the actual effectiveness of blame. This is
-exactly the role of the exception modes for Natural and Transient; the
-exception modes allow us to determine what part of the success of Natural
-and Transient is due to the checks or due to blame (and also whether blame
-masks useful information from exceptions). For example, we see that
-Natural without blame but with the same checks outperforms Erasure in ~25%
-of the scenarios (figure 6).  Blame adds another ~11% on top of that. For
-Transient there is a small percentage of scenarios where exceptions due
-better than blame but still blame improves over exceptions by ~11%
-compared to Erasure. Finally, we introduce yet another mode, the random
-one, to serve as the baseline for the all modes and exclude that our
-observations are due to mere luck.
-
-As a final note, even though each of the above points individually may not
-have sufficient technical depth for a separate contribution (which we do
-not claim), together they exhibit the creative steps that are necessary to
-lift the scope and magnitude of the idea from Lazarek's POPL paper  from
-evaluating blame in a single behavioral contract system to a framework
-for evaluating and comparing blame in different gradual typing systems.
-
 Replies to individual points
@@
 
-An impedance mismatch denotes a type annotation that is detected to
-disagree with the actual type of the annotated code. We will adjust the
-prose to clarify.
+"Impedence mismatch" is our name for a run-time disagreement between a
+type annotation and an untyped value. The point of the name is to suggest
+that either side (type or value) could be at fault. We will adjust the prose.
 
@@
 ===========================================================================
+
+> Paper summary
+> -------------
+>
+> [....]
+>
+> To evaluate this approach, the paper also presents empirical results. Starting
+> with a small corpus of programs, the authors use mutations to systematically
+> (but synthetically) introduce type errors -- e.g., one mutation replaces a
+> constants with another of a different type, and another mutation swaps method
+> identifiers. They then conduct experiments to quantify the number of steps
+> required by a rational programmer for several blame assignment schemes: Natural
+> (used in Typed Racket, which tracks at most one location for blame), Transient
+> (used in Reticulated Python, which tracks a set of locations), and Erasure
+> (which does not actually assign blame, and is widely used in industry).
+
+chrdimo: TODO
+effort is part 2/2 of the metrics
+
 > But while there is a lot to like about the paper, it also has some
@@
 
-Could you clarify what you mean by fairly small with respect to the
-SIGPLAN checklist? We are the first to conduct an empirical study for
-blame and gradual typing, so we repurposed the standard performance
-benchmark suite for Typed Racket -- in particular, its largest and most
-complex benchmarks. Our inspection of the benchmarks indicates that they
-are representative set of Racket programs in terms of features and
-structure.
+The paper does follow the SIGPLAN checklist guidelines for Principled Benchmark
+Choice:
+
+- This paper is the first to conduct an empirical study for blame and gradual
+  typing, so there is no established benchmark suite to reuse.
+
+- The paper thus explains how we adapted a suite of gradual typing performance
+  benchmarks to a new setting (section 4).
+
+- The paper also explains how we selected pertinent benchmarks to work within
+  our resource constraints.
+
+We cannot find any minimum size guidelines in the SIGPLAN checklist.
 
@@
 
-There is no existing catalog of bugs in gradually typed programs.
-Furthermore, most type-level bugs do not survive  decent testing and thus
-do not make it to code repos. There are a few anecdotes from the
-literature but they are mostly artificial. Hence, since we need a large
-number of buggy scenarios for a meaningful study, a synthetic approach
-is the only way forward. In addition, mutation allows us to tune the set
-of scenarios to get a diverse set of interesting bugs.  We discuss in
-section 8 how our approach to bug injection is a threat to validity.
-ll: TODO the review already acknowledges that we discuss this point in the threats; do we need to repeat that?
+There is no existing catalog of bugs in gradually typed programs, nor any
+statistics about common trail lengths.  Furthermore, most type-level bugs do
+not survive  decent testing and thus do not make it to code repos. There are a
+few anecdotes from the literature but they are mostly artificial. Hence, since
+we need a large number of buggy scenarios for a meaningful study, a synthetic
+approach is the only way forward. In addition, mutation allows us to tune the
+set of scenarios to get a diverse set of interesting bugs.
 
@@
 
-We would like to point out that the experiment considers nine different
-modes of the rational programmer spread across three different gradual
-typing systems.
-
-ll: TODO are we calling the random modes for each of natural, transient, and erasure distinct? I'm not sure that makes sense, in which case it's seven modes.
+We would like to point out that the experiment considers six different
+modes of the rational programmer spread across three different gradual typing
+systems plus the random rational programmer. Figure 7 clearly shows them all.
 
@@
 > This seems to be "only" an engineering challenge.
-
+>
+> [...]
+>
 > *Custom Mutators (Section 4)* The paper suggests that the mutators
@@
 > 2020 paper.  ...
-
+>
+> [...]
+>
 > *Debugging Strategies (Section 5)*
@@
 
-We absolutely disagree with you. Please see our discussion of the
-conclusions form our experiment at the beginning of our reply.
+TODO chrdimo
+- see general comments for how to interpret results, their significance, and how they challenge existing perceptions
+- usefulness asks if one succeeds when another failed; no extent for this
+- if both succeed, can quantify difference, use trail length to do so
+- trail length is proxy, paper discusses limitations, allows automation ... no humans + large scale
 
-chrdimo: Should we be more assertive here about the metrics? I think this
-is a hard to defend position because of the non-constructive manner of the
-criticism --- it is plain dismissive without explanation.
+> The presentation of the results is very much around aggregates and summaries.
+> Indeed the whole method is about having run a huge compute job over a large
+> number of variants. Perhaps it's paranoia but I'd be interested to see some
+> specific examples walked through by hand (in the paper) and some smaller sample
+> manually classified (as results). That would add an extra sanity check that the
+> metrics do correspond to some meaningful reality.
 
-ll: TODO "absolutely disagree with *you*" seems too personal and confrontational. I'm not sure, but just referring to the beginning should speak for itself?
+TODO chrdimo
+....
+
+if you are asking whether our results
 
@@
 
-Noted, we will simplify the language where possible.
+Noted, we will rephrase.
 
@@
 
-> In the title, "evaluating blame" reads oddly. It is ambiguous, because
-> "evaluate" sometimes means to compute a result. A fuller phrase, like
-> "evaluating the usefulness of blame tracking...", would probably be
-> worth the words.
-
-> The abstract doesn't say much about the work. It would be better written
-> for experts to quickly gather an overview what the paper contributes.
-
-> Conversely, the main body of the paper skimps a bit on background. It
-> never explicitly covers what "blame" means and how it works in practice.
-> Similarly, it repeatedly talks about "impedance mismatches" without
-> defining them. Perhaps this phrase is now standard in the gradual typing
-> literature, which I haven't kept up with (hence my Z expertise). But I
-> did read the Wadler/Findler paper carefully at the time. It did not talk
-> about impedance mismatches. In any case, it is a fuzzy metaphor...
-> please say exactly what it means here.
-
 > line 313: it threw me that a program might not fail but "produce a wrong
@@
 
-There are two points that distinguish gradual typing from migratory
-typing: the existence of Dyn and the granularity of components.  We will
-clarify the prose.
+We will clarify. Migratory typing is not about granularity, but whether the
+type system forces rewrites upon untyped code. A gradual type system for Python
+would be migratory. A new gradual-from-the-start language would not be.
 
@@
 Yes, the approach could be applied.
-ll: TODO worth a sentence saying that this would probably not be worthwhile to do?
+ll: worth a sentence saying that this would probably not be worthwhile to do?
+bg: TODO try explaining in 2-3 words
 
@@
 value.  We will add a reminder of this fact to the prose here.
+
+> In the title, "evaluating blame" reads oddly. It is ambiguous, because
+> "evaluate" sometimes means to compute a result. A fuller phrase, like
+> "evaluating the usefulness of blame tracking...", would probably be
+> worth the words.
+
+> The abstract doesn't say much about the work. It would be better written
+> for experts to quickly gather an overview what the paper contributes.
+
+> Conversely, the main body of the paper skimps a bit on background. It
+> never explicitly covers what "blame" means and how it works in practice.
+> Similarly, it repeatedly talks about "impedance mismatches" without
+> defining them. Perhaps this phrase is now standard in the gradual typing
+> literature, which I haven't kept up with (hence my Z expertise). But I
+> did read the Wadler/Findler paper carefully at the time. It did not talk
+> about impedance mismatches. In any case, it is a fuzzy metaphor...
+> please say exactly what it means here.
 
@@
 > you did! It's really not clear at this point.
+
+> line 199: "blame set as another form of a stack trace" -- yes. I was hoping to
+> see a clearer example of debugging with and without blame, i.e. something
+> making explicit the similarities and differences between having blame info and
+> having only a stack trace at the error site.
 

### 2021-02-10 writing/reviews/pldi-rebuttal-high-level-first.md
touching up the first bullet

--- 2021-02-10 writing/reviews/pldi-rebuttal-high-level-first.md
@@
 
-We start off with a couple of common themes. Then we reply to all points
-individually.
+The first part of our response addresses a couple of common themes in the reviews; the second part addresses all remaining concerns individually.
 
-The conclusions from our experiment
+Are the empirical results conclusive or surprising?
 ===========================================================================
 
-There are three clear major conclusions from our experiment:
+Here are the two relevant major conclusions from our experiment:
 
-1) The Transient and Natural rational programmers are more effective than
-the Erasure one;
+1) Transient and Natural make rational programmers more effective than
+Erasure;
 
-2) Part of  the additional effectiveness of the former is due to runtime
-checks (that Erasure foregoes) but another part is due to blame;
+2) According to the rational programmer, the effectiveness of Natural is (merely)
+comparable to Transient.
 
-3) The effectiveness of the Natural and Transient rational programmers is
-comparable.
+As review `D` astutely points out, the second one is surprising: _While Theory predicts that Natural blame is more accurate than Transient blame, the experimental results do not support this prediction to a sufficient degree._
 
-The first two results confirm what the vast literature on gradual typing
-claims without systematic evidence: blame is useful. However, as review
-`D` astutely points out, the third conclusion is surprising in two
-different ways:
+Thus our results point to two possible explanations:
 
-(i) An important motivation for Transient is performance but our
-experiment provides evidence that performance issues, which stem from the
-design of Transient, negatively impact the effectiveness of blame.
+(i) a need for theoretical models with improved predictive power
 
-(ii) Theory claims that Natural blame is more accurate than Transient
-blame but our experiment provides evidence for the opposite;
-ll: TODO is "the opposite" the right phrasing? It seems more accurate to say evidence that doesn't support the prediction.
+The predictive power of gradual typing formal models fails when it comes to blame. In all likelihood, the current models are too small and do not include crucial
+features. (It is posisble that the current models are inherently limited.)
 
-Thus our results point to two directions for further research in gradual
-typing:
+(ii) a need for "blame data" from large systems
 
-(i) Implementors can experiment with checks and blame to find a sweet spot
-between type soundness, blame effectiveness and performance. Our method
+The presented data is based on systems with up to 15 modules, totalling 25,000 loc.
+While the chosen benchmarks represent a wide spectrum of realistic code, Natural may prove its worth only when the dependency chains are even deeper than the ones in the existing benchmarks.
+
+In both cases, our new method provides the tools for validating new theoretical results and/or gathering new data.
+
+Additionally, the experiment suggests that implementors can experiment with checks and blame to find a sweet spot between type soundness, blame effectiveness and performance. Our method
 provides an automatic tool to guide their exploration.
 
-(ii) The predictive power of gradual typing formal models (the reason
-d'etre of scientific models) fails when it comes to blame. This may be
-because the current models are too small and do not include crucial
-features or because the current models are inherently limited. In any case
-our results are a call to theoreticians to re-examine their assumptions
-about blame.  Our method provides an automatic tool to validate their
-models similar to the way the methods of experimental physicists help
-prove or disprove the  math of theoretical physicists.
-
-Challenges for our work compared to Lazarek's POPL paper
+Is the method novel? 
 ===========================================================================

### 2021-02-10 writing/reviews/pldi-rebuttal-high-level-first.md
more revisions

@@
 
-1) Transient and Natural make rational programmers more effective than
+1. Transient and Natural make rational programmers more effective than
 Erasure;
 
-2) According to the rational programmer, the effectiveness of Natural is (merely)
+2. According to the rational programmer, the effectiveness of Natural is (merely)
 comparable to Transient.
@@
 
-(i) a need for theoretical models with improved predictive power
+- a need for theoretical models with improved predictive power
 
@@
 
-(ii) a need for "blame data" from large systems
+- a need for "blame data" from large systems
 
@@
 
-Our work builds on the idea of Lazarek et al.'s paper that it is possible
-to evaluate the effectiveness of blame by simulating a debugging process.
-Both their work and our work is inspired by Greenman et al's work that
-uses a lattice of configurations to systematically explore variants of the
-same program with specifications of different precision (for performance
-reasons in Greenman et al).  Both the ideas of ``following`` blame while
-debugging and of gradually adding specifications to tweak the evaluation
-of the program originate from the practice of contracts (see the story in
-the intro of Lazarek's paper and similar anecdotes in Meyer's writings).
+Our inspiration is Lazarek et al.'s POPL paper, which shows the possibility of large-scale PL desiggn evaluations that do _not_ involve human sw devs.
 
-That said,
+The challenge is how to adapt this work to different problems than contracts.
+Given the close relationship between Natural and contracts, it is _natural_ to pursue an adaptation of the Lazarek method to gradual typing.
+Adaptation means retaining some ideas and injecting entirely new ones:
 
-1) In the context of Lazarek et al.'s work interesting bugs  are those
-detectable by behavioral contracts. Hence, they can use off-the-shelf
-mutators. In fact, their contracts do not detect the overwhelming majority
-of bugs except at the highest precision, where the contracts describe the full
-functional correctness of components.
+1.
+Lazarek et al.'s work can use off-the-shelf mutators because their programs are untyped.
+For typed prorgams, we must create new ones, for which we occasionally chose the same name as Lazarek --- a significant articulation mistake on our side.
+For the final version, we will use distinct names.
 
-2) Lazarek et al.'s method does not isolate the effect of blame from that
-of checks or mere luck. Thus they cannot really conclude that blame is the
-reason for the success of the debugging process --- only that blame
-shifting works. In other words, their work is lacking proper `controls'.
+2.
+Lazarek et al.'s method cannot really conclude that blame is the reason for the success of the debugging process because it fails to isolate the effect of blame from that of checks or mere luck.
+What they can assert is only that blame shifting works.
+Our work demands a proper `control', which we have developed.
 
-3) Lazarek et al. deal with a single contact system and examine a yes or
-no question: does blame-shifting find the bug for all debugging scenarios
-of a benchmark. They do not provide any guidance how to generalize to
-multiple systems with different notions of checks and/or blame in a
-uniform way that also allows comparisons.
+3.
+Lazarek et al.
+deal with a single contact system and examine a yes-or-no question, namely, whether  blame-shifting finds the bug for all debugging scenarios of a benchmark.
+Our experiment compares fundamentally different error-reporting schemes.
 
-This brings us to the three problems we had to solve in order to design
-our experiment:
+As to the specific complaints in reviews `c` and `d`:
 
-1) Natural and Transient under the same roof (section 3).
+* Natural and Transient under the same roof (section 3).
 
@@
 
-2) Custom mutators (section 4).
+* Custom mutators (section 4).
 
@@
 
-ll: TODO phrasing of "original set we tried" seems off
+* TODO phrasing of "original set we tried" seems off
 The sixteen mutators we include in the paper are not the original set we
@@
 
-3) Modes of the rational programmer (section 5).
+* Modes of the rational programmer (section 5).
 

### 2021-02-10 writing/reviews/pldi-rebuttal-high-level-first.md
rebuttal: compare half-typed lattice to library, add author names to a cite

--- 2021-02-10 writing/reviews/pldi-rebuttal-high-level-first.md
@@
 
-These are standard names for these three systems [9, 10, 37]. We didn't
-introduce them.
+We are reusing these names from Greenman et al. and Vitousek et al. [10, 37],
+but will explain more.
 
@@
 
-Yes, the approach could be applied.
-ll: worth a sentence saying that this would probably not be worthwhile to do?
-bg: TODO try explaining in 2-3 words
+Yes. In fact, we already do this for programs that depend on library code.
 

### 2021-02-11 writing/reviews/pldi-rebuttal-final.md
Combine top and bot into -final.md

--- 2021-02-10 writing/reviews/pldi-rebuttal-high-level-first.md
@@
+PLDI 2021 Paper #66 Author Response
+===========================================================================
+
+Thank you to all the reviewers for the helpful feedback.
+
+The first part of our response addresses two common themes in the
+reviews; the second part addresses all remaining concerns individually.
+
+Are the empirical results conclusive? Are they surprising?
+===========================================================================
+
+Here are the two relevant major conclusions from our experiment:
+
+1. Transient and Natural's blame makes rational programmers more effective than Erasure.
+
+2. The effectiveness of Natural is (merely) comparable to Transient.
+
+As review `D` points out, the second one is surprising: _While Theory
+predicts that Natural blames more accurately than Transient, our
+experimental results do not support this prediction._
+
+In analogy to physics research, this points to two possible explanations:
+
+- a need for theoretical models with improved predictive power
+
+In all likelihood, the current models are too small to include language features
+critical to blame assignment.
+
+- a need for "blame data" from large systems
+
+The presented data is based on systems with up to 14 modules (see table
+below).  Natural may prove its worth only when the dependency chains are
+deeper than the ones in the existing benchmarks. If Natural works better
+for programs with deeper dependency chains, this should guide future research
+on mixing approaches.
+
+Is the method novel? 
+===========================================================================
+
+Our inspiration is Lazarek et al.'s POPL paper, which shows the possibility
+of large-scale design evaluations that do _not_ involve human software devs.
+
+The challenge is how to adapt this work to problems other than
+contracts. Given the close relationship between Natural and contracts, it
+is "natural" to pursue an adaptation of the Lazarek method to gradual
+typing. Adaptation means retaining some ideas and injecting entirely new
+ones:
+
+1. Lazarek et al.'s work can use off-the-shelf mutators because they are only
+   looking for bugs that contracts can discover in an untyped world.  Our
+   work demands an entirely new judgment of which mutators are "good" or
+   "bad". This new judgment means cutting one mutator, changing four of them,
+   and adding nine. There was no other way to get an _interesting_ and
+   diverse body of bugs (see section 4).
+
+2. Lazarek et al. deal with a single contract system. Our experiment compares
+   fundamentally different checking and error-reporting schemata: the
+   industrial Erasure standard and the academic ones (Natural and Transient).
+
+3. Lazarek et al. examine a single yes-or-no question: whether blame-shifting
+   works. Our work asks (a) several yes-or-no questions and (b) asks
+   questions of degree; it yields statistically sound answers for the former
+   and evidence to shed light on the latter.
+
+4. Lazarek et al.'s method cannot really conclude that blame is the reason
+   for the success of the debugging process, because it fails to isolate the
+   effect of blame from that of checks or mere luck. It simply validates that
+   blame shifting works. Our work focuses on the value of blame and that kind
+   of work demands proper controls, which are entirely new.
+
+   Our work relies on two different baselines to understand the value of blame:
+
+   - _the "random" ("lottery") programmer_ ~~ An irrational programmer ignores
+     the error report and adds types to one randomly chosen component at a
+     time. Hence we use the "random mode" as a baseline to show that following
+     the error report is _not_ a question of luck.
+
+   - _the run-time checking of the underlying language_ ~~ All sound gradual
+     type systems employ run-time checks to guarantee the consistency of types,
+     but they do not have to track blame. If a run-time check fails, they can
+     raise an exception that reports a stack (trace).  Many programmers use
+     stack traces for debugging, but our work must show that blame assignment
+     is better than keeping checks and throwing out blame.
+
+5. One additional point from review `D` stands out, namely, the remark that
+   bringing Natural and Transient under the same roof (section 3) is "just
+   engineering."
+
+   Neither Reticulated Python nor the model of Transient provide answers
+   about what shallow checks correspond to the sophisticated types of Typed
+   Racket (or TypeScript), e.g.  polymorphism, row types for mixins,
+   occurrence typing) and what information is necessary to record in the
+   blame map.  But, we do not think that this paper is the proper place to
+   describe this aspect of the work; the large explanation would distract the
+   reader from the actual thesis of the paper.
+
 Replies to individual points
@@
 
-Please see our reply to the second part of the discussion at the beginning
-of the rebuttal.
+A bad mutator results in a type-level bug that is trivial to locate. For
+instance, a bug that leads to a type-level error during the evaluation of the
+component that contains it is trivial to locate. In contrast, a bug that is
+non-trivial to locate is one that causes the buggy component to evaluate without
+an error and to produce a value of the wrong type. An interesting bug is
+one where the value with the wrong type manages also to cross undetected to at
+least one other component before a type-level check flags it. Good mutators
+should be able to produce these interesting bugs.
 
@@
 
-No, the quote is not from [3]. We will rephrase to clarify that it is our
-statement of the research question.
+No, that text is not a quote from [3]. We will rephrase to clarify that it is
+our statement of the research question.
 
@@
 
-Sure. Each CPU has 28 doubly-threaded cores.
+Each CPU has 28 doubly-threaded cores. We will add it to the line.
 
@@
 limitations of the theory but we can't exclude that some of them aren't
-because of some unknown Typed Racket bug. Please see the first part of the
-discussion at the beginning of our response for more details and for our
-take on your last question.
+because of some unknown Typed Racket bug.
+
+We do not have any leads for a new blame strategy better than both
+Transient and Natural, but at least our method provides a way to examine
+ideas in that direction.
 
@@
 for each component rather than a single type. At the same time, our method
-  is parametric to the "size" of a component and adapts in a
-  straightforward manner to a setting where each definition can be
-  considered a component instead of a whole module. Of course both of
-  these adjustments lead to significantly larger scenario lattices than
-  the ones we investigate, so keeping the experiment computationally
-  feasible would require more aggressive sampling.
+is parametric to the "size" of a component and adapts in a straightforward
+manner to a setting where each definition can be considered a component
+instead of a whole module. Of course both of these adjustments lead to
+significantly larger scenario lattices than the ones we investigate, so
+keeping the experiment computationally feasible would require more
+aggressive sampling.
 
@@
 
-"Impedence mismatch" is our name for a run-time disagreement between a
+"Impedance mismatch" is our name for a run-time disagreement between a
 type annotation and an untyped value. The point of the name is to suggest
@@
 the random rational programmer discovers the buggy component on every step
-is $1/n$.  The missing piece is that we randomly sample from the set of
-scenarios -- and by extension the set of trails. So the shape of the
-random programmer's effort matches our sampling distribution. That said,
-as you observe, the distribution of the random programmer's effort is not
-relevant to our results, so we will remove the characterization.
+is $1/n$. The expected curve is not flat though, because we randomly
+sample from the set of scenarios --- and by extension the set of trails.
+That said, as you observe, the distribution of the random programmer's
+effort is not relevant to our results, so we will remove the
+characterization.
 
@@
 ===========================================================================
-
-> Paper summary
-> -------------
->
-> [....]
->
-> To evaluate this approach, the paper also presents empirical results. Starting
-> with a small corpus of programs, the authors use mutations to systematically
-> (but synthetically) introduce type errors -- e.g., one mutation replaces a
-> constants with another of a different type, and another mutation swaps method
-> identifiers. They then conduct experiments to quantify the number of steps
-> required by a rational programmer for several blame assignment schemes: Natural
-> (used in Typed Racket, which tracks at most one location for blame), Transient
-> (used in Reticulated Python, which tracks a set of locations), and Erasure
-> (which does not actually assign blame, and is widely used in industry).
-
-chrdimo: TODO
-effort is part 2/2 of the metrics
-
 > But while there is a lot to like about the paper, it also has some
@@
 
-If  an  "omniscient programmer"  is the programmer that always makes the
-best choice, i.e. finds the bug at the first try, then this corresponds to
-a trivial mode; It has 100% success rate and a constant blame trail length
+If an "omniscient programmer" is the programmer that always makes the best
+choice, i.e. finds the bug at the first try, then this corresponds to a
+trivial mode; it has 100% success rate and a constant blame trail length
 of 1. Thus, comparing its success to any other strategy amounts to
 counting the number of scenarios where the other strategy fails; comparing
-effort is similarly trivial. This information is in the paper implicitly,
-but we can certainly add an explicit mention if it would be useful.
+effort is similarly trivial. This information can be inferred from the
+definitions of section 5, but we can certainly add an explicit mention if
+it would be useful.
 
@@
 The quote from section 8 refers to the theory of blame rather than our
-results.  Please see our discussion of the first theme above for more
-details on this.
+results. Please see the first part of the discussion at the beginning of
+the response for more details on the conclusions of our results.
+
+That said, the results of the comparison between the three systems in
+section 7 is only one part of the contribution of the paper. The other
+part is the method itself (sections 3, 4, and 5).
 
@@
 its current design that we discovered exactly because of the experiment.
-
-> Most if not all of these shortcomings are discussed in the paper, which
-> is much appreciated. But they do risk undermine the value of the
-> contributions to some degree.
-
-We would like to point out that the results of the comparison between the
-three systems are only one part of the contribution of the paper. The
-other part is the method itself.
 
@@
 
-Our experiment does have a baseline. The random mode of the rational
-programmer serves as the "control" of our experiment. Furthermore the
-exception modes each serve as further "controls" to distinguish the effect
-of checks from that of blame within each semantics. Please also see our
-discussion of modes in the second theme at the beginning of our reply.
+Our experiment has several baselines. The random mode of the rational
+programmer described in section 5.5 serves as the "control" of the whole
+experiment. In addition, the exception modes described in sections 5.1 and
+5.2 each serve as further "controls" to distinguish the effect of checks
+from that of blame within each semantics. Please also see our discussion
+of baselines in the second theme at the beginning of our reply.
 
@@
 A side note here: Transient Typed Racket is faster than Natural Typed
-Racket when we turn off blame. Exactly as a sanity check that this is not
-an artifact of our implementation we have coded up benchmarks in Python
-and we confirmed that in Reticulated the performance of Transient with
-blame is even worse than what we have observed in Typed Racket.
+Racket when we turn off blame. As a sanity check that this is not an
+artifact of our implementation, we translated some benchmarks to Python
+and confirmed that in Reticulated the performance of Transient with blame
+is even worse than we have observed in Typed Racket.
 
-> *Natural and Transient Under the Same Roof (Section 3)*
->
-> This seems to be "only" an engineering challenge.
->
-> [...]
->
 > *Custom Mutators (Section 4)* The paper suggests that the mutators
@@
 > 2020 paper.  ...
->
-> [...]
->
+
+First of all, we need to clarify that the mutators used by Lazarek et al.
+are standard mutators from the the extensive literature on mutation
+testing. Naturally, we started with those to develop our mutators, but we
+quickly realized that only three of them work for our purposes. In
+hindsight, that is not surprising because they are designed to cause
+behavioral bugs rather than type-level ones. This realization led us to
+think carefully about interesting bugs in our setting and develop a new
+set of mutators based on Lazarek et al.'s that meets our needs. For
+example, the relational mutator from Lazarek et al. does not lead to
+type-level errors that Typed Racket's type system can detect, so we
+eliminated it. Similarly, the original arithmetic mutator does not
+reliably lead to type errors, so we adapted it to only make operator swaps
+that can affect the type of the result. For example, changing a `/` to `*`
+will not affect the type of an expression, but changing `*` to `/` may! In
+addition to those adjustments, we had to come up with entirely new
+mutators in order to get a sufficiently large and diverse set of bugs. For
+example, the new `-id` family of mutators are perfectly suited to creating
+type-level mistakes, and in fact most of our interesting bugs are
+generated by them. These mutators are completely different than swapping
+subexpressions; in particular, they are context-sensitive and nonlocal.
+Finally, we found mutators that we expected should do well but turned out
+to be ineffective at producing interesting bugs. For example, adding a new
+public method to a class reliably produces both static and dynamic type
+errors, but we found they were not interesting bugs.
+
+Besides all of these details about the mutators, we consider the biggest
+insight that allowed us to find the right set of mutators to be the
+definition of interesting bugs. Please see the high level comments at the
+start for more discussion of this.
+
 > *Debugging Strategies (Section 5)*
@@
 
-Please see our discussion of the challenges for our method at the
-beginning of our response.
+The key point of section 5 is the definition of the different blame
+strategies in a common precise framework as modes of the rational
+programmer. Modes enable the comparison between the different systems even
+when sampling is involved, thanks to the common substrate of blame trails.
+
+In addition, modes allow us to introduce in a uniform manner in the
+framework baselines that help us isolate confounding factors from the
+actual effectiveness of blame. This is exactly the role of the exception
+modes for Natural and Transient. Those two perform checks at runtime and
+the exception modes help us determine what part of the success of Natural
+and Transient is due to the extra checks or due to blame (and also whether
+blame masks useful information from exceptions). For example, we see that
+Natural without blame but with the same checks outperforms Erasure in ~25%
+of the scenarios (figure 6). Blame adds another ~11% on top of that. For
+Transient there is a small percentage of scenarios where exceptions do
+better than blame but still blame improves over exceptions by ~11%
+compared to Erasure.
 
@@
 > more explicit. As discussed above, there seem to be more similarities
-> than suggested.  L92: "use of _a_ higher-order contract system" L155:
-> "authors extensive" L253: "exports _it_ as" L280: Missing `untyped-`
-> L314: " . " L382: "Dyn," Fig 2: Presumably authors were included to
-> indicate the variety of sources, but I'm not sure that was necessary.
-> L597: "make us_e_ of" L664: "described in _Section_ 4" L980: "Figure
-> 5.4" ==> "Figure 6" L986: "2.8%" ==> "2.18%" ?  L1017: "33%" ==> "26%" ?
+> than suggested.
+> L92: "use of _a_ higher-order contract system"
+> L155: "authors extensive"
+> L253: "exports _it_ as"
+> L280: Missing `untyped-`
+> L314: " . "
+> L382: "Dyn,"
+> Fig 2: Presumably authors were included to indicate the variety of
+> sources, but I'm not sure that was necessary.
+> L597: "make us_e_ of"
+> L664: "described in _Section_ 4"
+> L980: "Figure 5.4" ==> "Figure 6"
+> L986: "2.8%" ==> "2.18%" ?
+> L1017: "33%" ==> "26%" ?
 > L1304: "Problem is ...."
@@
 
-TODO chrdimo
-- see general comments for how to interpret results, their significance, and how they challenge existing perceptions
-- usefulness asks if one succeeds when another failed; no extent for this
-- if both succeed, can quantify difference, use trail length to do so
-- trail length is proxy, paper discusses limitations, allows automation ... no humans + large scale
+Please see the discussion at the beginning of our response about the
+interpretation of our results, their significance, and how they challenge
+existing perceptions about blame.
+
+We are not sure what "extent" means in this comment. Usefulness asks
+whether one system succeeds in a debugging scenario while another system
+fails. If both succeed we use the length of the trail to quantify the
+difference between the two systems. Trail length may not be the most
+optimal proxy for programmer effort, but it also allows us to perform a
+large scale automatic experiment about blame without the practical
+limitations of user studies. We will add a discussion of this to the
+paper.
 
@@
 
-TODO chrdimo
-....
+We are not sure what the question is here. If the question is whether our
+implementation of the experiment is faithful to the definitions from
+section 5 and the discussion about statistics from section 6, then the
+answer is emphatically yes. We have validated manually a number of blame
+trails for each mode and moreover our implementation of the experiment
+comes with plenty of sanity checks.
 
-if you are asking whether our results
+If the question is whether our methods acts as a correct classifier for
+the detection of bugs a la machine learning, then we would like to stress
+that this is not what our work is about. Our work defines the rational
+programmer, an ideal programmer, as a deterministic algorithmic procedure
+and then uses the results of this procedure on a large number of debugging
+scenarios to determine if blame is a critical factor for the success or
+failure of the rational programmer.
 
@@
 
-We do think there is a misunderstanding here about what gradual typing is
-in general.
+We think there might be a misunderstanding here about gradual typing:
 
-1) Erasure is the established name of the semantics implemented by
-TypeScript, Flow etc. These languages come with an equally sophisticated
-type system as Typed Racket. Programmers add annotations to their program
-and the type checker does its best to type check the program in order to
-discover mistakes and help the IDE. Then types are erased and the program
-is run in the underlying dynamic language. That language's primitive
-operations have runtime checks that may detect an incompatible argument
-and raise an exception.
+1. Erasure is the established name of the standard semantics implemented
+   by industrial languages such as TypeScript, Flow etc. These languages
+   come with a type system just as sophisticated as Typed Racket.
+   Programmers add annotations to their program and the type checker does
+   its best to type check it in order to discover mistakes and help the
+   IDE. Then types are erased and the program is run in the underlying
+   dynamically typed language. That language's primitive operations have
+   runtime checks that may detect an incompatible argument and raise an
+   exception.
 
-2) The other two semantics introduce extra run time checks (in the form of
-proxies for Natural and inlined checks for Transient) to enforce that
-values untyped code provides to typed code have the expected type. when
-one of these checks fails they also issue blame (with different
-mechanisms).
+2. The two academic semantics introduce extra run time checks (in the form
+   of proxies for Natural and inlined checks for Transient, as section 3
+   describes). The checks enforce that values the untyped code provides to
+   typed code behave according to their expected type. When one of these
+   checks fails both Natural and Transient issue blame (though each with
+   different mechanisms).
 
-3) In Typed Racket programmers cannot add contracts. The precision of the
-checks is determined by the types and matches those. In turn the types of
-primitive operations match the checks that their Racket counterparts
-perform. Finally our benchmarks come with the most precise types that
-allow them to type check (where precision is based on subtyping).
+3. In Typed Racket, programmers cannot add contracts. The precision of the
+   checks matches the type annotations of the program.
 
-4) The Natural and Transient exception modes perform the same checks as
+4. The Natural and Transient exception modes perform the same checks as
    Natural and Transient correspondingly. The only difference is that they do
@@
    the failed check. We use these modes to separate the effectiveness of
-blame from that of the checks.
-
-chrdimo: I am not sure if I understand what this guy is trying to say
-here. It seems a bit like rambling to me sprinkled with bold statements.
+   blame from that of the checks. Please also see the second part of the
+   discussion at the beginning of our response.
 
@@
 
-For point 1 and 3 please see our discusssion at the beginning of the
+For points 1 and 3 please see our discussion at the beginning of the
 reply. For point 2, the idea of the rational programmer is a new one so
-far applied to blame in contracts (Lazarek et al. POPL 2020) and blame in
-gradual typing (our work). We don't know the future but we believe it can
-have a similar effect in programming languages as homo economicus in
-economics.
+far applied only to blame in contracts (Lazarek et al. POPL 2020) and blame in
+gradual typing (our work). We do believe it can have a similar effect in
+programming languages as homo economicus in economics.
 
-That said, as posed, the comment expresses a significant
+That said, as posed, this comment seems to express a significant
 mis-characterization of the submission. It is not the benefits of a
@@
 type system constant but allows to answer the question which of several
-run-time checking regimes (for enforcing types) provides the best
+checking and blaming regimes (for enforcing types) provides the best
 explanatory messages in case of violations. Hence the analogous question
 for purely static type systems would ask which of several reporting
-  schemes provides the best explanatory message for type errors.  In the
-  case of simply typed languages and even languages with local type
-  inference, this question is basically meaningless. In the case of
-  languages with HM type inference, the question has been implicitly
-  raised for four decades with the development of alternative ways of
-  finding the source of inference conflicts.
+schemes provides the best explanatory message for type errors. In the case
+of simply typed languages and even languages with local type inference,
+this question is basically meaningless. In the case of languages with HM
+type inference, the question has been implicitly raised for four decades
+with the development of alternative ways of finding the source of
+inference conflicts. One recent example is SHErrLoc (Zhang et al. TOPLAS
+2017).
 
-If the reviewer is indeed interested in the question of whether type
-systems help programmers---a question that this submission does _not_
-ask---the recent OOPSLA literature contains several user studies. A simple
-Google query will suggest a short list of these publications.
+If the reviewer is indeed interested in the question of whether type systems
+help programmers --- a question that this submission does _not_ ask --- then
+Hanenberg and Stefik's papers are a good starting point (e.g. ICSE 2014).
 
@@
 In Erasure, a program can type check and still have type-level errors
-because some type annotations may be missing. In contrast to Natural and
-Transient, types do not result in run time checks so the only way to
-detect type-level errors that the type checker couldn't find is the
+because some type annotations may be missing. Section 3 illustrates
+exactly this scenario by example. In contrast to Natural and Transient,
+types in Erasure do not result in run time checks so the only way to
+detect type-level errors that the type checker could not find is the
 built-in checks of the language's primitives.
@@
 
-Thanks we will add a more detailed comparison along the lines of the
+Thanks, we will add a more detailed comparison along the lines of the
 discussion at the beginning of our response.
@@
 
-The following sentences are meant to explain this, so we will adjust the
-prose to make that clear.
+The sentences that follow line 70 are meant to expand on the slogan. We
+will adjust the prose to make it clear.
 
@@
 
-It means debugging scenario. We will clarify.
+It means debugging scenario. We will clarify the prose.
 
@@
 
-In this context a component is a module.
-
-> line 199: "blame set as another form of a stack trace" -- yes. I was
-> hoping to see a clearer example of debugging with and without blame,
-> i.e. something making explicit the similarities and differences between
-> having blame info and having only a stack trace at the error site.
+In the context of Typed Racket, a component is a module.
 
@@
 
-Yes in this context. Blame refers to the unit of blame which in Typed
-Racket is a module.
+Yes in this context. Party and component refer to the entities that can be
+blamed, which in Typed Racket is modules.
 
@@
 
-Yes, communicating a value via state also constitutes a boundary crossing.
-We will clarify what a boundary crossing is in the prose.
+Yes, exchange of values via state also constitutes a boundary crossing.
+Some of our benchmarks do exactly that. We will clarify what a boundary
+crossing is in the prose.
 
@@
 
-Consider a function with a type annotation that says its a function from
-integers to string that formats its argument to a string and returns it.
-If we import the function in an untyped module under Erasure there are no
-checks to make sure that the function is going to be applied to integers.
-So we can apply it to character or a string and still get back a result
-except that it is not the string representation of a number.
-
-> line 380: from this I inferred that "migratory typing" means "gradual
-> typing applied at modulewise granularity". Assuming that's correct, it's
-> worth saying directly.
-
-We will clarify. Migratory typing is not about granularity, but whether the
-type system forces rewrites upon untyped code. A gradual type system for Python
-would be migratory. A new gradual-from-the-start language would not be.
+No. A buggy program can produce a value in Erasure where in Natural or
+Transient it produces a run time type error. Greenman et al. [10] prove
+formally this proposition but it is easy to construct an example that
+demonstrates the issue. For instance if the client in Figure 1 just
+formats the result of 'crypto-pack' to a string and returns it, then the
+client will terminate without an error but its string result will be
+wrong; it will be a string representation of a hash rather than a list.
+The reason is that the `format` primitive of Racket does not care what its
+argument is.
 
@@
 
-Yes please see our answer to a relevant comment above.
+Yes please see our answers to your questions about Erasure, line 34 and
+line 313.
 
@@
 
-Indeed, Transient checks only check the top-level type constructor of a
-value.  We will add a reminder of this fact to the prose here.
+Indeed, Transient checks only the top-level type constructor of a
+value. We will add a reminder of section 3's discussion of Transient
+checks to the prose here.
 
@@
 
+> line 380: from this I inferred that "migratory typing" means "gradual
+> typing applied at modulewise granularity". Assuming that's correct, it's
+> worth saying directly.
+
 > line 421: "type mistake" -- does this mean "feasible run-time type

### 2021-02-12 writing/challenges.tex
Minor rewordings in intro and challenges
touching up the prose

@@
 
-The most serious challenge for evaluating blame concerns 
-the nature of this research. Unlike ordinary
-work on programming languages, the question seems to call for empirical research
-similar to the one in the human-computer interaction area. At the same time, a
-significant evaluation result demands a large number of cases yielding a serious
-amount of data. But, as~\citet{lksfd-popl-2020} already demonstrated, the way
-around this dilemma is to postulate the equivalent of a {\it homo economicus\/}
-for programming languages---the {\em rational programmer\/}---and to implement
-this rational programmer as an algorithm.  As indicated in the introduction, in
-the context of gradual typing, the rational programmer translates the
+%% -----------------------------------------------------------------------------
+
+The challenge of evaluating blame is a methodological one. Unlike most research
+on programming languages, the question seems to call for empirical research
+similar to that of human-computer interaction. At the same time, a significant
+evaluation result demands a large number of cases yielding a serious amount of
+data. As~\citet{lksfd-popl-2020} recently demonstrated, the way around this
+dilemma is to postulate the equivalent of a {\it homo economicus\/} for
+programming languages---the {\em rational programmer\/}---and to implement this
+rational programmer as an algorithm.
+
+In the context of gradual typing, the rational programmer translates the
 Wadler-Findler slogan into a debugging method, searching for the source of the
 impedance mismatch in an incremental fashion. Measuring this simulated
-programmer's behavior on a large number of inputs yields relevant 
-data, similar to collecting empirical data in a human-focused discipline.
+programmer's behavior on a large number of debugging sessions yields data that
+is similar to empirical data collected in a human-focused discipline.
 
-A derivative challenge is the problem of simulating the programmer's
-mistakes or, more precisely, the origin of impedance mismatches. Before a
-programmer can try to find mistakes, mistakes have to be made.
-Specifically, a necessary step to test the rational programmer is to find
-a collection of  type mistakes in  mix-typed programs that is highly
-representantive of both mistakes and programs ``in the wild.''
-Unfortunately no such collection exists.  There is a good reason for that;
-the kind of mistakes the rational programmer targets are easily detected
-by unit tests and, even if their sources are hard to find, these mistakes do not make it
-into code repositories with appropriate commit messages.  In other words, the problem is how to generate a
-sufficiently large and representantive corpus of type mistakes for a meaningful
-experiment.  Following~\citet{lksfd-popl-2020} again, the solution is to
-use mutation analysis~\cite{lipton1971fault, demillo1978hints,
-jia2011analysis} to create mistakes in a synthetic manner. This process
-yields a large number of buggy programs whose mixed-typed versions are
-suitable subjects for our experiment.  The mutation analysis of Lazarek et
-al, is, however, useless for creating impedance mismatches.
+A derivative challenge is the problem of simulating the impedance mismatches
+between typed and untyped pieces of code that programmers may accidentally create
+and that the run-time checks of academic systems catch. In other words, the
+experiment calls for a collection of type mistakes in mix-typed programs that is
+representative of those ``in the wild.'' Unfortunately no such collection exists,
+and with good reason; the kind of mistakes the rational programmer targets are
+easily detected by unit tests and, even if their sources take some time to find,
+these mistakes do not make it into code repositories with appropriate commit
+messages.  Hence the experimenter must generate a sufficiently large and
+representative corpus of type mistakes for a suitable collection of benchmark
+programs.  Following~\citet{lksfd-popl-2020} again, the solution is to synthesize
+mistakes using mutation analysis~\cite{lipton1971fault, demillo1978hints,
+jia2011analysis}.  The mutation analysis of Lazarek et al, is, however, useless
+for creating impedance mismatches.  While mutation analysis traditionally aims to
+inject bugs that challenge test suites, it discards those that yield ill-typed
+mutants as \emph{incompetent}.  Indeed, mutation analysis frameworks are
+fine-tuned to avoid them---and yet it is precisely those that are needed for
+evaluating blame assignment strategies.
 
-While mutation analysis traditionally aims to inject bugs that challenge
-extensive test suites, the ones yielding ill-typed mutants are usually deemed
-\emph{incompetent}. As a matter of fact, mutation analysis frameworks are
-fine-tuned to avoid them---yet it is precisely those that are needed for
-evaluating blame assignment strategies. The mutators used here are newly designed,
-inspired by the authors extensive experience with programming in Typed Racket.
-They cause type errors that cover a wide range of Typed Racket's type system and moreover
-reliably cause runtime exceptions in mixed-typed versions of mutants.
+Thus, an experimental analysis of blame needs a mostly new set of mutators.
+Roughly speaking, the new mutators inject type errors into fully typed programs.
+Applying such a mutator to any typed component produces a mutatnt component.  A
+debugging scenario results from removing the types from the mutatnt. For the
+design of such mutators, the authors relied on their own extensive programming
+experience though not without discovering a major pitfall: some of their original
+mutators systematically produced programs that immediately revealed the source of
+the impedance mismatch. About two thirds of the original mutators survived this
+first scrutiny testing. All of the remaining ones yield interesting debugging
+sessions.
 
-The second big challenge is the disparity of gradually
-typed languages. While the approaches are well understood in theory,
-they come with vastly different type systems, accommodate seemingly
-incompatible language features, and apply to radically different coding
-styles in the untyped world.  To eliminate the effect of these differences 
-on its results, the application of an evaluation method to different
-gradual typing systems must use the same test programs and the same type
-system. Otherwise, results for different gradual typing systems 
-do not share a common baseline and, therefore, do not lend support to any
-conclusions about whether the value of blame changes from one system to
-another. For instance, the case study presented here builds on
-Greenman and Felleisen's Transient variant for functional Typed
-Racket~\cite{gf-icfp-2018} to obtain a Transient semantics that is
-comparable with Typed Racket's default Natural one. This Transient
-variant accommodates varied linguistic features and library calls
-in a way that is faithful to the Transient
+A somewhat coincidental challenge is the disparity of gradually typed
+languages. While the approaches are well understood in theory, they come with
+vastly different type systems, accommodate seemingly incompatible language
+features, and apply to radically different coding styles in the untyped world.
+To eliminate the effect of these differences on its results, the application of
+an evaluation method to different gradual typing systems must use the same test
+programs and the same type system. Otherwise, results for different gradual
+typing systems do not share a common baseline and, therefore, do not lend support
+to any conclusions about whether the value of blame changes from one system to
+another. For instance, the case study presented here builds on Greenman and
+Felleisen's Transient variant for functional Typed Racket~\cite{gf-icfp-2018} to
+obtain a Transient semantics that is comparable with Typed Racket's default
+Natural one. This Transient variant accommodates varied linguistic features and
+library calls in a way that is faithful to the Transient
 semantics~\cite{vss-popl-2017}.
 
-Beyond the implementation issue, the method also faces the problem that
-the three variants---the Natural semantics, the Transient semantics, and
-the Erasure semantics---come with three radically different blame
-assignment strategies.  For example, the Natural semantics assigns blame
-to one component while Transient assigns blame to a set of components. The
-Erasure semantics does not blame a component per se, but it comes with an
-exception location and the usual stack traces.
+Beyond the sheer implementation issue, the development of a method must also
+address the problem that the three semantic variants---Natural, Transient, and
+Erasure---come with three different blame assignment strategies.  While Natural
+assigns blame to {\em one\/} component, Transient assigns blame to a sequence of
+components. The Erasure semantics does not blame a component per se, but it comes
+with an exception location and a stack trace.
 
@@
 
-The final big challenge is that the different gradual type
-systems, the number of necessary modes for the rational programmer
-and the large number of mutants lead to an explosion of scenarios the
-method needs to examine. Thus a practical application of the
-method has no option but to carefully sample the space of scenarios while 
-making sure that it still produces reproducible conclusions.
+The final big challenge is the explosion of debugging scenarios that results from
+the combination of mutation-based scenario generation and mode-based analysis.
+Specifically, the experiment demands a large amount of computational resources
+due to the multiplicative effect of mixing the different gradual type systems,
+the number of debugging modes for the rational programmer, and the large number
+of mutants.  A practical application of the method has no option but to sample
+the space of scenarios, carefully ensuring that it still produces reproducible
+conclusions.
 
@@
 
-These challenges expose the subtlety of evaluating blame strategies but
-also suggest a recipe:
+Overcoming these challenges needs a systematic approach:
 
@@
 
-\item Implement all three gradual type systems on a common
-  basis (section~\ref{sec:landscape});
+\item a single-language implementation of all three gradual type systems (section~\ref{sec:landscape});
 
-\item Generate a large set of programs with interesting
-  type-level errors (section~\ref{sec:mutate});
+\item a suitable set of mutators for the generation of a large set of debugging
+  scenarios with interesting type-level errors (section~\ref{sec:mutate});
 
-\item Simulate the rational programmer on these programs with sufficient 
-  modes to make meaningful comparisons between modes of the same or
-    different systems
-    (section~\ref{sec:rational});
+\item an algorithmic simulation of the rational programmer on these debugging
+  scenarios with sufficient modes to make meaningful comparisons between modes of
+  the same or different systems (section~\ref{sec:rational});
 
-\item Sample the experimental space  to obtain a computationally feasible
-  but statistically sound experiment (section~\ref{sec:sample}).
+\item a careful sampling of the experimental space to obtain a computationally
+  feasible and statistically sound experiment (section~\ref{sec:sample}).
 

### 2021-02-12 writing/landscape.tex
touching up the prose

@@
 of the responsible party for each check in the corresponding proxy. Thus,
-in our example, when the call to \texttt{typed-pack} returns, the return
+in our example, as the call to \texttt{typed-pack} returns, the return
 type check fails and the Natural semantics blames \texttt{pack-lib}.
@@
 forms in the example. Since the failed check corresponds to the return
-type of \texttt{pack}, Transient blames the sources of the two boundary
-crosses, which in this case is \texttt{untyped/pack-lib} for both. In
-general though, Transient may blame more than one party. In fact, the
-theoretical work of~\citet{gfd-oopsla-2019} shows that there are programs
-where Transient constructs a blame set that includes modules that are not
-relevant to the check failure or that it even misses entirely the responsible
-party ---  our experiment aims to examine the pragmatic impact of exactly these
-issues on debugging type-level mistakes.
+type of \texttt{pack}, Transient blames the sources of the two
+boundary crosses, which in this case is \texttt{untyped/pack-lib} for
+both. In general though, Transient may blame more than one party. In
+fact, the theoretical work of~\citet{gfd-oopsla-2019} shows that there
+are programs where Transient constructs a blame set that includes
+modules that are not relevant to the check failure or that it even
+entirely misses the responsible party--our experiment aims to examine
+the pragmatic impact of exactly these issues on debugging type-level
+mistakes.
 
-Under the Erasure semantics, the example does not raise an error due to the
-gradual type system; Erasure performs no type checks at runtime. Thus the
-program may fail with an exception if \texttt{client} tries to use the
-result of \texttt{typed-pack} or \texttt{crypto-pack}, but it may also not
-fail at all and produce a wrong result that goes undetected by the
-gradual type system .  In either case, the Erasure semantics provides no
-help to the programmer for detecting the bug in our program.
+Under the Erasure semantics, the example does not raise an error due
+to the gradual type system; Erasure performs no type checks at
+runtime. Thus the program may fail with an exception if
+\texttt{client} tries to use the result of \texttt{typed-pack} or
+\texttt{crypto-pack}, but it may also not fail at all and produce a
+result of the proper kind---and yet is completely wrong.  In either
+case, the Erasure semantics provides no help to the programmer for
+detecting the bug in our program.
 
@@
 
-The first major ingredient of Transient Typed Racket is to determine the
+The first major ingredient to the construction of Transient Typed Racket is to determine the
 Racket implementation of the shallow checks that map to each type.  For
@@
 for elimination forms also extend the blame map with further information. For
-instance, the check of the result of {\tt (f x)} adds to the blame map
-that the expected type of the result is the return type of the type of
+instance, the check of the result of {\tt (f x)} adds an entry to the blame map
+that records the expected type of the result as the return type of the type of
 {\tt f}.  If a Transient check ever fails, Transient Typed Racket walks over 
@@
 support for type \texttt{Dyn}, components are either untyped or fully
-typed and the gradual type system does not affect the compilation of
-untyped components. Despite, these differences, Typed Racket is the most
+typed.
+%% MF: does it ever:
+%% and the gradual type system does not affect the compilation of untyped components. 
+Despite, these differences, Typed Racket is the most
 mature system for sound mix-typed code

### 2021-02-12 writing/mutate.tex
Introduce transient in intro, reword "mere typos"
touching up the prose

@@
 
-An important challenge from section~\ref{sec:challenges} is the generation of
-mix-typed programs with representative impedance mismatches between
-the typed components, their types, and the behavior of untyped components. In this section, we
+%% this is awkward, but what I removed was exponentially awkward
+
+In this section, we
 present a generation process and argue that, despite its synthetic
@@
 analysis~\cite{jia2011analysis,demillo1978hints,lipton1971fault}.  A
-mutation is a local syntactic change of the code of a program that may
+mutation is a local syntactic change of the code that may
 produce a bug. Usually, the outcome of mutating a program once, dubbed a
@@
 
-However,  unlike the hard-to-kill mistakes introduced by standard mutators, 
-gradual type systems flag mere typos.  Thus, in order to
+Unlike the hard-to-kill mistakes introduced by standard mutators, 
+gradual type systems catch shallow errors.  Thus, in order to
 exercise the rational programmer in the context of gradual typing, 
@@
 
-The astute reader must have observed that mutators that produce 
+The reader may have observed that mutators that produce 
 type-level mistakes do not necessarily turn our benchmarks into
@@
 compile time.  Hence, we apply the mutators to versions, dubbed
-\dubbed{scenarios}, of the benchmarks without 
+dubbed{scenarios}, of the benchmarks without 
 some  of their type annotations.  Without the annotations, the type
@@
 The first criterion simply validates that the debugging scenario concerns
-an actual type-level mistake. However, The second and third criteria require 
-more explanation.  The second one stipulates that the mistake is
+an actual type-level mistake. The second one stipulates that the mistake is
 detectable under Erasure to ensure that it is detectable by all three
@@
 opt for tipping the scales in favor of the theoretically weakest
-semantics. We discuss further the implications of this choice in
+semantics. We further discuss the implications of this choice in
 section~\ref{sec:discussion}.

### 2021-02-21 writing/introduction.tex
strengthen introduction to results

--- 2020-11-21 writing/introduction.tex
@@
 
-Theoreticians of gradual typing have focused on
-blame theorems from the very beginning~\cite{mf-toplas-2009, tf-dls-2006}. ``Well-typed
+Theoreticians of gradual typing have focused on blame theorems from the very
+beginning~\cite{mf-toplas-2009, tf-dls-2006}. ``Well-typed
 [components]\footnote{The original authors got this word wrong.} can't be
-blamed'' turned the theorem into a slogan~\cite{wf-esop-2009}. Indeed,
-academic implementations---for example, Reticulated
-Python~\cite{vsc-dls-2019, vss-popl-2017, vksb-dls-2014} and Typed
-Racket~\cite{tf-dls-2006,tf-popl-2008,tfffgksst-snapl-2017,tf-icfp-2010}---come
-with sophisticated blame assignment strategies. The academics that
-created them firmly embrace the idea that blame can help practicing
-programmers find impedance mismatches between the types they added to a
-software system and the behavior of the remaining untyped components.
+blamed'' turned the theorem into a slogan~\cite{wf-esop-2009}. Academic systems
+(Reticulated Python~\cite{vsc-dls-2019, vss-popl-2017, vksb-dls-2014} and Typed
+Racket~\cite{tf-dls-2006,tf-popl-2008,tfffgksst-snapl-2017,tf-icfp-2010}) come
+with sophisticated blame assignment strategies. Their academic creators embrace
+the idea that blame can help practicing programmers find impedance mismatches
+between the types they added to a software system and the behavior of the
+remaining untyped components.
 
-Industrial implementors of  gradual typing systems have almost completely ignored blame assignment.
-Systems such as Flow,\footnote{\url{https://flow.org}}
+Industrial implementors of gradual typing systems have almost completely ignored
+blame assignment.  Systems such as Flow,\footnote{\url{https://flow.org}}
 Hack,\footnote{\url{https://hacklang.org}} or
-TypeScript\footnote{\url{https://www.typescriptlang.org}} exploit types for IDE actions and for finding typos in
-code. Then their compilers remove types and rely on the built-in safety checks of
-the underlying language to catch any problems. 
-Recently, \citet{cc-snapl-19} have turned this implicitly expressed
-skepticism into an explicit statement and challenge.
+TypeScript\footnote{\url{https://www.typescriptlang.org}} exploit types for IDE
+actions and for finding typos in code. Then their compilers remove types and
+rely on the built-in safety checks of the underlying language to catch any
+problems.
 
@@
 \end{quote}
-Given the long standing academic interest in blame  and its complete absence in
+Given the long-standing academic interest in blame and its complete absence in
 industrial systems, it comes as an even bigger surprise that the research
-literature and the industrial blog world do not present or discuss any
-possible answers.  Instead, when language designers decide to look into the
-question, they seem to answer it one way or another without any scientific
-justification. As a matter of fact, the research community simply does not have a
-method for evaluating blame, and given the question's practical
-relevance, this is a significant problem.
+literature and the industrial blog world do not discuss any possible answers.
+Instead, when language designers make relevant decisions, they seem to answer it
+one way or another without any scientific justification. Then again, the
+community has thus far failed to offer a method for evaluating blame assignment.
 
-This paper's first contribution is an automated method for evaluating the
-effectiveness of blame assignment strategies in the gradual typing world.
-%% (sections ~\ref{sec:landscape},~\ref{sec:mutate},~\ref{sec:rational} and \ref{sec:sample}). 
-%% MF: says you can't simulate a roadmap by listing all sections.
-%% MF: and for heaven's sake lookup the use of ~ in TeX
-Roughly speaking, the method, which is
-loosely based on the work of Lazarek et al.~\cite{lksfd-popl-2020},
-simulates a programmer who follows the Wadler-Findler slogan quoted above.
-Such a \emph{rational programmer} can find impedance-mismatch
-bugs by adding types to a partially typed program (section~\ref{sec:landscape}). That is, when the
-run-time system signals an impedance mismatch, a good blame message helps
-the programmer infer where to add an additional type specification.
-Components that pass the type checker cannot cause impedance mismatches,
-so if the type checker discovers these mismatches at compile time then it points
-directly to the faulty component. Otherwise, if the
-additional type does {\em not\/} cause a static type error, the program
-is run again to signal another violation. Then the programmer can heed
-blame one more time to add yet another type specification.
-Weighing whether we discover the mistake in this manner, and the number of
-steps it takes to do so, are measures of the effectiveness
-of the blame assignment strategy of the language.
+This paper's first contribution is {\em an automated method for evaluating the
+effectiveness of blame assignment strategies\/} in the gradual typing world.
+speaking, the method simulates a programmer who follows the Wadler-Findler
+slogan quoted above; its automation borrows elements of of Lazarek et
+al.~\cite{lksfd-popl-2020}'s recent work.\footnote{\citet{cc-snapl-19} use this
+tracing method to localize type errors reported by gradual typing systems based
+on inference.}  Such a \emph{rational programmer} can find impedance-mismatch
+bugs by adding types to a partially typed program (see
+section~\ref{sec:landscape}). That is, when the run-time system signals an
+impedance mismatch, a good blame message helps the programmer infer where to add
+an additional type specification because, by the basic theory, well-typed
+components cannot cause impedance mismatches. Otherwise, if the additional type
+does {\em not\/} cause a static type error, the program is run again to signal
+another violation. Then the programmer can heed blame one more time to add yet
+another type specification. Evaluating the quality of blame assignment thus
+becomes a question of whether the tracing the error messages finds the source of
+the mistake and how many steps it takes to get to this point.
 
-The paper's second contribution is a set of results from applying the evaluation
-method to three distinct blame assignment strategies and approximately
-722,000 cases (section~\ref{sec:results}): Vitousek et
-al.'s strategy of tracking typed/untyped boundaries in Reticulated, called Transient;
-Tobin-Hochstadt et al.'s use of higher-order contract system~\cite{ff-icfp-2002,
-mf-toplas-2009}; and the approach of industrial systems that forego blame
-and instead rely on error messages from the
-underlying programming language. A direct interpretation of the results
-(section~\ref{sec:discussion}) is that
-they validate the conjecture behind the work of theoreticians. A good blame assignment
-strategy helps the search for impedance mismatches between the specified types
-and the behavior of untyped components. Even better, the evaluation method 
-identifies the wrapper-based blame tracking of Typed Racket as the most
-useful academic system with that of Reticulated Python only marginally  behind.
+The paper's second contribution is a set of {\em results from applying the
+evaluation method to three distinct blame assignment strategies and
+approximately 72,200 cases\/} (see section~\ref{sec:results}): (1) {\it
+Transient\/}, i.e. Reticulated's tracking of typed/untyped boundaries; (2) {\it
+Natural\/}, i.e. the use of higher-order contract system and its blame
+assignment~\cite{ff-icfp-2002}; and (3) {\it Erasure\/}, i.e. the approach of
+industrial systems, which forego blame assignment and rely on error messages
+from the safety checks of the underlying programming language.
 
-Next, the paper describes the challenges of developing an evaluation
-method, which may explain why the question has been neglected for such
-a long time.
+The results of the method's first application (see section~\ref{sec:discussion})
+are at least somewhat surprising. In principle they validate the conjectures
+behind the work of theoreticians. First, a good blame assignment strategy helps
+with the search for impedance mismatches between the specified types and the
+behavior of untyped components.  Second, Natural's wrapper-based blame tracking
+is more useful than Transient's ``collective blame'' tracking algorithm, which
+in turn is superior to Erasure. {\em But\/}, the application of the method also
+indicates problems with the expectations of theoreticians. The first problem is
+that the cost of Transient's blame can be huge. The second one concerns the
+difference between the methods; neither is {\it Natural\/} vastly better than
+{\it Transient\/} nor are the two blame assignment methods clearly preferable to
+{\it Erasure\/}. The third one casts doubt on the current the trade-off between
+the cost of blame-assignment systems and the quality of their blame messages.
+In turn, these problems suggests that, on one hand, the existing theory does not
+predict practice properly, and on the other hand, the existing practice needs to
+be supported by additional data.

### 2021-02-21 writing/challenges.tex
describe challenges systematically, factor out the implementation challenge

@@
 
-The challenge of evaluating blame is a methodological one. Unlike most research
-on programming languages, the question seems to call for empirical research
-similar to that of human-computer interaction. At the same time, a significant
-evaluation result demands a large number of cases yielding a serious amount of
-data. As~\citet{lksfd-popl-2020} recently demonstrated, the way around this
-dilemma is to postulate the equivalent of a {\it homo economicus\/} for
-programming languages---the {\em rational programmer\/}---and to implement this
-rational programmer as an algorithm.
+The general challenge of evaluating blame is a methodological one. Unlike most
+current research on programming languages, the question seems to call for
+empirical research similar to that of human-computer interaction. At the same
+time, a significant evaluation result demands a large number of cases yielding a
+serious amount of data. As~\citet{lksfd-popl-2020} recently demonstrated, the
+way around this dilemma is to postulate the equivalent of a {\it homo
+economicus\/} for programming languages---the {\em rational programmer\/}---and
+to implement this rational programmer as an algorithm.
 
@@
 programmer's behavior on a large number of debugging sessions yields data that
-is similar to empirical data collected in a human-focused discipline.
+is similar to data collected in an empirical, human-facing manner.
 
-A derivative challenge is the problem of simulating the impedance mismatches
-between typed and untyped pieces of code that programmers may accidentally create
-and that the run-time checks of academic systems catch. In other words, the
-experiment calls for a collection of type mistakes in mix-typed programs that is
-representative of those ``in the wild.'' Unfortunately no such collection exists,
-and with good reason; the kind of mistakes the rational programmer targets are
-easily detected by unit tests and, even if their sources take some time to find,
-these mistakes do not make it into code repositories with appropriate commit
-messages.  Hence the experimenter must generate a sufficiently large and
-representative corpus of type mistakes for a suitable collection of benchmark
-programs.  Following~\citet{lksfd-popl-2020} again, the solution is to synthesize
-mistakes using mutation analysis~\cite{lipton1971fault, demillo1978hints,
-jia2011analysis}.  The mutation analysis of Lazarek et al, is, however, useless
-for creating impedance mismatches.  While mutation analysis traditionally aims to
-inject bugs that challenge test suites, it discards those that yield ill-typed
-mutants as \emph{incompetent}.  Indeed, mutation analysis frameworks are
-fine-tuned to avoid them---and yet it is precisely those that are needed for
-evaluating blame assignment strategies.
+The {\em first derivative challenge\/} is to find a representative collection of
+impedance mismatches between typed and untyped pieces of code that programmers
+may accidentally create and that the run-time checks of academic systems
+catch. In other words, the experiment calls for a collection of type mistakes in
+mix-typed programs that is representative of those ``in the wild.''
+Unfortunately no such collection exists, and with good reason. The kind of
+mistakes needed are typically detected by unit or integration tests; even if it
+takes some time to find their sources, these mistakes do not make it into code
+repositories with appropriate commit messages.
+
+Following~\citet{lksfd-popl-2020} again, the alternative is to {\em generate a
+sufficiently large and representative corpus of type mistakes \/} using mutation
+analysis~\cite{lipton1971fault, demillo1978hints, jia2011analysis}.  The
+mutation analysis of Lazarek et al, is, however, useless for creating impedance
+mismatches.  While mutation analysis traditionally aims to inject bugs that
+challenge test suites, it discards those that yield ill-typed mutants as
+\emph{incompetent}.  Indeed, mutation analysis frameworks are fine-tuned to
+avoid them---and yet it is precisely those that are needed for evaluating blame
+assignment strategies.
 
@@
 design of such mutators, the authors relied on their own extensive programming
-experience though not without discovering a major pitfall: some of their original
-mutators systematically produced programs that immediately revealed the source of
-the impedance mismatch. About two thirds of the original mutators survived this
-first scrutiny testing. All of the remaining ones yield interesting debugging
-sessions.
+experience though not without discovering a major pitfall: some of their
+original mutators systematically produced programs that immediately revealed the
+source of the impedance mismatch. About two thirds of the original mutators
+survived this first scrutiny testing. All of the remaining ones yield
+interesting (see section~{sub:interesting}) debugging sessions.
 
-A somewhat coincidental challenge is the disparity of gradually typed
-languages. While the approaches are well understood in theory, they come with
-vastly different type systems, accommodate seemingly incompatible language
-features, and apply to radically different coding styles in the untyped world.
-To eliminate the effect of these differences on its results, the application of
-an evaluation method to different gradual typing systems must use the same test
-programs and the same type system. Otherwise, results for different gradual
-typing systems do not share a common baseline and, therefore, do not lend support
-to any conclusions about whether the value of blame changes from one system to
-another. For instance, the case study presented here builds on Greenman and
-Felleisen's Transient variant for functional Typed Racket~\cite{gf-icfp-2018} to
-obtain a Transient semantics that is comparable with Typed Racket's default
-Natural one. This Transient variant accommodates varied linguistic features and
-library calls in a way that is faithful to the Transient
-semantics~\cite{vss-popl-2017}.
+The {\em second derivative challenge\/} concerns the differences among the blame
+assignment of the three semantic variants.  While Natural assigns blame to {\em
+one\/} component, Transient assigns blame to a sequence of components. The
+Erasure semantics does not blame a component {\it per se\/}, but it comes with
+an exception location and a stack trace, which implicitly suggests parties to
+blame.  The different strategies greatly affect the behavior of the rational
+programmer (and real programmers, too).
 
-Beyond the sheer implementation issue, the development of a method must also
-address the problem that the three semantic variants---Natural, Transient, and
-Erasure---come with three different blame assignment strategies.  While Natural
-assigns blame to {\em one\/} component, Transient assigns blame to a sequence of
-components. The Erasure semantics does not blame a component per se, but it comes
-with an exception location and a stack trace.
+One way to overcome this challenge is to equip the rational programmer with
+\emph{modes} that represent the alternative interpretations of blame.  For
+instance, a mode of the rational programmer using Transient may pick the oldest
+element of a blame set as the next component to add types because it corresponds
+to the earliest point in the program's evaluation that can discover an impedance
+mismatch.  Another mode may opt to treat the blame set as another form of a
+stack trace and add types to its newest element.  When both modes are equally
+successful in locating an impedance mismatch, the method can measure the
+rational programmer's effort to answer which mode is the most effective.
 
-Supporting these strategies is {\em not\/} just an issue of language
-implementation but affects greatly the behavior of the rational
-programmer.  Specifically, the rational programmer must come with
-\emph{modes} that represent the alternative interpretations of blame.  For
-instance, a mode of the rational programmer using Transient may pick the
-oldest element of a blame set as the next component to add types because
-it corresponds to the earliest point in the program's evaluation that can
-discover an impedance mismatch.  Another mode may opt to treat the blame
-set as another form of a stack trace and add types to its newest element.
-When both modes are equally successful in locating an impedance mismatch,
-the method needs to take into account the rational programmer's effort
-to answer which mode is the most effective.
+Different blame information is also due to differences in the run-time checks
+that the three gradual typing systems perform. This observation suggests another
+mode for the rational programmer, namely, a mode that ignores the blame
+information completely and instead inspects only the information in the
+exceptions of the safety checks in the underlying languages.  This
+\emph{exception} mode serves as the baseline for blame's value; if the
+programmer in this mode ``beats'' the others, blame systems might be useless.
 
-%Another mode does not distinguish between the elements of the
-%blame set and has the rational programmer type all blamed components
-%at once. Because this second kind of step 
-%is more laborious than the first one, the analysis of the results must take 
-%into account the rational programmer's effort.
-
-Besides the blame information they produce, the three gradual typing
-systems also differ in the runtime type checks they perform. Therefore an
-application of the evaluation method to a gradual typing system has to
-factor in the effect of their different type checks to determine whether blame helps the
-rational programmer find the cause of an impedance mismatch. A solution to this issue
-is yet another mode for the rational programmer that uses information only
-from the underlying languages exceptions to select the next component(s) to type.
-This \emph{exception} mode simulates a rational programmer that ignores
-blame and serves as the baseline for blame's value.
-
-The final big challenge is the explosion of debugging scenarios that results from
-the combination of mutation-based scenario generation and mode-based analysis.
-Specifically, the experiment demands a large amount of computational resources
-due to the multiplicative effect of mixing the different gradual type systems,
-the number of debugging modes for the rational programmer, and the large number
-of mutants.  A practical application of the method has no option but to sample
-the space of scenarios, carefully ensuring that it still produces reproducible
-conclusions.
-
-\smallskip
+The {\em third challenge\/} is the explosion of debugging scenarios that results
+from the combination of mutation-based scenario generation and mode-based
+analysis. All three factors---three different gradual typing systems, the large
+number of mutants, and the number of debugging modes---contribute possibly
+useful debugging data in multiplicative manner. Hence, carried out naively, the
+experiment would demand an unacceptably large amount of computational resources.
+A practical execution has no option but to sample the space of scenarios,
+carefully ensuring that it still produces reproducible conclusions.
 
 Overcoming these challenges needs a systematic approach: 
-
-%% for doing so correctly
-%% MF: what is correct about this recipe?
-
 \begin{itemize}
 
-\item a single-language implementation of all three gradual type systems (section~\ref{sec:landscape});
+\item a suitable set of mutators for the generation of interesting type-level
+ errors (section~\ref{sec:mutate});
 
-\item a suitable set of mutators for the generation of a large set of debugging
-  scenarios with interesting type-level errors (section~\ref{sec:mutate});
+\item an algorithmic simulation of the many modes of the rational programmer on
+ the corresponding debugging scenarios (section~\ref{sec:rational});
 
-\item an algorithmic simulation of the rational programmer on these debugging
-  scenarios with sufficient modes to make meaningful comparisons between modes of
-  the same or different systems (section~\ref{sec:rational});
+\item a careful sampling of the space (section~\ref{sec:sample}).
+\end{itemize}
+Getting all three aspects right is necessary to obtain a computationally
+feasible and statistically sound experiment
 
-\item a careful sampling of the experimental space to obtain a computationally
-  feasible and statistically sound experiment (section~\ref{sec:sample}).
+{\bf Note} A somewhat coincidental challenge is the disparity of gradually typed
+languages. While the three approaches to blame are well understood in theory,
+they have been implemented for rather different untyped languages and with
+vastly different type systems. To eliminate the effect of these differences, it
+is necessary to implement all three approaches in a single language so that an
+application of an evaluation method can use the same test programs, the same
+type system, and hence the same debugging scenarios.
 
-\end{itemize}
+The results in this paper are for a system that builds on Greenman and
+Felleisen's effort comparing the performance of Natural and Transient in
+Racket~\cite{gf-icfp-2018}. The extended system uses the type system of Typed
+Racket and comes with a Transient semantics that is comparable with the Natural
+one. The Transient semantics of the system also realizes the design of the blame
+assignment system proposed by~\cite{vss-popl-2017}. The Erasure semantics drops
+the type information and runs the programs as if they were plain Racket
+programs. The next section explains this system in detail.

### 2021-02-22 writing/challenges.tex
stylistic patches, typo

@@
 empirical research similar to that of human-computer interaction. At the same
-time, a significant evaluation result demands a large number of cases yielding a
-serious amount of data. As~\citet{lksfd-popl-2020} recently demonstrated, the
-way around this dilemma is to postulate the equivalent of a {\it homo
-economicus\/} for programming languages---the {\em rational programmer\/}---and
-to implement this rational programmer as an algorithm.
+time, a significant evaluation result demands a large amount of
+data. As~\citet{lksfd-popl-2020} recently demonstrated, the way around this
+dilemma is to postulate the {\em rational programmer\/}---the equivalent of a
+{\it homo economicus\/} for programming languages---and to implement it as an algorithm.
 
@@
 impedance mismatch in an incremental fashion. Measuring this simulated
-programmer's behavior on a large number of debugging sessions yields data that
+programmer's behavior on a large number of debugging scenarios yields data that
 is similar to data collected in an empirical, human-facing manner.
@@
 The {\em first derivative challenge\/} is to find a representative collection of
-impedance mismatches between typed and untyped pieces of code that programmers
-may accidentally create and that the run-time checks of academic systems
-catch. In other words, the experiment calls for a collection of type mistakes in
-mix-typed programs that is representative of those ``in the wild.''
-Unfortunately no such collection exists, and with good reason. The kind of
-mistakes needed are typically detected by unit or integration tests; even if it
-takes some time to find their sources, these mistakes do not make it into code
-repositories with appropriate commit messages.
+impedance mismatches. They must represent mistakes that programmers accidentally
+create and that the run-time checks of academic systems catch. In other words,
+the experiment calls for a collection of type mistakes in mix-typed programs
+that is representative of those ``in the wild.''  Unfortunately no such
+collection exists, and with good reason. The kind of mistakes needed are
+typically detected by unit or integration tests; even if it takes some time to
+find their sources, these mistakes do not make it into code repositories with
+appropriate commit messages.
 
-Following~\citet{lksfd-popl-2020} again, the alternative is to {\em generate a
-sufficiently large and representative corpus of type mistakes \/} using mutation
-analysis~\cite{lipton1971fault, demillo1978hints, jia2011analysis}.  The
-mutation analysis of Lazarek et al, is, however, useless for creating impedance
-mismatches.  While mutation analysis traditionally aims to inject bugs that
-challenge test suites, it discards those that yield ill-typed mutants as
-\emph{incompetent}.  Indeed, mutation analysis frameworks are fine-tuned to
-avoid them---and yet it is precisely those that are needed for evaluating blame
-assignment strategies.
+An alternative is to {\em generate a large and representative corpus of type
+mistakes \/} using mutation analysis~\cite{lipton1971fault, demillo1978hints,
+jia2011analysis}.  Conventional mutation analysis is useless, however. Mutation
+analysis traditionally aims to inject bugs that challenge test suites, and it
+discards those that yield ill-typed mutants as \emph{incompetent}.  Indeed,
+mutation analysis frameworks are fine-tuned to avoid them, and yet, it is
+precisely those that are needed for evaluating blame assignment strategies.
 
@@
 source of the impedance mismatch. About two thirds of the original mutators
-survived this first scrutiny testing. All of the remaining ones yield
-interesting (see section~{sub:interesting}) debugging sessions.
+survived this first scrutiny testing. All of the remaining ones yield {\em
+interesting debugging scenarios\/} (see section~\ref{sub:interesting}).
 
@@
 
-One way to overcome this challenge is to equip the rational programmer with
-\emph{modes} that represent the alternative interpretations of blame.  For
-instance, a mode of the rational programmer using Transient may pick the oldest
-element of a blame set as the next component to add types because it corresponds
-to the earliest point in the program's evaluation that can discover an impedance
-mismatch.  Another mode may opt to treat the blame set as another form of a
-stack trace and add types to its newest element.  When both modes are equally
-successful in locating an impedance mismatch, the method can measure the
-rational programmer's effort to answer which mode is the most effective.
+One way to overcome this challenge is to {\em equip the rational programmer with
+modes \/} that represent the alternative interpretations of blame.  For
+instance, one Transient mode may pick the oldest element of a blame set as
+the next component to add types because it corresponds to the earliest point in
+the program's evaluation that can discover an impedance mismatch.  Another mode
+may opt to treat the sequence as a stack and add types to its newest element.
+When both modes are equally successful in locating an impedance mismatch,
+measuring the rational programmer's effort may answer which mode is the most
+effective.
 
@@
 information completely and instead inspects only the information in the
-exceptions of the safety checks in the underlying languages.  This
-\emph{exception} mode serves as the baseline for blame's value; if the
-programmer in this mode ``beats'' the others, blame systems might be useless.
+exceptions of the safety checks in the underlying languages.  This exception
+mode serves as the {\em baseline\/} for blame's value; if the programmer in this
+mode ``beats'' the others, blame systems might be useless.
 
-The {\em third challenge\/} is the explosion of debugging scenarios that results
-from the combination of mutation-based scenario generation and mode-based
+The {\em third challenge\/} is the explosive number of debugging scenarios that
+result from the combination of mutation-based scenario generation and mode-based
 analysis. All three factors---three different gradual typing systems, the large
 number of mutants, and the number of debugging modes---contribute possibly
-useful debugging data in multiplicative manner. Hence, carried out naively, the
-experiment would demand an unacceptably large amount of computational resources.
-A practical execution has no option but to sample the space of scenarios,
-carefully ensuring that it still produces reproducible conclusions.
+useful experimental data in a multiplicative manner. Hence, carried out naively,
+the experiment would demand an unacceptably large amount of computational
+resources.  A practical execution has no option but to {\em sample the space of
+scenarios\/}, carefully ensuring that it still produces reproducible
+conclusions.
 
-Overcoming these challenges needs a systematic approach: 
+Overcoming these challenges yields a systematic approach: 
 \begin{itemize}
 
-\item a suitable set of mutators for the generation of interesting type-level
- errors (section~\ref{sec:mutate});
+\item interesting mutators generate interesting type-level errors and thus
+debugging scenarios (section~\ref{sec:mutate});
 
-\item an algorithmic simulation of the many modes of the rational programmer on
- the corresponding debugging scenarios (section~\ref{sec:rational});
+\item an algorithm simulates the modes of the rational programmer on these
+scenarios (section~\ref{sec:rational}); and
 
-\item a careful sampling of the space (section~\ref{sec:sample}).
+\item a careful sampling of the space yields experimental
+data about the bug finding process (section~\ref{sec:sample}).
 \end{itemize}
 Getting all three aspects right is necessary to obtain a computationally
-feasible and statistically sound experiment
+feasible and statistically sound experiment.
 
-{\bf Note} A somewhat coincidental challenge is the disparity of gradually typed
+{\bf Note} A coincidental challenge is the disparity of gradually typed
 languages. While the three approaches to blame are well understood in theory,
@@
 is necessary to implement all three approaches in a single language so that an
-application of an evaluation method can use the same test programs, the same
+application of the evaluation method can use the same test programs, the same
 type system, and hence the same debugging scenarios.
@@
 The results in this paper are for a system that builds on Greenman and
-Felleisen's effort comparing the performance of Natural and Transient in
+Felleisen's effort of comparing the performance of Natural and Transient in
 Racket~\cite{gf-icfp-2018}. The extended system uses the type system of Typed
-Racket and comes with a Transient semantics that is comparable with the Natural
-one. The Transient semantics of the system also realizes the design of the blame
-assignment system proposed by~\cite{vss-popl-2017}. The Erasure semantics drops
-the type information and runs the programs as if they were plain Racket
-programs. The next section explains this system in detail.
+Racket and comes with a Transient semantics that faithfully implements the
+design and the blame assignment strategy proposed by~\cite{vss-popl-2017}. The
+Erasure semantics drops the type information and runs the programs as if they
+were plain Racket programs. The next section explains this system in detail.

### 2021-02-22 writing/landscape.tex
create an editorial voice

@@
-\section{How to Bring Natural, Transient and Erasure Under the Same Roof} 
+\section{How to Think About All Three Systems} 
 \label{sec:landscape}
@@
 
-The  program in figure~\ref{fig:example} demonstrates the differences
-between Natural, Transient and Erasure using Typed Racket syntax.  It
-consists of four modules: the left and right ones are untyped while
-the two modules in the middle are typed.
+The program sketch in figure~\ref{fig:example} demonstrates the
+differences between Natural, Transient and Erasure using Typed Racket
+syntax.  The modules on the left and right are untyped while the two
+in the middle are typed.
 
-The \texttt{untyped/pack-lib} module (left) provides, among
-others, a function \texttt{pack} that consumes JSON data and packages it
-in a dictionary. However, by mistake \texttt{pack} returns
-a hash table instead of the association list its documentation promises.
-The \texttt{client} module (right) uses \texttt{pack} indirectly to
-package public and sensitive data it reads from the filesystem. It does so
-through two different intermediary typed modules (center). The first,
-\texttt{typed/pack-lib}, simply imports \texttt{pack} and immediately
-exports as \texttt{typed-pack} asserting that it is a function
-that consumes JSON and returns an association list that maps symbols to
-strings. The second typed module, \texttt{typed/crypto-pack-lib},
-also imports \texttt{pack} and assumes for it the same type as
-\texttt{typed/pack-lib} but instead of exporting it directly \texttt{typed/pack-lib} wraps it in
-another function, \texttt{crypto-pack}, that encrypts its input before
-passing it to \texttt{pack}.
+The \texttt{untyped/pack-lib} module (left) provides, among others, a function
+\texttt{pack} that consumes JSON data and packages it in a dictionary. However,
+by mistake \texttt{pack} returns a hash table instead of the association list
+its documentation promises.  The \texttt{client} module (right) uses
+\texttt{pack} indirectly to package public and sensitive data it reads from the
+filesystem. It does so through two different intermediary typed modules
+(center). The first, {\tt typed/pack-lib}, imports \texttt{pack} and re-exports
+it as \texttt{typed-pack} asserting that it is a function that consumes {\tt
+JSON} and returns a list associating {\tt Symbol}s with {\tt String}s. The
+second typed module, {\tt typed/crypto-pack-lib}, also imports \texttt{pack} and
+assumes for it the same type as {\tt typed/pack-lib}. Instead of exporting it,
+it uses the function in the exported {\tt crypto-pack} function, which encrypts
+its input before passing it to \texttt{pack}.
 
-Under the Natural semantics~\cite{tf-popl-2008,tfffgksst-snapl-2017}, when
-\texttt{pack} is imported in either \texttt{typed/pack-lib} or
-\texttt{typed/secure-pack-lib}, the runtime checks that \texttt{pack} is a
-function and wraps it in a proxy that enforces its  type.  Analogously
-the runtime wraps the exports of the two typed modules in proxies that
-check that untyped code applies \texttt{typed-pack} and
-\texttt{crypto-pack} to arguments of the expected type. In essence, the
-return type checks protect  the typed modules in case they use
-\texttt{pack} and, as is the case here, it does not return an association
-list. The check of the argument type  treats the
-exports of the two typed modules as typed values that need to be protected
-when they are used by untyped code.  As this analysis implies, if the
-return type check fails then the problem is with
-\texttt{untyped/pack-lib}, while if the argument type check fails
-responsibility lies with \texttt{client}. The Natural semantics keep track
-of the responsible party for each check in the corresponding proxy. Thus,
-in our example, as the call to \texttt{typed-pack} returns, the return
-type check fails and the Natural semantics blames \texttt{pack-lib}.
+Under the {\it Natural\/} semantics~\cite{tf-popl-2008,tfffgksst-snapl-2017},
+functions imported into and exported from typed modules are wrapped in proxies
+that enforce the static type discipline with run-time checks and track
+responsibilities. Thus, when \texttt{pack} is imported into a typed module, the
+run-time system checks that it is a function and wraps it in a protective proxy,
+which in turn, enforces the type of the function with run-time checks.  In
+essence, this check protects typed modules when they call an untyped function
+because it may not produce the promised kind of value, as is the case here with
+{\tt pack} from {\tt untyped/pack-lib}. Analogously the run-time system wraps
+exported functions such as {\tt crypto-pack} of a typed modules in a proxy that
+checks its arguments.  This check protects functions exported from type modules
+against applications in untyped code that supply the wrong kind of argument
+value.
 
-Under the Transient semantics~\cite{vss-popl-2017}, the call to \texttt{typed-pack} does not
-produce an error. In contrast to Natural, Transient does not rely on
-proxies. Instead, it rewrites typed modules to inline checks at appropriate
-places. Furthermore, these checks are
-partial, that is, they check only the top constructor of a type. In other words
-they check types in a  \emph{shallow} manner. In our
-example, Transient's inlined checks make sure that the imported \texttt{pack} 
-is a function and that the result of its call in \texttt{typed/crypto-pack-lib} is a list.   Exactly this
-last check fails when \texttt{client} calls to \texttt{crypto-pack}. In
-order to pinpoint the responsible party for the failed check, Transient
-keeps on the side a map from values to the boundaries between modules they
-cross and the corresponding types. In the example, the map records that
-\texttt{pack} crosses from \texttt{untyped/pack-lib} to
-\texttt{typed/pack-lib} and from \texttt{typed/pack-lib} to
-\texttt{client} with the type that appears in the \texttt{required/typed}
-forms in the example. Since the failed check corresponds to the return
-type of \texttt{pack}, Transient blames the sources of the two
-boundary crosses, which in this case is \texttt{untyped/pack-lib} for
-both. In general though, Transient may blame more than one party. In
-fact, the theoretical work of~\citet{gfd-oopsla-2019} shows that there
-are programs where Transient constructs a blame set that includes
-modules that are not relevant to the check failure or that it even
-entirely misses the responsible party--our experiment aims to examine
-the pragmatic impact of exactly these issues on debugging type-level
-mistakes.
+As this analysis implies, if the return-type check fails, the problem is with
+the untyped module that supplied an incorrect function, here {\tt
+untyped/pack-lib}. If the argument-type check fails, responsibility lies with
+{\tt client}. Due to the proxies wrapped around functions, Natural can easily
+track the responsible party for each check. Thus, in the example of
+figure~\ref{fig:example}, as \texttt{typed-pack} returns, the return-type check
+fails, and Natural blames \texttt{pack-lib}.
 
-Under the Erasure semantics, the example does not raise an error due
-to the gradual type system; Erasure performs no type checks at
-runtime. Thus the program may fail with an exception if
-\texttt{client} tries to use the result of \texttt{typed-pack} or
-\texttt{crypto-pack}, but it may also not fail at all and produce a
-result of the proper kind---and yet is completely wrong.  In either
-case, the Erasure semantics provides no help to the programmer for
-detecting the bug in our program.
+Under the Transient semantics~\cite{vss-popl-2017}, typed code is compiled so
+that all entry points to functions check their arguments at run time and all
+function calls check their return values against the expected type.
+Furthermore, Transient uses \emph{shallow} checks, meaning they check only the
+top constructor of a value. Since retrieving a value from within a structure (or
+list, array, hash table etc.) is performed via a function call, the innards of a
+complex value are checked on a piecemeal basis.
 
-Since, the goal of our experiment is to compare the effectiveness of blame
-across the three semantics, it is critical to isolate their
-distinguishing characteristics from superficial implementation
-differences. Thus we implement all three systems on top of the common
-basis of Racket. First, we use the Typed Racket syntax and type
-checker for all three semantics. Second, for Natural we use Typed Racket as
-is while for Erasure we run programs with plain Racket after erasing any
-type annotations. Third, for Transient we implement a variant of Typed
-Racket that follows closely the semantics from~\citet{vss-popl-2017}.
+As a result, the call to \texttt{typed-pack} does not signal an error because
+the call takes place in the untyped {\tt client} module, which is compiled in
+the usual manner. Because {\tt pack} is called in the {\tt
+typed/crypto-pack-lib} module, Transient's inlined checks make sure that the
+imported \texttt{pack} is a function and that the result of its call in is a
+list. This last check fails when \texttt{client} calls to {\tt crypto-pack}.
 
-The first major ingredient to the construction of Transient Typed Racket is to determine the
-Racket implementation of the shallow checks that map to each type.  For
-certain types, such as {\tt String} or {\tt (Vectorof Byte)}, Transient
-can re-use a predicate from the Racket language.  Other types require custom
-predicates; for example the type {\tt Index} of valid array indices allows
-a range of positive exact integers.  Many custom predicates run in
-constant time, but some are more involved.  One linear-time check is for
-{\tt (Listof T)} for any element type {\tt T}, because it must distinguish
-proper from improper list-like structures.  The
-key point, however, is that no shallow check inspects beyond the top
+In order to pinpoint the responsible party for failed checks, Transient
+maintains a map from values to the boundaries between typed/untyped modules that
+they cross, plus the corresponding types. In the example, the map records that
+\texttt{pack} crosses from \texttt{untyped/pack-lib} to {\tt typed/pack-lib} and
+from {\tt typed/pack-lib} to {\tt client} with the type that appears in the {\tt
+required/typed} forms in the example. Since the failed check corresponds to the
+return type of \texttt{pack}, Transient blames the sources of the two boundary
+crosses, which in this particular case is \texttt{untyped/pack-lib} for both. In
+general though, Transient blames more than one party. In fact, the theoretical
+work of~\citet{gfd-oopsla-2019} shows that for some programs Transient
+constructs a blame set that excludes responsible parties and includes modules
+irrelevant to the failing check.
+
+Under the Erasure semantics, the compiler checks the specified types and then
+discards them when it generates code. The generated code includes run-time
+checks that ensure the dynamic safety of all operations as specified in the
+underlying untyped language. Hence, neither the call to {\tt typed-pack} nor the
+call to {\tt crypto-pack} signals an error due to the gradual type system.
+
+When run, the program may fail with an exception if \texttt{client} tries to
+inspect the elements of the list that \texttt{typed-pack} and
+\texttt{crypto-pack} are supposed to produce. The information in this exception,
+plus its stack trace, or may not help the programmer find the source of the
+impedance mismatch between the specified return types of {\tt pack} in the two
+typed modules and the actual return value of {\tt pack}.
+
+All three systems used here are implemented on top of the common basis of
+Racket. The use the Typed Racket syntax and type checker for all three
+semantics. The Natural semantics is inherited as is, while the Erasure semantics
+is the one of plain Racket.  The Transient system implements the design
+of~\citet{vss-popl-2017}, though constructing it demands two design decisions:
+which shallow checks to use and where, plus how to maintain a blame map in a
+full-fledged run-time system.
+
+For certain types, such as {\tt String} or {\tt (Vectorof Byte)}, Transient can
+realize shallow checks with the predicate from Racket.  Other types require
+custom predicates; for example the type {\tt Index} of valid array indices
+allows a range of positive exact integers.  Many custom predicates run in
+constant time, but some, such as the one for {\tt (Union T1 ... Tn)}, run in
+linear time. The key point is that no shallow check inspects beyond the top
 constructor of each type.
 
-The second major ingredient of Transient Typed Racket  is the addition of
-a pass to the compiler of Typed Racket that rewrites all typed code.
-Rewriting modifies all elimination forms to shallow-check their result.
-For example, if the variable {\tt f} has the static type {\tt (-> String
-String)} then the application {\tt (f x)} must be rewritten to check for a
-string result because {\tt f} could be an untyped function.  Rewriting
-also inserts hooks to update the blame map.  When a value crosses a module
-boundary, the blame map records a type assumption.  The inlined checks
-for elimination forms also extend the blame map with further information. For
-instance, the check of the result of {\tt (f x)} adds an entry to the blame map
-that records the expected type of the result as the return type of the type of
-{\tt f}.  If a Transient check ever fails, Transient Typed Racket walks over 
-the information in the blame map and reports a set of blamed components.
+Once the type checker has completed its work, Transient translates the source
+into an alternative form and in the process inserts run-time checks. Technically
+speaking, the rewriting pass wraps all elimination in applications of shallow
+checks.  For example, if the variable {\tt f} has the static type {\tt (->
+String String)}, the application {\tt (f x)} must be rewritten to check for a
+string result because {\tt f} could be an untyped function.
 
-\smallskip
-\noindent
-{\bf Note.} As the common basis for the three semantics, we choose 
-Typed Racket's migratory typing. In contrast to the initial formulation of
-gradual typing~\cite{st-sfp-2006}, migratory typing does not provide
-support for type \texttt{Dyn}, components are either untyped or fully
-typed.
-%% MF: does it ever:
-%% and the gradual type system does not affect the compilation of untyped components. 
-Despite, these differences, Typed Racket is the most
-mature system for sound mix-typed code
-and recent theoretical work has established that
-its migratory typing offers a setting for comparing different approaches to gradual
-typing~\cite{gf-icfp-2018,gfd-oopsla-2019}.
+The rewriting pass also inserts hooks for updating the blame map.  When a value
+crosses a module boundary, such a hook records a type assumption in the blame
+map.  The in-lined checks for elimination forms extend the blame map with
+further information. For instance, the check of the result of {\tt (f x)} adds
+an entry to the blame map that records the expected type of the result as the
+return type of the type of {\tt f}.  If a Transient check ever fails, the
+run-time checking system of Transient Racket walks over the information in the
+blame map and reports a sequence of blamed components.
+
+%% MF: I have no clue what the following contributes:
+
+% \smallskip
+% \noindent
+% {\bf Note.} As the common basis for the three semantics, we choose Typed
+% Racket's migratory typing. In contrast to the initial formulation of gradual
+% typing~\cite{st-sfp-2006}, migratory typing does not provide support for type
+% \texttt{Dyn}, components are either untyped or fully typed.
+% %% MF: does it ever:
+% %% and the gradual type system does not affect the compilation of untyped components. 
+% Despite, these differences, Typed Racket is the most
+% mature system for sound mix-typed code
+% and recent theoretical work has established that
+% its migratory typing offers a setting for comparing different approaches to gradual
+% typing~\cite{gf-icfp-2018,gfd-oopsla-2019}.

### 2021-02-22 writing/landscape.tex
organize sections/titles/labels in paper
overfull box

@@
-\section{How to Think About All Three Systems} 
-\label{sec:landscape}
-
 %% -----------------------------------------------------------------------------
@@
 
-When run, the program may fail with an exception if \texttt{client} tries to
+When run, the program may raise an exception if \texttt{client} tries to
 inspect the elements of the list that \texttt{typed-pack} and

### 2021-02-22 writing/mutate-benchmarks.tex
factor mutate

--- 2021-02-12 writing/mutate.tex
@@
-\section{How to Make Lots of Mistakes} 
-\label{sec:mutate}
-
-%% this is awkward, but what I removed was exponentially awkward
-
-In this section, we
-present a generation process and argue that, despite its synthetic
-nature, it yields a large corpus of programs with  realistic and diverse type mistakes.
+The starting point for our corpus of programs
+is~\citet{gtnffvf-jfp-2019}'s gradual typing benchmark suite. The
+benchmark suite consists of fully typed correct programs that are written by different authors
+and have been used, maintained and evolved by their authors and others over a
+number of years.  They range widely in the size, complexity, purpose and
+features of Typed Racket they employ.  Furthermore the benchmarks use advanced aspects
+of Typed Racket's type system such as occurrence
+typing~\cite{tf-icfp-2010}, types for mutable and immutable data
+structures, polymorphic types, types for first-class classes and objects, and types for
+Racket's numeric tower~\cite{stathff-padl-12}. Without loss of the diversity of the benchmarks,
+we select the ten largest in terms of numbers of components (between 6 and 14 components).
+These benchmarks also have the most complex dependency graphs and, hence,
+can make the debugging process the hardest for the rational programmer.
+The table in figure~\ref{table:benchmark-descriptions} shows our selection
+together with a short description for each benchmark.
 
@@
 \hline
-
   \texttt{acquire} & Object-oriented board game implementation (M. Felleisen)  \\%[1em]
@@
 \end{figure}
-
-The starting point for our corpus of programs
-is~\citet{gtnffvf-jfp-2019}'s gradual typing benchmark suite. The
-benchmark suite consists of fully typed correct programs that are written by different authors
-and have been used, maintained and evolved by their authors and others over a
-number of years.  They range widely in the size, complexity, purpose and
-features of Typed Racket they employ.  Furthermore the benchmarks use advanced aspects
-of Typed Racket's type system such as occurrence
-typing~\cite{tf-icfp-2010}, types for mutable and immutable data
-structures, polymorphic types, types for first-class classes and objects, and types for
-Racket's numeric tower~\cite{stathff-padl-12}. Without loss of the diversity of the benchmarks,
-we select the ten largest in terms of numbers of components (between 6 and 14 components).
-These benchmarks also have the most complex dependency graphs and, hence,
-can make the debugging process the hardest for the rational programmer.
-The table in figure~\ref{table:benchmark-descriptions} shows our selection
-together with a short description for each benchmark.
 
@@
 attempt to confuse occurrence typing.
-
-\begin{figure*}
-  \begin{tabular}{p{2.1cm} | p{7cm}  | p{7.5cm} }
-    {\bf name} & {\bf description} & {\bf example} \\
-\hline
-
-    \texttt{constant}
-& Swaps a constant with another of the same value but different type.
-& \texttt{5.6} $\rightarrow$ \texttt{5.6+0.0i} \\
-\hline
-
-    \texttt{deletion}
-    & Deletes the result expression of a sequence.
-    & \texttt{(begin x y z)} $\rightarrow$ \texttt{(begin x y)} \\
-\hline
-
-   \texttt{position}
- & Swaps two subexpressions.
-  &  \texttt{(f a 42 "b" 0)} $\rightarrow$\texttt{(f a 42 0 "b")} \\
-\hline
-
-     \texttt{list}
-    & Replaces \texttt{append} with \texttt{cons}.
-    &  \begin{tabular}[t]{@{}l}
-      \texttt{append} $\rightarrow$ \texttt{cons}
-    \end{tabular}\\
-\hline
-
-    \texttt{top-level-id}
- & Swaps identifiers defined in the same module.
- &\texttt{(f x 42)} $\rightarrow$ \texttt{(g x 42)}\\
-\hline
-
-    \texttt{imported-id}
- & Swaps identifiers imported from the same module.
- &\texttt{(f x 42)} $\rightarrow$ \texttt{(g x 42)}\\
-\hline
-
-    \texttt{method-id}
- & Swaps two method identifiers.
- &\texttt{(send o f x 42)} $\rightarrow$ \texttt{(send o g x 42)}\\
-\hline
-
-    \texttt{field-id}
- & Swaps two field identifiers.
-    &  \texttt{(get-field o f)} $\rightarrow$  \texttt{(get-field o g)}\\
-\hline
-
-     \texttt{class:init}
-  & Swaps values of class initializers.
-   &  \begin{tabular}[t]{@{}l}
-     \texttt{(new c [a 5] [b "hello"])} $\rightarrow$\\
-     \texttt{(new c [a "hello"] [b 5])}\\
-    \end{tabular}\\
-\hline
-
-     \texttt{class:parent}
-    & Replaces the parent of classes with \texttt{object\%}.
-    &  \begin{tabular}[t]{@{}l}
-     \texttt{(class foo\% (super-new))} $\rightarrow$\\
-     \texttt{(class object\% (super-new))}\\
-    \end{tabular}\\
-\hline
-
-     \texttt{class:public}
-    & Makes a public method private and vice versa.
-    & \begin{tabular}[t]{@{}l}
-      \texttt{(class object (define/public (m x) x))} $\rightarrow$\\
-      \texttt{(class object (define/private (m x) x))}\\
-    \end{tabular}\\
-\hline
-
-     \texttt{class:super}
-    & Removes \texttt{super-new} calls from class definitions.
-    & \begin{tabular}[t]{@{}l}
-      \texttt{(class foo\% (super-new))} $\rightarrow$ \\
-      \texttt{(class foo\% (void))}\\
-       \end{tabular}\\
-\hline
-
-     \texttt{arithmetic}
-& Swaps arithmetic operators.
-    & \texttt{+} $\rightarrow$ \texttt{-} \\
-\hline
-
-     \texttt{boolean}
-& Swaps \texttt{and} and \texttt{or}.
-& \texttt{and} $\rightarrow$ \texttt{or} \\
-\hline
-
-     \texttt{negate-cond}
-    & Negates conditional test expressions.
-    &  \texttt{(if (= x 0) t e)} $\rightarrow$ \texttt{(if (not (= x 0)) t e)}\\
-\hline
-
-     \texttt{force-cond}
-    & Replaces conditional test expressions with \texttt{\#t}.
-    & \texttt{(if (= x 0) t e)} $\rightarrow$ \texttt{(if \#t t e)}
-\end{tabular}
-  \caption{Summary of mutators.}
-  \label{table:mutation-ops}
-\end{figure*}
-
-\subsection{Are these mutators any good?}
-
-The reader may have observed that mutators that produce 
-type-level mistakes do not necessarily turn our benchmarks into
-interesting debugging scenarios for the rational programmer.  The
-benchmarks are fully typed, and type checking detects each mistake at
-compile time.  Hence, we apply the mutators to versions, dubbed
-dubbed{scenarios}, of the benchmarks without 
-some  of their type annotations.  Without the annotations, the type
-checker may not detect a mistake at compile time and, instead, the run
-time checks
-of the gradual type system signal an error.  Scenarios
-with type-level mistakes that result in run time errors are the ideal
-ground to exercise the rational programmer's ability to locate errors.
-
-Specifically, we define an \emph{interesting debugging scenario} as a
-scenario that
-\begin{enumerate} 
-  \item is rejected by the type checker when we restore all its missing type annotations, 
-  \item produces a run time error under Erasure as-is, and 
-  \item the stack trace of the error mentions at least three distinct components.  
-\end{enumerate}
-
-The first criterion simply validates that the debugging scenario concerns
-an actual type-level mistake. The second one stipulates that the mistake is
-detectable under Erasure to ensure that it is detectable by all three
-semantics.   Whether a scenario produces a run time error largely depends on 
-the underlying semantics but \citet{gfd-oopsla-2019}
-show that a run time error under Erasure implies a run time
-error under both Transient and Natural. Thus the criterion aligns
-interesting debugging scenarios with our lowest denominator, Erasure, 
-and excludes  from consideration the very interesting scenarios indeed where Natural or
-Transient can detect a mistake but where Erasure is powerless. Of course this choice favors
-Erasure over Transient and Natural and, for the same reason, Transient over
-Natural. We consider some form of bias towards one or the other semantics
-unavoidable no matter how we define interesting debugging scenarios and we
-opt for tipping the scales in favor of the theoretically weakest
-semantics. We further discuss the implications of this choice in
-section~\ref{sec:discussion}.
-
-The third criterion aims to capture debugging scenarios with a specific
-kind of run time error: one produced because of the interaction between the buggy
-portion of the program with the rest of the program.  In these cases, the rational programmer
-may need to examine several pieces of the program to locate the source of the faulty interaction.
-In technical terms, we reliably mark these scenarios by
-analyzing the stack trace of a run time error.  If the stack trace mentions at
-least two distinct components, then this indicates interaction.  We require a third component because our benchmarks come
-with a driver component which is present in the stack trace of all run
-time errors. In essence, this last criterion excludes a large number of
-trivial to debug, and thus ineffective, scenarios that error immediately when their mutated component is
-evaluated.
-
-%\begin{figure*}
-%  \centering
-%  \includegraphics[scale=0.35]{./plots/mutant-breakdown}
-%  \caption{Interesting debugging scenarios produced by our mutators. Counts are cut off at 50.}
-%  \label{fig:mutant-breakdown}
-%\end{figure*}
-
-With the definition of interesting debugging scenarios in hand, we analyze how effective our mutators
-are for turning the benchmarks into interesting debugging scenarios.
-At a high level, the mutators produce 16,800 mutants with at least one interesting debugging scenario each across all of our benchmarks.
-Broken down by benchmark, the mutators produce at least 40 interesting
-scenarios for every benchmark, and these scenarios originate from at
-least four different mutators per benchmark.  Thus the mutators result in 
-a sizeable and diverse population of scenarios for every benchmark.
-Furthermore, every mutator contributes scenarios to at least one
-benchmark.
-That being said, some mutators only apply to a small subset of the benchmarks;
-those mutators target specific features which only those benchmarks make us of, so we include them despite their narrow applicability.
-For instance, the class-focused mutators are mainly effective in \texttt{take5}, since it is the benchmark with the most extensive use of object-oriented features.

### 2021-02-22 writing/mutate-interesting.tex
factor mutate

--- 2021-02-12 writing/mutate.tex
@@
-\section{How to Make Lots of Mistakes} 
-\label{sec:mutate}
-
-%% this is awkward, but what I removed was exponentially awkward
-
-In this section, we
-present a generation process and argue that, despite its synthetic
-nature, it yields a large corpus of programs with  realistic and diverse type mistakes.
-
-\begin{figure}
-\begin{tabular}{p{1.75cm} | p{6cm} }
-  {\bf  name} & {\bf description (author)}  \\
-
-\hline
-
-  \texttt{acquire} & Object-oriented board game implementation (M. Felleisen)  \\%[1em]
-
-\hline
-  \texttt{gregor} & Utilities for calendar dates (J. Zeppieri) \\%[1em]
-
-\hline
-  \texttt{kcfa} & Functional implementation of 2CFA for the lambda calculus (M. Might) \\%[1em]
-
-\hline
-  \texttt{quadT} & Converter from S-expression source code to PDF format (M. Butterick)\\%[1em]
-
-\hline
-  \texttt{quadU} & Converter from S-expression source code to PDF format  (B. Greenman) \\%[1em]
-
-\hline
-  \texttt{snake} & Functional implementation of the  Snake video game (D. Van Horn) \\%[1em]
-
-\hline
-  \texttt{suffixtree} & Algorithm for common longest subsequences between strings. (D. Yoo) \\%[1em]
-
-\hline
-  \texttt{synth} & Converter of notes and drum beats to WAV (V. St-Amour \& N. Toronto) \\%[1em]
-
-\hline
-  \texttt{take5} & Mixin-based card game simulator (M.Felleisen)  \\%[1em]
-
-\hline
-  \texttt{tetris} & Functional implementation of Tetris (D. Van Horn) \\%[1em]
-
-\end{tabular}
-  \caption{Benchmarks summary.}
-  \label{table:benchmark-descriptions}
-\end{figure}
-
-The starting point for our corpus of programs
-is~\citet{gtnffvf-jfp-2019}'s gradual typing benchmark suite. The
-benchmark suite consists of fully typed correct programs that are written by different authors
-and have been used, maintained and evolved by their authors and others over a
-number of years.  They range widely in the size, complexity, purpose and
-features of Typed Racket they employ.  Furthermore the benchmarks use advanced aspects
-of Typed Racket's type system such as occurrence
-typing~\cite{tf-icfp-2010}, types for mutable and immutable data
-structures, polymorphic types, types for first-class classes and objects, and types for
-Racket's numeric tower~\cite{stathff-padl-12}. Without loss of the diversity of the benchmarks,
-we select the ten largest in terms of numbers of components (between 6 and 14 components).
-These benchmarks also have the most complex dependency graphs and, hence,
-can make the debugging process the hardest for the rational programmer.
-The table in figure~\ref{table:benchmark-descriptions} shows our selection
-together with a short description for each benchmark.
-
-Of course, the gradual typing benchmarks have no mistakes for the rational
-programmer to debug. Therefore, we follow ~\citet{lksfd-popl-2020} and we
-inject bugs with mutation
-analysis~\cite{jia2011analysis,demillo1978hints,lipton1971fault}.  A
-mutation is a local syntactic change of the code that may
-produce a bug. Usually, the outcome of mutating a program once, dubbed a
-mutant, can be distinguished from the original program with a test case.
-The better the test suite of a program, the more mutants it can ``kill.''
-Mutants that are hard to ``kill'' usually correspond to subtle programming
-errors, and mutation frameworks typically come with standard mutators
-carefully designed to create such mutants.
-
-Unlike the hard-to-kill mistakes introduced by standard mutators, 
-gradual type systems catch shallow errors.  Thus, in order to
-exercise the rational programmer in the context of gradual typing, 
-we need to develop alternative mutators
-that are fine-tuned to produce the type-level mistakes gradual type systems 
-can detect. Our inspiration is our
-more-than-a-decade long experience of making type-level mistakes in Typed
-Racket. Some of these mistakes often take non-trivial effort to debug and
-have become the seed for our mutators, which we summarize in
-figure~\ref{table:mutation-ops}.  For each one, the figure provides a
-short description and an example mutation.
-
-Most of our mutators are self-explanatory.
-The first twelve apply to most gradually typed languages, including those with classes.
-The last four mutators target distinguishing features of Typed Racket.
-Specifically, \texttt{arithmetic} may replace a \texttt{+} with a \texttt{-}
-in an attempt to change the type of the result of the arithmetic
-operation. In Typed Racket, \texttt{+}'s result is a
-\texttt{Positive-Integer} when all arguments are
-positive integers. However, the result of \texttt{-} is
-\texttt{Integer}. In the same spirit, \texttt{boolean} aims to take
-advantage of Typed Racket's ``truthiness.'' Finally, \texttt{negate-cond} and \texttt{force-cond}
-attempt to confuse occurrence typing.
+The reader may have observed that mutators that produce 
+type-level mistakes do not necessarily turn our benchmarks into
+interesting debugging scenarios for the rational programmer.  The
+benchmarks are fully typed, and type checking detects each mistake at
+compile time.  Hence, we apply the mutators to versions, dubbed
+dubbed{scenarios}, of the benchmarks without 
+some  of their type annotations.  Without the annotations, the type
+checker may not detect a mistake at compile time and, instead, the run
+time checks
+of the gradual type system signal an error.  Scenarios
+with type-level mistakes that result in run time errors are the ideal
+ground to exercise the rational programmer's ability to locate errors.
 
@@
 
-\subsection{Are these mutators any good?}
-
-The reader may have observed that mutators that produce 
-type-level mistakes do not necessarily turn our benchmarks into
-interesting debugging scenarios for the rational programmer.  The
-benchmarks are fully typed, and type checking detects each mistake at
-compile time.  Hence, we apply the mutators to versions, dubbed
-dubbed{scenarios}, of the benchmarks without 
-some  of their type annotations.  Without the annotations, the type
-checker may not detect a mistake at compile time and, instead, the run
-time checks
-of the gradual type system signal an error.  Scenarios
-with type-level mistakes that result in run time errors are the ideal
-ground to exercise the rational programmer's ability to locate errors.
-
 Specifically, we define an \emph{interesting debugging scenario} as a

### 2021-02-22 writing/mutate.tex
label
organize sections/titles/labels in paper
factor mutate

@@
-\section{How to Make Lots of Mistakes} 
-\label{sec:mutate}
+%% -----------------------------------------------------------------------------
 
-%% this is awkward, but what I removed was exponentially awkward
+As section~\ref{sec:challenges} explains, the first challenge is to
+generate a representative mistakes in representative software. The
+following two subsections explain in reverse order.
 
-In this section, we
-present a generation process and argue that, despite its synthetic
-nature, it yields a large corpus of programs with  realistic and diverse type mistakes.
+%% -----------------------------------------------------------------------------
+\def\sub#1#2{\subsection{#2} \label{sub:mutate-#1} \input{mutate-#1.tex}}
 
-\begin{figure}
-\begin{tabular}{p{1.75cm} | p{6cm} }
-  {\bf  name} & {\bf description (author)}  \\
-
-\hline
-
-  \texttt{acquire} & Object-oriented board game implementation (M. Felleisen)  \\%[1em]
-
-\hline
-  \texttt{gregor} & Utilities for calendar dates (J. Zeppieri) \\%[1em]
-
-\hline
-  \texttt{kcfa} & Functional implementation of 2CFA for the lambda calculus (M. Might) \\%[1em]
-
-\hline
-  \texttt{quadT} & Converter from S-expression source code to PDF format (M. Butterick)\\%[1em]
-
-\hline
-  \texttt{quadU} & Converter from S-expression source code to PDF format  (B. Greenman) \\%[1em]
-
-\hline
-  \texttt{snake} & Functional implementation of the  Snake video game (D. Van Horn) \\%[1em]
-
-\hline
-  \texttt{suffixtree} & Algorithm for common longest subsequences between strings. (D. Yoo) \\%[1em]
-
-\hline
-  \texttt{synth} & Converter of notes and drum beats to WAV (V. St-Amour \& N. Toronto) \\%[1em]
-
-\hline
-  \texttt{take5} & Mixin-based card game simulator (M.Felleisen)  \\%[1em]
-
-\hline
-  \texttt{tetris} & Functional implementation of Tetris (D. Van Horn) \\%[1em]
-
-\end{tabular}
-  \caption{Benchmarks summary.}
-  \label{table:benchmark-descriptions}
-\end{figure}
-
-The starting point for our corpus of programs
-is~\citet{gtnffvf-jfp-2019}'s gradual typing benchmark suite. The
-benchmark suite consists of fully typed correct programs that are written by different authors
-and have been used, maintained and evolved by their authors and others over a
-number of years.  They range widely in the size, complexity, purpose and
-features of Typed Racket they employ.  Furthermore the benchmarks use advanced aspects
-of Typed Racket's type system such as occurrence
-typing~\cite{tf-icfp-2010}, types for mutable and immutable data
-structures, polymorphic types, types for first-class classes and objects, and types for
-Racket's numeric tower~\cite{stathff-padl-12}. Without loss of the diversity of the benchmarks,
-we select the ten largest in terms of numbers of components (between 6 and 14 components).
-These benchmarks also have the most complex dependency graphs and, hence,
-can make the debugging process the hardest for the rational programmer.
-The table in figure~\ref{table:benchmark-descriptions} shows our selection
-together with a short description for each benchmark.
-
-Of course, the gradual typing benchmarks have no mistakes for the rational
-programmer to debug. Therefore, we follow ~\citet{lksfd-popl-2020} and we
-inject bugs with mutation
-analysis~\cite{jia2011analysis,demillo1978hints,lipton1971fault}.  A
-mutation is a local syntactic change of the code that may
-produce a bug. Usually, the outcome of mutating a program once, dubbed a
-mutant, can be distinguished from the original program with a test case.
-The better the test suite of a program, the more mutants it can ``kill.''
-Mutants that are hard to ``kill'' usually correspond to subtle programming
-errors, and mutation frameworks typically come with standard mutators
-carefully designed to create such mutants.
-
-Unlike the hard-to-kill mistakes introduced by standard mutators, 
-gradual type systems catch shallow errors.  Thus, in order to
-exercise the rational programmer in the context of gradual typing, 
-we need to develop alternative mutators
-that are fine-tuned to produce the type-level mistakes gradual type systems 
-can detect. Our inspiration is our
-more-than-a-decade long experience of making type-level mistakes in Typed
-Racket. Some of these mistakes often take non-trivial effort to debug and
-have become the seed for our mutators, which we summarize in
-figure~\ref{table:mutation-ops}.  For each one, the figure provides a
-short description and an example mutation.
-
-Most of our mutators are self-explanatory.
-The first twelve apply to most gradually typed languages, including those with classes.
-The last four mutators target distinguishing features of Typed Racket.
-Specifically, \texttt{arithmetic} may replace a \texttt{+} with a \texttt{-}
-in an attempt to change the type of the result of the arithmetic
-operation. In Typed Racket, \texttt{+}'s result is a
-\texttt{Positive-Integer} when all arguments are
-positive integers. However, the result of \texttt{-} is
-\texttt{Integer}. In the same spirit, \texttt{boolean} aims to take
-advantage of Typed Racket's ``truthiness.'' Finally, \texttt{negate-cond} and \texttt{force-cond}
-attempt to confuse occurrence typing.
-
-\begin{figure*}
-  \begin{tabular}{p{2.1cm} | p{7cm}  | p{7.5cm} }
-    {\bf name} & {\bf description} & {\bf example} \\
-\hline
-
-    \texttt{constant}
-& Swaps a constant with another of the same value but different type.
-& \texttt{5.6} $\rightarrow$ \texttt{5.6+0.0i} \\
-\hline
-
-    \texttt{deletion}
-    & Deletes the result expression of a sequence.
-    & \texttt{(begin x y z)} $\rightarrow$ \texttt{(begin x y)} \\
-\hline
-
-   \texttt{position}
- & Swaps two subexpressions.
-  &  \texttt{(f a 42 "b" 0)} $\rightarrow$\texttt{(f a 42 0 "b")} \\
-\hline
-
-     \texttt{list}
-    & Replaces \texttt{append} with \texttt{cons}.
-    &  \begin{tabular}[t]{@{}l}
-      \texttt{append} $\rightarrow$ \texttt{cons}
-    \end{tabular}\\
-\hline
-
-    \texttt{top-level-id}
- & Swaps identifiers defined in the same module.
- &\texttt{(f x 42)} $\rightarrow$ \texttt{(g x 42)}\\
-\hline
-
-    \texttt{imported-id}
- & Swaps identifiers imported from the same module.
- &\texttt{(f x 42)} $\rightarrow$ \texttt{(g x 42)}\\
-\hline
-
-    \texttt{method-id}
- & Swaps two method identifiers.
- &\texttt{(send o f x 42)} $\rightarrow$ \texttt{(send o g x 42)}\\
-\hline
-
-    \texttt{field-id}
- & Swaps two field identifiers.
-    &  \texttt{(get-field o f)} $\rightarrow$  \texttt{(get-field o g)}\\
-\hline
-
-     \texttt{class:init}
-  & Swaps values of class initializers.
-   &  \begin{tabular}[t]{@{}l}
-     \texttt{(new c [a 5] [b "hello"])} $\rightarrow$\\
-     \texttt{(new c [a "hello"] [b 5])}\\
-    \end{tabular}\\
-\hline
-
-     \texttt{class:parent}
-    & Replaces the parent of classes with \texttt{object\%}.
-    &  \begin{tabular}[t]{@{}l}
-     \texttt{(class foo\% (super-new))} $\rightarrow$\\
-     \texttt{(class object\% (super-new))}\\
-    \end{tabular}\\
-\hline
-
-     \texttt{class:public}
-    & Makes a public method private and vice versa.
-    & \begin{tabular}[t]{@{}l}
-      \texttt{(class object (define/public (m x) x))} $\rightarrow$\\
-      \texttt{(class object (define/private (m x) x))}\\
-    \end{tabular}\\
-\hline
-
-     \texttt{class:super}
-    & Removes \texttt{super-new} calls from class definitions.
-    & \begin{tabular}[t]{@{}l}
-      \texttt{(class foo\% (super-new))} $\rightarrow$ \\
-      \texttt{(class foo\% (void))}\\
-       \end{tabular}\\
-\hline
-
-     \texttt{arithmetic}
-& Swaps arithmetic operators.
-    & \texttt{+} $\rightarrow$ \texttt{-} \\
-\hline
-
-     \texttt{boolean}
-& Swaps \texttt{and} and \texttt{or}.
-& \texttt{and} $\rightarrow$ \texttt{or} \\
-\hline
-
-     \texttt{negate-cond}
-    & Negates conditional test expressions.
-    &  \texttt{(if (= x 0) t e)} $\rightarrow$ \texttt{(if (not (= x 0)) t e)}\\
-\hline
-
-     \texttt{force-cond}
-    & Replaces conditional test expressions with \texttt{\#t}.
-    & \texttt{(if (= x 0) t e)} $\rightarrow$ \texttt{(if \#t t e)}
-\end{tabular}
-  \caption{Summary of mutators.}
-  \label{table:mutation-ops}
-\end{figure*}
-
-\subsection{Are these mutators any good?}
-
-The reader may have observed that mutators that produce 
-type-level mistakes do not necessarily turn our benchmarks into
-interesting debugging scenarios for the rational programmer.  The
-benchmarks are fully typed, and type checking detects each mistake at
-compile time.  Hence, we apply the mutators to versions, dubbed
-dubbed{scenarios}, of the benchmarks without 
-some  of their type annotations.  Without the annotations, the type
-checker may not detect a mistake at compile time and, instead, the run
-time checks
-of the gradual type system signal an error.  Scenarios
-with type-level mistakes that result in run time errors are the ideal
-ground to exercise the rational programmer's ability to locate errors.
-
-Specifically, we define an \emph{interesting debugging scenario} as a
-scenario that
-\begin{enumerate} 
-  \item is rejected by the type checker when we restore all its missing type annotations, 
-  \item produces a run time error under Erasure as-is, and 
-  \item the stack trace of the error mentions at least three distinct components.  
-\end{enumerate}
-
-The first criterion simply validates that the debugging scenario concerns
-an actual type-level mistake. The second one stipulates that the mistake is
-detectable under Erasure to ensure that it is detectable by all three
-semantics.   Whether a scenario produces a run time error largely depends on 
-the underlying semantics but \citet{gfd-oopsla-2019}
-show that a run time error under Erasure implies a run time
-error under both Transient and Natural. Thus the criterion aligns
-interesting debugging scenarios with our lowest denominator, Erasure, 
-and excludes  from consideration the very interesting scenarios indeed where Natural or
-Transient can detect a mistake but where Erasure is powerless. Of course this choice favors
-Erasure over Transient and Natural and, for the same reason, Transient over
-Natural. We consider some form of bias towards one or the other semantics
-unavoidable no matter how we define interesting debugging scenarios and we
-opt for tipping the scales in favor of the theoretically weakest
-semantics. We further discuss the implications of this choice in
-section~\ref{sec:discussion}.
-
-The third criterion aims to capture debugging scenarios with a specific
-kind of run time error: one produced because of the interaction between the buggy
-portion of the program with the rest of the program.  In these cases, the rational programmer
-may need to examine several pieces of the program to locate the source of the faulty interaction.
-In technical terms, we reliably mark these scenarios by
-analyzing the stack trace of a run time error.  If the stack trace mentions at
-least two distinct components, then this indicates interaction.  We require a third component because our benchmarks come
-with a driver component which is present in the stack trace of all run
-time errors. In essence, this last criterion excludes a large number of
-trivial to debug, and thus ineffective, scenarios that error immediately when their mutated component is
-evaluated.
-
-%\begin{figure*}
-%  \centering
-%  \includegraphics[scale=0.35]{./plots/mutant-breakdown}
-%  \caption{Interesting debugging scenarios produced by our mutators. Counts are cut off at 50.}
-%  \label{fig:mutant-breakdown}
-%\end{figure*}
-
-With the definition of interesting debugging scenarios in hand, we analyze how effective our mutators
-are for turning the benchmarks into interesting debugging scenarios.
-At a high level, the mutators produce 16,800 mutants with at least one interesting debugging scenario each across all of our benchmarks.
-Broken down by benchmark, the mutators produce at least 40 interesting
-scenarios for every benchmark, and these scenarios originate from at
-least four different mutators per benchmark.  Thus the mutators result in 
-a sizeable and diverse population of scenarios for every benchmark.
-Furthermore, every mutator contributes scenarios to at least one
-benchmark.
-That being said, some mutators only apply to a small subset of the benchmarks;
-those mutators target specific features which only those benchmarks make us of, so we include them despite their narrow applicability.
-For instance, the class-focused mutators are mainly effective in \texttt{take5}, since it is the benchmark with the most extensive use of object-oriented features.
+\sub{benchmarks} {Which Software to Use As Benchmarks}
+\sub{interesting}{Are These Mutators Interesting?}

### 2021-02-22 writing/paper.tex
Lots of minor changes and rewordings
organize sections/titles/labels in paper
factor mutate

@@
 
-\input{introduction.tex}    \label{sec:introduction}
-\input{challenges.tex}      \label{sec:challenges}
-\input{landscape.tex}       \label{sec:landscape}
-\input{mutate.tex}          \label{sec:mutate}
-\input{rational.tex}	      \label{sec:rational}
-\input{sample.tex}          \label{sec:sample}
-\input{results.tex}         \label{sec:results}
-\input{discussion.tex}      \label{sec:discussion}
-\input{related.tex}         \label{sec:related}
-\input{conclusion.tex}      \label{sec:conclusion}
+%% -----------------------------------------------------------------------------
+\def\sec#1#2{\section{#2} \label{sec:#1} \input{#1.tex}}
+
+\sec{introduction}{Does Blame Matter}
+\sec{challenges}  {Why Is It Hard to Evaluate Blame}
+\sec{landscape}   {How to Think About All Three Systems}
+\sec{mutate}      {How to Make Lots of Mistakes}
+\sec{rational}    {How to Test the Rational Programmer}
+\sec{sample}      {How to Sample the Experimental Space}
+\sec{results}     {What We Discovered}
+\sec{discussion}  {What We Learned}
+\sec{related}     {Who has Worked on this Problem Before}
+\sec{conclusion}  {What to Do Next}
+%% -----------------------------------------------------------------------------
 

### 2021-02-22 writing/mutate-benchmarks.tex
factor mutate

@@
 \end{figure}
-
-Of course, the gradual typing benchmarks have no mistakes for the rational
-programmer to debug. Therefore, we follow ~\citet{lksfd-popl-2020} and we
-inject bugs with mutation
-analysis~\cite{jia2011analysis,demillo1978hints,lipton1971fault}.  A
-mutation is a local syntactic change of the code that may
-produce a bug. Usually, the outcome of mutating a program once, dubbed a
-mutant, can be distinguished from the original program with a test case.
-The better the test suite of a program, the more mutants it can ``kill.''
-Mutants that are hard to ``kill'' usually correspond to subtle programming
-errors, and mutation frameworks typically come with standard mutators
-carefully designed to create such mutants.
-
-Unlike the hard-to-kill mistakes introduced by standard mutators, 
-gradual type systems catch shallow errors.  Thus, in order to
-exercise the rational programmer in the context of gradual typing, 
-we need to develop alternative mutators
-that are fine-tuned to produce the type-level mistakes gradual type systems 
-can detect. Our inspiration is our
-more-than-a-decade long experience of making type-level mistakes in Typed
-Racket. Some of these mistakes often take non-trivial effort to debug and
-have become the seed for our mutators, which we summarize in
-figure~\ref{table:mutation-ops}.  For each one, the figure provides a
-short description and an example mutation.
-
-Most of our mutators are self-explanatory.
-The first twelve apply to most gradually typed languages, including those with classes.
-The last four mutators target distinguishing features of Typed Racket.
-Specifically, \texttt{arithmetic} may replace a \texttt{+} with a \texttt{-}
-in an attempt to change the type of the result of the arithmetic
-operation. In Typed Racket, \texttt{+}'s result is a
-\texttt{Positive-Integer} when all arguments are
-positive integers. However, the result of \texttt{-} is
-\texttt{Integer}. In the same spirit, \texttt{boolean} aims to take
-advantage of Typed Racket's ``truthiness.'' Finally, \texttt{negate-cond} and \texttt{force-cond}
-attempt to confuse occurrence typing.

### 2021-02-22 writing/mutate-mutators.tex
factor mutate

Of course, the gradual typing benchmarks have no mistakes for the rational
programmer to debug. Therefore, we follow ~\citet{lksfd-popl-2020} and we
inject bugs with mutation
analysis~\cite{jia2011analysis,demillo1978hints,lipton1971fault}.  A
mutation is a local syntactic change of the code that may
produce a bug. Usually, the outcome of mutating a program once, dubbed a
mutant, can be distinguished from the original program with a test case.
The better the test suite of a program, the more mutants it can ``kill.''
Mutants that are hard to ``kill'' usually correspond to subtle programming
errors, and mutation frameworks typically come with standard mutators
carefully designed to create such mutants.

Unlike the hard-to-kill mistakes introduced by standard mutators, 
gradual type systems catch shallow errors.  Thus, in order to
exercise the rational programmer in the context of gradual typing, 
we need to develop alternative mutators
that are fine-tuned to produce the type-level mistakes gradual type systems 
can detect. Our inspiration is our
more-than-a-decade long experience of making type-level mistakes in Typed
Racket. Some of these mistakes often take non-trivial effort to debug and
have become the seed for our mutators, which we summarize in
figure~\ref{table:mutation-ops}.  For each one, the figure provides a
short description and an example mutation.

Most of our mutators are self-explanatory.
The first twelve apply to most gradually typed languages, including those with classes.
The last four mutators target distinguishing features of Typed Racket.
Specifically, \texttt{arithmetic} may replace a \texttt{+} with a \texttt{-}
in an attempt to change the type of the result of the arithmetic
operation. In Typed Racket, \texttt{+}'s result is a
\texttt{Positive-Integer} when all arguments are
positive integers. However, the result of \texttt{-} is
\texttt{Integer}. In the same spirit, \texttt{boolean} aims to take
advantage of Typed Racket's ``truthiness.'' Finally, \texttt{negate-cond} and \texttt{force-cond}
attempt to confuse occurrence typing.

### 2021-02-22 writing/landscape.tex
space war

@@
 
-All three systems used here are implemented on top of the common basis of
-Racket. The use the Typed Racket syntax and type checker for all three
-semantics. The Natural semantics is inherited as is, while the Erasure semantics
-is the one of plain Racket.  The Transient system implements the design
-of~\citet{vss-popl-2017}, though constructing it demands two design decisions:
-which shallow checks to use and where, plus how to maintain a blame map in a
-full-fledged run-time system.
+All three systems used for the experiment are variants of Racket. The
+use the Typed Racket syntax and type checker for all three
+semantics. The Natural semantics is inherited as is, while the Erasure
+semantics is the one of plain Racket.  The Transient system implements
+the design of~\citet{vss-popl-2017}, though constructing it demands
+two design decisions: which shallow checks to use and where, plus how
+to maintain a blame map in a full-fledged run-time system.
 
@@
 
-Once the type checker has completed its work, Transient translates the source
-into an alternative form and in the process inserts run-time checks. Technically
-speaking, the rewriting pass wraps all elimination in applications of shallow
-checks.  For example, if the variable {\tt f} has the static type {\tt (->
-String String)}, the application {\tt (f x)} must be rewritten to check for a
+Once the type checker has completed its work, Transient rewrites the
+source to insert run-time checks. Technically speaking, the rewriting
+pass wraps all elimination in applications of shallow checks.  For
+example, if the variable {\tt f} has the static type {\tt (-> String
+String)}, the application {\tt (f x)} must be rewritten to check for a
 string result because {\tt f} could be an untyped function.

### 2021-02-23 writing/mutate-benchmarks-figure.tex
factor out figures

--- 2021-02-22 writing/mutate-benchmarks.tex
@@
-The starting point for our corpus of programs
-is~\citet{gtnffvf-jfp-2019}'s gradual typing benchmark suite. The
-benchmark suite consists of fully typed correct programs that are written by different authors
-and have been used, maintained and evolved by their authors and others over a
-number of years.  They range widely in the size, complexity, purpose and
-features of Typed Racket they employ.  Furthermore the benchmarks use advanced aspects
-of Typed Racket's type system such as occurrence
-typing~\cite{tf-icfp-2010}, types for mutable and immutable data
-structures, polymorphic types, types for first-class classes and objects, and types for
-Racket's numeric tower~\cite{stathff-padl-12}. Without loss of the diversity of the benchmarks,
-we select the ten largest in terms of numbers of components (between 6 and 14 components).
-These benchmarks also have the most complex dependency graphs and, hence,
-can make the debugging process the hardest for the rational programmer.
-The table in figure~\ref{table:benchmark-descriptions} shows our selection
-together with a short description for each benchmark.
-
 \begin{figure}

### 2021-02-23 writing/mutate-benchmarks.tex
factor out figures

@@
-The starting point for our corpus of programs
-is~\citet{gtnffvf-jfp-2019}'s gradual typing benchmark suite. The
-benchmark suite consists of fully typed correct programs that are written by different authors
-and have been used, maintained and evolved by their authors and others over a
-number of years.  They range widely in the size, complexity, purpose and
-features of Typed Racket they employ.  Furthermore the benchmarks use advanced aspects
-of Typed Racket's type system such as occurrence
-typing~\cite{tf-icfp-2010}, types for mutable and immutable data
-structures, polymorphic types, types for first-class classes and objects, and types for
-Racket's numeric tower~\cite{stathff-padl-12}. Without loss of the diversity of the benchmarks,
-we select the ten largest in terms of numbers of components (between 6 and 14 components).
-These benchmarks also have the most complex dependency graphs and, hence,
-can make the debugging process the hardest for the rational programmer.
-The table in figure~\ref{table:benchmark-descriptions} shows our selection
-together with a short description for each benchmark.
+Over the past decade, \citet{gtnffvf-jfp-2019} have curated a collection of
+Typed Racket programs for systematically measuring the implementation's
+performance. The benchmark suite consists of fully typed, correct programs,
+written by a number of different authors, who have maintained and evolved these
+programs over several years. The programs range widely in size, complexity,
+purpose, and in programming style. The latter is mostly a reference to their
+reliance on many Typed Racket features: occurrence typing~\cite{tf-icfp-2010},
+types for mutable and immutable data structures~\cite{hpst-sfp-2010}, types for
+first-class classes and objects~\cite{tsdtf-oopsla-2012}, and types for Racket's
+numeric tower~\cite{stathff-padl-12}.
 
-\begin{figure}
-\begin{tabular}{p{1.75cm} | p{6cm} }
-  {\bf  name} & {\bf description (author)}  \\
-
-\hline
-  \texttt{acquire} & Object-oriented board game implementation (M. Felleisen)  \\%[1em]
-
-\hline
-  \texttt{gregor} & Utilities for calendar dates (J. Zeppieri) \\%[1em]
-
-\hline
-  \texttt{kcfa} & Functional implementation of 2CFA for the lambda calculus (M. Might) \\%[1em]
-
-\hline
-  \texttt{quadT} & Converter from S-expression source code to PDF format (M. Butterick)\\%[1em]
-
-\hline
-  \texttt{quadU} & Converter from S-expression source code to PDF format  (B. Greenman) \\%[1em]
-
-\hline
-  \texttt{snake} & Functional implementation of the  Snake video game (D. Van Horn) \\%[1em]
-
-\hline
-  \texttt{suffixtree} & Algorithm for common longest subsequences between strings. (D. Yoo) \\%[1em]
-
-\hline
-  \texttt{synth} & Converter of notes and drum beats to WAV (V. St-Amour \& N. Toronto) \\%[1em]
-
-\hline
-  \texttt{take5} & Mixin-based card game simulator (M.Felleisen)  \\%[1em]
-
-\hline
-  \texttt{tetris} & Functional implementation of Tetris (D. Van Horn) \\%[1em]
-
-\end{tabular}
-  \caption{Benchmarks summary.}
-  \label{table:benchmark-descriptions}
-\end{figure}
+Since mixing of typed and untyped code in Typed Racket takes place at the module
+level, the experiment calls for benchmarks with a decent number of modules and a
+variety of module dependency graphs. Filtering the benchmark suite using these
+criteria yields ten suitable programs. Figure~\ref{table:benchmark-descriptions}
+displays the selected benchmark programs and describes their basic
+characteristic.

### 2021-02-23 writing/mutate-interesting.tex
factor out figures

@@
-The reader may have observed that mutators that produce 
-type-level mistakes do not necessarily turn our benchmarks into
-interesting debugging scenarios for the rational programmer.  The
-benchmarks are fully typed, and type checking detects each mistake at
-compile time.  Hence, we apply the mutators to versions, dubbed
-dubbed{scenarios}, of the benchmarks without 
-some  of their type annotations.  Without the annotations, the type
-checker may not detect a mistake at compile time and, instead, the run
-time checks
-of the gradual type system signal an error.  Scenarios
-with type-level mistakes that result in run time errors are the ideal
-ground to exercise the rational programmer's ability to locate errors.
+Mutators that produce type-level mistakes do not necessarily turn
+benchmarks into interesting debugging scenarios.
 
-\begin{figure*}
-  \begin{tabular}{p{2.1cm} | p{7cm}  | p{7.5cm} }
-    {\bf name} & {\bf description} & {\bf example} \\
-\hline
-
-    \texttt{constant}
-& Swaps a constant with another of the same value but different type.
-& \texttt{5.6} $\rightarrow$ \texttt{5.6+0.0i} \\
-\hline
-
-    \texttt{deletion}
-    & Deletes the result expression of a sequence.
-    & \texttt{(begin x y z)} $\rightarrow$ \texttt{(begin x y)} \\
-\hline
-
-   \texttt{position}
- & Swaps two subexpressions.
-  &  \texttt{(f a 42 "b" 0)} $\rightarrow$\texttt{(f a 42 0 "b")} \\
-\hline
-
-     \texttt{list}
-    & Replaces \texttt{append} with \texttt{cons}.
-    &  \begin{tabular}[t]{@{}l}
-      \texttt{append} $\rightarrow$ \texttt{cons}
-    \end{tabular}\\
-\hline
-
-    \texttt{top-level-id}
- & Swaps identifiers defined in the same module.
- &\texttt{(f x 42)} $\rightarrow$ \texttt{(g x 42)}\\
-\hline
-
-    \texttt{imported-id}
- & Swaps identifiers imported from the same module.
- &\texttt{(f x 42)} $\rightarrow$ \texttt{(g x 42)}\\
-\hline
-
-    \texttt{method-id}
- & Swaps two method identifiers.
- &\texttt{(send o f x 42)} $\rightarrow$ \texttt{(send o g x 42)}\\
-\hline
-
-    \texttt{field-id}
- & Swaps two field identifiers.
-    &  \texttt{(get-field o f)} $\rightarrow$  \texttt{(get-field o g)}\\
-\hline
-
-     \texttt{class:init}
-  & Swaps values of class initializers.
-   &  \begin{tabular}[t]{@{}l}
-     \texttt{(new c [a 5] [b "hello"])} $\rightarrow$\\
-     \texttt{(new c [a "hello"] [b 5])}\\
-    \end{tabular}\\
-\hline
-
-     \texttt{class:parent}
-    & Replaces the parent of classes with \texttt{object\%}.
-    &  \begin{tabular}[t]{@{}l}
-     \texttt{(class foo\% (super-new))} $\rightarrow$\\
-     \texttt{(class object\% (super-new))}\\
-    \end{tabular}\\
-\hline
-
-     \texttt{class:public}
-    & Makes a public method private and vice versa.
-    & \begin{tabular}[t]{@{}l}
-      \texttt{(class object (define/public (m x) x))} $\rightarrow$\\
-      \texttt{(class object (define/private (m x) x))}\\
-    \end{tabular}\\
-\hline
-
-     \texttt{class:super}
-    & Removes \texttt{super-new} calls from class definitions.
-    & \begin{tabular}[t]{@{}l}
-      \texttt{(class foo\% (super-new))} $\rightarrow$ \\
-      \texttt{(class foo\% (void))}\\
-       \end{tabular}\\
-\hline
-
-     \texttt{arithmetic}
-& Swaps arithmetic operators.
-    & \texttt{+} $\rightarrow$ \texttt{-} \\
-\hline
-
-     \texttt{boolean}
-& Swaps \texttt{and} and \texttt{or}.
-& \texttt{and} $\rightarrow$ \texttt{or} \\
-\hline
-
-     \texttt{negate-cond}
-    & Negates conditional test expressions.
-    &  \texttt{(if (= x 0) t e)} $\rightarrow$ \texttt{(if (not (= x 0)) t e)}\\
-\hline
-
-     \texttt{force-cond}
-    & Replaces conditional test expressions with \texttt{\#t}.
-    & \texttt{(if (= x 0) t e)} $\rightarrow$ \texttt{(if \#t t e)}
-\end{tabular}
-  \caption{Summary of mutators.}
-  \label{table:mutation-ops}
-\end{figure*}
+The benchmarks are fully typed, and type checking detects each mistake
+at compile time.  Hence, we apply the mutators to versions, dubbed
+dubbed{scenarios}, of the benchmarks without some of their type
+annotations.  Without the annotations, the type checker may not detect
+a mistake at compile time and, instead, the run time checks of the
+gradual type system signal an error.  Scenarios with type-level
+mistakes that result in run time errors are the ideal ground to
+exercise the rational programmer's ability to locate errors.
 

### 2021-02-23 writing/mutate-mutators.tex
factor out figures

@@
-Of course, the gradual typing benchmarks have no mistakes for the rational
-programmer to debug. Therefore, we follow ~\citet{lksfd-popl-2020} and we
-inject bugs with mutation
-analysis~\cite{jia2011analysis,demillo1978hints,lipton1971fault}.  A
-mutation is a local syntactic change of the code that may
-produce a bug. Usually, the outcome of mutating a program once, dubbed a
-mutant, can be distinguished from the original program with a test case.
-The better the test suite of a program, the more mutants it can ``kill.''
-Mutants that are hard to ``kill'' usually correspond to subtle programming
-errors, and mutation frameworks typically come with standard mutators
-carefully designed to create such mutants.
+Generating debugging scenarios from the benchmarks demands the design
+of novel mutators. Generally speaking, a
+mutator~\cite{jia2011analysis,demillo1978hints,lipton1971fault}
+performs a localized syntactic change to a code basis. For the
+evaluation of a blame strategy, the mutator must produce type-level
+mistakes that the run-time checks of gradual typing systems or the
+safety checks of the underlying language can detect. Once detected,
+the rational programmer can attempt to detect the mistake by gradually
+adding types to blamed modules.
 
-Unlike the hard-to-kill mistakes introduced by standard mutators, 
-gradual type systems catch shallow errors.  Thus, in order to
-exercise the rational programmer in the context of gradual typing, 
-we need to develop alternative mutators
-that are fine-tuned to produce the type-level mistakes gradual type systems 
-can detect. Our inspiration is our
-more-than-a-decade long experience of making type-level mistakes in Typed
-Racket. Some of these mistakes often take non-trivial effort to debug and
-have become the seed for our mutators, which we summarize in
-figure~\ref{table:mutation-ops}.  For each one, the figure provides a
-short description and an example mutation.
+Figure~\ref{table:mutation-ops} describes the sixteen mutators used
+for the data reported here. They are inspired by the authors
+decades-long experience of making type-level mistakes in Typed Racket,
+some of which take a non-trivial effort to debug.  For each mutaror,
+the figure provides a short description and an example mutation.
 
-Most of our mutators are self-explanatory.
-The first twelve apply to most gradually typed languages, including those with classes.
-The last four mutators target distinguishing features of Typed Racket.
-Specifically, \texttt{arithmetic} may replace a \texttt{+} with a \texttt{-}
-in an attempt to change the type of the result of the arithmetic
-operation. In Typed Racket, \texttt{+}'s result is a
-\texttt{Positive-Integer} when all arguments are
-positive integers. However, the result of \texttt{-} is
-\texttt{Integer}. In the same spirit, \texttt{boolean} aims to take
-advantage of Typed Racket's ``truthiness.'' Finally, \texttt{negate-cond} and \texttt{force-cond}
-attempt to confuse occurrence typing.
+Most of these mutators are self-explanatory.  The first twelve apply
+to most gradually typed languages, including those with classes.  The
+last four target distinguishing features of Typed Racket.
+Specifically, {\tt arithmetic} may replace a {\tt +} with a {\tt -} in
+an attempt to change the type of the result of the arithmetic
+operation. In Typed Racket, {\tt +}'s result is a {\tt
+Positive-Integer} when all arguments are positive
+integers~\cite{stathff-padl-12}. However, the result of {\tt -}
+is {\tt Integer}. In the same spirit, the {\tt boolean} mutator aims
+to take advantage of Typed Racket faithfulness to Racket's
+``falsiness.'' Finally, {\tt negate-cond} and {\tt force-cond} attempt
+to confuse occurrence typing.

### 2021-02-23 writing/mutate.tex
factor mutate
factor out figures

@@
 As section~\ref{sec:challenges} explains, the first challenge is to
-generate a representative mistakes in representative software. The
-following two subsections explain in reverse order.
+generate a representative mistakes in representative software. This
+section explains the specific choices made for the first application
+of the method: the source and nature of the chosen benchmarks
+(sec.~\ref{sub:mutate-benchmarks}), the newly developed mutators
+(sec.~\ref{sub:mutate-mutators}), and why they are suitable for this
+experiment (sec.~\ref{sub:mutate-interesting}).
 
@@
 
+\input{mutate-benchmarks-figure.tex}
 \sub{benchmarks} {Which Software to Use As Benchmarks}
-\sub{interesting}{Are These Mutators Interesting?}
+\sub{mutators}   {How to Mutate the Benchmarks} 
+\input{mutate-mutators-figure.tex}
+\sub{interesting}{Are These Mutators Interesting}

### 2021-02-23 writing/mutate-interesting.tex
proper voice and style

@@
-Mutators that produce type-level mistakes do not necessarily turn
-benchmarks into interesting debugging scenarios.
+All mutators produce mutations with type-level mistakes, but not all such
+mutations necessarily represent interesting debugging scenarios. Running a
+mutation may not raise a run-time error, though produce the wrong
+outputs---something the rational programmer may not recognize and certainly
+cannot use for a debugging session. Even if the program execution ends in an
+exceptional state, it may do so too quickly; for example, the first-order export
+checks of Natural may catch the error and point directly to the faulty module.
 
-The benchmarks are fully typed, and type checking detects each mistake
-at compile time.  Hence, we apply the mutators to versions, dubbed
-dubbed{scenarios}, of the benchmarks without some of their type
-annotations.  Without the annotations, the type checker may not detect
-a mistake at compile time and, instead, the run time checks of the
-gradual type system signal an error.  Scenarios with type-level
-mistakes that result in run time errors are the ideal ground to
-exercise the rational programmer's ability to locate errors.
+To avoid these pitfalls, an \emph{interesting debugging scenario} is defined as
+a mutation that satisfies three conditions.
 
-Specifically, we define an \emph{interesting debugging scenario} as a
-scenario that
-\begin{enumerate} 
-  \item is rejected by the type checker when we restore all its missing type annotations, 
-  \item produces a run time error under Erasure as-is, and 
-  \item the stack trace of the error mentions at least three distinct components.  
-\end{enumerate}
+First, {\em the mutation is rejected by the type checker when all missing type
+annotations are restored\/}. This first criterion validates that the debugging
+scenario concerns an actual type-level mistake.
 
-The first criterion simply validates that the debugging scenario concerns
-an actual type-level mistake. The second one stipulates that the mistake is
-detectable under Erasure to ensure that it is detectable by all three
-semantics.   Whether a scenario produces a run time error largely depends on 
-the underlying semantics but \citet{gfd-oopsla-2019}
-show that a run time error under Erasure implies a run time
-error under both Transient and Natural. Thus the criterion aligns
-interesting debugging scenarios with our lowest denominator, Erasure, 
-and excludes  from consideration the very interesting scenarios indeed where Natural or
-Transient can detect a mistake but where Erasure is powerless. Of course this choice favors
-Erasure over Transient and Natural and, for the same reason, Transient over
-Natural. We consider some form of bias towards one or the other semantics
-unavoidable no matter how we define interesting debugging scenarios and we
-opt for tipping the scales in favor of the theoretically weakest
-semantics. We further discuss the implications of this choice in
-section~\ref{sec:discussion}.
+Second, {\em it produces a run-time exception under Erasure as-is\/}, which
+ensures that it is detectable under all three semantics.  Whether a scenario
+produces a run time error largely depends on the underlying semantics but
+\citet{gfd-oopsla-2019} show that a run-time error under Erasure implies a run
+time error under both Transient and Natural. Thus the criterion aligns
+interesting debugging scenarios with the lowest denominator, Erasure. It
+excludes from consideration the most complex scenarios where only Natural or
+Transient can detect a mistake and Erasure is powerless.
 
-The third criterion aims to capture debugging scenarios with a specific
-kind of run time error: one produced because of the interaction between the buggy
-portion of the program with the rest of the program.  In these cases, the rational programmer
-may need to examine several pieces of the program to locate the source of the faulty interaction.
-In technical terms, we reliably mark these scenarios by
-analyzing the stack trace of a run time error.  If the stack trace mentions at
-least two distinct components, then this indicates interaction.  We require a third component because our benchmarks come
-with a driver component which is present in the stack trace of all run
-time errors. In essence, this last criterion excludes a large number of
-trivial to debug, and thus ineffective, scenarios that error immediately when their mutated component is
-evaluated.
+{\bf Note} While this choice favors Erasure over Transient and Natural and, for
+the same reason, Transient over Natural, some form of bias towards one or the
+other semantics is unavoidable. Tipping the scales in favor of the theoretically
+weakest semantics yields the most stable results. See
+section~\ref{sec:discussion} for some further discussion of this choice.
+
+Third, {\em the exception's stack trace mentions at least three distinct
+components\/}, meaning the scenario demands a sufficiently sophisticated
+debugging effort: one due to the interaction between the buggy portion of the
+program with the rest of the program.  In these cases, the rational programmer
+may need to examine several modules of the program to locate the source of the
+faulty interaction.  In technical terms, if the stack trace mentions at least
+two distinct components, the bug should be revealed by the run-time checks
+governing an interaction. A third component is required because all benchmark
+programs come with a driver component, which is present in the stack trace of
+all run-time errors. In essence, this last criterion excludes a large number of
+trivial-to-debug scenarios that error immediately when their mutated component
+is evaluated.
 
@@
 
-With the definition of interesting debugging scenarios in hand, we analyze how effective our mutators
-are for turning the benchmarks into interesting debugging scenarios.
-At a high level, the mutators produce 16,800 mutants with at least one interesting debugging scenario each across all of our benchmarks.
-Broken down by benchmark, the mutators produce at least 40 interesting
-scenarios for every benchmark, and these scenarios originate from at
-least four different mutators per benchmark.  Thus the mutators result in 
-a sizeable and diverse population of scenarios for every benchmark.
-Furthermore, every mutator contributes scenarios to at least one
-benchmark.
-That being said, some mutators only apply to a small subset of the benchmarks;
-those mutators target specific features which only those benchmarks make us of, so we include them despite their narrow applicability.
-For instance, the class-focused mutators are mainly effective in \texttt{take5}, since it is the benchmark with the most extensive use of object-oriented features.
+The definition of interesting debugging scenarios creates a powerful filter. At
+a high level, the mutators produce 16,800 mutants with at least one interesting
+debugging scenario each across all of benchmarks.  Broken down by benchmark, the
+mutators produce at least 40 interesting scenarios for every benchmark, and
+these scenarios originate from at least four different mutators per benchmark.
+Thus, the mutators result in a sizable and diverse population of scenarios for
+every benchmark.  Furthermore, every mutator contributes scenarios to at least
+one benchmark.  Note, though, that some mutators apply only to a few benchmarks,
+because they target rather specific features.  For instance, the class-focused
+mutators are mainly effective in \texttt{take5}, because it makes the most
+extensive use of object-oriented Racket features.

### 2021-02-23 writing/mutate-interesting.tex
space war

@@
 
-To avoid these pitfalls, an \emph{interesting debugging scenario} is defined as
-a mutation that satisfies three conditions.
-
-First, {\em the mutation is rejected by the type checker when all missing type
-annotations are restored\/}. This first criterion validates that the debugging
-scenario concerns an actual type-level mistake.
+To avoid these pitfalls, an \emph{interesting debugging scenario} is
+defined as a mutation that satisfies three conditions.  First, {\em
+the mutation is rejected by the type checker when all missing type
+annotations are restored\/}. This first criterion validates that the
+debugging scenario concerns an impedance mismatch.
 
@@
 
-Third, {\em the exception's stack trace mentions at least three distinct
-components\/}, meaning the scenario demands a sufficiently sophisticated
-debugging effort: one due to the interaction between the buggy portion of the
-program with the rest of the program.  In these cases, the rational programmer
-may need to examine several modules of the program to locate the source of the
-faulty interaction.  In technical terms, if the stack trace mentions at least
-two distinct components, the bug should be revealed by the run-time checks
-governing an interaction. A third component is required because all benchmark
-programs come with a driver component, which is present in the stack trace of
-all run-time errors. In essence, this last criterion excludes a large number of
-trivial-to-debug scenarios that error immediately when their mutated component
-is evaluated.
+Third, {\em the exception's stack trace mentions at least three
+distinct components\/}, meaning the scenario demands a sufficiently
+sophisticated debugging effort: one due to the interaction between the
+buggy portion of the program with the rest of the program.  In these
+cases, the rational programmer must examine at least modules to locate
+the source of the faulty interaction.  In technical terms, if the
+stack trace mentions at least two distinct components, the bug should
+be revealed by the run-time checks governing an interaction. A third
+component is required because all benchmark programs include a driver
+component, which is guaranteed to be in the stack trace. In essence,
+this last criterion excludes a large number of trivial-to-debug
+scenarios that error immediately when their mutated component is
+evaluated.
 

### 2021-02-23 writing/rational.tex
small fix
organize sections/titles/labels in paper
simple 'we' cleanup

@@
-\section{How to Test the Rational Programmer} 
-\label{sec:rational}
-
 Code migration is the raison d'\^etre of Typed Racket, which is why the community
@@
 
+%% MF: I don't think we should honor the nonsense of ``true gt.''  We
+% are it as much as anyone!
+% 
+% \footnote{The method can be adapted to so-called true gradual typing
+% systems~\cite{svcb-snapl-2015} in a straightforward manner.}
+
 Based on this idea, migratory paths offer a concise setting for
-evaluating the effectiveness of a rational programmer for dealing with the
-interesting debugging scenarios  from section ~\ref{sec:mutate}.
-If we group debugging scenarios by mutation, then we obtain a set 
-of programs that all contain the same mistake but differ in their type
-annotations.  In the setting of migratory typing,\footnote{The method can be adapted to
-so-called true gradual typing systems~\cite{svcb-snapl-2015} in a
-straightforward manner.} each scenario differs from another in which
-components are typed. Formally, a program $\system$ is a set of
-components $\set{\component}$, and a scenario $\conf$ is the typed subset of
-$\set{\component}$ . Hence, the scenarios of $\system$ are ordered by the
-subset relation and, naturally,  form a lattice $\lattice{\system}$ with
-$2^{\size{\system}}$ elements, like
-the one introduced by ~\citet{tfgnvf-popl-2016} for performance
-evaluations. The bottom scenario of
-$\lattice{\system}$ is always $\emptyset$ and the top one is $\system$
-itself. In between are all the mix-typed scenarios.
+evaluating the effectiveness of a rational programmer for dealing with
+the interesting debugging scenarios from section ~\ref{sec:mutate}.
+Grouping debugging scenarios by mutation yields a set of programs that
+all contain the same mistake but differ in their type annotations.  In
+the setting of migratory typing, each scenario differs from another in
+which components are typed. Formally, a program $\system$ is a set of
+components $\set{\component}$, and a scenario $\conf$ is the typed
+subset of $\set{\component}$ . Hence, the scenarios of $\system$ are
+ordered by the subset relation and, naturally, form a lattice
+$\lattice{\system}$ with $2^{\size{\system}}$
+elements.\footnote{\citet{tfgnvf-popl-2016} use this lattice for
+performance evaluations.} The bottom scenario of $\lattice{\system}$
+is always $\emptyset$ and the top one is $\system$ itself. In between
+are all the mixed-typed scenarios.
 
-With these definitions, the actions of the rational programmer map to an
+In this context, the actions of the rational programmer map to an
 ascending chain---dubbed a \emph{trail}---in $\lattice{\system}$.  The
 starting scenario $\conf_0$ of a trail---also referred to as the
-\emph{root} (of the trail)---is the initial debugging scenario.  Herein we
-consider only roots that are interesting debugging scenarios as described
-in~\ref{sec:mutate} --- we expand on this point further on.  Intuitively,
-starting from such a root, the rational programmer  steps along a trail,
-and at each step, reduces the current scenario to an ``easier'' next
-scenario to debug.  The easiest scenarios are those that the type checker
-rejects outright, because a type error pinpoints an impedance mismatch
-between the broken code of one component and its typed interface to
-others. At this point, the programmer has identified the broken component.
-If the scenario type-checks, the programmers runs the program until it
-raises a run time error.  The rational programmer uses the error's blame
-assignment\footnote{Blame can result either directly from a run time type
-check or, in the absence of blame, from an exception that we track back to
-the component that causes it with the help of the stack trace.} to decide
-which component(s) to equip with types. This choice constructs scenario
-$\conf_{i+1}$ from $\conf_{i}$ and implies the creation of trail.
+\emph{root} (of the trail)---is the initial debugging scenario; only
+interesting (as described in~\ref{sec:mutate}) debugging scenarios
+serve as roots.  Intuitively, starting from such a root, the rational
+programmer steps along a trail, and at each step, reduces the current
+scenario to an ``easier'' next scenario to debug.  The easiest
+scenarios are those that the type checker rejects outright, because a
+type error pinpoints an impedance mismatch between the broken code of
+one component and its typed interface to others. At this point, the
+programmer has identified the broken component.  If the scenario
+type-checks, the programmers runs the program until it raises a run
+time error.  The rational programmer uses the error's blame
+assignment\footnote{Blame can result either directly from a run-time
+check or, in the absence of blame, from an exception that can be
+traced back to the component that causes it with the help of the stack
+trace.} to decide which component(s) to equip with types. This choice
+constructs scenario $\conf_{i+1}$ from $\conf_{i}$ and thus creates a trail.
 
@@
 
-The next three subsections specialize the trail-scenario idea to the
-various semantics of gradual typing. Once this machinery is in place,
-we use the fourth subsection to state the precise experimental
-research questions that correspond to the high-level question on the
-first page and to explain our experimental process. The final
-subsection extends the development with a discussion of programmer effort.
+The first three subsections specialize the trail-scenario idea to the
+three semantics of gradual typing. Once this machinery is in place,
+the fourth subsection presents a reformulation of the informal
+question on page~1 as precise experimental research questions. The
+final subsection adds a notion of programmer effort to the
+experimental setup.
 

### 2021-02-23 writing/rational-effort.tex
factor the rational programmer

--- 2021-02-23 writing/rational.tex
@@
-Code migration is the raison d'\^etre of Typed Racket, which is why the community
-uses ``migratory typing'' instead of gradual typing. Roughly speaking a
-programmer uses a gradually typed language to incrementally add types to a code
-base. From this perspective, a code base consists of some number of components,
-say modules, and a programmer converts a component at a time from its untyped
-state to a typed one.
-
-%% MF: I don't think we should honor the nonsense of ``true gt.''  We
-% are it as much as anyone!
-% 
-% \footnote{The method can be adapted to so-called true gradual typing
-% systems~\cite{svcb-snapl-2015} in a straightforward manner.}
-
-Based on this idea, migratory paths offer a concise setting for
-evaluating the effectiveness of a rational programmer for dealing with
-the interesting debugging scenarios from section ~\ref{sec:mutate}.
-Grouping debugging scenarios by mutation yields a set of programs that
-all contain the same mistake but differ in their type annotations.  In
-the setting of migratory typing, each scenario differs from another in
-which components are typed. Formally, a program $\system$ is a set of
-components $\set{\component}$, and a scenario $\conf$ is the typed
-subset of $\set{\component}$ . Hence, the scenarios of $\system$ are
-ordered by the subset relation and, naturally, form a lattice
-$\lattice{\system}$ with $2^{\size{\system}}$
-elements.\footnote{\citet{tfgnvf-popl-2016} use this lattice for
-performance evaluations.} The bottom scenario of $\lattice{\system}$
-is always $\emptyset$ and the top one is $\system$ itself. In between
-are all the mixed-typed scenarios.
-
-In this context, the actions of the rational programmer map to an
-ascending chain---dubbed a \emph{trail}---in $\lattice{\system}$.  The
-starting scenario $\conf_0$ of a trail---also referred to as the
-\emph{root} (of the trail)---is the initial debugging scenario; only
-interesting (as described in~\ref{sec:mutate}) debugging scenarios
-serve as roots.  Intuitively, starting from such a root, the rational
-programmer steps along a trail, and at each step, reduces the current
-scenario to an ``easier'' next scenario to debug.  The easiest
-scenarios are those that the type checker rejects outright, because a
-type error pinpoints an impedance mismatch between the broken code of
-one component and its typed interface to others. At this point, the
-programmer has identified the broken component.  If the scenario
-type-checks, the programmers runs the program until it raises a run
-time error.  The rational programmer uses the error's blame
-assignment\footnote{Blame can result either directly from a run-time
-check or, in the absence of blame, from an exception that can be
-traced back to the component that causes it with the help of the stack
-trace.} to decide which component(s) to equip with types. This choice
-constructs scenario $\conf_{i+1}$ from $\conf_{i}$ and thus creates a trail.
-
-% The successor of a scenario on a trail has more typed component;
-% formally, configuration $\conf_{i+1}$ on the trail differs from its
-% predecessor $\conf_{i}$ in that more components are typed.
-
-The first three subsections specialize the trail-scenario idea to the
-three semantics of gradual typing. Once this machinery is in place,
-the fourth subsection presents a reformulation of the informal
-question on page~1 as precise experimental research questions. The
-final subsection adds a notion of programmer effort to the
-experimental setup.
-
-%% -----------------------------------------------------------------------------
-\subsection{Modes of the Natural Rational Programmer} \label{sub:natural}
-
-The Natural semantics assigns blame to exactly one boundary.  A blame assignment
-has the following specific meaning: the typed component may make incorrect type
-assumptions about the untyped component in its interface, or the correct
-interface exposes a bug in the untyped component. Our setup rules out the first
-alternative (but see section~\ref{sec:conclusion}), and therefore the rational
-programmer extends the trail to a scenario that swaps out the untyped component
-for its typed counterpart.
-
-We can turn this description into a formal definition of the trail. 
-\begin{quote}
-\it A \emph{Natural blame trail} is a sequence of scenarios $\conf_0,...\conf_n$ of
-a program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
-\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =\{\blame{\system, \conf_i}\}$ where
-$\blame{\system, \conf}$ denotes the component that $\conf$ blames.
-\end{quote}
-
-When the buggy untyped component of a program is replaced by the typed counterpart,
-the type checker fails if this component causes the impedance mismatch. Hence, a
-trail that ends at an ill-typed scenarios successfully pinpoints the location of
-bugs. 
-\begin{quote}
-\it A Natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
-\emph{successful} iff its last scenario $\conf_n$ does not type check.  A Natural
-blame trail $\conf_0,..,\conf_n$ in a lattice $\lattice{\system}$ is \emph{failing}
-iff $\conf_n$ type checks and the trail cannot be extended further.
-\end{quote}
-That is, failing Natural blame trails are those that have not reached a scenario that
-reveals the bug at compile time and their last scenario does not blame an untyped component. Thus
-the rational programmer gets no hints from the gradual type system on how to
-continue the search for the bug.
-
-While a successful Natural blame trail indicates that it 
-pays off to heed blame assignments during debugging the trail's root, it does not answer whether
-blame is a critical piece of the rational programmer's process.  For instance,
-typing the top of stack trace of a failed run time type check, dubbed the
-location of the exception, might be as useful as typing the blamed one.
-
-To account for this situation, we devise a new mode of the Natural rational
-programmer that follows a migration process based on exceptions:
-\begin{quote}
-\it A Natural exception trail is a sequence of scenarios $\conf_0,...\conf_n$ of a
-program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
-\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i = \{\exception{\system, \conf_i}\}$
-where $\exception{\system, \conf}$ denotes the location of the exception of $\conf$.
-\end{quote}
-
-Same as for success for a Natural blame trail, we define success for a
-Natural exception trail.
-Putting the definitions for the two modes together, we can now compare the usefulness of blame 
-with that for mere exceptions for debugging a scenario in the context of Natural semantics.
-\begin{quote}
-\it 
-  Given a program $\system$ and a root $\conf_0$ in $\lattice{\system}$,
-  Natural blame is \emph{more useful} than Natural exceptions for
-  debugging $\conf_0$ iff 
-  the Natural blame trail 
-  that starts at $\conf_0$ is successful while the Natural exception trail that
-  starts at $\conf_0$ is failing.
-\end{quote}
-
-%% -----------------------------------------------------------------------------
-\subsection{Modes of the Transient Rational Programmer} \label{sub:transient}
-
-The transient semantics assigns blame to a set of components. In this setting, a
-blame assignment has the following meaning: the witness value to the impedance
-mismatch may have crossed the boundaries between elements in the set, and each crossing
-checked the value's type in a shallow manner.
-
-This ambiguity in Transient blame raises the question of how the rational programmer
-should react when the language produces a blame set. Our answer is that the rational
-programmer has at least two reasonable options. The first one is to select the component that is added
-to the blame set first and assign types to only that one---after all, if
-fully checked, the types of
-this first component should be able to detect an impedance mismatch earlier in the
-evaluation of a program than the later ones. The second option is to
-select the component that is added to the blame set last, effectively
-interpreting the blame set as a boundary-aware stack.
-
-These two modes of rationalizing give rise to two different notions of trail.
-\begin{quote}
-\it A \emph{Transient-first blame trail} is a sequence of scenarios
-$\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
-$\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
-\first{\mblame{\system, \conf_i}}$ where $\first{\mblame{\system, \conf}}$ is the
-first component Transient adds to the blame set for $\conf$.
-\end{quote}
-
-and
-
-\begin{quote}
-\it A \emph{Transient-last blame trail} is a sequence of scenarios
-$\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
-$\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
-  \last{\mblame{\system, \conf_i}}$ where $\last{\mblame{\system, \conf}}$ is the
-last component Transient adds to the blame set for $\conf$.
-\end{quote}
-
-The definition of \emph{Transient exception trails} 
-is analogous to that for Natural and, same as for Natural, we use it to  
-specify when  
-Transient-first and Transient-last blame are more useful than 
-Transient exceptions for debugging a scenario.
-
-%% -----------------------------------------------------------------------------
-\subsection{Modes of the Erasure Rational Programmer} \label{sub:erasure}
-
- Since gradually typed languages with Erasure semantics do not come with
- blame assignment, a rational programmer can only hope that the underlying
- safety checks and their exceptions are helpful.  Thus, the Erasure 
- rational programmer has a single mode, the Erasure exception mode, and
- its definition follows that for
- the Natural exception mode.
-
-%% -----------------------------------------------------------------------------
-\subsection{Experimental Process.}
-\label{subsec:experiment}
-
-Trails and their properties give us the tools for a rigorous examination
-of blame for Natural, Transient and Erasure Typed Racket in the setting of the 
-mutants from section~\ref{sec:mutate}. In line with the discussion so far, 
-our experimental process collects data to answer three initial questions
-for our interesting debugging scenarios:
-\begin{itemize}
-\item[$Q_1$] Is blame useful in the context of Natural?
-
-\item[$Q_2$] Is first blame useful in the context of Transient?
-
-\item[$Q_3$] Is last blame useful in the context of Transient?
-
-\end{itemize}
-
-Furthermore, our experiment can compare the relative usefulness of blame
-between any two of the three semantics. 
-\begin{itemize}
-\item[$Q_*$] Is blame for X more useful than blame for Y? (Where X and Y are any of Natural, Transient, or Erasure)
-\end{itemize}
-
-Table~\ref{fig:experiment-outline} summarizes how each question relates to
-different kinds of trails/modes of the rational programmer. For example, experimental
-question $Q_1$ asks whether blame is valuable for Natural and our experiment
-uses the Natural blame and exception trails to answer it.
-
-\begin{figure}[ht]
-\center
-{\begin{tabular}{l|c|c|c}
-%% ---------------------------------------------------------------------------------------------------------------
-                        & {\bf Natural}  & {\bf Transient} &  {\bf Erasure} \\ \hline 
-%% ---------------------------------------------------------------------------------------------------------------
-{\bf Blame}             &  $Q_1/Q_*$    &                  &                \\
-{\bf First blame}       &               &     $Q_2/Q_*$    &                 \\
-{\bf Last blame}        &               &     $Q_3/Q_*$    &                 \\
-{\bf Exceptions}        &      $Q_1$    &     $Q_2/Q_3$    &      $Q_*$      \\
-\end{tabular}}
-  \caption{ Experimental questions and rational programmer modes.}
-  \label{fig:experiment-outline}
-\end{figure}
-
-%% -----------------------------------------------------------------------------
-
-In detail we answer $Q_1$  by comparing the success of the 
-Natural blame and Natural exception trails for all interesting 
-debugging scenarios from section~\ref{sec:mutate}.
-  The first step is to construct each
-mutant's scenario lattice and identify their interesting debugging
-scenarios.
-Our test
-bed extends the trails that start from such roots according to the Natural-blame programmer.  If no scenarios
-can be added to the trail, we check if
-the last scenario of the trail type-checks or not. If it does, we record that the Natural-blame
-trail is successful, and otherwise that it fails. We repeat the process for the
-same roots again but for the other mode, Natural exceptions.
- Figure~\ref{fig:process}
-summarizes this
-experimental process for one mode of the rational programmer and connects
-it with the mutations from section~\ref{sec:mutate}.
-After completing the experiment, we calculate the
-success/failure results of the trails to determine for each root whether Natural blame
-is more useful than Natural exceptions. We
-obtain a positive answer for
-$Q_1$ when a root exists where the above is true since it is evidence that
-there is at least one interesting scenario that the rational programmer
-manages to debug because of blame. 
-The process is analogous for $Q_2$ and $Q_3$, employing the respective
-modes of the rational programmer.
-
-For $Q_*$, the process is a bit more
-involved. At a first level we proceed in a similar
-manner except that we use two modes 
-of the rational programmer from different semantics.
-However, answering this question requires us to compare the percentage of scenarios
-where one mode is more useful than the other and the inverse.
-For instance, to
-decide whether blame for Natural is more useful than first blame for Transient, 
-we compare the percentage of interesting scenarios where 
-Natural blame is more successful than Transient first blame 
-with the percentage of interesting scenarios where Transient first blame 
-is more successful than Natural blame. 
-Finally we repeat these steps one more time to compare Natural blame and Transient last blame
-and get a complete picture of the comparative usefulness of blame
-in the two semantics.
-
-\begin{figure}
-  \centering
-  \includegraphics[scale=0.36]{./Images/process}
-  \caption{The experimental process for one mode of the rational
-  programmer}
-  \label{fig:process}
-\end{figure}
-
-\subsection{Programmer Effort}
-\label{subsec:effort}
-
 In addition to successful and failure information for the various trails,  

### 2021-02-23 writing/rational-modes.tex
factor the rational programmer

--- 2021-02-23 writing/rational.tex
@@
-Code migration is the raison d'\^etre of Typed Racket, which is why the community
-uses ``migratory typing'' instead of gradual typing. Roughly speaking a
-programmer uses a gradually typed language to incrementally add types to a code
-base. From this perspective, a code base consists of some number of components,
-say modules, and a programmer converts a component at a time from its untyped
-state to a typed one.
-
-%% MF: I don't think we should honor the nonsense of ``true gt.''  We
-% are it as much as anyone!
-% 
-% \footnote{The method can be adapted to so-called true gradual typing
-% systems~\cite{svcb-snapl-2015} in a straightforward manner.}
-
-Based on this idea, migratory paths offer a concise setting for
-evaluating the effectiveness of a rational programmer for dealing with
-the interesting debugging scenarios from section ~\ref{sec:mutate}.
-Grouping debugging scenarios by mutation yields a set of programs that
-all contain the same mistake but differ in their type annotations.  In
-the setting of migratory typing, each scenario differs from another in
-which components are typed. Formally, a program $\system$ is a set of
-components $\set{\component}$, and a scenario $\conf$ is the typed
-subset of $\set{\component}$ . Hence, the scenarios of $\system$ are
-ordered by the subset relation and, naturally, form a lattice
-$\lattice{\system}$ with $2^{\size{\system}}$
-elements.\footnote{\citet{tfgnvf-popl-2016} use this lattice for
-performance evaluations.} The bottom scenario of $\lattice{\system}$
-is always $\emptyset$ and the top one is $\system$ itself. In between
-are all the mixed-typed scenarios.
-
-In this context, the actions of the rational programmer map to an
-ascending chain---dubbed a \emph{trail}---in $\lattice{\system}$.  The
-starting scenario $\conf_0$ of a trail---also referred to as the
-\emph{root} (of the trail)---is the initial debugging scenario; only
-interesting (as described in~\ref{sec:mutate}) debugging scenarios
-serve as roots.  Intuitively, starting from such a root, the rational
-programmer steps along a trail, and at each step, reduces the current
-scenario to an ``easier'' next scenario to debug.  The easiest
-scenarios are those that the type checker rejects outright, because a
-type error pinpoints an impedance mismatch between the broken code of
-one component and its typed interface to others. At this point, the
-programmer has identified the broken component.  If the scenario
-type-checks, the programmers runs the program until it raises a run
-time error.  The rational programmer uses the error's blame
-assignment\footnote{Blame can result either directly from a run-time
-check or, in the absence of blame, from an exception that can be
-traced back to the component that causes it with the help of the stack
-trace.} to decide which component(s) to equip with types. This choice
-constructs scenario $\conf_{i+1}$ from $\conf_{i}$ and thus creates a trail.
-
-% The successor of a scenario on a trail has more typed component;
-% formally, configuration $\conf_{i+1}$ on the trail differs from its
-% predecessor $\conf_{i}$ in that more components are typed.
-
-The first three subsections specialize the trail-scenario idea to the
-three semantics of gradual typing. Once this machinery is in place,
-the fourth subsection presents a reformulation of the informal
-question on page~1 as precise experimental research questions. The
-final subsection adds a notion of programmer effort to the
-experimental setup.
-
 %% -----------------------------------------------------------------------------
@@
 \end{quote}
-
-%% -----------------------------------------------------------------------------
-\subsection{Modes of the Transient Rational Programmer} \label{sub:transient}
-
-The transient semantics assigns blame to a set of components. In this setting, a
-blame assignment has the following meaning: the witness value to the impedance
-mismatch may have crossed the boundaries between elements in the set, and each crossing
-checked the value's type in a shallow manner.
-
-This ambiguity in Transient blame raises the question of how the rational programmer
-should react when the language produces a blame set. Our answer is that the rational
-programmer has at least two reasonable options. The first one is to select the component that is added
-to the blame set first and assign types to only that one---after all, if
-fully checked, the types of
-this first component should be able to detect an impedance mismatch earlier in the
-evaluation of a program than the later ones. The second option is to
-select the component that is added to the blame set last, effectively
-interpreting the blame set as a boundary-aware stack.
-
-These two modes of rationalizing give rise to two different notions of trail.
-\begin{quote}
-\it A \emph{Transient-first blame trail} is a sequence of scenarios
-$\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
-$\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
-\first{\mblame{\system, \conf_i}}$ where $\first{\mblame{\system, \conf}}$ is the
-first component Transient adds to the blame set for $\conf$.
-\end{quote}
-
-and
-
-\begin{quote}
-\it A \emph{Transient-last blame trail} is a sequence of scenarios
-$\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
-$\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
-  \last{\mblame{\system, \conf_i}}$ where $\last{\mblame{\system, \conf}}$ is the
-last component Transient adds to the blame set for $\conf$.
-\end{quote}
-
-The definition of \emph{Transient exception trails} 
-is analogous to that for Natural and, same as for Natural, we use it to  
-specify when  
-Transient-first and Transient-last blame are more useful than 
-Transient exceptions for debugging a scenario.
-
-%% -----------------------------------------------------------------------------
-\subsection{Modes of the Erasure Rational Programmer} \label{sub:erasure}
-
- Since gradually typed languages with Erasure semantics do not come with
- blame assignment, a rational programmer can only hope that the underlying
- safety checks and their exceptions are helpful.  Thus, the Erasure 
- rational programmer has a single mode, the Erasure exception mode, and
- its definition follows that for
- the Natural exception mode.
-
-%% -----------------------------------------------------------------------------
-\subsection{Experimental Process.}
-\label{subsec:experiment}
-
-Trails and their properties give us the tools for a rigorous examination
-of blame for Natural, Transient and Erasure Typed Racket in the setting of the 
-mutants from section~\ref{sec:mutate}. In line with the discussion so far, 
-our experimental process collects data to answer three initial questions
-for our interesting debugging scenarios:
-\begin{itemize}
-\item[$Q_1$] Is blame useful in the context of Natural?
-
-\item[$Q_2$] Is first blame useful in the context of Transient?
-
-\item[$Q_3$] Is last blame useful in the context of Transient?
-
-\end{itemize}
-
-Furthermore, our experiment can compare the relative usefulness of blame
-between any two of the three semantics. 
-\begin{itemize}
-\item[$Q_*$] Is blame for X more useful than blame for Y? (Where X and Y are any of Natural, Transient, or Erasure)
-\end{itemize}
-
-Table~\ref{fig:experiment-outline} summarizes how each question relates to
-different kinds of trails/modes of the rational programmer. For example, experimental
-question $Q_1$ asks whether blame is valuable for Natural and our experiment
-uses the Natural blame and exception trails to answer it.
-
-\begin{figure}[ht]
-\center
-{\begin{tabular}{l|c|c|c}
-%% ---------------------------------------------------------------------------------------------------------------
-                        & {\bf Natural}  & {\bf Transient} &  {\bf Erasure} \\ \hline 
-%% ---------------------------------------------------------------------------------------------------------------
-{\bf Blame}             &  $Q_1/Q_*$    &                  &                \\
-{\bf First blame}       &               &     $Q_2/Q_*$    &                 \\
-{\bf Last blame}        &               &     $Q_3/Q_*$    &                 \\
-{\bf Exceptions}        &      $Q_1$    &     $Q_2/Q_3$    &      $Q_*$      \\
-\end{tabular}}
-  \caption{ Experimental questions and rational programmer modes.}
-  \label{fig:experiment-outline}
-\end{figure}
-
-%% -----------------------------------------------------------------------------
-
-In detail we answer $Q_1$  by comparing the success of the 
-Natural blame and Natural exception trails for all interesting 
-debugging scenarios from section~\ref{sec:mutate}.
-  The first step is to construct each
-mutant's scenario lattice and identify their interesting debugging
-scenarios.
-Our test
-bed extends the trails that start from such roots according to the Natural-blame programmer.  If no scenarios
-can be added to the trail, we check if
-the last scenario of the trail type-checks or not. If it does, we record that the Natural-blame
-trail is successful, and otherwise that it fails. We repeat the process for the
-same roots again but for the other mode, Natural exceptions.
- Figure~\ref{fig:process}
-summarizes this
-experimental process for one mode of the rational programmer and connects
-it with the mutations from section~\ref{sec:mutate}.
-After completing the experiment, we calculate the
-success/failure results of the trails to determine for each root whether Natural blame
-is more useful than Natural exceptions. We
-obtain a positive answer for
-$Q_1$ when a root exists where the above is true since it is evidence that
-there is at least one interesting scenario that the rational programmer
-manages to debug because of blame. 
-The process is analogous for $Q_2$ and $Q_3$, employing the respective
-modes of the rational programmer.
-
-For $Q_*$, the process is a bit more
-involved. At a first level we proceed in a similar
-manner except that we use two modes 
-of the rational programmer from different semantics.
-However, answering this question requires us to compare the percentage of scenarios
-where one mode is more useful than the other and the inverse.
-For instance, to
-decide whether blame for Natural is more useful than first blame for Transient, 
-we compare the percentage of interesting scenarios where 
-Natural blame is more successful than Transient first blame 
-with the percentage of interesting scenarios where Transient first blame 
-is more successful than Natural blame. 
-Finally we repeat these steps one more time to compare Natural blame and Transient last blame
-and get a complete picture of the comparative usefulness of blame
-in the two semantics.
-
-\begin{figure}
-  \centering
-  \includegraphics[scale=0.36]{./Images/process}
-  \caption{The experimental process for one mode of the rational
-  programmer}
-  \label{fig:process}
-\end{figure}
-
-\subsection{Programmer Effort}
-\label{subsec:effort}
-
-In addition to successful and failure information for the various trails,  
-our experiment  records the number of components a rational programmer 
-has to type along each trail ($\lvert \conf_n \setminus \conf_0
-\rvert$). We use this number as the metric for the effort 
-of the rational programmer to debug an interesting scenario.
-
-A comparison of the effort of different modes of the rational programmer
-can help shed some further light to the effectiveness of the
-three gradual typing systems. For example, consider one debugging scenario.  
-If, for instance, both  Natural blame and Transient first
-blame trails are successful, the two modes of the 
-rational programmer can compete to see which one
-debugs the interesting scenario with less effort. In
-general, if the effort distribution for a mode of the rational programmer
-has a shorter tail and more
-volume around smaller values compared to the effort distribution of another
-mode, then this is evidence that the first mode is the more effective of the two.
-
-Finally, effort allows us to test whether the observed effectiveness 
-of the rational programmer is an artifact of pure chance or not.
-That is, we can compare the effort distribution for a mode of the
-rational programmer with another mode that ignores error information entirely
-and instead selects which component to type next randomly. 
-Of course, since each mutant has a finite number of components, 
-such random mode trails are always successful. At the same time though,
-the Natural, Transient, and Erasure random modes exhibit a normal effort
-distribution across scenarios that should be distinct from the effort distribution for all other modes of
-the rational programmer if the effectiveness of the modes are not
-coincidental.

### 2021-02-23 writing/rational-transient.tex
factor the rational programmer

--- 2021-02-23 writing/rational.tex
@@
-Code migration is the raison d'\^etre of Typed Racket, which is why the community
-uses ``migratory typing'' instead of gradual typing. Roughly speaking a
-programmer uses a gradually typed language to incrementally add types to a code
-base. From this perspective, a code base consists of some number of components,
-say modules, and a programmer converts a component at a time from its untyped
-state to a typed one.
-
-%% MF: I don't think we should honor the nonsense of ``true gt.''  We
-% are it as much as anyone!
-% 
-% \footnote{The method can be adapted to so-called true gradual typing
-% systems~\cite{svcb-snapl-2015} in a straightforward manner.}
-
-Based on this idea, migratory paths offer a concise setting for
-evaluating the effectiveness of a rational programmer for dealing with
-the interesting debugging scenarios from section ~\ref{sec:mutate}.
-Grouping debugging scenarios by mutation yields a set of programs that
-all contain the same mistake but differ in their type annotations.  In
-the setting of migratory typing, each scenario differs from another in
-which components are typed. Formally, a program $\system$ is a set of
-components $\set{\component}$, and a scenario $\conf$ is the typed
-subset of $\set{\component}$ . Hence, the scenarios of $\system$ are
-ordered by the subset relation and, naturally, form a lattice
-$\lattice{\system}$ with $2^{\size{\system}}$
-elements.\footnote{\citet{tfgnvf-popl-2016} use this lattice for
-performance evaluations.} The bottom scenario of $\lattice{\system}$
-is always $\emptyset$ and the top one is $\system$ itself. In between
-are all the mixed-typed scenarios.
-
-In this context, the actions of the rational programmer map to an
-ascending chain---dubbed a \emph{trail}---in $\lattice{\system}$.  The
-starting scenario $\conf_0$ of a trail---also referred to as the
-\emph{root} (of the trail)---is the initial debugging scenario; only
-interesting (as described in~\ref{sec:mutate}) debugging scenarios
-serve as roots.  Intuitively, starting from such a root, the rational
-programmer steps along a trail, and at each step, reduces the current
-scenario to an ``easier'' next scenario to debug.  The easiest
-scenarios are those that the type checker rejects outright, because a
-type error pinpoints an impedance mismatch between the broken code of
-one component and its typed interface to others. At this point, the
-programmer has identified the broken component.  If the scenario
-type-checks, the programmers runs the program until it raises a run
-time error.  The rational programmer uses the error's blame
-assignment\footnote{Blame can result either directly from a run-time
-check or, in the absence of blame, from an exception that can be
-traced back to the component that causes it with the help of the stack
-trace.} to decide which component(s) to equip with types. This choice
-constructs scenario $\conf_{i+1}$ from $\conf_{i}$ and thus creates a trail.
-
-% The successor of a scenario on a trail has more typed component;
-% formally, configuration $\conf_{i+1}$ on the trail differs from its
-% predecessor $\conf_{i}$ in that more components are typed.
-
-The first three subsections specialize the trail-scenario idea to the
-three semantics of gradual typing. Once this machinery is in place,
-the fourth subsection presents a reformulation of the informal
-question on page~1 as precise experimental research questions. The
-final subsection adds a notion of programmer effort to the
-experimental setup.
-
 %% -----------------------------------------------------------------------------
-\subsection{Modes of the Natural Rational Programmer} \label{sub:natural}
-
-The Natural semantics assigns blame to exactly one boundary.  A blame assignment
-has the following specific meaning: the typed component may make incorrect type
-assumptions about the untyped component in its interface, or the correct
-interface exposes a bug in the untyped component. Our setup rules out the first
-alternative (but see section~\ref{sec:conclusion}), and therefore the rational
-programmer extends the trail to a scenario that swaps out the untyped component
-for its typed counterpart.
-
-We can turn this description into a formal definition of the trail. 
-\begin{quote}
-\it A \emph{Natural blame trail} is a sequence of scenarios $\conf_0,...\conf_n$ of
-a program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
-\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =\{\blame{\system, \conf_i}\}$ where
-$\blame{\system, \conf}$ denotes the component that $\conf$ blames.
-\end{quote}
-
-When the buggy untyped component of a program is replaced by the typed counterpart,
-the type checker fails if this component causes the impedance mismatch. Hence, a
-trail that ends at an ill-typed scenarios successfully pinpoints the location of
-bugs. 
-\begin{quote}
-\it A Natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
-\emph{successful} iff its last scenario $\conf_n$ does not type check.  A Natural
-blame trail $\conf_0,..,\conf_n$ in a lattice $\lattice{\system}$ is \emph{failing}
-iff $\conf_n$ type checks and the trail cannot be extended further.
-\end{quote}
-That is, failing Natural blame trails are those that have not reached a scenario that
-reveals the bug at compile time and their last scenario does not blame an untyped component. Thus
-the rational programmer gets no hints from the gradual type system on how to
-continue the search for the bug.
-
-While a successful Natural blame trail indicates that it 
-pays off to heed blame assignments during debugging the trail's root, it does not answer whether
-blame is a critical piece of the rational programmer's process.  For instance,
-typing the top of stack trace of a failed run time type check, dubbed the
-location of the exception, might be as useful as typing the blamed one.
-
-To account for this situation, we devise a new mode of the Natural rational
-programmer that follows a migration process based on exceptions:
-\begin{quote}
-\it A Natural exception trail is a sequence of scenarios $\conf_0,...\conf_n$ of a
-program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
-\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i = \{\exception{\system, \conf_i}\}$
-where $\exception{\system, \conf}$ denotes the location of the exception of $\conf$.
-\end{quote}
-
-Same as for success for a Natural blame trail, we define success for a
-Natural exception trail.
-Putting the definitions for the two modes together, we can now compare the usefulness of blame 
-with that for mere exceptions for debugging a scenario in the context of Natural semantics.
-\begin{quote}
-\it 
-  Given a program $\system$ and a root $\conf_0$ in $\lattice{\system}$,
-  Natural blame is \emph{more useful} than Natural exceptions for
-  debugging $\conf_0$ iff 
-  the Natural blame trail 
-  that starts at $\conf_0$ is successful while the Natural exception trail that
-  starts at $\conf_0$ is failing.
-\end{quote}
-
-%% -----------------------------------------------------------------------------
-\subsection{Modes of the Transient Rational Programmer} \label{sub:transient}
 
@@
 Transient exceptions for debugging a scenario.
-
-%% -----------------------------------------------------------------------------
-\subsection{Modes of the Erasure Rational Programmer} \label{sub:erasure}
-
- Since gradually typed languages with Erasure semantics do not come with
- blame assignment, a rational programmer can only hope that the underlying
- safety checks and their exceptions are helpful.  Thus, the Erasure 
- rational programmer has a single mode, the Erasure exception mode, and
- its definition follows that for
- the Natural exception mode.
-
-%% -----------------------------------------------------------------------------
-\subsection{Experimental Process.}
-\label{subsec:experiment}
-
-Trails and their properties give us the tools for a rigorous examination
-of blame for Natural, Transient and Erasure Typed Racket in the setting of the 
-mutants from section~\ref{sec:mutate}. In line with the discussion so far, 
-our experimental process collects data to answer three initial questions
-for our interesting debugging scenarios:
-\begin{itemize}
-\item[$Q_1$] Is blame useful in the context of Natural?
-
-\item[$Q_2$] Is first blame useful in the context of Transient?
-
-\item[$Q_3$] Is last blame useful in the context of Transient?
-
-\end{itemize}
-
-Furthermore, our experiment can compare the relative usefulness of blame
-between any two of the three semantics. 
-\begin{itemize}
-\item[$Q_*$] Is blame for X more useful than blame for Y? (Where X and Y are any of Natural, Transient, or Erasure)
-\end{itemize}
-
-Table~\ref{fig:experiment-outline} summarizes how each question relates to
-different kinds of trails/modes of the rational programmer. For example, experimental
-question $Q_1$ asks whether blame is valuable for Natural and our experiment
-uses the Natural blame and exception trails to answer it.
-
-\begin{figure}[ht]
-\center
-{\begin{tabular}{l|c|c|c}
-%% ---------------------------------------------------------------------------------------------------------------
-                        & {\bf Natural}  & {\bf Transient} &  {\bf Erasure} \\ \hline 
-%% ---------------------------------------------------------------------------------------------------------------
-{\bf Blame}             &  $Q_1/Q_*$    &                  &                \\
-{\bf First blame}       &               &     $Q_2/Q_*$    &                 \\
-{\bf Last blame}        &               &     $Q_3/Q_*$    &                 \\
-{\bf Exceptions}        &      $Q_1$    &     $Q_2/Q_3$    &      $Q_*$      \\
-\end{tabular}}
-  \caption{ Experimental questions and rational programmer modes.}
-  \label{fig:experiment-outline}
-\end{figure}
-
-%% -----------------------------------------------------------------------------
-
-In detail we answer $Q_1$  by comparing the success of the 
-Natural blame and Natural exception trails for all interesting 
-debugging scenarios from section~\ref{sec:mutate}.
-  The first step is to construct each
-mutant's scenario lattice and identify their interesting debugging
-scenarios.
-Our test
-bed extends the trails that start from such roots according to the Natural-blame programmer.  If no scenarios
-can be added to the trail, we check if
-the last scenario of the trail type-checks or not. If it does, we record that the Natural-blame
-trail is successful, and otherwise that it fails. We repeat the process for the
-same roots again but for the other mode, Natural exceptions.
- Figure~\ref{fig:process}
-summarizes this
-experimental process for one mode of the rational programmer and connects
-it with the mutations from section~\ref{sec:mutate}.
-After completing the experiment, we calculate the
-success/failure results of the trails to determine for each root whether Natural blame
-is more useful than Natural exceptions. We
-obtain a positive answer for
-$Q_1$ when a root exists where the above is true since it is evidence that
-there is at least one interesting scenario that the rational programmer
-manages to debug because of blame. 
-The process is analogous for $Q_2$ and $Q_3$, employing the respective
-modes of the rational programmer.
-
-For $Q_*$, the process is a bit more
-involved. At a first level we proceed in a similar
-manner except that we use two modes 
-of the rational programmer from different semantics.
-However, answering this question requires us to compare the percentage of scenarios
-where one mode is more useful than the other and the inverse.
-For instance, to
-decide whether blame for Natural is more useful than first blame for Transient, 
-we compare the percentage of interesting scenarios where 
-Natural blame is more successful than Transient first blame 
-with the percentage of interesting scenarios where Transient first blame 
-is more successful than Natural blame. 
-Finally we repeat these steps one more time to compare Natural blame and Transient last blame
-and get a complete picture of the comparative usefulness of blame
-in the two semantics.
-
-\begin{figure}
-  \centering
-  \includegraphics[scale=0.36]{./Images/process}
-  \caption{The experimental process for one mode of the rational
-  programmer}
-  \label{fig:process}
-\end{figure}
-
-\subsection{Programmer Effort}
-\label{subsec:effort}
-
-In addition to successful and failure information for the various trails,  
-our experiment  records the number of components a rational programmer 
-has to type along each trail ($\lvert \conf_n \setminus \conf_0
-\rvert$). We use this number as the metric for the effort 
-of the rational programmer to debug an interesting scenario.
-
-A comparison of the effort of different modes of the rational programmer
-can help shed some further light to the effectiveness of the
-three gradual typing systems. For example, consider one debugging scenario.  
-If, for instance, both  Natural blame and Transient first
-blame trails are successful, the two modes of the 
-rational programmer can compete to see which one
-debugs the interesting scenario with less effort. In
-general, if the effort distribution for a mode of the rational programmer
-has a shorter tail and more
-volume around smaller values compared to the effort distribution of another
-mode, then this is evidence that the first mode is the more effective of the two.
-
-Finally, effort allows us to test whether the observed effectiveness 
-of the rational programmer is an artifact of pure chance or not.
-That is, we can compare the effort distribution for a mode of the
-rational programmer with another mode that ignores error information entirely
-and instead selects which component to type next randomly. 
-Of course, since each mutant has a finite number of components, 
-such random mode trails are always successful. At the same time though,
-the Natural, Transient, and Erasure random modes exhibit a normal effort
-distribution across scenarios that should be distinct from the effort distribution for all other modes of
-the rational programmer if the effectiveness of the modes are not
-coincidental.

### 2021-02-23 writing/rational.tex
factor the rational programmer

@@
 %% -----------------------------------------------------------------------------
-\subsection{Modes of the Natural Rational Programmer} \label{sub:natural}
+\def\rsub#1#2{\subsection{#2} \label{sub:#1} \input{rational-#1.tex}}
 
-The Natural semantics assigns blame to exactly one boundary.  A blame assignment
-has the following specific meaning: the typed component may make incorrect type
-assumptions about the untyped component in its interface, or the correct
-interface exposes a bug in the untyped component. Our setup rules out the first
-alternative (but see section~\ref{sec:conclusion}), and therefore the rational
-programmer extends the trail to a scenario that swaps out the untyped component
-for its typed counterpart.
-
-We can turn this description into a formal definition of the trail. 
-\begin{quote}
-\it A \emph{Natural blame trail} is a sequence of scenarios $\conf_0,...\conf_n$ of
-a program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
-\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =\{\blame{\system, \conf_i}\}$ where
-$\blame{\system, \conf}$ denotes the component that $\conf$ blames.
-\end{quote}
-
-When the buggy untyped component of a program is replaced by the typed counterpart,
-the type checker fails if this component causes the impedance mismatch. Hence, a
-trail that ends at an ill-typed scenarios successfully pinpoints the location of
-bugs. 
-\begin{quote}
-\it A Natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
-\emph{successful} iff its last scenario $\conf_n$ does not type check.  A Natural
-blame trail $\conf_0,..,\conf_n$ in a lattice $\lattice{\system}$ is \emph{failing}
-iff $\conf_n$ type checks and the trail cannot be extended further.
-\end{quote}
-That is, failing Natural blame trails are those that have not reached a scenario that
-reveals the bug at compile time and their last scenario does not blame an untyped component. Thus
-the rational programmer gets no hints from the gradual type system on how to
-continue the search for the bug.
-
-While a successful Natural blame trail indicates that it 
-pays off to heed blame assignments during debugging the trail's root, it does not answer whether
-blame is a critical piece of the rational programmer's process.  For instance,
-typing the top of stack trace of a failed run time type check, dubbed the
-location of the exception, might be as useful as typing the blamed one.
-
-To account for this situation, we devise a new mode of the Natural rational
-programmer that follows a migration process based on exceptions:
-\begin{quote}
-\it A Natural exception trail is a sequence of scenarios $\conf_0,...\conf_n$ of a
-program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
-\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i = \{\exception{\system, \conf_i}\}$
-where $\exception{\system, \conf}$ denotes the location of the exception of $\conf$.
-\end{quote}
-
-Same as for success for a Natural blame trail, we define success for a
-Natural exception trail.
-Putting the definitions for the two modes together, we can now compare the usefulness of blame 
-with that for mere exceptions for debugging a scenario in the context of Natural semantics.
-\begin{quote}
-\it 
-  Given a program $\system$ and a root $\conf_0$ in $\lattice{\system}$,
-  Natural blame is \emph{more useful} than Natural exceptions for
-  debugging $\conf_0$ iff 
-  the Natural blame trail 
-  that starts at $\conf_0$ is successful while the Natural exception trail that
-  starts at $\conf_0$ is failing.
-\end{quote}
-
-%% -----------------------------------------------------------------------------
-\subsection{Modes of the Transient Rational Programmer} \label{sub:transient}
-
-The transient semantics assigns blame to a set of components. In this setting, a
-blame assignment has the following meaning: the witness value to the impedance
-mismatch may have crossed the boundaries between elements in the set, and each crossing
-checked the value's type in a shallow manner.
-
-This ambiguity in Transient blame raises the question of how the rational programmer
-should react when the language produces a blame set. Our answer is that the rational
-programmer has at least two reasonable options. The first one is to select the component that is added
-to the blame set first and assign types to only that one---after all, if
-fully checked, the types of
-this first component should be able to detect an impedance mismatch earlier in the
-evaluation of a program than the later ones. The second option is to
-select the component that is added to the blame set last, effectively
-interpreting the blame set as a boundary-aware stack.
-
-These two modes of rationalizing give rise to two different notions of trail.
-\begin{quote}
-\it A \emph{Transient-first blame trail} is a sequence of scenarios
-$\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
-$\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
-\first{\mblame{\system, \conf_i}}$ where $\first{\mblame{\system, \conf}}$ is the
-first component Transient adds to the blame set for $\conf$.
-\end{quote}
-
-and
-
-\begin{quote}
-\it A \emph{Transient-last blame trail} is a sequence of scenarios
-$\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
-$\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
-  \last{\mblame{\system, \conf_i}}$ where $\last{\mblame{\system, \conf}}$ is the
-last component Transient adds to the blame set for $\conf$.
-\end{quote}
-
-The definition of \emph{Transient exception trails} 
-is analogous to that for Natural and, same as for Natural, we use it to  
-specify when  
-Transient-first and Transient-last blame are more useful than 
-Transient exceptions for debugging a scenario.
-
-%% -----------------------------------------------------------------------------
-\subsection{Modes of the Erasure Rational Programmer} \label{sub:erasure}
-
- Since gradually typed languages with Erasure semantics do not come with
- blame assignment, a rational programmer can only hope that the underlying
- safety checks and their exceptions are helpful.  Thus, the Erasure 
- rational programmer has a single mode, the Erasure exception mode, and
- its definition follows that for
- the Natural exception mode.
-
-%% -----------------------------------------------------------------------------
-\subsection{Experimental Process.}
-\label{subsec:experiment}
-
-Trails and their properties give us the tools for a rigorous examination
-of blame for Natural, Transient and Erasure Typed Racket in the setting of the 
-mutants from section~\ref{sec:mutate}. In line with the discussion so far, 
-our experimental process collects data to answer three initial questions
-for our interesting debugging scenarios:
-\begin{itemize}
-\item[$Q_1$] Is blame useful in the context of Natural?
-
-\item[$Q_2$] Is first blame useful in the context of Transient?
-
-\item[$Q_3$] Is last blame useful in the context of Transient?
-
-\end{itemize}
-
-Furthermore, our experiment can compare the relative usefulness of blame
-between any two of the three semantics. 
-\begin{itemize}
-\item[$Q_*$] Is blame for X more useful than blame for Y? (Where X and Y are any of Natural, Transient, or Erasure)
-\end{itemize}
-
-Table~\ref{fig:experiment-outline} summarizes how each question relates to
-different kinds of trails/modes of the rational programmer. For example, experimental
-question $Q_1$ asks whether blame is valuable for Natural and our experiment
-uses the Natural blame and exception trails to answer it.
-
-\begin{figure}[ht]
-\center
-{\begin{tabular}{l|c|c|c}
-%% ---------------------------------------------------------------------------------------------------------------
-                        & {\bf Natural}  & {\bf Transient} &  {\bf Erasure} \\ \hline 
-%% ---------------------------------------------------------------------------------------------------------------
-{\bf Blame}             &  $Q_1/Q_*$    &                  &                \\
-{\bf First blame}       &               &     $Q_2/Q_*$    &                 \\
-{\bf Last blame}        &               &     $Q_3/Q_*$    &                 \\
-{\bf Exceptions}        &      $Q_1$    &     $Q_2/Q_3$    &      $Q_*$      \\
-\end{tabular}}
-  \caption{ Experimental questions and rational programmer modes.}
-  \label{fig:experiment-outline}
-\end{figure}
-
-%% -----------------------------------------------------------------------------
-
-In detail we answer $Q_1$  by comparing the success of the 
-Natural blame and Natural exception trails for all interesting 
-debugging scenarios from section~\ref{sec:mutate}.
-  The first step is to construct each
-mutant's scenario lattice and identify their interesting debugging
-scenarios.
-Our test
-bed extends the trails that start from such roots according to the Natural-blame programmer.  If no scenarios
-can be added to the trail, we check if
-the last scenario of the trail type-checks or not. If it does, we record that the Natural-blame
-trail is successful, and otherwise that it fails. We repeat the process for the
-same roots again but for the other mode, Natural exceptions.
- Figure~\ref{fig:process}
-summarizes this
-experimental process for one mode of the rational programmer and connects
-it with the mutations from section~\ref{sec:mutate}.
-After completing the experiment, we calculate the
-success/failure results of the trails to determine for each root whether Natural blame
-is more useful than Natural exceptions. We
-obtain a positive answer for
-$Q_1$ when a root exists where the above is true since it is evidence that
-there is at least one interesting scenario that the rational programmer
-manages to debug because of blame. 
-The process is analogous for $Q_2$ and $Q_3$, employing the respective
-modes of the rational programmer.
-
-For $Q_*$, the process is a bit more
-involved. At a first level we proceed in a similar
-manner except that we use two modes 
-of the rational programmer from different semantics.
-However, answering this question requires us to compare the percentage of scenarios
-where one mode is more useful than the other and the inverse.
-For instance, to
-decide whether blame for Natural is more useful than first blame for Transient, 
-we compare the percentage of interesting scenarios where 
-Natural blame is more successful than Transient first blame 
-with the percentage of interesting scenarios where Transient first blame 
-is more successful than Natural blame. 
-Finally we repeat these steps one more time to compare Natural blame and Transient last blame
-and get a complete picture of the comparative usefulness of blame
-in the two semantics.
-
-\begin{figure}
-  \centering
-  \includegraphics[scale=0.36]{./Images/process}
-  \caption{The experimental process for one mode of the rational
-  programmer}
-  \label{fig:process}
-\end{figure}
-
-\subsection{Programmer Effort}
-\label{subsec:effort}
-
-In addition to successful and failure information for the various trails,  
-our experiment  records the number of components a rational programmer 
-has to type along each trail ($\lvert \conf_n \setminus \conf_0
-\rvert$). We use this number as the metric for the effort 
-of the rational programmer to debug an interesting scenario.
-
-A comparison of the effort of different modes of the rational programmer
-can help shed some further light to the effectiveness of the
-three gradual typing systems. For example, consider one debugging scenario.  
-If, for instance, both  Natural blame and Transient first
-blame trails are successful, the two modes of the 
-rational programmer can compete to see which one
-debugs the interesting scenario with less effort. In
-general, if the effort distribution for a mode of the rational programmer
-has a shorter tail and more
-volume around smaller values compared to the effort distribution of another
-mode, then this is evidence that the first mode is the more effective of the two.
-
-Finally, effort allows us to test whether the observed effectiveness 
-of the rational programmer is an artifact of pure chance or not.
-That is, we can compare the effort distribution for a mode of the
-rational programmer with another mode that ignores error information entirely
-and instead selects which component to type next randomly. 
-Of course, since each mutant has a finite number of components, 
-such random mode trails are always successful. At the same time though,
-the Natural, Transient, and Erasure random modes exhibit a normal effort
-distribution across scenarios that should be distinct from the effort distribution for all other modes of
-the rational programmer if the effectiveness of the modes are not
-coincidental.
+\rsub{natural}   {Modes of the Natural Rational Programmer} 
+\rsub{transient} {Modes of the Transient Rational Programmer} 
+\rsub{erasure}   {Modes of the Erasure Rational Programmer} 
+\rsub{experiment}{Experimental Process} 
+\rsub{effort}    {Programmer Effort} \label{subsec:effort}

### 2021-02-23 writing/results.tex
typos + small edits
organize sections/titles/labels in paper
factor the rational programmer

@@
-\section{What We Discovered}
-\label{sec:results}
-
 We run the experimental process from section~\ref{sec:rational} on the
@@
 Figure~\ref{fig:avo-matrix} summarizes our results and provides answers to
-the four questions from section~\ref{subsec:experiment}.  Specifically, it
+the four questions from section~\ref{sub:experiment}.  Specifically, it
 depicts a head-to-head comparison of the usefulness of each mode against
@@
 
-Figure~\ref{subsec:experiment} shows that the answers for questions $Q_1$
+Figure~\ref{sub:experiment} shows that the answers for questions $Q_1$
 to $Q_3$ are positive. In all three semantics, blame modes outperform
@@
   \caption{Programmer effort: Each plot depicts the distribution of trail
-  lengths for a given mode across all benchmarks, starting from trails
-  of length 0.}
+  lengths for a given mode across all benchmarks.}
   \label{fig:effort-table}

### 2021-02-28 writing/challenges.tex
organize sections/titles/labels in paper
factor mutate
revise intro to point to new 'why rational' section

@@
-\section{Why Is It Hard to Evaluate Blame?}
-\label{sec:challenges}
-
 %% -----------------------------------------------------------------------------
-
-The general challenge of evaluating blame is a methodological one. Unlike most
-current research on programming languages, the question seems to call for
-empirical research similar to that of human-computer interaction. At the same
-time, a significant evaluation result demands a large amount of
-data. As~\citet{lksfd-popl-2020} recently demonstrated, the way around this
-dilemma is to postulate the {\em rational programmer\/}---the equivalent of a
-{\it homo economicus\/} for programming languages---and to implement it as an algorithm.
-
-In the context of gradual typing, the rational programmer translates the
-Wadler-Findler slogan into a debugging method, searching for the source of the
-impedance mismatch in an incremental fashion. Measuring this simulated
-programmer's behavior on a large number of debugging scenarios yields data that
-is similar to data collected in an empirical, human-facing manner.
 
@@
 survived this first scrutiny testing. All of the remaining ones yield {\em
-interesting debugging scenarios\/} (see section~\ref{sub:interesting}).
+interesting debugging scenarios\/} (see section~\ref{sub:mutate-interesting}).
 

### 2021-02-28 writing/introduction.tex
organize sections/titles/labels in paper
revise intro to point to new 'why rational' section

@@
-\section{Does Blame Matter?}
-\label{sec:introduction}
-
 Theoreticians of gradual typing have focused on blame theorems from the very
@@
 
-This contrast between theoretical research and industrial applications of gradual typing raises the question 
+This contrast between theory and applications of gradual typing raises the question 
 \begin{quote}
@@
 effectiveness of blame assignment strategies\/} in the gradual typing world.
-speaking, the method simulates a programmer who follows the Wadler-Findler
-slogan quoted above; its automation borrows elements of of Lazarek et
-al.~\cite{lksfd-popl-2020}'s recent work.\footnote{\citet{cc-snapl-19} use this
-tracing method to localize type errors reported by gradual typing systems based
-on inference.}  Such a \emph{rational programmer} can find impedance-mismatch
-bugs by adding types to a partially typed program (see
-section~\ref{sec:landscape}). That is, when the run-time system signals an
-impedance mismatch, a good blame message helps the programmer infer where to add
-an additional type specification because, by the basic theory, well-typed
-components cannot cause impedance mismatches. Otherwise, if the additional type
-does {\em not\/} cause a static type error, the program is run again to signal
-another violation. Then the programmer can heed blame one more time to add yet
-another type specification. Evaluating the quality of blame assignment thus
-becomes a question of whether the tracing the error messages finds the source of
-the mistake and how many steps it takes to get to this point.
+The top-level innovation is the idea of a {\em rational programmer\/}, that is,
+a programmer that acts only in response to available information (see section~\ref{sec:why-rational}). In the case
+of an impedance mismatch, the available information consists of the error
+message and the current state of the program. The rational programmer can hence
+use the former to change the latter---and this systematic, information-driven
+process can be implemented.\footnote{\citet{cc-snapl-19} use this tracing
+method to localize type errors reported by gradual typing systems based on
+inference.} Implementing it means overcoming three major challenges: injecting
+representative impedance mismatches; putting the various kinds of error
+information to comparable use; and sampling the huge space of possibilities
+properly (see section~\ref{sec:challenges}).
+
+% The method simulates a programmer who follows the Wadler-Findler slogan quoted
+% above; its automation borrows elements of of Lazarek et
+% al.~\cite{lksfd-popl-2020}'s recent work.  Such a \emph{rational programmer} can find impedance-mismatch
+% bugs by adding types to a partially typed program (see
+% section~\ref{sec:landscape}). That is, when the run-time system signals an
+% impedance mismatch, a good blame message helps the programmer infer where to
+% add an additional type specification because, by the basic theory, well-typed
+% components cannot cause impedance mismatches. Otherwise, if the additional type
+% does {\em not\/} cause a static type error, the program is run again to signal
+% another violation. Then the programmer can heed blame one more time to add yet
+% another type specification. Evaluating the quality of blame assignment thus
+% becomes a question of whether the tracing the error messages finds the source
+% of the mistake and how many steps it takes to get to this point.
 
@@
 {\it Transient\/} nor are the two blame assignment methods clearly preferable to
-{\it Erasure\/}. The third one casts doubt on the current the trade-off between
-the cost of blame-assignment systems and the quality of their blame messages.
-In turn, these problems suggests that, on one hand, the existing theory does not
-predict practice properly, and on the other hand, the existing practice needs to
-be supported by additional data.
+{\it Erasure\/}.  In turn, these problems suggests that, on one hand, the
+existing theory does not predict practice properly, and on the other hand, the
+existing practice may need to find additional ways to collect data.

### 2021-02-28 writing/why-rational.tex
revise intro to point to new 'why rational' section

%% -----------------------------------------------------------------------------

The general challenge of evaluating blame is a methodological one. Unlike most
current research on programming languages, the question seems to call for
empirical research similar to that of human-computer interaction. At the same
time, a significant evaluation result demands a large amount of
data. As~\citet{lksfd-popl-2020} recently demonstrated, the way around this
dilemma is to postulate the {\em rational programmer\/}---the equivalent of a
{\it homo economicus\/} for programming languages---and to implement it as an algorithm.

In the context of gradual typing, the rational programmer translates the
Wadler-Findler slogan into a debugging method, searching for the source of the
impedance mismatch in an incremental fashion. Measuring this simulated
programmer's behavior on a large number of debugging scenarios yields data that
is similar to data collected in an empirical, human-facing manner.

\input{why-rational-case.tex}

### 2021-02-28 writing/results.tex
results: Add new figures with descriptions

@@
 We run the experimental process from section~\ref{sec:rational} on the
-sampled debugging scenarios from section~\ref{sec:sample} using a machine with two Intel Xeon Gold 6258R processors and 500GB of memory.
+sampled debugging scenarios from section~\ref{sec:sample} using a machine with two Intel Xeon Gold 6258R processors (28 doubly-threaded cores each) and 500GB of memory.
 Each debugging scenario had a timeout limit of 4 minutes and a
@@
 
-Figure~\ref{fig:avo-matrix} summarizes our results and provides answers to
-the four questions from section~\ref{sub:experiment}.  Specifically, it
-depicts a head-to-head comparison of the usefulness of each mode against
-every other mode.  The cell in row $R$ and column $C$ plots the percentage
-of interesting debugging scenarios for which $R$ is more useful $C$ ---
-we calculate this percentage as an estimated proportion from our sample.
-In effect, reading across a given row $R$ provides an
-overview of how $R$ compares to every other mode; the higher the bars, the
-more interesting debugging scenarios for which $R$ is strictly more useful
-than the other modes.  Reading down a given row $C$ provides an overview
-of how every other mode compares to $C$; the higher the bars, the more
-interesting debugging scenarios for which $C$ is strictly less useful than
-the other modes. For example, the second cell in the row of Natural blame
-shows that Natural is more useful than Natural exceptions for 12\% of all
-the interesting debugging scenarios we generate from our benchmarks and
-mutators.
-
-Figure~\ref{sub:experiment} shows that the answers for questions $Q_1$
-to $Q_3$ are positive. In all three semantics, blame modes outperform
-their corresponding exception mode (by 9.44\% to 12.7\%). The Natural
-exceptions mode is never more useful than Natural blame, and  Transient
-exceptions are more useful than Transient first and Transient last blame
-in 2.19\% and 2.8\% of the scenarios respectively.
-
-For the answers to $Q_*$, we compare the percentages in antisymmetric
-positions in the matrix. Blame for all three semantics is significantly
-more useful than Erasure exceptions (by 33\% to 35\%). Natural blame is
-more useful than both versions of Transient blame by a small percentage;
-in 4.05\% of scenarios Natural blame is more useful while in 1.67\% of the
-scenarios Transient blame is more useful. The Transient first and
-Transient last blame are practically indistinguishable. Finally, there is
-no clear winner between Natural exceptions and Transient exceptions
-despite the theoretically advantageous additional checks of the Natural
-semantics.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.45\textwidth]{./plots/avo-matrix}
-  \caption{Usefulness comparisons: Each cell depicts the estimated percentage of
-  interesting debugging scenarios for which the row mode is more useful
-  than the column mode.
-  The upper bound of error margin is 0.02\%}
-  \label{fig:avo-matrix}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.4\textwidth]{./plots/bt-lengths-table}
-  \caption{Programmer effort: Each plot depicts the distribution of trail
-  lengths for a given mode across all benchmarks.}
-  \label{fig:effort-table}
-\end{figure}
-
-Figure~\ref{fig:effort-table} shows the distribution of programmer effort
-for our sample of interesting debugging scenarios. Unlike the usefulness
-comparisons above, these proportions do not generalize to a representation of the full population.
-Still, as we discuss in section~\ref{subsec:effort}, they provide an
-alternative view of the workings of the rational programmer.
-
-There are
-three immediate take-aways from the figure. First, the effort for successfully
-debugging interesting scenarios (in green) for the random mode of the
-rational programmer follows  a normal distribution, as expected.\footnote{The
-random mode distribution is indistinguishable for all three semantics and thus the figure~\ref{fig:effort-table} 
-only shows one plot.} In contrast, in the other modes, successful effort coalesces at
-the left side of the plot, meaning that in most cases the programmer needs
-to type a single component to debug a scenario.
-
-A second point of interest is that the exceptions of
-the Erasure semantics either help the rational programmer immediately or 
-the rational programmer fails to debug a scenario altogether (in red).
-This is expected; no matter how many type annotations the programmer
-adds to an Erasure program, if the type checker doesn't reject the
-program, running the program always produces the same outcome. Thus an
-exception from the runtime has to point to the buggy module with
-the first try. Otherwise the rational programmer types an irrelevant
-module, runs the program again and the exception points again to the
-already typed module.
-
-A third notable piece of information from
-the figure is that every mode has failed debugging scenarios, not just
+Figure~\ref{fig:success-bars} summarizes the overall success rates of every mode.
+The success rates illustrate a few points that underly the rest of our analysis.
+The first notable piece of information from
+this figure is that every mode has failed debugging scenarios, not just
 Erasure. This should
@@
 does not contain frames from any module of the program, or only typed modules. This situation
-can even happen immediately, which explains failures with 0 effort.
+can happen several steps along a blame trail or even immediately (figure~\ref{fig:effort-table} dives into this below).
 
@@
 failure for Transient affect our experiment.
+
+The second key observation from figure~\ref{fig:success-bars} is that the modes that use blame all outperform those that do not.
+In particular, Natural and both of Transient's blame modes succeed in around 95\% of our scenarios, while their corresponding exception modes succeed in around 85\% of them and Erasure succeeds in only 60\% of them.
+The only exception is that the Random programmer always succeeds;
+this reflects the fact that every scenario has finitely many modules and this mode can never get stuck.
+
+Figure~\ref{fig:avo-bars} depicts a head-to-head comparison of every mode's performance against every other mode, helping to answer the four questions from section~\ref{sub:experiment}.
+Each plot shows a breakdown of the proportion of scenarios where one mode performs better or worse than each other mode.
+In particular, each bar above zero represents the proportion where the plot's entitled mode succeeds and the mode on the x-axis fails; the correspnding bar below zero represents the proportion of the opposite case.
+For example, the plot titled ``Natural'' shows that Natural outperforms Natural exceptions in about 10\% of the scenarios, and the inverse (Natural performs wrse than Natural exceptions) never happens.
+Similarly, the plot titled ``Transient last blame'' shows that Transient last blame outperforms Natural exceptions about in abut 10\% of the scenarios, but conversely it performs worse than Natural exceptions in about 2\% of the scenarios.
+
+Figure~\ref{fig:avo-bars} shows that the answers for questions $Q_1$ to $Q_3$ are all positive.
+In all three semantics, blame modes outperform their corresponding exception mode (by around 10\% for each one).
+The Natural exceptions mode is never more useful than Natural blame, and  Transient
+exceptions are more useful than Transient first and Transient last blame
+in a small percentage (about 2\%) of the scenarios.
+
+Figure~\ref{fig:avo-bars} also provides answers to $Q_*$.
+Blame for all three semantics is significantly
+more useful than Erasure exceptions (by around 30\%). Natural blame is
+more useful than both versions of Transient blame by a small percentage;
+in about 4\% of scenarios Natural blame is more useful while in about 1.5\% of the
+scenarios Transient blame is more useful. The Transient first and
+Transient last blame are practically indistinguishable. Finally, there is
+no clear winner between Natural exceptions and Transient exceptions
+despite the theoretically advantageous additional checks of the Natural
+semantics.
+
+In order to better understand questions $Q_1$ to $Q_3$, figure~\ref{fig:success-venns} depicts a breakdown of the success of each semantics in comparison to Erasure.
+Specifically, the figure shows one Venn diagram per mode of the rational programmer that uses blame.
+Each diagram breaks breaks down the overlap in the success of the blame mode, its corresponding exception mode, and Erasure.
+For example, in the breakdown of Natural, the diagram shows that all three modes succeed on the same scenario 57.3\% of the time, only Natural and Natural exceptions succeed on 29.1\% of the scenarios, only Natural and Erasure succeed on 2.1\%, and Natural alone succeeds on 9\%.
+This analysis highlights the success tradeoffs each semantics offers against Erasure, with and without blame, to answer questions about the tradeoffs in choosing to use blame or just exceptions for each semantics.
+For instance, the breakdown of Natural clearly illustrates that if choosing between Natural blame, Natural exceptions, and Erasure, Natural blame is the absolutely most successful: all of the successes of the other two modes are subsets of Natural's successful scenarios.
+On the other hand, Transient's blame modes are similar but not so clear-cut.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.30\textwidth]{./plots/success-bars}
+  \caption{Success baselines: The estimated percentage of scenarios for which each mode succeeds in locating the bug.
+  }
+  \label{fig:success-bars}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.9\textwidth]{./plots/avo-bars}
+  \caption{Usefulness comparisons: Each plot depicts a head-to-head comparison of the entitled mode against every other mode.
+  The portion above 0 (in green) is the estimated percentage of scenarios where the entitled mode is more useful than each of the other modes.
+  The portion below 0 (in red) is the estimated percentage of scenarios with the inverse situation.
+  The upper bound of error margin is 0.02\%.}
+  \label{fig:avo-bars}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.32\textwidth]{./plots/TR-TR-stack-first-venn}
+  \hfill
+  \includegraphics[width=0.32\textwidth]{./plots/transient-newest-transient-stack-first-venn}
+  \hfill
+  \includegraphics[width=0.32\textwidth]{./plots/transient-oldest-transient-stack-first-venn}
+
+  \caption{Blame usefulness analysis: Each diagram shows the breakdown of how three modes' successful scenarios overlap.
+  The percentage of scenarios covered by each circle are proportions for which the corresponding mode succeeds.
+  For example, in the breakdown of Natural, all three modes succeed on the same scenario 57.3\% of the time, only Natural and Natural exceptions succeed on 29.1\% of the scenarios, only Natural and Erasure succeed on 2.1\%, and Natural alone succeeds on 9\%.
+  }
+  \label{fig:success-venns}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.8\textwidth]{./plots/bt-lengths-table}
+  \caption{Programmer effort: Each plot depicts the distribution of trail
+  lengths for a given mode across all benchmarks.}
+  \label{fig:effort-table}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.9\textwidth]{./plots/bt-length-comparisons}
+  \caption{Effort comparisons: Each plot depicts the distribution of scenarios with trail length differences ranging from -3 to 3.
+  A negative difference means that the top mode's trail is three steps shorter than the bottom mode's trail for the same scenario, and a positive difference is the reverse.
+  Only scenarios where both modes succeed are considered.
+  The \texttt{.2|.8} on the y-axis indicates that the axis is truncated between 0.2 and 0.8.}
+  \label{fig:effort-comparisons}
+\end{figure}
+
+Turning to programmer effort, figure~\ref{fig:effort-table} shows the distribution of blame trail lengths
+for our sample of interesting debugging scenarios. Unlike the usefulness
+comparisons above, these proportions do not generalize to a representation of the full population.
+Still, as we discuss in section~\ref{subsec:effort}, they provide an
+alternative view of the workings of the rational programmer.
+
+There are
+two immediate take-aways from the figure. First, the effort for successfully
+debugging interesting scenarios (in green) for the random mode of the
+rational programmer follows  a normal distribution, as expected.\footnote{The
+random mode distribution is indistinguishable for all three semantics and thus the figure~\ref{fig:effort-table} 
+only shows one plot.} In contrast, in the other modes, successful effort coalesces at
+the left side of the plot, meaning that in most cases the programmer needs
+to type a single component to debug a scenario.
+
+A second point of interest is that the exceptions of
+the Erasure semantics either help the rational programmer immediately or 
+the rational programmer fails to debug a scenario altogether (in red).
+This is expected; no matter how many type annotations the programmer
+adds to an Erasure program, if the type checker doesn't reject the
+program, running the program always produces the same outcome. Thus an
+exception from the runtime has to point to the buggy module with
+the first try. Otherwise the rational programmer types an irrelevant
+module, runs the program again and the exception points again to the
+already typed module.
+
+Figure~\ref{fig:effort-comparisons} provides head-to-head comparisons of effort.
+To directly compare effort between modes, we compute the difference in trail length between two modes for all scenarios where they both succeed.
+Hence, each plot in the figure shows the distribution of scenarios with length differences ranging from -3 (the first mode's trail is 3 steps shorter than the second) to 3 (the first mode's trail is 3 steps longer than the second).
+The figure offers several insights about the comparative effort for each mode that complement the insights about comparitive success from figure~\ref{fig:avo-bars}.
+First, Natural blame never produces shorter trails than Natural exceptions, and sometimes (albeit rarely) produces slightly longer ones.
+Second, Natural relatively often (around 10\% of the scenarios) produces shorter trails than both Transient blame modes, and they are sometimes significantly shorter.
+Third, Transient last blame often produces shorter trails than Transient first blame, which is a particularly interesting distinction in light of their nearly undistinguishable success rates.
+Finally, Transient's blame modes also share the characteristic with Natural that blame sometimes produces longer trails than their corresponding exception modes.

### 2021-02-28 writing/results.tex
results: update figures and move them around

@@
 over 30,000 hours of compute hours.
+
+\begin{wrapfigure}{r}{.40\textwidth}
+  \centering
+  \includegraphics[width=0.40\textwidth]{./plots/success-bars}
+  \caption{Success baselines: The estimated percentage of scenarios for which each mode succeeds in locating the bug.
+  }
+  \label{fig:success-bars}
+\end{wrapfigure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{./plots/avo-bars}
+  \caption{Usefulness comparisons: Each plot depicts a head-to-head comparison of the entitled mode against every other mode.
+  The portion above 0 (in green) is the estimated percentage of scenarios where the entitled mode is more useful than each of the other modes.
+  The portion below 0 (in red) is the estimated percentage of scenarios with the inverse situation.
+  The upper bound of error margin is 0.02\%.}
+  \label{fig:avo-bars}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.32\textwidth]{./plots/TR-TR-stack-first-venn}
+  \hfill
+  \includegraphics[width=0.32\textwidth]{./plots/transient-newest-transient-stack-first-venn}
+  \hfill
+  \includegraphics[width=0.32\textwidth]{./plots/transient-oldest-transient-stack-first-venn}
+
+  \caption{Blame usefulness analysis: Each diagram shows the breakdown of how three modes' successful scenarios overlap.
+  The percentage of scenarios covered by each circle are proportions for which the corresponding mode succeeds.
+  For example, in the breakdown of Natural, all three modes succeed on the same scenario 57.3\% of the time, only Natural and Natural exceptions succeed on 29.1\% of the scenarios, only Natural and Erasure succeed on 2.1\%, and Natural alone succeeds on 9\%.
+  }
+  \label{fig:success-venns}
+\end{figure}
 
@@
 
-In order to better understand questions $Q_1$ to $Q_3$, figure~\ref{fig:success-venns} depicts a breakdown of the success of each semantics in comparison to Erasure.
-Specifically, the figure shows one Venn diagram per mode of the rational programmer that uses blame.
-Each diagram breaks breaks down the overlap in the success of the blame mode, its corresponding exception mode, and Erasure.
-For example, in the breakdown of Natural, the diagram shows that all three modes succeed on the same scenario 57.3\% of the time, only Natural and Natural exceptions succeed on 29.1\% of the scenarios, only Natural and Erasure succeed on 2.1\%, and Natural alone succeeds on 9\%.
-This analysis highlights the success tradeoffs each semantics offers against Erasure, with and without blame, to answer questions about the tradeoffs in choosing to use blame or just exceptions for each semantics.
-For instance, the breakdown of Natural clearly illustrates that if choosing between Natural blame, Natural exceptions, and Erasure, Natural blame is the absolutely most successful: all of the successes of the other two modes are subsets of Natural's successful scenarios.
-On the other hand, Transient's blame modes are similar but not so clear-cut.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.30\textwidth]{./plots/success-bars}
-  \caption{Success baselines: The estimated percentage of scenarios for which each mode succeeds in locating the bug.
-  }
-  \label{fig:success-bars}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.9\textwidth]{./plots/avo-bars}
-  \caption{Usefulness comparisons: Each plot depicts a head-to-head comparison of the entitled mode against every other mode.
-  The portion above 0 (in green) is the estimated percentage of scenarios where the entitled mode is more useful than each of the other modes.
-  The portion below 0 (in red) is the estimated percentage of scenarios with the inverse situation.
-  The upper bound of error margin is 0.02\%.}
-  \label{fig:avo-bars}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.32\textwidth]{./plots/TR-TR-stack-first-venn}
-  \hfill
-  \includegraphics[width=0.32\textwidth]{./plots/transient-newest-transient-stack-first-venn}
-  \hfill
-  \includegraphics[width=0.32\textwidth]{./plots/transient-oldest-transient-stack-first-venn}
-
-  \caption{Blame usefulness analysis: Each diagram shows the breakdown of how three modes' successful scenarios overlap.
-  The percentage of scenarios covered by each circle are proportions for which the corresponding mode succeeds.
-  For example, in the breakdown of Natural, all three modes succeed on the same scenario 57.3\% of the time, only Natural and Natural exceptions succeed on 29.1\% of the scenarios, only Natural and Erasure succeed on 2.1\%, and Natural alone succeeds on 9\%.
-  }
-  \label{fig:success-venns}
-\end{figure}
-
 \begin{figure}
@@
   \caption{Effort comparisons: Each plot depicts the distribution of scenarios with trail length differences ranging from -3 to 3.
-  A negative difference means that the top mode's trail is three steps shorter than the bottom mode's trail for the same scenario, and a positive difference is the reverse.
+  A negative difference means that the first mode's trail is three steps shorter than the second mode's trail for the same scenario, and a positive difference is the reverse.
   Only scenarios where both modes succeed are considered.
-  The \texttt{.2|.8} on the y-axis indicates that the axis is truncated between 0.2 and 0.8.}
+  The 10|85 on the y-axis indicates that the axis is truncated between 10 and 85\%.}
   \label{fig:effort-comparisons}
 \end{figure}
+
+In order to better understand questions $Q_1$ to $Q_3$, figure~\ref{fig:success-venns} depicts a breakdown of the success of each semantics in comparison to Erasure.
+Specifically, the figure shows one Venn diagram per mode of the rational programmer that uses blame.
+Each diagram breaks breaks down the overlap in the success of the blame mode, its corresponding exception mode, and Erasure.
+For example, in the breakdown of Natural, the diagram shows that all three modes succeed on the same scenario 57.3\% of the time, only Natural and Natural exceptions succeed on 29.1\% of the scenarios, only Natural and Erasure succeed on 2.1\%, and Natural alone succeeds on 9\%.
+This analysis highlights the success tradeoffs each semantics offers against Erasure, with and without blame, to answer questions about the tradeoffs in choosing to use blame or just exceptions for each semantics.
+For instance, the breakdown of Natural clearly illustrates that if choosing between Natural blame, Natural exceptions, and Erasure, Natural blame is the absolutely most successful: all of the successes of the other two modes are subsets of Natural's successful scenarios.
+On the other hand, Transient's blame modes are similar but not so clear-cut.
 

### 2021-03-01 writing/challenges.tex
re-organize and re-write secs. 1 thru 6

@@
 
-The {\em first derivative challenge\/} is to find a representative collection of
-impedance mismatches. They must represent mistakes that programmers accidentally
-create and that the run-time checks of academic systems catch. In other words,
-the experiment calls for a collection of type mistakes in mix-typed programs
-that is representative of those ``in the wild.''  Unfortunately no such
-collection exists, and with good reason. The kind of mistakes needed are
-typically detected by unit or integration tests; even if it takes some time to
-find their sources, these mistakes do not make it into code repositories with
-appropriate commit messages.
+The rational programmer works like an algorithm. Implementing it demands a large
+number of representative debugging scenarios; a reconciliation of three
+different kinds of error messages; and a proper sampling strategy for the huge
+space of possibilities. A coincidental challenge is the disparity of gradually
+typed languages. To eliminate this variable from the comparison, this paper uses
+Racket, which is thus far the only language in which all three major soundness
+mechanisms are available in a reasonably robust and directly comparable manner;
+for the implementation of the transient semantics, see the parallel submission
+of an experience report~\cite{ttt21}.
+
+The {\em first challenge\/} is to find a representative collection of impedance
+mismatches. They must represent mistakes that programmers accidentally create
+and that the run-time checks of academic systems catch. In other words, the
+experiment calls for a collection of type mistakes in mix-typed programs that is
+representative of those ``in the wild.''  Unfortunately no such collection
+exists, and with good reason. The kind of mistakes needed are typically detected
+by unit or integration tests; even if it takes some time to find their sources,
+these mistakes do not make it into code repositories with appropriate commit
+messages.
 
@@
 precisely those that are needed for evaluating blame assignment strategies.
+
+Based on a related experience, Gopinath and Walkingshaw~\cite{gw-mutation} write,
+``existing mutation frameworks \ldots\ do not generate the kinds of mutations
+needed to best evaluate type annotations'' and, worse, ``it is surprisingly
+difficult to come up with mutants that actually describe subtle type faults.''
+While the goal of their work---to evaluate the quality of types in
+Python---is unrelated to the one here, the mechanism is related. And their
+judgment confirms the experience of the authors.
 
@@
 
-The {\em second derivative challenge\/} concerns the differences among the blame
-assignment of the three semantic variants.  While Natural assigns blame to {\em
-one\/} component, Transient assigns blame to a sequence of components. The
-Erasure semantics does not blame a component {\it per se\/}, but it comes with
-an exception location and a stack trace, which implicitly suggests parties to
+The {\em second challenge\/} concerns the differences among the blame assignment
+of the three semantic variants.  While Natural assigns blame to {\em one\/}
+component, Transient assigns blame to a sequence of components. The Erasure
+semantics does not blame a component {\it per se\/}, but it comes with an
+exception location and a stack trace, which implicitly suggests parties to
 blame.  The different strategies greatly affect the behavior of the rational
@@
 
-Overcoming these challenges yields a systematic approach: 
-\begin{itemize}
+The next three sections explain how to overcome the three challenges
+in detail.
 
-\item interesting mutators generate interesting type-level errors and thus
-debugging scenarios (section~\ref{sec:mutate});
+% Overcoming these challenges yields a systematic approach: 
+% \begin{itemize}
 
-\item an algorithm simulates the modes of the rational programmer on these
-scenarios (section~\ref{sec:rational}); and
+% \item interesting mutators generate interesting type-level errors and thus
+% debugging scenarios (section~\ref{sec:mutate});
 
-\item a careful sampling of the space yields experimental
-data about the bug finding process (section~\ref{sec:sample}).
-\end{itemize}
-Getting all three aspects right is necessary to obtain a computationally
-feasible and statistically sound experiment.
+% \item an algorithm simulates the modes of the rational programmer on these
+% scenarios (section~\ref{sec:rational}); and
 
-{\bf Note} A coincidental challenge is the disparity of gradually typed
-languages. While the three approaches to blame are well understood in theory,
-they have been implemented for rather different untyped languages and with
-vastly different type systems. To eliminate the effect of these differences, it
-is necessary to implement all three approaches in a single language so that an
-application of the evaluation method can use the same test programs, the same
-type system, and hence the same debugging scenarios.
+% \item a careful sampling of the space yields experimental
+% data about the bug finding process (section~\ref{sec:sample}).
+% \end{itemize}
+% Getting all three aspects right is necessary to obtain a computationally
+% feasible and statistically sound experiment.
 
-The results in this paper are for a system that builds on Greenman and
-Felleisen's effort of comparing the performance of Natural and Transient in
-Racket~\cite{gf-icfp-2018}. The extended system uses the type system of Typed
-Racket and comes with a Transient semantics that faithfully implements the
-design and the blame assignment strategy proposed by~\cite{vss-popl-2017}. The
-Erasure semantics drops the type information and runs the programs as if they
-were plain Racket programs. The next section explains this system in detail.
+% {\bf Note} A coincidental challenge is the disparity of gradually typed
+% languages. While the three approaches to blame are well understood in theory,
+% they have been implemented for rather different untyped languages and with
+% vastly different type systems. To eliminate the effect of these differences, it
+% is necessary to implement all three approaches in a single language so that an
+% application of the evaluation method can use the same test programs, the same
+% type system, and hence the same debugging scenarios.
+
+% The results in this paper are for a system that builds on Greenman and
+% Felleisen's effort of comparing the performance of Natural and Transient in
+% Racket~\cite{gf-icfp-2018}. The extended system uses the type system of Typed
+% Racket and comes with a Transient semantics that faithfully implements the
+% design and the blame assignment strategy proposed by~\cite{vss-popl-2017}. The
+% Erasure semantics drops the type information and runs the programs as if they
+% were plain Racket programs. The next section explains this system in detail.

### 2021-03-01 writing/introduction.tex
re-organize and re-write secs. 1 thru 6

@@
 
-Industrial implementors of gradual typing systems have almost completely ignored
-blame assignment.  Systems such as Flow,\footnote{\url{https://flow.org}}
-Hack,\footnote{\url{https://hacklang.org}} or
-TypeScript\footnote{\url{https://www.typescriptlang.org}} exploit types for IDE
-actions and for finding typos in code. Then their compilers remove types and
-rely on the built-in safety checks of the underlying language to catch any
-problems.
+Industrial implementors of gradual typing systems have almost
+completely ignored blame assignment.  Systems such as Flow, Hack, or
+TypeScript\footnote{See \url{https://flow.org},
+\url{https://hacklang.org}, and \url{https://www.typescriptlang.org},
+respectively.} exploit types for IDE actions and for finding typos in
+code. Then their compilers remove types and rely on the built-in
+safety checks of the underlying language to catch any problems.
 
@@
 
-This paper's first contribution is {\em an automated method for evaluating the
+This paper's {\em first contribution\/} is {\em a method for evaluating the
 effectiveness of blame assignment strategies\/} in the gradual typing world.
 The top-level innovation is the idea of a {\em rational programmer\/}, that is,
-a programmer that acts only in response to available information (see section~\ref{sec:why-rational}). In the case
-of an impedance mismatch, the available information consists of the error
-message and the current state of the program. The rational programmer can hence
-use the former to change the latter---and this systematic, information-driven
-process can be implemented.\footnote{\citet{cc-snapl-19} use this tracing
-method to localize type errors reported by gradual typing systems based on
-inference.} Implementing it means overcoming three major challenges: injecting
-representative impedance mismatches; putting the various kinds of error
-information to comparable use; and sampling the huge space of possibilities
-properly (see section~\ref{sec:challenges}).
+a programmer that acts only in response to available information (see
+sec.~\ref{sec:why-rational}). In the case of an impedance mismatch, the
+available information consists of the error message and the current state of the
+program. The rational programmer can hence use the former to change the
+latter---and this systematic, information-driven process can be
+implemented.\footnote{\citet{cc-snapl-19} use this tracing method to localize
+type errors in inference-based gradual typing systems.}  Implementing it means
+overcoming major challenges: injecting representative impedance mismatches;
+putting the various kinds of error information to comparable use; and sampling
+the huge space of possibilities (see sec.~\ref{sec:challenges}).
 
-% The method simulates a programmer who follows the Wadler-Findler slogan quoted
-% above; its automation borrows elements of of Lazarek et
-% al.~\cite{lksfd-popl-2020}'s recent work.  Such a \emph{rational programmer} can find impedance-mismatch
-% bugs by adding types to a partially typed program (see
-% section~\ref{sec:landscape}). That is, when the run-time system signals an
-% impedance mismatch, a good blame message helps the programmer infer where to
-% add an additional type specification because, by the basic theory, well-typed
-% components cannot cause impedance mismatches. Otherwise, if the additional type
-% does {\em not\/} cause a static type error, the program is run again to signal
-% another violation. Then the programmer can heed blame one more time to add yet
-% another type specification. Evaluating the quality of blame assignment thus
-% becomes a question of whether the tracing the error messages finds the source
-% of the mistake and how many steps it takes to get to this point.
-
-The paper's second contribution is a set of {\em results from applying the
-evaluation method to three distinct blame assignment strategies and
-approximately 72,200 cases\/} (see section~\ref{sec:results}): (1) {\it
+The paper's {\it second contribution\/} is a set of {\em results from applying
+the evaluation method to three distinct blame assignment strategies and
+approximately 72,200 cases\/} (see sec.~\ref{sec:results}): (1) {\it
 Transient\/}, i.e. Reticulated's tracking of typed/untyped boundaries; (2) {\it
@@
 assignment~\cite{ff-icfp-2002}; and (3) {\it Erasure\/}, i.e. the approach of
-industrial systems, which forego blame assignment and rely on error messages
-from the safety checks of the underlying programming language.
-
-The results of the method's first application (see section~\ref{sec:discussion})
-are at least somewhat surprising. In principle they validate the conjectures
-behind the work of theoreticians. First, a good blame assignment strategy helps
-with the search for impedance mismatches between the specified types and the
-behavior of untyped components.  Second, Natural's wrapper-based blame tracking
-is more useful than Transient's ``collective blame'' tracking algorithm, which
-in turn is superior to Erasure. {\em But\/}, the application of the method also
-indicates problems with the expectations of theoreticians. The first problem is
-that the cost of Transient's blame can be huge. The second one concerns the
-difference between the methods; neither is {\it Natural\/} vastly better than
-{\it Transient\/} nor are the two blame assignment methods clearly preferable to
-{\it Erasure\/}.  In turn, these problems suggests that, on one hand, the
-existing theory does not predict practice properly, and on the other hand, the
-existing practice may need to find additional ways to collect data.
+industrial systems, which forgo blame in favor of error messages from the safety
+checks of the underlying language.---The results (see
+sec.~\ref{sec:discussion}) are at least somewhat surprising.  In principle
+they validate the conjectures behind the work of theoreticians.  First, a good
+blame assignment strategy helps with the search for impedance mismatches between
+the specified types and the behavior of untyped components.  Second, Natural's
+wrapper-based blame tracking is more useful than Transient's ``collective
+blame'' tracking algorithm, which in turn is superior to Erasure. {\em But\/},
+the application of the method also indicates problems with the expectations of
+theoreticians. The first problem is that the cost of Transient's blame can be
+huge. The second one concerns the difference between the methods; neither is
+{\it Natural\/} vastly better than {\it Transient\/} nor are the two blame
+assignment methods clearly preferable to {\it Erasure\/}.  In turn, these
+problems suggests that, on one hand, the existing theory does not predict
+practice properly, and on the other hand, the existing practice may need to find
+additional ways to collect data.

### 2021-03-01 writing/landscape.tex
re-organize and re-write secs. 1 thru 6

@@
-%% -----------------------------------------------------------------------------
-\def\fname#1#2{$\underline{\hbox to #2in{#1}}$\\[-2ex]}
-
-\begin{figure*}[hbt]\footnotesize
-  \vspace{-1mm}
-
-\begin{minipage}[t]{2.2in}
-\begin{alltt}\fname{untyped/pack-lib}{1.8}
-(provide pack _ _ _)
-
-_ _ _  dependencies     _ _ _
-_ _ _  and definitions  _ _ _
-
-(define (pack d)
-  ;; process JSON data and 
-  ;; package it in a dictionary
-  ;; (association list)
-  (make-hash _ _ _) ;; BUG!)
-
-\end{alltt}
-\end{minipage}\begin{minipage}[t]{2.2in}
-\begin{alltt}\fname{typed/pack-lib}{1.7}
-(provide typed-pack)
-
-(require/typed pack-lib
-  [pack 
-  (-> JSON 
-      (Listof 
-       (Pairof Symbol String)))])
-
-(define typed-pack pack) 
-\end{alltt}
-\smallskip
-\begin{alltt}\fname{typed/crypto-pack-lib}{1.7}
-(provide crypto-pack)
-
-(require/typed pack-lib
- [pack 
-  (-> JSON 
-      (Listof 
-       (Pairof Symbol String)))])
-
-(define (crypto-pack [d : JSON])
-  (pack (encrypt d _ _ _)))
-
- \end{alltt}
-\end{minipage}\begin{minipage}[t]{2.2in}
-\begin{alltt}\fname{untyped/client}{1.65}
-(require json 
-         typed/pack-lib
-           typed/crypto-pack-lib)
-
-_ _ _  other dependencies  _ _ _
-_ _ _  and definitions     _ _ _
-
-;; read data from files, pack 
-;; and share securely
-
-(define public-data 
-  (typed-pack 
-   (read-json 
-    "public-records")))
-
-(define secret-data 
-  (crypto-pack 
-   (read-json 
-    "medical-records")))
-
- _ _ _ rest of client _ _ _
-\end{alltt}\end{minipage}
-
-   \caption{One mix-typed program, three interpretations} \label{fig:example}
-
-  \vspace{-1mm}
-\end{figure*}
 %% -----------------------------------------------------------------------------
 
-The program sketch in figure~\ref{fig:example} demonstrates the
-differences between Natural, Transient and Erasure using Typed Racket
-syntax.  The modules on the left and right are untyped while the two
-in the middle are typed.
+The three blame strategies rely on three different ways of catching problems
+with types at run-time: {\em Natural\/}, {\em Transient\/}, and {\em
+Erasure\/}. This section summarizes these options with one illustrative example
+using Racket syntax. The informed reader may wish to merely scan the
+self-contained section.
 
-The \texttt{untyped/pack-lib} module (left) provides, among others, a function
-\texttt{pack} that consumes JSON data and packages it in a dictionary. However,
-by mistake \texttt{pack} returns a hash table instead of the association list
-its documentation promises.  The \texttt{client} module (right) uses
-\texttt{pack} indirectly to package public and sensitive data it reads from the
-filesystem. It does so through two different intermediary typed modules
-(center). The first, {\tt typed/pack-lib}, imports \texttt{pack} and re-exports
-it as \texttt{typed-pack} asserting that it is a function that consumes {\tt
-JSON} and returns a list associating {\tt Symbol}s with {\tt String}s. The
-second typed module, {\tt typed/crypto-pack-lib}, also imports \texttt{pack} and
-assumes for it the same type as {\tt typed/pack-lib}. Instead of exporting it,
-it uses the function in the exported {\tt crypto-pack} function, which encrypts
-its input before passing it to \texttt{pack}.
+Consider the program sketch in figure~\ref{fig:example}. Each box represents a
+module: the top bar lists the name and whether it is using typed (\typecolor) or
+untyped (\dyncolor) syntax:
+\begin{description}
 
-Under the {\it Natural\/} semantics~\cite{tf-popl-2008,tfffgksst-snapl-2017},
-functions imported into and exported from typed modules are wrapped in proxies
-that enforce the static type discipline with run-time checks and track
-responsibilities. Thus, when \texttt{pack} is imported into a typed module, the
-run-time system checks that it is a function and wraps it in a protective proxy,
-which in turn, enforces the type of the function with run-time checks.  In
-essence, this check protects typed modules when they call an untyped function
+\item[\texttt{pack-lib}] (at the top right) represents a library that provides,
+among others, a function {\tt pack}. As the documentation says, this function
+consumes JSON data and packages it in a dictionary. by mistake, though, the
+function returns a hash table instead of the association list its documentation
+promises.
+
+\item[\texttt{types}] (at the top left) is one of three modules that overlays
+types onto this library. This specific modules defines types in common to the
+two other typed libraries.
+
+\item[\texttt{typed-pack-lib}] (at the mid-level on the left), imports {\tt pack}
+and re-exports it as \texttt{typed-pack} asserting that it is a function that
+consumes {\tt JSON} and returns a list associating {\tt Symbol}s with {\tt
+String}s. In other words, it formalizes the comments in {\tt pack-lib}.
+
+\item[\texttt{crypto-pkac-lib}] (at the bottom left) also imports \texttt{pack}
+and assumes for it the same type as {\tt typed-pack-lib}. It applies the
+function in the definition of the exported {\tt crypto-pack} function, which
+encrypts its input before passing it to \texttt{pack}.
+
+\item[\texttt{client}] (at the bottom right) uses {\tt pack}
+indirectly. Specifically, it goes through the two intermediary typed modules to
+import them. Imagine a programmer who relies on the types in the \typecolor\
+modules as checked documentation but prototypes the client in an untyped manner.
+
+\end{description}
+Each of the three semantics assigns a different outcome to the client module and
+signals different kinds of errors.
+
+Under {\it the Natural semantics\/}, functions imported into and exported from
+typed modules are wrapped in proxies that enforce the static type discipline
+with run-time checks and track responsibilities~\cite{tf-popl-2008,
+tfffgksst-snapl-2017}. Thus, when {\tt pack} is imported into a typed module,
+the run-time system checks that it is a function and wraps it in a protective
+proxy, which in turn, enforces the type of the function with run-time checks.
+In essence, this check protects typed modules when they call an untyped function
 because it may not produce the promised kind of value, as is the case here with
-{\tt pack} from {\tt untyped/pack-lib}. Analogously the run-time system wraps
-exported functions such as {\tt crypto-pack} of a typed modules in a proxy that
-checks its arguments.  This check protects functions exported from type modules
-against applications in untyped code that supply the wrong kind of argument
-value.
+{\tt pack} from {\tt pack-lib}. Analogously the run-time system wraps exported
+functions such as {\tt crypto-pack} of a typed modules in a proxy that checks
+its arguments.  This check protects functions exported from type modules against
+applications in untyped code that supply the wrong kind of argument value.
 
 As this analysis implies, if the return-type check fails, the problem is with
-the untyped module that supplied an incorrect function, here {\tt
-untyped/pack-lib}. If the argument-type check fails, responsibility lies with
-{\tt client}. Due to the proxies wrapped around functions, Natural can easily
-track the responsible party for each check. Thus, in the example of
-figure~\ref{fig:example}, as \texttt{typed-pack} returns, the return-type check
-fails, and Natural blames \texttt{pack-lib}.
+the untyped module that supplied an incorrect function, here {\tt pack-lib}. If
+the argument-type check fails, responsibility lies with {\tt client}. Due to the
+proxies wrapped around functions, Natural can easily track the responsible party
+for each check. Thus, in the example of figure~\ref{fig:example}, as
+\texttt{typed-pack} returns, the return-type check fails, and Natural blames
+\texttt{typed-pack-lib}.
 
-Under the Transient semantics~\cite{vss-popl-2017}, typed code is compiled so
-that all entry points to functions check their arguments at run time and all
-function calls check their return values against the expected type.
-Furthermore, Transient uses \emph{shallow} checks, meaning they check only the
+Under {\it the Transient semantics\/}, typed code is compiled so that all entry
+points to functions check their arguments at run time and all function calls
+check their return values against the expected type~\cite{vss-popl-2017}.
+Furthermore, Transient uses \emph{shallow} checks, meaning they inspect only the
 top constructor of a value. Since retrieving a value from within a structure (or
@@
 
-As a result, the call to \texttt{typed-pack} does not signal an error because
-the call takes place in the untyped {\tt client} module, which is compiled in
-the usual manner. Because {\tt pack} is called in the {\tt
-typed/crypto-pack-lib} module, Transient's inlined checks make sure that the
-imported \texttt{pack} is a function and that the result of its call in is a
-list. This last check fails when \texttt{client} calls to {\tt crypto-pack}.
+As a result, the call to \texttt{typed-pack} does {\em not\/} signal an error
+because it takes place in the untyped {\tt client} module, which is compiled in
+the usual manner. Because {\tt pack} is called in the {\tt crypto-pack-lib}
+module, Transient's inlined checks make sure that the imported
+\texttt{pack} is a function and that the result of its call in is a list. This
+last check fails when \texttt{client} calls to {\tt crypto-pack}.
 
@@
 they cross, plus the corresponding types. In the example, the map records that
-\texttt{pack} crosses from \texttt{untyped/pack-lib} to {\tt typed/pack-lib} and
-from {\tt typed/pack-lib} to {\tt client} with the type that appears in the {\tt
-required/typed} forms in the example. Since the failed check corresponds to the
-return type of \texttt{pack}, Transient blames the sources of the two boundary
-crosses, which in this particular case is \texttt{untyped/pack-lib} for both. In
+\texttt{pack} crosses from \texttt{pack-lib} to {\tt crypto-pack-lib} and
+from {\tt crypto-pack-lib} to {\tt client} with the type that appears in the
+{\tt required/typed} forms in the example. Since the failed check corresponds to
+the return type of \texttt{pack}, Transient blames the sources of the two
+boundary crossings, which in this particular case is \texttt{pack-lib}. In
 general though, Transient blames more than one party. In fact, the theoretical
@@
 
-Under the Erasure semantics, the compiler checks the specified types and then
-discards them when it generates code. The generated code includes run-time
+Under {\it the Erasure semantics\/}, the compiler checks the specified types and
+then discards them when it generates code. The generated code includes run-time
 checks that ensure the dynamic safety of all operations as specified in the
 underlying untyped language. Hence, neither the call to {\tt typed-pack} nor the
-call to {\tt crypto-pack} signals an error due to the gradual type system.
+call to {\tt crypto-pack} signals an error due to the gradual type system. When,
+at some later point, {\tt client} tries to inspect the elements of the lists
+that \texttt{typed-pack} and {\tt crypto-pack} are supposed to produce, Racket's
+safety check signal a violation and point to some place in {\tt client}. The
+information in this exception, plus its stack trace, may help the programmer
+find the source of the impedance mismatch between the specified return types of
+{\tt pack} in the two typed modules and the actual return value of {\tt pack}.
 
-When run, the program may raise an exception if \texttt{client} tries to
-inspect the elements of the list that \texttt{typed-pack} and
-\texttt{crypto-pack} are supposed to produce. The information in this exception,
-plus its stack trace, or may not help the programmer find the source of the
-impedance mismatch between the specified return types of {\tt pack} in the two
-typed modules and the actual return value of {\tt pack}.
-
-All three systems used for the experiment are variants of Racket. The
-use the Typed Racket syntax and type checker for all three
-semantics. The Natural semantics is inherited as is, while the Erasure
-semantics is the one of plain Racket.  The Transient system implements
-the design of~\citet{vss-popl-2017}, though constructing it demands
-two design decisions: which shallow checks to use and where, plus how
-to maintain a blame map in a full-fledged run-time system.
-
-For certain types, such as {\tt String} or {\tt (Vectorof Byte)}, Transient can
-realize shallow checks with the predicate from Racket.  Other types require
-custom predicates; for example the type {\tt Index} of valid array indices
-allows a range of positive exact integers.  Many custom predicates run in
-constant time, but some, such as the one for {\tt (Union T1 ... Tn)}, run in
-linear time. The key point is that no shallow check inspects beyond the top
-constructor of each type.
-
-Once the type checker has completed its work, Transient rewrites the
-source to insert run-time checks. Technically speaking, the rewriting
-pass wraps all elimination in applications of shallow checks.  For
-example, if the variable {\tt f} has the static type {\tt (-> String
-String)}, the application {\tt (f x)} must be rewritten to check for a
-string result because {\tt f} could be an untyped function.
-
-The rewriting pass also inserts hooks for updating the blame map.  When a value
-crosses a module boundary, such a hook records a type assumption in the blame
-map.  The in-lined checks for elimination forms extend the blame map with
-further information. For instance, the check of the result of {\tt (f x)} adds
-an entry to the blame map that records the expected type of the result as the
-return type of the type of {\tt f}.  If a Transient check ever fails, the
-run-time checking system of Transient Racket walks over the information in the
-blame map and reports a sequence of blamed components.
-
-%% MF: I have no clue what the following contributes:
-
-% \smallskip
-% \noindent
-% {\bf Note.} As the common basis for the three semantics, we choose Typed
-% Racket's migratory typing. In contrast to the initial formulation of gradual
-% typing~\cite{st-sfp-2006}, migratory typing does not provide support for type
-% \texttt{Dyn}, components are either untyped or fully typed.
-% %% MF: does it ever:
-% %% and the gradual type system does not affect the compilation of untyped components. 
-% Despite, these differences, Typed Racket is the most
-% mature system for sound mix-typed code
-% and recent theoretical work has established that
-% its migratory typing offers a setting for comparing different approaches to gradual
-% typing~\cite{gf-icfp-2018,gfd-oopsla-2019}.
+The following table summarizes the illustration:
+%% -----------------------------------------------------------------------------
+\begin{center}
+\begin{tabular}{l|ll}
+                                 \multicolumn{3}{c}{the evaluation of} \\
+\relax                  & {\tt public-data}             & {\tt secret-data} in {\tt client}, \\
+{under the $\cdot$ semantics yields} & & \\ \hline
+%% ----------------------------------------------------------------------------------------
+\qquad {\it Natural\/}   & error, blaming              & error, blaming \\
+     		   	 & \quad {\tt typed-pack-lib}  & \quad {\tt typed-pack-lib} \\
+\qquad {\it Transient\/} & no error   		       & error, blaming the {\tt pack-lib}/ \\
+      		  	 &    			       & \quad {\tt crypto-pack-lib} boundary \\
+\qquad {\it Erasure\/}   & no error 		       & no error \\
+\end{tabular}
+\end{center}
+%% -----------------------------------------------------------------------------

### 2021-03-01 writing/mutate-mutators.tex
proper voice and style
re-organize and re-write secs. 1 thru 6

@@
-Generating debugging scenarios from the benchmarks demands the design
-of novel mutators. Generally speaking, a
-mutator~\cite{jia2011analysis,demillo1978hints,lipton1971fault}
-performs a localized syntactic change to a code basis. For the
-evaluation of a blame strategy, the mutator must produce type-level
-mistakes that the run-time checks of gradual typing systems or the
-safety checks of the underlying language can detect. Once detected,
-the rational programmer can attempt to detect the mistake by gradually
-adding types to blamed modules.
+% -----------------------------------------------------------------------------
 
-Figure~\ref{table:mutation-ops} describes the sixteen mutators used
-for the data reported here. They are inspired by the authors
-decades-long experience of making type-level mistakes in Typed Racket,
-some of which take a non-trivial effort to debug.  For each mutaror,
-the figure provides a short description and an example mutation.
+A {\it mutator\/} performs a localized syntactic change to a code base. The
+result is a {\em mutant\/}.
 
-Most of these mutators are self-explanatory.  The first twelve apply
-to most gradually typed languages, including those with classes.  The
-last four target distinguishing features of Typed Racket.
-Specifically, {\tt arithmetic} may replace a {\tt +} with a {\tt -} in
-an attempt to change the type of the result of the arithmetic
-operation. In Typed Racket, {\tt +}'s result is a {\tt
-Positive-Integer} when all arguments are positive
-integers~\cite{stathff-padl-12}. However, the result of {\tt -}
-is {\tt Integer}. In the same spirit, the {\tt boolean} mutator aims
-to take advantage of Typed Racket faithfulness to Racket's
-``falsiness.'' Finally, {\tt negate-cond} and {\tt force-cond} attempt
-to confuse occurrence typing.
+For the evaluation of a blame strategy, the mutator must produce type-level
+mistakes that the run-time checks of gradual typing systems or the safety
+checks of the underlying language can detect. Once detected, the rational
+programmer should be able to detect the mistake by gradually adding types to
+blamed modules.
+
+Figure~\ref{table:mutation-ops} describes the 16 mutators used for the
+experiment.  They are inspired by the authors decades-long experience
+of making type-level mistakes in Typed Racket, some of which take a
+non-trivial effort to debug.  For each mutaror, the figure provides a
+short description and an example mutation.
+
+Most mutators are self-explanatory.  The first twelve apply to most
+gradually typed languages, including those with classes.  The last
+four target distinguishing features of Typed Racket.  Specifically,
+{\tt arithmetic} may replace a {\tt +} with a {\tt -} in an attempt to
+change the type of the result of the arithmetic operation. In Typed
+Racket, {\tt +}'s result is a {\tt Positive-Integer} when all
+arguments are positive integers~\cite{stathff-padl-12}. However, the
+result of {\tt -} is {\tt Integer}. In the same spirit, the {\tt
+boolean} mutator aims to take advantage of Typed Racket faithfulness
+to Racket's ``falsiness.'' Finally, {\tt negate-cond} and {\tt
+force-cond} attempt to confuse occurrence typing.

### 2021-03-01 writing/mutate.tex
re-organize and re-write secs. 1 thru 6

--- 2021-02-22 writing/mutate.tex
@@
 
-As section~\ref{sec:challenges} explains, the first challenge is to
-generate a representative mistakes in representative software. The
-following two subsections explain in reverse order.
+Putting a rational programmer to work means generating many mutants and turning
+those into debugging scenarios. Since existing mutators do not generate useful
+mutants, the first step is to develop new mutators (see sec.~\ref{sub:mutate-mutators})
+and to validate their suitability for this application (see sec.~\ref{sub:mutate-interesting}).
+
+% As section~\ref{sec:challenges} explains, the first challenge is to
+% generate a representative mistakes in representative software. This
+% section explains the specific choices made for the first application
+% of the method: the source and nature of the chosen benchmarks
+% (sec.~\ref{sub:mutate-benchmarks}), the newly developed mutators
+% (sec.~\ref{sub:mutate-mutators}), and why they are suitable for this
+% experiment (sec.~\ref{sub:mutate-interesting}).
 
@@
 
-\sub{benchmarks} {Which Software to Use As Benchmarks}
-\sub{interesting}{Are These Mutators Interesting?}
+\input{mutate-mutators-figure.tex}
+\sub{mutators}   {How to Mutate Software} 
+\sub{interesting}{Are These Mutators Interesting}

### 2021-03-01 writing/paper.tex
Switch to icfp format
revise intro to point to new 'why rational' section
re-organize and re-write secs. 1 thru 6

@@
-\documentclass[sigplan,10pt,review,anonymous]{acmart}
+\documentclass[acmsmall,review,anonymous]{acmart}
 \settopmatter{printfolios=true,printccs=false,printacmref=false}
@@
 \usepackage{alltt}
+\usepackage{tcolorbox}
 \usepackage{wrapfig}
@@
 
+\input{landscape-program}
 \sec{introduction}{Does Blame Matter}
-\sec{challenges}  {Why Is It Hard to Evaluate Blame}
-\sec{landscape}   {How to Think About All Three Systems}
+\sec{landscape}   {How to Think About Three Blame Systems}
+\sec{why-rational}{What Is a Rational Programmer}
+\sec{challenges}  {Why It Is Hard to Implement the Rational Programmer}
 \sec{mutate}      {How to Make Lots of Mistakes}
-\sec{rational}    {How to Test the Rational Programmer}
-\sec{sample}      {How to Sample the Experimental Space}
+\sec{rational}    {How to Make Comparable Rational Programmers}
+\sec{sample}      {How to Make the Experimental Space}
 \sec{results}     {What We Discovered}

### 2021-03-01 writing/results.tex
re-organize and re-write secs. 1 thru 6

@@
+\input{mutate-benchmarks-figure.tex}
+\sub{benchmarks} {Which Software to Use As Benchmarks}
+
+\subsection{XXX}
 We run the experimental process from section~\ref{sec:rational} on the

### 2021-03-01 writing/why-rational.tex
re-organize and re-write secs. 1 thru 6

@@
 data. As~\citet{lksfd-popl-2020} recently demonstrated, the way around this
-dilemma is to postulate the {\em rational programmer\/}---the equivalent of a
-{\it homo economicus\/} for programming languages---and to implement it as an algorithm.
-
-In the context of gradual typing, the rational programmer translates the
-Wadler-Findler slogan into a debugging method, searching for the source of the
-impedance mismatch in an incremental fashion. Measuring this simulated
-programmer's behavior on a large number of debugging scenarios yields data that
-is similar to data collected in an empirical, human-facing manner.
+dilemma is to simulate a programmer's activities algorithmically on a large set
+of programming scenarios.
 
 \input{why-rational-case.tex}
+
+This paper dubs Lazarek et. alii's idea {\em rational programmer\/}. Like Stuart
+Mill's {\it homo economicus\/}, the rational programmer approximates the
+behavior of a software developer who reduces time spent on a task given the
+available information. In the context of gradual typing, the rational programmer
+has two pieces of information when an impedance mismatch signals exceptional
+behavior: the error message and the state of the program. Hence, the most
+rational procedure is to use the former to improve the latter. Put differently,
+a rational programmer translates the Wadler-Findler slogan into a debugging
+method, searching for the source of the impedance mismatch with additional type
+annotations. Measuring this simulated programmer's behavior on a large number of
+debugging scenarios yields data that is similar to data collected in an
+empirical, human-facing manner.
+
+The idea is most easily illustrated with an example from the Typed Racket
+world. In the context of Typed Racket, researchers speak of migratory
+typing. Imagine a code base consisting of dozens of modules in plain Racket.  A
+developer who modifies a module in this collection for maintenance purposes must
+study the module designs---reify the types that went into its creation---and to
+help future maintainers should report these insights as type annotations. Over
+time, the code base turns into a mix of typed and untyped modules.
+
+Now consider the concrete (and simplistic) example of figure~\ref{fig:rational}.
+Initially the code base consists of the two \dyncolor\ module on the left plus
+the \typecolor\ module at the bottom; \dyncolor\ indicates untyped, while
+\typecolor\ means typed. When a typed module imports an untyped module, it must
+assign types to the imported identifiers for the type checker's sake. Here {\tt
+main} specifies that {\tt na-client} consumes a {\tt Real} number and produces a
+non-positive one.\footnote{Typed Racket's type system reflects Racket's subset
+reasoning about numbers, not (only) machine-level
+representations~\cite{stathff-padl-12}.} A program execution ends in this error:
+\begin{verbatim}
+    na-client: broke its own contract
+      promised: (<=/c 0)
+      produced: 40
+      in: (-> any/c (<=/c 0))           
+      contract from: (interface for na-client)
+      blaming: (interface for na-client)
+       (assuming the contract is correct)
+\end{verbatim}
+The referenced contract is the compilation of the type of {\tt na-client}. The
+definitive hint is ``blaming: (interface for na-client)'' with the caveat
+``(assuming the contract is correct).''
+
+Assuming the rational programmer trust the type of {\tt na-client}, the next
+step is to inspect the {\tt layer} module and to equip it with type
+annotations. The result is the \typecolor\ module in the middle, and {\tt
+main}'s import is now re-directed there: {\tt (submod ".." layer)}. Running the
+program again yields this message:
+\begin{verbatim}
+    neg-abs: broke its own contract
+      promised: (<=/c 0)
+      produced: 10
+      in: (-> any/c (<=/c 0))
+      contract from: (interface for neg-abs)
+      blaming: (interface for neg-abs)
+       (assuming the contract is correct)
+\end{verbatim}
+The last step of the rational programmer is to assign types to the {\tt server}
+module. At this point the type checker objects to the conjecture type of {\tt
+neg-abs}---and the source of the impedance mismatch is found. How to fix it is a
+separate question.
+
+Like {\it homo economicus\/}, the rational programmer is an
+approximation.  People don't behave purely rationally as economic
+actors, and they also don't do so as software developers. The point is
+not to deny the existence of ``lucky hunches'' programmers or
+``tinkering works'' approaches and so on. It is also not to deny that
+equipping entire modules with types represents an always feasible
+approach.  But, the concept of studying the idea of an economically
+rational actor has produced benefits to political economics, and this
+paper suggests that implementing and studying the rational programmer
+will help language designers.

### 2021-03-01 writing/mutate-benchmarks.tex
adding prose for mutators design and benchmarks selection

@@
-Over the past decade, \citet{gtnffvf-jfp-2019} have curated a collection of
-Typed Racket programs for systematically measuring the implementation's
-performance. The benchmark suite consists of fully typed, correct programs,
-written by a number of different authors, who have maintained and evolved these
-programs over several years. The programs range widely in size, complexity,
-purpose, and in programming style. The latter is mostly a reference to their
-reliance on many Typed Racket features: occurrence typing~\cite{tf-icfp-2010},
-types for mutable and immutable data structures~\cite{hpst-sfp-2010}, types for
-first-class classes and objects~\cite{tsdtf-oopsla-2012}, and types for Racket's
-numeric tower~\cite{stathff-padl-12}.
+Since this is the first empirical attempt to evaluate blame in gradual
+typing, there is no standard benchmark suite that we can use off-the-self.
+In addition, there is no exsitng benchmark suite of gradually typed
+programs with type-level mistakes that we could
+repurpose.\footnote{Recently ~\citet{cc-oopsla-20} created a
+Reticulated Python program collection to evaluate their technique for
+fixing mistakes in type annotations; unfortunately their collection does
+not come with type-level mistakes in the code proper.} In fact, the bar
+for testing the rational programmer is much higher than the existence of a
+collection of programs with type-level errors. Testing the rational
+programmer demands a currated collection of programs that take advantage
+of the sophisticated features of gradually typed languages such as Typed
+Racket.
 
-Since mixing of typed and untyped code in Typed Racket takes place at the module
-level, the experiment calls for benchmarks with a decent number of modules and a
-variety of module dependency graphs. Filtering the benchmark suite using these
-criteria yields ten suitable programs. Figure~\ref{table:benchmark-descriptions}
+Over the past decade,  \citet{gtnffvf-jfp-2019} have curated such a
+collection of Typed Racket programs for systematically measuring the
+implementation's performance. The benchmark suite consists of fully typed,
+correct programs, written by a number of different authors, who have
+maintained and evolved these programs over several years. The programs
+range widely in size, complexity, purpose, origin and in programming
+style. The latter is mostly a reference to their reliance on many Typed
+Racket features: occurrence typing~\cite{tf-icfp-2010}, types for mutable
+and immutable data structures~\cite{hpst-sfp-2010}, types for first-class
+classes and objects~\cite{tsdtf-oopsla-2012}, and types for Racket's
+numeric tower~\cite{stathff-padl-12}. Hence they offer a diverse lignustic
+substrate suitable for injecting type-level errors and testing the
+rational programmer.
+
+Since mixing of typed and untyped code in Typed Racket takes place at the
+module level, the experiment calls for benchmarks with a decent number of
+modules and a variety of module dependency graphs. Filtering the benchmark
+suite using these criteria while preserving its linguistic diversity
+yields ten suitable programs. Figure~\ref{table:benchmark-descriptions}
 displays the selected benchmark programs and describes their basic

### 2021-03-01 writing/mutate-mutators.tex
adding prose for mutators design and benchmarks selection

@@
 
-For the evaluation of a blame strategy, the mutator must produce type-level
+For the evaluation of a blame strategy, mutators must produce type-level
 mistakes that the run-time checks of gradual typing systems or the safety
 checks of the underlying language can detect. Once detected, the rational
-programmer should be able to detect the mistake by gradually adding types to
+programmer should be able to locate the mistake by gradually adding types to
 blamed modules.
@@
 Figure~\ref{table:mutation-ops} describes the 16 mutators used for the
-experiment.  They are inspired by the authors decades-long experience
-of making type-level mistakes in Typed Racket, some of which take a
-non-trivial effort to debug.  For each mutaror, the figure provides a
-short description and an example mutation.
+experiment.  For each mutator, the figure provides a short description and
+an example mutation.  Most are self-explanatory.  The first twelve apply
+to most gradually typed languages, including those with classes.  The last
+four target distinguishing features of Typed Racket.  Specifically, {\tt
+arithmetic} may replace a {\tt +} with a {\tt -} in an attempt to change
+the type of the result of the arithmetic operation. In Typed Racket, {\tt
++}'s result is a {\tt Positive-Integer} when all arguments are positive
+integers~\cite{stathff-padl-12}. However, the result of {\tt -} is {\tt
+Integer}. In the same spirit, the {\tt boolean} mutator aims to take
+advantage of Typed Racket faithfulness to Racket's ``falsiness.'' Finally,
+{\tt negate-cond} and {\tt force-cond} attempt to confuse occurrence
+typing.
 
-Most mutators are self-explanatory.  The first twelve apply to most
-gradually typed languages, including those with classes.  The last
-four target distinguishing features of Typed Racket.  Specifically,
-{\tt arithmetic} may replace a {\tt +} with a {\tt -} in an attempt to
-change the type of the result of the arithmetic operation. In Typed
-Racket, {\tt +}'s result is a {\tt Positive-Integer} when all
-arguments are positive integers~\cite{stathff-padl-12}. However, the
-result of {\tt -} is {\tt Integer}. In the same spirit, the {\tt
-boolean} mutator aims to take advantage of Typed Racket faithfulness
-to Racket's ``falsiness.'' Finally, {\tt negate-cond} and {\tt
-force-cond} attempt to confuse occurrence typing.
+For the design of these mutators, the starting point is the mutators of
+~\citet{lksfd-popl-2020} which in turn are borrowed from   the vast
+literature on mutation testing~\cite{jia2011analysis}. However, only 2
+of the previous mutators transfer as is to the context of this study, 1 is
+entirely ineffective and 5 need significant modifications. For the
+latter  the authors relied on their decades-long experience of making type-level mistakes in Typed
+Racket, some of which take a non-trivial effort to debug.  In more detail,
+the relational mutator of ~\citet{lksfd-popl-2020} does not lead to
+type-level errors that Typed Racket's type system can detect, so it has
+been eliminated. Similarly, the original arithmetic mutator does not
+reliably lead to type errors, so it has been adapted to only make operator
+swaps that can affect the type of the result. For example, changing a `/`
+to `*` does not affect the type of an expression, but changing `*` to `/`
+may s division by zero is detectable by Typed Racket's type system. 
+In addition to careful tweaking of previous mutators, the final
+set of mutators comes with 9 new ones that collectively
+produce a sufficiently large and diverse set of type-level mistakes. 
+For example, unlike the traditional subexpression swap mutator, 
+the {\tt -id} family of mutators apply in different
+contexts and utilize different pieces of non-local information about the
+identifiers in a program.
+
+As a final remark, the 16 selected mutators not only produce  type-level
+mistakes  reliably but also interesting debugging scenarios. Most
+mutators,  including variants of the ones
+~\citet{w:good-types-mutation} propose, some times achieve the first but 
+rarely the second. 
+ For example,  adding a new public method to
+a class surprisingly leads to mutants that either terminate 
+successfully or to mutants that fail 
+ immediately during the evaluation of the mutated class.  
+ In contrast, the seemingly similar mutator that turns a private method into    
+ a public one affects inheritance and leads to subtle bugs.  
+Hence, due to the elusive nature of mutators, it is imperative to define
+precisely what makes a debugging scenario interesting.

### 2021-03-01 writing/mutate-interesting.tex
mutators

@@
-All mutators produce mutations with type-level mistakes, but not all such
-mutations necessarily represent interesting debugging scenarios. Running a
-mutation may not raise a run-time error, though produce the wrong
-outputs---something the rational programmer may not recognize and certainly
-cannot use for a debugging session. Even if the program execution ends in an
-exceptional state, it may do so too quickly; for example, the first-order export
-checks of Natural may catch the error and point directly to the faulty module.
+%% -----------------------------------------------------------------------------
 
-To avoid these pitfalls, an \emph{interesting debugging scenario} is
-defined as a mutation that satisfies three conditions.  First, {\em
-the mutation is rejected by the type checker when all missing type
-annotations are restored\/}. This first criterion validates that the
-debugging scenario concerns an impedance mismatch.
+A type-level mutator is {\em interesting\/} (1) if, upon removal of the type
+annotations from the mutant-module, it raises a run-time error, (2) whose stack
+trace contains source locations from at least three modules, and (3) if it
+applies to all of the to-be-evaluated semantics. The application of an
+interesting mutator yields a mutant that represents a {\em debugging scenario\/}
+where the rational programmer can exploit blame information.
 
-Second, {\em it produces a run-time exception under Erasure as-is\/}, which
-ensures that it is detectable under all three semantics.  Whether a scenario
-produces a run time error largely depends on the underlying semantics but
-\citet{gfd-oopsla-2019} show that a run-time error under Erasure implies a run
-time error under both Transient and Natural. Thus the criterion aligns
-interesting debugging scenarios with the lowest denominator, Erasure. It
-excludes from consideration the most complex scenarios where only Natural or
-Transient can detect a mistake and Erasure is powerless.
+Let us consider the three conditions in order:
+\begin{enumerate}
+
+\item By definition, an impedance mismatch is a class between type ascription of
+one module to another module's exports. Hence, the application of a mutator must
+create a mutant module that fails to type check. Removing the types from this
+module creates an impedance mismatch between it and the modules that import its
+services, directly or indirectly. This impedance mismatch is of interest if, and
+only if, running the program leads to an exception---otherwise it goes
+unobserved and cannot be debugged.
+
+\item If the evaluation o mutant module immediately raises an exception because
+of the mutation, there is no work for the rational programmer. Indeed, if the
+stack trace contains source pointers to two modules, the scenario is still
+uninteresting. Every ordinary benchmark program comes with a {\tt main} module
+that acts as a driver, whose source is guaranteed to be included in the stack
+trace.  Hence, the definition of interesting mutator insists on the presence of
+three different modules in the stack trace. This guarantees that the debugging
+scenario demands a sufficiently sophisticated effort, due to the interaction
+between the buggy module with its context.  In these cases, the rational
+programmer must examine at least modules to locate the source of the faulty
+interaction.
+
+\item The goal of a comparative evaluation is to give the rational programmer a
+chance to debug the same scenario using different pieces of information.  In the
+case of gradual typing semantics, a meta-theorem due to Greenman and
+Felleisen~\cite{gf-icfp-2018} says that if a program raises an exception under
+Erasure, it also errors under all other semantics.  Hence, a comparison of blame
+information insists that a debugging scenario {\em raises a run-time exception
+under Erasure\/}.
 
@@
 
-Third, {\em the exception's stack trace mentions at least three
-distinct components\/}, meaning the scenario demands a sufficiently
-sophisticated debugging effort: one due to the interaction between the
-buggy portion of the program with the rest of the program.  In these
-cases, the rational programmer must examine at least modules to locate
-the source of the faulty interaction.  In technical terms, if the
-stack trace mentions at least two distinct components, the bug should
-be revealed by the run-time checks governing an interaction. A third
-component is required because all benchmark programs include a driver
-component, which is guaranteed to be in the stack trace. In essence,
-this last criterion excludes a large number of trivial-to-debug
-scenarios that error immediately when their mutated component is
-evaluated.
+\end{enumerate}
+\mf{It seems to me that we're going out on a limb with this claim. Isn't it true
+that the ``three'' is really the result of how many modules are typed/untyped in
+the context of the mutant module?}
+%
+Although the creation of such scenarios depends on both the mutator and the
+context of the mutant module, it turns out that some of the discarded mutators 
+often generate uninterresting scenarios.
 
@@
 %\end{figure*}
-
-The definition of interesting debugging scenarios creates a powerful filter. At
-a high level, the mutators produce 16,800 mutants with at least one interesting
-debugging scenario each across all of benchmarks.  Broken down by benchmark, the
-mutators produce at least 40 interesting scenarios for every benchmark, and
-these scenarios originate from at least four different mutators per benchmark.
-Thus, the mutators result in a sizable and diverse population of scenarios for
-every benchmark.  Furthermore, every mutator contributes scenarios to at least
-one benchmark.  Note, though, that some mutators apply only to a few benchmarks,
-because they target rather specific features.  For instance, the class-focused
-mutators are mainly effective in \texttt{take5}, because it makes the most
-extensive use of object-oriented Racket features.

### 2021-03-01 writing/mutate-mutators.tex
mutators

@@
 
-Figure~\ref{table:mutation-ops} describes the 16 mutators used for the
-experiment.  For each mutator, the figure provides a short description and
-an example mutation.  Most are self-explanatory.  The first twelve apply
-to most gradually typed languages, including those with classes.  The last
-four target distinguishing features of Typed Racket.  Specifically, {\tt
-arithmetic} may replace a {\tt +} with a {\tt -} in an attempt to change
-the type of the result of the arithmetic operation. In Typed Racket, {\tt
-+}'s result is a {\tt Positive-Integer} when all arguments are positive
-integers~\cite{stathff-padl-12}. However, the result of {\tt -} is {\tt
-Integer}. In the same spirit, the {\tt boolean} mutator aims to take
-advantage of Typed Racket faithfulness to Racket's ``falsiness.'' Finally,
-{\tt negate-cond} and {\tt force-cond} attempt to confuse occurrence
-typing.
+Figure~\ref{table:mutation-ops} describes 16 mutators that satisfy this
+requirement.  For each mutator, the figure provides a short description and an
+example. The starting point is the mutator set of~\citet{lksfd-popl-2020}
+which in turn are borrowed from the vast literature on mutation
+testing~\cite{jia2011analysis}. But, only 2 of the previous mutators transfer
+as is to the context of this study, 1 is entirely ineffective and 5 need
+significant modifications. For the latter the authors relied on their
+decades-long experience of making type-level mistakes in Typed Racket, some of
+which take a non-trivial effort to debug.  In more detail, the relational
+mutator of ~\citet{lksfd-popl-2020} does not lead to type-level errors that
+Typed Racket's type system can detect, so it has been eliminated. Similarly,
+the original arithmetic mutator does not reliably lead to type errors, so it
+has been adapted to only make operator swaps that can affect the type of the
+result. For example, changing a `/` to `*` does not affect the type of an
+expression, but changing `*` to `/` may s division by zero is detectable by
+Typed Racket's type system.  In addition to careful tweaking of previous
+mutators, the final set of mutators comes with 9 new ones that collectively
+produce a sufficiently large and diverse set of type-level mistakes.  For
+example, unlike the traditional subexpression swap mutator, the {\tt -id}
+family of mutators apply in different contexts and utilize different pieces of
+non-local information about the identifiers in a program.
 
-For the design of these mutators, the starting point is the mutators of
-~\citet{lksfd-popl-2020} which in turn are borrowed from   the vast
-literature on mutation testing~\cite{jia2011analysis}. However, only 2
-of the previous mutators transfer as is to the context of this study, 1 is
-entirely ineffective and 5 need significant modifications. For the
-latter  the authors relied on their decades-long experience of making type-level mistakes in Typed
-Racket, some of which take a non-trivial effort to debug.  In more detail,
-the relational mutator of ~\citet{lksfd-popl-2020} does not lead to
-type-level errors that Typed Racket's type system can detect, so it has
-been eliminated. Similarly, the original arithmetic mutator does not
-reliably lead to type errors, so it has been adapted to only make operator
-swaps that can affect the type of the result. For example, changing a `/`
-to `*` does not affect the type of an expression, but changing `*` to `/`
-may s division by zero is detectable by Typed Racket's type system. 
-In addition to careful tweaking of previous mutators, the final
-set of mutators comes with 9 new ones that collectively
-produce a sufficiently large and diverse set of type-level mistakes. 
-For example, unlike the traditional subexpression swap mutator, 
-the {\tt -id} family of mutators apply in different
-contexts and utilize different pieces of non-local information about the
-identifiers in a program.
-
-As a final remark, the 16 selected mutators not only produce  type-level
-mistakes  reliably but also interesting debugging scenarios. Most
-mutators,  including variants of the ones
-~\citet{w:good-types-mutation} propose, some times achieve the first but 
-rarely the second. 
- For example,  adding a new public method to
-a class surprisingly leads to mutants that either terminate 
-successfully or to mutants that fail 
- immediately during the evaluation of the mutated class.  
- In contrast, the seemingly similar mutator that turns a private method into    
- a public one affects inheritance and leads to subtle bugs.  
-Hence, due to the elusive nature of mutators, it is imperative to define
-precisely what makes a debugging scenario interesting.
+Most of the mutators are self-explanatory.  The first six apply to all
+gradually typed languages; the next six to those that include classes and
+objects.  The last four target distinguishing features of Typed Racket,
+specifically is sophisticated type system. In particular, {\tt arithmetic}
+replaces a {\tt +} with a {\tt -} in an attempt to change the type of the
+arithmetic expression; {\tt +}'s result is a {\tt Positive-Integer} when all
+arguments are positive integers, while {\tt -} yields {\tt
+Integer}~\cite{stathff-padl-12}. Similarly, the other three also aim to usurp
+Typed Racket's occurrence type system.

### 2021-03-01 writing/mutate-mutators.tex
adjust prose

@@
 requirement.  For each mutator, the figure provides a short description and an
-example. The starting point is the mutator set of~\citet{lksfd-popl-2020}
-which in turn are borrowed from the vast literature on mutation
-testing~\cite{jia2011analysis}. But, only 2 of the previous mutators transfer
-as is to the context of this study, 1 is entirely ineffective and 5 need
-significant modifications. For the latter the authors relied on their
-decades-long experience of making type-level mistakes in Typed Racket, some of
-which take a non-trivial effort to debug.  In more detail, the relational
-mutator of ~\citet{lksfd-popl-2020} does not lead to type-level errors that
-Typed Racket's type system can detect, so it has been eliminated. Similarly,
-the original arithmetic mutator does not reliably lead to type errors, so it
-has been adapted to only make operator swaps that can affect the type of the
-result. For example, changing a `/` to `*` does not affect the type of an
-expression, but changing `*` to `/` may s division by zero is detectable by
-Typed Racket's type system.  In addition to careful tweaking of previous
-mutators, the final set of mutators comes with 9 new ones that collectively
-produce a sufficiently large and diverse set of type-level mistakes.  For
-example, unlike the traditional subexpression swap mutator, the {\tt -id}
-family of mutators apply in different contexts and utilize different pieces of
-non-local information about the identifiers in a program.
+example. As the explanation indicates, some specialize or
+generalize~\citet{lksfd-popl-2020}'s mutators, which in turn are borrowed from
+the vast literature on mutation testing~\cite{jia2011analysis}.\footnote{Many
+of~\citet{lksfd-popl-2020} mutators do not satisfy the constraints. For
+example, the relational mutator does not lead to type-level errors that Typed
+Racket's type system can detect, so it has been eliminated. Similarly, the
+original arithmetic mutator does not reliably lead to type errors, so it has
+been adapted to only make operator swaps that can affect the type of the
+result. Changing a `/` to `*` does not affect the type of an expression, but
+changing `*` to `/` may, because division by zero is detectable by Typed
+Racket's type system.}  Only two are directly inherited; many mutators are
+brand new. For the latter the authors relied on their decades-long experience
+of making type-level mistakes in Typed Racket, some of which take a non-trivial
+effort to debug.
 

### 2021-03-01 writing/rational.tex
refoctor
justify rational programmer section

@@
-Code migration is the raison d'\^etre of Typed Racket, which is why the community
-uses ``migratory typing'' instead of gradual typing. Roughly speaking a
-programmer uses a gradually typed language to incrementally add types to a code
-base. From this perspective, a code base consists of some number of components,
-say modules, and a programmer converts a component at a time from its untyped
-state to a typed one.
+%% -----------------------------------------------------------------------------
 
-%% MF: I don't think we should honor the nonsense of ``true gt.''  We
-% are it as much as anyone!
-% 
-% \footnote{The method can be adapted to so-called true gradual typing
-% systems~\cite{svcb-snapl-2015} in a straightforward manner.}
+Section~\ref{sec:why-rational} explains, and illustrates with an example, how a
+migratory type setting helps find the source of an impedance mismatch. Roughly
+speaking, it encourages the rational programmer equip a module with types if it
+is blamed in an error message. A sequence of such steps make up a portion of the
+migratory path in the lattice of type migration~\cite{tfgnvf-popl-2016} (see
+sec.~\ref{sub:stuff} for details).
 
-Based on this idea, migratory paths offer a concise setting for
-evaluating the effectiveness of a rational programmer for dealing with
-the interesting debugging scenarios from section ~\ref{sec:mutate}.
-Grouping debugging scenarios by mutation yields a set of programs that
-all contain the same mistake but differ in their type annotations.  In
-the setting of migratory typing, each scenario differs from another in
-which components are typed. Formally, a program $\system$ is a set of
-components $\set{\component}$, and a scenario $\conf$ is the typed
-subset of $\set{\component}$ . Hence, the scenarios of $\system$ are
-ordered by the subset relation and, naturally, form a lattice
-$\lattice{\system}$ with $2^{\size{\system}}$
-elements.\footnote{\citet{tfgnvf-popl-2016} use this lattice for
-performance evaluations.} The bottom scenario of $\lattice{\system}$
-is always $\emptyset$ and the top one is $\system$ itself. In between
-are all the mixed-typed scenarios.
+Unfortunately, the rational programmer receives different kinds of information
+and thus may construct different paths. As section~\ref{sec:landscape} shows,
+the three different semantics deal with the same program in three different
+ways, and the error messages point to different sources of a problem.
 
-In this context, the actions of the rational programmer map to an
-ascending chain---dubbed a \emph{trail}---in $\lattice{\system}$.  The
-starting scenario $\conf_0$ of a trail---also referred to as the
-\emph{root} (of the trail)---is the initial debugging scenario; only
-interesting (as described in~\ref{sec:mutate}) debugging scenarios
-serve as roots.  Intuitively, starting from such a root, the rational
-programmer steps along a trail, and at each step, reduces the current
-scenario to an ``easier'' next scenario to debug.  The easiest
-scenarios are those that the type checker rejects outright, because a
-type error pinpoints an impedance mismatch between the broken code of
-one component and its typed interface to others. At this point, the
-programmer has identified the broken component.  If the scenario
-type-checks, the programmers runs the program until it raises a run
-time error.  The rational programmer uses the error's blame
-assignment\footnote{Blame can result either directly from a run-time
-check or, in the absence of blame, from an exception that can be
-traced back to the component that causes it with the help of the stack
-trace.} to decide which component(s) to equip with types. This choice
-constructs scenario $\conf_{i+1}$ from $\conf_{i}$ and thus creates a trail.
-
-% The successor of a scenario on a trail has more typed component;
-% formally, configuration $\conf_{i+1}$ on the trail differs from its
-% predecessor $\conf_{i}$ in that more components are typed.
-
-The first three subsections specialize the trail-scenario idea to the
-three semantics of gradual typing. Once this machinery is in place,
-the fourth subsection presents a reformulation of the informal
-question on page~1 as precise experimental research questions. The
-final subsection adds a notion of programmer effort to the
-experimental setup.
+Hence the research problem is how to view the actions of a rational programmer
+comparable across all semantics. Sections~\ref{sub:natural}
+through~\ref{sub:erasure} describe the various rational programmers and explain
+to what extent their actions are comparable.
 
@@
 
-\rsub{natural}   {Modes of the Natural Rational Programmer} 
-\rsub{transient} {Modes of the Transient Rational Programmer} 
-\rsub{erasure}   {Modes of the Erasure Rational Programmer} 
-\rsub{experiment}{Experimental Process} 
-\rsub{effort}    {Programmer Effort} \label{subsec:effort}
+\rsub{stuff}     {Stuff}
+
+\rsub{natural}   {The Natural Rational Programmer} 
+\rsub{transient} {The Transient Rational Programmer} 
+\rsub{erasure}   {The Erasure Rational Programmer}

### 2021-03-01 writing/rational-stuff.tex
stuff

%% -----------------------------------------------------------------------------

\citet{tfgnvf-popl-2016} describe the set of all possible type migrations with a
lattice. Formally, a program $\system$ is a set of modules $\set{\component}$.
Let a configuration $\conf$ of $\system$ be the subset of $\set{\component}$ that
comes with (possibly type-incorrect) annotations. These configurations of
$\system$ are ordered by the subset relation and form a lattice
$\lattice{\system}$ with $2^{\size{\system}}$ elements. The bottom scenario of
$\lattice{\system}$ is $\emptyset$, the top one $\system$ itself. In between are
the mixed-typed scenarios.

Applying a mutator to module $\component^*$ of $\system$ acts like a homomorphism
on the generated lattice. The two lattices differ only in the mutant of
$\component^*$. Given this formulation, debugging scenarios are those points in
this new lattice that do not contain the mutant; at all other points, the type
checker points out the type-level mistake in the mutant of $\component^*$.

The actions of the rational programmer create an ascending chain---dubbed a
\emph{trail}---in $\lattice{\system}$.  The starting scenario $\conf_0$ of a
trail---also referred to as the \emph{root} (of the trail)---is the initial
interesting debugging scenario.  Starting from such a root, the rational
programmer steps along a trail. If the scenario type-checks, the programmers runs
the program until it raises a run time error.  The rational programmer uses the
information in the error to decide which component(s) to equip with types. This
choice constructs scenario $\conf_{i+1}$ from $\conf_{i}$ and thus creates a
trail. The trail ends when programmer reaches a scenarios that contain the mutant
module, because the type checker rejects it outright. At this point, the
programmer has identified the source of the impedance mismatch.

### 2021-03-01 writing/mutate-interesting.tex
fix terminology

@@
 
-A type-level mutator is {\em interesting\/} (1) if, upon removal of the type
-annotations from the mutant-module, it raises a run-time error, (2) whose stack
-trace contains source locations from at least three modules, and (3) if it
-applies to all of the to-be-evaluated semantics. The application of an
-interesting mutator yields a mutant that represents a {\em debugging scenario\/}
-where the rational programmer can exploit blame information.
+A type-level mutation is {\em interesting\/} (1) if, upon removal of the type
+annotations, running the program raises a run-time error, (2) whose stack trace
+contains source locations from at least three modules, and (3) if this happens
+in all of the to-be-evaluated semantics. An interesting mutation represents a
+{\em debugging scenario\/} that supplies the rational programmer with
+information.
 
@@
 \item By definition, an impedance mismatch is a class between type ascription of
-one module to another module's exports. Hence, the application of a mutator must
-create a mutant module that fails to type check. Removing the types from this
-module creates an impedance mismatch between it and the modules that import its
-services, directly or indirectly. This impedance mismatch is of interest if, and
-only if, running the program leads to an exception---otherwise it goes
-unobserved and cannot be debugged.
+one module to another module's exports. Hence, type checking fails for an
+interesting mutant module, and removing the types from this mutant module
+creates an impedance mismatch between it and the modules that import it,
+directly or indirectly. This impedance mismatch is of interest if, and only if,
+running the program leads to an exception---otherwise it goes unobserved and
+cannot be debugged.
 
 \item If the evaluation o mutant module immediately raises an exception because
-of the mutation, there is no work for the rational programmer. Indeed, if the
+of the changes, there is no work for the rational programmer. Indeed, if the
 stack trace contains source pointers to two modules, the scenario is still
@@
 that acts as a driver, whose source is guaranteed to be included in the stack
-trace.  Hence, the definition of interesting mutator insists on the presence of
+trace.  Hence, the definition of interesting mutation insists on the presence of
 three different modules in the stack trace. This guarantees that the debugging
@@
 \end{enumerate}
-\mf{It seems to me that we're going out on a limb with this claim. Isn't it true
-that the ``three'' is really the result of how many modules are typed/untyped in
-the context of the mutant module?}
-%
-Although the creation of such scenarios depends on both the mutator and the
-context of the mutant module, it turns out that some of the discarded mutators 
-often generate uninterresting scenarios.
 

### 2021-03-01 writing/results-benchmarks-figure.tex
name files properly

--- 2021-02-23 writing/mutate-benchmarks-figure.tex
@@
 \begin{figure}
-\begin{tabular}{p{1.75cm} | p{6cm} }
-  {\bf  name} & {\bf description (author)}  \\
+\begin{tabular}{p{1.75cm} | p{4.2cm} | p{2.2cm} | p{2.2cm}}
+  {\bf  name} & {\bf description (author)} & {\bf lines of code} & {\bf
+  components} \\
 
 \hline
-  \texttt{acquire} & Object-oriented board game implementation (M. Felleisen)  \\%[1em]
+  \texttt{acquire} & Object-oriented board game implementation (M.
+  Felleisen) & 1941 & 9 \\%[1em]
 
 \hline
-  \texttt{gregor} & Utilities for calendar dates (J. Zeppieri) \\%[1em]
+  \texttt{gregor} & Utilities for calendar dates (J. Zeppieri) & 2336 & 13\\%[1em]
 
 \hline
-  \texttt{kcfa} & Functional implementation of 2CFA for the lambda calculus (M. Might) \\%[1em]
+  \texttt{kcfa} & Functional implementation of 2CFA for the lambda
+  calculus (M. Might) & 328 & 7\\%[1em]
 
 \hline
-  \texttt{quadT} & Converter from S-expression source code to PDF format (M. Butterick)\\%[1em]
+  \texttt{quadT} & Converter from S-expression source code to PDF format
+  (M. Butterick) & 7396 & 14\\%[1em]
 
 \hline
-  \texttt{quadU} & Converter from S-expression source code to PDF format  (B. Greenman) \\%[1em]
+  \texttt{quadU} & Converter from S-expression source code to PDF format
+  (B. Greenman) & 7282 & 14 \\%[1em]
 
 \hline
-  \texttt{snake} & Functional implementation of the  Snake video game (D. Van Horn) \\%[1em]
+  \texttt{snake} & Functional implementation of the  Snake video game (D.
+  Van Horn) & 182 & 8 \\%[1em]
 
 \hline
-  \texttt{suffixtree} & Algorithm for common longest subsequences between strings. (D. Yoo) \\%[1em]
+  \texttt{suffixtree} & Algorithm for common longest subsequences between
+  strings. (D. Yoo) & 1500 & 6 \\%[1em]
 
 \hline
-  \texttt{synth} & Converter of notes and drum beats to WAV (V. St-Amour \& N. Toronto) \\%[1em]
+  \texttt{synth} & Converter of notes and drum beats to WAV (V. St-Amour
+  \& N. Toronto) & 871 & 10 \\%[1em]
 
 \hline
-  \texttt{take5} & Mixin-based card game simulator (M.Felleisen)  \\%[1em]
+  \texttt{take5} & Mixin-based card game simulator (M.Felleisen)  & 465 &
+  8\\%[1em]
 
 \hline
-  \texttt{tetris} & Functional implementation of Tetris (D. Van Horn) \\%[1em]
+  \texttt{tetris} & Functional implementation of Tetris (D. Van Horn) &
+  280 & 9 \\%[1em]
 

### 2021-03-01 writing/results-experiment.tex
name files properly

%% -----------------------------------------------------------------------------

Trails and their properties give us the tools for a rigorous examination
of blame for Natural, Transient and Erasure Typed Racket in the setting of the 
mutants from section~\ref{sec:mutate}. In line with the discussion so far, 
our experimental process collects data to answer three initial questions
for our interesting debugging scenarios:
\begin{itemize}
\item[$Q_1$] Is blame useful in the context of Natural?

\item[$Q_2$] Is first blame useful in the context of Transient?

\item[$Q_3$] Is last blame useful in the context of Transient?

\end{itemize}

Furthermore, our experiment can compare the relative usefulness of blame
between any two of the three semantics. 
\begin{itemize}
\item[$Q_*$] Is blame for X more useful than blame for Y? (Where X and Y are any of Natural, Transient, or Erasure)
\end{itemize}

Table~\ref{fig:experiment-outline} summarizes how each question relates to
different kinds of trails/modes of the rational programmer. For example, experimental
question $Q_1$ asks whether blame is valuable for Natural and our experiment
uses the Natural blame and exception trails to answer it.

\begin{figure}[ht]
\center
{\begin{tabular}{l|c|c|c}
%% ---------------------------------------------------------------------------------------------------------------
                        & {\bf Natural}  & {\bf Transient} &  {\bf Erasure} \\ \hline 
%% ---------------------------------------------------------------------------------------------------------------
{\bf Blame}             &  $Q_1/Q_*$    &                  &                \\
{\bf First blame}       &               &     $Q_2/Q_*$    &                 \\
{\bf Last blame}        &               &     $Q_3/Q_*$    &                 \\
{\bf Exceptions}        &      $Q_1$    &     $Q_2/Q_3$    &      $Q_*$      \\
\end{tabular}}
  \caption{ Experimental questions and rational programmer modes.}
  \label{fig:experiment-outline}
\end{figure}

%% -----------------------------------------------------------------------------

In detail we answer $Q_1$  by comparing the success of the 
Natural blame and Natural exception trails for all interesting 
debugging scenarios from section~\ref{sec:mutate}.
  The first step is to construct each
mutant's scenario lattice and identify their interesting debugging
scenarios.
Our test
bed extends the trails that start from such roots according to the Natural-blame programmer.  If no scenarios
can be added to the trail, we check if
the last scenario of the trail type-checks or not. If it does, we record that the Natural-blame
trail is successful, and otherwise that it fails. We repeat the process for the
same roots again but for the other mode, Natural exceptions.
 Figure~\ref{fig:process}
summarizes this
experimental process for one mode of the rational programmer and connects
it with the mutations from section~\ref{sec:mutate}.
After completing the experiment, we calculate the
success/failure results of the trails to determine for each root whether Natural blame
is more useful than Natural exceptions. We
obtain a positive answer for
$Q_1$ when a root exists where the above is true since it is evidence that
there is at least one interesting scenario that the rational programmer
manages to debug because of blame. 
The process is analogous for $Q_2$ and $Q_3$, employing the respective
modes of the rational programmer.

For $Q_*$, the process is a bit more
involved. At a first level we proceed in a similar
manner except that we use two modes 
of the rational programmer from different semantics.
However, answering this question requires us to compare the percentage of scenarios
where one mode is more useful than the other and the inverse.
For instance, to
decide whether blame for Natural is more useful than first blame for Transient, 
we compare the percentage of interesting scenarios where 
Natural blame is more successful than Transient first blame 
with the percentage of interesting scenarios where Transient first blame 
is more successful than Natural blame. 
Finally we repeat these steps one more time to compare Natural blame and Transient last blame
and get a complete picture of the comparative usefulness of blame
in the two semantics.

\begin{figure}
  \centering
  \includegraphics[scale=0.36]{./Images/process}
  \caption{The experimental process for one mode of the rational
  programmer}
  \label{fig:process}
\end{figure}

### 2021-03-01 writing/results.tex
results: use stratified proportion estimation for all proportions
results: fit effort figures to page width
refoctor
name files properly

--- 2021-02-28 writing/results.tex
@@
-We run the experimental process from section~\ref{sec:rational} on the
-sampled debugging scenarios from section~\ref{sec:sample} using a machine with two Intel Xeon Gold 6258R processors (28 doubly-threaded cores each) and 500GB of memory.
-Each debugging scenario had a timeout limit of 4 minutes and a
-memory limit of 6GB. The complete experiment required
-over 30,000 hours of compute hours.
+%% -----------------------------------------------------------------------------
 
-Figure~\ref{fig:success-bars} summarizes the overall success rates of every mode.
-The success rates illustrate a few points that underly the rest of our analysis.
-The first notable piece of information from
-this figure is that every mode has failed debugging scenarios, not just
-Erasure. This should
-not come as a surprise to the astute
-reader. After all, as we
-discuss above, some modes are more successful than others in certain
-scenarios. Indeed any mode can fail if running a scenario results in an
-exception and the exception carries no useful information about which
-module the programmer should type next, e.g., because the stack trace 
-does not contain frames from any module of the program, or only typed modules. This situation
-can happen several steps along a blame trail or even immediately (figure~\ref{fig:effort-table} dives into this below).
+\def\rsub#1#2{\subsection{#2} \label{sub:#1} \input{results-#1.tex}}
 
-While most failures to debug scenarios follow the above pattern, a few do not.
-Breaking down the reasons for failure for Natural blame (1748 in total)
-reveals an additional cause. For a small set of
-debugging scenarios (40), Natural produces a run time type error
-blaming a non-buggy already typed
-component. We tracked down all these cases to known open issues with Typed
-Racket and class contracts.
-
-In Transient, similar to Natural,
-most failures are due to unhelpful exception information (1851 for both
-Transient first and last blame).  
-However, Transient also has a substantial
-number of failures because scenarios hit the time and memory
-limits of our experiment (~ 770 scenarios).  Additionally, there are nearly a 1000 cases where
-Transient reports an empty blame set which leaves the rational programmer
-without hints about how to proceed.
-In~\ref{sec:threat:transient}, we discuss how both of these causes of
-failure for Transient affect our experiment.
-
-The second key observation from figure~\ref{fig:success-bars} is that the modes that use blame all outperform those that do not.
-In particular, Natural and both of Transient's blame modes succeed in around 95\% of our scenarios, while their corresponding exception modes succeed in around 85\% of them and Erasure succeeds in only 60\% of them.
-The only exception is that the Random programmer always succeeds;
-this reflects the fact that every scenario has finitely many modules and this mode can never get stuck.
-
-Figure~\ref{fig:avo-bars} depicts a head-to-head comparison of every mode's performance against every other mode, helping to answer the four questions from section~\ref{sub:experiment}.
-Each plot shows a breakdown of the proportion of scenarios where one mode performs better or worse than each other mode.
-In particular, each bar above zero represents the proportion where the plot's entitled mode succeeds and the mode on the x-axis fails; the correspnding bar below zero represents the proportion of the opposite case.
-For example, the plot titled ``Natural'' shows that Natural outperforms Natural exceptions in about 10\% of the scenarios, and the inverse (Natural performs wrse than Natural exceptions) never happens.
-Similarly, the plot titled ``Transient last blame'' shows that Transient last blame outperforms Natural exceptions about in abut 10\% of the scenarios, but conversely it performs worse than Natural exceptions in about 2\% of the scenarios.
-
-Figure~\ref{fig:avo-bars} shows that the answers for questions $Q_1$ to $Q_3$ are all positive.
-In all three semantics, blame modes outperform their corresponding exception mode (by around 10\% for each one).
-The Natural exceptions mode is never more useful than Natural blame, and  Transient
-exceptions are more useful than Transient first and Transient last blame
-in a small percentage (about 2\%) of the scenarios.
-
-Figure~\ref{fig:avo-bars} also provides answers to $Q_*$.
-Blame for all three semantics is significantly
-more useful than Erasure exceptions (by around 30\%). Natural blame is
-more useful than both versions of Transient blame by a small percentage;
-in about 4\% of scenarios Natural blame is more useful while in about 1.5\% of the
-scenarios Transient blame is more useful. The Transient first and
-Transient last blame are practically indistinguishable. Finally, there is
-no clear winner between Natural exceptions and Transient exceptions
-despite the theoretically advantageous additional checks of the Natural
-semantics.
-
-In order to better understand questions $Q_1$ to $Q_3$, figure~\ref{fig:success-venns} depicts a breakdown of the success of each semantics in comparison to Erasure.
-Specifically, the figure shows one Venn diagram per mode of the rational programmer that uses blame.
-Each diagram breaks breaks down the overlap in the success of the blame mode, its corresponding exception mode, and Erasure.
-For example, in the breakdown of Natural, the diagram shows that all three modes succeed on the same scenario 57.3\% of the time, only Natural and Natural exceptions succeed on 29.1\% of the scenarios, only Natural and Erasure succeed on 2.1\%, and Natural alone succeeds on 9\%.
-This analysis highlights the success tradeoffs each semantics offers against Erasure, with and without blame, to answer questions about the tradeoffs in choosing to use blame or just exceptions for each semantics.
-For instance, the breakdown of Natural clearly illustrates that if choosing between Natural blame, Natural exceptions, and Erasure, Natural blame is the absolutely most successful: all of the successes of the other two modes are subsets of Natural's successful scenarios.
-On the other hand, Transient's blame modes are similar but not so clear-cut.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.30\textwidth]{./plots/success-bars}
-  \caption{Success baselines: The estimated percentage of scenarios for which each mode succeeds in locating the bug.
-  }
-  \label{fig:success-bars}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.9\textwidth]{./plots/avo-bars}
-  \caption{Usefulness comparisons: Each plot depicts a head-to-head comparison of the entitled mode against every other mode.
-  The portion above 0 (in green) is the estimated percentage of scenarios where the entitled mode is more useful than each of the other modes.
-  The portion below 0 (in red) is the estimated percentage of scenarios with the inverse situation.
-  The upper bound of error margin is 0.02\%.}
-  \label{fig:avo-bars}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.32\textwidth]{./plots/TR-TR-stack-first-venn}
-  \hfill
-  \includegraphics[width=0.32\textwidth]{./plots/transient-newest-transient-stack-first-venn}
-  \hfill
-  \includegraphics[width=0.32\textwidth]{./plots/transient-oldest-transient-stack-first-venn}
-
-  \caption{Blame usefulness analysis: Each diagram shows the breakdown of how three modes' successful scenarios overlap.
-  The percentage of scenarios covered by each circle are proportions for which the corresponding mode succeeds.
-  For example, in the breakdown of Natural, all three modes succeed on the same scenario 57.3\% of the time, only Natural and Natural exceptions succeed on 29.1\% of the scenarios, only Natural and Erasure succeed on 2.1\%, and Natural alone succeeds on 9\%.
-  }
-  \label{fig:success-venns}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.8\textwidth]{./plots/bt-lengths-table}
-  \caption{Programmer effort: Each plot depicts the distribution of trail
-  lengths for a given mode across all benchmarks.}
-  \label{fig:effort-table}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.9\textwidth]{./plots/bt-length-comparisons}
-  \caption{Effort comparisons: Each plot depicts the distribution of scenarios with trail length differences ranging from -3 to 3.
-  A negative difference means that the top mode's trail is three steps shorter than the bottom mode's trail for the same scenario, and a positive difference is the reverse.
-  Only scenarios where both modes succeed are considered.
-  The \texttt{.2|.8} on the y-axis indicates that the axis is truncated between 0.2 and 0.8.}
-  \label{fig:effort-comparisons}
-\end{figure}
-
-Turning to programmer effort, figure~\ref{fig:effort-table} shows the distribution of blame trail lengths
-for our sample of interesting debugging scenarios. Unlike the usefulness
-comparisons above, these proportions do not generalize to a representation of the full population.
-Still, as we discuss in section~\ref{subsec:effort}, they provide an
-alternative view of the workings of the rational programmer.
-
-There are
-two immediate take-aways from the figure. First, the effort for successfully
-debugging interesting scenarios (in green) for the random mode of the
-rational programmer follows  a normal distribution, as expected.\footnote{The
-random mode distribution is indistinguishable for all three semantics and thus the figure~\ref{fig:effort-table} 
-only shows one plot.} In contrast, in the other modes, successful effort coalesces at
-the left side of the plot, meaning that in most cases the programmer needs
-to type a single component to debug a scenario.
-
-A second point of interest is that the exceptions of
-the Erasure semantics either help the rational programmer immediately or 
-the rational programmer fails to debug a scenario altogether (in red).
-This is expected; no matter how many type annotations the programmer
-adds to an Erasure program, if the type checker doesn't reject the
-program, running the program always produces the same outcome. Thus an
-exception from the runtime has to point to the buggy module with
-the first try. Otherwise the rational programmer types an irrelevant
-module, runs the program again and the exception points again to the
-already typed module.
-
-Figure~\ref{fig:effort-comparisons} provides head-to-head comparisons of effort.
-To directly compare effort between modes, we compute the difference in trail length between two modes for all scenarios where they both succeed.
-Hence, each plot in the figure shows the distribution of scenarios with length differences ranging from -3 (the first mode's trail is 3 steps shorter than the second) to 3 (the first mode's trail is 3 steps longer than the second).
-The figure offers several insights about the comparative effort for each mode that complement the insights about comparitive success from figure~\ref{fig:avo-bars}.
-First, Natural blame never produces shorter trails than Natural exceptions, and sometimes (albeit rarely) produces slightly longer ones.
-Second, Natural relatively often (around 10\% of the scenarios) produces shorter trails than both Transient blame modes, and they are sometimes significantly shorter.
-Third, Transient last blame often produces shorter trails than Transient first blame, which is a particularly interesting distinction in light of their nearly undistinguishable success rates.
-Finally, Transient's blame modes also share the characteristic with Natural that blame sometimes produces longer trails than their corresponding exception modes.
+\rsub{experiment}{The Experimental Process} 
+\input{results-benchmarks-figure.tex}
+\rsub{benchmarks}{The Experimental Benchmarks}
+\rsub{observed}  {The Observed Results}

### 2021-03-01 writing/results.tex
cleaning up prose

@@
 %% -----------------------------------------------------------------------------
+
+At this point all ingredients are in place to describe the first application of
+the evaluation method. Two things are needed: (1) the description of the
+experimental setup, including the experimental questions; and (2) a set of
+representative benchmark programs. The first two subsections address these
+issues in order. The last one presents the observed results.
 

### 2021-03-01 writing/sample.tex
Lots of minor changes and rewordings
organize sections/titles/labels in paper
cleaning up prose

@@
-\section{How to Sample the Experimental Space}
+%% -----------------------------------------------------------------------------
 
-The mutators from section~\ref{sec:mutate} generate 96,515,744 debugging
-scenarios; far too many to even identify the interesting ones among them.
-Furthermore this population is heterogeneous; different scenarios come
-from different mutants and different mutants are the result of a
-different mutator and benchmark pair. Hence, to make the experiment
-computationally feasible and statistically precise, we perform stratified
-uniform random sampling with three levels of stratification.  The first
-two levels of strata group mutants first by benchmark and then by mutator. 
-We randomly select from the mutants of
-each benchmark ensuring that, per
-benchmark, the sample reflects the diversity of mutants with respect to
-the mutators that generated them.  Specifically,  we sample 80 mutants
-with interesting scenarios per benchmark,  evenly-distributed across all
-of the mutators that contribute mutants for the benchmark.  The third
-level of strata for our sampling is the lattice of a single mutant and we draw 96 random
-interesting scenarios from each lattice with replacement.
+%  from section~\ref{sec:mutate} (96,515,744)
 
-In summary, our sampling process takes advantage of ancillary information
-to deliver a  sufficiently large representative set of scenarios.
-Concretely, we test each
-mode of the rational programmer on 72,192 randomly selected interesting
-debugging scenarios spread across both benchmarks and mutators.
-Furthermore, stratified random sampling allows us to derive precise conclusions
-about the proportion of the whole population that satisfies a given trait 
-from the sample despite the population's heterogeneity.
-That property is exactly what we need to project comparisons about the
-usefulness of various modes of the
-rational programmer from the sample to the set of all interesting debugging
-scenarios.
+As is, the chosen mutators generate a huge number of debugging scenarios,
+approximately 1,000,000 for a suitable benchmark suite such as the one used for
+the first experiment (see~sec.~\ref{sub:results-benchmarks}). This number of
+scenarios is far too large to even identify the interesting ones among them.
+Furthermore this population is heterogeneous; different scenarios come from
+different mutants and different mutants are the result of a different mutator
+and benchmark pair.
+
+To make the experiment computationally feasible and statistically precise,
+stratified uniform random sampling with three levels of stratification becomes
+necessary.  The first two levels group mutants first by benchmark and then by
+mutator.  The set of randomly select from the mutants of each benchmark is
+checked to make sure that, per benchmark, the sample reflects the diversity of
+mutants with respect to the mutators that generated them.  Specifically, the
+process samples 80 mutants with interesting scenarios per benchmark,
+evenly-distributed across all of the mutators that contribute mutants for the
+benchmark.  At the third level, the sampling process draws 96 random interesting
+scenarios from each configuration lattice with replacement.
+
+%% This looks like a repeat of the above, made concrete for the benchmark set
+
+% In summary, the sampling process takes advantage of ancillary information to
+% deliver a sufficiently large representative set of scenarios.  Concretely, we
+% test each mode of the rational programmer on 72,192 randomly selected
+% interesting debugging scenarios spread across both benchmarks and mutators.
+% Furthermore, stratified random sampling allows us to derive precise conclusions
+% about the proportion of the whole population that satisfies a given trait from
+% the sample despite the population's heterogeneity.  That property is exactly
+% what we need to project comparisons about the usefulness of various modes of the
+% rational programmer from the sample to the set of all interesting debugging
+% scenarios.
 

### 2021-03-01 writing/results-observed.tex
factored

We run the experimental process from section~\ref{sec:rational} on the
sampled debugging scenarios from section~\ref{sec:sample} using a machine with two Intel Xeon Gold 6258R processors (28 doubly-threaded cores each) and 500GB of memory.
Each debugging scenario had a timeout limit of 4 minutes and a
memory limit of 6GB. The complete experiment required
over 30,000 hours of compute hours.

\begin{wrapfigure}{r}{.40\textwidth}
  \centering
  \includegraphics[width=0.40\textwidth]{./plots/success-bars}
  \caption{Success baselines: The estimated percentage of scenarios for which each mode succeeds in locating the bug.
  The upper bound of error margin is 0.02\%.
  }
  \label{fig:success-bars}
\end{wrapfigure}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{./plots/avo-bars}
  \caption{Usefulness comparisons: Each plot depicts a head-to-head comparison of the entitled mode against every other mode.
  The portion above 0 (in green) is the estimated percentage of scenarios where the entitled mode is more useful than each of the other modes.
  The portion below 0 (in red) is the estimated percentage of scenarios with the inverse situation.
  The upper bound of error margin is 0.02\%.
  }
  \label{fig:avo-bars}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.32\textwidth]{./plots/TR-TR-stack-first-venn}
  \hfill
  \includegraphics[width=0.32\textwidth]{./plots/transient-newest-transient-stack-first-venn}
  \hfill
  \includegraphics[width=0.32\textwidth]{./plots/transient-oldest-transient-stack-first-venn}

  \caption{Blame usefulness analysis: Each diagram shows the breakdown of how three modes' successful scenarios overlap.
  The percentage of scenarios covered by each circle are proportions for which the corresponding mode succeeds.
  For example, in the breakdown of Natural, all three modes succeed on the same scenario 57.3\% of the time, only Natural and Natural exceptions succeed on 29.1\% of the scenarios, only Natural and Erasure succeed on 2.1\%, and Natural alone succeeds on 9\%.
  The upper bound of error margin is 0.02\%.
  }
  \label{fig:success-venns}
\end{figure}

Figure~\ref{fig:success-bars} summarizes the overall success rates of every mode.
The success rates illustrate a few points that underly the rest of our analysis.
The first notable piece of information from
this figure is that every mode has failed debugging scenarios, not just
Erasure. This should
not come as a surprise to the astute
reader. After all, as we
discuss above, some modes are more successful than others in certain
scenarios. Indeed any mode can fail if running a scenario results in an
exception and the exception carries no useful information about which
module the programmer should type next, e.g., because the stack trace 
does not contain frames from any module of the program, or only typed modules. This situation
can happen several steps along a blame trail or even immediately (figure~\ref{fig:effort-table} dives into this below).

While most failures to debug scenarios follow the above pattern, a few do not.
Breaking down the reasons for failure for Natural blame (1748 in total)
reveals an additional cause. For a small set of
debugging scenarios (40), Natural produces a run time type error
blaming a non-buggy already typed
component. We tracked down all these cases to known open issues with Typed
Racket and class contracts.

In Transient, similar to Natural,
most failures are due to unhelpful exception information (1851 for both
Transient first and last blame).  
However, Transient also has a substantial
number of failures because scenarios hit the time and memory
limits of our experiment (~ 770 scenarios).  Additionally, there are nearly a 1000 cases where
Transient reports an empty blame set which leaves the rational programmer
without hints about how to proceed.
In~\ref{sec:threat:transient}, we discuss how both of these causes of
failure for Transient affect our experiment.

The second key observation from figure~\ref{fig:success-bars} is that the modes that use blame all outperform those that do not.
In particular, Natural and both of Transient's blame modes succeed in around 95\% of our scenarios, while their corresponding exception modes succeed in around 85\% of them and Erasure succeeds in only 60\% of them.
The only exception is that the Random programmer always succeeds;
this reflects the fact that every scenario has finitely many modules and this mode can never get stuck.

Figure~\ref{fig:avo-bars} depicts a head-to-head comparison of every mode's performance against every other mode, helping to answer the four questions from section~\ref{sub:experiment}.
Each plot shows a breakdown of the proportion of scenarios where one mode performs better or worse than each other mode.
In particular, each bar above zero represents the proportion where the plot's entitled mode succeeds and the mode on the x-axis fails; the correspnding bar below zero represents the proportion of the opposite case.
For example, the plot titled ``Natural'' shows that Natural outperforms Natural exceptions in about 10\% of the scenarios, and the inverse (Natural performs wrse than Natural exceptions) never happens.
Similarly, the plot titled ``Transient last blame'' shows that Transient last blame outperforms Natural exceptions about in abut 10\% of the scenarios, but conversely it performs worse than Natural exceptions in about 2\% of the scenarios.

Figure~\ref{fig:avo-bars} shows that the answers for questions $Q_1$ to $Q_3$ are all positive.
In all three semantics, blame modes outperform their corresponding exception mode (by around 10\% for each one).
The Natural exceptions mode is never more useful than Natural blame, and  Transient
exceptions are more useful than Transient first and Transient last blame
in a small percentage (about 2\%) of the scenarios.

Figure~\ref{fig:avo-bars} also provides answers to $Q_*$.
Blame for all three semantics is significantly
more useful than Erasure exceptions (by around 30\%). Natural blame is
more useful than both versions of Transient blame by a small percentage;
in about 4\% of scenarios Natural blame is more useful while in about 1.5\% of the
scenarios Transient blame is more useful. The Transient first and
Transient last blame are practically indistinguishable. Finally, there is
no clear winner between Natural exceptions and Transient exceptions
despite the theoretically advantageous additional checks of the Natural
semantics.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{./plots/bt-lengths-table}
  \caption{Programmer effort: Each plot depicts the distribution of trail
  lengths for a given mode across all benchmarks.}
  \label{fig:effort-table}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{./plots/bt-length-comparisons}
  \caption{Effort comparisons: Each plot depicts the distribution of scenarios with trail length differences ranging from -3 to 3.
  A negative difference means that the first mode's trail is three steps shorter than the second mode's trail for the same scenario, and a positive difference is the reverse.
  Only scenarios where both modes succeed are considered.
  The 10|85 on the y-axis indicates that the axis is truncated between 10 and 85\%.}
  \label{fig:effort-comparisons}
\end{figure}

In order to better understand questions $Q_1$ to $Q_3$, figure~\ref{fig:success-venns} depicts a breakdown of the success of each semantics in comparison to Erasure.
Specifically, the figure shows one Venn diagram per mode of the rational programmer that uses blame.
Each diagram breaks breaks down the overlap in the success of the blame mode, its corresponding exception mode, and Erasure.
For example, in the breakdown of Natural, the diagram shows that all three modes succeed on the same scenario 57.3\% of the time, only Natural and Natural exceptions succeed on 29.1\% of the scenarios, only Natural and Erasure succeed on 2.1\%, and Natural alone succeeds on 9\%.
This analysis highlights the success tradeoffs each semantics offers against Erasure, with and without blame, to answer questions about the tradeoffs in choosing to use blame or just exceptions for each semantics.
For instance, the breakdown of Natural clearly illustrates that if choosing between Natural blame, Natural exceptions, and Erasure, Natural blame is the absolutely most successful: all of the successes of the other two modes are subsets of Natural's successful scenarios.
On the other hand, Transient's blame modes are similar but not so clear-cut.

Turning to programmer effort, figure~\ref{fig:effort-table} shows the distribution of blame trail lengths
for our sample of interesting debugging scenarios. Unlike the usefulness
comparisons above, these proportions do not generalize to a representation of the full population.
Still, as we discuss in section~\ref{subsec:effort}, they provide an
alternative view of the workings of the rational programmer.

There are
two immediate take-aways from the figure. First, the effort for successfully
debugging interesting scenarios (in green) for the random mode of the
rational programmer follows  a normal distribution, as expected.\footnote{The
random mode distribution is indistinguishable for all three semantics and thus the figure~\ref{fig:effort-table} 
only shows one plot.} In contrast, in the other modes, successful effort coalesces at
the left side of the plot, meaning that in most cases the programmer needs
to type a single component to debug a scenario.

A second point of interest is that the exceptions of
the Erasure semantics either help the rational programmer immediately or 
the rational programmer fails to debug a scenario altogether (in red).
This is expected; no matter how many type annotations the programmer
adds to an Erasure program, if the type checker doesn't reject the
program, running the program always produces the same outcome. Thus an
exception from the runtime has to point to the buggy module with
the first try. Otherwise the rational programmer types an irrelevant
module, runs the program again and the exception points again to the
already typed module.

Figure~\ref{fig:effort-comparisons} provides head-to-head comparisons of effort.
To directly compare effort between modes, we compute the difference in trail length between two modes for all scenarios where they both succeed.
Hence, each plot in the figure shows the distribution of scenarios with length differences ranging from -3 (the first mode's trail is 3 steps shorter than the second) to 3 (the first mode's trail is 3 steps longer than the second).
The figure offers several insights about the comparative effort for each mode that complement the insights about comparitive success from figure~\ref{fig:avo-bars}.
First, Natural blame never produces shorter trails than Natural exceptions, and sometimes (albeit rarely) produces slightly longer ones.
Second, Natural relatively often (around 10\% of the scenarios) produces shorter trails than both Transient blame modes, and they are sometimes significantly shorter.
Third, Transient last blame often produces shorter trails than Transient first blame, which is a particularly interesting distinction in light of their nearly undistinguishable success rates.
Finally, Transient's blame modes also share the characteristic with Natural that blame sometimes produces longer trails than their corresponding exception modes.

### 2021-03-01 writing/challenges.tex
re-established the old organization

@@
 
-The {\em first challenge\/} is to find a representative collection of impedance
-mismatches. They must represent mistakes that programmers accidentally create
-and that the run-time checks of academic systems catch. In other words, the
-experiment calls for a collection of type mistakes in mix-typed programs that is
-representative of those ``in the wild.''  Unfortunately no such collection
-exists, and with good reason. The kind of mistakes needed are typically detected
-by unit or integration tests; even if it takes some time to find their sources,
-these mistakes do not make it into code repositories with appropriate commit
-messages.
+The {\em first challenge\/} concerns the differences among the blame assignment
+of the three semantic variants.  While Natural assigns blame to {\em one\/}
+component, Transient assigns blame to a sequence of components. The Erasure
+semantics does not blame a component {\it per se\/}, but it comes with an
+exception location and a stack trace, which implicitly suggests parties to
+blame.  The different strategies greatly affect the behavior of the rational
+programmer (and real programmers, too).
+
+One way to overcome this challenge is to {\em equip the rational programmer with
+modes \/} that represent the alternative interpretations of blame.  For
+instance, one Transient mode may pick the oldest element of a blame set as
+the next component to add types because it corresponds to the earliest point in
+the program's evaluation that can discover an impedance mismatch.  Another mode
+may opt to treat the sequence as a stack and add types to its newest element.
+When both modes are equally successful in locating an impedance mismatch,
+measuring the rational programmer's effort may answer which mode is the most
+effective.
+
+Different blame information is also due to differences in the run-time checks
+that the three gradual typing systems perform. This observation suggests another
+mode for the rational programmer, namely, a mode that ignores the blame
+information completely and instead inspects only the information in the
+exceptions of the safety checks in the underlying languages.  This exception
+mode serves as the {\em baseline\/} for blame's value; if the programmer in this
+mode ``beats'' the others, blame systems might be useless.
+
+The {\em second challenge\/} is to find a representative collection of
+impedance mismatches. They must represent mistakes that programmers
+accidentally create and that the run-time checks of academic systems catch. In
+other words, the experiment calls for a collection of type mistakes in
+mix-typed programs that is representative of those ``in the wild.''
+Unfortunately no such collection exists, and with good reason. The kind of
+mistakes needed are typically detected by unit or integration tests; even if it
+takes some time to find their sources, these mistakes do not make it into code
+repositories with appropriate commit messages.
 
@@
 interesting debugging scenarios\/} (see section~\ref{sub:mutate-interesting}).
-
-The {\em second challenge\/} concerns the differences among the blame assignment
-of the three semantic variants.  While Natural assigns blame to {\em one\/}
-component, Transient assigns blame to a sequence of components. The Erasure
-semantics does not blame a component {\it per se\/}, but it comes with an
-exception location and a stack trace, which implicitly suggests parties to
-blame.  The different strategies greatly affect the behavior of the rational
-programmer (and real programmers, too).
-
-One way to overcome this challenge is to {\em equip the rational programmer with
-modes \/} that represent the alternative interpretations of blame.  For
-instance, one Transient mode may pick the oldest element of a blame set as
-the next component to add types because it corresponds to the earliest point in
-the program's evaluation that can discover an impedance mismatch.  Another mode
-may opt to treat the sequence as a stack and add types to its newest element.
-When both modes are equally successful in locating an impedance mismatch,
-measuring the rational programmer's effort may answer which mode is the most
-effective.
-
-Different blame information is also due to differences in the run-time checks
-that the three gradual typing systems perform. This observation suggests another
-mode for the rational programmer, namely, a mode that ignores the blame
-information completely and instead inspects only the information in the
-exceptions of the safety checks in the underlying languages.  This exception
-mode serves as the {\em baseline\/} for blame's value; if the programmer in this
-mode ``beats'' the others, blame systems might be useless.
 

### 2021-03-01 writing/mutate-benchmarks.tex
re-established the old organization

@@
 characteristic.
+
+The definition of interesting debugging scenarios creates a powerful filter. All
+together, the listed mutators produce 16,800 mutants with at least one
+interesting debugging scenario across all benchmarks.  Broken down by benchmark,
+the mutators produce at least 40 interesting scenarios for every benchmark, and
+these scenarios originate from at least four different mutators per benchmark.
+Thus, the mutators result in a sizable and diverse population of scenarios for
+every benchmark.  Furthermore, every mutator contributes scenarios to at least
+one benchmark.  Some mutators apply only to a few benchmarks, because they
+target rather specific features; for instance, the class-focused mutators are
+mainly effective in a program that makes extensive use of object-oriented
+features.

### 2021-03-01 writing/rational-experiment.tex
re-established the old organization

--- 2021-03-01 writing/results-experiment.tex
@@
 
-Furthermore, our experiment can compare the relative usefulness of blame
-between any two of the three semantics. 
+Furthermore, the experiment allows a comparison of the relative usefulness of blame
+information:
 \begin{itemize}
-\item[$Q_*$] Is blame for X more useful than blame for Y? (Where X and Y are any of Natural, Transient, or Erasure)
+\item[$Q_*$] Is blame for X more useful than blame for Y (for X, Y in Natural, Transient, or Erasure)?
 \end{itemize}
@@
 
-In detail we answer $Q_1$  by comparing the success of the 
+In detail the answer to $Q_1$ demands a comparison of the success of the 
 Natural blame and Natural exception trails for all interesting 
-debugging scenarios from section~\ref{sec:mutate}.
+debugging scenarios.
   The first step is to construct each
@@
 there is at least one interesting scenario that the rational programmer
-manages to debug because of blame. 
-The process is analogous for $Q_2$ and $Q_3$, employing the respective
+manages to debug with blame information. 
+The process is analogous for $Q_2$ and $Q_3$, using the respective
 modes of the rational programmer.

### 2021-03-01 writing/results.tex
re-established the old organization

--- 2021-03-01 writing/results.tex
@@
-\input{mutate-benchmarks-figure.tex}
-\sub{benchmarks} {Which Software to Use As Benchmarks}
+%% -----------------------------------------------------------------------------
 
-\subsection{XXX}
 We run the experimental process from section~\ref{sec:rational} on the
@@
   \caption{Success baselines: The estimated percentage of scenarios for which each mode succeeds in locating the bug.
+  The upper bound of error margin is 0.02\%.
   }
@@
   The portion below 0 (in red) is the estimated percentage of scenarios with the inverse situation.
-  The upper bound of error margin is 0.02\%.}
+  The upper bound of error margin is 0.02\%.
+  }
   \label{fig:avo-bars}
@@
   For example, in the breakdown of Natural, all three modes succeed on the same scenario 57.3\% of the time, only Natural and Natural exceptions succeed on 29.1\% of the scenarios, only Natural and Erasure succeed on 2.1\%, and Natural alone succeeds on 9\%.
+  The upper bound of error margin is 0.02\%.
   }
@@
   \centering
-  \includegraphics[width=0.8\textwidth]{./plots/bt-lengths-table}
+  \includegraphics[width=\textwidth]{./plots/bt-lengths-table}
   \caption{Programmer effort: Each plot depicts the distribution of trail
@@
   \centering
-  \includegraphics[width=0.9\textwidth]{./plots/bt-length-comparisons}
+  \includegraphics[width=\textwidth]{./plots/bt-length-comparisons}
   \caption{Effort comparisons: Each plot depicts the distribution of scenarios with trail length differences ranging from -3 to 3.

### 2021-03-01 writing/sample.tex
re-established the old organization

@@
 As is, the chosen mutators generate a huge number of debugging scenarios,
-approximately 1,000,000 for a suitable benchmark suite such as the one used for
-the first experiment (see~sec.~\ref{sub:results-benchmarks}). This number of
-scenarios is far too large to even identify the interesting ones among them.
-Furthermore this population is heterogeneous; different scenarios come from
-different mutants and different mutants are the result of a different mutator
-and benchmark pair.
+approximately 1,000,000 for the chosen benchmark. This number of scenarios is
+far too large to even identify the interesting ones among them.  Furthermore
+this population is heterogeneous; different scenarios come from different
+mutants and different mutants are the result of a different mutator and
+benchmark pair.
 

### 2021-03-01 writing/introduction.tex
intro, landscape, why-rational: typo, wording, and other small fixes

@@
 latter---and this systematic, information-driven process can be
-implemented.\footnote{\citet{cc-snapl-19} use this tracing method to localize
-type errors in inference-based gradual typing systems.}  Implementing it means
+implemented and tested at scale on real programs.\footnote{\citet{cc-snapl-19} use this tracing method to localize
+type errors in inference-based gradual typing systems.}  Testing it systematically requires
 overcoming major challenges: injecting representative impedance mismatches;
@@
 The paper's {\it second contribution\/} is a set of {\em results from applying
-the evaluation method to three distinct blame assignment strategies and
-approximately 72,200 cases\/} (see sec.~\ref{sec:results}): (1) {\it
-Transient\/}, i.e. Reticulated's tracking of typed/untyped boundaries; (2) {\it
-Natural\/}, i.e. the use of higher-order contract system and its blame
+the evaluation method to three distinct blame assignment strategies in
+approximately 72,200 different scenarios\/} (see sec.~\ref{sec:results}): (1)~{\it
+Transient\/}, i.e. Reticulated's tracking of typed/untyped boundaries~\cite{vss-popl-2017}; (2)~{\it
+Natural\/}, i.e. the use of a higher-order contract system and its blame
 assignment~\cite{ff-icfp-2002}; and (3) {\it Erasure\/}, i.e. the approach of
 industrial systems, which forgo blame in favor of error messages from the safety
-checks of the underlying language.---The results (see
+checks of the underlying language. The results (see
 sec.~\ref{sec:discussion}) are at least somewhat surprising.  In principle

### 2021-03-01 writing/landscape.tex
justify rational programmer section
intro, landscape, why-rational: typo, wording, and other small fixes

@@
 
-\item[\texttt{crypto-pkac-lib}] (at the bottom left) also imports \texttt{pack}
+\item[\texttt{crypto-pack-lib}] (at the bottom left) also imports \texttt{pack}
 and assumes for it the same type as {\tt typed-pack-lib}. It applies the
@@
 because it may not produce the promised kind of value, as is the case here with
-{\tt pack} from {\tt pack-lib}. Analogously the run-time system wraps exported
-functions such as {\tt crypto-pack} of a typed modules in a proxy that checks
-its arguments.  This check protects functions exported from type modules against
+{\tt pack} from {\tt pack-lib}. Analogously the run-time system wraps each exported
+function of a typed module (such as {\tt crypto-pack}) in a proxy that checks
+its arguments.  This check protects functions exported from typed modules against
 applications in untyped code that supply the wrong kind of argument value.
 
-As this analysis implies, if the return-type check fails, the problem is with
+As this analysis implies, if a return-type check fails, the problem is with
 the untyped module that supplied an incorrect function, here {\tt pack-lib}. If
-the argument-type check fails, responsibility lies with {\tt client}. Due to the
+an argument-type check fails, responsibility lies with {\tt client}. Due to the
 proxies wrapped around functions, Natural can easily track the responsible party
 for each check. Thus, in the example of figure~\ref{fig:example}, as
-\texttt{typed-pack} returns, the return-type check fails, and Natural blames
-\texttt{typed-pack-lib}.
+\texttt{pack} returns, the return-type check fails, and Natural blames
+\texttt{pack-lib}.
 
@@
 top constructor of a value. Since retrieving a value from within a structure (or
-list, array, hash table etc.) is performed via a function call, the innards of a
+list, array, hash table etc.) is performed via a function call, the contents of a
 complex value are checked on a piecemeal basis.
@@
 because it takes place in the untyped {\tt client} module, which is compiled in
-the usual manner. Because {\tt pack} is called in the {\tt crypto-pack-lib}
+the usual manner. Later, when {\tt pack} is called in the {\tt crypto-pack-lib}
 module, Transient's inlined checks make sure that the imported
-\texttt{pack} is a function and that the result of its call in is a list. This
-last check fails when \texttt{client} calls to {\tt crypto-pack}.
+\texttt{pack} is a function and that its result is a list. This
+last check fails in \texttt{client}'s call to {\tt crypto-pack}.
 
-In order to pinpoint the responsible party for failed checks, Transient
+In order to locate the responsible party for failed checks, Transient
 maintains a map from values to the boundaries between typed/untyped modules that
@@
 \begin{tabular}{l|ll}
-                                 \multicolumn{3}{c}{the evaluation of} \\
+                        &        \multicolumn{2}{|c}{the evaluation of} \\
 \relax                  & {\tt public-data}               & {\tt secret-data} in {\tt client}, \\
@@
 %% ----------------------------------------------------------------------------------------
-\qquad {\it Natural\/}   & error, blaming              & error, blaming \\
-     		   	 & \quad {\tt typed-pack-lib}  & \quad {\tt typed-pack-lib} \\
-\qquad {\it Transient\/} & no error   		       & error, blaming the {\tt pack-lib}/ \\
-      		  	 &    			       & \quad {\tt crypto-pack-lib} boundary \\
+\qquad {\it Natural\/}   & error, blaming {\tt pack-lib}  & error, blaming {\tt pack-lib} \\
+     		   	 &                                & \\
+\qquad {\it Transient\/} & no error   		          & error, blaming the boundary \\
+      		  	 &    			          & \quad {\tt pack-lib}/{\tt crypto-pack-lib} \\
 \qquad {\it Erasure\/}   & no error 		          & no error \\

### 2021-03-01 writing/why-rational.tex
intro, landscape, why-rational: typo, wording, and other small fixes

@@
 
-This paper dubs Lazarek et. alii's idea {\em rational programmer\/}. Like Stuart
+This paper dubs Lazarek et. all's idea {\em the rational programmer\/}. Like Stuart
 Mill's {\it homo economicus\/}, the rational programmer approximates the
@@
 behavior: the error message and the state of the program. Hence, the most
-rational procedure is to use the former to improve the latter. Put differently,
+rational procedure is to use the former to improve the latter. Specifically,
 a rational programmer translates the Wadler-Findler slogan into a debugging
-method, searching for the source of the impedance mismatch with additional type
-annotations. Measuring this simulated programmer's behavior on a large number of
+method, searching for the source of the impedance mismatch by adding type
+annotations to the parts of the program identified in the error.
+If the type checker rejects an annotation, the rational programmer has found the problem.
+Measuring this simulated programmer's behavior on a large number of
 debugging scenarios yields data that is similar to data collected in an
@@
 Now consider the concrete (and simplistic) example of figure~\ref{fig:rational}.
-Initially the code base consists of the two \dyncolor\ module on the left plus
+Initially the code base consists of the two \dyncolor\ modules on the left plus
 the \typecolor\ module at the bottom; \dyncolor\ indicates untyped, while
@@
 The referenced contract is the compilation of the type of {\tt na-client}. The
-definitive hint is ``blaming: (interface for na-client)'' with the caveat
-``(assuming the contract is correct).''
+definitive hint is ``{\tt blaming: (interface for na-client)}'' with the caveat
+``{\tt (assuming the contract is correct)}.''
 
-Assuming the rational programmer trust the type of {\tt na-client}, the next
+Assuming the rational programmer trusts the type of {\tt na-client}, the next
 step is to inspect the {\tt layer} module and to equip it with type
-annotations. The result is the \typecolor\ module in the middle, and {\tt
-main}'s import is now re-directed there: {\tt (submod ".." layer)}. Running the
-program again yields this message:
+annotations. This results in the \typecolor\ module in the middle, and {\tt
+main}'s import is now re-directed there by {\tt (submod ".." layer)}. Running the
+program again yields a new message:
 \begin{verbatim}
@@
 equipping entire modules with types represents an always feasible
-approach.  But, the concept of studying the idea of an economically
+approach.  But the concept of studying the idea of an economically
 rational actor has produced benefits to political economics, and this

### 2021-03-01 writing/rational-effort.tex
rational programmer with editorial voice

@@
-In addition to successful and failure information for the various trails,  
-our experiment  records the number of components a rational programmer 
-has to type along each trail ($\lvert \conf_n \setminus \conf_0
-\rvert$). We use this number as the metric for the effort 
-of the rational programmer to debug an interesting scenario.
+%% -----------------------------------------------------------------------------
 
-A comparison of the effort of different modes of the rational programmer
-can help shed some further light to the effectiveness of the
-three gradual typing systems. For example, consider one debugging scenario.  
-If, for instance, both  Natural blame and Transient first
-blame trails are successful, the two modes of the 
-rational programmer can compete to see which one
-debugs the interesting scenario with less effort. In
-general, if the effort distribution for a mode of the rational programmer
-has a shorter tail and more
-volume around smaller values compared to the effort distribution of another
-mode, then this is evidence that the first mode is the more effective of the two.
+In addition to success and failure information for the various trails, the
+experiment software records the number of components a rational programmer has
+to equip with types along each trail ($\lvert \conf_n \setminus \conf_0
+\rvert$). This number can serve as a metric.
 
-Finally, effort allows us to test whether the observed effectiveness 
-of the rational programmer is an artifact of pure chance or not.
-That is, we can compare the effort distribution for a mode of the
-rational programmer with another mode that ignores error information entirely
-and instead selects which component to type next randomly. 
-Of course, since each mutant has a finite number of components, 
-such random mode trails are always successful. At the same time though,
-the Natural, Transient, and Erasure random modes exhibit a normal effort
-distribution across scenarios that should be distinct from the effort distribution for all other modes of
-the rational programmer if the effectiveness of the modes are not
-coincidental.
+A comparison of the effort of different modes of the rational programmer can
+illuminate the comparison of the effectiveness of the three gradual typing
+systems. For example, consider a debugging scenario.  If, for instance, both
+Natural blame and Transient first blame trails are successful, the two modes of
+the rational programmer can compete to see which one debugs the interesting
+scenario with less effort. In general, if the effort distribution for a mode of
+the rational programmer has a shorter tail and more volume around smaller values
+compared to the effort distribution of another mode, then the first mode is
+probably the more effective of the two.
+
+Measuring effort can also tell whether the observed effectiveness of the
+rational programmer is an artifact of pure chance or not.  Doing so means
+comparing the effort distribution for a mode of the rational programmer with
+another mode that ignores error information entirely and instead selects which
+component to type next randomly.  Since each mutant has a finite number of
+components, such random mode trails are naturally always successful. At the same
+time though, the Natural, Transient, and Erasure random modes exhibit a normal
+effort distribution across scenarios that should be distinct from the effort
+distribution for all other modes of the rational programmer if the effectiveness
+of the modes are not coincidental.

### 2021-03-01 writing/rational-experiment.tex
rational programmer with editorial voice

@@
 
-Trails and their properties give us the tools for a rigorous examination
-of blame for Natural, Transient and Erasure Typed Racket in the setting of the 
-mutants from section~\ref{sec:mutate}. In line with the discussion so far, 
-our experimental process collects data to answer three initial questions
-for our interesting debugging scenarios:
+Trails and their properties provide the tools for a rigorous examination of
+blame for Natural, Transient and Erasure. In line with the discussion so far,
+the experiment software collects data to answer three initial questions for 
+interesting debugging scenarios:
 \begin{itemize}
@@
 different kinds of trails/modes of the rational programmer. For example, experimental
-question $Q_1$ asks whether blame is valuable for Natural and our experiment
+question $Q_1$ asks whether blame is valuable for Natural and the experiment
 uses the Natural blame and exception trails to answer it.
@@
 
-In detail the answer to $Q_1$ demands a comparison of the success of the 
-Natural blame and Natural exception trails for all interesting 
-debugging scenarios.
-  The first step is to construct each
-mutant's scenario lattice and identify their interesting debugging
-scenarios.
-Our test
-bed extends the trails that start from such roots according to the Natural-blame programmer.  If no scenarios
-can be added to the trail, we check if
-the last scenario of the trail type-checks or not. If it does, we record that the Natural-blame
-trail is successful, and otherwise that it fails. We repeat the process for the
-same roots again but for the other mode, Natural exceptions.
- Figure~\ref{fig:process}
-summarizes this
-experimental process for one mode of the rational programmer and connects
-it with the mutations from section~\ref{sec:mutate}.
-After completing the experiment, we calculate the
-success/failure results of the trails to determine for each root whether Natural blame
-is more useful than Natural exceptions. We
-obtain a positive answer for
-$Q_1$ when a root exists where the above is true since it is evidence that
-there is at least one interesting scenario that the rational programmer
-manages to debug with blame information. 
-The process is analogous for $Q_2$ and $Q_3$, using the respective
-modes of the rational programmer.
+In detail the answer to $Q_1$ demands a comparison of the success of the Natural
+blame and Natural exception trails for all interesting debugging scenarios.  The
+first step is to construct each mutant's scenario lattice and identify their
+interesting debugging scenarios.  The experiment software extends the trails that start
+from such roots according to the Natural-blame programmer.  If no scenarios can
+be added to the trail, the experiment software checks whether the last scenario
+of the trail type-checks or not. If it does, the software records that the
+Natural-blame trail is successful; otherwise it is recorded as failing. The
+process is repeated for the same roots again but for the other mode, Natural
+exceptions.  Figure~\ref{fig:process} summarizes this experimental process for
+one mode of the rational programmer and connects it with the mutations from
+section~\ref{sec:mutate}.  After completing the experiment, the experiment
+software reports the success/failure results of the trails to determine for each
+root whether Natural blame is more useful than Natural exceptions.  Question
+$Q_1$ has a positive answer if a root exists where the above is true because it
+is evidence that there is at least one interesting scenario that the rational
+programmer manages to debug with blame information.  The process is analogous
+for $Q_2$ and $Q_3$, using the respective modes of the rational programmer.
 
-For $Q_*$, the process is a bit more
-involved. At a first level we proceed in a similar
-manner except that we use two modes 
-of the rational programmer from different semantics.
-However, answering this question requires us to compare the percentage of scenarios
-where one mode is more useful than the other and the inverse.
-For instance, to
-decide whether blame for Natural is more useful than first blame for Transient, 
-we compare the percentage of interesting scenarios where 
-Natural blame is more successful than Transient first blame 
-with the percentage of interesting scenarios where Transient first blame 
-is more successful than Natural blame. 
-Finally we repeat these steps one more time to compare Natural blame and Transient last blame
-and get a complete picture of the comparative usefulness of blame
-in the two semantics.
+For $Q_*$, the process is a bit more involved. Answering this question calls for
+a comparison of the percentage of scenarios where one mode is more useful than
+the other and the inverse.  For instance, to decide whether blame for Natural is
+more useful than first blame for Transient, the software compares the percentage
+of interesting scenarios where Natural blame is more successful than Transient
+first blame with the percentage of interesting scenarios where Transient first
+blame is more successful than Natural blame.  Finally these steps are repeated
+one more time to compare Natural blame and Transient-last blame and get a
+complete picture of the comparative usefulness of blame in the two semantics.
 

### 2021-03-01 writing/rational-natural.tex
factor the rational programmer
rational programmer with editorial voice

--- 2021-02-23 writing/rational-modes.tex
@@
 %% -----------------------------------------------------------------------------
-\subsection{Modes of the Natural Rational Programmer} \label{sub:natural}
 
@@
 
-We can turn this description into a formal definition of the trail. 
+Here is a rigorous definition.
 \begin{quote}
@@
 \it A Natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
-\emph{successful} iff its last scenario $\conf_n$ does not type check.  A Natural
+\emph{successful} iff (the program for) its last scenario $\conf_n$ does not type check.  A Natural
 blame trail $\conf_0,..,\conf_n$ in a lattice $\lattice{\system}$ is \emph{failing}
-iff $\conf_n$ type checks and the trail cannot be extended further.
+iff (the program for) $\conf_n$ type checks and the trail cannot be extended further.
 \end{quote}
-That is, failing Natural blame trails are those that have not reached a scenario that
-reveals the bug at compile time and their last scenario does not blame an untyped component. Thus
-the rational programmer gets no hints from the gradual type system on how to
-continue the search for the bug.
+That is, failing Natural blame trails are those that have not reached a scenario
+that reveals the bug at compile time and ends in a scenario that does not blame
+an untyped component during program execution. Thus the rational programmer gets
+no hints from the gradual type system on how to continue the search for the bug.
 
@@
 \begin{quote}
-\it A Natural exception trail is a sequence of scenarios $\conf_0,...\conf_n$ of a
+\it A {\em Natural exception trail\/} is a sequence of scenarios $\conf_0,...\conf_n$ of a
 program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset

### 2021-03-01 writing/rational-stuff.tex
rational programmer with editorial voice

@@
 lattice. Formally, a program $\system$ is a set of modules $\set{\component}$.
-Let a configuration $\conf$ of $\system$ be the subset of $\set{\component}$ that
-comes with (possibly type-incorrect) annotations. These configurations of
-$\system$ are ordered by the subset relation and form a lattice
-$\lattice{\system}$ with $2^{\size{\system}}$ elements. The bottom scenario of
-$\lattice{\system}$ is $\emptyset$, the top one $\system$ itself. In between are
-the mixed-typed scenarios.
+Let a configuration $\conf$ of $\system$ be the subset of $\set{\component}$
+that comes with type annotations. These configurations of $\system$ are ordered
+by the subset relation and form a lattice $\lattice{\system}$ with
+$2^{\size{\system}}$ elements. The bottom scenario of $\lattice{\system}$ is
+$\emptyset$, the top one $\system$ itself; in between are the mixed-typed
+scenarios.
 
@@
 interesting debugging scenario.  Starting from such a root, the rational
-programmer steps along a trail. If the scenario type-checks, the programmers runs
-the program until it raises a run time error.  The rational programmer uses the
-information in the error to decide which component(s) to equip with types. This
-choice constructs scenario $\conf_{i+1}$ from $\conf_{i}$ and thus creates a
-trail. The trail ends when programmer reaches a scenarios that contain the mutant
-module, because the type checker rejects it outright. At this point, the
-programmer has identified the source of the impedance mismatch.
+programmer steps along a trail. If the (program for this) scenario type-checks,
+the program is run until it raises a run-time error.  The rational programmer
+uses the information in the error to decide which component to equip with
+types. This choice constructs scenario $\conf_{i+1}$ from $\conf_{i}$ and thus
+creates a trail. The trail's construction ends when it reaches a scenarios that
+contain the mutant module because the type checker rejects its typed version
+outright.  At this point, the source of the impedance mismatch is identified.

### 2021-03-01 writing/rational-transient.tex
rational programmer with editorial voice

@@
 
-The transient semantics assigns blame to a set of components. In this setting, a
-blame assignment has the following meaning: the witness value to the impedance
-mismatch may have crossed the boundaries between elements in the set, and each crossing
+The transient semantics assigns blame to a sequence of components. The blame
+assignment says that the value witnessing the impedance mismatch may have
+crossed the boundaries between elements in the sequence, and each crossing
 checked the value's type in a shallow manner.
 
-This ambiguity in Transient blame raises the question of how the rational programmer
-should react when the language produces a blame set. Our answer is that the rational
-programmer has at least two reasonable options. The first one is to select the component that is added
-to the blame set first and assign types to only that one---after all, if
-fully checked, the types of
-this first component should be able to detect an impedance mismatch earlier in the
-evaluation of a program than the later ones. The second option is to
-select the component that is added to the blame set last, effectively
-interpreting the blame set as a boundary-aware stack.
+This ambiguity in Transient blame raises the question of how the rational
+programmer should react when the language produces a blame sequence. Our answer
+is that the rational programmer has at least two reasonable options. The first
+one is to select the component that is added to the blame sequence first and
+assign types to only that one---after all, if fully checked, the types of this
+first component should be able to detect an impedance mismatch earlier in the
+evaluation of a program than the later ones. The second option is to select the
+component that is added to the blame sequence last, effectively interpreting the
+blame sequence as a boundary-aware stack.
 
 These two modes of rationalizing give rise to two different notions of trail.
+
 \begin{quote}
@@
 \first{\mblame{\system, \conf_i}}$ where $\first{\mblame{\system, \conf}}$ is the
-first component Transient adds to the blame set for $\conf$.
+first component Transient adds to the blame sequence for $\conf$.
 \end{quote}
-
-and
 
@@
   \last{\mblame{\system, \conf_i}}$ where $\last{\mblame{\system, \conf}}$ is the
-last component Transient adds to the blame set for $\conf$.
+last component Transient adds to the blame sequence for $\conf$.
 \end{quote}
 
-The definition of \emph{Transient exception trails} 
-is analogous to that for Natural and, same as for Natural, we use it to  
-specify when  
-Transient-first and Transient-last blame are more useful than 
-Transient exceptions for debugging a scenario.
+The definition of \emph{Transient exception trails} is analogous to that for
+Natural. It is used to specify when Transient-first and Transient-last blame are
+more useful than Transient exceptions for debugging a scenario.

### 2021-03-01 writing/rational.tex
stuff
name files properly
re-established the old organization
rational programmer with editorial voice

@@
 Section~\ref{sec:why-rational} explains, and illustrates with an example, how a
-migratory type setting helps find the source of an impedance mismatch. Roughly
-speaking, it encourages the rational programmer equip a module with types if it
-is blamed in an error message. A sequence of such steps make up a portion of the
-migratory path in the lattice of type migration~\cite{tfgnvf-popl-2016} (see
-sec.~\ref{sub:stuff} for details).
+migratory type setting helps with finding the source of an impedance
+mismatch. Roughly speaking, it encourages the rational programmer to equip a
+module with types if it is blamed in an error message. A sequence of such steps
+make up a portion of the migratory path in the lattice of type
+migration~\cite{tfgnvf-popl-2016} (see sec.~\ref{sub:stuff} for details).
 
@@
 
-Hence the research problem is how to view the actions of a rational programmer
+Hence the research problem is how to make the actions of a rational programmer
 comparable across all semantics. Sections~\ref{sub:natural}
 through~\ref{sub:erasure} describe the various rational programmers and explain
-to what extent their actions are comparable.
+to what extent their actions are comparable. Section\ref{sub:experiment}
+enhances comparability with the programmer-effort dimension. Finally, the last
+subsection states the experimental questions.
 
@@
 
-\rsub{stuff}     {Stuff}
-
+\rsub{stuff}     {The Lattice and the Debugging Scenario} 
 \rsub{natural}   {The Natural Rational Programmer} 
@@
 \rsub{erasure}   {The Erasure Rational Programmer} 
+\rsub{effort}    {The Programmer Effort} \label{subsec:effort}
+\rsub{experiment}{The Experimental Questions}

### 2021-03-02 writing/landscape.tex
landscape, why-rational: fix mistakes in my previous changes

@@
 because it takes place in the untyped {\tt client} module, which is compiled in
-the usual manner. Later, when {\tt pack} is called in the {\tt crypto-pack-lib}
+the usual manner. Because {\tt pack} is called in the {\tt crypto-pack-lib}
 module, Transient's inlined checks make sure that the imported
@@
 \qquad {\it Erasure\/}   & no error 		          & no error \\
+       	    		 &   \multicolumn{2}{|c}{but {\em Erasure} does signal an error on list access}
 \end{tabular}

### 2021-03-02 writing/mutate-benchmarks-figure.tex
editiroal voice thru 6

@@
-\begin{figure}
-\begin{tabular}{p{1.75cm} | p{6cm} }
-  {\bf  name} & {\bf description (author)}  \\
+\begin{figure*}
+\begin{tabular}{p{1.4cm} | p{6.9cm} | p{2.0cm} | r | r}
+             {\bf  name} & {\bf description} & {\bf author} & {\bf loc} & {\bf mod.} \\ \hline
+%% ---------------------------------------------------------------------------------------------------------------
 
-\hline
-  \texttt{acquire} & Object-oriented board game implementation (M. Felleisen)  \\%[1em]
+\texttt{acquire} & object-oriented board game implementation      & M. Felleisen & 1941 & 9 \\ \hline
 
-\hline
-  \texttt{gregor} & Utilities for calendar dates (J. Zeppieri) \\%[1em]
+\texttt{gregor}  & utilities for calendar dates                   & J. Zeppieri  & 2336 & 13\\ \hline
 
-\hline
-  \texttt{kcfa} & Functional implementation of 2CFA for the lambda calculus (M. Might) \\%[1em]
+\texttt{kcfa}    & functional implementation of 2CFA for LC       & M. Might & 328 & 7\\ \hline
 
-\hline
-  \texttt{quadT} & Converter from S-expression source code to PDF format (M. Butterick)\\%[1em]
+\texttt{quadT}   & converter from S-expression source code to PDF & M. Butterick & 7396 & 14\\ \hline
 
-\hline
-  \texttt{quadU} & Converter from S-expression source code to PDF format  (B. Greenman) \\%[1em]
+\texttt{quadU}   & converter from S-expression source code to PDF & B. Greenman  & 7282 & 14 \\ \hline
 
-\hline
-  \texttt{snake} & Functional implementation of the  Snake video game (D. Van Horn) \\%[1em]
+\texttt{snake}   & functional implementation of the Snake game    & D. Van Horn & 182 & 8 \\ \hline
 
-\hline
-  \texttt{suffixtree} & Algorithm for common longest subsequences between strings. (D. Yoo) \\%[1em]
+\texttt{synth}  & converter of notes and drum beats to WAV & V. St-Amour & 871 & 10 \\ \hline
 
-\hline
-  \texttt{synth} & Converter of notes and drum beats to WAV (V. St-Amour \& N. Toronto) \\%[1em]
+\texttt{take5} & Mixin-based card game simulator & M.Felleisen & 465 & 8\\ \hline
 
-\hline
-  \texttt{take5} & Mixin-based card game simulator (M.Felleisen)  \\%[1em]
+\texttt{tetris} & Functional implementation of Tetris & D. Van Horn &   280 & 9 \\ \hline
 
-\hline
-  \texttt{tetris} & Functional implementation of Tetris (D. Van Horn) \\%[1em]
+\texttt{suffix\-tree} & algorithm for common longest subsequences between strings. & D. Yoo & 1500 & 6 \\
 
 \end{tabular}
-  \caption{Benchmarks summary.}
-  \label{table:benchmark-descriptions}
-\end{figure}
+
+\caption{Benchmarks summary.} \label{table:benchmark-descriptions}
+
+\end{figure*}

### 2021-03-02 writing/mutate-benchmarks.tex
editiroal voice thru 6

--- 2021-03-01 writing/mutate-benchmarks.tex
@@
-Since this is the first empirical attempt to evaluate blame in gradual
-typing, there is no standard benchmark suite that we can use off-the-self.
-In addition, there is no exsitng benchmark suite of gradually typed
-programs with type-level mistakes that we could
-repurpose.\footnote{Recently ~\citet{cc-oopsla-20} created a
-Reticulated Python program collection to evaluate their technique for
-fixing mistakes in type annotations; unfortunately their collection does
-not come with type-level mistakes in the code proper.} In fact, the bar
-for testing the rational programmer is much higher than the existence of a
-collection of programs with type-level errors. Testing the rational
-programmer demands a currated collection of programs that take advantage
-of the sophisticated features of gradually typed languages such as Typed
-Racket.
+%% -----------------------------------------------------------------------------
+
+Since this is the first empirical attempt to evaluate blame in gradual typing,
+there is no standard benchmark suite that we can use off-the-self.  In addition,
+there is no exsitng benchmark suite of gradually typed programs with type-level
+mistakes.\footnote{\citet{cc-oopsla-20} created a collection of Reticulated
+Python programs to evaluate their technique of fixing mistakes in type
+annotations. Their collection does not come with type-level mistakes in the code
+itself.} In fact, the bar for testing the rational programmer is much higher
+than the existence of a collection of programs with type-level errors. Testing
+the rational programmer demands a currated collection of programs that take
+advantage of the sophisticated features of a gradually typed language such as
+Typed Racket.
 

### 2021-03-02 writing/mutate-interesting.tex
bring back interesting=plot
editiroal voice thru 6

@@
 
-Let us consider the three conditions in order:
+Here is the rationale for these three conditions:
 \begin{enumerate}
@@
 
-%\begin{figure*}
-%  \centering
-%  \includegraphics[scale=0.35]{./plots/mutant-breakdown}
-%  \caption{Interesting debugging scenarios produced by our mutators. Counts are cut off at 50.}
-%  \label{fig:mutant-breakdown}
-%\end{figure*}
+\begin{figure*} \footnotesize
+ \centering
+ \includegraphics[scale=0.33]{./plots/mutant-breakdown}
+  Counts are cut off at 50.
+
+\caption{The mutators produce Interesting debugging scenarios across all benchmarks.} 
+ \label{fig:mutant-breakdown}
+\end{figure*}
+
+The definition of interesting debugging scenarios creates a powerful filter. All
+together, the listed mutators produce 16,800 mutants with at least one
+interesting debugging scenario across all benchmarks; see
+figure~\ref{fig:mutant-breakdown} for an overview. Broken down by benchmark, the
+mutators produce at least 40 interesting scenarios for every benchmark, and
+these scenarios originate from at least four different mutators per benchmark.
+Thus, the mutators result in a sizable and diverse population of scenarios for
+every benchmark.  Furthermore, every mutator contributes scenarios to at least
+one benchmark.  Some mutators apply only to a few benchmarks, because they
+target rather specific features; for instance, the class-focused mutators are
+mainly effective in a program that makes extensive use of object-oriented
+features.

### 2021-03-02 writing/mutate-mutators.tex
editiroal voice thru 6

@@
 
-Figure~\ref{table:mutation-ops} describes 16 mutators that satisfy this
-requirement.  For each mutator, the figure provides a short description and an
-example. As the explanation indicates, some specialize or
+Figure~\ref{table:mutation-ops} describes 16 mutators that satisfy these
+constraints. As the explanation indicates, some specialize or
 generalize~\citet{lksfd-popl-2020}'s mutators, which in turn are borrowed from
 the vast literature on mutation testing~\cite{jia2011analysis}.\footnote{Many
-of~\citet{lksfd-popl-2020} mutators do not satisfy the constraints. For
-example, the relational mutator does not lead to type-level errors that Typed
-Racket's type system can detect, so it has been eliminated. Similarly, the
-original arithmetic mutator does not reliably lead to type errors, so it has
-been adapted to only make operator swaps that can affect the type of the
-result. Changing a `/` to `*` does not affect the type of an expression, but
-changing `*` to `/` may, because division by zero is detectable by Typed
-Racket's type system.}  Only two are directly inherited; many mutators are
-brand new. For the latter the authors relied on their decades-long experience
-of making type-level mistakes in Typed Racket, some of which take a non-trivial
-effort to debug.
+of~\citet{lksfd-popl-2020} mutators do not satisfy the constraints and had to
+be eliminated. For example, the relational mutator does not lead to type-level
+errors. Similarly, the arithmetic mutator does not reliably lead to type
+errors; it had to be adapted to make only operator swaps that can affect the
+type of the result, e.g., changing `*` to `/`.}  Only two are directly
+inherited; many mutators are brand new. For the latter the authors relied on
+their decades-long experience of making type-level mistakes in Typed Racket,
+some of which take a non-trivial effort to debug.
 

### 2021-03-02 writing/discussion.tex
organize sections/titles/labels in paper
results, discussion, conclusion: minor rewordings and fixes
Only change in conclusion is the wording "Problem is" that two of our PLDI
reviewers complained about.

@@
 %% -----------------------------------------------------------------------------
-\section{What We Learned}
-\label{sec:discussion}
-
 %% \begin{itemize}
@@
 Second, in many cases type checks that issue blame are more helpful than
-those that do not. However, our results indicate that in other cases blame
-is not critical and thus it is worth further investigation whether
+those that do not. However, our results indicate that blame
+is not critical in a majority of cases, and therefore it is worth further investigation whether
 disabling blame tracking for the sake of performance may be   an
-acceptable trade off. Third, from the three different blame strategies, the
+acceptable trade-off. Third, from the three different blame strategies, the
 Natural approach of Typed Racket fares better than the Transient approach
@@
 whether  the rational programmer follows first or the last element from a
-blame set. Hence, given the performance issues of Transient we discuss further on, 
+blame set, and that using the last element seems to produce shorter trails.
+Hence, given the performance issues of Transient we discuss further on,
 a possible optimization is to limit the size of blame sets
-using timestamps.
+to only contain the latest entries using timestamps.
 

### 2021-03-02 writing/results.tex
results, discussion, conclusion: minor rewordings and fixes
Only change in conclusion is the wording "Problem is" that two of our PLDI
reviewers complained about.

@@
 sampled debugging scenarios from section~\ref{sec:sample} using a machine with two Intel Xeon Gold 6258R processors (28 doubly-threaded cores each) and 500GB of memory.
-Each debugging scenario had a timeout limit of 4 minutes and a
+Each debugging scenario had a 4 minute timeout and a
 memory limit of 6GB. The complete experiment required
-over 30,000 hours of compute hours.
+over 30,000 compute hours.
 
@@
   \caption{Success baselines: The estimated percentage of scenarios for which each mode succeeds in locating the bug.
-  The upper bound of error margin is 0.02\%.
+  The upper bound margin of error is 0.02\%.
   }
@@
   The portion below 0 (in red) is the estimated percentage of scenarios with the inverse situation.
-  The upper bound of error margin is 0.02\%.
+  The upper bound margin of error is 0.02\%.
   }
@@
   For example, in the breakdown of Natural, all three modes succeed on the same scenario 57.3\% of the time, only Natural and Natural exceptions succeed on 29.1\% of the scenarios, only Natural and Erasure succeed on 2.1\%, and Natural alone succeeds on 9\%.
-  The upper bound of error margin is 0.02\%.
+  The upper bound margin of error is 0.02\%.
   }
@@
 discuss above, some modes are more successful than others in certain
-scenarios. Indeed any mode can fail if running a scenario results in an
-exception and the exception carries no useful information about which
-module the programmer should type next, e.g., because the stack trace 
+scenarios. Indeed any mode can fail if no untyped module from the program is identified in the error.
+For example, running a scenario may result in an
+exception that carries no useful information about which
+module the programmer should type next, e.g. because the stack trace
 does not contain frames from any module of the program, or only typed modules. This situation
-can happen several steps along a blame trail or even immediately (figure~\ref{fig:effort-table} dives into this below).
+can happen at any point along a blame trail (figure~\ref{fig:effort-table} dives into this below), causing it to fail.
 
@@
 number of failures because scenarios hit the time and memory
-limits of our experiment (~ 770 scenarios).  Additionally, there are nearly a 1000 cases where
-Transient reports an empty blame set which leaves the rational programmer
+limits of our experiment (\textasciitilde770 scenarios).  Additionally, there are nearly 1,000 cases where
+Transient reports an empty blame set, leaving the rational programmer
 without hints about how to proceed.
-In~\ref{sec:threat:transient}, we discuss how both of these causes of
+In section~\ref{sec:threat:transient}, we discuss how both of these causes of
 failure for Transient affect our experiment.
@@
 The only exception is that the Random programmer always succeeds;
-this reflects the fact that every scenario has finitely many modules and this mode can never get stuck.
+this reflects the facts that every scenario has finitely many modules and this mode can never get stuck, so it always eventually types the buggy module.
 
-Figure~\ref{fig:avo-bars} depicts a head-to-head comparison of every mode's performance against every other mode, helping to answer the four questions from section~\ref{sub:experiment}.
+Figure~\ref{fig:avo-bars} depicts a head-to-head comparison of every mode's performance against every other mode (except Random), helping to answer the four questions from section~\ref{sub:experiment}.
 Each plot shows a breakdown of the proportion of scenarios where one mode performs better or worse than each other mode.
-In particular, each bar above zero represents the proportion where the plot's entitled mode succeeds and the mode on the x-axis fails; the correspnding bar below zero represents the proportion of the opposite case.
-For example, the plot titled ``Natural'' shows that Natural outperforms Natural exceptions in about 10\% of the scenarios, and the inverse (Natural performs wrse than Natural exceptions) never happens.
-Similarly, the plot titled ``Transient last blame'' shows that Transient last blame outperforms Natural exceptions about in abut 10\% of the scenarios, but conversely it performs worse than Natural exceptions in about 2\% of the scenarios.
+In particular, each bar above zero represents the proportion where the plot's entitled mode succeeds and the mode on the x-axis fails; the corresponding bar below zero represents the proportion of the opposite case.
+For example, the plot titled ``Natural'' shows that Natural outperforms Natural exceptions in about 12\% of the scenarios, and the inverse (Natural performs worse than Natural exceptions) never happens.
+Similarly, the plot titled ``Transient last blame'' shows that Transient last blame outperforms Natural exceptions about in about 12\% of the scenarios, but conversely it performs worse than Natural exceptions in about 2\% of the scenarios.
 
 Figure~\ref{fig:avo-bars} shows that the answers for questions $Q_1$ to $Q_3$ are all positive.
-In all three semantics, blame modes outperform their corresponding exception mode (by around 10\% for each one).
+In all three semantics, blame modes outperform their corresponding exception mode (by \textasciitilde10\% for each one).
 The Natural exceptions mode is never more useful than Natural blame, and  Transient
@@
 Blame for all three semantics is significantly
-more useful than Erasure exceptions (by around 30\%). Natural blame is
+more useful than Erasure exceptions (by around 35\%). Natural blame is
 more useful than both versions of Transient blame by a small percentage;
@@
 Each diagram breaks breaks down the overlap in the success of the blame mode, its corresponding exception mode, and Erasure.
-For example, in the breakdown of Natural, the diagram shows that all three modes succeed on the same scenario 57.3\% of the time, only Natural and Natural exceptions succeed on 29.1\% of the scenarios, only Natural and Erasure succeed on 2.1\%, and Natural alone succeeds on 9\%.
-This analysis highlights the success tradeoffs each semantics offers against Erasure, with and without blame, to answer questions about the tradeoffs in choosing to use blame or just exceptions for each semantics.
+For example, in the breakdown of Natural, the diagram shows that all three modes succeed on the same scenario 59.5\% of the time, only Natural and Natural exceptions succeed on 25.7\% of the scenarios, only Natural and Erasure succeed on 2.1\%, and Natural alone succeeds on 9.7\%.
+This analysis highlights the success trade-offs each semantics offers against Erasure, with and without blame, to answer questions about choosing to use blame or just exceptions for each semantics.
 For instance, the breakdown of Natural clearly illustrates that if choosing between Natural blame, Natural exceptions, and Erasure, Natural blame is the absolutely most successful: all of the successes of the other two modes are subsets of Natural's successful scenarios.
@@
 debugging interesting scenarios (in green) for the random mode of the
-rational programmer follows  a normal distribution, as expected.\footnote{The
+rational programmer is highly spread out, as expected.\footnote{The
 random mode distribution is indistinguishable for all three semantics and thus the figure~\ref{fig:effort-table} 
@@
 the rational programmer fails to debug a scenario altogether (in red).
-This is expected; no matter how many type annotations the programmer
-adds to an Erasure program, if the type checker doesn't reject the
-program, running the program always produces the same outcome. Thus an
+This is expected; adding type annotations to an Erasure program doesn't
+change it's behavior, so if the type checker doesn't reject the
+program with the new annotations, the rational programmer is stuck.
+In other words, an
 exception from the runtime has to point to the buggy module with
 the first try. Otherwise the rational programmer types an irrelevant
-module, runs the program again and the exception points again to the
+module, runs the program again, and the same exception points again to the
 already typed module.
@@
 To directly compare effort between modes, we compute the difference in trail length between two modes for all scenarios where they both succeed.
-Hence, each plot in the figure shows the distribution of scenarios with length differences ranging from -3 (the first mode's trail is 3 steps shorter than the second) to 3 (the first mode's trail is 3 steps longer than the second).
+Hence, each plot in the figure shows the distribution of scenarios with length differences ranging from -3 (the first mode's trail is 3 steps shorter than the second's) to 3 (the first mode's trail is 3 steps longer than the second's).
 The figure offers several insights about the comparative effort for each mode that complement the insights about comparitive success from figure~\ref{fig:avo-bars}.
@@
 Second, Natural relatively often (around 10\% of the scenarios) produces shorter trails than both Transient blame modes, and they are sometimes significantly shorter.
-Third, Transient last blame often produces shorter trails than Transient first blame, which is a particularly interesting distinction in light of their nearly undistinguishable success rates.
+Third, Transient last blame often produces shorter trails than Transient first blame, which is a particularly interesting distinction in light of their nearly indistinguishable success rates.
 Finally, Transient's blame modes also share the characteristic with Natural that blame sometimes produces longer trails than their corresponding exception modes.

### 2021-03-02 writing/rational-methodology.tex
adding methodology subsection

Each time the rational programmer debugs a program, it is
validation for programming languages researchers; their blame assignment 
mechanisms provide information that points towards the source of the
problem. Moroeover, their slogans and theorems help programmers connect the dots.
In contrast, when the rational programmer fails, it speaks for the
limited predictive power of programming languages theory and the
misleading projections of language design.

In other words the rational programmer implies a methodology for
evaluating
the design of programming lanaguages features such as blame assignment mechanisms. 
In general terms, the methodology consists of four 
pieces:

\begin{itemize}

\item The description of a  rational programmer as an algorithmic
  procedure that examines a program and performs an action. 
    Typically the action of the rational program is consistent 
   with the theory of the features under examination.

  \item A hypothesis about whether a sequence of actions of the
    rational programmer leads to a desired outcome.

\item A large-scale experiment that tests the hypothesis 
  on a large number of representative scenarios.

\item The evaluation of the features based on the outcome of the
  experiment.

\end{itemize}

In specific terms for blame assignment mechanisms, the methodology
imagines:

\begin{itemize}
\item The rational programmer as a procedure that consumes a program and the error message 
  the evaluation of the program produces and, uses the error message to
    select the next component of the program to add types.

  \item The rational programmer hypothesis as  research question whether 
    for all programs with
    type-level mistakes a series of steps of the rational programmer always leads to a type
    error that points to the component that contains the mistake.

\item The experiment as the attempts of the rational programmer to debug a
  large number of programs with type-level mistakes.

\item The evaluation of a blame assignment mechanism based on whtehr the
  information it provides was critical for the outcome of the experiment   
\end{itemize}

### 2021-03-02 writing/why-rational.tex
landscape, why-rational: fix mistakes in my previous changes
adding methodology subsection

@@
 
-This paper dubs Lazarek et. all's idea {\em the rational programmer\/}. Like Stuart
+This paper dubs Lazarek et. al's idea {\em the rational programmer\/}. Like Stuart
 Mill's {\it homo economicus\/}, the rational programmer approximates the
@@
 step is to inspect the {\tt layer} module and to equip it with type
-annotations. This results in the \typecolor\ module in the middle, and {\tt
+annotations. The result is the \typecolor\ module in the middle, and {\tt
 main}'s import is now re-directed there by {\tt (submod ".." layer)}. Running the
@@
 will help language designers.
+
+\subsection{How to Turn the Idea of the Rational Programmer to a Methodology}
+
+\input{rational-methodology.tex}

### 2021-03-02 writing/rational-methodology.tex
adding methodology subsection

@@
 mechanisms provide information that points towards the source of the
-problem. Moroeover, their slogans and theorems help programmers connect the dots.
+problem. Moreover, their slogans and theorems help programmers connect the dots.
 In contrast, when the rational programmer fails, it speaks for the
@@
 evaluating
-the design of programming lanaguages features such as blame assignment mechanisms. 
+the design of programming languages features such as blame assignment mechanisms. 
 In general terms, the methodology consists of four 
@@
 
-\item The evaluation of a blame assignment mechanism based on whtehr the
+\item The evaluation of a blame assignment mechanism based on whether the
   information it provides was critical for the outcome of the experiment   

### 2021-03-02 writing/challenges.tex
pass over challenges section

@@
 
-The rational programmer works like an algorithm. Implementing it demands a large
-number of representative debugging scenarios; a reconciliation of three
-different kinds of error messages; and a proper sampling strategy for the huge
-space of possibilities. A coincidental challenge is the disparity of gradually
-typed languages. To eliminate this variable from the comparison, this paper uses
-Racket, which is thus far the only language in which all three major soundness
-mechanisms are available in a reasonably robust and directly comparable manner;
-for the implementation of the transient semantics, see the parallel submission
-of an experience report~\cite{ttt21}.
+At first glance implementing the methodology of
+section~\ref{sec:why-rational} seems straightforward. The rational
+programmer is a simple algorithm and testing whether the rational
+programmer hypothesis holds amounts to running the algorithm on a
+collection of scenarios.
 
-The {\em first challenge\/} concerns the differences among the blame assignment
-of the three semantic variants.  While Natural assigns blame to {\em one\/}
-component, Transient assigns blame to a sequence of components. The Erasure
-semantics does not blame a component {\it per se\/}, but it comes with an
-exception location and a stack trace, which implicitly suggests parties to
-blame.  The different strategies greatly affect the behavior of the rational
-programmer (and real programmers, too).
+However, three major challenges remain:  the isolation of the effect of
+blame on the rational programmer across three different mechanisms; the
+collection of a large number of representative debugging scenarios;  and a
+proper sampling strategy for the huge space of possibilities. A
+coincidental challenge is the disparity of gradually typed languages. To
+eliminate this variable from the comparison, this paper uses Racket, which
+is thus far the only language in which all three major soundness
+mechanisms are available in a reasonably robust and directly comparable
+manner; for the implementation of the transient semantics, see the
+parallel submission of an experience report~\cite{ttt21}.
 
-One way to overcome this challenge is to {\em equip the rational programmer with
-modes \/} that represent the alternative interpretations of blame.  For
-instance, one Transient mode may pick the oldest element of a blame set as
-the next component to add types because it corresponds to the earliest point in
-the program's evaluation that can discover an impedance mismatch.  Another mode
-may opt to treat the sequence as a stack and add types to its newest element.
-When both modes are equally successful in locating an impedance mismatch,
-measuring the rational programmer's effort may answer which mode is the most
+The {\em first challenge\/} stems from the interplay between blame and the
+run-time checks of a gradual type systems. When a check fails the Natural
+and Transient semantics assign blame and discard  the information in the
+exception from the run-time check. Nevertheless, the exception
+information may be equally or even more useful to the rational programmer
+than blame. If this is the case blame may not play a critical role for the
+rational programmer.\footnote{Indeed, precisely because they do not
+account for confounding factors, ~\citet{lksfd-popl-2020} cannot draw any
+conclusions about the pragmatics of blame despite advertisements for the
+opposite. Their experiment can only conclude that so-called blame-shifting
+works but not that this is because of blame.}
+
+One way to isolate blame from this confounding factor is to {\em equip the
+rational programmer with modes \/} that represent the different types of
+information the rational programmer takes into account when debugging a
+scenario. Specifically, the rational programmer comes with a blame and
+an exception mode. Only if the former succeeds debugging a program while
+the latter fails, it is safe to conclude that blame is indeed necessary
+for the rational programmer. Put differently, the exception mode serves as
+the {\em baseline\/} for blame's value; if the programmer in this mode
+``beats'' the  blame one, a blame assignment mechanism might be useless.
+
+Besides the run-time checks, pure luck might also amplify the usefulness
+of blame. Thus a proper evaluation of blame needs to factor out any random
+effects. A random rational programmer mode offers the necessary baseline
+using the same ingredients as for exceptions.
+
+Modes also offer a common vocabulary for the differences among the blame
+assignment of the three semantic variants.  While Natural assigns blame to
+{\em one\/} component, Transient assigns blame to a sequence of
+components. The Erasure semantics does not blame a component {\it per
+se\/}, but it comes with an exception location and a stack trace, which
+implicitly suggests parties to blame.  The different strategies greatly
+affect the behavior of the rational programmer (and real programmers,
+too).
+
+With modes, different strategy map to different ways the rational
+programmer  interprets blame.  For instance, one Transient mode may pick
+the oldest element of a blame set as the next component to add types
+because it corresponds to the earliest point in the program's evaluation
+that can discover an impedance mismatch.  Another mode may opt to treat
+the sequence as a stack and add types to its newest element.  When both
+modes are equally successful in locating an impedance mismatch, measuring
+the rational programmer's effort may answer which mode is the most
 effective.
 
-Different blame information is also due to differences in the run-time checks
-that the three gradual typing systems perform. This observation suggests another
-mode for the rational programmer, namely, a mode that ignores the blame
-information completely and instead inspects only the information in the
-exceptions of the safety checks in the underlying languages.  This exception
-mode serves as the {\em baseline\/} for blame's value; if the programmer in this
-mode ``beats'' the others, blame systems might be useless.
-
 The {\em second challenge\/} is to find a representative collection of
-impedance mismatches. They must represent mistakes that programmers
+programs with impedance mismatches. The latter must represent mistakes that programmers
 accidentally create and that the run-time checks of academic systems catch. In
@@
 Roughly speaking, the new mutators inject type errors into fully typed programs.
-Applying such a mutator to any typed component produces a mutatnt component.  A
-debugging scenario results from removing the types from the mutatnt. For the
+Applying such a mutator to any typed component produces a mutant component.  A
+debugging scenario results from removing the types from the mutant. For the
 design of such mutators, the authors relied on their own extensive programming
@@
 in detail.
-
-% Overcoming these challenges yields a systematic approach: 
-% \begin{itemize}
-
-% \item interesting mutators generate interesting type-level errors and thus
-% debugging scenarios (section~\ref{sec:mutate});
-
-% \item an algorithm simulates the modes of the rational programmer on these
-% scenarios (section~\ref{sec:rational}); and
-
-% \item a careful sampling of the space yields experimental
-% data about the bug finding process (section~\ref{sec:sample}).
-% \end{itemize}
-% Getting all three aspects right is necessary to obtain a computationally
-% feasible and statistically sound experiment.
-
-% {\bf Note} A coincidental challenge is the disparity of gradually typed
-% languages. While the three approaches to blame are well understood in theory,
-% they have been implemented for rather different untyped languages and with
-% vastly different type systems. To eliminate the effect of these differences, it
-% is necessary to implement all three approaches in a single language so that an
-% application of the evaluation method can use the same test programs, the same
-% type system, and hence the same debugging scenarios.
-
-% The results in this paper are for a system that builds on Greenman and
-% Felleisen's effort of comparing the performance of Natural and Transient in
-% Racket~\cite{gf-icfp-2018}. The extended system uses the type system of Typed
-% Racket and comes with a Transient semantics that faithfully implements the
-% design and the blame assignment strategy proposed by~\cite{vss-popl-2017}. The
-% Erasure semantics drops the type information and runs the programs as if they
-% were plain Racket programs. The next section explains this system in detail.

### 2021-03-02 writing/rational-methodology.tex
adding methodology subsection
pass over challenges section

@@
-Each time the rational programmer debugs a program, it is
-validation for programming languages researchers; their blame assignment 
-mechanisms provide information that points towards the source of the
-problem. Moreover, their slogans and theorems help programmers connect the dots.
-In contrast, when the rational programmer fails, it speaks for the
-limited predictive power of programming languages theory and the
-misleading projections of language design.
+Each time the rational programmer debugs a program, it is validation for
+programming languages researchers; their blame assignment mechanisms
+provide information that points towards the source of the problem.
+Moreover, their slogans and theorems help programmers connect the dots.
+In contrast, when the rational programmer fails, it speaks for the limited
+predictive power of programming languages theory and the misleading
+projections of language design.
 
 In other words the rational programmer implies a methodology for
-evaluating
-the design of programming languages features such as blame assignment mechanisms. 
-In general terms, the methodology consists of four 
+evaluating the design of programming languages features such as blame
+assignment mechanisms.  In general terms, the methodology consists of four
 pieces:
@@
   \item The description of a  rational programmer as an algorithmic
-  procedure that examines a program and performs an action. 
-    Typically the action of the rational program is consistent 
-   with the theory of the features under examination.
+    procedure that examines a program and performs an action.  Typically
+    the action of the rational program is consistent with the theory of
+    the features under examination.
 
-  \item A hypothesis about whether a sequence of actions of the
-    rational programmer leads to a desired outcome.
+  \item A hypothesis about whether a sequence of actions of the rational
+    programmer leads to a desired outcome.
 
-\item A large-scale experiment that tests the hypothesis 
-  on a large number of representative scenarios.
+  \item A large-scale experiment that tests the hypothesis on
+    representative programs.
 
@@
 \begin{itemize}
-\item The rational programmer as a procedure that consumes a program and the error message 
-  the evaluation of the program produces and, uses the error message to
-    select the next component of the program to add types.
+
+  \item The rational programmer as a procedure that consumes a program and
+    the error message from evaluation of the program and uses the error
+    message to select the next component to add types.
 
   \item The rational programmer hypothesis as  research question whether
-    for all programs with
-    type-level mistakes a series of steps of the rational programmer always leads to a type
-    error that points to the component that contains the mistake.
+    for all programs with type-level mistakes a series of steps of the
+    rational programmer always leads to a type error that points to the
+    component that contains the mistake.
 
-\item The experiment as the attempts of the rational programmer to debug a
-  large number of programs with type-level mistakes.
+  \item The experiment as the attempts of the rational programmer to debug
+    a large number of programs with type-level mistakes.
 
-\item The evaluation of a blame assignment mechanism based on whether the
-  information it provides was critical for the outcome of the experiment   
+  \item The evaluation of a blame assignment mechanism based on whether
+    the information it provides is critical for the outcome of the
+    experiment.
+
 \end{itemize}

### 2021-03-02 writing/challenges.tex
pass over challenges section

@@
 than blame. If this is the case blame may not play a critical role for the
-rational programmer.\footnote{Indeed, precisely because they do not
+rational programmer. Indeed, precisely because they do not
 account for confounding factors, ~\citet{lksfd-popl-2020} cannot draw any
@@
 opposite. Their experiment can only conclude that so-called blame-shifting
-works but not that this is because of blame.}
+works but not that this is because of blame.
 
@@
 information the rational programmer takes into account when debugging a
-scenario. Specifically, the rational programmer comes with a blame and
+scenario. Specifically, the rational programmer comes with a blame mode and
 an exception mode. Only if the former succeeds debugging a program while
@@
 of blame. Thus a proper evaluation of blame needs to factor out any random
-effects. A random rational programmer mode offers the necessary baseline
+effects. A random rational programmer mode provides the necessary baseline
 using the same ingredients as for exceptions.
@@
 
-With modes, different strategy map to different ways the rational
+With modes, different blame strategies map to different ways the rational
 programmer  interprets blame.  For instance, one Transient mode may pick

### 2021-03-02 writing/rational.tex
pass over section 5

@@
 
-Section~\ref{sec:why-rational} explains, and illustrates with an example, how a
-migratory type setting helps with finding the source of an impedance
-mismatch. Roughly speaking, it encourages the rational programmer to equip a
-module with types if it is blamed in an error message. A sequence of such steps
-make up a portion of the migratory path in the lattice of type
-migration~\cite{tfgnvf-popl-2016} (see sec.~\ref{sub:stuff} for details).
+Section~\ref{sec:why-rational} explains, and illustrates with an example,
+how a migratory type setting helps with finding the source of an impedance
+mismatch. Roughly speaking, it encourages the rational programmer to
+equip a module with types if it is blamed in an error message. A sequence
+of such steps make up a portion of the migratory path in the lattice of
+type migration~\cite{tfgnvf-popl-2016}. The lattice, described in Section
+~\ref{sub:stuff},  serves as the common substrate for the definition modes
+of the rational programmer mentioned in section~\ref{sec:challenges}.
 
-Unfortunately, the rational programmer receives different kinds of information
-and thus may construct different paths. As section~\ref{sec:landscape} shows,
-the three different semantics deal with the same program in three different
-ways, and the error messages point to different sources of a problem.
+Each mode receives different kinds of information and thus may construct
+different paths in the lattice. As discussed in~\ref{sec:challenges},
+evaluating blame relies on comparing modes of the rational programmer
+within the same semantics and across different semantics.
 
@@
 comparable across all semantics. Sections~\ref{sub:natural}
-through~\ref{sub:erasure} describe the various rational programmers and explain
-to what extent their actions are comparable. Section\ref{sub:experiment}
+through~\ref{sub:erasure} describe the various mode of the rational programmer and explain
+to what extent they are comparable. Section\ref{sub:experiment}
 enhances comparability with the programmer-effort dimension. Finally, the last
-subsection states the experimental questions.
+subsection states the experimental questions and gives an overview of the
+experimental process.
 

### 2021-03-02 writing/mutate-interesting.tex
pass over section 6

@@
 
-\item By definition, an impedance mismatch is a class between type ascription of
-one module to another module's exports. Hence, type checking fails for an
+\item By definition, an impedance mismatch is a clash between the type ascription of
+one module's imports and another module's exports. Hence, type checking fails for an
 interesting mutant module, and removing the types from this mutant module
@@
 
-\item If the evaluation o mutant module immediately raises an exception because
+\item If the evaluation od a mutant module immediately raises an exception because
 of the changes, there is no work for the rational programmer. Indeed, if the
@@
 between the buggy module with its context.  In these cases, the rational
-programmer must examine at least modules to locate the source of the faulty
+programmer must examine at least two modules to locate the source of the faulty
 interaction.

### 2021-03-02 writing/introduction.tex
touching up the intro

@@
 Racket~\cite{tf-dls-2006,tf-popl-2008,tfffgksst-snapl-2017,tf-icfp-2010}) come
-with sophisticated blame assignment strategies. Their academic creators embrace
+with sophisticated checking and blame assignment strategies (see sec.~\ref{sec:landscape}). Their academic creators embrace
 the idea that blame can help practicing programmers find impedance mismatches
@@
 program. The rational programmer can hence use the former to change the
-latter---and this systematic, information-driven process can be
-implemented and tested at scale on real programs.\footnote{\citet{cc-snapl-19} use this tracing method to localize
-type errors in inference-based gradual typing systems.}  Testing it systematically requires
-overcoming major challenges: injecting representative impedance mismatches;
-putting the various kinds of error information to comparable use; and sampling
-the huge space of possibilities (see sec.~\ref{sec:challenges}).
+latter---and this systematic, information-driven process can be implemented and
+tested at scale on real programs.  Turning this idea into a scientific
+experiment requires overcoming major challenges: injecting representative
+impedance mismatches; putting the various kinds of error information to
+comparable use; and sampling the huge space of possibilities (see
+sec.~\ref{sec:challenges} and, for details, secs.~\ref{sec:rational} through~\ref{sec:sample}).
 
@@
 problems suggests that, on one hand, the existing theory does not predict
-practice properly, and on the other hand, the existing practice may need to find
-additional ways to collect data.
+practice properly, and on the other hand, the existing practice may need to
+think of additional experiments to collect more data.

### 2021-03-02 writing/challenges.tex
touch up in challenges (section 4) and opening of rational (section 5)

@@
 proper sampling strategy for the huge space of possibilities. A
-coincidental challenge is the disparity of gradually typed languages. To
-eliminate this variable from the comparison, this paper uses Racket, which
-is thus far the only language in which all three major soundness
-mechanisms are available in a reasonably robust and directly comparable
-manner; for the implementation of the transient semantics, see the
-parallel submission of an experience report~\cite{ttt21}.
+coincidental challenge is the disparity of the implementations of
+gradually typed languages. To eliminate this variable from the comparison,
+this paper uses Racket, which is thus far the only language in which all
+three major soundness mechanisms are available in a reasonably robust and
+directly comparable manner; for the implementation of the transient
+semantics, see the parallel submission of an experience
+report~\cite{ttt21}.
 
-The {\em first challenge\/} stems from the interplay between blame and the
+The {\em first major challenge\/} stems from the interplay between blame and the
 run-time checks of a gradual type systems. When a check fails the Natural
@@
 
-Modes also offer a common vocabulary for the differences among the blame
-assignment of the three semantic variants.  While Natural assigns blame to
-{\em one\/} component, Transient assigns blame to a sequence of
-components. The Erasure semantics does not blame a component {\it per
-se\/}, but it comes with an exception location and a stack trace, which
-implicitly suggests parties to blame.  The different strategies greatly
-affect the behavior of the rational programmer (and real programmers,
-too).
+Modes also offer a solution to a derivative challenge; reconciling the
+differences among the blame assignment mechanisms of the three semantic variants.
+While Natural assigns blame to {\em one\/} component, Transient assigns
+blame to a sequence of components. The Erasure semantics does not blame a
+component {\it per se\/}, but it comes with an exception location and a
+stack trace, which implicitly suggests parties to blame.  The different
+strategies greatly affect the behavior of the rational programmer (and
+real programmers, too).
 
@@
 the sequence as a stack and add types to its newest element.  When both
-modes are equally successful in locating an impedance mismatch, measuring
-the rational programmer's effort may answer which mode is the most
-effective.
+modes are equally successful in locating an impedance mismatch,
+measuring the rational programmer's effort may answer which mode is the
+most effective.
 
-The {\em second challenge\/} is to find a representative collection of
+The {\em second major challenge\/} is to find a representative collection of
 programs with impedance mismatches. The latter must represent mistakes that programmers
@@
 
-The {\em third challenge\/} is the explosive number of debugging scenarios that
+The {\em third major challenge\/} is the explosive number of debugging scenarios that
 result from the combination of mutation-based scenario generation and mode-based

### 2021-03-02 writing/rational.tex
touch up in challenges (section 4) and opening of rational (section 5)

@@
 
-Hence the research problem is how to make the actions of a rational programmer
-comparable across all semantics. Sections~\ref{sub:natural}
-through~\ref{sub:erasure} describe the various mode of the rational programmer and explain
+Hence the research problem is how to make the modes of the rational programmer
+comparable even when they correspond to different semantics. Sections~\ref{sub:natural}
+through~\ref{sub:erasure} describe the various modes of the rational programmer and explain
 to what extent they are comparable. Section\ref{sub:experiment}

### 2021-03-02 writing/challenges.tex
a bit more work on challenges

@@
 
-However, three major challenges remain:  the isolation of the effect of
+However, three major challenges remain:  comparing the effect of
 blame on the rational programmer across three different mechanisms; the
@@
 
-The {\em first major challenge\/} stems from the interplay between blame and the
-run-time checks of a gradual type systems. When a check fails the Natural
-and Transient semantics assign blame and discard  the information in the
-exception from the run-time check. Nevertheless, the exception
-information may be equally or even more useful to the rational programmer
-than blame. If this is the case blame may not play a critical role for the
-rational programmer. Indeed, precisely because they do not
+The {\em first challenge\/} stems from the differences between the blame
+assignment mechanisms of the three semantic variants.  While Natural
+assigns blame to {\em one\/} component, Transient assigns blame to a
+sequence of components. The Erasure semantics does not blame a component
+{\it per se\/}, but it comes with an exception location and a stack trace,
+which implicitly suggests parties to blame.  The different strategies
+greatly affect the behavior of the rational programmer (and real
+programmers, too).
+
+One way to reconcile these differences is to {\em equip the rational
+programmer with modes \/} that represent the different types of
+information the rational programmer takes into account when debugging a
+scenario. Intuitively, with the common vocabulary of modes, different
+blame strategies become different ways the rational programmer  interprets
+blame.  For instance, one Transient mode may pick the oldest element of a
+blame set as the next component to add types because it corresponds to the
+earliest point in the program's evaluation that can discover an impedance
+mismatch.  Another mode may opt to treat the sequence as a stack and add
+types to its newest element.  When both modes are equally successful in
+locating an impedance mismatch, measuring the rational programmer's effort
+may answer which mode is the most effective.
+
+However, attributing the success of the rational programmer to this or
+that blame mechanism demands a careful look at the interplay between blame
+and the run-time checks of each gradual type system. When a check fails
+the Natural and Transient semantics assign blame and discard  the
+information in the exception from the run-time check. Nevertheless, the
+exception information may be equally or even more useful to the rational
+programmer than blame. If this is the case blame may not play a critical
+role for the rational programmer. Indeed, precisely because they do not
 account for confounding factors, ~\citet{lksfd-popl-2020} cannot draw any
@@
 
-One way to isolate blame from this confounding factor is to {\em equip the
-rational programmer with modes \/} that represent the different types of
-information the rational programmer takes into account when debugging a
-scenario. Specifically, the rational programmer comes with a blame mode and
-an exception mode. Only if the former succeeds debugging a program while
-the latter fails, it is safe to conclude that blame is indeed necessary
-for the rational programmer. Put differently, the exception mode serves as
-the {\em baseline\/} for blame's value; if the programmer in this mode
-``beats'' the  blame one, a blame assignment mechanism might be useless.
+Modes offer a uniform way to compare the different semantics and  isolate
+blame from the effect of the semantics's runtime checks. Specifically, the Natural and
+Transient rational programmer come with a blame mode and an exception
+mode. Only if the former succeeds debugging a program while the latter
+fails, it is safe to conclude that blame is indeed necessary for the
+rational programmer. Put differently, the exception mode serves as the
+{\em baseline\/} for blame's value within a given semantics; if the
+programmer in this mode ``beats'' the  blame one, a blame assignment
+mechanism might be useless.
 
@@
 of blame. Thus a proper evaluation of blame needs to factor out any random
-effects. A random rational programmer mode provides the necessary baseline
-using the same ingredients as for exceptions.
-
-Modes also offer a solution to a derivative challenge; reconciling the
-differences among the blame assignment mechanisms of the three semantic variants.
-While Natural assigns blame to {\em one\/} component, Transient assigns
-blame to a sequence of components. The Erasure semantics does not blame a
-component {\it per se\/}, but it comes with an exception location and a
-stack trace, which implicitly suggests parties to blame.  The different
-strategies greatly affect the behavior of the rational programmer (and
-real programmers, too).
-
-With modes, different blame strategies map to different ways the rational
-programmer  interprets blame.  For instance, one Transient mode may pick
-the oldest element of a blame set as the next component to add types
-because it corresponds to the earliest point in the program's evaluation
-that can discover an impedance mismatch.  Another mode may opt to treat
-the sequence as a stack and add types to its newest element.  When both
-modes are equally successful in locating an impedance mismatch,
-measuring the rational programmer's effort may answer which mode is the
-most effective.
+effects. A random rational programmer mode provides yet another necessary
+baseline using the same ingredients as for exceptions.
 

### 2021-03-02 writing/why-rational.tex
typo in title
basic changes to why-rational

@@
 Mill's {\it homo economicus\/}, the rational programmer approximates the
-behavior of a software developer who reduces time spent on a task given the
+behavior of a software developer who reduces time spent on a task by exploiting the
 available information. In the context of gradual typing, the rational programmer
@@
 method, searching for the source of the impedance mismatch by adding type
-annotations to the parts of the program identified in the error.
-If the type checker rejects an annotation, the rational programmer has found the problem.
-Measuring this simulated programmer's behavior on a large number of
-debugging scenarios yields data that is similar to data collected in an
-empirical, human-facing manner.
+annotations to the untyped parts of the program identified in the error.
+If the type checker rejects an annotation that is considered correct, the
+rational programmer has found the source of the problem.  Measuring this simulated
+programmer's behavior on a large number of debugging scenarios yields data that
+is similar to data collected in a human-facing study.
 
-The idea is most easily illustrated with an example from the Typed Racket
-world. In the context of Typed Racket, researchers speak of migratory
-typing. Imagine a code base consisting of dozens of modules in plain Racket.  A
-developer who modifies a module in this collection for maintenance purposes must
-study the module designs---reify the types that went into its creation---and to
-help future maintainers should report these insights as type annotations. Over
-time, the code base turns into a mix of typed and untyped modules.
+The idea is best illustrated with an example in Typed Racket's migratory type
+system. Imagine a code base with dozens of modules in plain Racket. A developer
+who opens a module for maintenance purposes must study the module's design and,
+as part of the process, is bound to re-construct the types that went into the
+module's creation.  To help future maintainers, the developer should report
+these insights as type annotations. Over time, the code base migrates into a mix
+of typed and untyped modules. As~\citet{tfffgksst-snapl-2017} reported, though,
+it is at least equally common that developers add typed modules that depend on
+the existing modules in the code base.
 
@@
 annotations. The result is the \typecolor\ module in the middle, and {\tt
-main}'s import is now re-directed there by {\tt (submod ".." layer)}. Running the
-program again yields a new message:
+main}'s import is now re-directed there by {\tt (submod ".." layer-typed)}. Running the
+modified program yields a different error message:
 \begin{verbatim}
@@
 The last step of the rational programmer is to assign types to the {\tt server}
-module. At this point the type checker objects to the conjecture type of {\tt
-neg-abs}---and the source of the impedance mismatch is found. How to fix it is a
-separate question.
+module and to re-direct the import of {\tt layer-typed} to {\tt (submod ".."
+server-typed)}. At this point the type checker objects to the conjectured type
+of {\tt neg-abs}---and the source of the impedance mismatch is found. How to fix
+it is a separate question.
 
-Like {\it homo economicus\/}, the rational programmer is an
-approximation.  People don't behave purely rationally as economic
-actors, and they also don't do so as software developers. The point is
-not to deny the existence of ``lucky hunches'' programmers or
-``tinkering works'' approaches and so on. It is also not to deny that
-equipping entire modules with types represents an always feasible
-approach.  But the concept of studying the idea of an economically
-rational actor has produced benefits to political economics, and this
-paper suggests that implementing and studying the rational programmer
-will help language designers.
+Like {\it homo economicus\/}, the rational programmer is an approximation.
+People do not behave in a purely rational manner as economic actors, and they
+also do not do so as software developers. The point is not to deny the existence
+of ``lucky hunches'' programmers or ``tinkering works'' approaches and so on. It
+is also not to deny that equipping entire modules with types represents an
+always feasible approach (though adding types at the expression level, say, as
+in TypeScript should be considered well within bounds).  {\em But\/}, the
+concept of studying the idea of an economically rational actor has produced
+benefits to the discipline of political economics, and this paper suggests that
+implementing and studying the rational programmer will help language designers.
 
-\subsection{How to Turn the Idea of the Rational Programmer to a Methodology}
+\subsection{How to Turn the Idea of the Rational Programmer into a Methodology} 
+\label{sub:methodology}
 

### 2021-03-02 writing/mutate-benchmarks.tex
restructuring benchmarks and moving earlier the experiment process figure

@@
 
-Since this is the first empirical attempt to evaluate blame in gradual typing,
-there is no standard benchmark suite that we can use off-the-self.  In addition,
-there is no exsitng benchmark suite of gradually typed programs with type-level
-mistakes.\footnote{\citet{cc-oopsla-20} created a collection of Reticulated
+The bar for testing the rational programmer is high: it demands a
+collection of realistic programs that
+\begin{itemize}
+
+  \item vary in size, complexity and purpose;
+
+  \item  have a decent number of modules and a variety of module
+    dependency graphs (since mixing of typed and untyped code in Typed
+    Racket takes place at the module level);
+
+  \item attempt to take  advantage of the sophisticated typing features of
+    a gradually typed language such as Typed Racket but;
+
+   \itme  fail and contain type-level mistakes. 
+\end{itemize}
+
+No such benchmark suite exists.  Worse, no benchmark suite of
+gradually typed programs with type-level mistakes exists at
+all.\footnote{\citet{cc-oopsla-20} created a collection of Reticulated
 Python programs to evaluate their technique of fixing mistakes in type
-annotations. Their collection does not come with type-level mistakes in the code
-itself.} In fact, the bar for testing the rational programmer is much higher
-than the existence of a collection of programs with type-level errors. Testing
-the rational programmer demands a currated collection of programs that take
-advantage of the sophisticated features of a gradually typed language such as
-Typed Racket.
+annotations. Their collection does not come with type-level mistakes
+in the code itself.}
 
-Over the past decade,  \citet{gtnffvf-jfp-2019} have curated such a
+However, over the past decade,  \citet{gtnffvf-jfp-2019} have curated a
 collection of Typed Racket programs for systematically measuring the
@@
 classes and objects~\cite{tsdtf-oopsla-2012}, and types for Racket's
-numeric tower~\cite{stathff-padl-12}. Hence they offer a diverse lignustic
-substrate suitable for injecting type-level errors and testing the
-rational programmer.
+numeric tower~\cite{stathff-padl-12}.
 
-Since mixing of typed and untyped code in Typed Racket takes place at the
-module level, the experiment calls for benchmarks with a decent number of
-modules and a variety of module dependency graphs. Filtering the benchmark
-suite using these criteria while preserving its linguistic diversity
-yields ten suitable programs. Figure~\ref{table:benchmark-descriptions}
-displays the selected benchmark programs and describes their basic
-characteristic.
+Hence, even if they do not come with type-level errors, they offer a
+diverse lignuistic substrate suitable for the careful introduction of
+type-level errors via mutations. Further filtering the benchmark suite to
+keep the benchmarks with the most sizeable and complex dependency graphs
+yields ten suitable programs while preserving the suite's linguistic
+diversity. Figure~\ref{table:benchmark-descriptions} displays the selected
+benchmark programs and describes their basic characteristic.

### 2021-03-02 writing/challenges.tex
a bit more work on challenges
small tweaks to emphasize interesting mutations

@@
 
-However, three major challenges remain:  comparing the effect of
+However, three challenges remain:  comparing the effect of
 blame on the rational programmer across three different mechanisms; the
@@
 of blame. Thus a proper evaluation of blame needs to factor out any random
-effects. A random rational programmer mode provides yet another necessary
+effects too. A random rational programmer mode provides yet another necessary
 baseline using the same ingredients as for exceptions.
 
-The {\em second major challenge\/} is to find a representative collection of
+The {\em second challenge\/} is to find a representative collection of
 programs with impedance mismatches. The latter must represent mistakes that programmers
@@
 
-The {\em third major challenge\/} is the explosive number of debugging scenarios that
+The {\em third challenge\/} is the explosive number of debugging scenarios that
 result from the combination of mutation-based scenario generation and mode-based

### 2021-03-02 writing/why-rational-methodology.tex
methodlogy

%% -----------------------------------------------------------------------------

Every time the rational programmer debugs a program, it is validation for
programming language researchers. It shows how their slogans and theorems help
programmers connect the dots. The example of blame assignment mechanisms makes
this point clearly. A blame-assignment mechanism provides information that,
according to programming language research, points toward the source of the
problem.

By contrast, when the rational programmer fails, it questions programming
language research. Specifically, it is indicative of the limited predictive
power of programming language theory with respect to the use of languages in
practice. Put differently, it suggests flaws in language design due to
misleading predictions.

In this way, the idea of a rational programmer implies an entire methodology for
evaluating the design of programming languages. At this point, this study of
methods supplies many questions whose answers might point to suitable evaluation
methods:

\begin{enumerate}

\item Does a language design provide information that guides the actions of a
 rational programmer?

\item Does the underlying theory suggest actions to the rational programmer?

\item Can this guidance be formulated as an algorithm on the program and the extra
information?

\item Does the underlying theory lead to a hypothesis about the effects of
 these actions?

\item Can this hypothesis be tested with a large-scale automated experiment?

\end{enumerate}

Here is how the answers to these methodological questions imagines an evaluation
method for blame-assignment mechanisms:

\begin{enumerate}

\item The design of blame-assignment mechanisms explicitly advertises the blame
information as helpful for debugging impedance mismatches.

\item The error messages of blame-assignment mechanisms include suspect
violations at the boundary of typed and untyped code fragments. The
Wadler-Findler slogan suggests the problem is a lack of types, meaning the
addition of types to the untyped fragment is the most natural action.

\item The step-by-step construction of paths based on error messages from
gradual-typing checks is clearly amenable to implementation, modulo the
ascription of types to modules.

\item The theory theories implies that blame assignment constrains the search
space that a developer must inspect to find the problem.

\end{enumerate}
Sadly, seeing an algorithm is different from using the algorithm to conduct a
data-gathering experiments. The specific challenges are spelled out in the next
section, and the following three sections explains ways of overcoming them.

### 2021-03-02 writing/challenges.tex
touching up 4.0

@@
 
-At first glance implementing the methodology of
-section~\ref{sec:why-rational} seems straightforward. The rational
-programmer is a simple algorithm and testing whether the rational
-programmer hypothesis holds amounts to running the algorithm on a
-collection of scenarios.
+At first glance implementing the methodology of section~\ref{sec:why-rational}
+seems straightforward. The rational programmer is a simple algorithm and testing
+whether the rational programmer hypothesis holds amounts to running the
+algorithm on a collection of scenarios.
 
-However, three challenges remain:  comparing the effect of
-blame on the rational programmer across three different mechanisms; the
-collection of a large number of representative debugging scenarios;  and a
-proper sampling strategy for the huge space of possibilities. A
-coincidental challenge is the disparity of the implementations of
-gradually typed languages. To eliminate this variable from the comparison,
-this paper uses Racket, which is thus far the only language in which all
-three major soundness mechanisms are available in a reasonably robust and
-directly comparable manner; for the implementation of the transient
-semantics, see the parallel submission of an experience
-report~\cite{ttt21}.
+Three challenges remain, however: comparing the effect of blame on the rational
+programmer across three different mechanisms; the collection of a large number
+of representative debugging scenarios; and a proper sampling strategy for the
+huge space of possibilities. A coincidental challenge is the disparity of the
+implementations of gradually typed languages. To eliminate this variable, this
+authors use Racket, which is thus far the only language in which all three major
+soundness mechanisms are available in a reasonably robust and comparable manner;
+for the implementation of the transient semantics, see the parallel submission
+of an experience report~\cite{ttt21}.
 
 The {\em first challenge\/} stems from the differences between the blame
-assignment mechanisms of the three semantic variants.  While Natural
-assigns blame to {\em one\/} component, Transient assigns blame to a
-sequence of components. The Erasure semantics does not blame a component
-{\it per se\/}, but it comes with an exception location and a stack trace,
-which implicitly suggests parties to blame.  The different strategies
-greatly affect the behavior of the rational programmer (and real
-programmers, too).
+assignment mechanisms of the three semantic variants.  While Natural assigns
+blame to {\em one\/} component, Transient assigns blame to a sequence of
+components. The Erasure semantics does not blame components {\it per se\/}. But,
+it comes with an exception location and a stack trace, which implicitly suggests
+parties to blame.  Each strategy affects the behavior of the rational programmer
+differently (and real ones, too).
 

### 2021-03-02 writing/results.tex
adjusting figures in section 8
results: move figure explanations from caption into figure

@@
 
-\begin{wrapfigure}{r}{.40\textwidth}
-  \centering
+\begin{figure}
   \includegraphics[width=0.40\textwidth]{./plots/success-bars}
-  \caption{Success baselines: The estimated percentage of scenarios for which each mode succeeds in locating the bug.
+
+  \vspace{1em}
+  \begin{minipage}{0.95\textwidth}
+  The estimated percentage of scenarios for which each mode succeeds in locating the bug.
   The upper bound margin of error is 0.02\%.
-  }
+  \end{minipage}
+
+  \caption{Success rates.}
   \label{fig:success-bars}
-\end{wrapfigure}
+\end{figure}
 
@@
   \includegraphics[width=\textwidth]{./plots/avo-bars}
-  \caption{Usefulness comparisons: Each plot depicts a head-to-head comparison of the entitled mode against every other mode.
+
+  \vspace{1em}
+  \begin{minipage}{0.95\textwidth}
+  Each plot depicts a head-to-head comparison of the entitled mode against every other mode.
   The portion above 0 (in green) is the estimated percentage of scenarios where the entitled mode is more useful than each of the other modes.
@@
   The upper bound margin of error is 0.02\%.
-  }
+  \end{minipage}
+
+  \caption{Usefulness comparisons.}
   \label{fig:avo-bars}
@@
 
-  \caption{Blame usefulness analysis: Each diagram shows the breakdown of how three modes' successful scenarios overlap.
+  \vspace{1em}
+  \begin{minipage}{0.95\textwidth}
+  Each diagram shows the breakdown of how three modes' successful scenarios overlap.
   The percentage of scenarios covered by each circle are proportions for which the corresponding mode succeeds.
@@
   The upper bound margin of error is 0.02\%.
-  }
+  \end{minipage}
+
+  \caption{Blame usefulness analysis.}
   \label{fig:success-venns}
@@
   \includegraphics[width=\textwidth]{./plots/bt-lengths-table}
-  \caption{Programmer effort: Each plot depicts the distribution of trail
-  lengths for a given mode across all benchmarks.}
+
+  \vspace{1em}
+  \begin{minipage}{0.95\textwidth}
+  Each plot depicts the distribution of trail lengths for a given mode across all benchmarks.
+  \end{minipage}
+
+  \caption{Programmer effort.}
   \label{fig:effort-table}
@@
   \includegraphics[width=\textwidth]{./plots/bt-length-comparisons}
-  \caption{Effort comparisons: Each plot depicts the distribution of scenarios with trail length differences ranging from -3 to 3.
+
+  \vspace{1em}
+  \begin{minipage}{0.95\textwidth}
+  Each plot depicts the distribution of scenarios with trail length differences ranging from -3 to 3.
   A negative difference means that the first mode's trail is three steps shorter than the second mode's trail for the same scenario, and a positive difference is the reverse.
   Only scenarios where both modes succeed are considered.
-  The 10|85 on the y-axis indicates that the axis is truncated between 10 and 85\%.}
+  The 10|85 on the y-axis indicates that the axis is truncated between 10 and 85\%.
+  \end{minipage}
+
+  \caption{Effort comparisons.}
   \label{fig:effort-comparisons}

### 2021-03-03 writing/discussion.tex
claeify language of threats, push experiment

@@
 
+%% -----------------------------------------------------------------------------
 \subsection{Threat: Is the Rational Programmer Realistic?}
@@
 
-\subsection{Threat: Is our Transient Correct?}
-\label{sec:threat:transient}
+%% -----------------------------------------------------------------------------
+\subsection{Threat: Can the Experiment Improve Transient?} \label{sec:threat:transient}
 
-As we discuss in section~\ref{sec:results}, Transient surprisingly
-produces
-an empty blame set in 967 scenarios.
-Such cases should never occur, at least in theory, because an empty set
- means the value has never crossed a boundary---if the value is indeed defined
- in the current module, then we have typed code blaming itself for a typed
- value as all the checks are inlined in typed code. Thus, the type checker must be unsound.
-After careful investigation of these empty blame cases, we found no soundness
- bugs in Transient Typed Racket.
-Instead, we found scenarios in which Transient lost track of the proper
- boundaries.
-These scenarios suggest ways to improve~\citet{vss-popl-2017}'s blame
-algorithm:
+The results in section~\ref{sec:results} show that Transient produces empty
+blame sequences for 967 scenarios. In theory, empty sequence must not occur. An
+empty blame sequence means a lack of boundary crossings for the witness value.
+By implication, a typed module is blaming itself---something that can
+happen only if the type checker (or system) is unsound.
+
+An investigation of these empty blame cases reveals that Vitousek et al.'s
+Transient blame assignment can lose track of the proper boundaries.  Despite
+these losses, the number of failures is a small fraction of the overall number
+of debugging scenarios. It is unlikely that any improvements to Vitousek et
+al.'s strategy would change the overall conclusion about Transient blame.
+
+{\bf Note} These scenarios suggest ways to improve Transient's blame map as
+follows:
 \begin{itemize}
-  \item
-    Entries in a blame map must point to several parent entries.
-    For example, if the function \texttt{f} receives bad input in the call
-    \texttt{(filter f xs)}, then blame should point to both \texttt{filter}
-    and the \texttt{xs} list.
-  \item
-    The construction of blame map entries must be guided by type-like specifications
-     instead of relying on syntax.
-    Aliasing the built-in \texttt{filter} function should not change the shape
-     of the blame map.
-  \item
-    The initial blame map must reflect the initial type environment.
-    For instance, Typed Racket trusts that untyped core-library functions, such as \texttt{filter},
-     behave correctly; such assumptions should appear in the blame map.
+
+\item Entries in a blame map should point to {\em several\/} parent entries.
+ For example, if \texttt{f} receives bad input in the context of {\tt (filter f
+ xs)}, the entry should point to both \texttt{filter} and \texttt{xs}.
+
+\item The construction of entries must be guided by type-like specifications for
+ primitives instead of special cases for built-in functions. Aliasing such
+ function changes the blame map.
+
+%% I don't see what this really means or contributes to the first two bullets:
+
+ % \item The initial blame map must reflect the initial type environment.  For
+ % instance, Typed Racket trusts that untyped core-library functions behave
+ % correctly.
+
 \end{itemize}
-\noindent{}Despite these known limitations, based on the relatively small
-number of debugging failures, we conjecture that these improvements
- to Transient blame-tracking do not affect our overall conclusions.
+This small example is indicative of how rational-programmer experiments may
+provide feedback.
 
-A second apparent flaw in our Transient implementation is the high cost of blame.
-Whereas \citet{vss-popl-2017} report an average slowdown of 2.5x and
-worst-case of 5.4x on fully-typed benchmarks in Reticulated,
-some of our configurations exceed our 4 minute timeout or 6GB memory limit with transient blame.
-To put those limits into context, the fully typed and fully untyped benchmarks all complete in a few seconds with minimal memory usage, and none of the mixed Natural configurations hit these limits.
-Unfortunately, our high cost for Transient seems closer to the truth;
- there are at least three broad issues that skew the earlier results.
-First, the 2017 implementation of Reticulated fails to insert certain
- soundness checks\footnote{Missing check: \url{https://github.com/mvitousek/reticulated/issues/36}}
- and blame-map updates\footnote{Missing cast: \url{https://github.com/mvitousek/reticulated/issues/43}}
- from the paper.
-Second, Reticulated's type checker assigns the dynamic type to local
- variables in the "fully typed" benchmarks---often because the type system
- cannot articulate a more precise type.
-Code that ends up with the dynamic type has fewer constraints to check at run-time.
-Third, \citet{vss-popl-2017} use relatively small benchmarks.
-Four have since been retired from the official Python benchmark suite
- because they are too small, unrealistic, and unstable.\footnote{Release notes: \url{https://pyperformance.readthedocs.io/changelog.html}}
-On larger programs, Reticulated suffers from high overhead.
-For example, a Reticulated version of our simplest benchmark runs in
- ~40 seconds without blame, and with blame times out after 10 minutes.
+%% -----------------------------------------------------------------------------
+\subsection{Threat: Is the Transient Blame Assignment Mechanism Realistic?}
+\label{sec:threat:transient2}
+
+The results in section~\ref{sec:results} also show that the cost of Transient
+blame is quite high. Under the Transient semantics, some of the debugging
+scenarios exceed the 4-minute timeout or the 6GB-memory limit. To put those
+limits into context, the fully typed and fully untyped benchmarks all complete
+in a few seconds with minimal memory usage. Furthermore, none of the mixed
+Natural configurations hit these limits, and with the blame map turned off,
+the Transient semantics also runs these programs in a short amount of time and
+well within the memory limits. In short, even though the Transient rational
+programmer appears to do well in the experiment, the implementation of the
+Transient blame strategy might be unrealistic.
+
+At first glance, these measurements seem to contradict the results
+of~\citet{vss-popl-2017}. They report an average slowdown of 2.5x and a
+worst-case of 5.4x on fully-typed benchmarks in Reticulated Python.
+Unfortunately, the high cost of Transient Racket seems closer to the truth.
+There are at least four broad issues that skew the earlier results:
+\begin{enumerate}
+
+\item The 2017 implementation of Reticulated fails to insert certain soundness
+checks\footnote{Missing check:
+\url{https://github.com/mvitousek/reticulated/issues/36}} and blame-map
+updates\footnote{Missing cast:
+\url{https://github.com/mvitousek/reticulated/issues/43}} from the paper.
+
+\item The types of almost all local variables in the "fully typed" benchmarks is
+{\tt Dynamic}, often due the impoverished nature of the type system, which does
+not allow the ascription of precise types.  Code that ends up with the type {\tt
+Dynamic} has fewer constraints to check at run-time---and much less information
+to track in the blame map.
+
+\item \citet{vss-popl-2017} use small benchmarks.  Four have since been retired
+from the official Python benchmark suite because they are too small,
+unrealistic, and unstable.\footnote{Release notes:
+\url{https://pyperformance.readthedocs.io/changelog.html}}
+
+\item On large programs, Reticulated suffers from high overhead.  For example, a
+Reticulated version of the simplest benchmark runs in ~40 seconds without blame,
+and with blame times out after 10 minutes. Typed Racket runs the same benchmark
+in a small fraction of the time.
+\end{enumerate}
+Future work on Transient blame is needed to make an informed decision. The
+evaluation process presented here can help guide this research.
 

### 2021-03-03 writing/related.tex
organize sections/titles/labels in paper
related

--- 2020-11-16 writing/related.tex
@@
-\section{Related Work}
+%% -----------------------------------------------------------------------------
 
-This paper builds on~\citet{lksfd-popl-2020}'s one-of-a-kind analysis of
-blame for higher-order contracts. They do not spell out the notion of the
-rational programmer but the basic ingredients of our technique derive
-directly from theirs. Specifically, we adapt their empirical technique
-to gradual typing which requires the development of custom mutators,
-careful selection of debugging scenarios and generalization of the
-technique to allow the comparison of different semantics.
+\citet{lksfd-popl-2020}'s human-out-of-the-loop empirical analysis of blame for
+higher-order contracts in the inspiration for this paper.  While they do not
+spell out the notion of the rational programmer, they present the basic
+ingredients of the technique.
 
-The literature on gradual typing presents many semantics beyond the three
-that we study, but only two additional blame strategies.
-Pyret (\shorturl{https://www.}{pyret.org}) treats fixed-size data types same as Natural
-and functions same as Transient. The Forgetful~\cite{cl-icfp-2017} and
-the Amnesic~\cite{gfd-oopsla-2019} semantics are similar to Transient but
-use wrappers instead of inlined checks.  Nom~\cite{mt-oopsla-2017} and
-other \emph{concrete\/} semantics~\cite{wnlov-popl-2010, rsfbv-popl-2015,
+Technically, this work and~\citet{lksfd-popl-2020}'s differ in many ways. The
+experiment presented here concerns three different modes of gradual typing,
+theirs a single notion of contract checking.  It also contributes the idea of
+creating three comparable rational programmers (with several modes). The
+type-level mutators represent another technical contribution of the work
+presented here; as explained in section~\ref{sec:mutate}, almost none
+of~\citet{lksfd-popl-2020}'s mutators is useful at the level of type mistakes.
+
+The literature on gradual typing presents many semantics beyond the three used
+here and only two additional blame strategies.  Pyret
+(\shorturl{https://www.}{pyret.org}) assigns fixed-size data types the Natural
+semantics and functions a Transient semantics. The Forgetful~\cite{cl-icfp-2017}
+and the Amnesic~\cite{gfd-oopsla-2019} semantics are similar to Transient but
+use wrappers instead of in-lined checks.  Nom~\cite{mt-oopsla-2017} and other
+\emph{concrete\/} semantics~\cite{wnlov-popl-2010, rsfbv-popl-2015,
 rzv-ecoop-2015, rat-oopsla-2017} assume that every value comes with a type tag
 and use tag checks to supervise the interactions between typed and untyped code.
-Monotonic
-references~\cite{svctg-esop-2015} and the semantics~\cite{tlt-popl-2019,
-etg-icfp-19, tt-scp-20, tgt-popl-18, tt-sas-17} derived with the
-Abstracting Gradual Typing
-technique~\cite{gct-popl-2016} are (optimized) variants of Natural.
-Among these representative semantics, only Amnesic and Nom present innovative
-blame strategies.
-Our study nevertheless excludes these strategies because Amnesic is a mere
-theoretical construction and Nom imposes restrictions on untyped
-code that would necessitate new data definitions across our benchmarks.
+Monotonic references~\cite{svctg-esop-2015} and the
+semantics~\cite{tlt-popl-2019, etg-icfp-19, tt-scp-20, tgt-popl-18, tt-sas-17}
+derived with the Abstracting Gradual Typing technique~\cite{gct-popl-2016} are
+variants of Natural.  Among these representative semantics, only Amnesic and Nom
+present innovative blame strategies.  The experiment excludes these strategies
+because they seem to be merely theoretical constructions and/or impose severe
+restrictions on programmers.

### 2021-03-03 writing/conclusion.tex
organize sections/titles/labels in paper
results, discussion, conclusion: minor rewordings and fixes
Only change in conclusion is the wording "Problem is" that two of our PLDI
reviewers complained about.
conclusion touched up

--- 2020-11-19 writing/conclusion.tex
@@
 %% -----------------------------------------------------------------------------
-\section{What to Do Next}
 
-The paper provides the first insight into the measurable value of blame
- in the gradually typed world. In an implicit manner researchers and
-language designers have answered this question one way or another, without any
-evidence. Now they may use the evaluation method of this paper to
-confirm or reject their answers. We already know from~\citet{tgpk-dls-2018}
-that, by intuition, programmers prefer the run-time checking of natural over
-other soundness methods. But, just because a random set of programmers expresses
-this view does not mean that blame assignment adds value. At this point we know
-that it does, in some circumstances, and that in others, the answer remains
-murky.
+The paper provides the first insight into the measurable value of blame in the
+gradually typed world. In an implicit manner researchers and language designers
+have answered this question one way or another, but without evidence. Now they
+may use the evaluation method of this paper to rationalize their answers.  The
+work of~\citet{tgpk-dls-2018} suggests that, by intuition, programmers prefer
+the run-time checking of Natural over other soundness methods. But, just because
+a random set of programmers expresses this view does not mean that blame
+assignment adds value. At this point, it is clear that, in some circumstances,
+it does and in others, the answer remains murky.
 
@@
 sta-nt-base-types, wmwz-ecoop-2017} and that has recently received theoretical
-attention~\cite{gfd-oopsla-2019, cc-oopsla-20}: mistakes in type annotations themselves.
-Developers use gradual typing to move an untyped code base into the typed realm,
-and to this end, they need typed APIs for the vast repositories of
+attention~\cite{gfd-oopsla-2019, cc-oopsla-20}: mistakes in type annotations
+themselves.  Developers use gradual typing to move an untyped code base into the
+typed realm, and to this end, they need typed APIs for the vast repositories of
 already-existing libraries. Because they can, language implementors merely
@@
 type annotations---instead of converting the libraries themselves. With those
-facades, the compiler can easily type-check typed components. Problem is
-that these retroactive additions of types to a library may result from a misunderstanding of the
-code. \emph{They may thus be a mistake themselves.}
+facades, the compiler can easily type-check typed components. The problem is
+that these retroactive additions of types to a library may result from a
+misunderstanding of the code. \emph{The ascribed types may thus be a mistake
+themselves.}
 
-The cited evidence suggests that this scenario is quite common and largely unadressed.
-%, and that a type soundness check does {\em not\/} at all address the problem.  
-Hence, a future evaluation must develop type-level mutators that produce incorrect type
-annotations without breaking the code itself. Some preliminary work suggests
-that such type mutators are much more difficult to develop than the code
-mutators of this paper. We conjecture that the natural semantics will
-usually discover such mistakes, while the transient and erasure semantics cannot
-help with such mistakes at all; indeed, we expect that these latter two raise
-misleading exceptions or produce plainly-wrong answers.  Of course, only further
-evaluation work can confirm or reject these conjectures.
+The cited evidence suggests that this scenario is quite common and largely
+unadressed.  A future evaluation must develop mutators that produce incorrect
+type annotations without breaking the code itself. Some preliminary work
+suggests that such type mutators are even more difficult to develop than the
+type-level code mutators presented here. Based on the gradual typing literature,
+the Natural semantics should usually discover such mistakes. In contrast, the
+Transient and Erasure semantics cannot help with such mistakes at all; indeed,
+we expect that these latter two raise misleading exceptions or produce
+plainly-wrong answers.  Of course, only further evaluation work can confirm or
+reject these conjectures.

### 2021-03-03 writing/mutate-interesting.tex
adding new para abput itneresting scenarios

@@
 
-\item If the evaluation od a mutant module immediately raises an exception because
+\item If the evaluation of a mutant module immediately raises an exception because
 of the changes, there is no work for the rational programmer. Indeed, if the
@@
 features.
+
+The definition of interesting scenarios is more than a filter for
+uninteresting mutants. It is also the guide for re-designing and refining
+mutators until they collectively produce a sizable and diverse population of
+interesting scenarios. For example, the {\tt wrap-cond} mutator casts the
+tests of conditional expressions to the {\tt Any} type and, hence,
+suppresses occurrence typing. However, an examination of the scenarios it
+produces through the lens of the definition of interesting scenarios
+reveals that none is interesting. The fully typed mutant does fails to
+typecheck but its variant without type annotations never raises a run-time
+error. In fact, it turns out 1/4 of the mutators originally designed for
+the experiment fail to live up to the standard of the definition while
+many of the rest of the mutators went thought rounds of tweaking
+to increase their potency.

### 2021-03-03 writing/sample.tex
adding total mutant and scenrio numbers in section 7

@@
 
-To make the experiment computationally feasible and statistically precise,
+To make the experiment computationally feasible and statistically sound,
 stratified uniform random sampling with three levels of stratification becomes
 necessary.  The first two levels group mutants first by benchmark and then by
-mutator.  The set of randomly select from the mutants of each benchmark is
+mutator.  The set of the randomly selected mutants of each benchmark is
 checked to make sure that, per benchmark, the sample reflects the diversity of
@@
 evenly-distributed across all of the mutators that contribute mutants for the
-benchmark.  At the third level, the sampling process draws 96 random interesting
-scenarios from each configuration lattice with replacement.
-
-%% This looks like a repeat of the above, made concrete for the benchmark set
-
-% In summary, the sampling process takes advantage of ancillary information to
-% deliver a sufficiently large representative set of scenarios.  Concretely, we
-% test each mode of the rational programmer on 72,192 randomly selected
-% interesting debugging scenarios spread across both benchmarks and mutators.
-% Furthermore, stratified random sampling allows us to derive precise conclusions
-% about the proportion of the whole population that satisfies a given trait from
-% the sample despite the population's heterogeneity.  That property is exactly
-% what we need to project comparisons about the usefulness of various modes of the
-% rational programmer from the sample to the set of all interesting debugging
-% scenarios.
-
-%% we use the following algorithm to select the 80 mutants evenly distributed across all of the mutators that do produce mutants of a benchmark:
-%% 1. Treat each mutator as a bucket containing the mutants produced by that mutator
-%% 2. Let the target-sample-size be 80
-%% 3. Let the bucket-sample-size be target-sample-size/number-of-buckets
-%% 4. For each bucket, do
-%%    1. Randomly remove min(bucket-sample-size, mutants-in(bucket)) mutants from the bucket
-%%    2. Set target-sample-size = target-sample-size - (the number of mutants removed)
-%% 5. If target-sample-size is empty, done; else, go back to 3.
+benchmark. Some benchmarks has less than  80 mutants with interesting scenarios
+in which case the process collects them all which results in a total of 756 mutants
+across all benchmarks.  
+At the third level, the sampling process draws 96 random interesting
+scenarios from each configuration lattice with replacement. In sum, the
+final sample consists of 72,200 interesting scenarios.

### 2021-03-03 writing/rational-natural.tex
removing wes

@@
 
-To account for this situation, we devise a new mode of the Natural rational
-programmer that follows a migration process based on exceptions:
+To account for this situation, a new mode of the Natural rational
+programmer follows a migration process based on exceptions:
 \begin{quote}
@@
 
-Same as for success for a Natural blame trail, we define success for a
-Natural exception trail.
-Putting the definitions for the two modes together, we can now compare the usefulness of blame 
+The definition of success for a Natural exception trail follows that for
+a Natural blame trail.
+Together the definitions for the two modes allow the comparison of the usefulness of blame 
 with that for mere exceptions for debugging a scenario in the context of Natural semantics.

### 2021-03-03 writing/results.tex
removing wes

@@
 
-We run the experimental process from section~\ref{sec:rational} on the
-sampled debugging scenarios from section~\ref{sec:sample} using a machine with two Intel Xeon Gold 6258R processors (28 doubly-threaded cores each) and 500GB of memory.
-Each debugging scenario had a 4 minute timeout and a
-memory limit of 6GB. The complete experiment required
-over 30,000 compute hours.
+The experimental process from section~\ref{sec:rational} was run on the
+sampled debugging scenarios from section~\ref{sec:sample} on a machine
+with two Intel Xeon Gold 6258R processors (28 doubly-threaded cores each)
+and 500GB of memory.  Each debugging scenario had a 4 minute timeout and a
+memory limit of 6GB. The complete experiment required over 30,000 compute
+hours.
 
@@
 not come as a surprise to the astute
-reader. After all, as we
-discuss above, some modes are more successful than others in certain
+reader. After all, as
+discussed above, some modes are more successful than others in certain
 scenarios. Indeed any mode can fail if no untyped module from the program is identified in the error.
@@
 without hints about how to proceed.
-In section~\ref{sec:threat:transient}, we discuss how both of these causes of
+Section~\ref{sec:threat:transient} explains how both of these causes of
 failure for Transient affect our experiment.

### 2021-03-03 writing/results.tex
removing wes

@@
 
-While most failures to debug scenarios follow the above pattern, a few do not.
-Breaking down the reasons for failure for Natural blame (1748 in total)
-reveals an additional cause. For a small set of
-debugging scenarios (40), Natural produces a run time type error
-blaming a non-buggy already typed
-component. We tracked down all these cases to known open issues with Typed
+While most failures to debug scenarios follow the above pattern, a few do
+not.  Breaking down the reasons for failure for Natural blame (1748 in
+total) reveals an additional cause. For a small set of debugging scenarios
+(40), Natural produces a run time type error blaming a non-buggy already
+typed component. All these cases are due to known open issues with Typed
 Racket and class contracts.
@@
 
-Figure~\ref{fig:effort-comparisons} provides head-to-head comparisons of effort.
-To directly compare effort between modes, we compute the difference in trail length between two modes for all scenarios where they both succeed.
-Hence, each plot in the figure shows the distribution of scenarios with length differences ranging from -3 (the first mode's trail is 3 steps shorter than the second's) to 3 (the first mode's trail is 3 steps longer than the second's).
-The figure offers several insights about the comparative effort for each mode that complement the insights about comparitive success from figure~\ref{fig:avo-bars}.
-First, Natural blame never produces shorter trails than Natural exceptions, and sometimes (albeit rarely) produces slightly longer ones.
-Second, Natural relatively often (around 10\% of the scenarios) produces shorter trails than both Transient blame modes, and they are sometimes significantly shorter.
-Third, Transient last blame often produces shorter trails than Transient first blame, which is a particularly interesting distinction in light of their nearly indistinguishable success rates.
-Finally, Transient's blame modes also share the characteristic with Natural that blame sometimes produces longer trails than their corresponding exception modes.
+Figure~\ref{fig:effort-comparisons} provides head-to-head comparisons of
+effort.  The effort comparison between modes boils down to the
+difference in trail length between two modes for all scenarios where they
+both succeed.  Hence, each plot in the figure shows the distribution of
+scenarios with length differences ranging from -3 (the first mode's trail
+is 3 steps shorter than the second's) to 3 (the first mode's trail is 3
+steps longer than the second's).  The figure offers several insights about
+the comparative effort for each mode that complement the insights about
+comparitive success from figure~\ref{fig:avo-bars}.  First, Natural blame
+never produces shorter trails than Natural exceptions, and sometimes
+(albeit rarely) produces slightly longer ones.  Second, Natural relatively
+often (around 10\% of the scenarios) produces shorter trails than both
+Transient blame modes, and they are sometimes significantly shorter.
+Third, Transient last blame often produces shorter trails than Transient
+first blame, which is a particularly interesting distinction in light of
+their nearly indistinguishable success rates.  Finally, Transient's blame
+modes also share the characteristic with Natural that blame sometimes
+produces longer trails than their corresponding exception modes.

### 2021-03-03 writing/challenges.tex
intro, landscape, why-rational: rewording
To clarify parts that confused me

@@
 
-At first glance implementing the methodology of section~\ref{sec:why-rational}
+At first glance implementing the method of section~\ref{sec:why-rational}
 seems straightforward. The rational programmer is a simple algorithm and testing
@@
 original mutators systematically produced programs that immediately revealed the
-source of the impedance mismatch. About two thirds of the original mutators
+source of the impedance mismatch. About four fifths of the original mutators
 survived this first scrutiny testing. All of the remaining ones yield {\em

### 2021-03-03 writing/introduction.tex
intro, landscape, why-rational: rewording
To clarify parts that confused me

@@
 with sophisticated checking and blame assignment strategies (see sec.~\ref{sec:landscape}). Their academic creators embrace
-the idea that blame can help practicing programmers find impedance mismatches
-between the types they added to a software system and the behavior of the
-remaining untyped components.
+the idea that blame can help practicing programmers find impedance mismatches: disagreements
+between the types they added to a software system and its behavior.
 
@@
 
-This contrast between theory and applications of gradual typing raises the question 
+This contrast between theory and applications of gradual typing raises the question of
 \begin{quote}
@@
 assignment methods clearly preferable to {\it Erasure\/}.  In turn, these
-problems suggests that, on one hand, the existing theory does not predict
-practice properly, and on the other hand, the existing practice may need to
-think of additional experiments to collect more data.
+problems suggest that, on one hand, the existing theory does not predict
+practice properly, and on the other hand, the existing practice may need
+additional experiments to collect more data.

### 2021-03-03 writing/landscape.tex
intro, landscape, why-rational: rewording
To clarify parts that confused me

@@
 with types at run-time: {\em Natural\/}, {\em Transient\/}, and {\em
-Erasure\/}. This section summarizes these options with one illustrative example
-using Racket syntax. The informed reader may wish to merely scan the
-self-contained section.
+Erasure\/}. This self-contained section summarizes these options with one illustrative example
+using Racket syntax. The informed reader may wish to merely scan it.
 
@@
 module: the top bar lists the name and whether it is using typed (\typecolor) or
-untyped (\dyncolor) syntax:
+untyped (\dyncolor) syntax.
 \begin{description}
@@
 among others, a function {\tt pack}. As the documentation says, this function
-consumes JSON data and packages it in a dictionary. by mistake, though, the
+consumes JSON data and packages it in a dictionary. By mistake, though, the
 function returns a hash table instead of the association list its documentation
@@
 indirectly. Specifically, it goes through the two intermediary typed modules to
-import them. Imagine a programmer who relies on the types in the \typecolor\
+use it. Imagine a programmer who relies on the types in the \typecolor\
 modules as checked documentation but prototypes the client in an untyped manner.
@@
 \end{description}
-Each of the three semantics assigns a different outcome to the client module and
-signals different kinds of errors.
+Due to the mistake in {\tt pack-lib}, each of the three semantics assigns a
+different outcome to the client module and signals different kinds of errors.
 
@@
 for each check. Thus, in the example of figure~\ref{fig:example}, as
-\texttt{pack} returns, the return-type check fails, and Natural blames
+\texttt{pack} returns, the return-type check fails and Natural blames
 \texttt{pack-lib}.
@@
 In order to locate the responsible party for failed checks, Transient
-maintains a map from values to the boundaries between typed/untyped modules that
+maintains a map from values to the boundaries between typed and untyped modules that
 they cross, plus the corresponding types. In the example, the map records that
@@
 underlying untyped language. Hence, neither the call to {\tt typed-pack} nor the
-call to {\tt crypto-pack} signals an error due to the gradual type system. When,
-at some later point, {\tt client} tries to inspect the elements of the lists
+call to {\tt crypto-pack} signals an error due to the gradual type system. If
+at some later point {\tt client} tries to inspect the elements of the lists
 that \texttt{typed-pack} and {\tt crypto-pack} are supposed to produce, Racket's
-safety check signal a violation and point to some place in {\tt client}. The
+safety checks signal a violation and point to some place in {\tt client}. The
 information in this exception, plus its stack trace, may help the programmer

### 2021-03-03 writing/why-rational-methodology.tex
intro, landscape, why-rational: rewording
To clarify parts that confused me

@@
 power of programming language theory with respect to the use of languages in
-practice. Put differently, it suggests flaws in language design due to
+practice. Put differently, it suggests flaws in language design revealed by
 misleading predictions.
@@
 
-\item Does a language design provide information that guides the actions of a
+\item Does a language design provide information that guides the actions of the
  rational programmer?
@@
 \item The error messages of blame-assignment mechanisms include suspect
-violations at the boundary of typed and untyped code fragments. The
-Wadler-Findler slogan suggests the problem is a lack of types, meaning the
-addition of types to the untyped fragment is the most natural action.
+locations at the boundary of typed and untyped code fragments.
+The Wadler-Findler slogan suggests that the source of the problem is concealed due to a lack of types, so
+adding types to the untyped fragment should lead to the source of the impendance mismatch.
 
@@
 
-\item The theory theories implies that blame assignment constrains the search
+\item The theory conjectures that blame assignment constrains the search
 space that a developer must inspect to find the problem.
@@
 \end{enumerate}
-Sadly, seeing an algorithm is different from using the algorithm to conduct a
+Unfortunately, seeing an algorithm is different from using the algorithm to conduct
 data-gathering experiments. The specific challenges are spelled out in the next

### 2021-03-03 writing/why-rational.tex
methodlogy
intro, landscape, why-rational: rewording
To clarify parts that confused me

@@
 rational procedure is to use the former to improve the latter. Specifically,
-a rational programmer translates the Wadler-Findler slogan into a debugging
+the rational programmer translates the Wadler-Findler slogan into a debugging
 method, searching for the source of the impedance mismatch by adding type
@@
 If the type checker rejects an annotation that is considered correct, the
-rational programmer has found the source of the problem.  Measuring this simulated
+rational programmer has found the source of the problem.
+Otherwise, the rational programmer concludes that the just-annotated parts are not the problem
+and re-runs the program;
+the Wadler-Findler slogan says the new version of the program must now blame a different location for the problem, which the rational programmer can check in the same fashion.
+Measuring this simulated
 programmer's behavior on a large number of debugging scenarios yields data that
@@
 of ``lucky hunches'' programmers or ``tinkering works'' approaches and so on. It
-is also not to deny that equipping entire modules with types represents an
+is also not to claim that equipping entire modules with types represents an
 always feasible approach (though adding types at the expression level, say, as
@@
 
+%% -----------------------------------------------------------------------------
 \subsection{How to Turn the Idea of the Rational Programmer into a Methodology} 
@@
 
-\input{rational-methodology.tex}
+\input{why-rational-methodology.tex}

### 2021-03-03 writing/results.tex
pass over section 8

@@
   Each plot depicts a head-to-head comparison of the entitled mode against every other mode.
-  The portion above 0 (in green) is the estimated percentage of scenarios where the entitled mode is more useful than each of the other modes.
-  The portion below 0 (in red) is the estimated percentage of scenarios with the inverse situation.
+  The portion above 0 (in green) is the estimated percentage of scenarios where the entitled mode is more useful
+    than the other mode.
+  The portion below 0 (in red) is the estimated percentage of scenarios
+    where the entitled mode is less useful than the other mode.
   The upper bound margin of error is 0.02\%.
@@
   \begin{minipage}{0.95\textwidth}
-  Each diagram shows the breakdown of how three modes' successful scenarios overlap.
-  The percentage of scenarios covered by each circle are proportions for which the corresponding mode succeeds.
-  For example, in the breakdown of Natural, all three modes succeed on the same scenario 57.3\% of the time, only Natural and Natural exceptions succeed on 29.1\% of the scenarios, only Natural and Erasure succeed on 2.1\%, and Natural alone succeeds on 9\%.
-  The upper bound margin of error is 0.02\%.
+  Each diagram shows the overlap of the successful scenarios for three modes.  
+     For example, in the leftmost diagram, all three modes succeed on the
+    same scenario 57.3\% of the time, only Natural and Natural exceptions
+    succeed on 29.1\% of the scenarios, only Natural and Erasure succeed
+    on 2.1\%, and Natural alone succeeds on 9\%.  The upper bound margin
+    of error is 0.02\%.
   \end{minipage}
@@
 
-Figure~\ref{fig:success-bars} summarizes the overall success rates of every mode.
-The success rates illustrate a few points that underly the rest of our analysis.
-The first notable piece of information from
-this figure is that every mode has failed debugging scenarios, not just
-Erasure. This should
-not come as a surprise to the astute
-reader. After all, as
-discussed above, some modes are more successful than others in certain
-scenarios. Indeed any mode can fail if no untyped module from the program is identified in the error.
-For example, running a scenario may result in an
-exception that carries no useful information about which
-module the programmer should type next, e.g. because the stack trace
-does not contain frames from any module of the program, or only typed modules. This situation
-can happen at any point along a blame trail (figure~\ref{fig:effort-table} dives into this below), causing it to fail.
+Figure~\ref{fig:success-bars} summarizes the overall success rates of
+every mode.  The success rates illustrate a few points that underlie the
+rest of our analysis.  The first notable piece of information from this
+figure is that every mode has failed debugging scenarios, not just
+Erasure. This should not come as a surprise to the astute reader.  Running
+a mode of the rational programmer on a scenario may result in an exception
+that carries no useful information about which module the programmer
+should type next. For instance the stack trace of the exception may not
+contain frames from any module of the program but only from the internals
+of the Typed Racket implementation. This situation can happen at any point
+along a blame trail, causing it to fail.
 
@@
 
-In Transient, similar to Natural,
-most failures are due to unhelpful exception information (1851 for both
-Transient first and last blame).  
-However, Transient also has a substantial
-number of failures because scenarios hit the time and memory
-limits of our experiment (\textasciitilde770 scenarios).  Additionally, there are nearly 1,000 cases where
-Transient reports an empty blame set, leaving the rational programmer
-without hints about how to proceed.
-Section~\ref{sec:threat:transient} explains how both of these causes of
-failure for Transient affect our experiment.
+In Transient, similar to Natural, most failures are due to unhelpful
+exception information (1851 for both Transient first and last blame).
+However, Transient also has a substantial number of failures because
+scenarios hit the time and memory limits of our experiment
+(\textasciitilde770 scenarios).  Additionally, there are nearly 1,000
+cases where Transient reports an empty blame set, leaving the rational
+programmer without hints about how to proceed.
+Sections~\label{sec:threat:transient} and ~\label{sec:threat:transient2}
+explain how both of these causes of failure for Transient affect our
+experiment.
 
-The second key observation from figure~\ref{fig:success-bars} is that the modes that use blame all outperform those that do not.
-In particular, Natural and both of Transient's blame modes succeed in around 95\% of our scenarios, while their corresponding exception modes succeed in around 85\% of them and Erasure succeeds in only 60\% of them.
-The only exception is that the Random programmer always succeeds;
-this reflects the facts that every scenario has finitely many modules and this mode can never get stuck, so it always eventually types the buggy module.
+The second key observation from figure~\ref{fig:success-bars} is that the
+modes that use blame all outperform those that do not.  In particular,
+Natural and both of Transient's blame modes succeed in around 95\% of the
+scenarios, while their corresponding exception modes succeed in around
+85\% of them and Erasure succeeds in only 60\%.  The only
+exception is that the Random programmer always succeeds; this reflects the
+fact that every scenario has finitely many modules, so Random programmer 
+eventually types the buggy module.
 
-Figure~\ref{fig:avo-bars} depicts a head-to-head comparison of every mode's performance against every other mode (except Random), helping to answer the four questions from section~\ref{sub:experiment}.
-Each plot shows a breakdown of the proportion of scenarios where one mode performs better or worse than each other mode.
-In particular, each bar above zero represents the proportion where the plot's entitled mode succeeds and the mode on the x-axis fails; the corresponding bar below zero represents the proportion of the opposite case.
-For example, the plot titled ``Natural'' shows that Natural outperforms Natural exceptions in about 12\% of the scenarios, and the inverse (Natural performs worse than Natural exceptions) never happens.
-Similarly, the plot titled ``Transient last blame'' shows that Transient last blame outperforms Natural exceptions about in about 12\% of the scenarios, but conversely it performs worse than Natural exceptions in about 2\% of the scenarios.
+Figure~\ref{fig:avo-bars} depicts a head-to-head comparison of every
+mode's performance against every other mode (except Random). The
+comparison  answers the four questions from section~\ref{sub:experiment}. 
+Each plot shows the proportion of scenarios where one mode performs
+better or worse than each other mode.  In particular, each bar above zero
+represents the proportion where the plot's entitled mode succeeds and the
+mode on the x-axis fails; the corresponding bar below zero represents the
+proportion of the inverse case.  For example, the plot titled ``Natural''
+shows that Natural outperforms Natural exceptions in about 12\% of the
+scenarios, and the inverse (Natural performs worse than Natural
+exceptions) never happens.  Similarly, the plot titled ``Transient last
+blame'' shows that Transient last blame outperforms Natural exceptions
+ in about 12\% of the scenarios, but conversely it performs worse
+than Natural exceptions in about 2\% of the scenarios.
 
-Figure~\ref{fig:avo-bars} shows that the answers for questions $Q_1$ to $Q_3$ are all positive.
-In all three semantics, blame modes outperform their corresponding exception mode (by \textasciitilde10\% for each one).
-The Natural exceptions mode is never more useful than Natural blame, and  Transient
-exceptions are more useful than Transient first and Transient last blame
-in a small percentage (about 2\%) of the scenarios.
+The figure answers questions $Q_1$, $Q_2$ and $Q_3$ affirmatively.  
+In all three semantics, blame modes outperform their
+corresponding exception mode by \textasciitilde10\%.  The
+Natural exceptions mode is never more useful than Natural blame, and
+Transient exceptions are more useful than Transient first and Transient
+last blame in a small percentage (about 2\%) of the scenarios.
 
-Figure~\ref{fig:avo-bars} also provides answers to $Q_*$.
+Figure~\ref{fig:avo-bars} also provides answers for $Q_*$.
 Blame for all three semantics is significantly
@@
   \begin{minipage}{0.95\textwidth}
-  Each plot depicts the distribution of scenarios with trail length differences ranging from -3 to 3.
-  A negative difference means that the first mode's trail is three steps shorter than the second mode's trail for the same scenario, and a positive difference is the reverse.
-  Only scenarios where both modes succeed are considered.
-  The 10|85 on the y-axis indicates that the axis is truncated between 10 and 85\%.
+  Each plot depicts the distribution of scenarios with trail length
+    differences ranging from -3 to 3.  A -x difference denotes that
+    the first mode's trail is x steps shorter than the second mode's
+    trail for the same scenario, and a positive difference denotes the
+    inverse.  Only scenarios where both modes succeed are considered.  The
+    10|85 on the y-axis indicates that the axis is truncated between 10
+    and 85\%.
   \end{minipage}
@@
 
-In order to better understand questions $Q_1$ to $Q_3$, figure~\ref{fig:success-venns} depicts a breakdown of the success of each semantics in comparison to Erasure.
-Specifically, the figure shows one Venn diagram per mode of the rational programmer that uses blame.
-Each diagram breaks breaks down the overlap in the success of the blame mode, its corresponding exception mode, and Erasure.
-For example, in the breakdown of Natural, the diagram shows that all three modes succeed on the same scenario 59.5\% of the time, only Natural and Natural exceptions succeed on 25.7\% of the scenarios, only Natural and Erasure succeed on 2.1\%, and Natural alone succeeds on 9.7\%.
-This analysis highlights the success trade-offs each semantics offers against Erasure, with and without blame, to answer questions about choosing to use blame or just exceptions for each semantics.
-For instance, the breakdown of Natural clearly illustrates that if choosing between Natural blame, Natural exceptions, and Erasure, Natural blame is the absolutely most successful: all of the successes of the other two modes are subsets of Natural's successful scenarios.
-On the other hand, Transient's blame modes are similar but not so clear-cut.
+In order to better understand the answers the experiment provides for
+questions $Q_1$ to $Q_3$, figure~\ref{fig:success-venns} depicts an
+analysis of the success of each semantics in comparison to Erasure.
+Specifically, the figure shows one Venn diagram per mode of the rational
+programmer that uses blame.  Each diagram shows the overlap of successful
+for a  the blame mode, its corresponding exception mode, and Erasure.  For
+example, for Natural, the leftmost diagram shows that all three modes
+succeed on 59.5\% of the  scenarios, only Natural and Natural exceptions
+succeed on 25.7\% of the scenarios, only Natural and Erasure succeed on
+2.1\%, and Natural alone succeeds on 9.7\%.  This analysis highlights
+the success trade-offs each semantics offers against Erasure, with and
+without blame. Hence, it answers questions about how to choose to between
+blame or just exceptions for each semantics.  For instance, the analysis
+for Natural clearly illustrates that, when choosing between Natural blame,
+Natural exceptions, and Erasure, Natural blame is the absolutely most
+successful: all of the successes of the other two modes are subsets of
+Natural's successful scenarios.  On the other hand, Transient's blame
+modes fare similarly but the choice is not so clear-cut.
 
-Turning to programmer effort, figure~\ref{fig:effort-table} shows the distribution of blame trail lengths
-for our sample of interesting debugging scenarios. Unlike the usefulness
-comparisons above, these proportions do not generalize to a representation of the full population.
-Still, as we discuss in section~\ref{subsec:effort}, they provide an
-alternative view of the workings of the rational programmer.
+Turning to programmer effort, figure~\ref{fig:effort-table} shows the
+distribution of blame trail lengths for the sampled interesting debugging
+scenarios. Unlike the usefulness comparisons above, these proportions do
+not generalize to a representation of the full population.  Still, as
+section~\ref{subsec:effort} discusses, they provide an alternative view of
+the workings of the rational programmer.
 
-There are
-two immediate take-aways from the figure. First, the effort for successfully
-debugging interesting scenarios (in green) for the random mode of the
-rational programmer is highly spread out, as expected.\footnote{The
-random mode distribution is indistinguishable for all three semantics and thus the figure~\ref{fig:effort-table} 
-only shows one plot.} In contrast, in the other modes, successful effort coalesces at
+There are two immediate take-aways from the figure. First, the effort for
+successfully debugging interesting scenarios (in green) for the random
+mode of the rational programmer is highly spread out, as
+expected.\footnote{The random mode distribution is indistinguishable for
+all three semantics and thus the figure~\ref{fig:effort-table} only shows
+one plot.} In contrast, in the other modes, successful effort coalesces at
 the left side of the plot, meaning that in most cases the programmer needs
@@
 
-A second point of interest is that the exceptions of
-the Erasure semantics either help the rational programmer immediately or 
-the rational programmer fails to debug a scenario altogether (in red).
-This is expected; adding type annotations to an Erasure program doesn't
-change it's behavior, so if the type checker doesn't reject the
-program with the new annotations, the rational programmer is stuck.
-In other words, an
-exception from the runtime has to point to the buggy module with
-the first try. Otherwise the rational programmer types an irrelevant
-module, runs the program again, and the same exception points again to the
-already typed module.
+A second point of interest is that the exceptions of the Erasure semantics
+either help the rational programmer immediately or the rational programmer
+fails to debug a scenario altogether (in red).  This is expected; adding
+type annotations to an Erasure program doesn't change it's behavior, so if
+the type checker doesn't reject the program with the new annotations, the
+rational programmer is stuck.  In other words, an exception from the
+runtime has to point to the buggy module with the first try. Otherwise the
+rational programmer types an irrelevant module, runs the program again,
+and the same exception points again to the already typed module.
 
 Figure~\ref{fig:effort-comparisons} provides head-to-head comparisons of
-effort.  The effort comparison between modes boils down to the
-difference in trail length between two modes for all scenarios where they
-both succeed.  Hence, each plot in the figure shows the distribution of
-scenarios with length differences ranging from -3 (the first mode's trail
-is 3 steps shorter than the second's) to 3 (the first mode's trail is 3
-steps longer than the second's).  The figure offers several insights about
-the comparative effort for each mode that complement the insights about
-comparitive success from figure~\ref{fig:avo-bars}.  First, Natural blame
-never produces shorter trails than Natural exceptions, and sometimes
-(albeit rarely) produces slightly longer ones.  Second, Natural relatively
-often (around 10\% of the scenarios) produces shorter trails than both
-Transient blame modes, and they are sometimes significantly shorter.
-Third, Transient last blame often produces shorter trails than Transient
-first blame, which is a particularly interesting distinction in light of
-their nearly indistinguishable success rates.  Finally, Transient's blame
-modes also share the characteristic with Natural that blame sometimes
-produces longer trails than their corresponding exception modes.
+effort.  The comparison between two modes boils down to the difference in
+length between their trails for all scenarios where they both succeed.
+Hence, each plot in the figure shows the distribution of scenarios with
+length differences ranging from -3 (the first mode's trail is 3 steps
+shorter than the second's) to 3 (the first mode's trail is 3 steps longer
+than the second's).  The figure offers several insights about how  modes
+compare in terms of effort that complement the insights about how they
+compare in terms of success rates from figure~\ref{fig:avo-bars}.  First,
+Natural blame never produces shorter trails than Natural exceptions, and
+sometimes (albeit rarely) produces slightly longer ones.  Hence, the
+experiment provides evidence that blame helps the rational programmer
+debug more scenarios but does not shorten the debugging process compared
+to exceptions. A second insight is that Natural relatively often (around 10\% of the
+scenarios) produces shorter trails than both Transient blame modes, and
+sometimes the trails are significantly shorter.  Third, Transient last
+blame often produces shorter trails than Transient first blame, which is a
+particularly interesting distinction in light of their nearly
+indistinguishable success rates.  Finally, Transient's blame modes also
+share the characteristic with Natural that blame sometimes produces longer
+trails than their corresponding exception modes.

### 2021-03-03 writing/discussion.tex
space war
pass over section 8

@@
 run time type checks have a large positive impact. This is the case
-regardless of whether these checks issue blame or throw plain exceptions.
-Second, in many cases type checks that issue blame are more helpful than
-those that do not. However, our results indicate that blame
-is not critical in a majority of cases, and therefore it is worth further investigation whether
-disabling blame tracking for the sake of performance may be   an
-acceptable trade-off. Third, from the three different blame strategies, the
-Natural approach of Typed Racket fares better than the Transient approach
-of Reticulated Python, but only by a small margin. Since Natural offers complete
-and sound path-based blame while Transient offers incomplete but sound heap-based
-blame~\cite{gfd-oopsla-2019}, our results call for a deeper understanding
-of the relative strengths of the two models for blame.  Fourth, given that
-Transient's sound but shallow run time type checks do not seem to hamper
-debugging, a version of Typed Racket that disables some of its wrappers
-may offer an answer to the performance issues of Natural. 
-Fifth, just as for Typed Racket, the experiment points to possible
-improvements for Reticulated Python. In particular, the experimental 
-data indicates that the usefulness of transient blame does not depend on
-whether  the rational programmer follows first or the last element from a
-blame set, and that using the last element seems to produce shorter trails.
-Hence, given the performance issues of Transient we discuss further on,
-a possible optimization is to limit the size of blame sets
-to only contain the latest entries using timestamps.
+regardless of whether these checks issue blame or throw plain
+exceptions.  Second, in many cases type checks that issue blame are more
+helpful than those that do not. However, the results indicate that blame
+is not critical in a majority of cases, and therefore it is worth further
+investigation whether disabling blame tracking for the sake of performance
+may be   an acceptable trade-off. Third, from the three different blame
+strategies, the Natural approach of Typed Racket fares better than the
+Transient approach of Reticulated Python, but only by a small margin.
+Since Natural offers complete and sound path-based blame while Transient
+offers incomplete but sound heap-based blame~\cite{gfd-oopsla-2019}, the
+results call for a deeper understanding of the relative strengths of the
+two models for blame.  Fourth, given that Transient's sound but shallow
+run time type checks do not seem to hamper debugging, a version of Typed
+Racket that disables some of its wrappers may offer an answer to the
+well-known performance issues of Natural~\cite{gtnffvf-jfp-2019}.  Fifth,
+just as for Typed Racket, the experiment points to possible improvements
+for Reticulated Python. In particular, the experimental data indicates
+that the usefulness of transient blame does not depend on whether  the
+rational programmer follows first or the last element from a blame set,
+and that using the last element seems to produce shorter trails.  Hence,
+given the performance issues of Transient discussed further on, a possible
+optimization is to limit the size of blame sets to only contain the latest
+entries using timestamps.
 
-There are a number of threats to the validity of our conclusions: (i) the
+There are a number of threats to the validity of these conclusions: (i) the
 representativeness of benchmarks; (ii) the relation between mutations and
 real programming mistakes; (iii) the definition of interesting debugging
-scenarios; and (iv) our approach to sampling scenarios. The previous sections
+scenarios; and (iv) the sampling strategy. The previous sections
 have described the various ways we attempt to mitigate these threats;
-we discuss two additional threats in the remainder of this section.
+the remainder of this section discussed three additional threats.
 
@@
 here. When this happens, the predicted benefits of blame assignment may not
-materialize. Indeed, our own personal experience suggests such deviations, and
+materialize. Indeed, the authors's personal experience suggests such deviations, and
 it also suggests that deviating is a mistake. To make a true judgment of the
@@
 
-Relatedly, our setup hides the type choice a rational programmer must make. When the
+Relatedly, the experimental setup hides the type choice a rational programmer must make. When the
 run-time checks signal an impedance mismatch in the real world, a programmer
@@
 many years has driven home this lesson, but fortunately, it has also shown that
-the types are in most cases reasonably canonical.  We therefore conjecture that
+the types are in most cases reasonably canonical.  The authors therefore conjecture that
 a rational programmer would in most cases come up with an equivalent type
-assignment for trail extension as our experimental setup describes.
+assignment in order to extend a trail as the experimental method describes.
 
 %% -----------------------------------------------------------------------------
-\subsection{Threat: Can the Experiment Improve Transient?} \label{sec:threat:transient}
+\subsection{Threat: Why Does Transient Lose Blame?} \label{sec:threat:transient}
 
@@
 
-An investigation of these empty blame cases reveals that Vitousek et al.'s
-Transient blame assignment can lose track of the proper boundaries.  Despite
-these losses, the number of failures is a small fraction of the overall number
-of debugging scenarios. It is unlikely that any improvements to Vitousek et
-al.'s strategy would change the overall conclusion about Transient blame.
+An investigation of these empty blame cases reveals that Transient is not
+unsound but that Vitousek et al.'s Transient blame assignment loses track
+of the proper boundaries due to an incomplete population of the blame map.
+Despite these losses, the number of failures is a small fraction of the
+overall number of debugging scenarios. It is unlikely that any
+improvements to Vitousek et al.'s strategy would change the overall
+conclusion about Transient blame.
 
-{\bf Note} These scenarios suggest ways to improve Transient's blame map as
-follows:
+{\bf Note} The scenarios where Transient returns empty blame 
+suggest ways to improve Transient's blame map as follows:
 \begin{itemize}
@@
 
-%% I don't see what this really means or contributes to the first two bullets:
-
- % \item The initial blame map must reflect the initial type environment.  For
- % instance, Typed Racket trusts that untyped core-library functions behave
- % correctly.
-
 \end{itemize}
 This small example is indicative of how rational-programmer experiments may
-provide feedback.
+provide actionable feedback for language design .
 
@@
 \end{enumerate}
-Future work on Transient blame is needed to make an informed decision. The
-evaluation process presented here can help guide this research.
+More work on Transient blame is needed to make an informed decision.
 

### 2021-03-03 writing/challenges.tex
challenges, start of rational: rewordings

@@
 
-At first glance implementing the method of section~\ref{sec:why-rational}
-seems straightforward. The rational programmer is a simple algorithm and testing
-whether the rational programmer hypothesis holds amounts to running the
-algorithm on a collection of scenarios.
-
-Three challenges remain, however: comparing the effect of blame on the rational
-programmer across three different mechanisms; the collection of a large number
-of representative debugging scenarios; and a proper sampling strategy for the
+While at first glance the method of section~\ref{sec:why-rational} seems straightforward to implement and test,
+three challenges remain. The first is comparing the effect of blame on the rational
+programmer across three different mechanisms; second is the collection of a large number
+of representative debugging scenarios; and third is finding a proper sampling strategy for the
 huge space of possibilities. A coincidental challenge is the disparity of the
-implementations of gradually typed languages. To eliminate this variable, this
+implementations of gradually typed languages. To eliminate this variable, the
 authors use Racket, which is thus far the only language in which all three major
@@
 blame to {\em one\/} component, Transient assigns blame to a sequence of
-components. The Erasure semantics does not blame components {\it per se\/}. But,
+components. The Erasure semantics does not blame components {\it per se\/}, but
 it comes with an exception location and a stack trace, which implicitly suggests
@@
 mismatch.  Another mode may opt to treat the sequence as a stack and add
-types to its newest element.  When both modes are equally successful in
-locating an impedance mismatch, measuring the rational programmer's effort
-may answer which mode is the most effective.
+types to its newest element.  If both modes are equally successful in
+locating an impedance mismatch, measuring the rational programmer's effort with each mode
+may answer which is the most effective.
 
@@
 and the run-time checks of each gradual type system. When a check fails
-the Natural and Transient semantics assign blame and discard  the
+the Natural and Transient semantics assign blame instead of using the
 information in the exception from the run-time check. Nevertheless, the
 exception information may be equally or even more useful to the rational
-programmer than blame. If this is the case blame may not play a critical
+programmer than blame. If this is the case, then blame may not play a critical
 role for the rational programmer. Indeed, precisely because they do not
-account for confounding factors, ~\citet{lksfd-popl-2020} cannot draw any
-conclusions about the pragmatics of blame despite advertisements for the
-opposite. Their experiment can only conclude that so-called blame-shifting
+account for confounding factors, \citet{lksfd-popl-2020} cannot draw any
+conclusions about blame specifically despite advertisements for the
+opposite. Their experiment can conclude that so-called blame-shifting
 works but not that this is because of blame.
@@
 Modes offer a uniform way to compare the different semantics and  isolate
-blame from the effect of the semantics's runtime checks. Specifically, the Natural and
+blame from the effect of the semantics' runtime checks. Specifically, the Natural and
 Transient rational programmer come with a blame mode and an exception
-mode. Only if the former succeeds debugging a program while the latter
-fails, it is safe to conclude that blame is indeed necessary for the
+mode. If the former succeeds debugging a program while the latter
+fails, then it is safe to conclude that blame is indeed necessary for the
 rational programmer. Put differently, the exception mode serves as the
 {\em baseline\/} for blame's value within a given semantics; if the
-programmer in this mode ``beats'' the  blame one, a blame assignment
+programmer in this mode performs as well or better than the  blame one, a blame assignment
 mechanism might be useless.
 
-Besides the run-time checks, pure luck might also amplify the usefulness
-of blame. Thus a proper evaluation of blame needs to factor out any random
-effects too. A random rational programmer mode provides yet another necessary
-baseline using the same ingredients as for exceptions.
+Besides the run-time checks, the usefulness of blame might be due to pure luck.
+Indeed, a proper evaluation of blame needs to consider the role of chance in blame assignment, too.
+A random rational programmer mode provides yet another necessary baseline using the same ingredients as for exceptions.
 
@@
 useful experimental data in a multiplicative manner. Hence, carried out naively,
-the experiment would demand an unacceptably large amount of computational
+the experiment would demand an infeasible amount of computational
 resources.  A practical execution has no option but to {\em sample the space of
-scenarios\/}, carefully ensuring that it still produces reproducible
+scenarios\/}, carefully ensuring that it still has reproducible
 conclusions.

### 2021-03-03 writing/rational-natural.tex
challenges, start of rational: rewordings

@@
 assumptions about the untyped component in its interface, or the correct
-interface exposes a bug in the untyped component. Our setup rules out the first
+interface exposes a bug in the untyped component (or its dependencies). Our setup rules out the first
 alternative (but see section~\ref{sec:conclusion}), and therefore the rational
@@
 
-Here is a rigorous definition.
+Formally,
 \begin{quote}
@@
 a program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
-\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =\{\blame{\system, \conf_i}\}$ where
+\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =\{\blame{\system, \conf_i}\}$, where
 $\blame{\system, \conf}$ denotes the component that $\conf$ blames.
@@
 When the buggy untyped component of a program is replaced by the typed counterpart,
-the type checker fails if this component causes the impedance mismatch. Hence, a
-trail that ends at an ill-typed scenarios successfully pinpoints the location of
-bugs. 
+the type checker fails because this component causes the impedance mismatch. Hence, a
+trail that ends at an ill-typed scenario successfully pinpoints the location of
+the bug. 
 \begin{quote}
@@
 \end{quote}
-That is, failing Natural blame trails are those that have not reached a scenario
-that reveals the bug at compile time and ends in a scenario that does not blame
-an untyped component during program execution. Thus the rational programmer gets
+That is, failing Natural blame trails are those that end in a scenario that does not reveal the bug statically, yet also does not blame
+an untyped component. Thus the rational programmer gets
 no hints from the gradual type system on how to continue the search for the bug.
@@
 blame is a critical piece of the rational programmer's process.  For instance,
-typing the top of stack trace of a failed run time type check, dubbed the
+typing the top of a failed run time type check's stack trace, dubbed the
 location of the exception, might be as useful as typing the blamed one.
@@
 a Natural blame trail.
-Together the definitions for the two modes allow the comparison of the usefulness of blame 
-with that for mere exceptions for debugging a scenario in the context of Natural semantics.
+Together, the definitions for the two modes allow the comparison of the usefulness of blame 
+with that of mere exceptions for debugging a scenario in the context of Natural semantics.
 \begin{quote}

### 2021-03-03 writing/rational-stuff.tex
challenges, start of rational: rewordings

@@
 trail---also referred to as the \emph{root} (of the trail)---is the initial
-interesting debugging scenario.  Starting from such a root, the rational
+debugging scenario.  Starting from such a root, the rational
 programmer steps along a trail. If the (program for this) scenario type-checks,
-the program is run until it raises a run-time error.  The rational programmer
+the rational programmer runs the program until it raises a run-time error.  The rational programmer
 uses the information in the error to decide which component to equip with
 types. This choice constructs scenario $\conf_{i+1}$ from $\conf_{i}$ and thus
-creates a trail. The trail's construction ends when it reaches a scenarios that
-contain the mutant module because the type checker rejects its typed version
+creates a trail. The trail's construction ends when it reaches a scenario that
+contains the mutant module because the type checker rejects its typed version
 outright.  At this point, the source of the impedance mismatch is identified.

### 2021-03-03 writing/rational.tex
challenges, start of rational: rewordings

--- 2021-03-02 writing/rational.tex
@@
 
-Hence the research problem is how to make the actions of a rational programmer
-comparable across all semantics. Sections~\ref{sub:natural}
-through~\ref{sub:erasure} describe the various mode of the rational programmer and explain
-to what extent they are comparable. Section\ref{sub:experiment}
-enhances comparability with the programmer-effort dimension. Finally, the last
+Hence the research problem is how to make the modes of the rational programmer
+comparable even when they correspond to different semantics and offer different kinds of information. Sections~\ref{sub:natural}
+through~\ref{sub:erasure} describe the various modes of the rational programmer and explain
+the extent to which they are comparable. Section~\ref{sub:experiment}
+enhances their comparability with the programmer-effort dimension. Finally, the last
 subsection states the experimental questions and gives an overview of the

### 2021-03-03 writing/why-rational-methodology.tex
challenges, start of rational: rewordings

@@
 \end{enumerate}
-Unfortunately, seeing an algorithm is different from using the algorithm to conduct
-data-gathering experiments. The specific challenges are spelled out in the next
-section, and the following three sections explains ways of overcoming them.
+
+That said, using the method to conduct data-gathering experiments entails
+several challenges. The specific challenges are spelled out in the next section,
+and the following three sections explain ways of overcoming them.

### 2021-03-03 writing/challenges.tex
blame sequences becomes blame sets

@@
 assignment mechanisms of the three semantic variants.  While Natural assigns
-blame to {\em one\/} component, Transient assigns blame to a sequence of
+blame to {\em one\/} component, Transient assigns blame to a set of
 components. The Erasure semantics does not blame components {\it per se\/}, but
@@
 earliest point in the program's evaluation that can discover an impedance
-mismatch.  Another mode may opt to treat the sequence as a stack and add
+mismatch.  Another mode may opt to treat the set as a stack and add
 types to its newest element.  If both modes are equally successful in

### 2021-03-03 writing/discussion.tex
blame sequences becomes blame sets

@@
 The results in section~\ref{sec:results} show that Transient produces empty
-blame sequences for 967 scenarios. In theory, empty sequence must not occur. An
-empty blame sequence means a lack of boundary crossings for the witness value.
+blame sets for 967 scenarios. In theory, empty set must not occur. An
+empty blame set means a lack of boundary crossings for the witness value.
 By implication, a typed module is blaming itself---something that can

### 2021-03-03 writing/results.tex
results: bring back wrapfigure for success rates
blame sequences becomes blame sets

@@
 
-\begin{figure}
+\begin{wrapfigure}{r}{.40\textwidth}
   \includegraphics[width=0.40\textwidth]{./plots/success-bars}
@@
   \vspace{1em}
-  \begin{minipage}{0.95\textwidth}
+  \begin{minipage}{0.4\textwidth}
   The estimated percentage of scenarios for which each mode succeeds in locating the bug.
@@
   \label{fig:success-bars}
-\end{figure}
+\end{wrapfigure}
 
@@
 programmer without hints about how to proceed.
-Sections~\label{sec:threat:transient} and ~\label{sec:threat:transient2}
+Sections~\ref{sec:threat:transient} and ~\ref{sec:threat:transient2}
 explain how both of these causes of failure for Transient affect our

### 2021-03-03 writing/related.tex
space war
emphasizing difference with POPL paper

@@
 
-Technically, this work and~\citet{lksfd-popl-2020}'s differ in many ways. The
-experiment presented here concerns three different modes of gradual typing,
-theirs a single notion of contract checking.  It also contributes the idea of
-creating three comparable rational programmers (with several modes). The
+At a technical level, this work and~\citet{lksfd-popl-2020}'s differ in
+many ways. The experiment presented here concerns three different modes of
+gradual typing, theirs a single notion of contract checking.  It also
+contributes the idea of creating three comparable rational programmers
+(with several modes). Thus the method of this paper answers whether blame
+adds value to a gradual type system and what gradual type system it
+benefits the most; while~\citet{lksfd-popl-2020} only establish a
+systematic relation between blame for contracts and behavioral bugs .The
 type-level mutators represent another technical contribution of the work
 presented here; as explained in section~\ref{sec:mutate}, almost none
-of~\citet{lksfd-popl-2020}'s mutators is useful at the level of type mistakes.
+of~\citet{lksfd-popl-2020}'s mutators is useful at the level of type
+mistakes.
 

### 2021-03-03 writing/rational-effort.tex
rational-*: rewording

@@
 In addition to success and failure information for the various trails, the
-experiment software records the number of components a rational programmer has
+experiment records the number of components a rational programmer has
 to equip with types along each trail ($\lvert \conf_n \setminus \conf_0
-\rvert$). This number can serve as a metric.
+\rvert$). This number can serve as an additional metric to compare modes.
 
-A comparison of the effort of different modes of the rational programmer can
-illuminate the comparison of the effectiveness of the three gradual typing
-systems. For example, consider a debugging scenario.  If, for instance, both
-Natural blame and Transient first blame trails are successful, the two modes of
-the rational programmer can compete to see which one debugs the interesting
+Comparing the effort of different modes of the rational programmer can
+illuminate the comparitive effectiveness of the three gradual typing
+systems. If, for instance, both
+Natural blame and Transient first blame trails are successful for the same scenario, the two modes of
+the rational programmer can compete to see which one debugs the
 scenario with less effort. In general, if the effort distribution for a mode of
@@
 
-Measuring effort can also tell whether the observed effectiveness of the
-rational programmer is an artifact of pure chance or not.  Doing so means
-comparing the effort distribution for a mode of the rational programmer with
-another mode that ignores error information entirely and instead selects which
-component to type next randomly.  Since each mutant has a finite number of
-components, such random mode trails are naturally always successful. At the same
-time though, the Natural, Transient, and Erasure random modes exhibit a normal
-effort distribution across scenarios that should be distinct from the effort
-distribution for all other modes of the rational programmer if the effectiveness
-of the modes are not coincidental.
+Measuring effort can also reveal whether the observed effectiveness of the
+rational programmer is an artifact of pure chance or not.
+In particular, the effort distribution for a mode of the rational programmer can be compared with that of
+the random mode that ignores error information entirely and instead selects which
+component to type randomly.  Since each mutant has a finite number of
+components, random mode trails are always successful. However, the random mode's effort
+distribution will be thinly spread out across the range of trail lengths possible
+in the set of debugging scenarios.
+In contrast, the effort distribution of other modes should be quite different if their effectiveness is not coincidental.

### 2021-03-03 writing/rational-experiment.tex
restructuring benchmarks and moving earlier the experiment process figure
rational-*: rewording

@@
 blame for Natural, Transient and Erasure. In line with the discussion so far,
-the experiment software collects data to answer three initial questions for 
+the experiment collects data to answer three initial questions for 
 interesting debugging scenarios:
@@
 question $Q_1$ asks whether blame is valuable for Natural and the experiment
-uses the Natural blame and exception trails to answer it.
+uses the Natural blame and exception trails to answer it, so $Q_1$ shows up in the cells for Natural blame and Natural exceptions.
 
@@
 
-In detail the answer to $Q_1$ demands a comparison of the success of the Natural
-blame and Natural exception trails for all interesting debugging scenarios.  The
+In detail, the answer to $Q_1$ demands a comparison of the success of the Natural
+blame and Natural exception trails for all debugging scenarios.  The
 first step is to construct each mutant's scenario lattice and identify their
-interesting debugging scenarios.  The experiment software extends the trails that start
-from such roots according to the Natural-blame programmer.  If no scenarios can
-be added to the trail, the experiment software checks whether the last scenario
-of the trail type-checks or not. If it does, the software records that the
-Natural-blame trail is successful; otherwise it is recorded as failing. The
-process is repeated for the same roots again but for the other mode, Natural
-exceptions.  Figure~\ref{fig:process} summarizes this experimental process for
+debugging scenarios.  The experiment then extends the trails that start
+from those roots according to the Natural-blame programmer.  If no scenarios can
+be added to the trail, the experiment checks whether the last scenario
+of the trail type-checks or not. If it does, the
+Natural-blame trail is successful; otherwise it is failing. 
+Figure~\ref{fig:process} summarizes this experimental process for
 one mode of the rational programmer and connects it with the mutations from
-section~\ref{sec:mutate}.  After completing the experiment, the experiment
-software reports the success/failure results of the trails to determine for each
-root whether Natural blame is more useful than Natural exceptions.  Question
+section~\ref{sec:mutate}.
+The experiment repeats the process for the same roots with the other mode, Natural
+exceptions.
+After completing, the experiment
+reports the success/failure results of the trails to determine the proportion of
+scenarios where Natural blame is more useful than Natural exceptions.  Question
 $Q_1$ has a positive answer if a root exists where the above is true because it
@@
 
-For $Q_*$, the process is a bit more involved. Answering this question calls for
-a comparison of the percentage of scenarios where one mode is more useful than
-the other and the inverse.  For instance, to decide whether blame for Natural is
-more useful than first blame for Transient, the software compares the percentage
-of interesting scenarios where Natural blame is more successful than Transient
-first blame with the percentage of interesting scenarios where Transient first
-blame is more successful than Natural blame.  Finally these steps are repeated
-one more time to compare Natural blame and Transient-last blame and get a
-complete picture of the comparative usefulness of blame in the two semantics.
-
-\begin{figure}
+\begin{figure}[bh]
   \centering
@@
 \end{figure}
+
+For $Q_*$, the process is a bit more involved. Answering this question calls for
+a comparison of the percentage of scenarios where one mode is more useful than
+the other and the inverse.  For instance, deciding whether blame for Natural is
+more useful than first blame for Transient requires comparing the percentage
+of scenarios where Natural blame is more successful than Transient
+first blame with the percentage where Transient first
+blame is more successful than Natural blame. Repeating the whole process
+for every pair of modes produces a
+complete picture of the comparative usefulness of blame between all of the modes.

### 2021-03-03 writing/rational-transient.tex
challenges, start of rational: rewordings
rational-*: rewording

--- 2021-02-23 writing/rational-transient.tex
@@
 
-The transient semantics assigns blame to a set of components. In this setting, a
-blame assignment has the following meaning: the witness value to the impedance
-mismatch may have crossed the boundaries between elements in the set, and each crossing
+The Transient semantics assigns blame to a set of components. The blame
+assignment says that the value witnessing the impedance mismatch may have
+crossed the boundaries between elements in the set, and each crossing
 checked the value's type in a shallow manner.
 
-This ambiguity in Transient blame raises the question of how the rational programmer
-should react when the language produces a blame set. Our answer is that the rational
-programmer has at least two reasonable options. The first one is to select the component that is added
-to the blame set first and assign types to only that one---after all, if
-fully checked, the types of
-this first component should be able to detect an impedance mismatch earlier in the
-evaluation of a program than the later ones. The second option is to
-select the component that is added to the blame set last, effectively
-interpreting the blame set as a boundary-aware stack.
+This ambiguity in Transient blame raises the question of how the rational
+programmer should react when the language produces a blame set. Our answer
+is that the rational programmer has at least two reasonable options. The first
+one is to select the component that is added to the blame set first and
+assign types to only that one---after all, if fully checked, the types of this
+first component should be able to detect an impedance mismatch earlier in the
+evaluation of a program than the later ones. The second option is to select the
+component that is added to the blame set last, effectively interpreting the
+blame set as a boundary-aware stack.
 
 These two modes of rationalizing give rise to two different notions of trail.
+
 \begin{quote}
@@
 \end{quote}
-
-and
 
@@
 
-The definition of \emph{Transient exception trails} 
-is analogous to that for Natural and, same as for Natural, we use it to  
-specify when  
-Transient-first and Transient-last blame are more useful than 
-Transient exceptions for debugging a scenario.
+The definition of \emph{Transient exception trails} is analogous to that for
+Natural. It is used to specify when Transient-first and Transient-last blame are
+more useful than Transient exceptions for debugging a scenario.

### 2021-03-03 writing/introduction.tex
touch ups in intro

@@
-Theoreticians of gradual typing have focused on blame theorems from the very
-beginning~\cite{mf-toplas-2009, tf-dls-2006}. ``Well-typed
+Theoreticians of gradual typing have focused on blame theorems from the
+very beginning~\cite{mf-toplas-2009, tf-dls-2006}. ``Well-typed
 [components]\footnote{The original authors got this word wrong.} can't be
-blamed'' turned the theorem into a slogan~\cite{wf-esop-2009}. Academic systems
-(Reticulated Python~\cite{vsc-dls-2019, vss-popl-2017, vksb-dls-2014} and Typed
-Racket~\cite{tf-dls-2006,tf-popl-2008,tfffgksst-snapl-2017,tf-icfp-2010}) come
-with sophisticated checking and blame assignment strategies (see sec.~\ref{sec:landscape}). Their academic creators embrace
-the idea that blame can help practicing programmers find impedance mismatches: disagreements
-between the types they added to a software system and its behavior.
+blamed'' turned the theorem into a slogan~\cite{wf-esop-2009}. Academic
+systems (Reticulated Python~\cite{vsc-dls-2019, vss-popl-2017,
+vksb-dls-2014} and Typed
+Racket~\cite{tf-dls-2006,tf-popl-2008,tfffgksst-snapl-2017,tf-icfp-2010})
+come with sophisticated checking and blame assignment strategies (see
+sec.~\ref{sec:landscape}). Their academic creators embrace the idea that
+blame can help practicing programmers find impedance mismatches:
+disagreements between the type ascriptions of a software system and its
+behavior.
 
@@
 Transient\/}, i.e. Reticulated's tracking of typed/untyped boundaries~\cite{vss-popl-2017}; (2)~{\it
-Natural\/}, i.e. the use of a higher-order contract system and its blame
+Natural\/}, i.e. Typed Racket's use of a higher-order contract system and its blame
 assignment~\cite{ff-icfp-2002}; and (3) {\it Erasure\/}, i.e. the approach of
@@
 huge. The second one concerns the difference between the methods; neither is
-{\it Natural\/} vastly better than {\it Transient\/} nor are the two blame
-assignment methods clearly preferable to {\it Erasure\/}.  In turn, these
+{\it Natural\/} vastly more useful than {\it Transient\/} nor are the two blame
+assignment methods always superior to {\it Erasure\/}.  In turn, these
 problems suggest that, on one hand, the existing theory does not predict

### 2021-03-03 writing/introduction.tex
touch ups in intro

@@
 
-The paper's {\it second contribution\/} is a set of {\em results from applying
-the evaluation method to three distinct blame assignment strategies in
-approximately 72,200 different scenarios\/} (see sec.~\ref{sec:results}): (1)~{\it
-Transient\/}, i.e. Reticulated's tracking of typed/untyped boundaries~\cite{vss-popl-2017}; (2)~{\it
-Natural\/}, i.e. Typed Racket's use of a higher-order contract system and its blame
-assignment~\cite{ff-icfp-2002}; and (3) {\it Erasure\/}, i.e. the approach of
-industrial systems, which forgo blame in favor of error messages from the safety
-checks of the underlying language. The results (see
-sec.~\ref{sec:discussion}) are at least somewhat surprising.  In principle
-they validate the conjectures behind the work of theoreticians.  First, a good
-blame assignment strategy helps with the search for impedance mismatches between
-the specified types and the behavior of untyped components.  Second, Natural's
-wrapper-based blame tracking is more useful than Transient's ``collective
-blame'' tracking algorithm, which in turn is superior to Erasure. {\em But\/},
-the application of the method also indicates problems with the expectations of
-theoreticians. The first problem is that the cost of Transient's blame can be
-huge. The second one concerns the difference between the methods; neither is
-{\it Natural\/} vastly more useful than {\it Transient\/} nor are the two blame
-assignment methods always superior to {\it Erasure\/}.  In turn, these
+The paper's {\it second contribution\/} is a set of {\em results from
+applying the evaluation method to three distinct blame assignment
+strategies in approximately 72,200 different scenarios\/} (see
+sec.~\ref{sec:results}): (1)~{\it Transient\/}, i.e. Reticulated's
+tracking of typed/untyped boundaries~\cite{vss-popl-2017}; (2)~{\it
+Natural\/}, i.e. Typed Racket's use of a higher-order contract system and
+its blame assignment~\cite{ff-icfp-2002}; and (3) {\it Erasure\/}, i.e.
+the approach of industrial systems, which forgo blame in favor of error
+messages from the safety checks of the underlying language. The results
+(see sec.~\ref{sec:discussion}) are at least somewhat surprising.  In
+principle they validate the conjectures behind the work of theoreticians.
+First, a good blame assignment strategy helps with the search for
+impedance mismatches between the specified types and the behavior of
+untyped components.  Second, Natural's wrapper-based blame tracking is
+more useful than Transient's ``collective blame'' tracking algorithm,
+which in turn is superior to Erasure. {\em But\/}, the application of the
+method also indicates problems with the expectations of theoreticians. The
+first problem concerns the difference between the methods; neither is {\it
+Natural\/} vastly more useful than {\it Transient\/} nor are the two
+blame assignment methods always superior to {\it Erasure\/}.  The second
+one is that the cost of Transient's blame can be huge.  In turn, these
 problems suggest that, on one hand, the existing theory does not predict

### 2021-03-03 writing/mutate-benchmarks.tex
bad latex
mutate-*: rewording

@@
 
-   \itme  fail and contain type-level mistakes. 
+   \item  fail and contain type-level mistakes. 
 \end{itemize}
@@
 Hence, even if they do not come with type-level errors, they offer a
-diverse lignuistic substrate suitable for the careful introduction of
+diverse linguistic substrate suitable for the careful introduction of
 type-level errors via mutations. Further filtering the benchmark suite to
@@
 diversity. Figure~\ref{table:benchmark-descriptions} displays the selected
-benchmark programs and describes their basic characteristic.
+benchmark programs and describes their basic characteristics.

### 2021-03-03 writing/mutate-interesting.tex
mutate-*: rewording

--- 2021-03-02 writing/mutate-interesting.tex
@@
 
-A type-level mutation is {\em interesting\/} (1) if, upon removal of the type
-annotations, running the program raises a run-time error, (2) whose stack trace
-contains source locations from at least three modules, and (3) if this happens
-in all of the to-be-evaluated semantics. An interesting mutation represents a
-{\em debugging scenario\/} that supplies the rational programmer with
-information.
+A type-level mutation is {\em interesting\/} (1) if the type checker rejects the fully typed version of the mutant,
+(2) running the mutant with all type annotations removed raises a run-time error, and
+(3) that error's stack trace contains source locations from at least three modules.
 
@@
 
-\item By definition, an impedance mismatch is a clash between the type ascription of
-one module's imports and another module's exports. Hence, type checking fails for an
-interesting mutant module, and removing the types from this mutant module
-creates an impedance mismatch between it and the modules that import it,
-directly or indirectly. This impedance mismatch is of interest if, and only if,
-running the program leads to an exception---otherwise it goes unobserved and
-cannot be debugged.
+\item By definition, an impedance mismatch is a clash between the type ascription of one module's imports and another module's exports. Hence, type checking should fail for an interesting mutant.
 
-\item If the evaluation od a mutant module immediately raises an exception because
+\item The goal of a comparative evaluation is to give the rational programmer a
+chance to debug the same scenario using different pieces of information.  In the
+case of gradual typing semantics, a meta-theorem due to Greenman and
+Felleisen~\cite{gf-icfp-2018} says that if a program raises an exception under
+Erasure, it also errors under all other semantics.  Hence, a comparison of blame
+information insists that an interesting mutant {\em raises a run-time exception
+under Erasure\/}.
+
+{\bf Note} While this choice favors Erasure over Transient and Natural and, for
+the same reason, Transient over Natural, some form of bias towards one or the
+other semantics is unavoidable. Tipping the scales in favor of the theoretically
+weakest semantics yields the most stable results. See
+section~\ref{sec:discussion} for some further discussion of this choice.
+
+\item If the evaluation of a mutant module immediately raises an exception because
 of the changes, there is no work for the rational programmer. Indeed, if the
@@
 between the buggy module with its context.  In these cases, the rational
-programmer must examine at least two modules to locate the source of the faulty
+programmer must contend with at least two modules involved in a faulty
 interaction.
-
-\item The goal of a comparative evaluation is to give the rational programmer a
-chance to debug the same scenario using different pieces of information.  In the
-case of gradual typing semantics, a meta-theorem due to Greenman and
-Felleisen~\cite{gf-icfp-2018} says that if a program raises an exception under
-Erasure, it also errors under all other semantics.  Hence, a comparison of blame
-information insists that a debugging scenario {\em raises a run-time exception
-under Erasure\/}.
-
-{\bf Note} While this choice favors Erasure over Transient and Natural and, for
-the same reason, Transient over Natural, some form of bias towards one or the
-other semantics is unavoidable. Tipping the scales in favor of the theoretically
-weakest semantics yields the most stable results. See
-section~\ref{sec:discussion} for some further discussion of this choice.
 
@@
 
-The definition of interesting debugging scenarios creates a powerful filter. All
-together, the listed mutators produce 16,800 mutants with at least one
-interesting debugging scenario across all benchmarks; see
+The definition of interesting mutants creates a powerful filter. All
+together, the listed mutators produce 16,800 interesting mutants across all benchmarks; see
 figure~\ref{fig:mutant-breakdown} for an overview. Broken down by benchmark, the
-mutators produce at least 40 interesting scenarios for every benchmark, and
-these scenarios originate from at least four different mutators per benchmark.
+mutators produce at least 40 interesting mutants for every benchmark, and
+these mutants originate from at least four different mutators per benchmark.
 Thus, the mutators result in a sizable and diverse population of scenarios for
-every benchmark.  Furthermore, every mutator contributes scenarios to at least
+every benchmark.  Furthermore, every mutator contributes interesting mutants in at least
 one benchmark.  Some mutators apply only to a few benchmarks, because they
@@
 features.
+
+The definition of interesting mutants is more than a filter for
+uninteresting mutants. It is also the guide for evaluating and refining
+mutators until they collectively produce a sizable and diverse population of
+interesting mutants. For example, one early mutator idea casts the
+tests of conditional expressions to the {\tt Any} type to
+suppress occurrence typing. However, an examination of the mutants it
+produces through the lens of the definition of interesting mutants
+reveals that none is interesting because they never raise run-time errors.
+Indeed, the interesting standard guided countless iterations of adding,
+removing, and refining the mutators in figure~\ref{table:mutation-ops}
+to maximize interesting mutants.

### 2021-03-03 writing/mutate-mutators.tex
small tweaks to emphasize interesting mutations
mutate-*: rewording

@@
 checks of the underlying language can detect. Once detected, the rational
-programmer should be able to locate the mistake by gradually adding types to
-blamed modules.
+programmer should be able to locate the mistake by gradually adding types
+to blamed modules. In other words, the suitability of the mutators hinges
+on their ability to generate interesting debugging scenarios (see
+sec.~\ref{sub:mutate-interesting}).
 
@@
 of~\citet{lksfd-popl-2020} mutators do not satisfy the constraints and had to
-be eliminated. For example, the relational mutator does not lead to type-level
+be eliminated or significantly adapted. For example, the relational mutator does not lead to type-level
 errors. Similarly, the arithmetic mutator does not reliably lead to type
@@
 their decades-long experience of making type-level mistakes in Typed Racket,
-some of which take a non-trivial effort to debug.
+some of which take non-trivial effort to debug.
 
@@
 objects. The last four target distinguishing features of Typed Racket,
-specifically is sophisticated type system. In particular, {\tt arithmetic}
+specifically its sophisticated type system. For example, one mutation produced by {\tt arithmetic}
 replaces a {\tt +} with a {\tt -} in an attempt to change the type of the
@@
 arguments are positive integers, while {\tt -} yields {\tt
-Integer}~\cite{stathff-padl-12}. Similarly, the other three also aim to usurp
+Integer}~\cite{stathff-padl-12}. Similarly, the other three aim to confound
 Typed Racket's occurrence type system.

### 2021-03-03 writing/mutate.tex
re-established the old organization
editiroal voice thru 6
mutate-*: rewording

@@
 
-Putting a rational programmer to work means generating many mutants and turning
-those into debugging scenarios. Since existing mutators do not generate useful
-mutants, the first step is to develop new mutators (see sec.~\ref{sub:mutate-mutators})
-and to validate their suitability for this application (see sec.~\ref{sub:mutate-interesting}).
+Putting the rational programmer to work means generating many mutants and turning
+those into debugging scenarios. The process must start with a suitable
+collection of representative programs (see sec.~\ref{sub:mutate-benchmarks}).
+Since existing mutators do not generate useful mutants, the first step is to
+develop new mutators (see sec.~\ref{sub:mutate-mutators}) and to validate their
+suitability for this application (see sec.~\ref{sub:mutate-interesting}).
 
@@
 
+\input{mutate-benchmarks-figure.tex}
+\sub{benchmarks}{The Experimental Benchmarks}
 \input{mutate-mutators-figure.tex}

### 2021-03-03 writing/landscape.tex
touch ups in section 2

@@
 proxy, which in turn, enforces the type of the function with run-time checks.
-In essence, this check protects typed modules when they call an untyped function
+In essence, these checks protect typed modules when they call an untyped function
 because it may not produce the promised kind of value, as is the case here with
@@
 function of a typed module (such as {\tt crypto-pack}) in a proxy that checks
-its arguments.  This check protects functions exported from typed modules against
-applications in untyped code that supply the wrong kind of argument value.
+its arguments.  These checks protect functions exported from typed modules against
+applications in untyped code that supply the wrong kind of arguments.
 
@@
 they cross, plus the corresponding types. In the example, the map records that
-\texttt{pack} crosses from \texttt{pack-lib} to {\tt crypto-pack-lib} and
-from {\tt crypto-pack-lib} to {\tt client} with the type that appears in the
+\texttt{pack} crosses from \texttt{pack-lib} to {\tt typed-pack-lib} and
+from {\tt pack-lib} to {\tt crypto-pack-lib} with the type that appears in the
 {\tt required/typed} forms in the example. Since the failed check corresponds to
@@
 \begin{center}
-\begin{tabular}{l|ll}
+  \begin{tabular}{l|ll@{}}
                         &        \multicolumn{2}{|c}{the evaluation of} \\
@@
      		   	 &                                & \\
-\qquad {\it Transient\/} & no error   		          & error, blaming the boundary \\
-      		  	 &    			          & \quad {\tt pack-lib}/{\tt crypto-pack-lib} \\
+\qquad {\it Transient\/} & no error   		          & error, blaming the boundary crossings \\
+      		  	 &    			          & \quad {\tt pack-lib} to {\tt typed-pack-lib} \\
+               &                     & \quad {\tt pack-lib} to {\tt crypto-pack-lib} \\
 \qquad {\it Erasure\/}   & no error 		          & no error \\

### 2021-03-03 writing/sample.tex
sample: rewording

@@
 
-As is, the chosen mutators generate a huge number of debugging scenarios,
-approximately 1,000,000 for the chosen benchmark. This number of scenarios is
+As is, the chosen mutators generate a huge number of debugging scenarios:
+approximately 1 million for the chosen benchmarks. This number of scenarios is
 far too large to even identify the interesting ones among them.  Furthermore
-this population is heterogeneous; different scenarios come from different
-mutants and different mutants are the result of a different mutator and
-benchmark pair.
+this population is heterogeneous; scenarios come from different
+mutants, and the mutants are the result of different mutators applied to each
+benchmark.
 
-To make the experiment computationally feasible and statistically sound,
-stratified uniform random sampling with three levels of stratification becomes
-necessary.  The first two levels group mutants first by benchmark and then by
-mutator.  The set of the randomly selected mutants of each benchmark is
-checked to make sure that, per benchmark, the sample reflects the diversity of
+To become computationally feasible and statistically sound, the experiment employs
+stratified uniform random sampling with three levels of stratification.
+The first two levels group mutants first by benchmark and then by
+mutator, making sure that the sample within each benchmark reflects the diversity of
 mutants with respect to the mutators that generated them.  Specifically, the
-process samples 80 mutants with interesting scenarios per benchmark,
+experiment samples 80 interesting mutants per benchmark,
 evenly-distributed across all of the mutators that contribute mutants for the
-benchmark. Some benchmarks has less than  80 mutants with interesting scenarios
-in which case the process collects them all which results in a total of 756 mutants
-across all benchmarks.  
-At the third level, the sampling process draws 96 random interesting
-scenarios from each configuration lattice with replacement. In sum, the
-final sample consists of 72,200 interesting scenarios.
+benchmark. Some benchmarks have less than  80 mutants with interesting scenarios,
+in which case the only choice is to take them all.
+The result is a total of 756 interesting mutants across all benchmarks.
+Finally, the third level of sampling draws 96 random interesting debugging
+scenarios from each configuration lattice with replacement. The
+final sample thus consists of 72,192 interesting scenarios.

### 2021-03-03 writing/challenges.tex
touch ups in section 4

@@
 authors use Racket, which is thus far the only language in which all three major
-soundness mechanisms are available in a reasonably robust and comparable manner;
-for the implementation of the transient semantics, see the parallel submission
+soundness mechanisms are available in a robust and comparable manner;
+for the implementation of Transient Typed Racket, see the parallel submission
 of an experience report~\cite{ttt21}.
@@
 role for the rational programmer. Indeed, precisely because they do not
-account for confounding factors, \citet{lksfd-popl-2020} cannot draw any
+account for such confounding factors, \citet{lksfd-popl-2020} cannot draw any
 conclusions about blame specifically despite advertisements for the
@@
 Modes offer a uniform way to compare the different semantics and  isolate
-blame from the effect of the semantics' runtime checks. Specifically, the Natural and
+blame from the effect of the semantics' run-time checks. Specifically, the Natural and
 Transient rational programmer come with a blame mode and an exception

### 2021-03-03 writing/rational-experiment.tex
touch ups in section 5

@@
 is evidence that there is at least one interesting scenario that the rational
-programmer manages to debug with blame information.  The process is analogous
+programmer manages to debug because of blame information.  The process is analogous
 for $Q_2$ and $Q_3$, using the respective modes of the rational programmer.

### 2021-03-03 writing/rational-natural.tex
touch ups in section 5

@@
 While a successful Natural blame trail indicates that it 
-pays off to heed blame assignments during debugging the trail's root, it does not answer whether
+pays off to heed blame assignments while debugging the trail's root, it does not answer whether
 blame is a critical piece of the rational programmer's process.  For instance,
-typing the top of a failed run time type check's stack trace, dubbed the
+typing the top of a failed run-time type check's stack trace, dubbed the
 location of the exception, might be as useful as typing the blamed one.

### 2021-03-03 writing/rational.tex
touch ups in section 5

@@
 Each mode receives different kinds of information and thus may construct
-different paths in the lattice. As discussed in~\ref{sec:challenges},
+different paths in the lattice. As discussed in section~\ref{sec:challenges},
 evaluating blame relies on comparing modes of the rational programmer

### 2021-03-03 writing/results.tex
results: rewording

@@
 
-The experimental process from section~\ref{sec:rational} was run on the
-sampled debugging scenarios from section~\ref{sec:sample} on a machine
-with two Intel Xeon Gold 6258R processors (28 doubly-threaded cores each)
-and 500GB of memory.  Each debugging scenario had a 4 minute timeout and a
+A machine with two Intel Xeon Gold 6258R processors (28 doubly-threaded cores each)
+and 500GB of memory ran the experimental process from section~\ref{sec:rational} on the
+sampled debugging scenarios from section~\ref{sec:sample}.  Each debugging scenario had a 4 minute timeout and a
 memory limit of 6GB. The complete experiment required over 30,000 compute
@@
 Figure~\ref{fig:success-bars} summarizes the overall success rates of
-every mode.  The success rates illustrate a few points that underlie the
+every mode.\footnote{The random mode behaves the same for
+all three semantics, so this analysis considers it a single mode.}
+The success rates illustrate a few points that underlie the
 rest of our analysis.  The first notable piece of information from this
@@
 should type next. For instance the stack trace of the exception may not
-contain frames from any module of the program but only from the internals
-of the Typed Racket implementation. This situation can happen at any point
+contain frames from any untyped module of the program. This situation can happen at any point
 along a blame trail, causing it to fail.
@@
 While most failures to debug scenarios follow the above pattern, a few do
-not.  Breaking down the reasons for failure for Natural blame (1748 in
+not.  Breaking down the failure reasons for Natural blame (1748 in
 total) reveals an additional cause. For a small set of debugging scenarios
-(40), Natural produces a run time type error blaming a non-buggy already
-typed component. All these cases are due to known open issues with Typed
+(40), Natural produces a run time type error blaming a non-buggy
+already-typed component. All these cases are due to known open issues with Typed
 Racket and class contracts.
@@
 Sections~\ref{sec:threat:transient} and ~\ref{sec:threat:transient2}
-explain how both of these causes of failure for Transient affect our
+explain both of these causes of failure for Transient in detail and how they affect our
 experiment.
@@
 scenarios, while their corresponding exception modes succeed in around
-85\% of them and Erasure succeeds in only 60\%.  The only
-exception is that the Random programmer always succeeds; this reflects the
-fact that every scenario has finitely many modules, so Random programmer 
+85\% of them, and Erasure succeeds in only 60\%.  The only
+exception is that the random programmer always succeeds; this reflects the
+fact that every scenario has finitely many modules, so the random programmer 
 eventually types the buggy module.
@@
 programmer that uses blame.  Each diagram shows the overlap of successful
-for a  the blame mode, its corresponding exception mode, and Erasure.  For
-example, for Natural, the leftmost diagram shows that all three modes
+scenarios for  the blame mode, its corresponding exception mode, and Erasure.  For
+example, the leftmost diagram (Natural) shows that all three modes
 succeed on 59.5\% of the  scenarios, only Natural and Natural exceptions
@@
 mode of the rational programmer is highly spread out, as
-expected.\footnote{The random mode distribution is indistinguishable for
-all three semantics and thus the figure~\ref{fig:effort-table} only shows
-one plot.} In contrast, in the other modes, successful effort coalesces at
+expected. In contrast, in the other modes, successful effort coalesces at
 the left side of the plot, meaning that in most cases the programmer needs

### 2021-03-03 writing/discussion.tex
discussion: start edits

@@
 have described the various ways we attempt to mitigate these threats;
-the remainder of this section discussed three additional threats.
+the remainder of this section discusses three additional threats.
 
@@
 here. When this happens, the predicted benefits of blame assignment may not
-materialize. Indeed, the authors's personal experience suggests such deviations, and
+materialize. Indeed, the authors' personal experience suggests such deviations, and
 it also suggests that deviating is a mistake. To make a true judgment of the
-usefulness of the rational-programmer idea, the community will need a lot more
+usefulness of the rational-programmer idea, the community will need much more
 experience with this form of evaluation and relating the evaluation to the
@@
 component in different ways. The creation and curation of the benchmarks over
-many years has driven home this lesson, but fortunately, it has also shown that
+many years has driven home this lesson but, fortunately, it has also shown that
 the types are in most cases reasonably canonical.  The authors therefore conjecture that

### 2021-03-03 writing/conclusion.tex
pass over section 11
rw and conclusion: small edits

@@
 work of~\citet{tgpk-dls-2018} suggests that, by intuition, programmers prefer
-the run-time checking of Natural over other soundness methods. But, just because
+the run time checking of Natural over other soundness methods. But just because
 a random set of programmers expresses this view does not mean that blame
-assignment adds value. At this point, it is clear that, in some circumstances,
-it does and in others, the answer remains murky.
+assignment adds value. At this point it is clear that in some circumstances
+it does, and in others the answer remains murky.
 

### 2021-03-03 writing/related.tex
rw and conclusion: small edits

--- 2021-03-03 writing/related.tex
@@
 \citet{lksfd-popl-2020}'s human-out-of-the-loop empirical analysis of blame for
-higher-order contracts in the inspiration for this paper.  While they do not
+higher-order contracts is the inspiration for this paper.  While they do not
 spell out the notion of the rational programmer, they present the basic
@@
 
-Technically, this work and~\citet{lksfd-popl-2020}'s differ in many ways. The
-experiment presented here concerns three different modes of gradual typing,
-theirs a single notion of contract checking.  It also contributes the idea of
-creating three comparable rational programmers (with several modes). The
+At a technical level, this work and~\citet{lksfd-popl-2020}'s differ in
+many ways. The experiment presented here concerns three different modes of
+gradual typing, theirs a single notion of contract checking.  It also
+contributes the idea of creating three comparable rational programmers
+(with several modes). Thus the method of this paper answers whether blame
+adds value to a gradual type system and what gradual type system it
+benefits the most; while~\citet{lksfd-popl-2020} only establish a
+systematic relation between blame for contracts and behavioral bugs. The
 type-level mutators represent another technical contribution of the work
 presented here; as explained in section~\ref{sec:mutate}, almost none
-of~\citet{lksfd-popl-2020}'s mutators is useful at the level of type mistakes.
+of~\citet{lksfd-popl-2020}'s mutators is useful at the level of type
+mistakes.
 

### 2021-03-03 writing/results.tex
touch ups in section 8

@@
 The success rates illustrate a few points that underlie the
-rest of our analysis.  The first notable piece of information from this
+rest of the analysis.  The first notable piece of information from this
 figure is that every mode has failed debugging scenarios, not just
 Erasure. This should not come as a surprise to the astute reader.  Running
-a mode of the rational programmer on a scenario may result in an exception
+a rational programmer mode on a scenario may result in an exception
 that carries no useful information about which module the programmer
 should type next. For instance the stack trace of the exception may not
-contain frames from any untyped module of the program. This situation can happen at any point
+contain frames from any untyped module of the program. This can happen at any point
 along a blame trail, causing it to fail.
 
-While most failures to debug scenarios follow the above pattern, a few do
+While most blame trail failures follow the above pattern, a few do
 not.  Breaking down the failure reasons for Natural blame (1748 in
@@
 However, Transient also has a substantial number of failures because
-scenarios hit the time and memory limits of our experiment
+scenarios hit the time and memory limits of the experiment
 (\textasciitilde770 scenarios).  Additionally, there are nearly 1,000
@@
 Sections~\ref{sec:threat:transient} and ~\ref{sec:threat:transient2}
-explain both of these causes of failure for Transient in detail and how they affect our
+explain both of these causes of failure for Transient in detail and how
+they affect the
 experiment.
@@
 the success trade-offs each semantics offers against Erasure, with and
-without blame. Hence, it answers questions about how to choose to between
-blame or just exceptions for each semantics.  For instance, the analysis
+without blame. For instance, the analysis
 for Natural clearly illustrates that, when choosing between Natural blame,

### 2021-03-03 writing/challenges.tex
touch ups for consistent use of mutant/mutated module and module/component

@@
 Roughly speaking, the new mutators inject type errors into fully typed programs.
-Applying such a mutator to any typed component produces a mutant component.  A
-debugging scenario results from removing the types from the mutant. For the
+Applying such a mutator to any typed component produces a mutated component.  A
+debugging scenario results from removing the types from the mutated component. For the
 design of such mutators, the authors relied on their own extensive programming

### 2021-03-03 writing/discussion.tex
discussion: last edit
touch ups for consistent use of mutant/mutated module and module/component

@@
 run-time checks signal an impedance mismatch in the real world, a programmer
-does not have a typed component ready to swap in. Instead, the programmer must
+does not have a typed module ready to swap in. Instead, the programmer must
 come up with the next set of types---and this means the programmer must make
 choices. It is normally possible to assign consistent types to variables in a
-component in different ways. The creation and curation of the benchmarks over
+module in different ways. The creation and curation of the benchmarks over
 many years has driven home this lesson but, fortunately, it has also shown that
@@
 scenarios exceed the 4-minute timeout or the 6GB-memory limit. To put those
-limits into context, the fully typed and fully untyped benchmarks all complete
+limits into context, the fully typed and fully untyped benchmarks all normally complete
 in a few seconds with minimal memory usage. Furthermore, none of the mixed

### 2021-03-03 writing/rational-effort.tex
touch ups for consistent use of mutant/mutated module and module/component

@@
 In addition to success and failure information for the various trails, the
-experiment records the number of components a rational programmer has
+experiment records the number of modules a rational programmer has
 to equip with types along each trail ($\lvert \conf_n \setminus \conf_0
@@
 the random mode that ignores error information entirely and instead selects which
-component to type randomly.  Since each mutant has a finite number of
-components, random mode trails are always successful. However, the random mode's effort
+module to type randomly.  Since each mutant has a finite number of
+modules, random mode trails are always successful. However, the random mode's effort
 distribution will be thinly spread out across the range of trail lengths possible

### 2021-03-03 writing/rational-natural.tex
touch ups for consistent use of mutant/mutated module and module/component

@@
 The Natural semantics assigns blame to exactly one boundary.  A blame assignment
-has the following specific meaning: the typed component may make incorrect type
-assumptions about the untyped component in its interface, or the correct
-interface exposes a bug in the untyped component (or its dependencies). Our setup rules out the first
+has the following specific meaning: the typed module may make incorrect type
+assumptions about the untyped module in its interface, or the correct
+interface exposes a bug in the untyped module (or its dependencies). Our setup rules out the first
 alternative (but see section~\ref{sec:conclusion}), and therefore the rational
-programmer extends the trail to a scenario that swaps out the untyped component
+programmer extends the trail to a scenario that swaps out the untyped
+module
 for its typed counterpart.
@@
 \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =\{\blame{\system, \conf_i}\}$, where
-$\blame{\system, \conf}$ denotes the component that $\conf$ blames.
+$\blame{\system, \conf}$ denotes the module that $\conf$ blames.
 \end{quote}
 
-When the buggy untyped component of a program is replaced by the typed counterpart,
-the type checker fails because this component causes the impedance mismatch. Hence, a
+When the buggy untyped module of a program is replaced by the typed counterpart,
+the type checker fails because this module causes the impedance mismatch. Hence, a
 trail that ends at an ill-typed scenario successfully pinpoints the location of
@@
 That is, failing Natural blame trails are those that end in a scenario that does not reveal the bug statically, yet also does not blame
-an untyped component. Thus the rational programmer gets
+an untyped module. Thus the rational programmer gets
 no hints from the gradual type system on how to continue the search for the bug.

### 2021-03-03 writing/rational-stuff.tex
touch ups for consistent use of mutant/mutated module and module/component

@@
 Applying a mutator to module $\component^*$ of $\system$ acts like a homomorphism
-on the generated lattice. The two lattices differ only in the mutant of
+on the generated lattice. The two lattices differ only in the mutated
 $\component^*$. Given this formulation, debugging scenarios are those points in
-this new lattice that do not contain the mutant; at all other points, the type
-checker points out the type-level mistake in the mutant of $\component^*$.
+this new lattice that do not contain the mutated module; at all other points, the type
+checker points out the type-level mistake in the mutated $\component^*$.
 
@@
 the rational programmer runs the program until it raises a run-time error.  The rational programmer
-uses the information in the error to decide which component to equip with
+uses the information in the error to decide which module to equip with
 types. This choice constructs scenario $\conf_{i+1}$ from $\conf_{i}$ and thus
 creates a trail. The trail's construction ends when it reaches a scenario that
-contains the mutant module because the type checker rejects its typed version
+contains the mutated module because the type checker rejects its typed version
 outright.  At this point, the source of the impedance mismatch is identified.

### 2021-03-03 writing/rational-transient.tex
touch ups for consistent use of mutant/mutated module and module/component

@@
 
-The Transient semantics assigns blame to a set of components. The blame
+The Transient semantics assigns blame to a set of modules. The blame
 assignment says that the value witnessing the impedance mismatch may have
@@
 is that the rational programmer has at least two reasonable options. The first
-one is to select the component that is added to the blame set first and
+one is to select the module that is added to the blame set first and
 assign types to only that one---after all, if fully checked, the types of this
-first component should be able to detect an impedance mismatch earlier in the
+first module should be able to detect an impedance mismatch earlier in the
 evaluation of a program than the later ones. The second option is to select the
-component that is added to the blame set last, effectively interpreting the
+module that is added to the blame set last, effectively interpreting the
 blame set as a boundary-aware stack.
@@
 \first{\mblame{\system, \conf_i}}$ where $\first{\mblame{\system, \conf}}$ is the
-first component Transient adds to the blame set for $\conf$.
+first module Transient adds to the blame set for $\conf$.
 \end{quote}
@@
   \last{\mblame{\system, \conf_i}}$ where $\last{\mblame{\system, \conf}}$ is the
-last component Transient adds to the blame set for $\conf$.
+last module Transient adds to the blame set for $\conf$.
 \end{quote}

### 2021-03-03 writing/results.tex
touch ups for consistent use of mutant/mutated module and module/component

@@
 (40), Natural produces a run time type error blaming a non-buggy
-already-typed component. All these cases are due to known open issues with Typed
+already-typed module. All these cases are due to known open issues with Typed
 Racket and class contracts.
@@
 the left side of the plot, meaning that in most cases the programmer needs
-to type a single component to debug a scenario.
+to type a single module to debug a scenario.
 

### 2021-03-03 writing/rational-experiment.tex
'the experiment' does _not_ 'do'

@@
 blame for Natural, Transient and Erasure. In line with the discussion so far,
-the experiment collects data to answer three initial questions for 
+the test bed collects data to answer three initial questions for 
 interesting debugging scenarios:
@@
 first step is to construct each mutant's scenario lattice and identify their
-debugging scenarios.  The experiment then extends the trails that start
+debugging scenarios.  The test bed then extends the trails that start
 from those roots according to the Natural-blame programmer.  If no scenarios can
-be added to the trail, the experiment checks whether the last scenario
+be added to the trail, the test bed checks whether the last scenario
 of the trail type-checks or not. If it does, the
@@
 section~\ref{sec:mutate}.
-The experiment repeats the process for the same roots with the other mode, Natural
+The process is repeated for the same roots with the other mode, Natural
 exceptions.
-After completing, the experiment
+After completing, the test bed
 reports the success/failure results of the trails to determine the proportion of
@@
   \includegraphics[scale=0.36]{./Images/process}
-  \caption{The experimental process for one mode of the rational
-  programmer}
+  \caption{The experimental process for one mode of the rational programmer}
   \label{fig:process}

### 2021-03-03 writing/challenges.tex
dangling lines

@@
 
-Besides the run-time checks, the usefulness of blame might be due to pure luck.
-Indeed, a proper evaluation of blame needs to consider the role of chance in blame assignment, too.
-A random rational programmer mode provides yet another necessary baseline using the same ingredients as for exceptions.
+An experiment must rule out that the usefulness of blame assignment is sheer
+luck.  Hence, a random rational programmer mode provides yet another necessary
+baseline using the same ingredients as for exceptions.
 

### 2021-03-03 writing/mutate-benchmarks.tex
touch ups in section 6
dangling lines

@@
 
-The bar for testing the rational programmer is high: it demands a
-collection of realistic programs that
-\begin{itemize}
+The benchmark programs for a rational-programmer experiment must 
+\begin{enumerate}
 
@@
 
-  \item  have a decent number of modules and a variety of module
-    dependency graphs (since mixing of typed and untyped code in Typed
-    Racket takes place at the module level);
+\item have a decent number of modules and a variety of module dependency graphs
+ (since mixing of typed and untyped code in Typed Racket takes place at the
+ module level);
 
@@
 \item  fail and contain type-level mistakes. 
-\end{itemize}
-
+\end{enumerate}
+\noindent
 No such benchmark suite exists.  Worse, no benchmark suite of
@@
 
-However, over the past decade,  \citet{gtnffvf-jfp-2019} have curated a
-collection of Typed Racket programs for systematically measuring the
-implementation's performance. The benchmark suite consists of fully typed,
-correct programs, written by a number of different authors, who have
-maintained and evolved these programs over several years. The programs
-range widely in size, complexity, purpose, origin and in programming
-style. The latter is mostly a reference to their reliance on many Typed
-Racket features: occurrence typing~\cite{tf-icfp-2010}, types for mutable
-and immutable data structures~\cite{hpst-sfp-2010}, types for first-class
-classes and objects~\cite{tsdtf-oopsla-2012}, and types for Racket's
+Fortunately, \citet{gtnffvf-jfp-2019} have curated a collection of Typed Racket
+programs for systematically measuring the implementation's performance. The
+benchmark suite consists of fully typed, correct programs, written by a number
+of different authors, who have maintained and evolved these programs over
+several years. The programs range widely in size, complexity, purpose, origin
+and in programming style. The latter is mostly a reference to their reliance on
+many Typed Racket features: occurrence typing~\cite{tf-icfp-2010}, types for
+mutable and immutable data structures~\cite{hpst-sfp-2010}, types for
+first-class classes and objects~\cite{tsdtf-oopsla-2012}, and types for Racket's
 numeric tower~\cite{stathff-padl-12}.

### 2021-03-03 writing/mutate-interesting.tex
touch ups in section 6
dangling lines

@@
 
-\item By definition, an impedance mismatch is a clash between the type ascription of one module's imports and another module's exports. Hence, type checking should fail for an interesting mutant.
+\item An impedance mismatch is a clash between the type ascription of one
+module's imports and another module's exports. Hence, type checking should fail
+for an interesting mutant.
 
@@
 
-\item If the evaluation of a mutant module immediately raises an exception because
+\item If the evaluation of a mutated module immediately raises an exception because
 of the changes, there is no work for the rational programmer. Indeed, if the

### 2021-03-03 writing/results.tex
dangling lines

@@
 a rational programmer mode on a scenario may result in an exception
-that carries no useful information about which module the programmer
-should type next. For instance the stack trace of the exception may not
+that carries no useful information about which module to equip with types next. For instance the stack trace of the exception may not
 contain frames from any untyped module of the program. This can happen at any point
@@
 no clear winner between Natural exceptions and Transient exceptions
-despite the theoretically advantageous additional checks of the Natural
-semantics.
+despite the theoretically advantageous additional checks of Natural.
 

### 2021-03-03 writing/why-rational-methodology.tex
touch ups in section 3
dangling lines

@@
 
-Every time the rational programmer debugs a program, it is validation for
-programming language researchers. It shows how their slogans and theorems help
-programmers connect the dots. The example of blame assignment mechanisms makes
-this point clearly. A blame-assignment mechanism provides information that,
-according to programming language research, points toward the source of the
-problem.
+Every time the rational programmer succeeds, it is validation for programming
+language researchers. It shows how their theorems and slogans help
+programmers>. The example of blame assignment mechanisms makes this point
+clearly. A blame-assignment mechanism provides information that, according to
+programming language research, points toward the source of the problem.
 
@@
 
-\item Does a language design provide information that guides the actions of the
+\item Does a language design provide information that can guide a
  rational programmer?
@@
 
-Here is how the answers to these methodological questions imagines an evaluation
+Here is how the answers to these methodological questions lead to an evaluation
 method for blame-assignment mechanisms:
@@
 \end{enumerate}
-
 That said, using the method to conduct data-gathering experiments entails

### 2021-03-04 writing/challenges.tex
uniform editorial voice, mistakes, typos, size of text in figures, captions

@@
 
-While at first glance the method of section~\ref{sec:why-rational} seems straightforward to implement and test,
-three challenges remain. The first is comparing the effect of blame on the rational
-programmer across three different mechanisms; second is the collection of a large number
-of representative debugging scenarios; and third is finding a proper sampling strategy for the
+While at first glance the method of section~\ref{sec:why-rational} seems
+straightforward to implement and test, three challenges remain. The first is
+comparing the effect of blame on the rational programmer across three different
+mechanisms; second is the collection of a large number of representative
+debugging scenarios; and third is finding a proper sampling strategy for the
 huge space of possibilities. A coincidental challenge is the disparity of the
@@
 authors use Racket, which is thus far the only language in which all three major
-soundness mechanisms are available in a robust and comparable manner;
-for the implementation of Transient Typed Racket, see the parallel submission
-of an experience report~\cite{ttt21}.
+soundness mechanisms are available in a robust and comparable manner; for the
+implementation of Transient Typed Racket, see the parallel submission of an
+experience report~\cite{ttt21}.
 
@@
 assignment mechanisms of the three semantic variants.  While Natural assigns
-blame to {\em one\/} component, Transient assigns blame to a set of
+blame to {\em one\/} component, Transient assigns blame to a sequence of
 components. The Erasure semantics does not blame components {\it per se\/}, but
@@
 blame.  For instance, one Transient mode may pick the oldest element of a
-blame set as the next component to add types because it corresponds to the
+blame sequence as the next component to add types because it corresponds to the
 earliest point in the program's evaluation that can discover an impedance
-mismatch.  Another mode may opt to treat the set as a stack and add
+mismatch.  Another mode may opt to treat the sequence as a stack and add
 types to its newest element.  If both modes are equally successful in
@@
 information in the exception from the run-time check. Nevertheless, the
-exception information may be equally or even more useful to the rational
-programmer than blame. If this is the case, then blame may not play a critical
+exception information may be as useful to the rational
+programmer as a blame assignment. If this is the case, then blame may not play a critical
 role for the rational programmer. Indeed, precisely because they do not
 account for such confounding factors, \citet{lksfd-popl-2020} cannot draw any
-conclusions about blame specifically despite advertisements for the
-opposite. Their experiment can conclude that so-called blame-shifting
-works but not that this is because of blame.
+conclusions about blame specifically, despite advertisements for the
+opposite. Their experiment may conclude only that so-called blame-shifting
+works, but they cannot attribute this conclusion to blame.
 
@@
 Transient rational programmer come with a blame mode and an exception
-mode. If the former succeeds debugging a program while the latter
+mode. If the former succeeds in debugging a program while the latter
 fails, then it is safe to conclude that blame is indeed necessary for the
@@
 mutation analysis frameworks are fine-tuned to avoid them, and yet, it is
-precisely those that are needed for evaluating blame assignment strategies.
+precisely those mutators that are needed for evaluating blame assignment strategies.
 
@@
 
-Thus, an experimental analysis of blame needs a mostly new set of mutators.
+An experimental analysis of blame needs a mostly new set of mutators.
 Roughly speaking, the new mutators inject type errors into fully typed programs.
 Applying such a mutator to any typed component produces a mutated component.  A
-debugging scenario results from removing the types from the mutated component. For the
-design of such mutators, the authors relied on their own extensive programming
-experience though not without discovering a major pitfall: some of their
-original mutators systematically produced programs that immediately revealed the
-source of the impedance mismatch. About four fifths of the original mutators
-survived this first scrutiny testing. All of the remaining ones yield {\em
-interesting debugging scenarios\/} (see section~\ref{sub:mutate-interesting}).
+debugging scenario results from removing the types from the mutated
+component. For the design of such mutators, the authors relied on their own
+extensive programming experience though not without discovering a major pitfall:
+some of their original mutators systematically produced programs that
+immediately revealed the source of the impedance mismatch.  All of the remaining
+ones yield {\em interesting debugging scenarios\/} (see
+sec.~\ref{sub:mutate-interesting}).
 

### 2021-03-04 writing/discussion.tex
dangling lines
uniform editorial voice, mistakes, typos, size of text in figures, captions

@@
 The results in section~\ref{sec:results} show that Transient produces empty
-blame sets for 967 scenarios. In theory, empty set must not occur. An
-empty blame set means a lack of boundary crossings for the witness value.
+blame sequences for 967 scenarios. In theory, empty set must not occur. An
+empty blame sequence means a lack of boundary crossings for the witness value.
 By implication, a typed module is blaming itself---something that can
@@
 
-{\bf Note} The scenarios where Transient returns empty blame 
-suggest ways to improve Transient's blame map as follows:
+{\bf Note} The scenarios where Transient returns no blame 
+suggest ways to improve its blame map:
 \begin{itemize}

### 2021-03-04 writing/introduction.tex
uniform editorial voice, mistakes, typos, size of text in figures, captions

@@
+%% ------------------------------------------------------------------------------
+
 Theoreticians of gradual typing have focused on blame theorems from the
@@
 sec.~\ref{sec:landscape}). Their academic creators embrace the idea that
-blame can help practicing programmers find impedance mismatches:
+blame can help practicing programmers find impedance mismatches, that is,
 disagreements between the type ascriptions of a software system and its
@@
 literature and the industrial blog world do not discuss any possible answers.
-Instead, when language designers make relevant decisions, they seem to answer it
-one way or another without any scientific justification. Then again, the
+Instead, when language designers make decisions concerning this aspect, they seem
+to go one way or another without any scientific justification. Then again, the
 community has thus far failed to offer a method for evaluating blame assignment.
@@
 latter---and this systematic, information-driven process can be implemented and
-tested at scale on real programs.  Turning this idea into a scientific
+tested, at scale, on real programs.  Turning this idea into a scientific
 experiment requires overcoming major challenges: injecting representative
@@
 applying the evaluation method to three distinct blame assignment
-strategies in approximately 72,200 different scenarios\/} (see
+strategies in approximately 72,000 different scenarios\/} (see
 sec.~\ref{sec:results}): (1)~{\it Transient\/}, i.e. Reticulated's
@@
 Natural\/} vastly more useful than {\it Transient\/} nor are the two
-blame assignment methods always superior to {\it Erasure\/}.  The second
+blame assignment methods highly superior to {\it Erasure\/}.  The second
 one is that the cost of Transient's blame can be huge.  In turn, these

### 2021-03-04 writing/landscape.tex
uniform editorial voice, mistakes, typos, size of text in figures, captions

@@
 among others, a function {\tt pack}. As the documentation says, this function
-consumes JSON data and packages it in a dictionary. By mistake, though, the
+consumes JSON data and packages it in an association list (like a dictionary). By mistake, though, the
 function returns a hash table instead of the association list its documentation
@@
 the run-time system checks that it is a function and wraps it in a protective
-proxy, which in turn, enforces the type of the function with run-time checks.
-In essence, these checks protect typed modules when they call an untyped function
-because it may not produce the promised kind of value, as is the case here with
-{\tt pack} from {\tt pack-lib}. Analogously the run-time system wraps each exported
-function of a typed module (such as {\tt crypto-pack}) in a proxy that checks
-its arguments.  These checks protect functions exported from typed modules against
-applications in untyped code that supply the wrong kind of arguments.
+proxy, which in turn, enforces the type of the function arguments with run-time checks.
+Analogously the run-time system wraps each exported function of a typed module
+(such as {\tt crypto-pack}) in a proxy that checks its arguments.  These checks
+protect functions exported from typed modules against applications in untyped
+code that supply the wrong kind of arguments.
 
@@
 \texttt{pack} returns, the return-type check fails and Natural blames
-\texttt{pack-lib}.
+\texttt{pack-lib} for both {\tt define}s in {\tt client}.
 
@@
 list, array, hash table etc.) is performed via a function call, the contents of a
-complex value are checked on a piecemeal basis.
+complex value is checked on a piecemeal basis.
 
@@
 work of~\citet{gfd-oopsla-2019} shows that for some programs Transient
-constructs a blame set that excludes responsible parties and includes modules
+constructs a blame sequence that excludes responsible parties and includes modules
 irrelevant to the failing check.

### 2021-03-04 writing/mutate-benchmarks.tex
uniform editorial voice, mistakes, typos, size of text in figures, captions

@@
 \item have a decent number of modules and a variety of module dependency graphs
- (since mixing of typed and untyped code in Typed Racket takes place at the
+ (because mixing of typed and untyped code in Typed Racket takes place at the
  module level);
 
-\item attempt to take  advantage of the sophisticated typing features of
- a gradually typed language such as Typed Racket but;
+\item take advantage of the variety of typing features of a gradually typed
+ language; but
 
@@
 \end{enumerate}
-\noindent
 No such benchmark suite exists.  Worse, no benchmark suite of
@@
 
-Fortunately, \citet{gtnffvf-jfp-2019} have curated a collection of Typed Racket
-programs for systematically measuring the implementation's performance. The
-benchmark suite consists of fully typed, correct programs, written by a number
-of different authors, who have maintained and evolved these programs over
-several years. The programs range widely in size, complexity, purpose, origin
-and in programming style. The latter is mostly a reference to their reliance on
-many Typed Racket features: occurrence typing~\cite{tf-icfp-2010}, types for
-mutable and immutable data structures~\cite{hpst-sfp-2010}, types for
-first-class classes and objects~\cite{tsdtf-oopsla-2012}, and types for Racket's
-numeric tower~\cite{stathff-padl-12}.
+What exists is \citet{gtnffvf-jfp-2019}'s collection of Typed Racket programs
+for systematically measuring the implementation's performance. The benchmark
+suite consists of fully typed, correct programs, written by a number of
+different authors, who have maintained and evolved these programs over several
+years. The programs range widely in size, complexity, purpose, origin and in
+programming style. The latter is mostly a reference to their reliance on many
+Typed Racket features: occurrence typing~\cite{tf-icfp-2010}, types for mutable
+and immutable data structures~\cite{hpst-sfp-2010}, types for first-class
+classes and objects~\cite{tsdtf-oopsla-2012}, and types for Racket's numeric
+tower~\cite{stathff-padl-12}.
 
-Hence, even if they do not come with type-level errors, they offer a
-diverse linguistic substrate suitable for the careful introduction of
-type-level errors via mutations. Further filtering the benchmark suite to
-keep the benchmarks with the most sizeable and complex dependency graphs
-yields ten suitable programs while preserving the suite's linguistic
-diversity. Figure~\ref{table:benchmark-descriptions} displays the selected
+Hence, even if they do not come with type-level errors, they offer a diverse
+starter suite, suitable for the careful introduction of type-level errors via
+mutations. Further filtering the benchmark suite to keep the benchmarks with the
+most sizeable and complex dependency graphs yields ten suitable programs while
+preserving the suite's linguistic diversity. Figure~\ref{table:benchmark-descriptions} displays the selected
 benchmark programs and describes their basic characteristics.

### 2021-03-04 writing/mutate-interesting.tex
uniform editorial voice, mistakes, typos, size of text in figures, captions

@@
 other semantics is unavoidable. Tipping the scales in favor of the theoretically
-weakest semantics yields the most stable results. See
-section~\ref{sec:discussion} for some further discussion of this choice.
+weakest semantics yields the most stable results. 
+Section~\ref{sec:discussion} includes some further discussion of this choice.
 
@@
 
-\caption{The mutators produce Interesting debugging scenarios across all benchmarks.} 
+\caption{The generated debugging scenarios across all benchmarks} 
  \label{fig:mutant-breakdown}
@@
 
-The definition of interesting mutants is more than a filter for
-uninteresting mutants. It is also the guide for evaluating and refining
-mutators until they collectively produce a sizable and diverse population of
-interesting mutants. For example, one early mutator idea casts the
-tests of conditional expressions to the {\tt Any} type to
-suppress occurrence typing. However, an examination of the mutants it
-produces through the lens of the definition of interesting mutants
-reveals that none is interesting because they never raise run-time errors.
-Indeed, the interesting standard guided countless iterations of adding,
-removing, and refining the mutators in figure~\ref{table:mutation-ops}
-to maximize interesting mutants.
+The definition of interesting mutants is more than a filter for uninteresting
+mutants. It is also the guide for evaluating and refining mutators until they
+collectively produce a sizable and diverse population of interesting
+mutants. Consider a candidate for the set of mutators that casts the tests of conditional
+expressions to the {\tt Any} type, which would suppress occurrence typing. An
+examination of its mutants through the lens of ``interesting'' reveals that none
+of them is interesting, because they never raise run-time errors. Hence Indeed,
+the interesting standard guided countless iterations of adding and removing
+mutators and of refining the chosen mutators in figure~\ref{table:mutation-ops}.

### 2021-03-04 writing/mutate-mutators.tex
uniform editorial voice, mistakes, typos, size of text in figures, captions

@@
 
+%% MF: I don't see what this contributes:
+% \footnote{Many of~\citet{lksfd-popl-2020} mutators do not satisfy the
+% constraints and had to be eliminated or significantly adapted. For example, the
+% relational mutator does not lead to type-level errors. Similarly, the
+% arithmetic mutator does not reliably lead to type errors; it had to be adapted
+% to make only operator swaps that can affect the type of the result, e.g.,
+% changing `*` to `/`.}
+
 Figure~\ref{table:mutation-ops} describes 16 mutators that satisfy these
-constraints. As the explanation indicates, some specialize or
+constraints. As the last column indicates, some specialize or
 generalize~\citet{lksfd-popl-2020}'s mutators, which in turn are borrowed from
-the vast literature on mutation testing~\cite{jia2011analysis}.\footnote{Many
-of~\citet{lksfd-popl-2020} mutators do not satisfy the constraints and had to
-be eliminated or significantly adapted. For example, the relational mutator does not lead to type-level
-errors. Similarly, the arithmetic mutator does not reliably lead to type
-errors; it had to be adapted to make only operator swaps that can affect the
-type of the result, e.g., changing `*` to `/`.}  Only two are directly
-inherited; many mutators are brand new. For the latter the authors relied on
-their decades-long experience of making type-level mistakes in Typed Racket,
-some of which take non-trivial effort to debug.
+the vast literature on mutation testing~\cite{jia2011analysis}.  Only two are
+directly inherited; many mutators are brand new. For the latter, the authors
+relied on their decades-long experience of making type-level mistakes in Typed
+Racket, some of which take non-trivial effort to debug.
 

### 2021-03-04 writing/mutate.tex
uniform editorial voice, mistakes, typos, size of text in figures, captions

@@
 develop new mutators (see sec.~\ref{sub:mutate-mutators}) and to validate their
-suitability for this application (see sec.~\ref{sub:mutate-interesting}).
+suitability on the benchmarks (see sec.~\ref{sub:mutate-interesting}).
 

### 2021-03-04 writing/paper.tex
re-established the old organization
editiroal voice thru 6
pass over challenges section
adjusting title for section 7
uniform editorial voice, mistakes, typos, size of text in figures, captions

@@
 method for blame assignment strategies and (2) the results from applying
-it to three different gradual type systems. These results confirm that blame assignment
+it to three different gradually typed languages. These results confirm that blame assignment
 is helpful to a rationally acting programmer.
@@
 
+\sec{introduction}{Does Blame Matter}
 \input{landscape-program}
-\sec{introduction}{Does Blame Matter}
 \sec{landscape}   {How to Think About Three Blame Systems}
 \sec{why-rational}{What Is a Rational Programmer}
-\sec{challenges}  {Why It Is Hard to Implement the Rational Programmer}
+\sec{challenges}  {Why It Is Hard to Evaluate Blame}
+\sec{rational}    {How to Make Comparable Rational Programmers}
 \sec{mutate}      {How to Make Lots of Mistakes}
-\sec{rational}    {How to Make Comparable Rational Programmers}
-\sec{sample}      {How to Make the Experimental Space}
-\sec{results}     {What We Discovered}
-\sec{discussion}  {What We Learned}
+\sec{sample}      {How to Make the Experimental Space Tractable}
+\sec{results}     {What Are the Outcomes of the Experiment}
+\sec{discussion}  {What Can Programmers Learn}
 \sec{related}     {Who has Worked on this Problem Before}

### 2021-03-04 writing/rational-effort.tex
dangling lines
uniform editorial voice, mistakes, typos, size of text in figures, captions

@@
 In addition to success and failure information for the various trails, the
-experiment records the number of modules a rational programmer has
+experimental test bed can record the number of modules a rational programmer has
 to equip with types along each trail ($\lvert \conf_n \setminus \conf_0
@@
 Measuring effort can also reveal whether the observed effectiveness of the
-rational programmer is an artifact of pure chance or not.
+rational programmer is an artifact of pure chance.
 In particular, the effort distribution for a mode of the rational programmer can be compared with that of
@@
 modules, random mode trails are always successful. However, the random mode's effort
-distribution will be thinly spread out across the range of trail lengths possible
+distribution should be thinly spread out across the range of trail lengths possible
 in the set of debugging scenarios.

### 2021-03-04 writing/rational-experiment.tex
uniform editorial voice, mistakes, typos, size of text in figures, captions

@@
 
-Table~\ref{fig:experiment-outline} summarizes how each question relates to
-different kinds of trails/modes of the rational programmer. For example, experimental
-question $Q_1$ asks whether blame is valuable for Natural and the experiment
-uses the Natural blame and exception trails to answer it, so $Q_1$ shows up in the cells for Natural blame and Natural exceptions.
-
-\begin{figure}[ht]
+\begin{figure}[ht] \footnotesize
 \center
@@
 \end{tabular}}
-  \caption{ Experimental questions and rational programmer modes.}
   \label{fig:experiment-outline}
 \end{figure}
+
+The above table summarizes how each question relates to
+different kinds of trails/modes of the rational programmer. For example,
+experimental question $Q_1$ asks whether blame is valuable for Natural and the
+experiment uses the Natural blame and exception trails to answer it, so $Q_1$
+shows up in the cells for Natural blame and Natural exceptions.
 
@@
 
-\begin{figure}[bh]
+\begin{figure}[bh] \footnotesize
   \centering

### 2021-03-04 writing/rational-natural.tex
dangling lines
uniform editorial voice, mistakes, typos, size of text in figures, captions

@@
 interface exposes a bug in the untyped module (or its dependencies). Our setup rules out the first
-alternative (but see section~\ref{sec:conclusion}), and therefore the rational
+alternative (but see sec.~\ref{sec:conclusion}), and therefore the rational
 programmer extends the trail to a scenario that swaps out the untyped
@@
 
-Formally,
+% Formally, %% <-- MF: this is called a ``dangling def.'' and is considered bad style
+
+Here is a rigorous definition. 
 \begin{quote}
@@
 To account for this situation, a new mode of the Natural rational
-programmer follows a migration process based on exceptions:
+programmer follows a migration process based on exceptions.
 \begin{quote}

### 2021-03-04 writing/rational-transient.tex
dangling lines
uniform editorial voice, mistakes, typos, size of text in figures, captions

@@
 
-The Transient semantics assigns blame to a set of modules. The blame
+The Transient semantics assigns blame to a sequence of modules. The blame
 assignment says that the value witnessing the impedance mismatch may have
@@
 is that the rational programmer has at least two reasonable options. The first
-one is to select the module that is added to the blame set first and
+one is to select the module that is added to the blame sequence first and
 assign types to only that one---after all, if fully checked, the types of this
@@
 evaluation of a program than the later ones. The second option is to select the
-module that is added to the blame set last, effectively interpreting the
-blame set as a boundary-aware stack.
+module that is added to the blame sequence last, effectively interpreting the
+blame sequence as a boundary-aware stack.
 
@@
 \begin{quote}
-\it A \emph{Transient-first blame trail} is a sequence of scenarios
+\it A \emph{Transient-first blame trail}/\emph{Transient-last blame trail} is a sequence of scenarios
 $\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
 $\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
-\first{\mblame{\system, \conf_i}}$ where $\first{\mblame{\system, \conf}}$ is the
-first module Transient adds to the blame set for $\conf$.
+\first{\mblame{\system, \conf_i}}$ where $\first{\mblame{\system,
+\conf}}$/$\last{\mblame{\system, \conf}}$ is the
+first/last module, respectively,  that Transient adds to the blame sequence for $\conf$.
 \end{quote}
 
-\begin{quote}
-\it A \emph{Transient-last blame trail} is a sequence of scenarios
-$\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
-$\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
-  \last{\mblame{\system, \conf_i}}$ where $\last{\mblame{\system, \conf}}$ is the
-last module Transient adds to the blame set for $\conf$.
-\end{quote}
+% \begin{quote}
+% \it A  is a sequence of scenarios
+% $\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
+% $\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
+%   \last{\mblame{\system, \conf_i}}$ where $\last{\mblame{\system, \conf}}$ is the
+% last module Transient adds to the blame sequence for $\conf$.
+% \end{quote}
 
 The definition of \emph{Transient exception trails} is analogous to that for
-Natural. It is used to specify when Transient-first and Transient-last blame are
-more useful than Transient exceptions for debugging a scenario.
+Natural. It is used to as a baseline for Transient-first and Transient-last
+blame.

### 2021-03-04 writing/rational.tex
uniform editorial voice, mistakes, typos, size of text in figures, captions

@@
 
-Section~\ref{sec:why-rational} explains, and illustrates with an example,
-how a migratory type setting helps with finding the source of an impedance
-mismatch. Roughly speaking, it encourages the rational programmer to
-equip a module with types if it is blamed in an error message. A sequence
-of such steps make up a portion of the migratory path in the lattice of
-type migration~\cite{tfgnvf-popl-2016}. The lattice, described in Section
-~\ref{sub:stuff},  serves as the common substrate for the definition modes
-of the rational programmer mentioned in section~\ref{sec:challenges}.
+Section~\ref{sec:why-rational} explains how a migratory type setting helps with
+finding the source of an impedance mismatch. Roughly speaking, it encourages the
+rational programmer to equip a module with types if it is blamed in an error
+message. A sequence of such steps make up a portion of the migratory path in the
+lattice of type migration~\cite{tfgnvf-popl-2016}. The lattice serves as the
+common substrate for the definition modes of the rational programmer (see
+sec.~\ref{sub:stuff}).
 
@@
 
-Hence the research problem is how to make the modes of the rational programmer
-comparable even when they correspond to different semantics and offer different kinds of information. Sections~\ref{sub:natural}
-through~\ref{sub:erasure} describe the various modes of the rational programmer and explain
-the extent to which they are comparable. Section~\ref{sub:experiment}
-enhances their comparability with the programmer-effort dimension. Finally, the last
-subsection states the experimental questions and gives an overview of the
-experimental process.
+Hence the research problem is develop modes for the rational programmer and to make
+them comparable even when they correspond to different semantics and offer
+different kinds of information (see secs.~\ref{sub:natural}
+through~\ref{sub:erasure}). Programmer effort relative to a fixed debugging
+scenario introduces another dimension along which the modes become may be
+compared (see sec.~\ref{sub:effort}). With these notions in place, it
+becomes possible to state the experimental questions and gives an overview of
+the experimental process (see sec.~\ref{sub:experiment}).
 

### 2021-03-04 writing/results.tex
uniform editorial voice, mistakes, typos, size of text in figures, captions

@@
 
-A machine with two Intel Xeon Gold 6258R processors (28 doubly-threaded cores each)
-and 500GB of memory ran the experimental process from section~\ref{sec:rational} on the
-sampled debugging scenarios from section~\ref{sec:sample}.  Each debugging scenario had a 4 minute timeout and a
-memory limit of 6GB. The complete experiment required over 30,000 compute
-hours.
-
-\begin{wrapfigure}{r}{.40\textwidth}
-  \includegraphics[width=0.40\textwidth]{./plots/success-bars}
-
+\begin{wrapfigure}{r}{.40\textwidth} \footnotesize 
+  \vspace{-1.5em}
+  \includegraphics[width=0.38\textwidth]{./plots/success-bars}
   \vspace{1em}
-  \begin{minipage}{0.4\textwidth}
-  The estimated percentage of scenarios for which each mode succeeds in locating the bug.
+  \begin{minipage}{0.35\textwidth} \raggedright
+%   The estimated percentage of scenarios for which each mode succeeds in locating the bug.
    The upper bound margin of error is 0.02\%.
   \end{minipage}
-
-  \caption{Success rates.}
-  \label{fig:success-bars}
+\vspace{-2em}
+  \caption{Percentage rates of success} \label{fig:success-bars}
 \end{wrapfigure}
 
-\begin{figure}
+The test bed for executing the experimental process utilizes a machine with two
+Intel Xeon Gold 6258R processors (28 doubly-threaded cores each) and 500GB of
+memory.  Each debugging scenario had a 4 minute timeout and a 6GB memory
+limit. Running the experiment on all debugging scenarios took over
+30,000 compute hours.
+
+\begin{figure} \footnotesize 
   \centering
@@
   \begin{minipage}{0.95\textwidth}
-  Each plot depicts a head-to-head comparison of the entitled mode against every other mode.
-  The portion above 0 (in green) is the estimated percentage of scenarios where the entitled mode is more useful
-    than the other mode.
-  The portion below 0 (in red) is the estimated percentage of scenarios
-    where the entitled mode is less useful than the other mode.
-  The upper bound margin of error is 0.02\%.
+      Each plot depicts a head-to-head comparison of the mode named above the
+      plot vs. every other mode.  The (green) portion above 0 is the estimated
+      percentage of scenarios where the named mode is more useful than the
+      other.  The (red) portion below 0 is the estimated percentage of scenarios
+      where the named mode is less useful than the other.  The upper bound
+      margin of error is 0.02\%.
   \end{minipage}
 
-  \caption{Usefulness comparisons.}
+  \caption{Usefulness comparisons}
   \label{fig:avo-bars}
@@
 
-\begin{figure}
+\begin{figure}\footnotesize 
   \centering
@@
   \begin{minipage}{0.95\textwidth}
+
      Each diagram shows the overlap of the successful scenarios for three modes.
-     For example, in the leftmost diagram, all three modes succeed on the
-    same scenario 57.3\% of the time, only Natural and Natural exceptions
-    succeed on 29.1\% of the scenarios, only Natural and Erasure succeed
-    on 2.1\%, and Natural alone succeeds on 9\%.  The upper bound margin
-    of error is 0.02\%.
+     For example, in the leftmost diagram, all three modes succeed on the same
+     scenario 57.3\% of the time, only Natural and Natural exceptions succeed on
+     29.1\% of the scenarios, only Natural and Erasure succeed on 2.1\%, and
+     Natural alone succeeds on 9\%.  \\ The upper bound margin of error is 0.02\%.
+
   \end{minipage}
 
-  \caption{Blame usefulness analysis.}
+  \caption{Blame usefulness analysis}
   \label{fig:success-venns}
@@
 
+% \footnote{The random mode behaves the same for all three semantics, so this analysis considers it a single mode.}
+
 Figure~\ref{fig:success-bars} summarizes the overall success rates of
-every mode.\footnote{The random mode behaves the same for
-all three semantics, so this analysis considers it a single mode.}
+every mode.
 The success rates illustrate a few points that underlie the
@@
 
-In Transient, similar to Natural, most failures are due to unhelpful
-exception information (1851 for both Transient first and last blame).
-However, Transient also has a substantial number of failures because
-scenarios hit the time and memory limits of the experiment
-(\textasciitilde770 scenarios).  Additionally, there are nearly 1,000
-cases where Transient reports an empty blame set, leaving the rational
-programmer without hints about how to proceed.
-Sections~\ref{sec:threat:transient} and ~\ref{sec:threat:transient2}
-explain both of these causes of failure for Transient in detail and how
-they affect the
-experiment.
+In Transient, similar to Natural, most failures are due to unhelpful exception
+information (1851 for both Transient first and last blame).  However, Transient
+also has a substantial number of failures because scenarios hit the time and/or
+memory limits of the experiment (\textasciitilde770 scenarios).  Additionally,
+there are nearly 1,000 cases where Transient reports an empty blame set, leaving
+the rational programmer without hints about how to proceed.
+Sections~\ref{sec:threat:transient} and ~\ref{sec:threat:transient2} address
+these causes of failure for Transient and how they affect the experiment.
 
@@
 better or worse than each other mode.  In particular, each bar above zero
-represents the proportion where the plot's entitled mode succeeds and the
+represents the proportion where the plot's named mode succeeds and the
 mode on the x-axis fails; the corresponding bar below zero represents the
@@
 
-\begin{figure}
-  \centering
+\begin{figure} \footnotesize \centering
   \includegraphics[width=\textwidth]{./plots/bt-lengths-table}
@@
 
-  \caption{Programmer effort.}
-  \label{fig:effort-table}
+  \caption{Programmer effort} \label{fig:effort-table}
 \end{figure}
 
-\begin{figure}
-  \centering
+\begin{figure}\footnotesize \centering
   \includegraphics[width=\textwidth]{./plots/bt-length-comparisons}
@@
   \begin{minipage}{0.95\textwidth}
+
     Each plot depicts the distribution of scenarios with trail length
-    differences ranging from -3 to 3.  A -x difference denotes that
-    the first mode's trail is x steps shorter than the second mode's
-    trail for the same scenario, and a positive difference denotes the
-    inverse.  Only scenarios where both modes succeed are considered.  The
-    10|85 on the y-axis indicates that the axis is truncated between 10
-    and 85\%.
+    differences ranging from -3 to 3.  A -x difference denotes that the first
+    mode's trail is x steps {\em shorter\/} than the second mode's trail for the
+    same scenario; a positive difference denotes the inverse.  Only
+    scenarios where both modes succeed are considered.  The 10|85 on the y-axis
+    indicates that the axis is truncated between 10 and 85\%.
+
   \end{minipage}
 
-  \caption{Effort comparisons.}
-  \label{fig:effort-comparisons}
+  \caption{Effort comparisons} \label{fig:effort-comparisons}
 \end{figure}
 
-In order to better understand the answers the experiment provides for
-questions $Q_1$ to $Q_3$, figure~\ref{fig:success-venns} depicts an
-analysis of the success of each semantics in comparison to Erasure.
+An alternative way to understand the answers for questions $Q_1$ to $Q_3$, is to
+analyze the success of each semantics in comparison to
+Erasure. Figure~\ref{fig:success-venns} depicts the results of this analysis.
 Specifically, the figure shows one Venn diagram per mode of the rational
 programmer that uses blame.  Each diagram shows the overlap of successful
-scenarios for  the blame mode, its corresponding exception mode, and Erasure.  For
-example, the leftmost diagram (Natural) shows that all three modes
-succeed on 59.5\% of the  scenarios, only Natural and Natural exceptions
-succeed on 25.7\% of the scenarios, only Natural and Erasure succeed on
-2.1\%, and Natural alone succeeds on 9.7\%.  This analysis highlights
-the success trade-offs each semantics offers against Erasure, with and
-without blame. For instance, the analysis
-for Natural clearly illustrates that, when choosing between Natural blame,
-Natural exceptions, and Erasure, Natural blame is the absolutely most
-successful: all of the successes of the other two modes are subsets of
-Natural's successful scenarios.  On the other hand, Transient's blame
-modes fare similarly but the choice is not so clear-cut.
+scenarios for the blame mode, its corresponding exception mode, and Erasure.
+For example, the leftmost diagram (Natural) shows that all three modes succeed
+on 59.5\% of the scenarios, only Natural and Natural exceptions succeed on
+25.7\% of the scenarios, only Natural and Erasure succeed on 2.1\%, and Natural
+alone succeeds on 9.7\%.  This analysis highlights the success trade-offs each
+semantics offers against Erasure, with and without blame. For instance, the
+analysis for Natural clearly illustrates that, when choosing between Natural
+blame, Natural exceptions, and Erasure, Natural blame is the absolutely most
+successful: all of the successes of the other two modes are subsets of Natural's
+successful scenarios.  On the other hand, Transient's blame modes fare similarly
+but the choice is not so clear-cut.
 
@@
 
-A second point of interest is that the exceptions of the Erasure semantics
-either help the rational programmer immediately or the rational programmer
-fails to debug a scenario altogether (in red).  This is expected; adding
-type annotations to an Erasure program doesn't change it's behavior, so if
-the type checker doesn't reject the program with the new annotations, the
-rational programmer is stuck.  In other words, an exception from the
-runtime has to point to the buggy module with the first try. Otherwise the
-rational programmer types an irrelevant module, runs the program again,
-and the same exception points again to the already typed module.
+Second, the exceptions of the Erasure semantics either help the rational
+programmer immediately or the rational programmer fails to debug a scenario
+altogether (in red).  This is expected; adding type annotations to an Erasure
+program does not change its behavior, so if the type checker does not reject the
+program with the new annotations, the rational programmer is stuck.  In other
+words, an exception from the runtime has to point to the buggy module with the
+first try. Otherwise the rational programmer types an irrelevant module, runs
+the program again, and the same exception points again to the already typed
+module.
 
@@
 debug more scenarios but does not shorten the debugging process compared
-to exceptions. A second insight is that Natural relatively often (around 10\% of the
+to exceptions. Second, Natural relatively often (around 10\% of the
 scenarios) produces shorter trails than both Transient blame modes, and

### 2021-03-04 writing/sample.tex
touch ups in section 7
uniform editorial voice, mistakes, typos, size of text in figures, captions

@@
 
-As is, the chosen mutators generate a huge number of debugging scenarios:
-approximately 1 million for the chosen benchmarks. This number of scenarios is
-far too large to even identify the interesting ones among them.  Furthermore
-this population is heterogeneous; scenarios come from different
-mutants, and the mutants are the result of different mutators applied to each
-benchmark.
+As is, the chosen mutators generate approximately one million debugging
+scenarios for the chosen benchmarks. This number of scenarios is far too large
+to even identify the interesting ones among them.  Furthermore this population
+is heterogeneous; scenarios come from different mutants, and the mutants are the
+result of different mutators applied to each benchmark.
 
@@
 benchmark. Some benchmarks have less than  80 mutants with interesting scenarios,
-in which case the only choice is to take them all.
+in which case the only choice is to include them all.
 The result is a total of 756 interesting mutants across all benchmarks.
@@
 final sample thus consists of 72,192 interesting scenarios.
+
+%% MF: I don't get this number ^^^^
+% > (* 756 96)
+% 72576

### 2021-03-04 writing/why-rational.tex
uniform editorial voice, mistakes, typos, size of text in figures, captions

@@
 current research on programming languages, the question seems to call for
-empirical research similar to that of human-computer interaction. At the same
+empirical studies similar to those of human-computer interaction. At the same
 time, a significant evaluation result demands a large amount of
@@
 these insights as type annotations. Over time, the code base migrates into a mix
-of typed and untyped modules. As~\citet{tfffgksst-snapl-2017} reported, though,
-it is at least equally common that developers add typed modules that depend on
+of typed and untyped modules. As~\citet{tfffgksst-snapl-2017} report though,
+it is equally common that developers add typed modules that depend on
 the existing modules in the code base.
@@
 annotations. The result is the \typecolor\ module in the middle, and {\tt
-main}'s import is now re-directed there by {\tt (submod ".." layer-typed)}. Running the
-modified program yields a different error message:
+main}'s import is now re-directed there by {\tt (submod ".." layer-typed)}. As
+predicted by the theory, running the modified program yields a different error message:
 \begin{verbatim}

### 2021-03-04 writing/challenges.tex
issues with sections 9, 10, and 11, plus triggers for global searches

@@
 authors use Racket, which is thus far the only language in which all three major
-soundness mechanisms are available in a robust and comparable manner; for the
-implementation of Transient Typed Racket, see the parallel submission of an
-experience report~\cite{ttt21}.
+soundness mechanisms are available in a robust and comparable manner:
+Typed Racket implements Natural, Shallow Racket~\cite{ttt21}
+Transient, and plain Racket for Erasure.
 

### 2021-03-04 writing/conclusion.tex
touch ups for consistent use of mutant/mutated module and module/component
issues with sections 9, 10, and 11, plus triggers for global searches

--- 2021-03-03 writing/conclusion.tex
@@
 
-The paper provides the first insight into the measurable value of blame in the
-gradually typed world. In an implicit manner researchers and language designers
-have answered this question one way or another, but without evidence. Now they
-may use the evaluation method of this paper to rationalize their answers.  The
-work of~\citet{tgpk-dls-2018} suggests that, by intuition, programmers prefer
-the run-time checking of Natural over other soundness methods. But, just because
-a random set of programmers expresses this view does not mean that blame
-assignment adds value. At this point, it is clear that, in some circumstances,
-it does and in others, the answer remains murky.
+The work of~\citet{tgpk-dls-2018} suggests that, by intuition, programmers
+prefer the run-time checking of Natural over other soundness methods. But just
+because a random set of programmers expresses this view does not mean that blame
+assignment adds value. Similarly, researchers and language designers have
+implicitly answered this question one way or another without evidence for the
+blame-strategy dimension. The experiment presented in this paper provides some
+justification for the programmers' leanings and helps language creators revisit
+their decisions. Of course, the design choice remains a trade-off, and the
+the experiment contribute to only one of its many facets.
 
@@
 typed realm, and to this end, they need typed APIs for the vast repositories of
-already-existing libraries. Because they can, language implementors merely
-create typed facade modules that import untyped functions and export them with
-type annotations---instead of converting the libraries themselves. With those
-facades, the compiler can easily type-check typed components. The problem is
-that these retroactive additions of types to a library may result from a
-misunderstanding of the code. \emph{The ascribed types may thus be a mistake
-themselves.}
+already-existing libraries. Instead of converting the libraries themselves,
+language implementors merely create typed facade modules that import untyped
+functions and export them with type annotations, like {\tt typed-pack-lib} in
+figure~\ref{fig:example}.  With those facades, the compiler can easily
+type-check typed modules. The problem is that these retroactive additions of
+types to a library may result from a misunderstanding of the code. \emph{Any
+retroactively ascribed type may thus be a mistake itself.}
 
@@
 we expect that these latter two raise misleading exceptions or produce
-plainly-wrong answers.  Of course, only further evaluation work can confirm or
-reject these conjectures.
+plainly-wrong answers.  Of course, only additional experimental work can confirm
+or reject these conjectures.

### 2021-03-04 writing/discussion.tex
issues with sections 9, 10, and 11, plus triggers for global searches

@@
 %% -----------------------------------------------------------------------------
-%% \begin{itemize}
-%% \item Why does the data look like this?
-%%   - why tr fails
-%%     - bad runtime errors
-%%     - TR bug
-%%   - why transient fails
-%%     - bad runtime errors
-%%     - timeouts/ooms: give context wrt fully untyped/typed versions
-%%     - empty blame: trim Ben's prose
 
-%%   \item Any kind of dynamic checking has a large positive impact, whether blame or stacktraces are used.
-%%     \begin{itemize}
-%%     \item Both in terms of usefulness and programmer effort
-%%     \item Erasure extremely often fails to provide any helpful information.
-%%     \item That said, there are no long trails for Erasure.
-%%     \end{itemize}
+The results of the experiment suggests a number of high-level conclusions about
+blame strategies in the gradual typed world.  First, run-time type checks have a
+large positive impact, regardless of whether these checks assign blame or throw
+plain exceptions.  Second, error messages with blame assignments are more
+helpful than those without. The results also indicate, though, that blame is not
+critical in a majority of cases, and therefore they suggest investigating
+whether blame tracking is worth the performance cost.  Third, the Natural
+approach fares better than the Transient approach, but only by a small margin.
+Since Natural offers complete and sound path-based blame while Transient offers
+incomplete but sound heap-based blame~\cite{gfd-oopsla-2019}, the results call
+for a study concerning the relative strengths of the two models of blame.
+Fourth, given that Transient's sound but shallow run-time type checks do not
+seem to hamper debugging, a version of the Natural semantics of Typed Racket
+that disables some of its wrappers may offer an answer to the well-known
+performance issues of Natural~\cite{gtnffvf-jfp-2019}.  Fifth, given that both
+modes of the Transient rational programmer are equally successful and that
+Transient-last produces shorter trails, Reticulated Python could limit the size
+of blame sequences to the last few entries to reduce its serious performance
+problems (see below).
 
-%%   \item Blame helps significantly over just using stacktraces for both Natural and Transient.
-%%     \begin{itemize}
-%%     \item Natural's blame improves over exceptions by nearly 20\%
-%%     \item Transient's blame improves over exceptions half as much
-%%     \end{itemize}
-
-%%   \item Natural clearly offers the most utility, but Transient is not too far behind.
-%%     \begin{itemize}
-%%     \item Natural clearly improves over all other modes
-%%     \item Transient improves significantly over all other modes except Natural.
-%%     \item Keep in mind that our selection of scenarios is biased toward Erasure & Transient.
-%%     \end{itemize}
-
-%%   \item Discussion of how confidence / margin of error should be interpreted wrt these results?
-
-%%    \item Threats to validity:
-%%         \begin{itemize}
-%%           \item Racket stack traces
-%%           \item Mutant sampling
-%%           \item Blame trails do not necessarily capture all aspects of error reporting
-%%           \item Transient blame adaptations
-%%           \item More?
-
-%%    \end{itemize}
-%% \end{itemize}
-
-The analysis in the previous section suggests a number of interesting
-high-level conclusions about blame in the gradual typing setting.  First,
-run time type checks have a large positive impact. This is the case
-regardless of whether these checks issue blame or throw plain
-exceptions.  Second, in many cases type checks that issue blame are more
-helpful than those that do not. However, the results indicate that blame
-is not critical in a majority of cases, and therefore it is worth further
-investigation whether disabling blame tracking for the sake of performance
-may be   an acceptable trade-off. Third, from the three different blame
-strategies, the Natural approach of Typed Racket fares better than the
-Transient approach of Reticulated Python, but only by a small margin.
-Since Natural offers complete and sound path-based blame while Transient
-offers incomplete but sound heap-based blame~\cite{gfd-oopsla-2019}, the
-results call for a deeper understanding of the relative strengths of the
-two models for blame.  Fourth, given that Transient's sound but shallow
-run time type checks do not seem to hamper debugging, a version of Typed
-Racket that disables some of its wrappers may offer an answer to the
-well-known performance issues of Natural~\cite{gtnffvf-jfp-2019}.  Fifth,
-just as for Typed Racket, the experiment points to possible improvements
-for Reticulated Python. In particular, the experimental data indicates
-that the usefulness of transient blame does not depend on whether  the
-rational programmer follows first or the last element from a blame set,
-and that using the last element seems to produce shorter trails.  Hence,
-given the performance issues of Transient discussed further on, a possible
-optimization is to limit the size of blame sets to only contain the latest
-entries using timestamps.
-
-There are a number of threats to the validity of these conclusions: (i) the
-representativeness of benchmarks; (ii) the relation between mutations and
-real programming mistakes; (iii) the definition of interesting debugging
-scenarios; and (iv) the sampling strategy. The previous sections
-have described the various ways we attempt to mitigate these threats;
-the remainder of this section discusses three additional threats.
+The validity of these conclusions is threatened in several ways: (i) the
+representativeness of benchmarks; (ii) the relation between mutations and real
+programming mistakes; (iii) the definition of interesting debugging scenarios;
+and (iv) the sampling strategy. While these threats have been mitigated by the
+experimental setup, the remainder of this section discusses three additional
+ones.
 
@@
 developer for the sake of a systematic, large-scale analysis. This decoupling
-comes with advantages and disadvantages. In the economic realm, mathematical models
-have provided some predictive insights into the market's behavior; but as
+comes with advantages and disadvantages. In the economic realm, mathematical
+models have provided some predictive insights into the market's behavior; but as
 behavioral economics has shown more recently, the mathematical abstraction of a
 rational actor makes predictions also quite unreliable in some situations.
-\footnote{It has also misled economists to focus on just the mathematics, though
-this problem is not relevant here.}  Just like an ordinary consumer or producer,
-an actual software developer is unlikely to stick to the exact strategy proposed
-here. When this happens, the predicted benefits of blame assignment may not
-materialize. Indeed, the authors' personal experience suggests such deviations, and
-it also suggests that deviating is a mistake. To make a true judgment of the
-usefulness of the rational-programmer idea, the community will need much more
-experience with this form of evaluation and relating the evaluation to the
-behavior of working programmers.
+\footnote{It has misled economists to focus on just mathematics, though
+this problem is not relevant here---tongue firmly in cheek.}  Just like an
+ordinary consumer or producer, an actual software developer is unlikely to stick
+to the exact strategy proposed here. When this happens, the predicted benefits
+of blame assignment may not materialize. Indeed, the authors' personal
+experience suggests such deviations, and it also suggests that deviating is a
+mistake. To make a true judgment of the usefulness of the rational-programmer
+idea, the community will need much more experience with this form of evaluation
+and relating the evaluation to the behavior of working programmers.
 
-Relatedly, the experimental setup hides the type choice a rational programmer must make. When the
-run-time checks signal an impedance mismatch in the real world, a programmer
-does not have a typed module ready to swap in. Instead, the programmer must
-come up with the next set of types---and this means the programmer must make
-choices. It is normally possible to assign consistent types to variables in a
-module in different ways. The creation and curation of the benchmarks over
-many years has driven home this lesson but, fortunately, it has also shown that
-the types are in most cases reasonably canonical.  The authors therefore conjecture that
-a rational programmer would in most cases come up with an equivalent type
-assignment in order to extend a trail as the experimental method describes.
+Relatedly, the experimental setup hides how a rational programmer ascribes types
+to extend a trail. When the run-time checks signal an impedance mismatch in the
+real world, a real-world programmer does not have a typed module ready to swap
+in. Instead, the programmer must come up with the next set of types, which means
+making choices. It is usually possible to consistently assign types to variables
+in a module in different ways. The maintenance of the benchmarks over many years
+has driven home this lesson but, fortunately, it has also shown that the types
+are in somewhat canonical.  The authors therefore conjecture that different
+real-world programmers would often come up with equivalent type assignments
+during debugging sessions.
 
@@
 
-The results in section~\ref{sec:results} show that Transient produces empty
-blame sequences for 967 scenarios. In theory, empty set must not occur. An
-empty blame sequence means a lack of boundary crossings for the witness value.
-By implication, a typed module is blaming itself---something that can
-happen only if the type checker (or system) is unsound.
+The execution of the experiment reveals that Transient produces empty blame
+sequences for 967 scenarios. In theory, empty set must not occur. An empty blame
+sequence means a lack of boundary crossings for the witness value.  By
+implication, a typed module is blaming itself---something that can happen only
+if the type checker (or system) is unsound.
 
@@
 unsound but that Vitousek et al.'s Transient blame assignment loses track
-of the proper boundaries due to an incomplete population of the blame map.
+of the proper boundaries due to an incomplete initialization of the blame map.
 Despite these losses, the number of failures is a small fraction of the
@@
 
-{\bf Note} The scenarios where Transient returns no blame 
-suggest ways to improve its blame map:
-\begin{itemize}
-
-\item Entries in a blame map should point to {\em several\/} parent entries.
- For example, if \texttt{f} receives bad input in the context of {\tt (filter f
- xs)}, the entry should point to both \texttt{filter} and \texttt{xs}.
-
-\item The construction of entries must be guided by type-like specifications for
- primitives instead of special cases for built-in functions. Aliasing such
- function changes the blame map.
-
-\end{itemize}
-This small example is indicative of how rational-programmer experiments may
-provide actionable feedback for language design .
+{\bf Note} The scenarios where Transient returns no blame suggest two ways to
+improve its blame map.  First, entries in a blame map should point to {\em
+several\/} parent entries.  For example, if \texttt{f} receives bad input in the
+context of {\tt (filter f xs)}, the entry should point to both \texttt{filter}
+and \texttt{xs}. Second, the construction of entries must be guided by type-like
+specifications for primitives instead of special cases for built-in
+functions. Aliasing such function changes the blame map.
 
@@
 worst-case of 5.4x on fully-typed benchmarks in Reticulated Python.
-Unfortunately, the high cost of Transient Racket seems closer to the truth.
+Unfortunately, the high cost of Shallow Racket seems closer to the truth.
 There are at least four broad issues that skew the earlier results:
@@
 
-\item On large programs, Reticulated suffers from high overhead.  For example, a
-Reticulated version of the simplest benchmark runs in ~40 seconds without blame,
-and with blame times out after 10 minutes. Typed Racket runs the same benchmark
-in a small fraction of the time.
+\item On non-trivial programs, Reticulated Python suffers from the same high
+overhead as Shallow Racket.  It runs the simplest benchmark from the suite 
+in approximately 40 seconds without blame and times out after 10 minutes when
+the blame map is enabled.
+
 \end{enumerate}

### 2021-03-04 writing/mutate-mutators.tex
issues with sections 9, 10, and 11, plus triggers for global searches

@@
 gradually typed languages; the next six to those that include classes and
-objects. The last four target distinguishing features of Typed Racket,
+objects. The last four target distinguishing features of Typed
+Racket's type system,
 specifically its sophisticated type system. For example, one mutation produced by {\tt arithmetic}
@@
 Integer}~\cite{stathff-padl-12}. Similarly, the other three aim to confound
-Typed Racket's occurrence type system.
+the occurrence type system.

### 2021-03-04 writing/rational-modes.tex
issues with sections 9, 10, and 11, plus triggers for global searches

@@
 blame is a critical piece of the rational programmer's process.  For instance,
-typing the top of stack trace of a failed run time type check, dubbed the
+typing the top of stack trace of a failed run-time type check, dubbed the
 location of the exception, might be as useful as typing the blamed one.

### 2021-03-04 writing/related.tex
issues with sections 9, 10, and 11, plus triggers for global searches

@@
 The literature on gradual typing presents many semantics beyond the three used
-here and only two additional blame strategies.  Pyret
-(\shorturl{https://www.}{pyret.org}) assigns fixed-size data types the Natural
-semantics and functions a Transient semantics. The Forgetful~\cite{cl-icfp-2017}
-and the Amnesic~\cite{gfd-oopsla-2019} semantics are similar to Transient but
-use wrappers instead of in-lined checks.  Nom~\cite{mt-oopsla-2017} and other
-\emph{concrete\/} semantics~\cite{wnlov-popl-2010, rsfbv-popl-2015,
-rzv-ecoop-2015, rat-oopsla-2017} assume that every value comes with a type tag
-and use tag checks to supervise the interactions between typed and untyped code.
-Monotonic references~\cite{svctg-esop-2015} and the
-semantics~\cite{tlt-popl-2019, etg-icfp-19, tt-scp-20, tgt-popl-18, tt-sas-17}
-derived with the Abstracting Gradual Typing technique~\cite{gct-popl-2016} are
-variants of Natural.  Among these representative semantics, only Amnesic and Nom
-present innovative blame strategies.  The experiment excludes these strategies
-because they seem to be merely theoretical constructions and/or impose severe
-restrictions on programmers.
+here and two additional blame strategies.  Pyret (pyret.org) assigns fixed-size
+data types the Natural semantics and functions a Transient semantics. The
+Forgetful~\cite{cl-icfp-2017} and the Amnesic~\cite{gfd-oopsla-2019} semantics
+are similar to Transient but use wrappers instead of in-lined checks.
+Nom~\cite{mt-oopsla-2017} and other \emph{concrete\/}
+semantics~\cite{wnlov-popl-2010, rsfbv-popl-2015, rzv-ecoop-2015,
+rat-oopsla-2017} assume that every value comes with a type tag and use tag
+checks to supervise the interactions between typed and untyped code.  Monotonic
+references~\cite{svctg-esop-2015} and the semantics~\cite{tlt-popl-2019,
+etg-icfp-19, tt-scp-20, tgt-popl-18, tt-sas-17} derived with the Abstracting
+Gradual Typing technique~\cite{gct-popl-2016} are variants of Natural.  Among
+these representative semantics, only Amnesic and Nom present innovative blame
+strategies.  The experiment excludes these strategies because they seem to be
+merely theoretical constructions and/or impose severe restrictions on
+programmers.

### 2021-03-04 writing/results.tex
issues with sections 9, 10, and 11, plus triggers for global searches

@@
 total) reveals an additional cause. For a small set of debugging scenarios
-(40), Natural produces a run time type error blaming a non-buggy
+(40), Natural produces a run-time type error blaming a non-buggy
 already-typed module. All these cases are due to known open issues with Typed
@@
 program with the new annotations, the rational programmer is stuck.  In other
-words, an exception from the runtime has to point to the buggy module with the
+words, an exception from the run-time system has to point to the buggy module with the
 first try. Otherwise the rational programmer types an irrelevant module, runs

### 2021-03-05 writing/challenges.tex
Fix citation style

@@
 soundness mechanisms are available in a robust and comparable manner:
-Typed Racket implements Natural, Shallow Racket~\cite{ttt21}
+Typed Racket implements Natural, Shallow Racket~\citep{ttt21}
 Transient, and plain Racket for Erasure.
@@
 An alternative is to {\em generate a large and representative corpus of type
-mistakes \/} using mutation analysis~\cite{lipton1971fault, demillo1978hints,
+mistakes \/} using mutation analysis~\citep{lipton1971fault, demillo1978hints,
 jia2011analysis}. Conventional mutation analysis is useless, however.  Mutation
@@
 
-Based on a related experience, Gopinath and Walkingshaw~\cite{gw-mutation} write,
+Based on a related experience, Gopinath and Walkingshaw~\citep{gw-mutation} write,
 ``existing mutation frameworks \ldots\ do not generate the kinds of mutations

### 2021-03-05 writing/conclusion.tex
Fix citation style

@@
 The paper does {\em not\/} address a problem in the gradually typed world that
-was pointed out early on by practical researchers~\cite{incorrect-ts,
+was pointed out early on by practical researchers~\citep{incorrect-ts,
 sta-nt-base-types, wmwz-ecoop-2017} and that has recently received theoretical
-attention~\cite{gfd-oopsla-2019, cc-oopsla-20}: mistakes in type annotations
+attention~\citep{gfd-oopsla-2019, cc-oopsla-20}: mistakes in type annotations
 themselves.  Developers use gradual typing to move an untyped code base into the

### 2021-03-05 writing/discussion.tex
Fix citation style

@@
 Since Natural offers complete and sound path-based blame while Transient offers
-incomplete but sound heap-based blame~\cite{gfd-oopsla-2019}, the results call
+incomplete but sound heap-based blame~\citep{gfd-oopsla-2019}, the results call
 for a study concerning the relative strengths of the two models of blame.
@@
 that disables some of its wrappers may offer an answer to the well-known
-performance issues of Natural~\cite{gtnffvf-jfp-2019}.  Fifth, given that both
+performance issues of Natural~\citep{gtnffvf-jfp-2019}.  Fifth, given that both
 modes of the Transient rational programmer are equally successful and that

### 2021-03-05 writing/ideas.tex
uniform editorial voice, mistakes, typos, size of text in figures, captions
Fix citation style

@@
 approach,\footnote{The method can be adapted to so-called true gradual typing
-systems~\cite{svcb-snapl-2015} in a straightforward manner.}, hence, each configuration differs from another
+systems~\citep{svcb-snapl-2015} in a straightforward manner.}, hence, each configuration differs from another
 in which of its components are typed and which are not. Formally, we can
@@
 We say a blametrail is final iff its last configuration, $\conf_n$
-doesn't blame a component, $\blame{\conf_i} = \emptyset$. We call a
+does not blame a component, $\blame{\conf_i} = \emptyset$. We call a
 blametrail \emph{successful} iff it is final and its last configuration

### 2021-03-05 writing/introduction.tex
Fix citation style

@@
 Theoreticians of gradual typing have focused on blame theorems from the
-very beginning~\cite{mf-toplas-2009, tf-dls-2006}. ``Well-typed
+very beginning~\citep{mf-toplas-2009, tf-dls-2006}. ``Well-typed
 [components]\footnote{The original authors got this word wrong.} can't be
-blamed'' turned the theorem into a slogan~\cite{wf-esop-2009}. Academic
-systems (Reticulated Python~\cite{vsc-dls-2019, vss-popl-2017,
+blamed'' turned the theorem into a slogan~\citep{wf-esop-2009}. Academic
+systems (Reticulated Python~\citep{vsc-dls-2019, vss-popl-2017,
 vksb-dls-2014} and Typed
-Racket~\cite{tf-dls-2006,tf-popl-2008,tfffgksst-snapl-2017,tf-icfp-2010})
+Racket~\citep{tf-dls-2006,tf-popl-2008,tfffgksst-snapl-2017,tf-icfp-2010})
 come with sophisticated checking and blame assignment strategies (see
@@
 sec.~\ref{sec:results}): (1)~{\it Transient\/}, i.e. Reticulated's
-tracking of typed/untyped boundaries~\cite{vss-popl-2017}; (2)~{\it
+tracking of typed/untyped boundaries~\citep{vss-popl-2017}; (2)~{\it
 Natural\/}, i.e. Typed Racket's use of a higher-order contract system and
-its blame assignment~\cite{ff-icfp-2002}; and (3) {\it Erasure\/}, i.e.
+its blame assignment~\citep{ff-icfp-2002}; and (3) {\it Erasure\/}, i.e.
 the approach of industrial systems, which forgo blame in favor of error

### 2021-03-05 writing/landscape.tex
issues with sections 9, 10, and 11, plus triggers for global searches
Fix citation style

@@
 with types at run-time: {\em Natural\/}, {\em Transient\/}, and {\em
-Erasure\/}. This self-contained section summarizes these options with one illustrative example
-using Racket syntax. The informed reader may wish to merely scan it.
+Erasure\/}. This self-contained section summarizes these options with one
+illustrative example using (Typed) Racket syntax. The informed reader may wish to merely
+scan it.
 
@@
 typed modules are wrapped in proxies that enforce the static type discipline
-with run-time checks and track responsibilities~\cite{tf-popl-2008,
+with run-time checks and track responsibilities~\citep{tf-popl-2008,
 tfffgksst-snapl-2017}. Thus, when {\tt pack} is imported into a typed module,
@@
 points to functions check their arguments at run time and all function calls
-check their return values against the expected type~\cite{vss-popl-2017}.
+check their return values against the expected type~\citep{vss-popl-2017}.
 Furthermore, Transient uses \emph{shallow} checks, meaning they inspect only the

### 2021-03-05 writing/mutate-benchmarks.tex
Fix citation style

@@
 programming style. The latter is mostly a reference to their reliance on many
-Typed Racket features: occurrence typing~\cite{tf-icfp-2010}, types for mutable
-and immutable data structures~\cite{hpst-sfp-2010}, types for first-class
-classes and objects~\cite{tsdtf-oopsla-2012}, and types for Racket's numeric
-tower~\cite{stathff-padl-12}.
+Typed Racket features: occurrence typing~\citep{tf-icfp-2010}, types for mutable
+and immutable data structures~\citep{hpst-sfp-2010}, types for first-class
+classes and objects~\citep{tsdtf-oopsla-2012}, and types for Racket's numeric
+tower~\citep{stathff-padl-12}.
 

### 2021-03-05 writing/mutate-interesting.tex
Fix citation style

@@
 case of gradual typing semantics, a meta-theorem due to Greenman and
-Felleisen~\cite{gf-icfp-2018} says that if a program raises an exception under
+Felleisen~\citep{gf-icfp-2018} says that if a program raises an exception under
 Erasure, it also errors under all other semantics.  Hence, a comparison of blame

### 2021-03-05 writing/mutate-mutators.tex
Fix citation style

@@
 generalize~\citet{lksfd-popl-2020}'s mutators, which in turn are borrowed from
-the vast literature on mutation testing~\cite{jia2011analysis}.  Only two are
+the vast literature on mutation testing~\citep{jia2011analysis}.  Only two are
 directly inherited; many mutators are brand new. For the latter, the authors
@@
 arguments are positive integers, while {\tt -} yields {\tt
-Integer}~\cite{stathff-padl-12}. Similarly, the other three aim to confound
+Integer}~\citep{stathff-padl-12}. Similarly, the other three aim to confound
 the occurrence type system.

### 2021-03-05 writing/related.tex
Fix citation style

@@
 data types the Natural semantics and functions a Transient semantics. The
-Forgetful~\cite{cl-icfp-2017} and the Amnesic~\cite{gfd-oopsla-2019} semantics
+Forgetful~\citep{cl-icfp-2017} and the Amnesic~\citep{gfd-oopsla-2019} semantics
 are similar to Transient but use wrappers instead of in-lined checks.
-Nom~\cite{mt-oopsla-2017} and other \emph{concrete\/}
-semantics~\cite{wnlov-popl-2010, rsfbv-popl-2015, rzv-ecoop-2015,
+Nom~\citep{mt-oopsla-2017} and other \emph{concrete\/}
+semantics~\citep{wnlov-popl-2010, rsfbv-popl-2015, rzv-ecoop-2015,
 rat-oopsla-2017} assume that every value comes with a type tag and use tag
 checks to supervise the interactions between typed and untyped code.  Monotonic
-references~\cite{svctg-esop-2015} and the semantics~\cite{tlt-popl-2019,
+references~\citep{svctg-esop-2015} and the semantics~\citep{tlt-popl-2019,
 etg-icfp-19, tt-scp-20, tgt-popl-18, tt-sas-17} derived with the Abstracting
-Gradual Typing technique~\cite{gct-popl-2016} are variants of Natural.  Among
+Gradual Typing technique~\citep{gct-popl-2016} are variants of Natural.  Among
 these representative semantics, only Amnesic and Nom present innovative blame

### 2021-03-05 writing/why-rational.tex
Fix citation style

@@
 reasoning about numbers, not (only) machine-level
-representations~\cite{stathff-padl-12}.} A program execution ends in this error:
+representations~\citep{stathff-padl-12}.} A program execution ends in this error:
 \begin{verbatim}

### 2021-03-08 writing/mutate-interesting.tex
Add explanation to interesting mutant breakdown figure (6)

@@
 
-\begin{figure*} \footnotesize
+\begin{figure}
  \centering
  \includegraphics[scale=0.33]{./plots/mutant-breakdown}
-  Counts are cut off at 50.
 
-\caption{The generated debugging scenarios across all benchmarks} 
+ \vspace{1em}
+
+ \begin{minipage}{\textwidth}
+ Each plot shows a breakdown of interesting mutants by mutator.
+ Each mutator corresponds to a bar representing the number of interesting mutants generated by that mutator.
+ The counts are cut off at 50, so those bars reaching the edge of the plot represent 50 or more interesting mutants.
+ \end{minipage}
+
+\caption{Breakdown of interesting mutants by mutator, per benchmark.}
  \label{fig:mutant-breakdown}
-\end{figure*}
+\end{figure}
 

### 2021-03-08 writing/challenges.tex
Fix incorrect uses of parenthetical citations, replace with textual

@@
 
-Based on a related experience, Gopinath and Walkingshaw~\citep{gw-mutation} write,
+Based on a related experience, \citet{gw-mutation} write,
 ``existing mutation frameworks \ldots\ do not generate the kinds of mutations

### 2021-03-08 writing/mutate-interesting.tex
Fix incorrect uses of parenthetical citations, replace with textual

@@
 chance to debug the same scenario using different pieces of information.  In the
-case of gradual typing semantics, a meta-theorem due to Greenman and
-Felleisen~\citep{gf-icfp-2018} says that if a program raises an exception under
+case of gradual typing semantics, a meta-theorem due to \citet{gf-icfp-2018} says that if a program raises an exception under
 Erasure, it also errors under all other semantics.  Hence, a comparison of blame

### 2021-04-21 writing/reviews/icfp21-rebuttal.md
Add icfp21 rebuttal draft

Thank you to all the reviewers for the helpful feedback.

The first part of our response addresses big themes in the reviews;
the second part addresses all remaining concerns individually.

What do the results say about blame for gradual typing in practice?
================================================================================
The introduction says that the experimental results lead to the following conclusions about blame in practice:
1. Blame helps locate impedance mismatches, though not immensely more than simple stacktraces, and
2. Natural blame helps more than Transient, and Transient more than Erasure's stacktraces, *but*
3. Natural blame is only marginally better than Transient in practice, and both not immensely better than Erasure's stacktraces.
The phrase "in practice" refers to the way that blame behaves in real programs with an impedance mismatch, using real implementations of each semantics.

In light of the data in figures 8, 9, and 10, the benefits of blame in in practice do not measure up to the expectations established by theory.
Although blame does provide useful debugging information, the results suggest that in practice blame is only necessary to debug (as compared with simple exceptions) in a small fraction of scenarios.
Furthermore, the theory suggests that Natural's blame should be more precise than Transient because Natural precisely tracks all possible interactions while Transient approximates (see [11]: Greenman et al. OOPSLA'19), but in practice Natural's blame has comparable utility to Transient.

The experiment in this paper reveals that mismatch, but it necessarily considers only a limited view of the space of possible impedance mismatch scenarios that might arise in reality.
Section 11 points to another space of scenarios that deserves analysis: scenarios where mismatches arise due to mistakes in annotations rather than (or as well as) code.
Indeed, review B astutely points out two more spaces worth exploring (programs where some components cannot be typed, and non-deterministic programs).
<!-- As another example, the paper only considers scenarios known to be detectable by all three systems, but theory makes it clear that Natural and Erasure can detect some scenarios that Erasure cannot. -->
<!-- How common are those scenarios for impedance mismatch bugs? -->
<!-- Does blame provide more benefit than simple exceptions in those scenarios? -->
Additional experimentation is necessary to develop an understanding of these different spaces.

The Rational Programmer vs Lazarek et al.
================================================================================
Lazarek et al. (POPL'20) provide a kernel insight, which this paper develops into the novel Rational Programmer for evaluating blame in gradual typing as its first contribution.
The key distinction is the idea of the rational programmer as an entity to study and how it can be parameterized by modes to enable cross-system comparison.

Mutators and benchmarks
================================================================================
Some more description of the mutators and benchmarks the paper uses would clearly be useful, so we will expand the explanation and examples in the paper.
For now, we have responded to the reviewers specific questions inline below.

Failing blame trails
================================================================================
The paper does not make clear enough how Natural blame can have failing scenarios, so we will update it to clarify.
In short, Natural blame trails can by stymied by unhelpful run-time errors.
Because debugging scenarios are only partially typed, some scenarios produce an error from the runtime (e.g. `+` receiving a non-number) before any of Natural's run-time type checks can discover a problem.
Runtime errors do not carry blame, only stacktraces, so the only way to proceed (regardless of mode) is by using the stack information.
If the stack proves unhelpful, then the trail fails.
Such failures are the main source of failing blame trails in the `Natural blame` mode.

Replies to individual points
==================================================================

Review #45A
===========================================================================
> the paper talks in the introduction of its top-level innovation being the idea of the rational programmer. But then they go on to say that the idea is actually taken from Lazarek. Which is it then?
See above.

> The table in section 2 looks rather ugly
Thanks for letting us know, we will find a new way to present this information.

> p5. one thing I wonder about is how this all hinges on the rational programmer not only taking the right action, but also taking it right? Is that exactly what the rational programmer idea presupposes? (I guess this reflects on your remark in Sec. 11 that one problem is with errors in the ascribed types themselves.).
Yes, you hit the nail on the head. 
One of the central parts of the rational programmer is that it fixes decision of how to respond to a given scenario; 
so evaluating different ways of taking action boils down to defining and testing the corresponding rational programmers.

> When the rational programmer "fixes" a module and reruns, the implicit assumption seems to be that he runs the program in the exact same way. That would make the most sense to me, but you do not seem to come out and say so.
Yes you're right. Thanks, we will clarify this in the prose.

> It would help if you could explain how a mutator like negate-cond actually leads to a program
> that has more type-level mistakes than before. It does not seem to me this part of the code
> itself will now fail, but that changing the conditions makes certain previously excluded paths in the code feasible.
In the specific cases of `negate-cond` and `force-cond`, these mutators lead to type-level mistakes when the program uses occurrence typing. 
For example, they could both lead to a mistake in a program like this:
```
(: use-int  (-> Integer Void))
(: use-bool (-> Boolean Void))
(: x        (U Integer Boolean))

(if (integer? x) (use-int x) (use-bool x))
```

> How do you arrive at the number 72192 in Section 7? It is not equal to 756 times 96.
Indeed, 756 is a typo: it should be 752.

Review #45B
===========================================================================
>  - Which blame assignment strategy is "good" and which is not "good"?
TODO

>  - At first glance, the conclusion in lines 55-56 (starting with "Second, ...")
>    seems to contradict that in line 99-100 (starting with "neither is ...").  It would
>    be nicer to address them in a clearer manner.
Thanks for the feedback, we will rephrase the conclusion.

> - The paper often mixes gradual typing and migratory typing.  It is unclear how
>   it distinguishes them.
TODO

> - It is not fully explained how the given mutators change programs.  For
>   example, I cannot completely predict changes by the mutators deletion and
>   class:super.
In the specific case of `deletion`, the other most common place that this mutator applies are so-called "implicit begins": places where sequences are allowed implicitly by enclosing forms like `define` or `cond`.
So here are two more examples of `deletion` mutations:
```
(define (do-something) (step-1!) (step-2!))
~>
(define (do-something) (step-1!))

(cond [(do-something?) (step-1!) (step-2!)]
      [else (void)])
~>
(cond [(do-something?) (step-1!)]
      [else (void)])
```

In the case of `class:super`, the mutator simply replaces any occurrence of `super-new` in the body of a class definition with `void`.

> - The paper says that the experiment uses 72,192 sampled scenarios (line 802).
>   However, the total number of mutants (i.e., configuration lattices) is 756
>   and, from each lattice, 96 scenarios are sampled.  Thus, I guess 756*96 =
>   72576 scenarios are used.  Why does this mismatch happen?
Indeed, 756 is a typo: it should be 752.

> - The experiment samples multiple scenarios from a configuration lattice (line
>   800).  Does this mean the debugging process may start with a mix-typed program
>   and may not with a fully untyped one?  If so, why does the paper take such an
>   approach?
The bottom of the configuration lattice is fully untyped, so that is a possible starting scenario to be sampled.

> - The paper often says that a blame system is (un)sound, but it is difficult for
>   me to identify what it precisely means.
The term refers to the soundness defined by [11] (Greenman et al. OOPSLA'19), but we agree that for clarity the paper should define it in addition to the reference.
Intuitively, a blame system is sound if it can only blame components that have had control of the value witnessing a violation.
In other words, the system cannot blame arbitrary components unrelated to the witness of a contract violation.

> - The paragraph starting at line 1122 is quite difficult to understand for me.
>   [...]
>   - What does the paper mean by "incomplete population of the blame map"?
>   - To understand the idea on the improvements of Transit, more explanations
>     on the usage of blame maps in Transient would be needed.
TODO

>   - What "sophisticated typing features" are considered (line 589)?  It would be
>     crucial to confirm whether the proposed mutators are enough.
The benchmarks employ many of Typed Racket's sophisticated feaures, most notably (from L641) occurrence typing, types for mutable and immutable data structures, types for first-class classes and objects, and types for Racket's numeric tower.
We absolutely agree that it is crucial to analyze the mutators in light of these features;
perhaps adding a breakdown of features by benchmark to section 6.3 would help clarify this?

>   - The benchmarks are selected (line 647), but why?  The GPT benchmark suite of
>     Racket provides more examples.
The benchmarks in figure 5 are the ten largest dependency graphs of the GTP suite. 
We will rephrase L646-648 to make this clearer.

> - It seems that the experiment implicitly supposes a few assumptions on
>   benchmarks.  First, the original benchmarks must be fully typed.  Thus more
>   experiments on mix-typed programs where some components cannot be typed may be
>   needed.  Second, the benchmarks are supposed to be deterministic, which I find
>   from the definitions of trails.  If it is the case, while I do not think these
>   restrictions have to be lifted in the paper, it would be nice to expose them.
You are absolutely right, we will clarify these assumptions about the benchmarks in the paper.

> - As an alternative of Transient first and last blame, perhaps determining
>   blamed modules by voting (i.e., blaming a module that has been most often
>   added to the blame map) might be promising (if the blame map has a large
>   population).  Have the authors considered it or another approach for Transient
>   blame?
We have since thought of a few alternative strategies, but not that one: good idea!
Ultimately, Transient does not offer any interpretation of its blame, so the paper chooses first and last as two obvious reasonable choices.

> - Perhaps it is valuable for followers to share the experience on developing
>   mutators that are not interesting, .e.g., in the supplementary material.
We agree and would be happy to do so.

> - Why doesn't the paper adopt the author-year citation style?
This was a mistake typesetting the paper, we will fix it.

> - Page 7: How is the question (5) answered?
Thanks for pointing out that this is confusing. 
As it is, the rest of the paper is the implicit answer to (5), but we will clarify the prose.

> - L356 "despite advertisements for the opposite":  I cannot find what this means.
This is meant to convey that although Lazarek et al. aim to evaluate blame, they fail to adequately account for confounding factors.
We will reword for clarity.

> - L370 "the latter must represent":  What does the "latter" specify?
It refers to impedance mismatches from the end of the preceding sentence;
we will clarify the wording here.

> - L424 "Let a configuration s of P":  Are configurations the same as scenarios?
Scenarios are a specific kind of configuration (that doesn't type the buggy module).
We will adjust these two paragraphs to make this more clear.

> - L431 "type-level mistake":  Is it the same as a impedance mismatch?
A type-level mistake is a mistake that causes an impedance mismatch at a boundary between typed and untyped code.
We will clarify this point in the prose.

> - Are failing Natural blame trails produced even for programs with impedance
>   mismatch?  (This question is related to the above issue with Natural blame.)
Yes, please see the start of this response.

> - L556: It would be helpful to describe how to extend trails and how to
>   determine if there is no scenario to be added.  (This comment is also related
>   to the issue with Natural blame).
Thanks, we will expand on this in the prose.

> - L557: "if it does" --> "if it does not"?
You're right, thanks, we will fix it.

> - L685 "e.g., changing '*' to '/'":  Is this an example of the mutator that
>   "does not reliably lead to type errors"?
This is a refinement of the arithmetic mutator that does lead to type errors, because given two integers `*` always produces an integer, but `/` may produce a rational, which have different types in Typed Racket.
We will clarify this comment in the footnote.

> - L785 "the interesting standard guided countless iterations":  I fail to find
>   what this intends.
The measure of interesting mutants provided a standard by which we could judge potential mutators and iteratively develop an effective set of them.
We will make this comment clearer.

> - What is the total number of failing trails in Transient?
3594 out of 72192 trails fail for transient last blame, and 3591 for first.

> - L1099 "as behavioral economics has shown more recently":  Is there a reference
>   to be cited?
TODO

> - L1104 "deviating is a mistake":  Why?
Our experience of ignoring blame as opposed to heeding it suggests that ignoring it is usually unwise.

> - L1154 "in the "fully typed" benchmarks":  Are these benchmarks on Python?
The slowdowns the paper quotes here are from Python benchmarks.

> - L1161 "the simplest benchmark":  Is this on Racket?  Which benchmark
>   does it intend?
This paragraph refers to a Python translation of the Racket `Sieve` benchmark;
it is confusingly worded, we will adjust the phrasing.

> - L118: "(at the mid-level on the left)[] imports" (the comma is removed)
> - L332 "soundness mechanisms":  What they mean is somewhat unclear.  Please
>   consider clarifying.
> - L347 "the rational programmer's effort": This was unclear at the first
>   reading.  It would be nice to briefly explain it here.
> - L451: "denotes the module [of P] that s blames"
> - L970: "the experiment provides [an] evidence that"
> - L1198 "But just because...":  It is difficult for me to find what this
>   sentence wants to say.  Please consider rephrasing.
Thanks for these comments, we will incorporate them to improve the prose.

### 2021-04-21 writing/reviews/icfp21-rebuttal.md
small edits

@@
-Thank you to all the reviewers for the helpful feedback.
-
 The first part of our response addresses big themes in the reviews;
@@
 ================================================================================
-The introduction says that the experimental results lead to the following conclusions about blame in practice:
+
+The introduction says that the experimental results lead to the
+following conclusions about blame _in practice_:
+
 1. Blame helps locate impedance mismatches, though not immensely more than simple stacktraces, and
@@
 3. Natural blame is only marginally better than Transient in practice, and both not immensely better than Erasure's stacktraces.
-The phrase "in practice" refers to the way that blame behaves in real programs with an impedance mismatch, using real implementations of each semantics.
 
-In light of the data in figures 8, 9, and 10, the benefits of blame in in practice do not measure up to the expectations established by theory.
-Although blame does provide useful debugging information, the results suggest that in practice blame is only necessary to debug (as compared with simple exceptions) in a small fraction of scenarios.
-Furthermore, the theory suggests that Natural's blame should be more precise than Transient because Natural precisely tracks all possible interactions while Transient approximates (see [11]: Greenman et al.
-OOPSLA'19), but in practice Natural's blame has comparable utility to Transient.
+The phrase "in practice" refers to the way that blame behaves in real
+programs with an impedance mismatch, using real implementations of
+each semantics.
 
-The experiment in this paper reveals that mismatch, but it necessarily considers only a limited view of the space of possible impedance mismatch scenarios that might arise in reality.
-Section 11 points to another space of scenarios that deserves analysis: scenarios where mismatches arise due to mistakes in annotations rather than (or as well as) code.
-Indeed, review B astutely points out two more spaces worth exploring (programs where some components cannot be typed, and non-deterministic programs).
-<!-- As another example, the paper only considers scenarios known to be detectable by all three systems, but theory makes it clear that Natural and Erasure can detect some scenarios that Erasure cannot.
--->
+In light of the data in figures 8, 9, and 10, the benefits of blame in
+practice do not measure up to the expectations established by theory.
+Concretely, blame does provide useful debugging information, but the
+results suggest that in practice blame is only necessary to debug (as
+compared with simple exceptions) in a small fraction of scenarios.
+
+- The theory suggests that Natural's blame should be more precise than
+  Transient because Natural precisely tracks all possible interactions
+  while Transient approximates (see [11]: Greenman et al. OOPSLA'19).
+
+- In practice, Natural's blame has utility comparable to Transient's.
+
+The experiment in this paper reveals that mismatch, but it necessarily
+considers only a limited view of the space of possible impedance
+mismatch scenarios that might arise in reality.
+
+Section 11 points to another space of scenarios that deserves
+analysis: scenarios where mismatches arise due to mistakes in type
+annotations rather than (or as well as) code.  Indeed, review B
+astutely points out two more spaces worth exploring:
+
+- programs where some components _cannot be typed_
+- non-deterministic programs, which inject ephemeral bugs.
+
+<!-- As another example, the paper only considers scenarios known to
+be detectable by all three systems, but theory makes it clear that
+Natural and Erasure can detect some scenarios that Erasure cannot. -->
+
 <!-- How common are those scenarios for impedance mismatch bugs? -->
 <!-- Does blame provide more benefit than simple exceptions in those scenarios? -->
+
 Additional experimentation is necessary to develop an understanding of these different spaces.
@@
 ================================================================================
-Lazarek et al. (POPL'20) provide a kernel insight, which this paper develops into the novel Rational Programmer for evaluating blame in gradual typing as its first contribution.
-The key distinction is the idea of the rational programmer as an entity to study and how it can be parameterized by modes to enable cross-system comparison.
+
+Lazarek et al. (POPL'20) provide a kernel insight, which this paper
+develops into the novel idea of the Rational Programmer. It is the
+first paper to explain the idea properly and use the phrase. We
+therefore consider it its first contribution.
+
+The key distinction is the idea of the rational programmer as an
+entity to study and how it can be parameterized by modes to enable
+cross-system comparison.
 
@@
 ================================================================================
-Some more description of the mutators and benchmarks the paper uses would clearly be useful, so we will expand the explanation and examples in the paper.
-For now, we have responded to the reviewers specific questions inline below.
+
+Some more description of the mutators and benchmarks the paper uses
+would clearly be useful, so we will expand the explanation and
+examples in the final paper.  For now, we have responded to the
+reviewers specific questions inline below.
 
@@
 ================================================================================
-The paper does not make clear enough how Natural blame can have failing scenarios, so we will update it to clarify.
-In short, Natural blame trails can by stymied by unhelpful run-time errors.
-Because debugging scenarios are only partially typed, some scenarios produce an error from the runtime (e.g.
-`+` receiving a non-number) before any of Natural's run-time type checks can discover a problem.
-Runtime errors do not carry blame, only stacktraces, so the only way to proceed (regardless of mode) is by using the stack information.
-If the stack proves unhelpful, then the trail fails.
-Such failures are the main source of failing blame trails in the `Natural blame` mode.
+
+The paper fails to clarify how Natural blame can have failing
+scenarios, so we will update it to clarify.  In short, Natural blame
+trails can by stymied by unhelpful run-time errors.
+
+Because debugging scenarios are only partially typed, some scenarios
+produce an error from the runtime (e.g. `+` receiving a non-number)
+before any of Natural's run-time type checks can discover a problem.
+
+Run-time errors do not carry blame, only stacktraces, so the only way
+to proceed (regardless of mode) is by using the stack information.  If
+the stack proves unhelpful, then the trail fails.  Such failures are
+the main source of failing blame trails in the `Natural blame` mode.
 
@@
 ===========================================================================
-> the paper talks in the introduction of its top-level innovation being the idea of the rational programmer. But then they go on to say that the idea is actually taken from Lazarek. Which is it then?
-See above.
-
-> The table in section 2 looks rather ugly
-Thanks for letting us know, we will find a new way to present this information.
 
@@
 ===========================================================================
+
 >  - Which blame assignment strategy is "good" and which is not "good"?
@@
 >   class:super.
-In the specific case of `deletion`, the other most common place that this mutator applies are so-called "implicit begins": places where sequences are allowed implicitly by enclosing forms like `define` or `cond`.
+In the specific case of `deletion`, the other most common place that
+>   this mutator applies are so-called "implicit begins": places where
+>   sequences are allowed implicitly by enclosing forms like `define`
+>   or `cond`. 
 So here are two more examples of `deletion` mutations:
@@
 >   72576 scenarios are used.  Why does this mismatch happen?
+
 Indeed, 756 is a typo: it should be 752.
@@
 >   approach?
-The bottom of the configuration lattice is fully untyped, so that is a possible starting scenario to be sampled.
+
+The bottom of the configuration lattice is completely untyped, so that
+is a possible starting scenario to be sampled.
 
@@
 >   me to identify what it precisely means.
-The term refers to the soundness defined by [11] (Greenman et al. OOPSLA'19), but we agree that for clarity the paper should define it in addition to the reference.
-Intuitively, a blame system is sound if it can only blame components that have had control of the value witnessing a violation.
-In other words, the system cannot blame arbitrary components unrelated to the witness of a contract violation.
+
+The term refers to the soundness defined by [11] (Greenman et
+al. OOPSLA'19), but we agree that for clarity the paper should define
+it in addition to the reference.  Intuitively, a blame system is sound
+if it can only blame components that have had control of the value
+witnessing a violation.  In other words, the system cannot blame
+arbitrary components unrelated to the witness of a contract violation.
 
@@
 >     crucial to confirm whether the proposed mutators are enough.
-The benchmarks employ many of Typed Racket's sophisticated feaures, most notably (from L641) occurrence typing, types for mutable and immutable data structures, types for first-class classes and objects, and types for Racket's numeric tower.
-We absolutely agree that it is crucial to analyze the mutators in light of these features;
-perhaps adding a breakdown of features by benchmark to section 6.3 would help clarify this?
+
+The benchmarks employ many of Typed Racket's sophisticated features,
+most notably (from L641) occurrence typing, types for mutable and
+immutable data structures, types for first-class classes and objects,
+and types for Racket's numeric tower.  We absolutely agree that it is
+crucial to analyze the mutators in light of these features; perhaps
+adding a breakdown of features by benchmark to section 6.3 would help
+clarify this?
 
@@
 >     Racket provides more examples.
+
 The benchmarks in figure 5 are the ten largest dependency graphs of the GTP suite. 
@@
 >   restrictions have to be lifted in the paper, it would be nice to expose them.
-You are absolutely right, we will clarify these assumptions about the benchmarks in the paper.
+
+You are absolutely right, we will clarify these assumptions about the
+benchmarks in the paper. They are inherited from the several
+publications on the gradual benchmark suite.
 
@@
 >   blame?
-We have since thought of a few alternative strategies, but not that one: good idea!
-Ultimately, Transient does not offer any interpretation of its blame, so the paper chooses first and last as two obvious reasonable choices.
+
+We have since thought of a few alternative strategies, but not that
+one: good idea!  Ultimately, Transient does not offer any
+interpretation of its blame, so the paper chooses first and last as
+two obvious reasonable choices.
 
@@
 >   mutators that are not interesting, .e.g., in the supplementary material.
+
 We agree and would be happy to do so.
@@
 > - Why doesn't the paper adopt the author-year citation style?
+
 This was a mistake typesetting the paper, we will fix it.
@@
 > - Page 7: How is the question (5) answered?
-Thanks for pointing out that this is confusing. 
-As it is, the rest of the paper is the implicit answer to (5), but we will clarify the prose.
+
+Thanks for pointing out that this is confusing.  As it is, the rest of
+the paper is the implicit answer to (5), but we will clarify the
+prose.
 
 > - L356 "despite advertisements for the opposite":  I cannot find what this means.
-This is meant to convey that although Lazarek et al. aim to evaluate blame, they fail to adequately account for confounding factors.
-We will reword for clarity.
+
+This is meant to convey that although Lazarek et al. aim to evaluate
+blame, they fail to adequately account for confounding factors.  We
+will reword for clarity.
 
 > - L370 "the latter must represent":  What does the "latter" specify?
-It refers to impedance mismatches from the end of the preceding sentence;
-we will clarify the wording here.
+
+It refers to impedance mismatches from the end of the preceding
+sentence; we will clarify the wording here.
 
 > - L424 "Let a configuration s of P":  Are configurations the same as scenarios?
-Scenarios are a specific kind of configuration (that doesn't type the buggy module).
-We will adjust these two paragraphs to make this more clear.
+
+Scenarios are a specific kind of configuration (that doesn't type the
+buggy module).  We will adjust these two paragraphs to make this more
+clear.
 
 > - L431 "type-level mistake":  Is it the same as a impedance mismatch?
-A type-level mistake is a mistake that causes an impedance mismatch at a boundary between typed and untyped code.
-We will clarify this point in the prose.
+
+A type-level mistake is a mistake that causes an impedance mismatch at
+a boundary between typed and untyped code.  We will clarify this point
+in the prose.
 
@@
 >   mismatch?  (This question is related to the above issue with Natural blame.)
+
 Yes, please see the start of this response.
@@
 >   to the issue with Natural blame).
+
 Thanks, we will expand on this in the prose.
@@
 > - L557: "if it does" --> "if it does not"?
+
 You're right, thanks, we will fix it.
@@
 >   "does not reliably lead to type errors"?
-This is a refinement of the arithmetic mutator that does lead to type errors, because given two integers `*` always produces an integer, but `/` may produce a rational, which have different types in Typed Racket.
-We will clarify this comment in the footnote.
+
+This is a refinement of the arithmetic mutator that does lead to type
+errors, because given two integers `*` always produces an integer, but
+`/` may produce a rational, which have different types in Typed
+Racket.  We will clarify this comment in the footnote.
 
@@
 >   what this intends.
-The measure of interesting mutants provided a standard by which we could judge potential mutators and iteratively develop an effective set of them.
-We will make this comment clearer.
+
+The measure of interesting mutants provided a standard by which we
+could judge potential mutators and iteratively develop an effective
+set of them.  We will make this comment clearer.
 
 > - What is the total number of failing trails in Transient?
-3594 out of 72192 trails fail for transient last blame, and 3591 for first.
+
+3594 out of 72192 trails fail for transient last blame, and 3591 for
+first.
 
@@
 >   to be cited?
+
 TODO
@@
 > - L1104 "deviating is a mistake":  Why?
-Our experience of ignoring blame as opposed to heeding it suggests that ignoring it is usually unwise.
+
+Our long experience of ignoring blame as opposed to heeding it
+suggests that ignoring it is usually unwise.
 
 > - L1154 "in the "fully typed" benchmarks":  Are these benchmarks on Python?
+
 The slowdowns the paper quotes here are from Python benchmarks.
@@
 >   does it intend?
-This paragraph refers to a Python translation of the Racket `Sieve` benchmark;
-it is confusingly worded, we will adjust the phrasing.
+
+This paragraph refers to a Python translation of the Racket `Sieve`
+benchmark; it is confusingly worded, we will adjust the phrasing.
 
@@
 >   sentence wants to say.  Please consider rephrasing.
+
 Thanks for these comments, we will incorporate them to improve the prose.

### 2021-04-22 writing/reviews/icfp21-rebuttal.md
Update icfp21 rebuttal with my tasks from yesterdays meeting

@@
 
-Mutators and benchmarks
-================================================================================
-
-Some more description of the mutators and benchmarks the paper uses
-would clearly be useful, so we will expand the explanation and
-examples in the final paper.  For now, we have responded to the
-reviewers specific questions inline below.
-
-Failing blame trails
-================================================================================
-
-The paper fails to clarify how Natural blame can have failing
-scenarios, so we will update it to clarify.  In short, Natural blame
-trails can by stymied by unhelpful run-time errors.
-
-Because debugging scenarios are only partially typed, some scenarios
-produce an error from the runtime (e.g. `+` receiving a non-number)
-before any of Natural's run-time type checks can discover a problem.
-
-Run-time errors do not carry blame, only stacktraces, so the only way
-to proceed (regardless of mode) is by using the stack information.  If
-the stack proves unhelpful, then the trail fails.  Such failures are
-the main source of failing blame trails in the `Natural blame` mode.
-
 Replies to individual points
@@
 11 that one problem is with errors in the ascribed types themselves.).
+
 Yes, you hit the nail on the head. 
@@
 > itself will now fail, but that changing the conditions makes certain previously excluded paths in the code feasible.
+
+Thanks for pointing this out, we will expand the explanation and examples of mutators in the final paper.
 In the specific cases of `negate-cond` and `force-cond`, these mutators lead to type-level mistakes when the program uses occurrence typing. 
@@
 > How do you arrive at the number 72192 in Section 7? It is not equal to 756 times 96.
+
 Indeed, 756 is a typo: it should be 752.
@@
 >  - Which blame assignment strategy is "good" and which is not "good"?
-TODO
+
+TODO - Christos will do it.
 
@@
 >    be nicer to address them in a clearer manner.
+
 Thanks for the feedback, we will rephrase the conclusion.
@@
 >   it distinguishes them.
-TODO
+
+The paper uses the two terms interchangeably after the introduction, because the distinction between the two does not change the evaluation method.
+Specifically, in the gradual typing setting the components are smaller than whole modules, and there are more possible intermediate types (incorporating Dyn) between untyped and fully-typed, both of which just end up making the configuration lattice larger.
+
+> - In principle, Natural blame should always point out the faulty components
+>   by the Wadler-Findler slogan.  However, the experiment shows it is not the
+>   case.  Why?  Is it the same reason as in Lazarek et al.?  (Perhaps it is due
+>   to a gap between theory and practice, but exposing a reason is crucial.)
+
+(Combining this question with the related one from below next.)
+
+> - Are failing Natural blame trails produced even for programs with impedance
+>   mismatch?  (This question is related to the above issue with Natural blame.)
+
+We agree that the paper needs a bit more explanation to clarify how Natural blame can have failing
+scenarios, we will update it in the final paper.  In short, Natural blame
+trails can by stymied by unhelpful run-time errors.
+
+Because debugging scenarios are only partially typed, some scenarios
+produce an error from the runtime (e.g. `+` receiving a non-number)
+before any of Natural's run-time type checks can discover a problem.
+Run-time errors do not carry blame, only stacktraces, so the only way
+to proceed (regardless of mode) is by using the stack information.  If
+the stack proves unhelpful, then the trail fails.
 
@@
 >   class:super.
+
+Thanks for the comment, we will expand the explanation and examples of mutators in the final paper.
 In the specific case of `deletion`, the other most common place that
->   this mutator applies are so-called "implicit begins": places where
->   sequences are allowed implicitly by enclosing forms like `define`
->   or `cond`. 
-So here are two more examples of `deletion` mutations:
+this mutator applies are so-called "implicit begins": places where
+sequences are allowed implicitly by enclosing forms like `define`
+or `cond`.
+Here are two more examples of `deletion` mutations:
 ```
@@
 >     on the usage of blame maps in Transient would be needed.
-TODO
+
+We will improve the prose in this section to make it clearer, thank you for pointing it out.
+The paper is referring here to the incomplete population of the blame map described by Greenman et al. ([11]: OOPSLA'19) .
+For more details, there is a parallel submission by the same set of authors describing the implementation of Transient in depth.
 
@@
 
-> - Are failing Natural blame trails produced even for programs with impedance
->   mismatch?  (This question is related to the above issue with Natural blame.)
-
-Yes, please see the start of this response.
-
 > - L556: It would be helpful to describe how to extend trails and how to
@@
 
-TODO
+The best reference we might use here is Kahnemann's popular book entitled "Thinking Fast and Slow", which explains the problem well.
+Notably with respect to this work, the book criticizes classical (mathematical) economic theory for assuming that people act rationally all of the time, which demonstrates how our work is complementary to human studies; both are necessary.
 

### 2021-04-22 writing/reviews/icfp21-rebuttal.md
Small touchups of point-by-point responses

@@
 Yes, you hit the nail on the head. 
-One of the central parts of the rational programmer is that it fixes decision of how to respond to a given scenario; 
+One of the central parts of the rational programmer is that it fixes the decision of how to respond to a given scenario; 
 so evaluating different ways of taking action boils down to defining and testing the corresponding rational programmers.
@@
 That would make the most sense to me, but you do not seem to come out and say so.
+
 Yes you're right. Thanks, we will clarify this in the prose.
@@
 Thanks for the comment, we will expand the explanation and examples of mutators in the final paper.
-In the specific case of `deletion`, the other most common place that
-this mutator applies are so-called "implicit begins": places where
+In the specific case of `deletion`, the most common place that this mutator applies
+(beyond the literal `begin` expressions the example illustrates)
+are so-called "implicit begins": places where
 sequences are allowed implicitly by enclosing forms like `define`
 or `cond`.
-Here are two more examples of `deletion` mutations:
+Here are two more examples of `deletion` mutations for these situations:
 ```
@@
 
-In the case of `class:super`, the mutator simply replaces any occurrence of `super-new` in the body of a class definition with `void`.
+In the case of `class:super`, the mutator simply removes calls to `super-new`, causing them to no longer instantiate the superclass.
 
@@
 crucial to analyze the mutators in light of these features; perhaps
-adding a breakdown of features by benchmark to section 6.3 would help
+adding a breakdown of features per benchmark to section 6.3 would help
 clarify this?
@@
 
-The benchmarks in figure 5 are the ten largest dependency graphs of the GTP suite. 
+The benchmarks in figure 5 are the ten with the largest dependency graphs of the GTP suite.
 We will rephrase L646-648 to make this clearer.

### 2021-04-22 writing/reviews/icfp21-rebuttal.md
Minor revisions of point-by-point replies

@@
 
-The benchmarks employ many of Typed Racket's sophisticated features,
-most notably (from L641) occurrence typing, types for mutable and
-immutable data structures, types for first-class classes and objects,
-and types for Racket's numeric tower.  We absolutely agree that it is
-crucial to analyze the mutators in light of these features; perhaps
-adding a breakdown of features per benchmark to section 6.3 would help
-clarify this?
+Broadly speaking, this phrase refers to features beyond basic functions and
+primitive data. In terms of this paper's evaluation, the benchmarks employ many
+of Typed Racket's sophisticated features (with those listed on L641 being the
+most notable).
+
+We absolutely agree that it is crucial to analyze the mutators in light of these
+features; perhaps adding a breakdown of features per benchmark to the analysis
+of section 6.3 would help clarify this?
 

### 2021-04-22 writing/reviews/icfp21-rebuttal.md
rebuttal revisions

--- 2021-04-22 writing/reviews/icfp21-rebuttal.md
@@
-The first part of our response addresses big themes in the reviews;
-the second part addresses all remaining concerns individually.
+The first part of our response addresses one selected theme from each review;
+the second part addresses all remaining comments and questions individually.
 
-What do the results say about blame for gradual typing in practice?
+What the experimental results say about the theory vs the practice of blame
 ================================================================================
 
-The introduction says that the experimental results lead to the
-following conclusions about blame _in practice_:
+Theoretical work that compares different checking and blame strategies for
+gradual types (Greenman et al. ICFP 2018 and OOPSLA 2019) declares a clear
+winner: Natural. Natural is the only semantics that is strongly type sound,
+a complete monitor and, most importantly in the context of this paper, has
+sound and complete blame assignment. The latter means that Natural blames
+all and only those components that are relevant to a type impedance. In
+contrast Transient may fail to blame some relevant components and may
+blame irrelevant ones.
 
-1. Blame helps locate impedance mismatches, though not immensely more than simple stacktraces, and
-2. Natural blame helps more than Transient, and Transient more than Erasure's stacktraces, *but*
-3. Natural blame is only marginally better than Transient in practice, and both not immensely better than Erasure's stacktraces.
+At first glance, the experiment of this paper seems to validate the
+theoretical conclusions:
 
-The phrase "in practice" refers to the way that blame behaves in real
-programs with an impedance mismatch, using real implementations of
-each semantics.
+1. Natural's blame assignment is as good as the theory predicts; 
+Natural blame helps locate impedance mismatches in almost all 
+debugging scenarios. The only exceptions are (i) due to bugs in Typed
+Racket and (ii) debugging scenarios that result in runtime errors with
+unhelpful stacktraces from the underlying checks of Racket rather than 
+Typed Racket's checks.
 
-In light of the data in figures 8, 9, and 10, the benefits of blame in
-practice do not measure up to the expectations established by theory.
-Concretely, blame does provide useful debugging information, but the
-results suggest that in practice blame is only necessary to debug (as
-compared with simple exceptions) in a small fraction of scenarios.
+2. In general, Natural blame is better that Transient blame and both are
+better than Erasure's stacktraces.
 
-- The theory suggests that Natural's blame should be more precise than
-  Transient because Natural precisely tracks all possible interactions
-  while Transient approximates (see [11]: Greenman et al. OOPSLA'19).
+However, a closer inspection of the empirical results reveals that 
+they paint a picture with gray zones that the black and white theoretical 
+results are  not refined enough to capture:
 
-- In practice, Natural's blame has utility comparable to Transient's.
+1. There is a small but substantial number of debugging scenarios where 
+Natural blame is not helpful but Transient blame is. These scenarios are
+not all due to bugs in Typed Racket. Hence, the theory that deems natural
+blame superior to that of Natural cannot explain them away.
 
-The experiment in this paper reveals that mismatch, but it necessarily
-considers only a limited view of the space of possible impedance
-mismatch scenarios that might arise in reality.
+2. While the theory claims that language designers should select Natural and
+   its blames for maximum benefits, the results of the experiment indicate
+   that Transient and Erasure are good enough for locating impedance
+   mismatches in most cases. Hence, given the cost of Natural, the theory's answer
+   fails to provide good enough guidance to language designers.
+
+In sum the experiment in this paper reveals the mismatch between the theory and the practice
+of blame. The latter refers to blame's actual utility in real
+programs that run in a real implementation rather than artificial
+examples in an idealized model.
+
+As a result, there are two intertwined directions forward:
+
+1. Revisiting the theory of blame to match the practice of blame.
+
+2. Analyzing further the practice of blame beyond the scenarios this paper
+   considers.
 
@@
 
-<!-- As another example, the paper only considers scenarios known to
-be detectable by all three systems, but theory makes it clear that
-Natural and Erasure can detect some scenarios that Erasure cannot. -->
-
-<!-- How common are those scenarios for impedance mismatch bugs? -->
-<!-- Does blame provide more benefit than simple exceptions in those scenarios? -->
-
-Additional experimentation is necessary to develop an understanding of these different spaces.
-
-The Rational Programmer vs Lazarek et al.
+What is the origin of the rational programmer
 ================================================================================
 
-Lazarek et al. (POPL'20) provide a kernel insight, which this paper
-develops into the novel idea of the Rational Programmer. It is the
-first paper to explain the idea properly and use the phrase. We
-therefore consider it its first contribution.
-
-The key distinction is the idea of the rational programmer as an
-entity to study and how it can be parameterized by modes to enable
-cross-system comparison.
+Lazarek et al. (POPL'20) provide the kernel insight that one can simulate
+an idealized programmer on a large corpus of programs to test the behavior
+of a language feature. This paper hones this kernel into an experimental
+framework that allows the isolation of confounding factors and,
+furthermore, the comparison between different semantics for a feature in
+the form of different but comparable experimental modes. The paper
+christens the experimental framework the Rational Programmer.
 
@@
 11 that one problem is with errors in the ascribed types themselves.).
-
 Yes, you hit the nail on the head. 
@@
 > itself will now fail, but that changing the conditions makes certain previously excluded paths in the code feasible.
-
-Thanks for pointing this out, we will expand the explanation and examples of mutators in the final paper.
 In the specific cases of `negate-cond` and `force-cond`, these mutators lead to type-level mistakes when the program uses occurrence typing. 
@@
 > How do you arrive at the number 72192 in Section 7? It is not equal to 756 times 96.
-
 Indeed, 756 is a typo: it should be 752.
@@
 >  - Which blame assignment strategy is "good" and which is not "good"?
-
-TODO - Christos will do it.
+TODO
 
@@
 >    be nicer to address them in a clearer manner.
-
 Thanks for the feedback, we will rephrase the conclusion.
@@
 >   it distinguishes them.
-
-The paper uses the two terms interchangeably after the introduction, because the distinction between the two does not change the evaluation method.
-Specifically, in the gradual typing setting the components are smaller than whole modules, and there are more possible intermediate types (incorporating Dyn) between untyped and fully-typed, both of which just end up making the configuration lattice larger.
-
-> - In principle, Natural blame should always point out the faulty components
->   by the Wadler-Findler slogan.  However, the experiment shows it is not the
->   case.  Why?  Is it the same reason as in Lazarek et al.?  (Perhaps it is due
->   to a gap between theory and practice, but exposing a reason is crucial.)
-
-(Combining this question with the related one from below next.)
-
-> - Are failing Natural blame trails produced even for programs with impedance
->   mismatch?  (This question is related to the above issue with Natural blame.)
-
-We agree that the paper needs a bit more explanation to clarify how Natural blame can have failing
-scenarios, we will update it in the final paper.  In short, Natural blame
-trails can by stymied by unhelpful run-time errors.
-
-Because debugging scenarios are only partially typed, some scenarios
-produce an error from the runtime (e.g. `+` receiving a non-number)
-before any of Natural's run-time type checks can discover a problem.
-Run-time errors do not carry blame, only stacktraces, so the only way
-to proceed (regardless of mode) is by using the stack information.  If
-the stack proves unhelpful, then the trail fails.
+TODO
 
@@
 >   class:super.
-
-Thanks for the comment, we will expand the explanation and examples of mutators in the final paper.
 In the specific case of `deletion`, the other most common place that
-this mutator applies are so-called "implicit begins": places where
-sequences are allowed implicitly by enclosing forms like `define`
-or `cond`.
-Here are two more examples of `deletion` mutations:
+>   this mutator applies are so-called "implicit begins": places where
+>   sequences are allowed implicitly by enclosing forms like `define`
+>   or `cond`. 
+So here are two more examples of `deletion` mutations:
 ```
@@
 >     on the usage of blame maps in Transient would be needed.
-
-We will improve the prose in this section to make it clearer, thank you for pointing it out.
-The paper is referring here to the incomplete population of the blame map described by Greenman et al. ([11]: OOPSLA'19) .
-For more details, there is a parallel submission by the same set of authors describing the implementation of Transient in depth.
+TODO
 
@@
 
+> - Are failing Natural blame trails produced even for programs with impedance
+>   mismatch?  (This question is related to the above issue with Natural blame.)
+
+Yes, please see the start of this response.
+
 > - L556: It would be helpful to describe how to extend trails and how to
@@
 
-The best reference we might use here is Kahnemann's popular book entitled "Thinking Fast and Slow", which explains the problem well.
-Notably with respect to this work, the book criticizes classical (mathematical) economic theory for assuming that people act rationally all of the time, which demonstrates how our work is complementary to human studies; both are necessary.
+TODO
 

### 2021-04-22 writing/reviews/icfp21-rebuttal.md

@@
 christens the experimental framework the Rational Programmer.
+=======
 
@@
 11 that one problem is with errors in the ascribed types themselves.).
+
 Yes, you hit the nail on the head. 
-One of the central parts of the rational programmer is that it fixes decision of how to respond to a given scenario; 
+One of the central parts of the rational programmer is that it fixes the decision of how to respond to a given scenario; 
 so evaluating different ways of taking action boils down to defining and testing the corresponding rational programmers.
@@
 >   class:super.
-In the specific case of `deletion`, the other most common place that
->   this mutator applies are so-called "implicit begins": places where
->   sequences are allowed implicitly by enclosing forms like `define`
->   or `cond`. 
-So here are two more examples of `deletion` mutations:
+
+Thanks for the comment, we will expand the explanation and examples of mutators in the final paper.
+In the specific case of `deletion`, the most common place that this mutator applies
+(beyond the literal `begin` expressions the example illustrates)
+are so-called "implicit begins": places where
+sequences are allowed implicitly by enclosing forms like `define`
+or `cond`.
+Here are two more examples of `deletion` mutations for these situations:
 ```
@@
 
-The benchmarks employ many of Typed Racket's sophisticated features,
-most notably (from L641) occurrence typing, types for mutable and
-immutable data structures, types for first-class classes and objects,
-and types for Racket's numeric tower.  We absolutely agree that it is
-crucial to analyze the mutators in light of these features; perhaps
-adding a breakdown of features by benchmark to section 6.3 would help
-clarify this?
+Broadly speaking, this phrase refers to features beyond basic functions and
+primitive data. In terms of this paper's evaluation, the benchmarks employ many
+of Typed Racket's sophisticated features (with those listed on L641 being the
+most notable).
+
+We absolutely agree that it is crucial to analyze the mutators in light of these
+features; perhaps adding a breakdown of features per benchmark to the analysis
+of section 6.3 would help clarify this?
 

### 2021-04-22 writing/reviews/icfp21-rebuttal.md
rebuttal edits

@@
-The first part of our response addresses one selected theme from each review;
-the second part addresses all remaining comments and questions individually.
+The first part of our response addresses one major comment from each
+reviewer.
+
+What is the origin of the rational programmer
+================================================================================
+
+Reviewer A says:
+> the paper talks in the introduction of its top-level innovation being the
+> idea of the rational programmer. But then they go on to say that the idea
+> is actually taken from Lazarek. Which is it then?
+
+Lazarek et al. (POPL'20) provide the kernel insight that one can simulate
+an idealized programmer on a large corpus of programs to test the behavior
+of a language feature. This paper hones this kernel into an experimental
+framework. With the framework we can now:
+1. isolate confounding factors. 
+2. compare different semantics for gradual type boundaries in the form of 
+   experimental modes.  
+The paper christens this experimental framework the Rational Programmer.
 
@@
 
+Reviewer B says:
+> The conclusions of the second contribution described in the introduction
+>  (around lines 41-103) are confusing for me.
+
 Theoretical work that compares different checking and blame strategies for
 gradual types (Greenman et al. ICFP 2018 and OOPSLA 2019) declares a clear
-winner: Natural. Natural is the only semantics that is strongly type sound,
-a complete monitor and, most importantly in the context of this paper, has
-sound and complete blame assignment. The latter means that Natural blames
-all and only those components that are relevant to a type impedance. In
-contrast Transient may fail to blame some relevant components and may
-blame irrelevant ones.
+winner: Natural. Natural is the only semantics that is strongly type
+sound, a complete monitor and, most importantly in the context of this
+paper, Natural blames all and only those components that are relevant to a
+type impedance mismatch (sound and complete blame). In contrast Transient
+may fail to blame some relevant components and may blame irrelevant ones.
+Hence, theory predicts that Natural is superior to Transient and Transient
+is superior to Erasure.
 
-At first glance, the experiment of this paper seems to validate the
-theoretical conclusions:
+At first glance, our experiments seem to validate these predictions:
 
-1. Natural's blame assignment is as good as the theory predicts; 
-Natural blame helps locate impedance mismatches in almost all 
-debugging scenarios. The only exceptions are (i) due to bugs in Typed
-Racket and (ii) debugging scenarios that result in runtime errors with
-unhelpful stacktraces from the underlying checks of Racket rather than 
-Typed Racket's checks.
+1. When Natural issues blame it is helpful in almost all debugging
+   scenarios. The problem is that many debugging scenarios result in
+   runtime errors with unhelpful stacktraces from the underlying checks of
+   Racket rather than Typed Racket boundary checks.
 
-2. In general, Natural blame is better that Transient blame and both are
-better than Erasure's stacktraces.
+2. Natural blame looks better that Transient blame and both look better
+   than Erasure's stacktraces.
 
-However, a closer inspection of the empirical results reveals that 
-they paint a picture with gray zones that the black and white theoretical 
-results are  not refined enough to capture:
+However, a close inspection of the empirical data reveals that 
+ a gray-shaded picture not the black and white print of theory:
 
-1. There is a small but substantial number of debugging scenarios where 
-Natural blame is not helpful but Transient blame is. These scenarios are
-not all due to bugs in Typed Racket. Hence, the theory that deems natural
-blame superior to that of Natural cannot explain them away.
+1. There are some debugging scenarios where Natural blame is not helpful
+   but Transient blame is. The theory cannot explain these scenarios.
 
-2. While the theory claims that language designers should select Natural and
-   its blames for maximum benefits, the results of the experiment indicate
-   that Transient and Erasure are good enough for locating impedance
-   mismatches in most cases. Hence, given the cost of Natural, the theory's answer
-   fails to provide good enough guidance to language designers.
+2. While the theory claims that language designers should select Natural
+   and its blame for maximum benefits, the data indicates that both
+   Transient blame and Erasure stacktrace often suffice. Hence, given the
+   cost of Natural (Takikawa et al. POPL 2016), the theory's suggestion
+   has shown to be of questionable value.
 
-In sum the experiment in this paper reveals the mismatch between the theory and the practice
-of blame. The latter refers to blame's actual utility in real
-programs that run in a real implementation rather than artificial
-examples in an idealized model.
+In sum our experiments reveal the mismatch between the theory and the
+practice of blame. The latter refers to blame's actual utility in real
+programs that run in a real implementation rather than artificial examples
+in an idealized model.
 
-As a result, there are two intertwined directions forward:
+As a result, there are two intertwined directions going forward:
 
@@
 
-2. Analyzing further the practice of blame beyond the scenarios this paper
-   considers.
-
-Section 11 points to another space of scenarios that deserves
-analysis: scenarios where mismatches arise due to mistakes in type
-annotations rather than (or as well as) code.  Indeed, review B
-astutely points out two more spaces worth exploring:
-
-- programs where some components _cannot be typed_
-- non-deterministic programs, which inject ephemeral bugs.
-
-What is the origin of the rational programmer
-================================================================================
-
-Lazarek et al. (POPL'20) provide the kernel insight that one can simulate
-an idealized programmer on a large corpus of programs to test the behavior
-of a language feature. This paper hones this kernel into an experimental
-framework that allows the isolation of confounding factors and,
-furthermore, the comparison between different semantics for a feature in
-the form of different but comparable experimental modes. The paper
-christens the experimental framework the Rational Programmer.
-=======
+2. Empirically analyzing the practice of blame beyond the scenarios this
+   paper considers. Section 11 points to another space of scenarios that
+   deserves analysis: scenarios where mismatches arise due to mistakes in
+   type annotations rather than (or as well as) code.  Indeed, review B
+   astutely points out two more directions worth exploring in this
+   space:
+    - programs where some components _cannot be typed_;
+    - non-deterministic programs.
 
@@
 
-Yes, you hit the nail on the head. 
-One of the central parts of the rational programmer is that it fixes the decision of how to respond to a given scenario; 
-so evaluating different ways of taking action boils down to defining and testing the corresponding rational programmers.
+Yes, you hit the nail on the head.  One of the central parts of the
+rational programmer is that it fixes the decision of how to respond to a
+given scenario; so evaluating different ways of taking action boils down
+to defining and testing the corresponding rational programmers.
 
@@
 
-Thanks for pointing this out, we will expand the explanation and examples of mutators in the final paper.
-In the specific cases of `negate-cond` and `force-cond`, these mutators lead to type-level mistakes when the program uses occurrence typing. 
-For example, they could both lead to a mistake in a program like this:
+Thanks for pointing this out, we will expand the explanation and examples
+of mutators in the final paper.  In the specific cases of `negate-cond`
+and `force-cond`, these mutators lead to type-level mistakes when the
+program uses occurrence typing.  For example, they could both lead to a
+mistake in a program like this:
+
 ```
@@
 
-TODO - Christos will do it.
+In this context, good means Natural and Transient. We will rephrase.
 
@@
 
-The paper uses the two terms interchangeably after the introduction, because the distinction between the two does not change the evaluation method.
-Specifically, in the gradual typing setting the components are smaller than whole modules, and there are more possible intermediate types (incorporating Dyn) between untyped and fully-typed, both of which just end up making the configuration lattice larger.
+The paper uses the two terms interchangeably after the introduction,
+because the distinction between the two does not change the evaluation
+method.  Specifically, in the gradual typing setting the components are
+smaller than whole modules, and there are more possible intermediate types
+(incorporating Dyn) between untyped and fully-typed, both of which just
+end up making the configuration lattice much much larger.
 
@@
 
-We agree that the paper needs a bit more explanation to clarify how Natural blame can have failing
-scenarios, we will update it in the final paper.  In short, Natural blame
-trails can by stymied by unhelpful run-time errors.
-
-Because debugging scenarios are only partially typed, some scenarios
-produce an error from the runtime (e.g. `+` receiving a non-number)
-before any of Natural's run-time type checks can discover a problem.
-Run-time errors do not carry blame, only stacktraces, so the only way
-to proceed (regardless of mode) is by using the stack information.  If
-the stack proves unhelpful, then the trail fails.
+We agree that the paper needs a bit more explanation to clarify how
+Natural blame can have failing scenarios, we will update it in the final
+paper.  In short, Natural blame trails can by stymied by unhelpful
+run-time errors from Racket. Because debugging scenarios are only
+partially typed, some scenarios produce an error from the runtime (e.g.
+`+` receiving a non-number) before any of Natural's run-time type checks
+can discover a problem.  Runtime errors do not carry blame, only
+stacktraces.
 
@@
 
-Thanks for the comment, we will expand the explanation and examples of mutators in the final paper.
-In the specific case of `deletion`, the most common place that this mutator applies
-(beyond the literal `begin` expressions the example illustrates)
-are so-called "implicit begins": places where
-sequences are allowed implicitly by enclosing forms like `define`
-or `cond`.
-Here are two more examples of `deletion` mutations for these situations:
+Thanks for the comment, we will expand the explanation and examples of
+mutators in the final paper.  In the specific case of `deletion`, the most
+common place that this mutator applies (beyond the literal `begin`
+expressions the example illustrates) are so-called "implicit begins":
+places where sequences are allowed implicitly by enclosing forms like
+`define` or `cond`.  Here are two more examples of `deletion` mutations
+for these situations:
+
 ```
@@
 
-In the case of `class:super`, the mutator simply removes calls to `super-new`, causing them to no longer instantiate the superclass.
+In the case of `class:super`, the mutator simply removes calls to
+`super-new`, causing the class to no longer instantiate the superclass.
 
@@
 
-The bottom of the configuration lattice is completely untyped, so that
-is a possible starting scenario to be sampled.
+The debugging process can start with the fully untyped program but it
+doesn't have to. The lattice contains the completely untyped program (bottom
+configuration) and we sample from all the configurations where
+the buggy module is untyped (including the fully untyped one).
 
@@
 
-The term refers to the soundness defined by [11] (Greenman et
-al. OOPSLA'19), but we agree that for clarity the paper should define
-it in addition to the reference.  Intuitively, a blame system is sound
-if it can only blame components that have had control of the value
-witnessing a violation.  In other words, the system cannot blame
-arbitrary components unrelated to the witness of a contract violation.
+The term refers to blame soundness from Greenman et al. (OOPSLA'19), but
+we agree that for clarity the paper should define it in addition to the
+reference.  Intuitively, a blame system is sound if it can only blame
+components that have had control of the value witnessing a violation.  In
+other words, the system cannot blame arbitrary components unrelated to the
+witness of a contract violation.
 
@@
 
-We will improve the prose in this section to make it clearer, thank you for pointing it out.
-The paper is referring here to the incomplete population of the blame map described by Greenman et al. ([11]: OOPSLA'19) .
-For more details, there is a parallel submission by the same set of authors describing the implementation of Transient in depth.
+ We will improve the prose in this paper to clarify. Thank you for
+ pointing this out.  The paper is referring here to the incomplete
+ population of the blame map described by Greenman et al. ([11]:
+ OOPSLA'19).
+
+ For the details of how we have implemented the blame map and its
+ population, there is a parallel experience report submission by the same
+ set of authors describing the implementation of Transient in depth.
 
@@
 
-Broadly speaking, this phrase refers to features beyond basic functions and
-primitive data. In terms of this paper's evaluation, the benchmarks employ many
-of Typed Racket's sophisticated features (with those listed on L641 being the
-most notable).
-
-We absolutely agree that it is crucial to analyze the mutators in light of these
-features; perhaps adding a breakdown of features per benchmark to the analysis
-of section 6.3 would help clarify this?
+The features we have in mind appear in line 641. We will add a breakdown
+of features per benchmark to the analysis of section 6.3. Our mutators do
+target all of these.
 
@@
 
-The benchmarks in figure 5 are the ten with the largest dependency graphs of the GTP suite.
-We will rephrase L646-648 to make this clearer.
+The benchmarks in figure 5 are the ten with the largest dependency graphs
+of the GTP suite. We selected those because small benchmarks have
+short blame trails. We will rephrase L646-648 to make this clearer.
 
@@
 
-This was a mistake typesetting the paper, we will fix it.
+This was a mistake typesetting the paper, we have fixed it. It didn't
+affect the length of the paper
 
@@
 This is a refinement of the arithmetic mutator that does lead to type
-errors, because given two integers `*` always produces an integer, but
-`/` may produce a rational, which have different types in Typed
-Racket.  We will clarify this comment in the footnote.
+errors, because given two integers `*` always produces an integer, but `/`
+may produce a rational, which have different types in Typed Racket's
+sophisticated numeric tower (St-Amour et al. PADL 2011).  We will clarify
+this comment in the footnote.
 
@@
 
-The best reference we might use here is Kahnemann's popular book entitled "Thinking Fast and Slow", which explains the problem well.
-Notably with respect to this work, the book criticizes classical (mathematical) economic theory for assuming that people act rationally all of the time, which demonstrates how our work is complementary to human studies; both are necessary.
+The best reference we might use here is Kahnemann's popular book entitled
+"Thinking Fast and Slow", which explains the problem well.  Notably with
+respect to this work, the book criticizes classical (mathematical)
+economic theory for assuming that people act rationally all of the time,
+which demonstrates how our work is complementary to human studies and
+theory. All are necessary.
 

### 2021-04-23 writing/reviews/icfp21-rebuttal.md
small edits

@@
 framework. With the framework we can now:
-1. isolate confounding factors. 
+
+1. isolate confounding factors;
 2. compare different semantics for gradual type boundaries in the form of 
    experimental modes.
+
 The paper christens this experimental framework the Rational Programmer.
+We have already revised the prose accordingly.
 
@@
 gradual types (Greenman et al. ICFP 2018 and OOPSLA 2019) declares a clear
-winner: Natural. Natural is the only semantics that is strongly type
-sound, a complete monitor and, most importantly in the context of this
-paper, Natural blames all and only those components that are relevant to a
-type impedance mismatch (sound and complete blame). In contrast Transient
-may fail to blame some relevant components and may blame irrelevant ones.
-Hence, theory predicts that Natural is superior to Transient and Transient
-is superior to Erasure.
+winner: Natural.
 
-At first glance, our experiments seem to validate these predictions:
+- _Natural_ is the only semantics that is type sound in the conventional sense,
+  completely monitors boundaries, and, most importantly here, blames all and
+  only those components that are relevant to a type impedance mismatch (sound
+  and complete blame).
 
-1. When Natural issues blame it is helpful in almost all debugging
-   scenarios. The problem is that many debugging scenarios result in
-   runtime errors with unhelpful stacktraces from the underlying checks of
-   Racket rather than Typed Racket boundary checks.
+- By contrast, _Transient_ may fail to blame some relevant components and may
+  blame irrelevant ones.
 
-2. Natural blame looks better that Transient blame and both look better
-   than Erasure's stacktraces.
+- Erasure doesn't even assign checks to type boundaries.
+
+Hence, theory predicts that Natural is superior to Transient and Transient is
+superior to Erasure.
+
+Our experiments only appear to validate these predictions:
+
+1. When Natural issues blame it is helpful in almost all debugging scenarios.
+
+2. Natural blame looks better that Transient blame and both look better than
+   Erasure's stacktraces.
 
@@
 
-1. There are some debugging scenarios where Natural blame is not helpful
-   but Transient blame is. The theory cannot explain these scenarios.
+1. One problem is that many debugging scenarios result in runtime errors with
+   unhelpful stacktraces from the underlying checks of Racket rather than Typed
+   Racket boundary checks.
 
-2. While the theory claims that language designers should select Natural
-   and its blame for maximum benefits, the data indicates that both
-   Transient blame and Erasure stacktrace often suffice. Hence, given the
-   cost of Natural (Takikawa et al. POPL 2016), the theory's suggestion
-   has shown to be of questionable value.
+1. Additionally, for some debugging scenarios Natural blame is not helpful but
+   Transient blame is. The theory cannot explain these scenarios at this point.
 
-In sum our experiments reveal the mismatch between the theory and the
-practice of blame. The latter refers to blame's actual utility in real
-programs that run in a real implementation rather than artificial examples
-in an idealized model.
+2. Finally, while the theory claims that language designers should select
+   Natural and its blame for maximum benefits, the data indicates that both
+   Transient blame information and Erasure stacktraces often suffice.  Given the
+   large cost of Natural (Takikawa et al. POPL 2016), theory has has shown
+   itself to be of questionable value -- at this time.
 
-As a result, there are two intertwined directions going forward:
+In sum our experiments reveal a mismatch between theory and practice where blame
+may or may not matter for real programs running on real implementations. This is
+in contrast to theoretical papers, which symbol-push small terms thru a reduction 
+semantics in a tiny model of PLs (even if these models exhibit all essential 
+attributes).
+
+As a result, we expect to invest in two intertwined research efforts going forward:
 
@@
    deserves analysis: scenarios where mismatches arise due to mistakes in
-   type annotations rather than (or as well as) code.  Indeed, review B
-   astutely points out two more directions worth exploring in this
-   space:
+   type annotations rather than (or as well as) code.
+
+   Indeed, the review astutely points out two more directions worth exploring:
+
     - programs where some components _cannot be typed_;

### 2021-04-23 writing/reviews/icfp21-rebuttal.md
Small typo and grammar fixes, and revive quote of B's main concerns

@@
-The first part of our response addresses one major comment from each
-reviewer.
+The first part of our response addresses one major comment from each review.
 
-What is the origin of the rational programmer
+The origin of the rational programmer
 ================================================================================
 
-Reviewer A says:
+Review A says:
 > the paper talks in the introduction of its top-level innovation being the
@@
 
-What the experimental results say about the theory vs the practice of blame
+What the results say about the theory vs practice of blame
 ================================================================================
 
-Reviewer B says:
+Review B says:
 > The conclusions of the second contribution described in the introduction
@@
 
-- Erasure doesn't even assign checks to type boundaries.
+- _Erasure_ doesn't even assign checks to type boundaries.
 
@@
 
-However, a close inspection of the empirical data reveals that 
- a gray-shaded picture not the black and white print of theory:
+However, a close inspection of the empirical data reveals a gray-shaded picture
+instead of the black and white print of theory:
 
@@
 
-In sum our experiments reveal a mismatch between theory and practice where blame
+In sum, our experiments reveal a mismatch between theory and practice where blame
 may or may not matter for real programs running on real implementations. This is
-in contrast to theoretical papers, which symbol-push small terms thru a reduction 
+in contrast to theoretical papers, which symbol-push small terms through a reduction
 semantics in a tiny model of PLs (even if these models exhibit all essential
@@
 ===========================================================================
+
+> The table in section 2 looks rather ugly
+
+We will find another way to present the information.
 
@@
 
+>  - Why does the paper conclude that "the existing theory does not predict
+>    practice properly"?  Figure 8 shows that the theory (Natural) works well.
+>  - What does the paper mean by "The existing practice may need additional
+>    experiments"?  I have not found any evidence that confirms this claim in the
+>    paper.
+>  - What does "practice" means here?
+
+We hope the start of the response addresses these questions adequately.
+
 >  - At first glance, the conclusion in lines 55-56 (starting with "Second, ...")
@@
 (incorporating Dyn) between untyped and fully-typed, both of which just
-end up making the configuration lattice much much larger.
+end up making the configuration lattice much larger.
 
@@
 
-We agree that the paper needs a bit more explanation to clarify how
-Natural blame can have failing scenarios, we will update it in the final
-paper.  In short, Natural blame trails can by stymied by unhelpful
-run-time errors from Racket. Because debugging scenarios are only
-partially typed, some scenarios produce an error from the runtime (e.g.
-`+` receiving a non-number) before any of Natural's run-time type checks
-can discover a problem.  Runtime errors do not carry blame, only
-stacktraces.
+We agree that the paper needs a bit more explanation to clarify how Natural
+blame can have failing scenarios, we will update it in the final paper. In
+short, Natural blame trails can by stymied by unhelpful errors from Racket's
+runtime. Because debugging scenarios are only partially typed, some scenarios
+produce an error from the runtime (e.g. `+` receiving a non-number) before any
+of Natural's run-time type checks can discover a problem. Errors from the
+runtime do not carry blame, only stacktraces.
 
@@
 doesn't have to. The lattice contains the completely untyped program (bottom
-configuration) and we sample from all the configurations where
+configuration) and we sample trail roots from all the configurations where
 the buggy module is untyped (including the fully untyped one).
@@
 
- We will improve the prose in this paper to clarify. Thank you for
- pointing this out.  The paper is referring here to the incomplete
- population of the blame map described by Greenman et al. ([11]:
- OOPSLA'19).
+We will improve the prose in this paper to clarify. Thank you for pointing this
+out. The paper is referring here to the incomplete population of the blame map
+described by Greenman et al. (OOPSLA'19).
 
- For the details of how we have implemented the blame map and its
- population, there is a parallel experience report submission by the same
- set of authors describing the implementation of Transient in depth.
+For the details of how we have implemented the blame map and its population,
+there is a parallel experience report submission by the same set of authors
+describing the implementation of Transient in depth.
 
@@
 
-The features we have in mind appear in line 641. We will add a breakdown
-of features per benchmark to the analysis of section 6.3. Our mutators do
-target all of these.
+The features we have in mind appear in line 641. We will add a breakdown of
+features per benchmark to the analysis of section 6.3 to clarify that our
+mutators target all of these.
 
@@
 You are absolutely right, we will clarify these assumptions about the
-benchmarks in the paper. They are inherited from the several
+benchmarks in the paper. They are inherited from the other
 publications on the gradual benchmark suite.

### 2021-04-23 writing/reviews/icfp21-rebuttal.md
Revert to old "in sum" paragraph

@@
 
-In sum, our experiments reveal a mismatch between theory and practice where blame
-may or may not matter for real programs running on real implementations. This is
-in contrast to theoretical papers, which symbol-push small terms through a reduction
-semantics in a tiny model of PLs (even if these models exhibit all essential
-attributes).
+In sum, our experiments reveal a mismatch between the theory and the
+practice of blame, where practice refers to blame's actual utility in real
+programs that run in a real implementation rather than artificial examples
+in an idealized model.
 

### 2021-04-23 writing/reviews/icfp21-rebuttal.md
reviewer -> review

--- 2021-04-23 writing/reviews/icfp21-rebuttal.md
@@
-The first part of our response addresses one major comment from each
-reviewer.
+The first part of our response addresses one major comment from each review.
 
@@
 
-Reviewer A says:
+Review A says:
 > the paper talks in the introduction of its top-level innovation being the
@@
 
-Reviewer B says:
+Review B says:
 > The conclusions of the second contribution described in the introduction

### 2021-04-24 writing/reviews/icfp21-rebuttal.md
Add icfp21 review #C

--- 2021-04-23 writing/reviews/icfp21-rebuttal.md
@@
 Thanks for these comments, we will incorporate them to improve the prose.
+
+Review #45C
+===========================================================================
+
+> My only concern at this point is that we need to believe that the authors have
+> implemented the Transiet semantics correctly in Racket, though no details are
+> offered. The paper refers to another submission, so perhaps some details can
+> be found in there.
+
+TODO
+
+> I am wondering whether the approach that the paper calls Transient First could
+> be regarded as Transient Early, since it points to an earlier part of the
+> code. Similarly Transient Last could be Transient Late, but you would have to
+> see whether this terminology really fits.
+
+TODO
+
+> The only thing that I am concerned about at this point is that, later in the
+> paper, it is shown that "most cases the programmer need to type a single
+> module to debug a scenario," which makes me wonder whether the authors did not
+> generate difficult enough scenarios. If that is the case, what is the problem?
+
+TODO
+
+> My only issue at this point is that the reader is left with numnbers only,
+> while I think a reader would like to see selected examples, for example about
+> which blame labels have been provided by Natural and which by Transient, so to
+> see that one approach took a longer path, as well as similar examples the
+> reader can learn from. Numbers do not seem to teach the whole story in this
+> part of the paper.
+
+TODO

### 2021-04-24 writing/reviews/icfp21-rebuttal.md
Add responses to review C, touch up the introductory reply

@@
 
-1. isolate confounding factors;
+1. isolate confounding factors, such as whether the utility of blame is due
+   to blame tracking or just due the checks;
 2. compare different semantics for gradual type boundaries in the form of
@@
     - non-deterministic programs.
+
+Are the debugging scenarios difficult?
+================================================================================
+
+Review C says:
+> The only thing that I am concerned about at this point is that, later in the
+> paper, it is shown that "most cases the programmer need to type a single
+> module to debug a scenario," which makes me wonder whether the authors did not
+> generate difficult enough scenarios. If that is the case, what is the problem?
+
+The bugs that we inject are non-trivial. To provide some evidence, we sampled a
+few (~24,000) scenarios from three middle-of-the-road benchmarks (acquire,
+take5, kcfa) and the median stack has 7 program modules, with the sizes ranging
+from 3 to 10. This indicates that these bugs involve significant interaction
+among different parts of the program.
+
+One could argue that the benchmark programs are too small, because they consist
+of about 10-14 modules. This is a decent size but not a truly large one, so
+perhaps even larger benchmark programs are called for, though we conjecture that
+results will remain roughly the same because the programming style is
+representative (it includes a good number of higher-order idioms, for both
+functions and classes; it includes a fair number of call-back scenarios across
+module boundaries and storage of objects to be called later -- after the storage
+call returns -- but again not too many).
 
@@
 
-TODO
+Our implementation of Transient is faithful to Reticulated Python; the parallel
+experience report submission by the same set of authors describes the
+implementation of Transient in depth.
 
@@
 
-TODO
-
-> The only thing that I am concerned about at this point is that, later in the
-> paper, it is shown that "most cases the programmer need to type a single
-> module to debug a scenario," which makes me wonder whether the authors did not
-> generate difficult enough scenarios. If that is the case, what is the problem?
-
-TODO
+Thanks, we will consider this alternative name for the final paper.
 
@@
 
-TODO
+This is a good idea, thanks. In the final paper we will try to incorporate one
+or more examples that illustrate the difference in trail between the three
+mechanisms.

### 2021-04-24 writing/reviews/icfp21-rebuttal.md
Reflows and 2 typo fixes in the new replies

@@
 
-> p5.
-one thing I wonder about is how this all hinges on the rational programmer not only taking the right action, but also taking it right?
-Is that exactly what the rational programmer idea presupposes?
-(I guess this reflects on your remark in Sec.
-11 that one problem is with errors in the ascribed types themselves.).
+> p5. one thing I wonder about is how this all hinges on the rational programmer
+> not only taking the right action, but also taking it right? Is that exactly
+> what the rational programmer idea presupposes? (I guess this reflects on your
+> remark in Sec. 11 that one problem is with errors in the ascribed types
+> themselves.).
 
@@
 
-> When the rational programmer "fixes" a module and reruns, the implicit assumption seems to be that he runs the program in the exact same way.
-That would make the most sense to me, but you do not seem to come out and say so.
+> When the rational programmer "fixes" a module and reruns, the implicit
+> assumption seems to be that he runs the program in the exact same way. That
+> would make the most sense to me, but you do not seem to come out and say so.
 
@@
 
-> It would help if you could explain how a mutator like negate-cond actually leads to a program
-> that has more type-level mistakes than before. It does not seem to me this part of the code
-> itself will now fail, but that changing the conditions makes certain previously excluded paths in the code feasible.
+> It would help if you could explain how a mutator like negate-cond actually
+> leads to a program that has more type-level mistakes than before. It does not
+> seem to me this part of the code itself will now fail, but that changing the
+> conditions makes certain previously excluded paths in the code feasible.
 
-Thanks for pointing this out, we will expand the explanation and examples
-of mutators in the final paper.  In the specific cases of `negate-cond`
-and `force-cond`, these mutators lead to type-level mistakes when the
-program uses occurrence typing.  For example, they could both lead to a
-mistake in a program like this:
+Thanks for pointing this out, we will expand the explanation and examples of
+mutators in the final paper. In the specific cases of `negate-cond` and
+`force-cond`, these mutators lead to type-level mistakes when the program uses
+occurrence typing. For example, they could both lead to a mistake in a program
+like this:
 
@@
 
-You are absolutely right, we will clarify these assumptions about the
-benchmarks in the paper. They are inherited from the other
-publications on the gradual benchmark suite.
+You are absolutely right, we will clarify these assumptions about the benchmarks
+in the paper. They are inherited from the other publications on the gradual
+benchmark suite.
 
@@
 
-Thanks, we will consider this alternative name for the final paper.
+Thanks, we will consider these alternative names for the final paper.
 
@@
 This is a good idea, thanks. In the final paper we will try to incorporate one
-or more examples that illustrate the difference in trail between the three
+or more examples that illustrate the difference in trails between the three
 mechanisms.

### 2021-04-30 writing/rational.tex
Fix citation style
Typo fixes

@@
 message. A sequence of such steps make up a portion of the migratory path in the
-lattice of type migration~\cite{tfgnvf-popl-2016}. The lattice serves as the
+lattice of type migration~\citep{tfgnvf-popl-2016}. The lattice serves as the
 common substrate for the definition modes of the rational programmer (see
@@
 
-Hence the research problem is develop modes for the rational programmer and to make
+Hence the research problem is to develop modes for the rational programmer and to make
 them comparable even when they correspond to different semantics and offer

### 2021-04-30 writing/why-rational-methodology.tex
uniform editorial voice, mistakes, typos, size of text in figures, captions
Typo fixes

@@
 language researchers. It shows how their theorems and slogans help
-programmers>. The example of blame assignment mechanisms makes this point
+programmers. The example of blame assignment mechanisms makes this point
 clearly. A blame-assignment mechanism provides information that, according to
@@
 \end{enumerate}
-That said, using the method to conduct data-gathering experiments entails
+That said, using the method to conduct data-gathering experiments poses
 several challenges. The specific challenges are spelled out in the next section,

### 2021-04-30 writing/introduction.tex
Say what practice means at the end of intro

@@
 one is that the cost of Transient's blame can be huge.  In turn, these
-problems suggest that, on one hand, the existing theory does not predict
-practice properly, and on the other hand, the existing practice may need
+problems suggest that, on one hand, the existing theory does not properly
+predict practice (i.e. the behavior of blame in real systems with real programs),
+and on the other hand, the existing practice may need
 additional experiments to collect more data.

### 2021-04-30 writing/discussion.tex
In Transient threats, clarify when we talk about Python benchmarks

@@
 of~\citet{vss-popl-2017}. They report an average slowdown of 2.5x and a
-worst-case of 5.4x on fully-typed benchmarks in Reticulated Python.
+worst-case of 5.4x on the fully-typed Python benchmarks in Reticulated Python.
 Unfortunately, the high cost of Shallow Racket seems closer to the truth.
@@
 \item On non-trivial programs, Reticulated Python suffers from the same high
-overhead as Shallow Racket.  It runs the simplest benchmark from the suite 
+overhead as Shallow Racket.  It runs the simplest benchmark from the GTP suite (translated into Python)
 in approximately 40 seconds without blame and times out after 10 minutes when

### 2021-04-30 writing/mutate-benchmarks.tex
Try to note assumptions about benchmarks typability and determinism

@@
 
-\item vary in size, complexity and purpose;
+\item vary in size, complexity and purpose;\footnote{A finer point worth noting is that since the rational programmer assumes observable changes in program behavior are due to its actions, the programs should be deterministic.}
 
-\item have a decent number of modules and a variety of module dependency graphs
+\item have a decent number of type-able modules and a variety of module dependency graphs
  (because mixing of typed and untyped code in Typed Racket takes place at the

### 2021-04-30 writing/introduction.tex
Reword conclusions in intro to clarify practice/theory contrast

@@
 principle they validate the conjectures behind the work of theoreticians.
-First, a good blame assignment strategy helps with the search for
+First, a blame assignment strategy helps with the search for
 impedance mismatches between the specified types and the behavior of
@@
 which in turn is superior to Erasure. {\em But\/}, the application of the
-method also indicates problems with the expectations of theoreticians. The
-first problem concerns the difference between the methods; neither is {\it
-Natural\/} vastly more useful than {\it Transient\/} nor are the two
-blame assignment methods highly superior to {\it Erasure\/}.  The second
-one is that the cost of Transient's blame can be huge.  In turn, these
+method also indicates problems with the expectations of theoreticians. First,
+in contrast with the large theoretrical differences between the methods,
+{\it Natural\/} is only marginally more useful than {\it Transient\/}, and neither of the two
+blame assignment methods are highly superior to {\it Erasure\/}.  Second,
+the cost of Transient's blame can be huge.  In turn, these
 problems suggest that, on one hand, the existing theory does not properly

### 2021-04-30 writing/rational-stuff.tex
uniform editorial voice, mistakes, typos, size of text in figures, captions
Use configuration consistently in lattice definitions

@@
 on the generated lattice. The two lattices differ only in the mutated
-$\component^*$. Given this formulation, debugging scenarios are those points in
-this new lattice that do not contain the mutated module; at all other points, the type
+$\component^*$. Given this formulation, debugging scenarios are those configurations in
+this new lattice that do not contain the mutated module; at all other configurations, the type
 checker points out the type-level mistake in the mutated $\component^*$.
@@
 The actions of the rational programmer create an ascending chain---dubbed a
-\emph{trail}---in $\lattice{\system}$.  The starting scenario $\conf_0$ of a
-trail---also referred to as the \emph{root} (of the trail)---is the initial
-debugging scenario.  Starting from such a root, the rational
+\emph{trail}---in $\lattice{\system}$.  The {\em root\/} $\conf_0$ of a trail is
+the initial debugging scenario.  Starting from such a root, the rational
 programmer steps along a trail. If the (program for this) scenario type-checks,
-the rational programmer runs the program until it raises a run-time error.  The rational programmer
-uses the information in the error to decide which module to equip with
-types. This choice constructs scenario $\conf_{i+1}$ from $\conf_{i}$ and thus
-creates a trail. The trail's construction ends when it reaches a scenario that
-contains the mutated module because the type checker rejects its typed version
-outright.  At this point, the source of the impedance mismatch is identified.
+the rational programmer runs the program until it raises a run-time error.  The
+rational programmer uses the information in the error to decide which module to
+equip with types. This choice constructs scenario $\conf_{i+1}$ from $\conf_{i}$
+and thus creates a trail. The trail's construction ends when it reaches a
+scenario that contains the mutated module because the type checker rejects its
+typed version outright.  At this point, the source of the impedance mismatch is
+identified.

### 2021-04-30 writing/landscape.tex
Typo fixes

@@
 
-\item[\texttt{typed-pack-lib}] (at the mid-level on the left), imports {\tt pack}
+\item[\texttt{typed-pack-lib}] (at the mid-level on the left) imports {\tt pack}
 and re-exports it as \texttt{typed-pack} asserting that it is a function that

### 2021-04-30 writing/challenges.tex
Use "semantic variants" instead of "soundness mechanisms"
To be consistent

@@
 authors use Racket, which is thus far the only language in which all three major
-soundness mechanisms are available in a robust and comparable manner:
+semantic variants are available in a robust and comparable manner:
 Typed Racket implements Natural, Shallow Racket~\citep{ttt21}

### 2021-04-30 writing/challenges.tex
Add parenthetical specifying what we mean by effort to challenges

@@
 locating an impedance mismatch, measuring the rational programmer's effort with each mode
-may answer which is the most effective.
+(how many components it must type) may answer which is the most effective.
 

### 2021-04-30 writing/mutate-benchmarks-figure.tex
uniform editorial voice, mistakes, typos, size of text in figures, captions
Clean up the two tables (figs 4 and 5)

@@
-\begin{figure*}
-\begin{tabular}{p{1.4cm} | p{6.9cm} | p{2.0cm} | r | r}
+\begin{figure*} \footnotesize
+\begin{tabular}{p{1.5cm} | p{6.9cm} | p{1.9cm} | r | r}
              {\bf  name} & {\bf description} & {\bf author} & {\bf loc} & {\bf mod.} \\ \hline
@@
 
-\texttt{take5} & Mixin-based card game simulator & M.Felleisen & 465 & 8\\ \hline
+\texttt{take5} & mixin-based card game simulator & M.Felleisen & 465 & 8\\ \hline
 
-\texttt{tetris} & Functional implementation of Tetris & D. Van Horn &   280 & 9 \\ \hline
+\texttt{tetris} & functional implementation of Tetris & D. Van Horn &   280 & 9 \\ \hline
 
-\texttt{suffix\-tree} & algorithm for common longest subsequences between strings. & D. Yoo & 1500 & 6 \\
+\texttt{suffix\-tree} & algorithm for common longest subsequences between strings & D. Yoo & 1500 & 6 \\
 
@@
 
-\caption{Benchmarks summary.} \label{table:benchmark-descriptions}
+\caption{Summary of benchmarks} \label{table:benchmark-descriptions}
 

### 2021-04-30 writing/rational-natural.tex
Typo fixes
Extend definition of blame modes to use stack in absence of blame
And add a paragraph explaining why.

@@
 programmer extends the trail to a scenario that swaps out the untyped
-module
-for its typed counterpart.
+module for its typed counterpart.
 
@@
 a program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
-\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =\{\blame{\system, \conf_i}\}$, where
-$\blame{\system, \conf}$ denotes the module that $\conf$ blames.
+\conf_{i+1}$ and
+\[ \conf_{i+1} \setminus \conf_i = %
+\left\{ %
+\begin{tabular}{ll} %
+$\{\blame{\system, \conf_i}\}$      & if $\conf_i$ produces blame \\ %
+$\{\exception{\system, \conf_i}\}$  & otherwise \\ %
+\end{tabular} %
+\]
+where $\blame{\system, \conf}$ denotes the module (of $\system$) that $\conf$ blames,
+and $\exception{\system, \conf}$ denotes the first untyped location in the stacktrace produced by $\conf$.
 \end{quote}
+
+This definition rests on the two options for $\conf_{i+1} \setminus \conf_i$.
+The first part reflects the rational programmer's use of blame to extend the
+trail. However, blame may not be available in some scenarios. The second part
+accounts for those scenarios that produce errors from the runtime system before
+any impedance mismatch can be detected. For example, the fully untyped
+configuration can only produce such an error (e.g. from {\tt +} receiving a
+string). Exceptions from the runtime do not carry blame, so
+the rational programmer proceeds using the accompanying stacktrace instead.
 
@@
 To account for this situation, a new mode of the Natural rational
-programmer follows a migration process based on exceptions.
+programmer follows a migration process based on entirely on exceptions.
 \begin{quote}
@@
 program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
-\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i = \{\exception{\system, \conf_i}\}$
-where $\exception{\system, \conf}$ denotes the location of the exception of $\conf$.
+\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i = \{\exception{\system, \conf_i}\}$.
 \end{quote}

### 2021-04-30 writing/rational-transient.tex
Extend definition of blame modes to use stack in absence of blame
And add a paragraph explaining why.

@@
 $\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
-$\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
-\first{\mblame{\system, \conf_i}}$ where $\first{\mblame{\system,
-\conf}}$/$\last{\mblame{\system, \conf}}$ is the
+$\conf_i \subset \conf_{i+1}$ and
+\[ \conf_{i+1} \setminus \conf_i = %
+\left\{ %
+\begin{tabular}{ll} %
+$\{\first{\mblame{\system, \conf_i}}\}$  & if $\conf_i$ produces blame \\ %
+$\{\exception{\system, \conf_i}\}$     & otherwise \\ %
+\end{tabular} %
+\]
+where $\first{\mblame{\system, \conf}}$/$\last{\mblame{\system, \conf}}$ is the
 first/last module, respectively,  that Transient adds to the blame sequence for $\conf$.
 \end{quote}
-
-% \begin{quote}
-% \it A  is a sequence of scenarios
-% $\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
-% $\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
-%   \last{\mblame{\system, \conf_i}}$ where $\last{\mblame{\system, \conf}}$ is the
-% last module Transient adds to the blame sequence for $\conf$.
-% \end{quote}
 

### 2021-05-01 writing/reviews/icfp21-rebuttal.md
Add draft replies to the review D's points in icfp21-rebuttal

@@
 mechanisms.
+
+Review #45D [R1]
+===========================================================================
+
+> First, the performance figures cited from [38] (worst case 5.4x
+> slowdown) are used incorrectly: those numbers are about transient
+> /without/ blame (see Section 6.2 of [38]), and given that "with the
+> blame map turned off, the Transient semantics also runs these programs
+> in a short amount of time and well within the memory limits" there
+> appears to be no contradiction here. When blame /is/ enabled, [38]
+> reports a worst-case 18x slowdown, which may be more in line with the
+> results reported in this work, and as a result it's not actually clear
+> that the earlier results are skewed in a way that demands explanation.
+
+Yes you are right, this is our mistake: we meant to use the number with blame
+here. After looking at the plot in Section 6.2 of [38] we believe the right
+numbers are 6.2x and 17.2x (the 18x is an upper bound).
+
+These numbers are still orders-of-magnitude better than Shallow. We believe this
+gap needs an explanation.
+
+> Claim (2), that local variables in Reticulated Python are dynamically
+> typed because no types are ascribed to them, is false; Reticulated
+> Python uses a local type inference system to give types to local
+> variables (see Section 3.1.3 of [36]).
+
+Correct, Reticulated uses local type inference. We need to acknowledge this
+fact. Our overall claim stands, however, because Reticulated's inference often
+picks the dynamic type for local variables.
+
+This behavior came as a surprise to us too, and we had to manually inspect the
+output of `retic --mgd-transient --print file.py` for several benchmarks to
+convince ourselves.
+
+Section 5.4.4 of Greenman's dissertation presents a tiny example where
+Reticulated infers `Dyn` for a local variable.
+
+> Claim (4) that "on large programs, Reticulated suffers from high
+> overhead" is unsubstantiated here or in cited prior work, and is also
+> unclear -- overhead compared to what? The text compares it (vaguely) to
+> Typed Racket (without specifying which semantics), but this isn't a
+> useful comparison given the different baseline performance of Racket and
+> CPython. It's not clear if the point is that performance of this
+> benchmark in blame-free Reticulated is seriously slow compared to
+> regular Python, or if the program itself is just slow in Python (in
+> which case, what's the relevance to this paper?).
+
+You are right, the prose here does not explain the comparison here well.
+Especially, Typed Racket has no place in this bullet.
+
+First off, our goal was to run Reticulated on one large benchmark to see if it
+ever had a huge "timeout-level" slowdown.
+
+To this end, we converted our fully-typed `sieve` benchmark from Racket to
+Reticulated. Then we ran this fully-typed Reticulated program twice: without
+blame and with blame. Adding blame gave a huge slowdown.
+
+This little experiment suggests that Reticulated does not have any crucial
+implementation techniques that Shallow Racket is missing.
+
+> The characterization of the Monotonic semantics for references [22] as a
+> variant of natural isn't correct--its exception-raising and blame
+> behavior for references (though not for functions) is quite different
+> from that of Natural. For example, a monotonic reference imported from
+> untyped code into two incompatible typed contexts will error
+> immediately, before being used.
+
+Indeed this characterization is not right, monotonic is more than a mere variant
+of Natural. We will fix this description in the final paper.

### 2021-05-02 writing/reviews/icfp21-rebuttal.md
editing reply to last review

@@
 
-These numbers are still orders-of-magnitude better than Shallow. We believe this
-gap needs an explanation.
+These numbers, however, are still orders-of-magnitude better than Shallow. Hence, 
+we believe this gap needs an explanation.
 
@@
 
-Correct, Reticulated uses local type inference. We need to acknowledge this
-fact. Our overall claim stands, however, because Reticulated's inference often
+You are right and we should had been more precise here. Reticulated does attempt to 
+infer types for local variables. However, our experience indicates Reticulated's inference often
 picks the dynamic type for local variables.
@@
 
-Section 5.4.4 of Greenman's dissertation presents a tiny example where
-Reticulated infers `Dyn` for a local variable.
+Section 5.4.4 of Greenman's dissertation presents a tiny example that
+explains how Reticulated can infer `Dyn` for a local variable.
+
+chrdimo: is it worth it expanding on this here?
 
@@
 
-You are right, the prose here does not explain the comparison here well.
-Especially, Typed Racket has no place in this bullet.
+Thank for pointing this out. The prose here does not explain clearly our
+evidence for this claim.
 
-First off, our goal was to run Reticulated on one large benchmark to see if it
-ever had a huge "timeout-level" slowdown.
+First off, the overall goal was to run Reticulated (with and without blame) on one 
+large benchmark to see if it ever had a huge "timeout-level" slowdown,
+similar to the ones we observe in our ``rational programmer'' experiment.
 
-To this end, we converted our fully-typed `sieve` benchmark from Racket to
+To this end, we converted our fully-typed `sieve` benchmark from TR to
 Reticulated. Then we ran this fully-typed Reticulated program twice: without
@@
 
-This little experiment suggests that Reticulated does not have any crucial
-implementation techniques that Shallow Racket is missing.
+This little experiment suggests that Reticulated does not employ any crucial
+implementation techniques that can mitigate the slowdown we observe with Shallow Racket.
+
+chrdimo: Is the translation faithful in terms of types?
+
+chrdimo: I think there is a valid point here that the claim comes out as
+too strong but what we have a single data point. I think we should also
+admit this and say that we will tone down the claim.
 
@@
 
-Indeed this characterization is not right, monotonic is more than a mere variant
-of Natural. We will fix this description in the final paper.
+Indeed this characterization is not right, when it comes to boxes
+monotonic is more than a mere variant of Natural. We will fix this
+description in the final paper.
+
+chrdimo: Is monotonic implemented? What happens with hashes, objects and
+other containers?

### 2021-05-03 writing/reviews/icfp21-rebuttal.md
typos, grammar, claims

@@
 
-You are right and we should had been more precise here. Reticulated does attempt to 
-infer types for local variables. However, our experience indicates Reticulated's inference often
-picks the dynamic type for local variables.
+You are right and we should have been more precise here. Reticulated
+does attempt to infer types for local variables. However, our
+experience indicates Reticulated's inference often picks type Dynamic
+for local variables. Section 5.4.4 of Greenman's dissertation presents
+a tiny example that explains how Reticulated can infer `Dyn` for a
+local variable.
 
-This behavior came as a surprise to us too, and we had to manually inspect the
+This behavior came as a surprise to us too. We manually inspected the
 output of `retic --mgd-transient --print file.py` for several benchmarks to
 convince ourselves.
-
-Section 5.4.4 of Greenman's dissertation presents a tiny example that
-explains how Reticulated can infer `Dyn` for a local variable.
-
-chrdimo: is it worth it expanding on this here?
 
@@
 
-First off, the overall goal was to run Reticulated (with and without blame) on one 
-large benchmark to see if it ever had a huge "timeout-level" slowdown,
-similar to the ones we observe in our ``rational programmer'' experiment.
+The goal was to run Reticulated (with and without blame) on one
+reasonably large benchmark to see if it ever had a huge
+"timeout-level" slowdown, similar to the ones we observe in our
+``rational programmer'' experiment.
 
 To this end, we converted our fully-typed `sieve` benchmark from TR to
-Reticulated. Then we ran this fully-typed Reticulated program twice: without
-blame and with blame. Adding blame gave a huge slowdown.
+Reticulated. Then we ran this fully-typed Reticulated program twice:
+with and without blame. The `with blame` variant experienced a huge
+slowdown.
 
-This little experiment suggests that Reticulated does not employ any crucial
-implementation techniques that can mitigate the slowdown we observe with Shallow Racket.
-
-chrdimo: Is the translation faithful in terms of types?
-
-chrdimo: I think there is a valid point here that the claim comes out as
-too strong but what we have a single data point. I think we should also
-admit this and say that we will tone down the claim.
+This little experiment suggests that Reticulated does not employ any
+crucial implementation techniques that can mitigate the slowdown we
+observe with Shallow Racket. Of course, all we have is a single data
+point, and this must come thru more clearly in our prose. We would
+hope that this submission and the parallel submission on Shallow
+Racket's construction inspires the Reticulaed team to report data on
+larger benchmarks than the ones in their published papers.
 
@@
 monotonic is more than a mere variant of Natural. We will fix this
-description in the final paper.
-
-chrdimo: Is monotonic implemented? What happens with hashes, objects and
-other containers?
+description.

### 2021-05-03 writing/reviews/icfp21-rebuttal.md
fixing a typo
editing reply to last review

@@
 
-Thank for pointing this out. The prose here does not explain clearly our
+Thank you for pointing this out. The prose here does not explain clearly our
 evidence for this claim.
@@
 hope that this submission and the parallel submission on Shallow
-Racket's construction inspires the Reticulaed team to report data on
+Racket's construction inspires the Reticulated team to report data on
 larger benchmarks than the ones in their published papers.

### 2021-05-03 writing/reviews/icfp21-rebuttal.md
Switch to Christos's suggestion about making Transient viable

@@
 observe with Shallow Racket. Of course, all we have is a single data
-point, and this must come thru more clearly in our prose. We would
-hope that this submission and the parallel submission on Shallow
-Racket's construction inspires the Reticulated team to report data on
-larger benchmarks than the ones in their published papers.
+point, and this must come through more clearly in our prose.
+
+We hope that this submission and the parallel submission on Shallow
+Racket's construction will inspire further investigation of Transient as a
+viable production-level approach.
 

### 2021-05-04 writing/landscape.tex
Remove the clever table layout in lieu of a sentence description

@@
 
-The following table summarizes the illustration:
+The following table summarizes the illustration. Each cell describes the result
+of evaluating the column's definition (in {\tt client}) under the row's
+semantics.
 %% -----------------------------------------------------------------------------
 \begin{center}
-  \begin{tabular}{l|ll@{}}
-                        &        \multicolumn{2}{|c}{the evaluation of} \\
-\relax                  & {\tt public-data}               & {\tt secret-data} in {\tt client}, \\
-{under the $\cdot$ semantics yields} & & \\ \hline
-%% ----------------------------------------------------------------------------------------
-\qquad {\it Natural\/}   & error, blaming {\tt pack-lib}  & error, blaming {\tt pack-lib} \\
+  \begin{tabular}{l|ll}
+                  & {\tt public-data}              & {\tt secret-data}                                 \\
+\hline %% ----------------------------------------------------------------------------------------
+{\it Natural\/}   & error, blaming {\tt pack-lib}  & error, blaming {\tt pack-lib}                     \\
                   &                                &                                                   \\
-\qquad {\it Transient\/} & no error   		          & error, blaming the boundary crossings \\
-      		  	 &    			          & \quad {\tt pack-lib} to {\tt typed-pack-lib} \\
+{\it Transient\/} & no error                       & error, blaming the boundary crossings             \\
+                  &                                & \quad {\tt pack-lib} to {\tt typed-pack-lib}, and \\
                   &                                & \quad {\tt pack-lib} to {\tt crypto-pack-lib}     \\
-\qquad {\it Erasure\/}   & no error 		          & no error \\
-       	    		 &   \multicolumn{2}{|c}{but {\em Erasure} does signal an error on list access}
+                  &                                &                                                   \\
+{\it Erasure\/}   & no error*                      & no error*                                         \\
+                  &   \multicolumn{2}{|c}{*but {\em Erasure} does signal an error on list access}
 \end{tabular}

### 2021-05-04 writing/discussion.tex
Add parenthetical defining sound and complete blame

@@
 approach fares better than the Transient approach, but only by a small margin.
-Since Natural offers complete and sound path-based blame while Transient offers
+Since Natural offers complete and sound path-based blame (i.e. it blames all and only those components
+that are relevant to a type impedance mismatch) while Transient offers
 incomplete but sound heap-based blame~\citep{gfd-oopsla-2019}, the results call

### 2021-05-07 writing/introduction.tex
Touch up previous edits

@@
 method also indicates problems with the expectations of theoreticians. First,
-in contrast with the large theoretrical differences between the methods,
+in contrast with the theoretical differences between the methods,
 {\it Natural\/} is only marginally more useful than {\it Transient\/}, and neither of the two

### 2021-05-07 writing/mutate-benchmarks.tex
Touch up previous edits

@@
 
-\item vary in size, complexity and purpose;\footnote{A finer point worth noting is that since the rational programmer assumes observable changes in program behavior are due to its actions, the programs should be deterministic.}
+\item vary in size, complexity and purpose;
 
@@
 tower~\citep{stathff-padl-12}.
+Finally, all of the programs are deterministic, so any changes in the programs' behavior
+between runs can be solely attributed to the actions of the rational programmer.
 

### 2021-05-07 writing/discussion.tex
Touch up previous edits
Add reference for Mill's homo economicus

@@
 models have provided some predictive insights into the market's behavior; but as
-behavioral economics has shown more recently, the mathematical abstraction of a
+behavioral economics has shown more recently (e.g.~\cite{henrich2001search}), the mathematical abstraction of a
 rational actor makes predictions also quite unreliable in some situations.
@@
 of blame assignment may not materialize. Indeed, the authors' personal
-experience suggests such deviations, and it also suggests that deviating is a
-mistake. To make a true judgment of the usefulness of the rational-programmer
+experience suggests such deviations, and it also suggests that deviating often leads to dead-ends.
+To make a true judgment of the usefulness of the rational-programmer
 idea, the community will need much more experience with this form of evaluation

### 2021-05-07 writing/why-rational.tex
Note that RP re-runs program in the same way
Add reference for Mill's homo economicus

@@
 
-This paper dubs Lazarek et. al's idea {\em the rational programmer\/}. Like Stuart
-Mill's {\it homo economicus\/}, the rational programmer approximates the
+This paper dubs Lazarek et. al's idea {\em the rational programmer\/}.
+Like~\citet{mill1874essays}'s {\it homo economicus\/}, the rational programmer approximates the
 behavior of a software developer who reduces time spent on a task by exploiting the
@@
 main}'s import is now re-directed there by {\tt (submod ".." layer-typed)}. As
-predicted by the theory, running the modified program yields a different error message:
+predicted by the theory, running the modified program (in the same way as before) yields a different error message:
 \begin{verbatim}

### 2021-05-12 writing/discussion.tex
Add blame trail example to discussion, in new subsection

@@
 %% -----------------------------------------------------------------------------
+\subsection{Contextualizing the results through an example scenario}
+\begin{figure}
+\input{trail-example-table}
+\caption{An example scenario from take5, with every mode's resulting trail.}
+\label{fig:example-trails}
+\end{figure}
 
+Interpreting the numeric summaries and aggregations described in the last section demands an intuitive understanding of how blame trails look in practice.
+Thus, before drawing high level conclusions in the next subsection, this one breaks down one illustrative debugging scenario from the take5 benchmark.
+Figure~\ref{fig:example-trails} shows the blame trails produced by every mode of the rational programmer for the debugging scenario in two different ways.
+At top left, the trails are depicted diagrammatically as paths through the configuration lattice starting at the root (leftmost) configuration.
+Each configuration is represented by a bit string where each digit corresponds to a module in the program and indicates if it is typed.
+To contextualize this module information, the module graph of take5 is at top right.
+The table at the bottom of the figure expands upon the diagram with the details of every step in each trail.
+Specifically, each mode's trail is represented by a row in the table.
+The \textbf{Root} column describes the result of running the root configuration in the row's mode.
+Using that information, the rational programmer for each mode constructs the successor configuration in the next column (\textbf{Step 1}) if possible.
+Repeating the same process from that configuration produces the next column (\textbf{Step 2}).
+Finally, the \textbf{Success?} column summarizes whether the trail reaches a type error, and therefore succeeds.
+For example, the first row shows that running the root configuration under the Natural-blame mode results in blame on the \texttt{player} module;
+typing that module and running again then results in a type error, and hence the trail is successful.
+In contrast, the Natural-exceptions row shows that the stack information produced by the root configuration under the Natural-exceptions mode is unhelpful (it only identifies \texttt{main}, which is already typed), and hence the trail immediately gets stuck.
+In sum, this example provides a concrete demonstration of how the different modes of the rational programmer behave (and diverge from one another), giving context to the numeric results of the last section.
+
+\subsection{Interpreting the results}
 The results of the experiment suggests a number of high-level conclusions about

### 2021-05-12 writing/trail-example-table.tex
Add blame trail example to discussion, in new subsection

{
  \newcommand\runtimeError{\begin{minipage}{0.03\textwidth}\centering\includegraphics[width=\linewidth]{Images/runtime-error}\end{minipage}}
  \newcommand\checkFailure{\begin{minipage}{0.03\textwidth}\centering\includegraphics[width=\linewidth]{Images/check-failure}\end{minipage}}
  \newcommand\blameFinger{\begin{minipage}{0.02\textwidth}\centering\vspace{0.25em}\includegraphics[width=\linewidth]{Images/finger}\end{minipage}}
  \newcommand\typeError{\scalebox{1.5}{$\tau_{\hspace{-0.1em}\times}$}}
  \newcommand\success{\textcolor{black!30!green}{\checkmark}}
  \newcommand\fail{\textcolor{red}{\sffamily x}}
  \footnotesize
\centering
\begin{tabular}{l|rcl|rcl|rc|c}
\multicolumn{7}{c}{\begin{minipage}{0.65\textwidth}\centering\includegraphics[width=\linewidth]{Images/trails-example}\end{minipage}} & %
\multicolumn{3}{c}{\begin{minipage}{0.25\textwidth}\centering\includegraphics[width=\linewidth]{Images/take5-module-graph}\end{minipage}} \\
\multicolumn{7}{c}{\begin{minipage}{0.65\textwidth}\centering The paths taken by each mode through the configuration lattice.\end{minipage}} & %
\multicolumn{3}{c}{\begin{minipage}{0.25\textwidth}\centering The module graph of take5.\end{minipage}}
\vspace{1em} \\
 & \textbf{Root} &  &  & \textbf{Step 1} &  &  & \textbf{Step 2} &  & \textbf{Success?}\\
\textbf{Mode} & config & result & stack & config & result & stack & config & result & \\
\hline
Natural & \textcolor{red}{1}000\textcolor{red}{1}\textcolor{red}{1}00 & \blameFinger & \texttt{main} & \textcolor{red}{1}000\textcolor{red}{1}\textcolor{red}{1}\textcolor{red}{1}0 & \typeError &  &  &  & \success\\
-blame &  & \texttt{player} & \texttt{main} &  &  &  &  &  & \\
\hline
Transient & \textcolor{red}{1}000\textcolor{red}{1}\textcolor{red}{1}00 & \runtimeError & \texttt{dealer} & \textcolor{red}{1}00\textcolor{red}{1}\textcolor{red}{1}\textcolor{red}{1}00 & \blameFinger & \texttt{dealer} & \textcolor{red}{1}00\textcolor{red}{1}\textcolor{red}{1}\textcolor{red}{1}\textcolor{red}{1}0 & \typeError & \success\\
-first-blame &  &  & \texttt{dealer} &  & \texttt{player} &  &  &  & \\
\emph{and} &  &  & \texttt{dealer} &  & \texttt{dealer} &  &  &  & \\
-last-blame &  &  & \texttt{main} &  &  &  &  &  & \\
\hline
Erasure & \textcolor{red}{1}000\textcolor{red}{1}\textcolor{red}{1}00 & \runtimeError & \texttt{dealer} & \textcolor{red}{1}00\textcolor{red}{1}\textcolor{red}{1}\textcolor{red}{1}00 & \runtimeError & \texttt{dealer} &  &  & \fail\\
 &  &  & \texttt{dealer} &  &  & \texttt{dealer} &  &  & \\
 &  &  & \texttt{dealer} &  &  & \texttt{dealer} &  &  & \\
 &  &  & \texttt{main} &  &  & \texttt{main} &  &  & \\
\hline
Natural & \textcolor{red}{1}000\textcolor{red}{1}\textcolor{red}{1}00 & \checkFailure & \texttt{main} &  &  &  &  &  & \fail\\
-exceptions &  &  & \texttt{main} &  &  &  &  &  & \\
\hline
Transient & \textcolor{red}{1}000\textcolor{red}{1}\textcolor{red}{1}00 & \runtimeError & \texttt{dealer} & \textcolor{red}{1}00\textcolor{red}{1}\textcolor{red}{1}\textcolor{red}{1}00 & \checkFailure & \texttt{dealer} &  &  & \fail\\
-exceptions &  &  & \texttt{dealer} &  &  &  &  &  & \\
 &  &  & \texttt{dealer} &  &  &  &  &  & \\
 &  &  & \texttt{main} &  &  &  &  &  & \\
\multicolumn{10}{c}{\begin{minipage}{0.95\textwidth}\vspace{0.5em} \textit{Legend.} Configurations: represented by a bit string, each digit corresponds to a module and indicates if it is typed. Results: \blameFinger~~denotes that the configuration results in a run-time type check failure blaming the module(s) below it, \checkFailure~~a run-time type check failure for which blame is ignored, \runtimeError~~an error from the runtime, and \typeError~~a type-checker error.\end{minipage}}  \\
\end{tabular}
}

### 2021-05-13 writing/trail-example-table.tex
Incorporate Ben's suggestions for the trails example figure

@@
 {
+  \definecolor{trans-fb}{RGB}{215, 48, 31}
+  \definecolor{trans-lb}{RGB}{252, 141, 89}
+  \definecolor{trans-exn}{RGB}{253, 204, 0}
+  \definecolor{erasure}{RGB}{119, 87, 254}
+
+  \newcommand\config[1]{\begin{minipage}{0.09\textwidth}\centering\includegraphics[width=\linewidth]{Images/#1}\end{minipage}}
   \newcommand\runtimeError{\begin{minipage}{0.03\textwidth}\centering\includegraphics[width=\linewidth]{Images/runtime-error}\end{minipage}}
-  \newcommand\checkFailure{\begin{minipage}{0.03\textwidth}\centering\includegraphics[width=\linewidth]{Images/check-failure}\end{minipage}}
+  \newcommand\checkFailure{\begin{minipage}{0.03\textwidth}\centering\includegraphics[width=\linewidth]{Images/check-failure-plain}\end{minipage}}
   \newcommand\blameFinger{\begin{minipage}{0.02\textwidth}\centering\vspace{0.25em}\includegraphics[width=\linewidth]{Images/finger}\end{minipage}}
@@
 \centering
-\begin{tabular}{l|rcl|rcl|rc|c}
+\begin{tabular}{l|ccl|ccl|cc|c}
 \multicolumn{7}{c}{\begin{minipage}{0.65\textwidth}\centering\includegraphics[width=\linewidth]{Images/trails-example}\end{minipage}} & %
@@
 \hline
-Natural & \textcolor{red}{1}000\textcolor{red}{1}\textcolor{red}{1}00 & \blameFinger & \texttt{main} & \textcolor{red}{1}000\textcolor{red}{1}\textcolor{red}{1}\textcolor{red}{1}0 & \typeError &  &  &  & \success\\
+Natural & \config{10001100} & \blameFinger & \texttt{main} & \config{10001110} & \typeError &  &  &  & \success\\
 -blame &  & \texttt{player} & \texttt{main} &  &  &  &  &  & \\
 \hline
-Transient & \textcolor{red}{1}000\textcolor{red}{1}\textcolor{red}{1}00 & \runtimeError & \texttt{dealer} & \textcolor{red}{1}00\textcolor{red}{1}\textcolor{red}{1}\textcolor{red}{1}00 & \blameFinger & \texttt{dealer} & \textcolor{red}{1}00\textcolor{red}{1}\textcolor{red}{1}\textcolor{red}{1}\textcolor{red}{1}0 & \typeError & \success\\
--first-blame &  &  & \texttt{dealer} &  & \texttt{player} &  &  &  & \\
+\textcolor{trans-fb}{Transient} & \config{10001100} & \runtimeError & \texttt{dealer} & \config{10011100} & \blameFinger & \texttt{dealer} & \config{10011110} & \typeError & \success\\
+\textcolor{trans-fb}{-first-blame} &  &  & \texttt{dealer} &  & \texttt{player} &  &  &  & \\
 \emph{and} &  &  & \texttt{dealer} &  & \texttt{dealer} &  &  &  & \\
--last-blame &  &  & \texttt{main} &  &  &  &  &  & \\
+\textcolor{trans-lb}{-last-blame} &  &  & \texttt{main} &  &  &  &  &  & \\
 \hline
-Erasure & \textcolor{red}{1}000\textcolor{red}{1}\textcolor{red}{1}00 & \runtimeError & \texttt{dealer} & \textcolor{red}{1}00\textcolor{red}{1}\textcolor{red}{1}\textcolor{red}{1}00 & \runtimeError & \texttt{dealer} &  &  & \fail\\
+\textcolor{erasure}{Erasure} & \config{10001100} & \runtimeError & \texttt{dealer} & \config{10011100} & \runtimeError & \texttt{dealer} &  &  & \fail\\
  &  &  & \texttt{dealer} &  &  & \texttt{dealer} &  &  & \\
@@
 \hline
-Natural & \textcolor{red}{1}000\textcolor{red}{1}\textcolor{red}{1}00 & \checkFailure & \texttt{main} &  &  &  &  &  & \fail\\
--exceptions &  &  & \texttt{main} &  &  &  &  &  & \\
+\textcolor{gray}{Natural} & \config{10001100} & \checkFailure & \texttt{main} &  &  &  &  &  & \fail\\
+\textcolor{gray}{-exceptions} &  &  & \texttt{main} &  &  &  &  &  & \\
 \hline
-Transient & \textcolor{red}{1}000\textcolor{red}{1}\textcolor{red}{1}00 & \runtimeError & \texttt{dealer} & \textcolor{red}{1}00\textcolor{red}{1}\textcolor{red}{1}\textcolor{red}{1}00 & \checkFailure & \texttt{dealer} &  &  & \fail\\
--exceptions &  &  & \texttt{dealer} &  &  &  &  &  & \\
+\textcolor{trans-exn}{Transient} & \config{10001100} & \runtimeError & \texttt{dealer} & \config{10011100} & \checkFailure & \texttt{dealer} &  &  & \fail\\
+\textcolor{trans-exn}{-exceptions} &  &  & \texttt{dealer} &  &  &  &  &  & \\
  &  &  & \texttt{dealer} &  &  &  &  &  & \\
  &  &  & \texttt{main} &  &  &  &  &  & \\
-\multicolumn{10}{c}{\begin{minipage}{0.95\textwidth}\vspace{0.5em} \textit{Legend.} Configurations: represented by a bit string, each digit corresponds to a module and indicates if it is typed.
+\multicolumn{10}{c}{\begin{minipage}{0.95\textwidth}\vspace{0.5em} \textit{Legend.} Configurations: each box corresponds to a module and indicates (with an x) if it is typed.
+The mutated module is gray.
 Results: \blameFinger~~denotes that the configuration results in a run-time type check failure blaming the module(s) below it, \checkFailure~~a run-time type check failure for which blame is ignored, \runtimeError~~an error from the runtime, and \typeError~~a type-checker error.\end{minipage}}  \\

### 2021-05-14 writing/trail-example-table.tex
clean up the figure

@@
 \begin{tabular}{l|ccl|ccl|cc|c}
-\multicolumn{7}{c}{\begin{minipage}{0.65\textwidth}\centering\includegraphics[width=\linewidth]{Images/trails-example}\end{minipage}} & %
-\multicolumn{3}{c}{\begin{minipage}{0.25\textwidth}\centering\includegraphics[width=\linewidth]{Images/take5-module-graph}\end{minipage}} \\
-\multicolumn{7}{c}{\begin{minipage}{0.65\textwidth}\centering The paths taken by each mode through the configuration lattice.\end{minipage}} & %
-\multicolumn{3}{c}{\begin{minipage}{0.25\textwidth}\centering The module graph of take5.\end{minipage}}
+\multicolumn{3}{c}{\begin{minipage}{0.25\textwidth}\centering\includegraphics[width=\linewidth]{Images/take5-module-graph}\end{minipage}} &
+\multicolumn{7}{c}{\begin{minipage}{0.69\textwidth}\centering\includegraphics[width=\linewidth]{Images/trails-example}\end{minipage}} \\
+\multicolumn{3}{c}{\begin{minipage}{0.25\textwidth}\centering the dependency graph\end{minipage}} &
+\multicolumn{7}{c}{\begin{minipage}{0.69\textwidth}\centering the paths taken by each mode through the configuration lattice\end{minipage}} 
 \vspace{1em} \\
+\end{tabular}
+
+\begin{tabular}{l|ccl|ccl|cc|c}
  & \textbf{Root} &  &  & \textbf{Step 1} &  &  & \textbf{Step 2} &  & \textbf{Success?}\\
@@
  &  &  & \texttt{main} &  &  &  &  &  & \\
-\multicolumn{10}{c}{\begin{minipage}{0.95\textwidth}\vspace{0.5em} \textit{Legend.} Configurations: each box corresponds to a module and indicates (with an x) if it is typed.
+\end{tabular}
+
+\begin{minipage}{0.95\textwidth}
+\vspace{0.5em}
+\centerline{\it Legend}
+
+\noindent{\bf config} Each box corresponds to a module and indicates (with x) if it is typed. 
 The mutated module is gray.
-Results: \blameFinger~~denotes that the configuration results in a run-time type check failure blaming the module(s) below it, \checkFailure~~a run-time type check failure for which blame is ignored, \runtimeError~~an error from the runtime, and \typeError~~a type-checker error.\end{minipage}}  \\
+
+\medskip
+
+\noindent{\bf result}\\
+\begin{center}
+\begin{tabular}{l@{\quad}l}
+symbol        & denotation \\ \hline 
+\blameFinger  & the configuration results in a dynamic type-check failure, blaming the module(s) below \\
+\typeError    & the configuration does not type check\\
+\runtimeError & the configuration fails a check to the runtime system\\
+\checkFailure & the configuration signals a dynamic type check failure for which blame is ignored\\
 \end{tabular}
-}
+\end{center}
+
+\end{minipage}}

### 2021-05-14 writing/discussion.tex
fix the prose explanation of the figure

--- 2021-05-07 writing/discussion.tex
@@
 %% -----------------------------------------------------------------------------
+% \subsection{Contextualizing the results through an example scenario}
+
+\begin{figure}
+\input{trail-example-table}
+\caption{An example scenario from take5, with every mode's resulting trail.}
+\label{fig:example-trails}
+\end{figure}
+
+Interpreting the numeric summaries and aggregations of the last section demands an intuitive understanding of what blame trails look like in practice.
+Thus, before drawing high level conclusions, let us examine one set of scenarios with a typical debugging scenario from the {\tt take5} benchmark,
+whose module dependency graph is shown in the top left of figure~\ref{fig:example-trails}.
+
+The rest of figure~\ref{fig:example-trails} illustrates the blame trails produced by every mode of the rational programmer for the debugging scenario in
+two different ways: 
+\begin{itemize}
+
+\item The top right shows the blame trails produced by every mode of the rational programmer as paths through the configuration lattice starting at the
+root (leftmost) configuration. Each configuration is represented by a bit string where each digit corresponds to a module in the program and indicates if
+it is typed.
+
+\item The table in the middle expands the information in the diagram with the details of every step in each trail.
+Every row of the table represents the trail of one mode. The middle-three columns depict the steps of a blame trail:
+\begin{description}
+\item[Root] describes the result of running the root configuration in this row's mode.
+\item[Step 1] is the result of the rational programmer's reaction to the outcome of running the root. 
+\item[Step 2] shows the result of reacting to the outcome of running step~1 configuration, if any. 
+\end{description}
+Finally, the \textbf{Success?} column summarizes whether exploring the trail succeeds.
+
+\end{itemize}
+
+Let us contrast rows 1 and~4. The first one shows that running the root
+configuration under the Natural-blame mode fails due a dynamic type check and
+blames the {\tt player} module; typing that module and running again then
+results in a type error, and hence the trail is successful.  By contrast, the
+Natural-exceptions mode (row 4) yields stack information for the root
+configuration that is unhelpful; it identifies only {\tt main}, which is already
+typed. Hence, this trail immediately gets stuck.
+
+In short, this figure concretely demonstrates how the rational programmer
+behaves in different modes. In this case, the behaviors differ from each other
+in all five modes. The reader may keep these scenarios in mind as we next
+discuss the numeric results of the last section.
+
+\subsection{Interpreting the results}
 

### 2021-05-14 writing/discussion.tex
trail examples: touch up figure and new prose

@@
 Interpreting the numeric summaries and aggregations of the last section demands an intuitive understanding of what blame trails look like in practice.
-Thus, before drawing high level conclusions, let us examine one set of scenarios with a typical debugging scenario from the {\tt take5} benchmark,
+Thus, before drawing high level conclusions, let us examine one set of trails from a typical debugging scenario from the {\tt take5} benchmark,
 whose module dependency graph is shown in the top left of figure~\ref{fig:example-trails}.
 
-The rest of figure~\ref{fig:example-trails} illustrates the blame trails produced by every mode of the rational programmer for the debugging scenario in
+The rest of figure~\ref{fig:example-trails} illustrates the blame trails produced by every mode of the rational programmer (except Random) for the debugging scenario in
 two different ways: 
@@
 \item The top right shows the blame trails produced by every mode of the rational programmer as paths through the configuration lattice starting at the
-root (leftmost) configuration. Each configuration is represented by a bit string where each digit corresponds to a module in the program and indicates if
-it is typed.
+root (leftmost) configuration. Each configuration is represented by a sequence of boxes corresponding to modules in the program, with an x indicating that the module is typed.
+The mutated module has a gray box.
 
@@
 
-Let us contrast rows 1 and~4. The first one shows that running the root
+In illustration of how to read the table, let us contrast rows 1 and~4. The first one shows that running the root
 configuration under the Natural-blame mode fails due a dynamic type check and
@@
 behaves in different modes. In this case, the behaviors differ from each other
-in all five modes. The reader may keep these scenarios in mind as we next
+in five of the six modes (the two Transient blame modes behave the same).
+The reader may keep these scenarios in mind as we next
 discuss the numeric results of the last section.

### 2021-05-14 writing/trail-example-table.tex
trail examples: touch up figure and new prose

@@
 \begin{center}
+\vspace{-2.5em}
 \begin{tabular}{l@{\quad}l}
 symbol        & denotation \\ \hline 
-\blameFinger  & the configuration results in a dynamic type-check failure, blaming the module(s) below \\
+\blameFinger  & the configuration signals a dynamic type check failure, blaming the module(s) below \\
 \typeError    & the configuration does not type check\\
-\runtimeError & the configuration fails a check to the runtime system\\
+\runtimeError & the configuration fails a check by the runtime system\\
 \checkFailure & the configuration signals a dynamic type check failure for which blame is ignored\\

### 2021-05-14 writing/trail-example-table.tex
trails example: A few more small touches

@@
 \multicolumn{3}{c}{\begin{minipage}{0.25\textwidth}\centering the dependency graph\end{minipage}} &
-\multicolumn{7}{c}{\begin{minipage}{0.69\textwidth}\centering the paths taken by each mode through the configuration lattice\end{minipage}} 
-\vspace{1em} \\
+\multicolumn{7}{c}{\begin{minipage}{0.69\textwidth}\centering the paths taken by each mode through the configuration lattice\end{minipage}} \\
 \end{tabular}
+\vspace{1em}
 
@@
 
-\noindent{\bf config} Each box corresponds to a module and indicates (with x) if it is typed. 
+\noindent{\bf config} \hspace{0.5em} Each box corresponds to a module and indicates (with x) if it is typed. 
 The mutated module is gray.
@@
 \vspace{-2.5em}
-\begin{tabular}{l@{\quad}l}
+\begin{tabular}{c@{\quad}l}
 symbol        & denotation \\ \hline 

### 2021-05-14 writing/discussion.tex
fix language, add explanation of mistake in take5

@@
 
-Interpreting the numeric summaries and aggregations of the last section demands an intuitive understanding of what blame trails look like in practice.
-Thus, before drawing high level conclusions, let us examine one set of trails from a typical debugging scenario from the {\tt take5} benchmark,
-whose module dependency graph is shown in the top left of figure~\ref{fig:example-trails}.
+Interpreting the numeric summaries and aggregations of the last section demands
+an intuitive understanding of what blame trails look like in practice.  A
+concrete example of blame trails and programmer modes is a good basis for
+synthesizing this kind of intuition.
 
-The rest of figure~\ref{fig:example-trails} illustrates the blame trails produced by every mode of the rational programmer (except Random) for the debugging scenario in
-two different ways: 
+Figure~\ref{fig:example-trails} summarizes one particularly interesting
+debugging scenario from the {\tt take5} benchmark. The module dependency graph
+of this benchmark is shown in the top left of the figure. Its mutated {\tt
+player} module provides a method under a different name than the client module,
+{\tt dealer}, expects. In Typed Racket's gradual type system, this mistake
+corresponds to a type-impedance mismatch---and all rational-programmer modes
+come to different conclusions.
+
+The rest of figure~\ref{fig:example-trails} illustrates the blame trails for
+every mode of the rational programmer (except Random) for the debugging scenario
+in two different ways:
 \begin{itemize}
@@
 
-In illustration of how to read the table, let us contrast rows 1 and~4. The first one shows that running the root
-configuration under the Natural-blame mode fails due a dynamic type check and
-blames the {\tt player} module; typing that module and running again then
-results in a type error, and hence the trail is successful.  By contrast, the
-Natural-exceptions mode (row 4) yields stack information for the root
-configuration that is unhelpful; it identifies only {\tt main}, which is already
-typed. Hence, this trail immediately gets stuck.
+To make this table concrete, compare rows 1 and~4. The first one shows that
+running the root configuration under the Natural-blame mode fails due a dynamic
+type check and blames the {\tt player} module; typing that module and running
+again then results in a type error, and hence the trail is successful.  By
+contrast, the Natural-exceptions mode (row 4) yields stack information for the
+root configuration that is unhelpful; it identifies only {\tt main}, which is
+already typed. Hence, this trail immediately gets stuck.
 

### 2021-05-17 writing/related.tex
edits to related work to fix monotoinc references discussion

@@
 The literature on gradual typing presents many semantics beyond the three used
-here and two additional blame strategies.  Pyret (pyret.org) assigns fixed-size
+here and three additional blame strategies.  Pyret (pyret.org) assigns fixed-size
 data types the Natural semantics and functions a Transient semantics. The
@@
 rat-oopsla-2017} assume that every value comes with a type tag and use tag
-checks to supervise the interactions between typed and untyped code.  Monotonic
-references~\citep{svctg-esop-2015} and the semantics~\citep{tlt-popl-2019,
+checks to supervise the interactions between typed and untyped code.  The semantics~\citep{tlt-popl-2019,
 etg-icfp-19, tt-scp-20, tgt-popl-18, tt-sas-17} derived with the Abstracting
-Gradual Typing technique~\citep{gct-popl-2016} are variants of Natural.  Among
-these representative semantics, only Amnesic and Nom present innovative blame
-strategies.  The experiment excludes these strategies because they seem to be
-merely theoretical constructions and/or impose severe restrictions on
-programmers.
+Gradual Typing technique~\citep{gct-popl-2016} are variants of Natural.
+ Monotonic References~\citep{svctg-esop-2015, rsfbv-popl-2015, sfrbcsb-popl-2014, kas-pldi-2019} 
+differ from Natural in the treatment of mutable data by
+associating heap locations with a type and permitting only updates that 
+make that type more precise.  
+Among these representative semantics, only Amnesic, Nom and Monotonic References present innovative blame
+strategies.  The experiment excludes the first because it is
+merely a theoretical construction, the second because it imposes severe restrictions on
+programmers and the third because it would require a complete
+re-engineering of the Racket runtime. That said, the paper offers a
+detailed chart that the creators of these blame strategies can transplant to their
+linguistic setting and determine the value of their blame strategies.

### 2021-05-18 writing/discussion.tex
editng discussion about reticulated

@@
 At first glance, these measurements seem to contradict the results
-of~\citet{vss-popl-2017}. They report an average slowdown of 2.5x and a
-worst-case of 5.4x on the fully-typed Python benchmarks in Reticulated Python.
-Unfortunately, the high cost of Shallow Racket seems closer to the truth.
-There are at least four broad issues that skew the earlier results:
+of~\citet{vss-popl-2017}. They report an average slowdown of 6.2x and a
+worst-case of 17.2x on the fully-typed Python benchmarks in Reticulated
+Python with the blame map enabled.  Unfortunately, the orders-of-magnitude
+higher cost of Shallow Racket seems closer to the truth.  There are at
+least four broad issues that skew the earlier results:
 \begin{enumerate}
@@
 
-\item The types of almost all local variables in the "fully typed" benchmarks is
-{\tt Dynamic}, often due the impoverished nature of the type system, which does
-not allow the ascription of precise types.  Code that ends up with the type {\tt
+\item While Reticulated attempts to infer types for local variables,
+  often local variables end up with type
+{\tt Dynamic} due the impoverished nature of the type system,  which does
+    not allow the ascription of precise
+    types~\citet{g-dissertation-2020}. Code with the type {\tt
 Dynamic} has fewer constraints to check at run-time---and much less information
@@
 
-\item On non-trivial programs, Reticulated Python suffers from the same high
-overhead as Shallow Racket.  It runs the simplest benchmark from the GTP suite (translated into Python)
-in approximately 40 seconds without blame and times out after 10 minutes when
-the blame map is enabled.
+\item Reticulated Python does not seem to employ any crucial
+  implementation techniques that can mitigate the significant slowdown due to
+    the blame map observed in Shallow Racket. For example Reticulated Python  runs the simplest
+    benchmark from the GTP suite (translated into Python) in approximately
+    40 seconds without blame but times out after 10 minutes with blame.
 
 \end{enumerate}
-More work on Transient blame is needed to make an informed decision.
+More work on Transient blame is needed to make an informed decision about
+its prospects as a viable production-level approach.
 

### 2021-05-18 writing/introduction.tex
improve prose, mark up 2 questionable phrases

--- 2021-04-30 writing/introduction.tex
@@
 blame can help practicing programmers find impedance mismatches, that is,
-disagreements between the type ascriptions of a software system and its
+disagreements between the type ascriptions of a software component and its
 behavior.
@@
 principle they validate the conjectures behind the work of theoreticians.
-First, a good blame assignment strategy helps with the search for
+\mf{don't we also want to say the run-time checks alone help?}
+A blame assignment strategy helps with the search for
 impedance mismatches between the specified types and the behavior of
-untyped components.  Second, Natural's wrapper-based blame tracking is
+untyped components.  Natural's wrapper-based blame tracking is
 more useful than Transient's ``collective blame'' tracking algorithm,
 which in turn is superior to Erasure. {\em But\/}, the application of the
-method also indicates problems with the expectations of theoreticians. The
-first problem concerns the difference between the methods; neither is {\it
-Natural\/} vastly more useful than {\it Transient\/} nor are the two
-blame assignment methods highly superior to {\it Erasure\/}.  The second
-one is that the cost of Transient's blame can be huge.  In turn, these
+method also indicates problems with the expectations of theoreticians. 
+In contrast with the theoretical differences between the methods,
+{\it Natural\/} is only marginally more useful than {\it Transient\/}, and neither of the two
+blame assignment methods are highly superior to {\it Erasure\/}. Additionally, 
+the cost of Transient's blame can be huge.  In turn, these
 problems suggest that, on one hand, the existing theory does not properly
-predict practice (i.e. the behavior of blame in real systems with real programs),
-and on the other hand, the existing practice may need
-additional experiments to collect more data.
+predict the behavior of blame in real systems with real programs,
+and on the other hand, the existing practice lacks data and alternative
+experiments to assess the entire landscape of blame assignment.

### 2021-05-18 writing/landscape.tex
improve prose, mark up 2 questionable phrases

@@
 \item[\texttt{pack-lib}] (at the top right) represents a library that provides,
-among others, a function {\tt pack}. As the documentation says, this function
-consumes JSON data and packages it in an association list (like a dictionary). By mistake, though, the
-function returns a hash table instead of the association list its documentation
-promises.
+among others, a function {\tt pack}. The documentation says this function
+consumes JSON data and packages it in an association list. In reality, though, the
+function returns a hash table instead of the association list.
 
@@
 use it. Imagine a programmer who relies on the types in the \typecolor\
-modules as checked documentation but prototypes the client in an untyped manner.
+modules as checked documentation but prototypes the client in the untyped language.
 
 \end{description}
-Due to the mistake in {\tt pack-lib}, each of the three semantics assigns a
-different outcome to the client module and signals different kinds of errors.
+The mistaken comment in {\tt pack-lib} causes an impedance mismatch,
+with which each of the three semantics deals differently.
 
@@
 the run-time system checks that it is a function and wraps it in a protective
-proxy, which in turn, enforces the type of the function arguments with run-time checks.
+proxy, which in turn, enforces the type of the function result with run-time checks.
 Analogously the run-time system wraps each exported function of a typed module
-(such as {\tt crypto-pack}) in a proxy that checks its arguments.  These checks
-protect functions exported from typed modules against applications in untyped
-code that supply the wrong kind of arguments.
+such as {\tt crypto-pack} in a proxy that checks its arguments.  These checks
+protect functions exported from typed modules against applications to
+wrong arguments in untyped code.
 
@@
 an argument-type check fails, responsibility lies with {\tt client}. Due to the
-proxies wrapped around functions, Natural can easily track the responsible party
+proxies wrapped around functions, {\it Natural\/} can easily track the responsible party
 for each check. Thus, in the example of figure~\ref{fig:example}, as
 \texttt{pack} returns, the return-type check fails and Natural blames
-\texttt{pack-lib} for both {\tt define}s in {\tt client}.
+the boundary between \texttt{pack-lib} with {\tt typed-pack-lib} and
+{\tt crypti-packed-lib}, respectively, for the two {\tt define}s in {\tt client}.
 
-Under {\it the Transient semantics\/}, typed code is compiled so that all entry
-points to functions check their arguments at run time and all function calls
-check their return values against the expected type~\citep{vss-popl-2017}.
-Furthermore, Transient uses \emph{shallow} checks, meaning they inspect only the
-top constructor of a value. Since retrieving a value from within a structure (or
-list, array, hash table etc.) is performed via a function call, the contents of a
-complex value is checked on a piecemeal basis.
+Under {\it Transient\/}, typed code is compiled so that all entry points to
+functions check their arguments at run time and all function calls check their
+return values against the expected type~\citep{vss-popl-2017}.  Furthermore,
+Transient uses \emph{shallow} checks, meaning they inspect only the top
+constructor of a value. Since retrieving a value from within a structure (or
+list, array, hash table etc.) is performed via a function call, the contents of
+a complex value is checked on a piecemeal basis.
 
@@
 
-Under {\it the Erasure semantics\/}, the compiler checks the specified types and
+Under {\it Erasure\/}, the compiler checks the specified types and
 then discards them when it generates code. The generated code includes run-time
@@
 \begin{center}
-  \begin{tabular}{l|ll}
+  \begin{tabular}{l|l@{\qquad\quad}l}
                   & {\tt public-data}              & {\tt secret-data}                                 \\
 \hline %% ----------------------------------------------------------------------------------------
-{\it Natural\/}   & error, blaming {\tt pack-lib}  & error, blaming {\tt pack-lib}                     \\
+{\it Natural\/}   & error, blaming the boundary between & error, blaming the boundary between          \\
+                  & {\tt pack-lib} and {\tt typed-pack-lib} & {\tt pack-lib} and {\tt crypto-pack-lib} \\
                   &                                &                                                   \\

### 2021-05-18 writing/why-rational-methodology.tex
Add explicit answer to question (5) in sec 3.1
improve prose, mark up 2 questionable phrases

@@
 
-By contrast, when the rational programmer fails, it questions programming
-language research. Specifically, it is indicative of the limited predictive
+When the rational programmer fails, it questions programming
+language research. Specifically, it indicates limited predictive
 power of programming language theory with respect to the use of languages in
-practice. Put differently, it suggests flaws in language design revealed by
-misleading predictions.
+practice. Indeed, misleading predictions may even suggest flaws in language design.
 
@@
 
-\item Can this guidance be formulated as an algorithm on the program and the extra
-information?
+\item Can this guidance be formulated as an algorithm?
 
@@
 
+\item Based on these insights, the remaining sections detail a large-scale automated experiment.
+
 \end{enumerate}

### 2021-05-18 writing/why-rational.tex
improve prose, mark up 2 questionable phrases

@@
 current research on programming languages, the question seems to call for
-empirical studies similar to those of human-computer interaction. At the same
-time, a significant evaluation result demands a large amount of
+empirical studies similar to those of the human-computer interaction research
+area. At the same time, a significant result demands a large amount of
 data. As~\citet{lksfd-popl-2020} recently demonstrated, the way around this
-dilemma is to simulate a programmer's activities algorithmically on a large set
+dilemma is to simulate a programmer algorithmically on a large set
 of programming scenarios.
@@
 
-This paper dubs Lazarek et. al's idea {\em the rational programmer\/}.
+This paper generalizes Lazarek et. al.'s idea to {\em the rational programmer\/}.
 Like~\citet{mill1874essays}'s {\it homo economicus\/}, the rational programmer approximates the
@@
 method, searching for the source of the impedance mismatch by adding type
-annotations to the untyped parts of the program identified in the error.
-If the type checker rejects an annotation that is considered correct, the
+annotations to some of the untyped parts of the program identified in the error.
+If the type checker rejects an annotation derived from the context, the
 rational programmer has found the source of the problem.
@@
 and re-runs the program;
-the Wadler-Findler slogan says the new version of the program must now blame a different location for the problem, which the rational programmer can check in the same fashion.
+the Wadler-Findler slogan implies the new version of the program must now blame
+a different location for the problem, which the rational programmer can check in
+the same fashion. 
 Measuring this simulated
-programmer's behavior on a large number of debugging scenarios yields data that
+behavior on a large number of debugging scenarios yields data that
 is similar to data collected in a human-facing study.
@@
 main} specifies that {\tt na-client} consumes a {\tt Real} number and produces a
-non-positive one.\footnote{Typed Racket's type system reflects Racket's subset
-reasoning about numbers, not (only) machine-level
+non-positive one.\footnote{Racket's type system reifies
+reasoning about subsets of numbers, not machine-level
 representations~\citep{stathff-padl-12}.} A program execution ends in this error:
@@
 \end{verbatim}
-The last step of the rational programmer is to assign types to the {\tt server}
-module and to re-direct the import of {\tt layer-typed} to {\tt (submod ".."
-server-typed)}. At this point the type checker objects to the conjectured type
-of {\tt neg-abs}---and the source of the impedance mismatch is found. How to fix
-it is a separate question.
+Lastly, the rational programmer assigns types to {\tt server} and to re-direct
+the import of {\tt layer-typed} to {\tt (submod ".." server-typed)}. Now the
+type checker objects to the conjectured type of {\tt neg-abs}, i.e.\ the source
+of the impedance mismatch is found. How to fix it is a separate question.
 
@@
 is also not to claim that equipping entire modules with types represents an
-always feasible approach (though adding types at the expression level, say, as
-in TypeScript should be considered well within bounds).  {\em But\/}, the
+always feasible approach\footnote{Adding types at the expression level, say, as
+in TypeScript should be considered well within bounds.}  {\em But\/}, the
 concept of studying the idea of an economically rational actor has produced

### 2021-05-19 writing/discussion.tex
local edits here and there

@@
 of~\citet{vss-popl-2017}. They report an average slowdown of 6.2x and a
-worst-case of 17.2x on the fully-typed Python benchmarks in Reticulated
-Python with the blame map enabled.  Unfortunately, the orders-of-magnitude
-higher cost of Shallow Racket seems closer to the truth.  There are at
-least four broad issues that skew the earlier results:
+worst-case slowdown of 17.2x on the fully-typed Python benchmarks in Reticulated
+Python when the blame map gets enabled.  Unfortunately, the average slowdown of 
+133x and the
+worst-case slowdown of 560x  due to to blame  in Shallow Racket seems closer to the truth.  There are at
+least three broad issues that skew the earlier results:
 \begin{enumerate}
@@
 
-\item While Reticulated attempts to infer types for local variables,
-  often local variables end up with type
-{\tt Dynamic} due the impoverished nature of the type system,  which does
-    not allow the ascription of precise
-    types~\citet{g-dissertation-2020}. Code with the type {\tt
-Dynamic} has fewer constraints to check at run-time---and much less information
+\item While Reticulated attempts to infer types for local variables, the impoverished nature of its type
+    system does not allow the ascription of precise types and often resorts
+    to type {\tt Dynamic}~\citep[section~5.4.4]{g-dissertation-2020}. 
+    Code with the type {\tt Dynamic}
+    has fewer constraints to check at run-time---and much less information
     to track in the blame map.
@@
 unrealistic, and unstable.\footnote{Release notes:
-\url{https://pyperformance.readthedocs.io/changelog.html}}
-
-\item Reticulated Python does not seem to employ any crucial
-  implementation techniques that can mitigate the significant slowdown due to
-    the blame map observed in Shallow Racket. For example Reticulated Python  runs the simplest
-    benchmark from the GTP suite (translated into Python) in approximately
-    40 seconds without blame but times out after 10 minutes with blame.
+\url{https://pyperformance.readthedocs.io/changelog.html}}  On the flip
+    side, all the benchmarks in the GTP suite are larger than the official Python
+    benchmarks. Reticulated Python  runs the
+    translation of the smallest GTP benchmark in
+    approximately 40 seconds without blame but times out after 10 minutes
+    with blame.
 

### 2021-05-19 writing/introduction.tex
local edits here and there

@@
 The paper's {\it second contribution\/} is a set of {\em results from
-applying the evaluation method to three distinct blame assignment
+applying the evaluation method to three distinct checking regimes for
+gradual types and their blame assignment
 strategies in approximately 72,000 different scenarios\/} (see
 sec.~\ref{sec:results}): (1)~{\it Transient\/}, i.e. Reticulated's
-tracking of typed/untyped boundaries~\citep{vss-popl-2017}; (2)~{\it
+inlined type assertions and collaborative tracking of typed/untyped boundaries~\citep{vss-popl-2017}; (2)~{\it
 Natural\/}, i.e. Typed Racket's use of a higher-order contract system and
 its blame assignment~\citep{ff-icfp-2002}; and (3) {\it Erasure\/}, i.e.
-the approach of industrial systems, which forgo blame in favor of error
+the approach of industrial systems, which forgo type checks and blame in favor of error
 messages from the safety checks of the underlying language. The results
@@
 principle they validate the conjectures behind the work of theoreticians.
-\mf{don't we also want to say the run-time checks alone help?}
-A blame assignment strategy helps with the search for
+Type checks and blame work together to help with the search for
 impedance mismatches between the specified types and the behavior of
-untyped components.  Natural's wrapper-based blame tracking is
-more useful than Transient's ``collective blame'' tracking algorithm,
-which in turn is superior to Erasure. {\em But\/}, the application of the
+untyped components.  Natural's wrapper-based type checks and blame tracking are
+more useful than Transient's type in-lined assertions and ``collective blame'' tracking algorithm,
+which in turn are superior to Erasure. {\em But\/}, the application of the
 method also indicates problems with the expectations of theoreticians. 
@@
 {\it Natural\/} is only marginally more useful than {\it Transient\/}, and neither of the two
-blame assignment methods are highly superior to {\it Erasure\/}. Additionally, 
+checking and blame assignment methods are highly superior to {\it Erasure\/}. Additionally, 
 the cost of Transient's blame can be huge.  In turn, these
@@
 and on the other hand, the existing practice lacks data and alternative
-experiments to assess the entire landscape of blame assignment.
+experiments to assess the entire landscape of the pragmatics of blame.

### 2021-05-19 writing/paper.tex
Fix citation style
Add blame trail example to discussion, in new subsection
authors
turn on ACM thingies
improve prose, mark up 2 questionable phrases
local edits here and there

@@
-\documentclass[acmsmall,review,anonymous]{acmart}
-\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[acmsmall]{acmart}
+% \settopmatter{printfolios=true,printccs=false,printacmref=false}
+
+\citestyle{acmauthoryear}
 
@@
 
-\author{Nobody In Particular}
-% \affiliation{\institution{PLT} \city{Evanston} \country{Illinois} }
-\affiliation{\relax} % \institution{PLT} \city{Evanston} \country{Illinois} }
-% \email{chrdim@northwestern.edu}          %% \email is recommended
+\author{Lukas Lazarek}
+\affiliation{%
+  \institution{PLT @ Northwestern University}
+  \city{Evanston}
+  \state{Illinois}
+  \country{USA}
+}
+\email{lukas.lazarek@eecs.northwestern.edu}
+
+\author{Ben Greenman}
+\orcid{0000-0001-7078-9287}
+\affiliation{%
+  \institution{PLT @ Northeastern University}
+  \city{Boston}
+  \state{Massachusetts}
+  \country{USA}
+}
+\email{benjaminlgreenman@gmail.com}
+
+\author{Matthias Felleisen}
+\orcid{0000-0001-6678-1004}
+\affiliation{%
+  \institution{PLT @ Northeastern University}
+  \city{Boston}
+  \state{Massachusetts}
+  \country{USA}
+}
+\email{matthias@ccs.neu.edu}
+
+\author{Christos Dimoulas}
+\orcid{0000-0002-9338-7034}
+\affiliation{%
+  \institution{PLT @ Northwestern University}
+  \city{Evanston}
+  \state{Illinois}
+  \country{USA}
+}
+\email{chrdimo@northwestern.edu}
+\renewcommand{\shortauthors}{Lazarek, Greenmanm Felleisen, Dimoulas}
 
@@
 \usepackage{upgreek}
+\usepackage{multirow}
 
@@
 %% use 'none' for review submission.
-\setcopyright{none}
+%% \setcopyright{none}
 
@@
 \begin{abstract}
-Programming language theoreticians develop blame assignment systems
-and prove blame theorems for gradually typed programming languages.
-Practical implementations of gradual typing almost completely ignore
-the idea of blame assignment.  This contrast raises the question
-whether blame assignment provides any value to the working programmer
-and poses the challenge of how to evaluate the effectiveness of blame
-assignment strategies. This paper contributes (1) the first evaluation
-method for blame assignment strategies and (2) the results from applying
-it to three different gradually typed languages. These results confirm that blame assignment
-is helpful to a rationally acting programmer.
+Programming language theoreticians develop blame assignment systems and
+  prove blame theorems for gradually typed programming languages.
+  Practical implementations of gradual typing almost completely ignore the
+  idea of blame assignment.  This contrast raises the question whether
+  blame provides any value to the working programmer and poses
+  the challenge of how to evaluate the effectiveness of blame assignment
+  strategies. This paper contributes (1) the first evaluation method for
+  blame assignment strategies and (2) the results from applying it to
+  three different semantics for gradual typing. These results cast doubt to
+  the theoretical effectiveness of blame. In most scenarios,  strategies
+  with imprecise blame assignment are as helpful to a rationally acting
+  programmer as strategies with provably correct blame.
 \end{abstract}

### 2021-05-19 writing/related.tex
editng discussion about reticulated
local edits here and there

@@
 Gradual Typing technique~\citep{gct-popl-2016} are variants of Natural.
- Monotonic References~\citep{svctg-esop-2015, rsfbv-popl-2015, sfrbcsb-popl-2014, kas-pldi-2019} 
-differ from Natural in the treatment of mutable data by
+The Monotonic semantics~\citep{svctg-esop-2015, rsfbv-popl-2015, sfrbcsb-popl-2014, kas-pldi-2019} 
+differs from Natural in the treatment of mutable data by
 associating heap locations with a type and permitting only updates that 
-make that type more precise.  
-Among these representative semantics, only Amnesic, Nom and Monotonic References present innovative blame
+do not make the type less precise.  
+Among these representative semantics, only Amnesic, Nom, and Monotonic present innovative blame
 strategies.  The experiment excludes the first because it is
 merely a theoretical construction, the second because it imposes severe restrictions on
-programmers and the third because it would require a complete
-re-engineering of the Racket runtime. That said, the paper offers a
-detailed chart that the creators of these blame strategies can transplant to their
-linguistic setting and determine the value of their blame strategies.
+programmers, and the third because it would require a complete
+re-engineering of the Racket runtime. 
+%That said, the paper offers a
+%detailed chart that the creators of these blame strategies can transplant to their
+%linguistic setting and determine the value of their blame strategies.

### 2021-05-19 writing/results.tex
moving broken wrap figure
Update plots and discussion of results with new data from re-run

@@
 %% -----------------------------------------------------------------------------
-
-\begin{wrapfigure}{r}{.40\textwidth} \footnotesize 
-  \vspace{-1.5em}
-  \includegraphics[width=0.38\textwidth]{./plots/success-bars}
-  \vspace{1em}
-  \begin{minipage}{0.35\textwidth} \raggedright
-%   The estimated percentage of scenarios for which each mode succeeds in locating the bug.
-   The upper bound margin of error is 0.02\%.
-  \end{minipage}
-\vspace{-2em}
-  \caption{Percentage rates of success} \label{fig:success-bars}
-\end{wrapfigure}
 
@@
 
+\
 \begin{figure} \footnotesize 
@@
      29.1\% of the scenarios, only Natural and Erasure succeed on 2.1\%, and
-     Natural alone succeeds on 9\%.  \\ The upper bound margin of error is 0.02\%.
+     Natural alone succeeds on 9\%. The upper bound margin of error is 0.02\%.
 
@@
 
+\begin{wrapfigure}{r}{.40\textwidth} \footnotesize 
+  \vspace{-1.5em}
+  \includegraphics[width=0.38\textwidth]{./plots/success-bars}
+  \vspace{1em}
+  \begin{minipage}{0.35\textwidth} \raggedright
+%   The estimated percentage of scenarios for which each mode succeeds in locating the bug.
+   The upper bound margin of error is 0.02\%.
+  \end{minipage}
+\vspace{-2em}
+  \caption{Percentage rates of success} \label{fig:success-bars}
+\end{wrapfigure}
+
 While most blame trail failures follow the above pattern, a few do
@@
 
-The second key observation from figure~\ref{fig:success-bars} is that the
-modes that use blame all outperform those that do not.  In particular,
-Natural and both of Transient's blame modes succeed in around 95\% of the
-scenarios, while their corresponding exception modes succeed in around
-85\% of them, and Erasure succeeds in only 60\%.  The only
-exception is that the random programmer always succeeds; this reflects the
-fact that every scenario has finitely many modules, so the random programmer 
-eventually types the buggy module.
+The second key observation from figure~\ref{fig:success-bars} is that the modes
+that use blame all outperform those that do not. In particular, Natural and both
+of Transient's blame modes succeed in 85 - 90\% of the scenarios, while their
+corresponding exception modes succeed in less than 80\% of them, and so too for
+Erasure. The only exception is that the random programmer always succeeds; the
+figure omits this mode because it just reflects the fact that every scenario has
+finitely many modules, so the random programmer eventually types the buggy
+module.
 
@@
 proportion of the inverse case.  For example, the plot titled ``Natural''
-shows that Natural outperforms Natural exceptions in about 12\% of the
+shows that Natural outperforms Natural exceptions in about 11\% of the
 scenarios, and the inverse (Natural performs worse than Natural
@@
 blame'' shows that Transient last blame outperforms Natural exceptions
- in about 12\% of the scenarios, but conversely it performs worse
+ in about 9\% of the scenarios, but conversely it performs worse
 than Natural exceptions in about 2\% of the scenarios.
@@
 Transient exceptions are more useful than Transient first and Transient
-last blame in a small percentage (about 2\%) of the scenarios.
+last blame in a small percentage (less than 1\%) of the scenarios.
 
-Figure~\ref{fig:avo-bars} also provides answers for $Q_*$.
-Blame for all three semantics is significantly
-more useful than Erasure exceptions (by around 35\%). Natural blame is
-more useful than both versions of Transient blame by a small percentage;
-in about 4\% of scenarios Natural blame is more useful while in about 1.5\% of the
-scenarios Transient blame is more useful. The Transient first and
-Transient last blame are practically indistinguishable. Finally, there is
-no clear winner between Natural exceptions and Transient exceptions
-despite the theoretically advantageous additional checks of Natural.
+Figure~\ref{fig:avo-bars} also provides answers for $Q_*$. Blame for all three
+semantics is significantly more useful than Erasure exceptions --- by almost
+12\% for Natural and almost 9\% for Transient. Natural blame is more useful than
+both versions of Transient blame by a small percentage (about 4\%). The
+Transient first and Transient last blame are practically indistinguishable.
+Finally, Natural exceptions are more useful than Transient exceptions, although
+only in a small percentage of scenarios (about 2.5\%). A rare few scenarios
+(about 0.5\%) show the opposite, despite the theoretically advantageous
+additional checks of Natural.
 
@@
     Each plot depicts the distribution of trail lengths for a given mode across all benchmarks.
+    The upper bound margin of error is 0.05\%.
   \end{minipage}
@@
     Each plot depicts the distribution of scenarios with trail length
-    differences ranging from -3 to 3.  A -x difference denotes that the first
-    mode's trail is x steps {\em shorter\/} than the second mode's trail for the
-    same scenario; a positive difference denotes the inverse.  Only
-    scenarios where both modes succeed are considered.  The 10|85 on the y-axis
-    indicates that the axis is truncated between 10 and 85\%.
+    differences ranging from -3 to 3. A $-x$ difference denotes that the first
+    mode's trail is $x$ steps {\em shorter\/} than the second mode's trail for the
+    same scenario; a positive difference denotes the inverse. A difference of
+    $\infty$ indicates one mode's trail succeeds while other mode's fails. The
+    15|60 on the y-axis indicates that the axis is truncated between 15 and
+    60\%.
+    The upper bound margin of error is 0.03\%.
 
@@
 For example, the leftmost diagram (Natural) shows that all three modes succeed
-on 59.5\% of the scenarios, only Natural and Natural exceptions succeed on
-25.7\% of the scenarios, only Natural and Erasure succeed on 2.1\%, and Natural
-alone succeeds on 9.7\%.  This analysis highlights the success trade-offs each
+on 75.7\% of the scenarios, only Natural and Natural exceptions succeed on
+11.6\% of the scenarios, only Natural and Erasure succeed on 1.8\%, and Natural
+alone succeeds on 9.2\%.  This analysis highlights the success trade-offs each
 semantics offers against Erasure, with and without blame. For instance, the
@@
 Turning to programmer effort, figure~\ref{fig:effort-table} shows the
-distribution of blame trail lengths for the sampled interesting debugging
-scenarios. Unlike the usefulness comparisons above, these proportions do
-not generalize to a representation of the full population.  Still, as
-section~\ref{subsec:effort} discusses, they provide an alternative view of
-the workings of the rational programmer.
-
-There are two immediate take-aways from the figure. First, the effort for
+estimated distribution of blame trail lengths for the interesting debugging
+scenarios. There are two immediate take-aways from the figure. First, the effort for
 successfully debugging interesting scenarios (in green) for the random
@@
 
-Figure~\ref{fig:effort-comparisons} provides head-to-head comparisons of
-effort.  The comparison between two modes boils down to the difference in
-length between their trails for all scenarios where they both succeed.
-Hence, each plot in the figure shows the distribution of scenarios with
-length differences ranging from -3 (the first mode's trail is 3 steps
-shorter than the second's) to 3 (the first mode's trail is 3 steps longer
-than the second's).  The figure offers several insights about how  modes
-compare in terms of effort that complement the insights about how they
-compare in terms of success rates from figure~\ref{fig:avo-bars}.  First,
-Natural blame never produces shorter trails than Natural exceptions, and
-sometimes (albeit rarely) produces slightly longer ones.  Hence, the
-experiment provides evidence that blame helps the rational programmer
-debug more scenarios but does not shorten the debugging process compared
-to exceptions. Second, Natural relatively often (around 10\% of the
-scenarios) produces shorter trails than both Transient blame modes, and
-sometimes the trails are significantly shorter.  Third, Transient last
-blame often produces shorter trails than Transient first blame, which is a
-particularly interesting distinction in light of their nearly
-indistinguishable success rates.  Finally, Transient's blame modes also
-share the characteristic with Natural that blame sometimes produces longer
-trails than their corresponding exception modes.
+Figure~\ref{fig:effort-comparisons} provides head-to-head comparisons of effort.
+The comparison between two modes boils down to the difference in length between
+their trails for all scenarios where they both succeed. Hence, each plot in the
+figure shows the distribution of scenarios with length differences ranging from
+-3 (the first mode's trail is 3 steps shorter than the second's) to 3 (the first
+mode's trail is 3 steps longer than the second's). The figure offers several
+insights about how modes compare in terms of effort that complement the insights
+about how they compare in terms of success rates from figure~\ref{fig:avo-bars}.
+First, Natural blame never produces shorter trails than Natural exceptions, and
+sometimes (albeit rarely) produces slightly longer ones. Hence, the experiment
+provides evidence that blame helps the rational programmer debug more scenarios
+but does not shorten the debugging process compared to exceptions. Second,
+Natural relatively often (close to 8\% of the scenarios) produces shorter trails
+than both Transient blame modes, and sometimes the trails are significantly
+shorter. Finally, Transient's blame modes also share the characteristic with
+Natural that blame sometimes produces longer trails than their corresponding
+exception modes.

### 2021-05-19 writing/discussion.tex
Update plots and discussion of results with new data from re-run
Add draft threat subsection about natural-biased scenarios

@@
 that disables some of its wrappers may offer an answer to the well-known
-performance issues of Natural~\citep{gtnffvf-jfp-2019}.  Fifth, given that both
-modes of the Transient rational programmer are equally successful and that
-Transient-last produces shorter trails, Reticulated Python could limit the size
-of blame sequences to the last few entries to reduce its serious performance
-problems (see below).
+performance issues of Natural~\citep{gtnffvf-jfp-2019}.
 
@@
 models have provided some predictive insights into the market's behavior; but as
-behavioral economics has shown more recently (e.g.~\cite{henrich2001search}), the mathematical abstraction of a
+behavioral economics has shown more recently (e.g.~\cite{henrich2001search}),
+the mathematical abstraction of a
 rational actor makes predictions also quite unreliable in some situations.
@@
 during debugging sessions.
+
+%% -----------------------------------------------------------------------------
+\subsection{Threat: Is the Definition of Interesting Scenarios Reasonable?} \label{sec:threat:erasure-bias}
+
+Section~\ref{sub:mutate-interesting} defines criteria for interesting mutations,
+one of which limits the scenarios under consideration to those with mistakes
+detectable by Erasure. In other words, the experiment only considers the utility
+of blame when no checking at all is necessary to detect a problem. In reality,
+checking is necessary to detect some mistakes, and it is possible that blame has
+more or less to offer for these kinds of mistakes. If that is the case, then the
+results of the experiment on a population of scenarios including such mistakes
+should be quite different.
+
+The authors have collected preliminary data that suggests this difference may be
+large. In particular, the data comes from running the experiment with a
+population including any scenario detectable by Natural, regardless of the other
+semantics; the population therefore also includes scenarios where Transient or
+Erasure may not even detect the mistake.
+%% While that detection disparity reflects reality in that such mistakes can be made in practice, it also biases the population to favor Natural in those new scenarios where Transient and Erasure have no chance to succeed.
+Under this broader population of scenarios, the preliminary experimental results
+suggest that Natural blame may be much more useful than all of the other modes.
 

### 2021-05-19 writing/introduction.tex
fix small words, vspace for table on pg. 4

@@
 principle they validate the conjectures behind the work of theoreticians.
-Type checks and blame work together to help with the search for
+Run-time type checks and blame work together to help with the search for
 impedance mismatches between the specified types and the behavior of

### 2021-05-19 writing/challenges.tex
tighten prose

--- 2021-04-30 writing/challenges.tex
@@
 
-While at first glance the method of section~\ref{sec:why-rational} seems
-straightforward to implement and test, three challenges remain. The first is
-comparing the effect of blame on the rational programmer across three different
-mechanisms; second is the collection of a large number of representative
-debugging scenarios; and third is finding a proper sampling strategy for the
+Implementing the method of the preceding section poses three challenges.  The
+first concerns the comparison of the effect of blame on the rational programmer
+across three different mechanisms; the second challenge is about finding a large
+number of representative debugging scenarios; and the third is the resulting
 huge space of possibilities. A coincidental challenge is the disparity of the
@@
 authors use Racket, which is thus far the only language in which all three major
-semantic variants are available in a robust and comparable manner:
-Typed Racket implements Natural, Shallow Racket~\citep{ttt21}
-Transient, and plain Racket for Erasure.
+semantic variants are available in a robust and comparable manner: Typed Racket
+implements Natural, Shallow Racket~\citep{ttt21} Transient, and plain Racket
+Erasure.
 
@@
 it comes with an exception location and a stack trace, which implicitly suggests
-parties to blame.  Each strategy affects the behavior of the rational programmer
-differently (and real ones, too).
+fixes.  Each strategy triggers different reactions by the rational programmer
+(and real ones, too).
 
@@
 information the rational programmer takes into account when debugging a
-scenario. Intuitively, with the common vocabulary of modes, different
-blame strategies become different ways the rational programmer  interprets
-blame.  For instance, one Transient mode may pick the oldest element of a
-blame sequence as the next component to add types because it corresponds to the
-earliest point in the program's evaluation that can discover an impedance
+scenario. Intuitively, different
+blame strategies correspond to different modes of operation.
+For instance, one Transient mode may assign types to the oldest element of a
+blame sequence because it corresponds to the
+earliest point in the execution that can discover an impedance
 mismatch.  Another mode may opt to treat the sequence as a stack and add
@@
 However, attributing the success of the rational programmer to this or
-that blame mechanism demands a careful look at the interplay between blame
-and the run-time checks of each gradual type system. When a check fails
+that blame mechanism demands a careful analysis of the interplay between blame
+and the run-time checks of each gradual type system. When a check fails,
 the Natural and Transient semantics assign blame instead of using the
-information in the exception from the run-time check. Nevertheless, the
+information in the exception from the run-time check. But, the
 exception information may be as useful to the rational
-programmer as a blame assignment. If this is the case, then blame may not play a critical
+programmer as a blame assignment. If this is the case, then blame {\em per se\/} may not play a critical
 role for the rational programmer. Indeed, precisely because they do not
 account for such confounding factors, \citet{lksfd-popl-2020} cannot draw any
-conclusions about blame specifically, despite advertisements for the
+conclusions about blame specifically, despite advertisements to the
 opposite. Their experiment may conclude only that so-called blame-shifting
-works, but they cannot attribute this conclusion to blame.
+works, but they cannot attribute this conclusion to blame alone.
 
-Modes offer a uniform way to compare the different semantics and  isolate
-blame from the effect of the semantics' run-time checks. Specifically, the Natural and
-Transient rational programmer come with a blame mode and an exception
-mode. If the former succeeds in debugging a program while the latter
-fails, then it is safe to conclude that blame is indeed necessary for the
-rational programmer. Put differently, the exception mode serves as the
-{\em baseline\/} for blame's value within a given semantics; if the
-programmer in this mode performs as well or better than the  blame one, a blame assignment
+Modes offer a uniform way to compare the different semantics and isolate blame
+from the effect of the semantics' run-time checks. Specifically, the rational
+programmer comes with a blame mode and an exception mode for Natural and
+Transient. If the blame mode succeeds in debugging a program while the exception
+one fails, it is safe to conclude that blame is indeed beneficial for the
+rational programmer. Put differently, the exception mode serves as the {\em
+baseline\/} for blame's value within a given semantics; if the programmer in
+this mode performs as well or better than the blame one, a blame assignment
 mechanism might be useless.
 
-An experiment must rule out that the usefulness of blame assignment is sheer
-luck.  Hence, a random rational programmer mode provides yet another necessary
-baseline using the same ingredients as for exceptions.
+An experiment must also rule out that the usefulness of blame assignment is sheer
+luck.  Hence, a completely random mode provides yet another necessary baseline.
 
@@
 accidentally create and that the run-time checks of academic systems catch. In
-other words, the experiment calls for a collection of type mistakes in
+other words, the experiment calls for a collection of mistakes in
 mix-typed programs that is representative of those ``in the wild.''
@@
 
-An alternative is to {\em generate a large and representative corpus of type
+An alternative is to {\em generate a corpus of 
 mistakes \/} using mutation analysis~\citep{lipton1971fault, demillo1978hints,
-jia2011analysis}. Conventional mutation analysis is useless, however.  Mutation
+jia2011analysis}, but conventional mutation analysis is useless.  Mutation
 analysis traditionally aims to inject bugs that challenge test suites, and it
@@
 While the goal of their work---to evaluate the quality of types in
-Python---is unrelated to the one here, the mechanism is related. And their
+Python---is unrelated to blame, the mechanism is related. And their
 judgment confirms the experience of the authors.
@@
 resources.  A practical execution has no option but to {\em sample the space of
-scenarios\/}, carefully ensuring that it still has reproducible
-conclusions.
+scenarios\/}, carefully ensuring reproducibility.
 

### 2021-05-20 writing/rational-effort.tex
tighten prose, get fig. 3 on same page as first reference

@@
 systems. If, for instance, both
-Natural blame and Transient first blame trails are successful for the same scenario, the two modes of
+Natural blame and Transient-first blame trails are successful for the same scenario, the two modes of
 the rational programmer can compete to see which one debugs the
@@
 rational programmer is an artifact of pure chance.
-In particular, the effort distribution for a mode of the rational programmer can be compared with that of
+In particular, the effort distribution for one mode can be compared with that of
 the random mode that ignores error information entirely and instead selects which

### 2021-05-20 writing/rational-experiment.tex
Typo fixes
tighten prose, get fig. 3 on same page as first reference

@@
 
-\begin{figure}[ht] \footnotesize
+\begin{wrapfigure}{r}{.450\textwidth}
+% \begin{figure}[ht]
+\footnotesize
 \center
@@
   \label{fig:experiment-outline}
-\end{figure}
+\end{wrapfigure}
 
-The above table summarizes how each question relates to
+The nearby table summarizes how each question relates to
 different kinds of trails/modes of the rational programmer. For example,
@@
 be added to the trail, the test bed checks whether the last scenario
-of the trail type-checks or not. If it does, the
+of the trail type-checks or not. If it does not, the
 Natural-blame trail is successful; otherwise it is failing. 
@@
 section~\ref{sec:mutate}.
-The process is repeated for the same roots with the other mode, Natural
-exceptions.
+The process is repeated for the same roots with the Natural-exceptions mode.
 After completing, the test bed
 reports the success/failure results of the trails to determine the proportion of
-scenarios where Natural blame is more useful than Natural exceptions.  Question
+scenarios where Natural-blame is more useful than Natural-exceptions.  Question
 $Q_1$ has a positive answer if a root exists where the above is true because it
@@
 programmer manages to debug because of blame information.  The process is analogous
-for $Q_2$ and $Q_3$, using the respective modes of the rational programmer.
+for $Q_2$ and $Q_3$, using the respective modes. % of the rational programmer.
 
@@
 the other and the inverse.  For instance, deciding whether blame for Natural is
-more useful than first blame for Transient requires comparing the percentage
-of scenarios where Natural blame is more successful than Transient
-first blame with the percentage where Transient first
-blame is more successful than Natural blame. Repeating the whole process
+more useful than Transient-first requires comparing the percentage
+of scenarios where the first is more successful than the second with
+the percentage where the second is more successful than the first. Repeating the whole process
 for every pair of modes produces a
-complete picture of the comparative usefulness of blame between all of the modes.
+complete picture of the comparative usefulness of blame. % between all of the modes.

### 2021-05-20 writing/rational-natural.tex
tex problems fixed
tighten prose, get fig. 3 on same page as first reference

@@
 
+\mf{Technically a Natural blame trail could be a Natural exception trail. Hmph? }
+
 Here is a rigorous definition. 
@@
 $\{\exception{\system, \conf_i}\}$  & otherwise \\ %
-\end{tabular} %
+\end{tabular} \right. %
 \]
@@
 any impedance mismatch can be detected. For example, the fully untyped
-configuration can only produce such an error (e.g. from {\tt +} receiving a
-string). Exceptions from the runtime do not carry blame, so
+configuration can only produce such an error, e.g. from {\tt +} receiving a
+string. Exceptions from the runtime do not carry blame, so
 the rational programmer proceeds using the accompanying stacktrace instead.
@@
 That is, failing Natural blame trails are those that end in a scenario that does not reveal the bug statically, yet also does not blame
-an untyped module. Thus the rational programmer gets
-no hints from the gradual type system on how to continue the search for the bug.
+an untyped module. Thus the rational programmer has
+no further hints on how to continue the search for the bug.
 
@@
 To account for this situation, a new mode of the Natural rational
-programmer follows a migration process based on entirely on exceptions.
+programmer follows a migration process based entirely on exceptions.
 \begin{quote}

### 2021-05-20 writing/rational-stuff.tex
tighten prose, get fig. 3 on same page as first reference

@@
 \citet{tfgnvf-popl-2016} describe the set of all possible type migrations with a
-lattice. Formally, a program $\system$ is a set of modules $\set{\component}$.
+lattice. A program $\system$ is a set of modules $\set{\component}$.
 Let a configuration $\conf$ of $\system$ be the subset of $\set{\component}$
@@
 by the subset relation and form a lattice $\lattice{\system}$ with
-$2^{\size{\system}}$ elements. The bottom scenario of $\lattice{\system}$ is
-$\emptyset$, the top one $\system$ itself; in between are the mixed-typed
-scenarios.
+$2^{\size{\system}}$ elements. The bottom of $\lattice{\system}$ is
+$\emptyset$, the top one $\set{\component}$ itself; in between are the mixed-typed ones.
 
@@
 \emph{trail}---in $\lattice{\system}$.  The {\em root\/} $\conf_0$ of a trail is
-the initial debugging scenario.  Starting from such a root, the rational
-programmer steps along a trail. If the (program for this) scenario type-checks,
+the initial debugging scenario. If the program for some scenario  $\conf_i$ type-checks,
 the rational programmer runs the program until it raises a run-time error.  The
-rational programmer uses the information in the error to decide which module to
+rational programmer then uses the information in the error to decide which module to
 equip with types. This choice constructs scenario $\conf_{i+1}$ from $\conf_{i}$

### 2021-05-20 writing/rational-transient.tex
tex problems fixed
tighten prose, get fig. 3 on same page as first reference

@@
 assignment says that the value witnessing the impedance mismatch may have
-crossed the boundaries between elements in the set, and each crossing
+crossed the boundaries between elements in the sequence, and each crossing
 checked the value's type in a shallow manner.
@@
 is that the rational programmer has at least two reasonable options. The first
-one is to select the module that is added to the blame sequence first and
+one is to select the untyped module that is added to the blame sequence first and
 assign types to only that one---after all, if fully checked, the types of this
@@
 $\{\exception{\system, \conf_i}\}$     & otherwise \\ %
-\end{tabular} %
+\end{tabular} \right. %
 \]
@@
 Natural. It is used to as a baseline for Transient-first and Transient-last
-blame.
+blame.\footnote{A reader may wonder whether the rational programmer
+should not just equip {\em all\/} modules in the Transient blame set
+with types. Doing so may accelerate the search for the impedance
+mismatch but it would also impose a large migration cost.}

### 2021-05-20 writing/rational.tex
tighten prose, get fig. 3 on same page as first reference

--- 2021-03-04 writing/rational.tex
@@
 rational programmer to equip a module with types if it is blamed in an error
-message. A sequence of such steps make up a portion of the migratory path in the
-lattice of type migration~\cite{tfgnvf-popl-2016}. The lattice serves as the
-common substrate for the definition modes of the rational programmer (see
+message. A sequence of such steps makes up a path in the
+lattice of type migration~\citep{tfgnvf-popl-2016}. The lattice serves as the
+universe \mf{what does this sentence really mean? what does it contribute?} for the definition modes of the rational programmer (see
 sec.~\ref{sub:stuff}).
@@
 within the same semantics and across different semantics.  
-
-Hence the research problem is develop modes for the rational programmer and to make
-them comparable even when they correspond to different semantics and offer
+Hence the research problem is to develop modes for the rational programmer and to make
+them comparable even when they correspond to different semantics and process
 different kinds of information (see secs.~\ref{sub:natural}
@@
 compared (see sec.~\ref{sub:effort}). With these notions in place, it
-becomes possible to state the experimental questions and gives an overview of
-the experimental process (see sec.~\ref{sub:experiment}).
+becomes possible to state the experimental questions and describes the process (see sec.~\ref{sub:experiment}).
 

### 2021-05-20 writing/challenges.tex
move Louisiana footnote so that the intro to the benchmark section makes sense

@@
 
-The {\em second challenge\/} is to find a representative collection of
-programs with impedance mismatches. The latter must represent mistakes that programmers
-accidentally create and that the run-time checks of academic systems catch. In
-other words, the experiment calls for a collection of mistakes in
-mix-typed programs that is representative of those ``in the wild.''
-Unfortunately no such collection exists, and with good reason. The kind of
-mistakes needed are typically detected by unit or integration tests; even if it
-takes some time to find their sources, these mistakes do not make it into code
-repositories with appropriate commit messages.
+The {\em second challenge\/} is to find a representative collection of programs
+with impedance mismatches.\footnote{\citet{cc-oopsla-20} created a collection of
+Reticulated Python programs to evaluate their technique of fixing mistakes in
+type annotations. Their collection does not come with type-level mistakes in the
+code itself.}  The latter must represent mistakes that programmers accidentally
+create and that the run-time checks of academic systems catch. In other words,
+the experiment calls for a collection of mistakes in mix-typed programs that is
+representative of those ``in the wild.''  Unfortunately no such collection
+exists, and with good reason. The kind of mistakes needed are typically detected
+by unit or integration tests; even if it takes some time to find their sources,
+these mistakes do not make it into code repositories with appropriate commit
+messages.
 

### 2021-05-20 writing/mutate-benchmarks.tex
move Louisiana footnote so that the intro to the benchmark section makes sense

@@
 
+\item be fully typed so that the choice of types is fixed;
+
+\item take advantage of the variety of typing features of a gradually
+typed; and
+
 \item have a decent number of type-able modules and a variety of module dependency graphs
- (because mixing of typed and untyped code in Typed Racket takes place at the
- module level);
+ because mixing of typed and untyped code in Typed Racket takes place at the
+ module level.
 
-\item take advantage of the variety of typing features of a gradually typed
- language; but
+\end{enumerate}
 
-\item fail and contain type-level mistakes. 
-\end{enumerate}
-No such benchmark suite exists.  Worse, no benchmark suite of
-gradually typed programs with type-level mistakes exists at
-all.\footnote{\citet{cc-oopsla-20} created a collection of Reticulated
-Python programs to evaluate their technique of fixing mistakes in type
-annotations. Their collection does not come with type-level mistakes
-in the code itself.}
-
-What exists is \citet{gtnffvf-jfp-2019}'s collection of Typed Racket programs
-for systematically measuring the implementation's performance. The benchmark
-suite consists of fully typed, correct programs, written by a number of
-different authors, who have maintained and evolved these programs over several
-years. The programs range widely in size, complexity, purpose, origin and in
-programming style. The latter is mostly a reference to their reliance on many
-Typed Racket features: occurrence typing~\citep{tf-icfp-2010}, types for mutable
-and immutable data structures~\citep{hpst-sfp-2010}, types for first-class
-classes and objects~\citep{tsdtf-oopsla-2012}, and types for Racket's numeric
-tower~\citep{stathff-padl-12}.
-Finally, all of the programs are deterministic, so any changes in the programs' behavior
-between runs can be solely attributed to the actions of the rational programmer.
+\citet{gtnffvf-jfp-2019}'s collection of Typed Racket programs for
+systematically measuring the implementation's performance satisfies these
+criteria. The benchmark suite consists of fully typed, correct programs, written
+by a number of different authors, who have maintained and evolved these programs
+over several years. The programs range widely in size, complexity, purpose,
+origin and in programming style. The latter is mostly a reference to their
+reliance on many Typed Racket features: occurrence typing~\citep{tf-icfp-2010},
+types for mutable and immutable data structures~\citep{hpst-sfp-2010}, types for
+first-class classes and objects~\citep{tsdtf-oopsla-2012}, and types for
+Racket's numeric tower~\citep{stathff-padl-12}.  Finally, all of the programs
+are deterministic, so any changes in the programs' behavior between runs can be
+solely attributed to the actions of the rational programmer.
 

### 2021-05-20 writing/mutate.tex
move Louisiana footnote so that the intro to the benchmark section makes sense

@@
 collection of representative programs (see sec.~\ref{sub:mutate-benchmarks}).
-Since existing mutators do not generate useful mutants, the first step is to
+Since existing mutators do not generate useful mutants, the next step is to
 develop new mutators (see sec.~\ref{sub:mutate-mutators}) and to validate their
 suitability on the benchmarks (see sec.~\ref{sub:mutate-interesting}).
-
-% As section~\ref{sec:challenges} explains, the first challenge is to
-% generate a representative mistakes in representative software. This
-% section explains the specific choices made for the first application
-% of the method: the source and nature of the chosen benchmarks
-% (sec.~\ref{sub:mutate-benchmarks}), the newly developed mutators
-% (sec.~\ref{sub:mutate-mutators}), and why they are suitable for this
-% experiment (sec.~\ref{sub:mutate-interesting}).
 

### 2021-05-20 writing/why-rational.tex
move Louisiana footnote so that the intro to the benchmark section makes sense

@@
 
-This paper generalizes Lazarek et. al.'s idea to {\em the rational programmer\/}.
-Like~\citet{mill1874essays}'s {\it homo economicus\/}, the rational programmer approximates the
-behavior of a software developer who reduces time spent on a task by exploiting the
-available information. In the context of gradual typing, the rational programmer
-has two pieces of information when an impedance mismatch signals exceptional
-behavior: the error message and the state of the program. Hence, the most
-rational procedure is to use the former to improve the latter. Specifically,
-the rational programmer translates the Wadler-Findler slogan into a debugging
-method, searching for the source of the impedance mismatch by adding type
-annotations to some of the untyped parts of the program identified in the error.
-If the type checker rejects an annotation derived from the context, the
-rational programmer has found the source of the problem.
-Otherwise, the rational programmer concludes that the just-annotated parts are not the problem
-and re-runs the program;
-the Wadler-Findler slogan implies the new version of the program must now blame
-a different location for the problem, which the rational programmer can check in
-the same fashion. 
-Measuring this simulated
-behavior on a large number of debugging scenarios yields data that
-is similar to data collected in a human-facing study.
+This paper generalizes Lazarek et. al.'s idea to {\em the rational
+programmer\/}.  Like~\citet{mill1874essays}'s {\it homo economicus\/}, the
+rational programmer approximates the behavior of a software developer who
+reduces time spent on a task by exploiting the available information. In the
+context of gradual typing, the rational programmer has two pieces of information
+when an impedance mismatch signals exceptional behavior: the error message and
+the state of the program. Hence, the most rational procedure is to use the
+former to improve the latter. Specifically, the rational programmer translates
+the Wadler-Findler slogan into a debugging method, searching for the source of
+the impedance mismatch by adding type annotations to some of the untyped parts
+of the program identified in the error.  If the type checker rejects an
+annotation derived from the context, the rational programmer has found the
+source of the problem.  Otherwise, the rational programmer concludes that the
+just-annotated parts are not the problem and re-runs the program---which must,
+by the slogan, blame a different location for the problem. At this point, the
+rational programmer can iterate the process.  Measuring this simulated behavior
+on a large number of scenarios yields data that is similar to data collected in
+a human-facing study.
 

### 2021-05-20 writing/mutate-benchmarks.tex
move Louisiana footnote so that the intro to the benchmark section makes sense

@@
 over several years. The programs range widely in size, complexity, purpose,
-origin and in programming style. The latter is mostly a reference to their
-reliance on many Typed Racket features: occurrence typing~\citep{tf-icfp-2010},
-types for mutable and immutable data structures~\citep{hpst-sfp-2010}, types for
-first-class classes and objects~\citep{tsdtf-oopsla-2012}, and types for
-Racket's numeric tower~\citep{stathff-padl-12}.  Finally, all of the programs
-are deterministic, so any changes in the programs' behavior between runs can be
-solely attributed to the actions of the rational programmer.
+origin and in programming style. They rely on many Typed Racket features:
+occurrence typing~\citep{tf-icfp-2010}, types for mutable and immutable data
+structures~\citep{hpst-sfp-2010}, types for first-class classes and
+objects~\citep{tsdtf-oopsla-2012}, and types for Racket's numeric
+tower~\citep{stathff-padl-12}.  Finally, all of the programs are deterministic,
+so any changes in the programs' behavior between runs can be solely attributed
+to the actions of the rational programmer.
 

### 2021-05-20 writing/mutate-benchmarks.tex
section 6, rest

@@
 
-Hence, even if they do not come with type-level errors, they offer a diverse
-starter suite, suitable for the careful introduction of type-level errors via
-mutations. Further filtering the benchmark suite to keep the benchmarks with the
-most sizeable and complex dependency graphs yields ten suitable programs while
-preserving the suite's linguistic diversity. Figure~\ref{table:benchmark-descriptions} displays the selected
-benchmark programs and describes their basic characteristics.
+Figure~\ref{table:benchmark-descriptions} describes the benchmark programs that
+meet all the criteria and come with a sizeable and complex dependency graphs.
+This additional filter is needed because Typed Racket demands type assignments
+for entire modules. For benchmarks with less than five modules, finding errors
+would almost be trivial.

### 2021-05-20 writing/mutate-interesting.tex
section 6, rest

@@
 
-The definition of interesting mutants is more than a filter for uninteresting
-mutants. It is also the guide for evaluating and refining mutators until they
-collectively produce a sizable and diverse population of interesting
-mutants. Consider a candidate for the set of mutators that casts the tests of conditional
-expressions to the {\tt Any} type, which would suppress occurrence typing. An
-examination of its mutants through the lens of ``interesting'' reveals that none
-of them is interesting, because they never raise run-time errors. Hence Indeed,
-the interesting standard guided countless iterations of adding and removing
-mutators and of refining the chosen mutators in figure~\ref{table:mutation-ops}.
+The goal of filtering for interesting mutants guided countless iterations of
+adding, removing, and refining mutators in figure~\ref{table:mutation-ops}.  For
+an illustrative example, Consider a candidate mutator that casts the tests of
+conditionals to the {\tt Any} type. Like the example explained at the end of the
+preceding subsection, this mutant would suppress occurrence typing. But, it
+would not be interesting because an execution would not raise a run-time error;
+instead the function would process its input as if nothing had changed. Hence
+this candidate mutator is not included in the final set.

### 2021-05-20 writing/mutate-mutators.tex
Add negate-cond mutator example to section 6.2
section 6, rest

@@
 
+\begin{wrapfigure}{r}{.5\textwidth} \footnotesize
+\hspace{0.2cm}
+\begin{module}{}\typecolor
+(: deal-with [(U Real False) -> Real])
+(define (deal-with optional-result)
+  (if optional-result
+      (+ optional-result OFFSET)
+      DEFAULT))
+
+(define DEFAULT 40)
+(define OFFSET  11)
+\end{module}
+\caption{Example program using occurrence typing}
+\label{fig:negate-cond-example}
+\end{wrapfigure}
+
 Most of the mutators are self-explanatory.  The first six apply to all
 gradually typed languages; the next six to those that include classes and
-objects. The last four target distinguishing features of Typed
-Racket's type system,
-specifically its sophisticated type system. For example, one mutation produced by {\tt arithmetic}
-replaces a {\tt +} with a {\tt -} in an attempt to change the type of the
-arithmetic expression; {\tt +}'s result is a {\tt Positive-Integer} when all
-arguments are positive integers, while {\tt -} yields {\tt
-Integer}~\citep{stathff-padl-12}. Similarly, the other three aim to confound
-the occurrence type system.
+objects. The last four target distinguishing features of Typed Racket's type
+system, specifically its sophisticated type system. For example, one mutation
+produced by {\tt arithmetic} replaces a {\tt +} with a {\tt -} in an attempt to
+change the type of the arithmetic expression; {\tt +}'s result is a {\tt
+Positive-Integer} when all arguments are positive integers, while {\tt -}
+yields {\tt Integer}~\citep{stathff-padl-12}. Similarly, the other three aim to
+confound the occurrence type system. Figure~\ref{fig:negate-cond-example}
+illustrates how this confusion works. The function deals with an input that is
+either a {\tt Real} or {\tt \#false}; the conditional deals with the first type
+in the {\tt then} branch and the second type in the {\tt else} branch.  If a
+mutator wraps {\tt (not $\cdot$)} around the test of the conditional, the
+resulting mutant no longer type checks and, when run, this function eventually
+causes a run-time type check to signal an impedance mismatch.

### 2021-05-20 writing/paper.tex
fix small words, vspace for table on pg. 4
move overfullrule to top of paper file so it's easy to turn off
turn of Acknws so we don't forget

@@
 \documentclass[acmsmall]{acmart}
-% \settopmatter{printfolios=true,printccs=false,printacmref=false}
+\overfullrule=.1mm
 
@@
   blame assignment strategies and (2) the results from applying it to
-  three different semantics for gradual typing. These results cast doubt to
-  the theoretical effectiveness of blame. In most scenarios,  strategies
+  three different semantics for gradual typing. These results cast doubt on
+  the theoretical effectiveness of blame in gradual typing. In most scenarios, strategies
   with imprecise blame assignment are as helpful to a rationally acting
@@
 %% Acknowledgments
-%%\begin{acks}                 
-  %% acks environment is optional
-  %% contents suppressed with 'anonymous'
-  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
-  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
-  %% acknowledge financial support and will be used by metadata
-  %% extraction tools.
-%%\end{acks}
+\begin{acks}                 
+Felleisen and Greenman were partly supported by NSF grants SHF
+1763922. Greenman also received support from a CRA-NSF Post-doctoral
+fellowship grant. Dimoulas and Felleisen wish to thank Max New for
+extensive discussions of the early ides in this paper. 
+\end{acks}
 

### 2021-05-20 writing/conclusion.tex
Rephrase start of conclusion section
the rest, minus the revisions of the results and discussion sections

@@
 The work of~\citet{tgpk-dls-2018} suggests that, by intuition, programmers
-prefer the run-time checking of Natural over other soundness methods. But just
-because a random set of programmers expresses this view does not mean that blame
+prefer the run-time checking of Natural over other soundness methods. But of course,
+the opinion of a random set of programmers does not mean that blame
 assignment adds value. Similarly, researchers and language designers have
@@
 justification for the programmers' leanings and helps language creators revisit
-their decisions. Of course, the design choice remains a trade-off, and the
-the experiment contribute to only one of its many facets.
+their decisions. Of course, the design choice remains a trade-off
+along several dimensions while the presented experiment sheds light on
+only one of them.
 
@@
 already-existing libraries. Instead of converting the libraries themselves,
-language implementors merely create typed facade modules that import untyped
-functions and export them with type annotations, like {\tt typed-pack-lib} in
-figure~\ref{fig:example}.  With those facades, the compiler can easily
-type-check typed modules. The problem is that these retroactive additions of
-types to a library may result from a misunderstanding of the code. \emph{Any
-retroactively ascribed type may thus be a mistake itself.}
+language implementors merely create facade modules that import untyped functions
+and export them with type annotations, like {\tt typed-pack-lib} in
+figure~\ref{fig:example}.  With those facades, the compiler can 
+type-check typed modules. But, these retroactive additions of types to a library
+may result from a misunderstanding of the code. \emph{Any retroactively ascribed
+type may thus be a mistake itself.}
 
@@
 type-level code mutators presented here. Based on the gradual typing literature,
-the Natural semantics should usually discover such mistakes. In contrast, the
-Transient and Erasure semantics cannot help with such mistakes at all; indeed,
-we expect that these latter two raise misleading exceptions or produce
-plainly-wrong answers.  Of course, only additional experimental work can confirm
-or reject these conjectures.
+the Natural semantics should discover such mistakes. In contrast, the Transient
+and Erasure semantics cannot help with such mistakes at all; indeed, we expect
+that these latter two raise misleading exceptions or produce wrong answers
+without warning.  Only additional experimental work can confirm or reject these
+conjectures.

### 2021-05-20 writing/discussion.tex
the rest, minus the revisions of the results and discussion sections

@@
 %% -----------------------------------------------------------------------------
-% \subsection{Contextualizing the results through an example scenario}
+Interpreting the numeric summaries and aggregations of the preceding section
+demands an intuitive understanding of what blame trails look like in practice.
+A concrete example of blame trails and programmer modes is a good basis for
+synthesizing this kind of intuition.
 
@@
 \end{figure}
-
-Interpreting the numeric summaries and aggregations of the last section demands
-an intuitive understanding of what blame trails look like in practice.  A
-concrete example of blame trails and programmer modes is a good basis for
-synthesizing this kind of intuition.
 
@@
 in five of the six modes (the two Transient blame modes behave the same).
-The reader may keep these scenarios in mind as we next
-discuss the numeric results of the last section.
+The reader may keep these scenarios for the following discussion of the numeric results.
 

### 2021-05-20 writing/related.tex
the rest, minus the revisions of the results and discussion sections

--- 2021-05-17 writing/related.tex
@@
 
-\citet{lksfd-popl-2020}'s human-out-of-the-loop empirical analysis of blame for
-higher-order contracts is the inspiration for this paper.  While they do not
-spell out the notion of the rational programmer, they present the basic
-ingredients of the technique.
+At a philosophical level, \citet{lksfd-popl-2020} present the first empirical
+analysis {\em without\/} involving humans in this general area.  While they do
+not spell out the notion of the rational programmer, they present many of the
+basic ingredients.
 
-At a technical level, this work and~\citet{lksfd-popl-2020}'s differ in
-many ways. The experiment presented here concerns three different modes of
-gradual typing, theirs a single notion of contract checking.  It also
-contributes the idea of creating three comparable rational programmers
-(with several modes). Thus the method of this paper answers whether blame
-adds value to a gradual type system and what gradual type system it
-benefits the most; while~\citet{lksfd-popl-2020} only establish a
-systematic relation between blame for contracts and behavioral bugs. The
-type-level mutators represent another technical contribution of the work
-presented here; as explained in section~\ref{sec:mutate}, almost none
-of~\citet{lksfd-popl-2020}'s mutators is useful at the level of type
-mistakes.
+At a technical level, the two pieces of work differ in many ways.  First the
+experiment presented here involves three different modes of gradual typing,
+theirs a single notion of contract checking.  Second, this paper also
+contributes the idea of creating three comparable rational programmers, with
+several modes. Hence it can answer whether blame adds value to a gradual type
+system and which gradual type system it benefits the most, while theirs
+establishes only a systematic relation between blame for contracts and
+behavioral bugs. Finally, this paper also contributes type-level mutators. As
+explained in section~\ref{sec:mutate}, almost none of~\citet{lksfd-popl-2020}'s
+mutators is useful in the context of gradual typing.
 
@@
 Gradual Typing technique~\citep{gct-popl-2016} are variants of Natural.
- Monotonic References~\citep{svctg-esop-2015, rsfbv-popl-2015, sfrbcsb-popl-2014, kas-pldi-2019} 
-differ from Natural in the treatment of mutable data by
-associating heap locations with a type and permitting only updates that 
-make that type more precise.  
-Among these representative semantics, only Amnesic, Nom and Monotonic References present innovative blame
+The Monotonic semantics~\citep{svctg-esop-2015, rsfbv-popl-2015, sfrbcsb-popl-2014, kas-pldi-2019} 
+differs from Natural in the treatment of mutable data. It 
+associates every heap location with a type and permits only updates that 
+do not make the type less precise.  
+Among these semantics, only Amnesic, Nom, and Monotonic present interesting blame
 strategies.  The experiment excludes the first because it is 
 merely a theoretical construction, the second because it imposes severe restrictions on
-programmers and the third because it would require a complete
-re-engineering of the Racket runtime. That said, the paper offers a
-detailed chart that the creators of these blame strategies can transplant to their
-linguistic setting and determine the value of their blame strategies.
+programmers, and the third because it requires a re-engineering of the Racket runtime.

### 2021-05-20 writing/sample.tex
L says it's a typo
the rest, minus the revisions of the results and discussion sections

@@
 
-To become computationally feasible and statistically sound, the experiment employs
-stratified uniform random sampling with three levels of stratification.
-The first two levels group mutants first by benchmark and then by
-mutator, making sure that the sample within each benchmark reflects the diversity of
-mutants with respect to the mutators that generated them.  Specifically, the
-experiment samples 80 interesting mutants per benchmark,
+To render the experiment computationally feasible and statistically sound, it
+becomes necessary to sample this large space in a uniform and stratified manner.
+The first two levels of stratification group mutants first by benchmark and then
+by mutator, making sure that the sample within each benchmark reflects the
+diversity of mutants with respect to the mutators that generated them.
+Specifically, the experiment samples 80 interesting mutants per benchmark,
 evenly-distributed across all of the mutators that contribute mutants for the
 benchmark. Some benchmarks have less than 80 mutants with interesting scenarios,
-in which case the only choice is to include them all.
-The result is a total of 756 interesting mutants across all benchmarks.
-Finally, the third level of sampling draws 96 random interesting debugging
-scenarios from each configuration lattice with replacement. The
-final sample thus consists of 72,192 interesting scenarios.
-
-%% MF: I don't get this number ^^^^
-% > (* 756 96)
-% 72576
+in which case the only choice is to include them all.  The result is a total of
+752 interesting mutants across all benchmarks.  Finally, the third level of
+sampling randomly draws 96 debugging scenarios from each configuration lattice
+\mf{is ``with replacement'' correct here? Do you put it back for the next draw?}
+with replacement. The final sample thus consists of 72,192 interesting
+scenarios.

### 2021-05-21 writing/discussion.tex
Tweak my previous edits in discussion

--- 2021-05-19 writing/discussion.tex
@@
 that disables some of its wrappers may offer an answer to the well-known
-performance issues of Natural~\citep{gtnffvf-jfp-2019}.  Fifth, given that both
-modes of the Transient rational programmer are equally successful and that
-Transient-last produces shorter trails, Reticulated Python could limit the size
-of blame sequences to the last few entries to reduce its serious performance
-problems (see below).
+performance issues of Natural~\citep{gtnffvf-jfp-2019}.  Fifth, the fact that both
+modes of the Transient rational programmer are equally successful suggests that
+returning the whole blame sequence may not be beneficial. If so, Reticulated
+Python could limit the size of blame sequences to attempt to mitigate its
+serious performance problems (see below).
 
@@
 during debugging sessions.
+
+%% -----------------------------------------------------------------------------
+\subsection{Threat: Is the Definition of Interesting Scenarios Reasonable?} \label{sec:threat:erasure-bias}
+
+Section~\ref{sub:mutate-interesting} defines criteria for interesting mutations,
+one of which limits the scenarios under consideration to those with mistakes
+that raise a run-time error under Erasure. In other words, the experiment is
+Erasure-biased: it only considers the usefulness of blame when the safety checks
+of the underlying language alone are sufficient to detect the mistake. In
+reality, some mistakes require run-time type checks to be
+detected~\cite{gfd-oopsla-2019}, and it is possible that blame has more or less
+to offer for these kinds of mistakes. If that is the case, then the results of
+the experiment on a population of scenarios including such mistakes should be
+different.
+
+In fact, the authors have conducted a variation of the experiment that provides
+some preliminary evidence that this difference may be significant. In
+particular, the variation of the experiment includes only three of the
+benchmarks but broadens the selection of scenarios to include any that raise
+run-time errors under Natural, regardless of other semantics. The preliminary
+results from this experiment suggest that without Erasure-bias, Natural blame
+may be much more useful than all of the other modes.
 

### 2021-05-21 writing/conclusion.tex
Respond to and add side notes

@@
 blame-strategy dimension. The experiment presented in this paper provides some
-justification for the programmers' leanings and helps language creators revisit
+justification for the programmers' leanings
+\ll{Does it really?}
+and helps language creators revisit
 their decisions. Of course, the design choice remains a trade-off

### 2021-05-21 writing/rational.tex
Respond to and add side notes

@@
 lattice of type migration~\citep{tfgnvf-popl-2016}. The lattice serves as the
-universe \mf{what does this sentence really mean? what does it contribute?} for the definition modes of the rational programmer (see
+common vocabulary for defining the modes of the rational programmer (see
 sec.~\ref{sub:stuff}).
+\mf{what does this sentence really mean? what does it contribute?}
+\ll{Addressed}
 

### 2021-05-22 writing/conclusion.tex
fixing prose

@@
 
-The work of~\citet{tgpk-dls-2018} suggests that, by intuition, programmers
-prefer the run-time checking of Natural over other soundness methods. But of course,
+The interviews of~\citet{tgpk-dls-2018} suggests that programmers
+prefer the run-time checking of Natural over other soundness methods. But,
 the opinion of a random set of programmers does not mean that blame
@@
 justification for the programmers' leanings
-\ll{Does it really?}
+\ll{Does it really?} \mf{addressed}
 and helps language creators revisit
 their decisions. Of course, the design choice remains a trade-off
-along several dimensions while the presented experiment sheds light on
+along several dimensions, and  the presented experiment sheds light on
 only one of them.
@@
 figure~\ref{fig:example}.  With those facades, the compiler can 
-type-check typed modules. But, these retroactive additions of types to a library
-may result from a misunderstanding of the code. \emph{Any retroactively ascribed
+type-check typed modules, but these retroactive additions of types to a library
+may result from a misunderstanding of the code. \emph{In short, any retroactively ascribed
 type may thus be a mistake itself.}

### 2021-05-22 writing/discussion.tex
fixing prose

@@
 approach fares better than the Transient approach, but only by a small margin.
-Since Natural offers complete and sound path-based blame (i.e. it blames all and only those components
-that are relevant to a type impedance mismatch) while Transient offers
+Since Natural offers complete and sound path-based blame  while Transient offers
 incomplete but sound heap-based blame~\citep{gfd-oopsla-2019}, the results call
@@
 Fourth, given that Transient's sound but shallow run-time type checks do not
-seem to hamper debugging, a version of the Natural semantics of Typed Racket
-that disables some of its wrappers may offer an answer to the well-known
-performance issues of Natural~\citep{gtnffvf-jfp-2019}.  Fifth, the fact that both
+seem to hamper debugging, a language that supports {\em both\/} Natural and Transient 
+might help reduce the number of wrappers and thus address the well-known
+performance issues of sound gradual typing~\citep[chapter~6]{g-dissertation-2020}.  Fifth, the fact that both
 modes of the Transient rational programmer are equally successful suggests that
@@
 
-Like {\em homo economicus\/}, which decouples the actual behavior of a
+Like {\em homo economicus\/}, which idealizes the actual behavior of a
 participant in an economy for the sake of mathematical modeling, the model of a
-rational programmer decouples the actual debugging behavior of a software
-developer for the sake of a systematic, large-scale analysis. This decoupling
+rational programmer idealizes the actual debugging behavior of a software
+developer for the sake of a systematic, large-scale analysis. This idealization 
 comes with advantages and disadvantages. In the economic realm, mathematical
 models have provided some predictive insights into the market's behavior; but as
-behavioral economics has shown more recently (e.g.~\cite{henrich2001search}),
+behavioral economics has shown more recently~\cite{henrich2001search},
 the mathematical abstraction of a
@@
 to extend a trail. When the run-time checks signal an impedance mismatch in the
-real world, a real-world programmer does not have a typed module ready to swap
+real world, a programmer does not have a typed module ready to swap
 in. Instead, the programmer must come up with the next set of types, which means
 making choices. It is usually possible to consistently assign types to variables
-in a module in different ways. The maintenance of the benchmarks over many years
+in a module in several different ways. The maintenance of the benchmarks over many years
 has driven home this lesson but, fortunately, it has also shown that the types
@@
 reality, some mistakes require run-time type checks to be
-detected~\cite{gfd-oopsla-2019}, and it is possible that blame has more or less
+detected~\cite{gfd-oopsla-2019}, and it is possible that blame has more 
 to offer for these kinds of mistakes. If that is the case, then the results of
@@
 
-In fact, the authors have conducted a variation of the experiment that provides
-some preliminary evidence that this difference may be significant. In
-particular, the variation of the experiment includes only three of the
+In fact, a variation of the experiment provides
+some preliminary evidence that this difference may be significant. Thus far,
+the variation of the experiment covers only three of the
 benchmarks but broadens the selection of scenarios to include any that raise
-run-time errors under Natural, regardless of other semantics. The preliminary
-results from this experiment suggest that without Erasure-bias, Natural blame
+run-time errors under Natural, regardless of other semantics. The 
+results from this small experiment suggest that without Erasure-bias, Natural blame
 may be much more useful than all of the other modes.
@@
 The execution of the experiment reveals that Transient produces empty blame
-sequences for 967 scenarios. In theory, empty set must not occur. An empty blame
-sequence means a lack of boundary crossings for the witness value.  By
-implication, a typed module is blaming itself---something that can happen only
+sequences for 967 scenarios. An empty blame sequence means a lack of boundary
+crossings for the witness value. In theory, an empty sequence should not occur,
+because it means a typed module is blaming itself---something that can happen only
 if the type checker (or system) is unsound.
@@
 the Transient semantics also runs these programs in a short amount of time and
-well within the memory limits. In short, even though the Transient rational
+well within the memory limit. In short, even though the Transient rational
 programmer appears to do well in the experiment, the implementation of the
@@
 Python when the blame map gets enabled.  Unfortunately, the average slowdown of
-133x and the
-worst-case slowdown of 560x  due to to blame  in Shallow Racket seems closer to the truth.  There are at
-least three broad issues that skew the earlier results:
+133x and the worst-case slowdown of 560x due to blame in Shallow Racket seems
+closer to the truth.  There are at least three broad factors that skew Vitousek
+et. al.'s results:
 \begin{enumerate}
@@
     to type {\tt Dynamic}~\citep[section~5.4.4]{g-dissertation-2020}. 
-    Code with the type {\tt Dynamic}
+    Code with type {\tt Dynamic}
     has fewer constraints to check at run-time---and much less information

### 2021-05-22 writing/rational-natural.tex
fixing prose

@@
 \mf{Technically a Natural blame trail could be a Natural exception trail. Hmph? }
+\mf{I added a sentence below this definition and the next to address this issue.}
 
@@
 \end{quote}
+Note how this definition interprets ``blame information'' broadly, as
+any information from a failed run-time check in a blame-oriented semantics.
 
@@
 \end{quote}
+Using Natural exception trails it becomes possible to factor out
+``blame information'' in the narrow sense from the broad one of the
+above definition.
 

### 2021-05-23 writing/introduction.tex
intro: explain why 'programs' is wrong vs 'components'

@@
 very beginning~\citep{mf-toplas-2009, tf-dls-2006}. ``Well-typed
-[components]\footnote{The original authors got this word wrong.} can't be
+[components]\footnote{The original authors got this word wrong. A program has many components; blame helps identify a faulty one.} can't be
 blamed'' turned the theorem into a slogan~\citep{wf-esop-2009}. Academic

### 2021-05-23 writing/rational-natural.tex
TeX setting

@@
 \begin{quote}
-\it A Natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
-\emph{successful} iff (the program for) its last scenario $\conf_n$ does not type check.  A Natural
-blame trail $\conf_0,..,\conf_n$ in a lattice $\lattice{\system}$ is \emph{failing}
+\it A {\em Natural blame trail\/} $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
+\emph{successful}\/ iff (the program for) its last scenario $\conf_n$ does not type check.  A Natural
+blame trail $\conf_0,..,\conf_n$ in a lattice $\lattice{\system}$ is \emph{failing\/}
 iff (the program for) $\conf_n$ type checks and the trail cannot be extended further.
@@
   Given a program $\system$ and a root $\conf_0$ in $\lattice{\system}$,
-  Natural blame is \emph{more useful} than Natural exceptions for
+  Natural blame is \emph{more useful\/} than Natural exceptions for
   debugging $\conf_0$ iff 

### 2021-05-23 writing/rational-transient.tex
TeX setting

@@
 \begin{quote}
-\it A \emph{Transient-first blame trail}/\emph{Transient-last blame trail} is a sequence of scenarios
+\it A \emph{Transient-first blame trail}{\lrdef}\emph{Transient-last blame trail} is a sequence of scenarios
 $\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
@@
 \]
-where $\first{\mblame{\system, \conf}}$/$\last{\mblame{\system, \conf}}$ is the
-first/last module, respectively,  that Transient adds to the blame sequence for $\conf$.
+where $\first{\mblame{\system, \conf}}${\lrdef}$\last{\mblame{\system, \conf}}$ is the
+first{\lrdef}last module, respectively,  that Transient adds to the blame sequence for $\conf$.
 \end{quote}

### 2021-05-24 writing/conclusion.tex
Small fixes from meeting last week

@@
 blame-strategy dimension. The experiment presented in this paper provides some
-justification for the programmers' leanings
-\ll{Does it really?} \mf{addressed}
-and helps language creators revisit
+justification for the programmers' leanings and helps language creators revisit
 their decisions. Of course, the design choice remains a trade-off

### 2021-05-24 writing/results.tex
the rest, minus the revisions of the results and discussion sections
Small fixes from meeting last week

@@
 limit. Running the experiment on all debugging scenarios took over
-30,000 compute hours.
+30,000 compute hours or roughly three-and-a-half compute years.
 
@@
 Figure~\ref{fig:avo-bars} also provides answers for $Q_*$. Blame for all three
-semantics is significantly more useful than Erasure exceptions --- by almost
+semantics is significantly more useful than Erasure exceptions---by almost
 12\% for Natural and almost 9\% for Transient. Natural blame is more useful than

### 2021-05-25 writing/landscape.tex
fix small words, vspace for table on pg. 4
clarifying blame and boundaries + a small edit about narrow vs broad blame information

@@
 \end{description}
-The mistaken comment in {\tt pack-lib} causes an impedance mismatch,
-with which each of the three semantics deals differently.
 
-Under {\it the Natural semantics\/}, functions imported into and exported from
-typed modules are wrapped in proxies that enforce the static type discipline
-with run-time checks and track responsibilities~\citep{tf-popl-2008,
-tfffgksst-snapl-2017}. Thus, when {\tt pack} is imported into a typed module,
-the run-time system checks that it is a function and wraps it in a protective
-proxy, which in turn, enforces the type of the function result with run-time checks.
-Analogously the run-time system wraps each exported function of a typed module
-such as {\tt crypto-pack} in a proxy that checks its arguments.  These checks
-protect functions exported from typed modules against applications to
-wrong arguments in untyped code.
+The mistaken comment in {\tt pack-lib} causes an impedance mismatch, with which each
+of the three semantics deals differently.  Under {\it the Natural semantics\/},
+functions imported into and exported from typed modules are wrapped in proxies that
+enforce the static type discipline with run-time checks and track
+responsibilities~\citep{tf-popl-2008, tfffgksst-snapl-2017}. Thus, when {\tt pack}
+is imported into a typed module, the run-time system checks that it is a function
+and wraps it in a protective proxy, which in turn, enforces the type of the function
+result with run-time checks.  Analogously the run-time system wraps each exported
+function of a typed module such as {\tt crypto-pack} in a proxy that checks its
+arguments.  These checks protect functions exported from typed modules against
+applications to wrong arguments in untyped code.
 
-As this analysis implies, if a return-type check fails, the problem is with
-the untyped module that supplied an incorrect function, here {\tt pack-lib}. If
-an argument-type check fails, responsibility lies with {\tt client}. Due to the
-proxies wrapped around functions, {\it Natural\/} can easily track the responsible party
-for each check. Thus, in the example of figure~\ref{fig:example}, as
+As this analysis implies, if a return-type check fails, the problem is
+that the untyped module supplied a function, here {\tt
+pack-lib}, that is not a  match for the type ascribed by the typed module. 
+Hence either the type at the boundary between the two modules is wrong or, 
+if the programmer trusts the type, the untyped module is at fault.
+If an argument-type check fails, responsibility  lies with
+{\tt client}. After all either the type it ascribes to the argument is
+wrong or the argument it produced clashes with the type. Due to 
+proxies, {\it Natural\/} can easily track the 
+boundary, type and responsible parties that correspond to each check. 
+Thus, in the example of figure~\ref{fig:example}, as
 \texttt{pack} returns, the return-type check fails and Natural blames
 the boundary between \texttt{pack-lib} with {\tt typed-pack-lib} and
-{\tt crypti-packed-lib}, respectively, for the two {\tt define}s in {\tt client}.
+{\tt crypto-pack-lib}, respectively, for the two {\tt define}s in {\tt client}.
 
@@
 
-In order to locate the responsible party for failed checks, Transient
+In order to locate the corresponding boundaries for failed checks, Transient
 maintains a map from values to the boundaries between typed and untyped modules that
@@
 {\tt required/typed} forms in the example. Since the failed check corresponds to
-the return type of \texttt{pack}, Transient blames the sources of the two
-boundary crossings, which in this particular case is \texttt{pack-lib}. In
-general though, Transient blames more than one party. In fact, the theoretical
+the return type of \texttt{pack}, assuming that the type is correct, 
+the responsible parties are the sources of the two
+boundary crossings, \texttt{pack-lib}. In
+general though, Transient blames more than one boundary. In fact, the theoretical
 work of~\citet{gfd-oopsla-2019} shows that for some programs Transient
@@
 information in this exception, plus its stack trace, may help the programmer
-find the source of the impedance mismatch between the specified return types of
-{\tt pack} in the two typed modules and the actual return value of {\tt pack}.
+find the source of the impedance mismatch between the specified types of
+{\tt pack} in the two typed modules and its actual results.
 
@@
 \begin{center}
-  \begin{tabular}{l|l@{\qquad\quad}l}
+  \begin{tabular}{l|l@{\qquad}l}
                   & {\tt public-data}              & {\tt secret-data}                                 \\
@@
                   & {\tt pack-lib} and {\tt typed-pack-lib} & {\tt pack-lib} and {\tt crypto-pack-lib} \\
-                  &                                &                                                   \\
-{\it Transient\/} & no error                       & error, blaming the boundary crossings             \\
-                  &                                & \quad {\tt pack-lib} to {\tt typed-pack-lib}, and \\
-                  &                                & \quad {\tt pack-lib} to {\tt crypto-pack-lib}     \\
-                  &                                &                                                   \\
+\hline
+{\it Transient\/} & no error                       & error, blaming the boundaries between            \\
+                  &                                & {\tt pack-lib}  and {\tt typed-pack-lib}, and \\
+                  &                                & {\tt  pack-lib} and {\tt crypto-pack-lib}     \\
+\hline
  {\it Erasure\/}   & no error*                      & no error*                                         \\

### 2021-05-25 writing/rational-natural.tex
clarifying blame and boundaries + a small edit about narrow vs broad blame information

@@
 \mf{I added a sentence below this definition and the next to address this issue.}
+\cd{I revised a bit.}
 
@@
 \end{quote}
-Note how this definition interprets ``blame information'' broadly, as
-any information from a failed run-time check in a blame-oriented semantics.
+Note how, in the absence of ``blame information'' in the narrow sense, 
+this definition interprets ``blame information'' broadly, as
+any information from a failing run-time check.
 

### 2021-05-25 writing/rational.tex
Small fixes from meeting last week
clarifying blame and boundaries + a small edit about narrow vs broad blame information

--- 2021-05-20 writing/rational.tex
@@
 message. A sequence of such steps makes up a path in the
-lattice of type migration~\citep{tfgnvf-popl-2016}. The lattice serves as the
-universe \mf{what does this sentence really mean? what does it contribute?} for the definition modes of the rational programmer (see
+lattice of type migration~\citep{tfgnvf-popl-2016}. The lattice describes the
+space in which the modes of the rational programmer search for bugs (see
 sec.~\ref{sub:stuff}).
+%\mf{what does this sentence really mean? what does it contribute?}
+%\ll{Addressed}
 

### 2021-05-25 writing/paper.tex
turn of Acknws so we don't forget
Add Robby and NWU PLT to acknowledgements

@@
 \begin{acks}                 
-Felleisen and Greenman were partly supported by NSF grants SHF
+Felleisen and Greenman were partly supported by NSF grant SHF
 1763922. Greenman also received support from a CRA-NSF Post-doctoral
 fellowship grant. Dimoulas and Felleisen wish to thank Max New for
-extensive discussions of the early ides in this paper. 
+extensive discussions of the early ideas in this paper. Thanks also to
+Robby Findler and Northwestern PLT for their valuable feedback at
+various stages of this work.
 \end{acks}

### 2021-05-25 writing/discussion.tex
Small fixes from meeting last week
new subsection on threats

@@
 
-The validity of these conclusions is threatened in several ways: (i) the
+%% -----------------------------------------------------------------------------
+\subsection{What Are the Threats to Validity?}
+
+The validity of these conclusions is threatened in two fundamentally distinct
+ways. The first sets of threats have been mentioned before: (i) the
 representativeness of benchmarks; (ii) the relation between mutations and real
 programming mistakes; (iii) the definition of interesting debugging scenarios;
-and (iv) the sampling strategy. While these threats have been mitigated by the
-experimental setup, the remainder of this section discusses three additional
-ones.
+and (iv) the sampling strategy.  While the experimental setup attempts to
+mitigate against these threats, the reader must definitely take into account
+these limitations of the experiment and must understand that the collected data
+not support any further generalization.
+
+The second set of threats connects the experiment with reality or, more
+precisely, assess how close the connection is. To start with, it questions the
+realism of the rational programmer itself (sec.~\ref{sub:rational}), followed by
+the definition of ``interesting scenarios''
+(sec.~\ref{sec:threat:erasure-bias}).  The remaining questions address the
+accuracy and the cost of Transient blame, respectively (see
+secs.~\ref{sec:threat:transient} and~\ref{sec:threat:transient2}).
 
 %% -----------------------------------------------------------------------------
-\subsection{Threat: Is the Rational Programmer Realistic?}
+\subsection{Threat: Is the Rational Programmer Realistic?} \label{sub:rational}
 

### 2021-05-26 writing/discussion.tex
commenting out soem margin notes + removing duplicate threat to validity listing

@@
 representativeness of benchmarks; (ii) the relation between mutations and real
-programming mistakes; (iii) the definition of interesting debugging scenarios;
-and (iv) the sampling strategy.  While the experimental setup attempts to
+programming mistakes; and (iii) the sampling strategy.  While the experimental setup attempts to
 mitigate against these threats, the reader must definitely take into account

### 2021-05-26 writing/landscape.tex
commenting out soem margin notes + removing duplicate threat to validity listing

@@
 If an argument-type check fails, responsibility  lies with
-{\tt client}. After all either the type it ascribes to the argument is
+{\tt client}. After all, either the type it ascribes to the argument is
 wrong or the argument it produced clashes with the type. Due to 
@@
 {\it Transient\/} & no error                       & error, blaming the boundaries between            \\
-                  &                                & {\tt pack-lib}  and {\tt typed-pack-lib}, and \\
+                  &                                & {\tt pack-lib}  and {\tt typed-pack-lib}/ \\
                   &                                & {\tt  pack-lib} and {\tt crypto-pack-lib}     \\

### 2021-05-26 writing/rational-natural.tex
commenting out soem margin notes + removing duplicate threat to validity listing

--- 2021-05-23 writing/rational-natural.tex
@@
 
-\mf{Technically a Natural blame trail could be a Natural exception trail. Hmph? }
-\mf{I added a sentence below this definition and the next to address this issue.}
+%\mf{Technically a Natural blame trail could be a Natural exception trail. Hmph? }
+%\mf{I added a sentence below this definition and the next to address this issue.}
+%\cd{I revised a bit.}
 
@@
 \end{quote}
-Note how this definition interprets ``blame information'' broadly, as
-any information from a failed run-time check in a blame-oriented semantics.
-
-This definition rests on the two options for $\conf_{i+1} \setminus \conf_i$.
+Note how, in the absence of ``blame information'' in the narrow sense, 
+this definition interprets ``blame information'' broadly, as in
+any information from a failing run-time check.  
+In particular, the definition rests on the two options for $\conf_{i+1} \setminus \conf_i$.
 The first part reflects the rational programmer's use of blame to extend the
@@
 \end{quote}
-Using Natural exception trails it becomes possible to factor out
+Using Natural exception trails, it becomes possible to factor out
 ``blame information'' in the narrow sense from the broad one of the

### 2021-05-26 writing/discussion.tex
revision of general threats discussiom

@@
 
-The validity of these conclusions is threatened in two fundamentally distinct
-ways. The first sets of threats have been mentioned before: (i) the
-representativeness of benchmarks; (ii) the relation between mutations and real
+The validity of these conclusions is threatened in two distinct
+ways. The first set of threats concerns easy-to-tune parameters of the
+experiment and have been mentioned before: (i) the
+representativeness of benchmarks; (ii) the relation between the mutations and real
 programming mistakes; and (iii) the sampling strategy.  While the experimental setup attempts to
 mitigate against these threats, the reader must definitely take into account
-these limitations of the experiment and must understand that the collected data
-not support any further generalization.
+these limitations of the experiment when generalizing its conclusions.
 
-The second set of threats connects the experiment with reality or, more
+The second set of threats connects the fundamental ideas and infrastructure of the experiment with reality or, more
 precisely, assess how close the connection is. To start with, it questions the

### 2021-05-26 writing/discussion.tex
9.2 weakened

@@
 
-The validity of these conclusions is threatened in two distinct
-ways. The first set of threats concerns easy-to-tune parameters of the
-experiment and have been mentioned before: (i) the
-representativeness of benchmarks; (ii) the relation between the mutations and real
-programming mistakes; and (iii) the sampling strategy.  While the experimental setup attempts to
-mitigate against these threats, the reader must definitely take into account
-these limitations of the experiment when generalizing its conclusions.
+The validity of these conclusions is threatened in two distinct ways. The first
+set of threats concerns aspects of the experimental setup discussed in preceding
+sections: (i) the representativity of benchmarks; (ii) the relation between the
+mutations and real programming mistakes; and (iii) the sampling strategy.
+Although the experimental setup attempts to mitigate against these threats, the
+reader must definitely keep these limitations in mind when drawing conclusions.
 
-The second set of threats connects the fundamental ideas and infrastructure of the experiment with reality or, more
-precisely, assess how close the connection is. To start with, it questions the
-realism of the rational programmer itself (sec.~\ref{sub:rational}), followed by
-the definition of ``interesting scenarios''
-(sec.~\ref{sec:threat:erasure-bias}).  The remaining questions address the
-accuracy and the cost of Transient blame, respectively (see
-secs.~\ref{sec:threat:transient} and~\ref{sec:threat:transient2}).
+The second set of threats questions four rather different aspects. To start
+with, the realism of the rational programmer itself is questionable
+(sec.~\ref{sub:rational}) as is the definition of ``interesting scenarios''
+(sec.~\ref{sec:threat:erasure-bias}).  The remaining threats are about the
+accuracy and cost of Transient blame, respectively
+(secs.~\ref{sec:threat:transient} and~\ref{sec:threat:transient2}).
 

### 2021-05-27 writing/discussion.tex
9.2 grammar / wording

@@
 set of threats concerns aspects of the experimental setup discussed in preceding
-sections: (i) the representativity of benchmarks; (ii) the relation between the
+sections: (i) the representativeness of the benchmarks; (ii) the relation between the
 mutations and real programming mistakes; and (iii) the sampling strategy.
-Although the experimental setup attempts to mitigate against these threats, the
-reader must definitely keep these limitations in mind when drawing conclusions.
+Although the experimental setup attempts to mitigate these threats, the
+reader must keep these limitations in mind when drawing conclusions.
 
@@
 with, the realism of the rational programmer itself is questionable
-(sec.~\ref{sub:rational}) as is the definition of ``interesting scenarios''
+(sec.~\ref{sub:rational}), as is the definition of ``interesting scenarios''
 (sec.~\ref{sec:threat:erasure-bias}).  The remaining threats are about the

### 2021-05-30 writing/challenges.tex
Prettify changelog for sending, and some minor prose changes for it

@@
 type annotations. Their collection does not come with type-level mistakes in the
-code itself.}  The latter must represent mistakes that programmers accidentally
+code itself.}  The impedance mismatches must represent mistakes that programmers accidentally
 create and that the run-time checks of academic systems catch. In other words,

### 2021-05-30 writing/mutate-benchmarks.tex
Prettify changelog for sending, and some minor prose changes for it

@@
 
-Figure~\ref{table:benchmark-descriptions} describes the benchmark programs that
-meet all the criteria and come with a sizeable and complex dependency graphs.
-This additional filter is needed because Typed Racket demands type assignments
-for entire modules. For benchmarks with less than five modules, finding errors
+Figure~\ref{table:benchmark-descriptions} describes the ten benchmark programs that
+meet all the criteria, and furthermore come with the largest dependency graphs.
+This additional filter reflects that Typed Racket demands type assignments
+for entire modules, and so finding errors in benchmarks with small dependency graphs
 would almost be trivial.

### 2021-05-31 writing/discussion.tex
discussion: reword 'why transient lose blame'

@@
 
-An investigation of these empty blame cases reveals that Transient is not
-unsound but that Vitousek et al.'s Transient blame assignment loses track
-of the proper boundaries due to an incomplete initialization of the blame map.
-Despite these losses, the number of failures is a small fraction of the
-overall number of debugging scenarios. It is unlikely that any
-improvements to Vitousek et al.'s strategy would change the overall
-conclusion about Transient blame.
-
-{\bf Note} The scenarios where Transient returns no blame suggest two ways to
-improve its blame map.  First, entries in a blame map should point to {\em
-several\/} parent entries.  For example, if \texttt{f} receives bad input in the
-context of {\tt (filter f xs)}, the entry should point to both \texttt{filter}
-and \texttt{xs}. Second, the construction of entries must be guided by type-like
-specifications for primitives instead of special cases for built-in
-functions. Aliasing such function changes the blame map.
+An investigation of these empty blame cases reveals problems with tracking blame
+for higher-order functions, but conversely suggests three improvements for the
+Transient algorithm. To illustrate, consider the call \texttt{(filter f xs)}.
+The first suggestion is that the blame map should know that inputs to
+\texttt{f} may have come from the \texttt{xs} list; concretely, the blame-map
+entry for \texttt{f} should point to \texttt{xs} as a parent.
+Second, there should be two parents for \texttt{f} instead of one, because
+both \texttt{xs} and \texttt{filter} are responsible for sending correct input
+to \texttt{f}. Third, the blame map should work equally well in programs
+that rename \texttt{filter} or that replace the identifier with
+an expression. This third point suggests a need for type-like specifications
+that guide the construction of the blame map, instead of the identifier-based
+matching in Reticulated~\cite{vss-popl-2017} and Shallow Racket.
 

### 2021-05-31 writing/discussion.tex
minor fix

@@
 An investigation of these empty blame cases reveals problems with tracking blame
-for higher-order functions, but conversely suggests three improvements for the
+for higher-order functions and conversely suggests three improvements for the
 Transient algorithm. To illustrate, consider the call \texttt{(filter f xs)}.

### 2021-05-31 writing/discussion.tex
discussion: remove unneeded popl'17 cite
its purpose was to prove they do ID matching, but that's clear enough and
a real proof should offer a section/page number

@@
 that guide the construction of the blame map, instead of the identifier-based
-matching in Reticulated~\cite{vss-popl-2017} and Shallow Racket.
+matching in Reticulated and Shallow Racket.
 

### 2021-06-01 writing/challenges.tex
ben edits
- delete "see" as in "see sec. 1"; in most places, there's no need for this
  extra direction
- "mix-typed" -> "mixed-typed"
- move ",here ..." clause to directly follow target
- add oxford comma
- plural -> singular b/c pack-lib is the one and only responsible party
- use "--" for Wadler--Findler
- rational-stuff: **A successful** trail
- rational-transient: "and each" -> "and that each" to clarify the assignment
  says two things .... but it's a little strange to say that shallow checking
  is part of what the blame says
- "blame set" -> "blame sequence"
- benchmarks programs were

@@
 components. The Erasure semantics does not blame components {\it per se\/}, but
-it comes with an exception location and a stack trace, which implicitly suggests
+it comes with an exception location and a stack trace, which implicitly suggest
 fixes.  Each strategy triggers different reactions by the rational programmer
@@
 create and that the run-time checks of academic systems catch. In other words,
-the experiment calls for a collection of mistakes in mix-typed programs that is
+the experiment calls for a collection of mistakes in mixed-typed programs that is
 representative of those ``in the wild.''  Unfortunately no such collection

### 2021-06-01 writing/discussion.tex
ben edits
- delete "see" as in "see sec. 1"; in most places, there's no need for this
  extra direction
- "mix-typed" -> "mixed-typed"
- move ",here ..." clause to directly follow target
- add oxford comma
- plural -> singular b/c pack-lib is the one and only responsible party
- use "--" for Wadler--Findler
- rational-stuff: **A successful** trail
- rational-transient: "and each" -> "and that each" to clarify the assignment
  says two things .... but it's a little strange to say that shallow checking
  is part of what the blame says
- "blame set" -> "blame sequence"
- benchmarks programs were

@@
 the mathematical abstraction of a
-rational actor makes predictions also quite unreliable in some situations.
-\footnote{It has misled economists to focus on just mathematics, though
+rational actor makes predictions also quite unreliable in some situations.\footnote{It
+has misled economists to focus on just mathematics, though
 this problem is not relevant here---tongue firmly in cheek.}  Just like an

### 2021-06-01 writing/introduction.tex
ben edits
- delete "see" as in "see sec. 1"; in most places, there's no need for this
  extra direction
- "mix-typed" -> "mixed-typed"
- move ",here ..." clause to directly follow target
- add oxford comma
- plural -> singular b/c pack-lib is the one and only responsible party
- use "--" for Wadler--Findler
- rational-stuff: **A successful** trail
- rational-transient: "and each" -> "and that each" to clarify the assignment
  says two things .... but it's a little strange to say that shallow checking
  is part of what the blame says
- "blame set" -> "blame sequence"
- benchmarks programs were

@@
 Racket~\citep{tf-dls-2006,tf-popl-2008,tfffgksst-snapl-2017,tf-icfp-2010})
-come with sophisticated checking and blame assignment strategies (see
-sec.~\ref{sec:landscape}). Their academic creators embrace the idea that
+come with sophisticated checking and blame assignment strategies
+(sec.~\ref{sec:landscape}). Their academic creators embrace the idea that
 blame can help practicing programmers find impedance mismatches, that is,
@@
 The top-level innovation is the idea of a {\em rational programmer\/}, that is,
-a programmer that acts only in response to available information (see
-sec.~\ref{sec:why-rational}). In the case of an impedance mismatch, the
+a programmer that acts only in response to available information
+(sec.~\ref{sec:why-rational}). In the case of an impedance mismatch, the
 available information consists of the error message and the current state of the
@@
 gradual types and their blame assignment
-strategies in approximately 72,000 different scenarios\/} (see
-sec.~\ref{sec:results}): (1)~{\it Transient\/}, i.e. Reticulated's
+strategies in approximately 72,000 different scenarios\/}
+(sec.~\ref{sec:results}): (1)~{\it Transient\/}, i.e. Reticulated's
 inlined type assertions and collaborative tracking of typed/untyped boundaries~\citep{vss-popl-2017}; (2)~{\it
@@
 messages from the safety checks of the underlying language. The results
-(see sec.~\ref{sec:discussion}) are at least somewhat surprising.  In
+(sec.~\ref{sec:discussion}) are at least somewhat surprising.  In
 principle they validate the conjectures behind the work of theoreticians.

### 2021-06-01 writing/landscape.tex
ben edits
- delete "see" as in "see sec. 1"; in most places, there's no need for this
  extra direction
- "mix-typed" -> "mixed-typed"
- move ",here ..." clause to directly follow target
- add oxford comma
- plural -> singular b/c pack-lib is the one and only responsible party
- use "--" for Wadler--Findler
- rational-stuff: **A successful** trail
- rational-transient: "and each" -> "and that each" to clarify the assignment
  says two things .... but it's a little strange to say that shallow checking
  is part of what the blame says
- "blame set" -> "blame sequence"
- benchmarks programs were

@@
 As this analysis implies, if a return-type check fails, the problem is
-that the untyped module supplied a function, here {\tt
-pack-lib}, that is not a  match for the type ascribed by the typed module. 
+that the untyped module, here {\tt pack-lib}, supplied a function that is not a
+match for the type ascribed by the typed module.
 Hence either the type at the boundary between the two modules is wrong or, 
@@
 proxies, {\it Natural\/} can easily track the 
-boundary, type and responsible parties that correspond to each check. 
+boundary, type, and responsible parties that correspond to each check. 
 Thus, in the example of figure~\ref{fig:example}, as
@@
 the return type of \texttt{pack}, assuming that the type is correct,
-the responsible parties are the sources of the two
-boundary crossings, \texttt{pack-lib}. In
+the responsible party is the source of the two
+boundary crossings: \texttt{pack-lib}. In
 general though, Transient blames more than one boundary. In fact, the theoretical

### 2021-06-01 writing/mutate-benchmarks.tex
ben edits
- delete "see" as in "see sec. 1"; in most places, there's no need for this
  extra direction
- "mix-typed" -> "mixed-typed"
- move ",here ..." clause to directly follow target
- add oxford comma
- plural -> singular b/c pack-lib is the one and only responsible party
- use "--" for Wadler--Findler
- rational-stuff: **A successful** trail
- rational-transient: "and each" -> "and that each" to clarify the assignment
  says two things .... but it's a little strange to say that shallow checking
  is part of what the blame says
- "blame set" -> "blame sequence"
- benchmarks programs were

@@
 criteria. The benchmark suite consists of fully typed, correct programs, written
-by a number of different authors, who have maintained and evolved these programs
-over several years. The programs range widely in size, complexity, purpose,
+by a number of different authors who had maintained and evolved these programs
+over time. The programs range widely in size, complexity, purpose,
 origin and in programming style. They rely on many Typed Racket features:
@@
 for entire modules, and so finding errors in benchmarks with small dependency graphs
-would almost be trivial.
+would almost be trivial for the rational programmer.

### 2021-06-01 writing/mutate.tex
ben edits
- delete "see" as in "see sec. 1"; in most places, there's no need for this
  extra direction
- "mix-typed" -> "mixed-typed"
- move ",here ..." clause to directly follow target
- add oxford comma
- plural -> singular b/c pack-lib is the one and only responsible party
- use "--" for Wadler--Findler
- rational-stuff: **A successful** trail
- rational-transient: "and each" -> "and that each" to clarify the assignment
  says two things .... but it's a little strange to say that shallow checking
  is part of what the blame says
- "blame set" -> "blame sequence"
- benchmarks programs were

@@
 those into debugging scenarios. The process must start with a suitable
-collection of representative programs (see sec.~\ref{sub:mutate-benchmarks}).
+collection of representative programs (sec.~\ref{sub:mutate-benchmarks}).
 Since existing mutators do not generate useful mutants, the next step is to
-develop new mutators (see sec.~\ref{sub:mutate-mutators}) and to validate their
-suitability on the benchmarks (see sec.~\ref{sub:mutate-interesting}).
+develop new mutators (sec.~\ref{sub:mutate-mutators}) and to validate their
+suitability on the benchmarks (sec.~\ref{sub:mutate-interesting}).
 

### 2021-06-01 writing/rational-stuff.tex
ben edits
- delete "see" as in "see sec. 1"; in most places, there's no need for this
  extra direction
- "mix-typed" -> "mixed-typed"
- move ",here ..." clause to directly follow target
- add oxford comma
- plural -> singular b/c pack-lib is the one and only responsible party
- use "--" for Wadler--Findler
- rational-stuff: **A successful** trail
- rational-transient: "and each" -> "and that each" to clarify the assignment
  says two things .... but it's a little strange to say that shallow checking
  is part of what the blame says
- "blame set" -> "blame sequence"
- benchmarks programs were

@@
 equip with types. This choice constructs scenario $\conf_{i+1}$ from $\conf_{i}$
-and thus creates a trail. The trail's construction ends when it reaches a
+and thus lengthens the trail. A trail's construction ends successfully when it reaches a
 scenario that contains the mutated module because the type checker rejects its

### 2021-06-01 writing/rational-transient.tex
ben edits
- delete "see" as in "see sec. 1"; in most places, there's no need for this
  extra direction
- "mix-typed" -> "mixed-typed"
- move ",here ..." clause to directly follow target
- add oxford comma
- plural -> singular b/c pack-lib is the one and only responsible party
- use "--" for Wadler--Findler
- rational-stuff: **A successful** trail
- rational-transient: "and each" -> "and that each" to clarify the assignment
  says two things .... but it's a little strange to say that shallow checking
  is part of what the blame says
- "blame set" -> "blame sequence"
- benchmarks programs were

@@
 assignment says that the value witnessing the impedance mismatch may have
-crossed the boundaries between elements in the sequence, and each crossing
+crossed the boundaries between elements in the sequence, and that each crossing
 checked the value's type in a shallow manner.
@@
 This ambiguity in Transient blame raises the question of how the rational
-programmer should react when the language produces a blame set. Our answer
+programmer should react when the language produces a blame sequence. Our answer
 is that the rational programmer has at least two reasonable options. The first

### 2021-06-01 writing/rational.tex
ben edits
- delete "see" as in "see sec. 1"; in most places, there's no need for this
  extra direction
- "mix-typed" -> "mixed-typed"
- move ",here ..." clause to directly follow target
- add oxford comma
- plural -> singular b/c pack-lib is the one and only responsible party
- use "--" for Wadler--Findler
- rational-stuff: **A successful** trail
- rational-transient: "and each" -> "and that each" to clarify the assignment
  says two things .... but it's a little strange to say that shallow checking
  is part of what the blame says
- "blame set" -> "blame sequence"
- benchmarks programs were

@@
 lattice of type migration~\citep{tfgnvf-popl-2016}. The lattice describes the
-space in which the modes of the rational programmer search for bugs (see
-sec.~\ref{sub:stuff}).
+space in which the modes of the rational programmer search for bugs
+(sec.~\ref{sub:stuff}).
 %\mf{what does this sentence really mean? what does it contribute?}
@@
 them comparable even when they correspond to different semantics and process
-different kinds of information (see secs.~\ref{sub:natural}
+different kinds of information (secs.~\ref{sub:natural}
 through~\ref{sub:erasure}). Programmer effort relative to a fixed debugging
 scenario introduces another dimension along which the modes become may be
-compared (see sec.~\ref{sub:effort}). With these notions in place, it
-becomes possible to state the experimental questions and describes the process (see sec.~\ref{sub:experiment}).
+compared (sec.~\ref{sub:effort}). With these notions in place, it
+becomes possible to state the experimental questions and describes the process (sec.~\ref{sub:experiment}).
 

### 2021-06-01 writing/related.tex
fixing prose
ben edits
- delete "see" as in "see sec. 1"; in most places, there's no need for this
  extra direction
- "mix-typed" -> "mixed-typed"
- move ",here ..." clause to directly follow target
- add oxford comma
- plural -> singular b/c pack-lib is the one and only responsible party
- use "--" for Wadler--Findler
- rational-stuff: **A successful** trail
- rational-transient: "and each" -> "and that each" to clarify the assignment
  says two things .... but it's a little strange to say that shallow checking
  is part of what the blame says
- "blame set" -> "blame sequence"
- benchmarks p

@@
+\newcommand{\pyret}{\href{https://www.pyret.org}{pyret.org}}
 %% -----------------------------------------------------------------------------
@@
 not spell out the notion of the rational programmer, they present many of the
-basic ingredients.
-
-At a technical level, the two pieces of work differ in many ways.  First the
-experiment presented here involves three different modes of gradual typing,
-theirs a single notion of contract checking.  Second, this paper also
-contributes the idea of creating three comparable rational programmers, with
-several modes. Hence it can answer whether blame adds value to a gradual type
-system and which gradual type system it benefits the most, while theirs
+basic ingredients. At a technical level, the two pieces of work differ in many
+ways.  First the experiment presented here involves three different modes of
+gradual typing, theirs a single notion of contract checking.  Second, this paper
+also contributes the idea of creating three comparable rational programmers,
+with several modes. Hence it can answer whether blame adds value to a gradual
+type system and which gradual type system it benefits the most, while theirs
 establishes only a systematic relation between blame for contracts and
@@
 explained in section~\ref{sec:mutate}, almost none of~\citet{lksfd-popl-2020}'s
-mutators is useful in the context of gradual typing.
+mutators are useful in the context of gradual typing.
 
 The literature on gradual typing presents many semantics beyond the three used
-here and three additional blame strategies.  Pyret (pyret.org) assigns fixed-size
+here, and three additional blame strategies.  Pyret (\pyret{}) assigns fixed-size
 data types the Natural semantics and functions a Transient semantics. The
@@
 rat-oopsla-2017} assume that every value comes with a type tag and use tag
-checks to supervise the interactions between typed and untyped code.  The semantics~\citep{tlt-popl-2019,
-etg-icfp-19, tt-scp-20, tgt-popl-18, tt-sas-17} derived with the Abstracting
+checks to supervise the interactions between typed and untyped code.  The semantics derived with the Abstracting
 Gradual Typing technique~\citep{gct-popl-2016} are variants of Natural.
@@
 differs from Natural in the treatment of mutable data. It 
-associates every heap location with a type and permits only updates that 
-do not make the type less precise.  
+associates every heap location with a type and rejects updates that lower the
+precision of types.
 Among these semantics, only Amnesic, Nom, and Monotonic present interesting blame

### 2021-06-01 writing/why-rational-methodology.tex
ben edits
- delete "see" as in "see sec. 1"; in most places, there's no need for this
  extra direction
- "mix-typed" -> "mixed-typed"
- move ",here ..." clause to directly follow target
- add oxford comma
- plural -> singular b/c pack-lib is the one and only responsible party
- use "--" for Wadler--Findler
- rational-stuff: **A successful** trail
- rational-transient: "and each" -> "and that each" to clarify the assignment
  says two things .... but it's a little strange to say that shallow checking
  is part of what the blame says
- "blame set" -> "blame sequence"
- benchmarks programs were

@@
 locations at the boundary of typed and untyped code fragments.
-The Wadler-Findler slogan suggests that the source of the problem is concealed due to a lack of types, so
+The Wadler--Findler slogan suggests that the source of the problem is concealed due to a lack of types, so
 adding types to the untyped fragment should lead to the source of the impendance mismatch.

### 2021-06-01 writing/why-rational.tex
ben edits
- delete "see" as in "see sec. 1"; in most places, there's no need for this
  extra direction
- "mix-typed" -> "mixed-typed"
- move ",here ..." clause to directly follow target
- add oxford comma
- plural -> singular b/c pack-lib is the one and only responsible party
- use "--" for Wadler--Findler
- rational-stuff: **A successful** trail
- rational-transient: "and each" -> "and that each" to clarify the assignment
  says two things .... but it's a little strange to say that shallow checking
  is part of what the blame says
- "blame set" -> "blame sequence"
- benchmarks programs were

@@
 former to improve the latter. Specifically, the rational programmer translates
-the Wadler-Findler slogan into a debugging method, searching for the source of
+the Wadler--Findler slogan into a debugging method, searching for the source of
 the impedance mismatch by adding type annotations to some of the untyped parts
@@
 is also not to claim that equipping entire modules with types represents an
-always feasible approach\footnote{Adding types at the expression level, say, as
+always feasible approach.\footnote{Adding types at the expression level, say, as
 in TypeScript should be considered well within bounds.}  {\em But\/}, the

### 2021-06-22 writing/paper.tex
Add CRI grant to acknowledgements
ben edits
- delete "see" as in "see sec. 1"; in most places, there's no need for this
  extra direction
- "mix-typed" -> "mixed-typed"
- move ",here ..." clause to directly follow target
- add oxford comma
- plural -> singular b/c pack-lib is the one and only responsible party
- use "--" for Wadler--Findler
- rational-stuff: **A successful** trail
- rational-transient: "and each" -> "and that each" to clarify the assignment
  says two things .... but it's a little strange to say that shallow checking
  is part of what the blame says
- "blame set" -> "blame seq

@@
 \email{chrdimo@northwestern.edu}
-\renewcommand{\shortauthors}{Lazarek, Greenmanm Felleisen, Dimoulas}
+\renewcommand{\shortauthors}{Lazarek, Greenman, Felleisen, Dimoulas}
 
@@
 \input{def.tex}
+
+\pagenumbering{gobble}
 
@@
 
-%\begin{CCSXML}
-%<ccs2012>
-%<concept>
-%<concept_id></concept_id>
-%<concept_desc></concept_desc>
-%<concept_significance></concept_significance>
-%</concept>
-%<concept>
-%<concept_id></concept_id>
-%<concept_desc></concept_desc>
-%<concept_significance></concept_significance>
-%</concept>
-%</ccs2012>
-%\end{CCSXML}
+\begin{CCSXML}
+<ccs2012>
+   <concept>
+       <concept_id>10011007.10011074.10011099.10011693</concept_id>
+       <concept_desc>Software and its engineering~Empirical software validation</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10003752.10010124.10010138.10010140</concept_id>
+       <concept_desc>Theory of computation~Program specifications</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+ </ccs2012>
+\end{CCSXML}
 
-%\ccsdesc[]{}
-%\ccsdesc[]{}
+\ccsdesc[500]{Software and its engineering~Empirical software validation}
+\ccsdesc[500]{Theory of computation~Program specifications}
 
@@
 %% comma separated list
-%\keywords{}  %% \keywords are mandatory in final camera-ready submission
+\keywords{gradual typing, blame}  %% \keywords are mandatory in final camera-ready submission
 
@@
 Felleisen and Greenman were partly supported by NSF grant SHF
-1763922. Greenman also received support from a CRA-NSF Post-doctoral
-fellowship grant. Dimoulas and Felleisen wish to thank Max New for
+1763922.
+Greenman also received support from NSF grant 2030859 to the CRA for the
+\href{https://cifellows2020.org}{CIFellows} project.
+Dimoulas and Felleisen wish to thank Max New for
 extensive discussions of the early ideas in this paper. Thanks also to
@@
 various stages of this work.
+This paper was supported in part by the NSF grant CNS-1823244.
 \end{acks}

### 2021-06-23 writing/introduction.tex
Add footnote about artifacts
Then again -> Indeed
I've always read the phrase as contradicting or somehow nullifying the prior
statement. Seems like it can be used in both ways. Indeed seems clearer?
https://dictionary.cambridge.org/us/dictionary/english/then-again
https://idioms.thefreedictionary.com/then+again
https://ell.stackexchange.com/questions/29881/meaning-of-then-again

@@
 Instead, when language designers make decisions concerning this aspect, they seem
-to go one way or another without any scientific justification. Then again, the
+to go one way or another without any scientific justification. Indeed, the
 community has thus far failed to offer a method for evaluating blame assignment.
@@
 the approach of industrial systems, which forgo type checks and blame in favor of error
-messages from the safety checks of the underlying language. The results
+messages from the safety checks of the underlying language.
+The results\footnote{The full data set is too large to be hosted on the web, so it is freely available upon request -- along with the infrastructure to obtain it.}
 (sec.~\ref{sec:discussion}) are at least somewhat surprising.  In

### 2021-06-25 writing/introduction.tex
Add email to note about artifact

@@
 messages from the safety checks of the underlying language.
-The results\footnote{The full data set is too large to be hosted on the web, so it is freely available upon request -- along with the infrastructure to obtain it.}
+The results\footnote{The full data set is too large to be hosted on the web, so it is freely available upon request -- along with the infrastructure to obtain it.
+Please reach out at \texttt{lukas.lazarek@eecs.northwestern.edu}.}
 (sec.~\ref{sec:discussion}) are at least somewhat surprising.  In

### 2021-06-27 writing/discussion.tex
Fix grammar and typos

@@
 in five of the six modes (the two Transient blame modes behave the same).
-The reader may keep these scenarios for the following discussion of the numeric results.
+The reader may keep the illustration in mind for the following discussion of the numeric results.
 
@@
 has driven home this lesson but, fortunately, it has also shown that the types
-are in somewhat canonical.  The authors therefore conjecture that different
+are in practice somewhat canonical.  The authors therefore conjecture that different
 real-world programmers would often come up with equivalent type assignments

### 2021-06-27 writing/results.tex
results: Fix wrapfigure (10)
Fix wrapfigure placement
Fix grammar and typos
Fix old leftover discussion in results section

@@
 %% -----------------------------------------------------------------------------
+
+\begin{wrapfigure}{r}{.40\textwidth} \footnotesize
+  \vspace{-1.5em}
+  \includegraphics[width=0.38\textwidth]{./plots/success-bars}
+  \vspace{1em}
+  \begin{minipage}{0.35\textwidth} \raggedright
+%   The estimated percentage of scenarios for which each mode succeeds in locating the bug.
+   The upper bound margin of error is 0.02\%.
+  \end{minipage}
+\vspace{-2em}
+  \caption{Percentage rates of success} \label{fig:success-bars}
+\end{wrapfigure}
 
@@
 
-\
 \begin{figure} \footnotesize 
@@
 
-\begin{wrapfigure}{r}{.40\textwidth} \footnotesize 
-  \vspace{-1.5em}
-  \includegraphics[width=0.38\textwidth]{./plots/success-bars}
-  \vspace{1em}
-  \begin{minipage}{0.35\textwidth} \raggedright
-%   The estimated percentage of scenarios for which each mode succeeds in locating the bug.
-   The upper bound margin of error is 0.02\%.
-  \end{minipage}
-\vspace{-2em}
-  \caption{Percentage rates of success} \label{fig:success-bars}
-\end{wrapfigure}
-
 While most blame trail failures follow the above pattern, a few do
@@
 
-The figure answers questions $Q_1$, $Q_2$ and $Q_3$ affirmatively.  
+The figure answers questions $Q_1$, $Q_2$, and $Q_3$ affirmatively.
 In all three semantics, blame modes outperform their
@@
 
-Second, the exceptions of the Erasure semantics either help the rational
-programmer immediately or the rational programmer fails to debug a scenario
-altogether (in red).  This is expected; adding type annotations to an Erasure
-program does not change its behavior, so if the type checker does not reject the
-program with the new annotations, the rational programmer is stuck.  In other
-words, an exception from the run-time system has to point to the buggy module with the
-first try. Otherwise the rational programmer types an irrelevant module, runs
-the program again, and the same exception points again to the already typed
-module.
-
 Figure~\ref{fig:effort-comparisons} provides head-to-head comparisons of effort.
@@
 about how they compare in terms of success rates from figure~\ref{fig:avo-bars}.
-First, Natural blame never produces shorter trails than Natural exceptions, and
-sometimes (albeit rarely) produces slightly longer ones. Hence, the experiment
+First, Natural blame rarely produces shorter trails than Natural exceptions, and
+occasionally produces slightly longer ones. Hence, the experiment
 provides evidence that blame helps the rational programmer debug more scenarios
@@
 than both Transient blame modes, and sometimes the trails are significantly
-shorter. Finally, Transient's blame modes also share the characteristic with
+shorter. Finally, Transient's blame modes share the characteristic with
 Natural that blame sometimes produces longer trails than their corresponding

### 2021-06-27 writing/challenges.tex
Small grammar + typo fixes, and minor wording changes

@@
 types to its newest element.  If both modes are equally successful in
-locating an impedance mismatch, measuring the rational programmer's effort with each mode
+locating an impedance mismatch, measuring the rational programmer's debugging effort with each mode
 may answer which is the most effective.

### 2021-06-27 writing/landscape.tex
Small grammar + typo fixes, and minor wording changes

@@
 is imported into a typed module, the run-time system checks that it is a function
-and wraps it in a protective proxy, which in turn, enforces the type of the function
-result with run-time checks.  Analogously the run-time system wraps each exported
+and wraps it in a protective proxy, which in turn enforces the type of the function
+result with run-time checks.  Analogously, the run-time system wraps each exported
 function of a typed module such as {\tt crypto-pack} in a proxy that checks its
@@
 \texttt{pack} returns, the return-type check fails and Natural blames
-the boundary between \texttt{pack-lib} with {\tt typed-pack-lib} and
+the boundary of \texttt{pack-lib} with {\tt typed-pack-lib} and
 {\tt crypto-pack-lib}, respectively, for the two {\tt define}s in {\tt client}.
@@
 list, array, hash table etc.) is performed via a function call, the contents of
-a complex value is checked on a piecemeal basis.
+a complex value are checked on a piecemeal basis.
 

### 2021-06-27 writing/rational-transient.tex
Small grammar + typo fixes, and minor wording changes

@@
 The definition of \emph{Transient exception trails} is analogous to that for
-Natural. It is used to as a baseline for Transient-first and Transient-last
+Natural. It is used as a baseline for Transient-first and Transient-last
 blame.\footnote{A reader may wonder whether the rational programmer
-should not just equip {\em all\/} modules in the Transient blame set
-with types. Doing so may accelerate the search for the impedance
-mismatch but it would also impose a large migration cost.}
+should just equip {\em all\/} modules in the Transient blame set
+with types. That might accelerate the search for the impedance
+mismatch, but if so it would also impose a large migration cost.}

### 2021-06-27 writing/why-rational.tex
Small grammar + typo fixes, and minor wording changes

@@
 \end{verbatim}
-Lastly, the rational programmer assigns types to {\tt server} and to re-direct
+Lastly, the rational programmer assigns types to {\tt server} and re-directs
 the import of {\tt layer-typed} to {\tt (submod ".." server-typed)}. Now the

### 2021-06-27 writing/rational-transient.tex
Fix the transient first/last definition

@@
 \begin{tabular}{ll} %
-$\{\first{\mblame{\system, \conf_i}}\}$  & if $\conf_i$ produces blame \\ %
+$\{\first{\mblame{\system, \conf_i}}$ \lrdef $\last{\mblame{\system, \conf_i}}\}$  & if $\conf_i$ produces blame \\ %
 $\{\exception{\system, \conf_i}\}$     & otherwise \\ %

### 2021-06-29 writing/mutate-mutators.tex
Add Christos's touch ups

@@
 mutator wraps {\tt (not $\cdot$)} around the test of the conditional, the
-resulting mutant no longer type checks and, when run, this function eventually
+resulting mutant is ill-typed and, when run, this function eventually
 causes a run-time type check to signal an impedance mismatch.

### 2021-07-01 writing/mutate-benchmarks.tex
Fix grammar and typos
Change table "figures" to tables, with caption at top

@@
 over time. The programs range widely in size, complexity, purpose,
-origin and in programming style. They rely on many Typed Racket features:
+origin, and in programming style. They rely on many Typed Racket features:
 occurrence typing~\citep{tf-icfp-2010}, types for mutable and immutable data
@@
 
-Figure~\ref{table:benchmark-descriptions} describes the ten benchmark programs that
+Table~\ref{table:benchmark-descriptions} describes the ten benchmark programs that
 meet all the criteria, and furthermore come with the largest dependency graphs.

### 2021-07-01 writing/mutate-interesting.tex
Update changelog
Change table "figures" to tables, with caption at top

@@
 The goal of filtering for interesting mutants guided countless iterations of
-adding, removing, and refining mutators in figure~\ref{table:mutation-ops}.  For
-an illustrative example, Consider a candidate mutator that casts the tests of
+adding, removing, and refining mutators in table~\ref{table:mutation-ops}.  For
+an illustrative example, consider a candidate mutator that casts the tests of
 conditionals to the {\tt Any} type. Like the example explained at the end of the

### 2021-07-24 writing/why-rational-methodology.tex
Small grammar + typo fixes, and minor wording changes
Changes to satisfy the publisher

@@
 
+\vspace{1.5em}
+\textbf{How to Turn the Idea of the Rational Programmer into a Methodology.}
 Every time the rational programmer succeeds, it is validation for programming
-language researchers. It shows how their theorems and slogans help
+language researchers. It shows how their theorems, slogans, and tools help
 programmers. The example of blame assignment mechanisms makes this point
